### 2017-11-27 paper/manuscript.tex
Added manuscript latex document

% Use only LaTeX2e, calling the article.cls class and 12-point type.

\documentclass[11pt]{article}
\usepackage[round,semicolon]{natbib}
\usepackage[margin=1.4in]{geometry}
\usepackage{kpfonts}

\usepackage{seqsplit}
\usepackage{placeins}

\usepackage{newfloat}
\usepackage[labelfont=bf]{caption}
\usepackage{nameref}
\usepackage{rotating}
\usepackage{color}
\usepackage{float}

\setcounter{topnumber}{8}
\setcounter{bottomnumber}{8}
\setcounter{totalnumber}{8}
\renewcommand{\topfraction}{1}
\renewcommand{\bottomfraction}{1}
\renewcommand{\textfraction}{0}
\renewcommand{\floatpagefraction}{1}

\usepackage[font=small,labelfont=bf]{caption}

\usepackage{newfloat}
\DeclareFloatingEnvironment[name={Figure}]{suppfigure}
\renewcommand{\thesuppfigure}{S\arabic{suppfigure}}
\DeclareFloatingEnvironment[name={Table}]{supptable}
\renewcommand{\thesupptable}{S\arabic{supptable}}
\DeclareFloatingEnvironment[name={File}]{suppfile}
\renewcommand{\thesuppfile}{S\arabic{suppfile}}

\definecolor{darkblue}{rgb}{0, 0.0, 0.6}

\usepackage{hyperref}
\hypersetup{colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue}

\usepackage{seqsplit}

\usepackage{array}
\newcolumntype{R}[1]{>{\raggedright\arraybackslash}p{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\newcommand{\comment}[1]{{\color{red}[\textsl{#1}]}}

\usepackage{setspace}

\renewcommand{\topfraction}{1}
\renewcommand{\bottomfraction}{1}
\renewcommand{\textfraction}{0}
\renewcommand{\floatpagefraction}{1}

\title{TITLE}

\author
{Juhye M. Lee$^{1,3,4,\dagger}$, John Huddleston$^{2,4,\dagger}$, Michael B. Doud$^{1,3,4}$, Kathryn A. Hooper$^{1,4}$, Trevor Bedford,$^{2}$ and Jesse D. Bloom$^{1,3,5*}$\\
\\
\scriptsize{$^1$Basic Sciences and Computational Biology Divisions, Fred Hutchinson Cancer Research Center}\\
\scriptsize{$^2$Vaccine and Infectious Diseases Division, Fred Hutchinson Cancer Research Center} \\
\scriptsize{$^3$Department of Genome Sciences and $^4$Medical Scientist Training Program, University of Washington} \\
\scriptsize{$^5$Department of Molecular and Cellular Biology, University of Washington} \\
\scriptsize{Seattle, WA, USA} \\
\scriptsize{$^{\dagger}$These authors contributed equally} \\
\scriptsize{$^*$Correspondence: \href{jbloom@fredhutch.org}{jbloom@fredhutch.org}}
}

\date{}

\begin{document}

\maketitle
\onehalfspacing

\begin{abstract}
\comment{Some title ideas:
- Determination of lineage-specific mutational effects in hemagglutinin of seasonal H3N2 influenza virus evolution
- Experimental measurements of mutational effects to hemagglutinin reveal lineage-specific evolution of seasonal H3N2 influenza virus
- Understanding the effects of mutation to hemagglutinin in the context of seasonal H3N2 influenza virus evolution}
\end{abstract}

\end{document}

### 2017-11-27 paper/manuscript.tex
clean up title page and add title suggestion

@@
 
-\title{TITLE}
+\title{Deep mutational scanning of an H3 hemagglutinin can inform evolutionary forecasting of human H3N2 influenza virus}
 
-\author
-{Juhye M. Lee$^{1,3,4,\dagger}$, John Huddleston$^{2,4,\dagger}$, Michael B. Doud$^{1,3,4}$, Kathryn A. Hooper$^{1,4}$, Trevor Bedford,$^{2}$ and Jesse D. Bloom$^{1,3,5*}$\\
+\author{
+Juhye M. Lee$^{1,4,5,\dagger}$ \and 
+John Huddleston$^{2,6,\dagger}$ \and 
+Michael B. Doud$^{1,4,5}$ \and 
+Kathryn A. Hooper$^{1,6}$ \and
+Trevor Bedford,$^{2,3}$ \and 
+Jesse D. Bloom$^{1,3,4*}$
 \\
-\scriptsize{$^1$Basic Sciences and Computational Biology Divisions, Fred Hutchinson Cancer Research Center}\\
-\scriptsize{$^2$Vaccine and Infectious Diseases Division, Fred Hutchinson Cancer Research Center} \\
-\scriptsize{$^3$Department of Genome Sciences and $^4$Medical Scientist Training Program, University of Washington} \\
-\scriptsize{$^5$Department of Molecular and Cellular Biology, University of Washington} \\
-\scriptsize{Seattle, WA, USA} \\
+\scriptsize{$^1$Basic Sciences Division, $^2$Vaccine and Infectious Diseases Division, and $^3$Computational Biology Program,} \\
+\scriptsize{Fred Hutchinson Cancer Research Center, Seattle, WA, USA} \\
+\scriptsize{$^4$Department of Genome Sciences, $^5$Medical Scientist Training Program, and $^6$Molecular and Cellular Biology Program,} \\
+\scriptsize{University of Washington, Seattle, WA, USA} \\
 \scriptsize{$^{\dagger}$These authors contributed equally} \\

### 2017-12-03 paper/manuscript.tex
Add section headers to `.tex` file

@@
 \begin{abstract}
-\comment{Some title ideas:
-- Determination of lineage-specific mutational effects in hemagglutinin of seasonal H3N2 influenza virus evolution
-- Experimental measurements of mutational effects to hemagglutinin reveal lineage-specific evolution of seasonal H3N2 influenza virus
-- Understanding the effects of mutation to hemagglutinin in the context of seasonal H3N2 influenza virus evolution}
+Abstract text.
 \end{abstract}
 
+\section*{INTRODUCTION}
+
+\section*{RESULTS}
+\label{sec:results}
+
+\subsection*{Strategy for deep mutational scanning of an H3 hemagglutinin}
+
+\subsection*{H3 site-specific amino-acid preferences}
+
+\subsection*{Estimating mutational effects from an H3N2 phylogeny}
+
+\subsection*{Comparing H1 and H3 preferences}
+
+\section*{DISCUSSION}
+
+\clearpage
+\small
+
+\section*{METHODS}
+\label{sec:methods}
+\subsection*{HA numbering}
+
+\subsection*{Generation of HA codon mutant plasmid libraries}
+
+\subsection*{Generation and passaging of mutant viruses}
+
+\subsection*{Barcoded subamplicon sequencing}
+
+\subsection*{Analysis of deep sequencing data}
+
+\subsection*{Quantification of mutational effects and sequence preferences from an H3N2 phylogeny}
+
+\subsection*{Data availability and source code}
+Deep sequencing data are available from the Sequence Read Archive under BioSample accession \comment{add accession}.
+
+\subsection*{ACKNOWLEDGMENTS}
+We thank Sarah Hilton, Hugh Haddox...the Fred Hutch Genomics Core
+Funding...
+
 \end{document}

### 2017-12-05 paper/manuscript.tex
Start integrating figures into manuscript

@@
 
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/dms_overview/dms_overview.pdf}}
+\caption{\label{fig:dms_overview}
+{\bf Overview of deep mutational scanning experiments of H3 hemagglutinin}
+Figure caption text
+}
+\end{figure}
+
 \subsection*{H3 site-specific amino-acid preferences}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/mut_tolerance/entropy_heatmap.pdf}}
+\caption{\label{fig:mut_tolerance}
+{\bf The mutational tolerance of HA}
+Figure caption text
+}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}}
+\caption{\label{fig:logoplot}
+{\bf The site-specific amino-acid preferences of H3 hemagglutinin}
+Figure caption text
+}
+\end{figure}
 
@@
 
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/trunkvssidebranch/trunkvssidebranch.pdf}}
+\caption{\label{fig:trunkvssidebranch}
+{\bf The trunk of a human H3N2 phylogeny has higher mutational effects than those of side branches}
+Figure caption text
+}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/muteffects_sliding_window/muteffects_sliding_window.pdf}}
+\caption{\label{fig:muteffects_sliding_window}
+{\bf Sliding window analysis of mutational effects of trunk vs side branches.}
+Figure caption text
+}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}}
+\caption{\label{fig:WSN_trunkvssidebranch}
+{\bf Trunk vs side branch mutational effects calculated using WSN preferences}
+Figure caption text
+}
+\end{figure}
+
 \subsection*{Comparing H1 and H3 preferences}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/distance_distribution/distance_distribution.pdf}}
+\caption{\label{fig:distance_distribution}
+{\bf Distribution of preference shifts for protein homologs}
+Figure caption text
+}
+\end{figure}
 

### 2017-12-07 paper/manuscript.tex
Add draft of RMSD_heatmap figure
Small tweak to latex
Added captions to first set of figures

--- 2017-12-03 paper/manuscript.tex
@@
 \usepackage[round,semicolon]{natbib}
-\usepackage[margin=1.4in]{geometry}
+\usepackage[margin=1.25in]{geometry}
 \usepackage{kpfonts}
@@
 \subsection*{Strategy for deep mutational scanning of an H3 hemagglutinin}
+Previously, we measured the effect of all possible single amino-acid mutations to a highly lab-adapted H1 hemagglutinin from the A/WSN/1933 (H1N1) strain.
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/dms_overview/dms_overview.pdf}}
+\caption{\label{fig:dms_overview}
+{\bf Overview of deep mutational scanning experiments of H3 hemagglutinin.}
+(A) We used a helper virus approach previously described in Doud (2016) to generate the mutant virus libraries. 
+We transfected MDCK-SIAT1-TMPRSS2 cells with the mutant plasmid library carrying all possible amino-acid mutations to the A/Perth/16/2009 (H3N2) HA, in addition to protein expression plasmids encoding the viral ribonucleoprotein complex.
+After transfection, we infected the cells with an HA-deficient helper virus carrying all of the WSN/1933 influenza genes. 
+We then passage the initially generated pool of mutant viruses at low MOI to establish a genotype-phenotype linkage and select for functional HA variants. 
+(B) All of the experiments were completed in biological triplicate, starting from independent preps of the wildtype HA genes to create the mutant plasmids. 
+In addition, we passaged and deep sequenced library 3 in technical replicate, denoted 3-1 and 3-2, to gauge the amount of experimental noise within a single biological replicate.
+(C) Mutation frequencies of nonsynonymous, stop, and synonymous mutations for the mutant DNA, mutant virus, wildtype DNA, and wildtype virus samples. 
+There is selection against nonsynonymous and stop codons in the mutant viruses. 
+The percentages signify the frequency of stop codons remaining in the passaged mutant viruses relative to their starting frequency in the mutant plasmid libraries after correcting for the stop codon frequencies in the wildtype DNA and viruses.
+(D) The correlations and the Pearson correlation coefficient for the amino-acid preferences between each pair of replicates are shown. 
+The biological replicates are fairly well-correlated, as are the technical replicates, indicating some degree of bottlenecking of variants during the viral passage which can contribute to experimental noise.
+}
+\end{figure}
 
@@
 
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/mut_tolerance/entropy_heatmap.pdf}}
+\caption{\label{fig:mut_tolerance}
+{\bf Different regions of HA exhibit varying degrees of mutational tolerance.}
+Mutational tolerance as calculated by the Shannon entropy of a given site's amino-acid preferences are mapped onto the structure of the H3 trimer (PDB 4O5N, \comment{cite}) and the H1 trimer (PDB 1RVX, \comment{cite}), with both trimers in approximately the same orientation.
+The site entropies were calculated from the preferences measured in the Perth/2009 H3 (left panel) or the preferences measured in the WSN/1933 H1 (right panel). 
+Lighter shades of blue or red signify low mutational tolerance, while darker shades of blue or red signify high mutational tolerance. 
+For each HA, the structure on the left side colors the full HA trimer, while the structure on the right side colors only one of the monomers.
+The sialic acid receptor is shown as black sticks.
+The Perth/2009 H3 shows relatively high mutational tolerance in the stalk region compared to the head region. 
+The head region of the WSN/1933 H1 is mutationally tolerant compared to the relatively intolerant stalk region. 
+}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}}
+\caption{\label{fig:logoplot}
+{\bf The site-specific amino-acid preferences of H3 hemagglutinin.}
+This logoplot shows the site-specific amino-acid preferences for the averaged replicates rescaled by the stringency parameter (Table 1) estimated by phydms.
+The height of each letter is proportional to its preference at that site, and the preferences for all sites are normalized to sum to 1.
+The sites are in H3 numbering.
+The top overlay bar shows the relative solvent accessibility.
+The bottom overlay bar is colored by the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
+The letters directly above each logo indicate the wildtype amino acid at that site.
+}
+\end{figure}
+
 \subsection*{Estimating mutational effects from an H3N2 phylogeny}
 
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/trunkvssidebranch/trunkvssidebranch.pdf}}
+\caption{\label{fig:trunkvssidebranch}
+{\bf The trunk of a human H3N2 phylogeny has higher mutational effects than those of side branches.}
+(A) Phylogenetic tree of human H3N2 influenza virus from 1968-present. 
+\comment{We downloaded X sequences from the Influenza Virus Resource ?.... etc. inferred the tree, ancestral state reconstruction, visualized the tree. Mark Perth/2009 on the tree}
+To parse out trunk mutations from side branch mutations, we first defined a set of recent nodes sampled on or after Jan. 1, 2017, and traced these nodes back to their most recent common ancestor. 
+All branches ancestral to the MRCA of the recent nodes were defined as the trunk (shown in red), and all other branches were defined as side branches (shown in blue).
+(B) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel).
+To estimate significance, we performed 10,000 randomizations of the preferences to calculate the median difference in trunk vs side branch mutational effects, and counted how many of the randomizations exceeded the true difference in median trunk vs side branch mutational effects.
+The effects of trunk mutations are higher than side branch, internal and terminal node, mutations.
+}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/muteffects_sliding_window/muteffects_sliding_window.pdf}}
+\caption{\label{fig:muteffects_sliding_window}
+{\bf Trunk mutational effects are consistently higher than side branch mutational effects across time.}
+We calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
+The median log$_{2}$ mutational effect in a given window is shown as circles for trunk mutations and triangles for side branch mutations. 
+The shaded region demarcates the interquartile range of trunk and side branch mutational effects.
+The median trunk mutational effects are consistently higher than the median side branch mutational effects for all windows.
+}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}}
+\caption{\label{fig:WSN_trunkvssidebranch}
+{\bf The WSN/1933 H1 preferences do not reveal differences in trunk vs side branch mutational effects}
+(A) We calculated the log$_{2}$ mutational effects of the same set of trunk and side branch mutations from the inferred H3N2 phylogeny in Figure~\ref{fig:trunkvssidebranch} using the WSN/1933 H1 preferences.
+We again randomized the preferences for a total of 10,000 iterations and counted the number of randomizations that exceeded that true difference in median trunk vs side branch mutational effects to calculate a p-value.
+Using the WSN/1933 H1 preferences, there is not a significant difference in trunk vs side branch mutational effects.
+(B) We also performed the same sliding window analysis shown in Figure~\ref{fig:muteffects_sliding_window}, but using the WSN/1933 H1 preferences.
+There is not a distinct difference in trunk and side branch mutational effects.
+}
+\end{figure}
+
 \subsection*{Comparing H1 and H3 preferences}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/distance_distribution/distance_distribution.pdf}}
+\caption{\label{fig:distance_distribution}
+{\bf The HA homologs exhibit many large shifts in preference compared to shifts for other viral protein homologs}
+(A) A phylogenetic tree of the HA subtypes, with the two HA's, WSN/1933 H1 and Perth/2009 H3, for which we have measured amino-acid preferences denoted on the tree. 
+The WSN/1933 H1 and the Perth/2009 H3 share $\sim$42\% amino-acid identity.
+(B) The distribution of shifts in preference for various homolog pairs are shown.
+}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/RMSD_heatmap/RMSD_heatmap.pdf}}
+\caption{\label{fig:RMSD_heatmap}
+{\bf Shifts in preferences mapped onto the structure of HA}
+(A) The preference shifts as calculated by $RMSD_{corrected}$ between the two HA homologs is mapped onto the structure of HA (PDB 4O5N, citation). 
+The left structure shows the HA trimer, and the right structure colors one of the monomers. 
+The sialic acid receptor is in black sticks.
+Gray indicates little shifts in preference, while red indicates large shifts in preference.
+The top ten most shifted sites are shown in spheres on the monomer.
+(B)
+}
+\end{figure}
 
@@
 \subsection*{HA numbering}
+Unless otherwise indicated, all sites are in H3 numbering.
 
@@
 \subsection*{ACKNOWLEDGMENTS}
-We thank Sarah Hilton, Hugh Haddox...the Fred Hutch Genomics Core
+We thank Sarah Hilton, Hugh Haddox, Sidney Bell...the Fred Hutch Genomics Core...
 Funding...

### 2017-12-10 paper/manuscript.tex
Some small changes to figure captions

@@
 We transfected MDCK-SIAT1-TMPRSS2 cells with the mutant plasmid library carrying all possible amino-acid mutations to the A/Perth/16/2009 (H3N2) HA, in addition to protein expression plasmids encoding the viral ribonucleoprotein complex.
-After transfection, we infected the cells with an HA-deficient helper virus carrying all of the WSN/1933 influenza genes. 
-We then passage the initially generated pool of mutant viruses at low MOI to establish a genotype-phenotype linkage and select for functional HA variants. 
+After transfection, we infected the cells with an HA-deficient helper virus carrying all of the WSN/1933 (H1N1) influenza genes. 
+We then passaged the initially generated pool of mutant viruses at low MOI to establish a genotype-phenotype linkage and select for functional HA variants. 
 (B) All of the experiments were completed in biological triplicate, starting from independent preps of the wildtype HA genes to create the mutant plasmids. 
@@
 All branches ancestral to the MRCA of the recent nodes were defined as the trunk (shown in red), and all other branches were defined as side branches (shown in blue).
-(B) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel).
+(B) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
+The median log$_{2}$ mutational effect in a given window is shown as circles for trunk mutations and triangles for side branch mutations. 
+The shaded region demarcates the interquartile range of trunk and side branch mutational effects.
+The median trunk mutational effects are consistently higher than the median side branch mutational effects for all windows.
+(C) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel) are shown.
 To estimate significance, we performed 10,000 randomizations of the preferences to calculate the median difference in trunk vs side branch mutational effects, and counted how many of the randomizations exceeded the true difference in median trunk vs side branch mutational effects.
@@
 \begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/muteffects_sliding_window/muteffects_sliding_window.pdf}}
-\caption{\label{fig:muteffects_sliding_window}
-{\bf Trunk mutational effects are consistently higher than side branch mutational effects across time.}
-We calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
-The median log$_{2}$ mutational effect in a given window is shown as circles for trunk mutations and triangles for side branch mutations. 
-The shaded region demarcates the interquartile range of trunk and side branch mutational effects.
-The median trunk mutational effects are consistently higher than the median side branch mutational effects for all windows.
+\centerline{\includegraphics[width=\textwidth]{figs/sequence_preference/sequence_preference.pdf}}
+\caption{\label{fig:sequence_preference}
+{\bf The trunk exhibits higher sequence preference than side branches.}
+figure text
 }

### 2017-12-12 paper/manuscript.tex
Start working on first results section

@@
 \subsection*{Strategy for deep mutational scanning of an H3 hemagglutinin}
-Previously, we measured the effect of all possible single amino-acid mutations to a highly lab-adapted H1 hemagglutinin from the A/WSN/1933 (H1N1) strain.
+Previously, we measured the effect of all possible single amino-acid mutations to a highly lab-adapted H1 hemagglutinin from the A/WSN/1933 (H1N1) strain~\citep{thyagarajan2014inherent,doud2016accurate}.
+To better understand the effects of mutations to a hemagglutinin of more direct relevance to influenza strains currently circulating in the human population, we chose to study H3 HA from the A/Perth/16/2009 (H3N2) strain.
+This strain was the H3N2 component of the influenza vaccine from 2010-2012.
+\comment{mention serial passages and TC-adaptation mutations...}
+
+We mutagenized the Perth/2009 HA gene to create mutant plasmid HA libraries encompassing all 567 codons and harboring an average of $\sim$1.4 codon mutations per clone.
+We then used a helper-virus system previously established in~\cite{doud2016accurate} to bypass the bottlenecks associated with reverse genetics to generate complex mutant virus libraries from the mutant plasmids (Figure~\ref{fig:dms_overview}A).
+In order to maximize viral titers and further avoid bottlenecking the diversity of the initially generated mutant viruses, we used an HA-deficient helper virus carrying WSN/1933 internal and NA genes to rescue the mutant viruses, thereby pairing the mutant HA's with the WSN NA.
+Additionally, we used MDCK-SIAT1 cells constitutively expressing the TMPRSS2 protease, which is found endogenously in the human airway and facilitates HA cleavage and activation \comment{cite B\"{o}ttcher 2006 JVI, B\"{o}ttcher-Friebertsh\"{a}user 2010 JVI}.
+All of the experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B). 
+We also passaged and deep sequenced library 3 in technical replicate (denoted as library 3-1 and 3-2) to gauge to the amount of experimental noise occurring \textit{within} a single biological replicate.
+
+Deep sequencing shows...
 
@@
 (B) All of the experiments were completed in biological triplicate, starting from independent preps of the wildtype HA genes to create the mutant plasmids. 
-In addition, we passaged and deep sequenced library 3 in technical replicate, denoted 3-1 and 3-2, to gauge the amount of experimental noise within a single biological replicate.
+In addition, we passaged and deep sequenced library 3 in technical replicate, denoted 3-1 and 3-2, to estimate the amount of experimental noise within a single biological replicate.
 (C) Mutation frequencies of nonsynonymous, stop, and synonymous mutations for the mutant DNA, mutant virus, wildtype DNA, and wildtype virus samples. 
@@
 The left structure shows the HA trimer, and the right structure colors one of the monomers. 
-The sialic acid receptor is in black sticks.
+The sialic acid receptor is shown in black sticks.
 Gray indicates little shifts in preference, while red indicates large shifts in preference.
@@
 \subsection*{HA numbering}
-Unless otherwise indicated, all sites are in H3 numbering.
+Unless otherwise indicated, all sites are in H3 numbering, with the signal peptide in negative numbers, the HA1 domain in plain numbers, and the HA2 domain denoted with "(HA2)".
+
+\subsection*{Creation of MDCK-SIAT1-TMPRSS2 cell line}
+The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector...etc etc
+\comment{Need to look at Katie's notebooks for this...}
 
@@
 
+\subsection*{Inference of phylogenetic trees}
+
 \subsection*{Quantification of mutational effects and sequence preferences from an H3N2 phylogeny}
@@
 \subsection*{Data availability and source code}
-Deep sequencing data are available from the Sequence Read Archive under BioSample accession \comment{add accession}.
+Deep sequencing data are available from the Sequence Read Archive under BioSample accessions \comment{add accession}.
 
@@
 
+\bibliographystyle{mbe}
+\bibliography{references.bib}
+
+\clearpage
+\normalsize
+
 \end{document}

### 2017-12-13 paper/manuscript.tex
Some more progress on 1st results section and a bit of methods

@@
 
-Deep sequencing shows...
+Figure~\ref{fig:dms_overview}C shows the mutation frequencies of the mutant plasmids, mutant viruses, wildtype plasmids and wildtype viruses determined from deep sequencing. 
+There is selection against non-functional HA variants as revealed by the reduced mutation frequencies of the mutant viruses compared to their starting frequencies in the mutant plasmids.
+Specifically, stop codons were purged to 20-45\% of their original starting frequencies, after correcting for error rates from the wildtype controls.
+Although the majority of stop codons were purged, incomplete purging of stop codons is likely due to complementation of defective viral particles by infectious virions. 
+We also observed purging of most nonsynonymous mutations to 30-40\% of their initial starting frequencies after error correction, suggesting strong selection against deleterious HA variants.
+
+We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
+We inferred the amino-acid preferences at each site in the Perth/2009 HA using the method described in~\cite{bloom2015software} and implemented in the \texttt{dms\_tools2} software [\url{https://jbloomlab.github.io/dms_tools2/}].
+These preferences represent an estimate of the 567 sites $\times$ 20 amino acids $=$ 11340 experimental measurements and are normalized to sum to one at each site.
+The correlations of the amino-acid preferences between each pair of replicates is shown in Figure~\ref{fig:dms_overview}D.
+The biological replicates are fairly well-correlated, with a Pearson's $R$ ranging from 0.69 to 0.78. 
+Replicate 1 exhibits the least amount of correlation with the other biological replicates, consistent with the observation that this replicate showed the weakest selection against stop and nonsynonymous mutations and might therefore be subject to more experimental noise.
+Of note, the two technical replicates 3-1 and 3-2 were only slightly more reproducible than that between biological replicates.
+This suggests that bottlenecking of the virus library during the low MOI passage contributes to much of the noise observed in our experiments, as we are only able to passage a finite number of viral particles.
 
@@
 \subsection*{H3 site-specific amino-acid preferences}
+How well do the Perth/2009 HA preferences inferred from experimental measurements describe the evolution of H3N2 influenza virus in nature?
+\comment{phydms analysis here}
 
@@
 The WSN/1933 H1 and the Perth/2009 H3 share $\sim$42\% amino-acid identity.
-(B) The distribution of shifts in preference for various homolog pairs are shown.
+(B) The correlation for the amino-acid preferences for replicates both within and between the two HA homologs.
+(C) The distribution of shifts in preference for various homolog pairs are shown.
 }
@@
 \subsection*{HA numbering}
-Unless otherwise indicated, all sites are in H3 numbering, with the signal peptide in negative numbers, the HA1 domain in plain numbers, and the HA2 domain denoted with "(HA2)".
+Unless otherwise indicated, all sites are in H3 numbering, with the signal peptide in negative numbers, the HA1 subunit in plain numbers, and the HA2 subunit denoted with "(HA2)".
+The conversion between sequential numbering of the A/Perth/16/2009 HA and H3 numbering was performed using an HA numbering Python script (available at \url{https://github.com/jbloomlab/HA_numbering}).
 
 \subsection*{Creation of MDCK-SIAT1-TMPRSS2 cell line}
-The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector...etc etc
+The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter and attached to mCherry through an IRES...etc etc
 \comment{Need to look at Katie's notebooks for this...}
@@
 \subsection*{Generation of HA codon mutant plasmid libraries}
+Recombinant A/Perth/16/2009 (HA, NA) $\times$ A/Puerto Rico/8/1934 influenza virus, NIB-64, NR-41803 was ordered from BEI Resources, NIAID, NIH. 
+Bulk RNA from the viral sample was extracted using the QIAamp Viral RNA Mini Kit (QIAGEN) according to manufacturer's instructions.
+The Perth/2009 HA and NA genes were then reverse transcribed, PCR amplified, and cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2 \comment{cite?} plasmid backbones.
+
+The codon-mutant libraries were generated using a PCR-based approach described in~\cite{dingens2017comprehensive}.
 
 \subsection*{Generation and passaging of mutant viruses}
+The mutant virus libraries were generated and passaged using the approach described in~\cite{doud2016accurate} with several modifications.
 
@@
 \subsection*{Data availability and source code}
-Deep sequencing data are available from the Sequence Read Archive under BioSample accessions \comment{add accession}.
+Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
+Computer code used to analyze the data and produce the results in the paper are in...
 

### 2017-12-15 paper/manuscript.tex
Added phydms table to results section, changed fig 2 & 3 order

@@
 How well do the Perth/2009 HA preferences inferred from experimental measurements describe the evolution of H3N2 influenza virus in nature?
-\comment{phydms analysis here}
+This question can be addressed by evaluating how well experimentally informed codon substitution models (ExpCM's) constructed from our laboratory measurements improve phylogenetic fit of H3N2 evolution~\citep{hilton2017phydms}.
+The results shown in Table~\ref{tab:phydms} show that the ExpCM outperforms conventional substitution models in describing the evolution of human H3N2 HA. 
+The ExpCM also optimizes a stringency parameter ($\beta$) for the preferences to more closely reflect the strength of selection in nature.
+The stringency parameter in the ExpCM is equal to 2.44, which indicates that although the same amino acids are preferred, the strength of selection is more stringent in nature than in our experiments.
+Figure~\ref{fig:logoplot} shows a logo plot of the Perth/2009 HA amino-acid preferences rescaled by this stringency parameter.
+
+\begin{table}
+\centering
+\begin{tabular}{cccc}
+\hline
+\bf{Model} & \bf{$\Delta$AIC} & \bf{Log Likelihood} & \bf{Parameters} \\ \hline
+ExpCM & 0.0 & -8439.33 & $\beta = 2.44, \kappa = 5.78, \omega = 0.91$ \\
+Goldman-Yang M5 & 2166.06 & -9516.36 & $\omega_{\alpha} = 0.30, \omega_{\beta} = 0.84, \kappa = 5.10$ \\
+ExpCM, averaged across sites & 2504.18 & -9691.42 & $\beta = 0.68, \kappa = 5.58, \omega = 0.32$ \\
+Goldman-Yang M0 & 2607.92 & -9738.29 & $\kappa = 5.05, \omega = 0.31$ \\
+\hline
+\end{tabular}
+\caption{\label{tab:phydms}
+{\bf The site-specific amino-acid preferences are informative for describing human H3N2 evolution in nature.}
+We implemented several codon substitution models for phylogenetic fitting of an alignment of human H3N2 HA sequences. 
+The maximum likelihood values for each model were compared using the Akaike information criteria ($\Delta$AIC)~\citep{posada2004model}.
+An experimentally-informed codon substitution model (ExpCM) built from the preferences averaged across all replicates performs better than conventional substitution models, specifically the M0 and M5 models in~\cite{yang2000codon}.
+A non-site-specific ExpCM informed by preferences averaged across all sites does not perform as well as the M5 model, but still has a higher log likelihood than M0.
+The optimized parameters for each model is also shown.
+}
+\end{table}
+
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}}
+\caption{\label{fig:logoplot}
+{\bf The site-specific amino-acid preferences of H3 hemagglutinin.}
+This logoplot shows the site-specific amino-acid preferences for the averaged replicates rescaled by the stringency parameter (Table 1) estimated by phydms.
+The height of each letter is proportional to its preference at that site, and the preferences for all sites are normalized to sum to 1.
+The sites are in H3 numbering.
+The top overlay bar shows the relative solvent accessibility.
+The bottom overlay bar is colored by the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
+The letters directly above each logo indicate the wildtype amino acid at that site.
+}
+\end{figure}
 
@@
 The head region of the WSN/1933 H1 is mutationally tolerant compared to the relatively intolerant stalk region. 
-}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}}
-\caption{\label{fig:logoplot}
-{\bf The site-specific amino-acid preferences of H3 hemagglutinin.}
-This logoplot shows the site-specific amino-acid preferences for the averaged replicates rescaled by the stringency parameter (Table 1) estimated by phydms.
-The height of each letter is proportional to its preference at that site, and the preferences for all sites are normalized to sum to 1.
-The sites are in H3 numbering.
-The top overlay bar shows the relative solvent accessibility.
-The bottom overlay bar is colored by the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
-The letters directly above each logo indicate the wildtype amino acid at that site.
 }

### 2017-12-17 paper/manuscript.tex
Some progress on section 2 of results

@@
 
+A closer examination of the logo plot reveals that the preferences generally agree with existing knowledge about HA's biochemistry.
+For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\citep{waterfield1981disulphide} possess high preference for cysteine.
+At residues involved in receptor binding, there are strong preferences for the amino acids at sites Y98, D190, W153, and S228.
+A positively charged amino acid at site 329 is important for cleavage activation of the HA0 precursor, and indeed this site exhibits a high preference for arginine~\citep{kido1992isolation, stech2005new}.
+However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine. 
+This codon is part of the signal peptide and is cleaved from the mature HA protein.
+One possible explanation for why we do not see a strong preference for Met at this site is due to alternative translation initiation occurring at a downstream or upstream start site, as has been described in \comment{need to list citations here, inc.
+Girard 2011 Infection Genetics and Evolution, Chen 2001 Nat Med}.
+% No upstream or downstream "ATG"'s are present.
+I did find an in-frame "ATC" one codon upstream of the canonical start site (in the non-coding region), and interestingly, all Perth/2009 HA sequences deposited in GenBank have an "ATT" in this non-coding position, including Seema's Perth HA.
+% I6M(HA2) mutant slightly increases the fusion pH (Daniels 1985 Cell)
+% We see high preference for T,S,Q over the wildtype Ile at site 226.
+Although Q226 in an H3 background is characterized as having preference for avian-type receptors, most of these studies were done in older H3 strains such as Aichi.
+Wu 2017 Cell Host Microbe has shown extensive epistasis in the RBS, specifically in the 220-loop, so perhaps the amino acids tolerated in this loop have shifted over time.
+Indeed the wildtype amino acids at 225 and 226 are different in Perth compared to HK68 and Aichi.
+
+We next sought to investigate the inherent mutational tolerance of the Perth/2009 HA. 
+Figure~\ref{fig:mut_tolerance} shows the mutational tolerance as calculated from the rescaled Perth/2009 H3 preferences and the rescaled WSN/1933 H1 preferences mapped onto the HA crystal structures.
+Interestingly, the Perth/2009 H3 stalk is relatively mutationally tolerant compared to the tolerance of the globular head domain. 
+In particular, the shorter $\alpha$-helix (helix A) of the stalk domain, antigenic site C, and the most distal portion of the globular head near antigenic site B are all highly mutationally tolerant.
+On the other hand, the head domain of the WSN/1933 H1 is more mutationally tolerant relative to its stalk domain.
+
 \begin{figure}
@@
 {\bf Different regions of HA exhibit varying degrees of mutational tolerance.}
-Mutational tolerance as calculated by the Shannon entropy of a given site's amino-acid preferences are mapped onto the structure of the H3 trimer (PDB 4O5N, \comment{cite}) and the H1 trimer (PDB 1RVX, \comment{cite}), with both trimers in approximately the same orientation.
+Mutational tolerance as calculated by the Shannon entropy of a given site's amino-acid preferences are mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}) and the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}), with both trimers in approximately the same orientation.
 The site entropies were calculated from the preferences measured in the Perth/2009 H3 (left panel) or the preferences measured in the WSN/1933 H1 (right panel). 

### 2017-12-18 paper/manuscript.tex
More progress on results sections 1 and 2

@@
 Previously, we measured the effect of all possible single amino-acid mutations to a highly lab-adapted H1 hemagglutinin from the A/WSN/1933 (H1N1) strain~\citep{thyagarajan2014inherent,doud2016accurate}.
-To better understand the effects of mutations to a hemagglutinin of more direct relevance to influenza strains currently circulating in the human population, we chose to study H3 HA from the A/Perth/16/2009 (H3N2) strain.
+To examine the effects of mutations to a hemagglutinin of more direct relevance to influenza strains currently circulating in the human population, we chose to study H3 HA from the A/Perth/16/2009 (H3N2) strain.
 This strain was the H3N2 component of the influenza vaccine from 2010-2012.
-\comment{mention serial passages and TC-adaptation mutations...}
+We first rescued virus carrying the wildtype Perth/2009 HA and NA and the WSN internal genes by reverse genetics, and serially passaged the virus in cell culture a total of six times.
+%This passaging was completed to allow the virus to adapt to growth in cell culture and to find cell-culture adaptation mutations that emerged during the course of passaging.
+Two mutations, G78D and T212I, arose together from passaging, and the G78D-T212I Perth/2009 HA mutant supported viral growth to $\sim$15- to 20-fold higher titers than the wildtype HA \comment{(supplementary data?)}.
+We decided to perform all deep mutational scanning experiments in the background of the Perth/2009 HA carrying these two mutations to support higher viral growth and to limit these cell culture adaptation mutations from swamping out the signal from other mutations.
 
@@
 This question can be addressed by evaluating how well experimentally informed codon substitution models (ExpCM's) constructed from our laboratory measurements improve phylogenetic fit of H3N2 evolution~\citep{hilton2017phydms}.
-The results shown in Table~\ref{tab:phydms} show that the ExpCM outperforms conventional substitution models in describing the evolution of human H3N2 HA. 
+The results in Table~\ref{tab:phydms} show that the ExpCM outperforms conventional substitution models in describing the evolution of human H3N2 HA. 
 The ExpCM also optimizes a stringency parameter ($\beta$) for the preferences to more closely reflect the strength of selection in nature.
@@
 A non-site-specific ExpCM informed by preferences averaged across all sites does not perform as well as the M5 model, but still has a higher log likelihood than M0.
-The optimized parameters for each model is also shown.
+The optimized parameters for each model are also shown.
 }
@@
 In particular, the shorter $\alpha$-helix (helix A) of the stalk domain, antigenic site C, and the most distal portion of the globular head near antigenic site B are all highly mutationally tolerant.
+% may add more details here
 On the other hand, the head domain of the WSN/1933 H1 is more mutationally tolerant relative to its stalk domain.
+The receptor binding pockets of both HA's are relatively mutationally intolerant, although the residues surrounding the receptor binding pocket are fairly tolerant of mutations.
+These findings are consistent with the high functional constraint of the receptor binding pocket, while the mutability of sites surrounding the pocket that are under strong immune pressure may contribute to antigenic evolution~\citep{wilson1981structure, wiley1981structural}.
 
@@
 Mutational tolerance as calculated by the Shannon entropy of a given site's amino-acid preferences are mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}) and the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}), with both trimers in approximately the same orientation.
-The site entropies were calculated from the preferences measured in the Perth/2009 H3 (left panel) or the preferences measured in the WSN/1933 H1 (right panel). 
+The site entropies were calculated from the preferences measured in the Perth/2009 H3 (left panel) from this study, or the preferences measured in the WSN/1933 H1 (right panel) from~\cite{doud2016accurate}.
 Lighter shades of blue or red signify low mutational tolerance, while darker shades of blue or red signify high mutational tolerance. 
@@
 Using the WSN/1933 H1 preferences, there is not a significant difference in trunk vs side branch mutational effects.
-(B) We also performed the same sliding window analysis shown in Figure~\ref{fig:muteffects_sliding_window}, but using the WSN/1933 H1 preferences.
+(B) We also performed the same sliding window analysis shown in Figure~\ref{fig:trunkvssidebranch}B, but using the WSN/1933 H1 preferences.
 There is not a distinct difference in trunk and side branch mutational effects.

### 2017-12-20 paper/manuscript.tex
Start working on results section 4

@@
 Kathryn A. Hooper$^{1,6}$ \and
-Trevor Bedford,$^{2,3}$ \and 
+Trevor Bedford$^{2,3*}$ and \and 
 Jesse D. Bloom$^{1,3,4*}$
@@
 \scriptsize{$^{\dagger}$These authors contributed equally} \\
-\scriptsize{$^*$Correspondence: \href{jbloom@fredhutch.org}{jbloom@fredhutch.org}}
+\scriptsize{$^*$Correspondence: \href{trevor@bedford.io}{trevor@bedford.io}, \href{jbloom@fredhutch.org}{jbloom@fredhutch.org}}
 }
@@
 \section*{INTRODUCTION}
+Very rough outline:
+\begin{itemize}
+\item Mutations are rampant in the evolution of human influenza virus. Seasonal H3N2 influenza virus in particular rapidly accumulates mutations.
+\item The evolution of H3N2 is also characterized by clade competition and population turnover.
+\item There have been efforts to predict evolutionary success.
+\item Mutations that contribute to antigenic evolution largely determine strain success.
+\item However, mutations in HA that impact viral growth may also influence strain success.
+\item We need to understand the functional impact of mutations in HA.
+\item Previously, we measured the effect of all possible single amino-acid mutations to an H1 hemagglutinin from the A/WSN/1933 (H1N1) strain~\citep{thyagarajan2014inherent,doud2016accurate}. 
+\item However, this is a highly lab-adapted strain, and the measurements in this strain may not be particularly relevant for studying mutational processes of more contemporaneous strains circulating in the human population.
+\item We chose to study the Perth/2009 H3 HA.
+\item This also enabled a comparison of how the preferences have shifted for two diverged HA's.
+\end{itemize}
 
@@
 
-\subsection*{Strategy for deep mutational scanning of an H3 hemagglutinin}
-Previously, we measured the effect of all possible single amino-acid mutations to a highly lab-adapted H1 hemagglutinin from the A/WSN/1933 (H1N1) strain~\citep{thyagarajan2014inherent,doud2016accurate}.
-To examine the effects of mutations to a hemagglutinin of more direct relevance to influenza strains currently circulating in the human population, we chose to study H3 HA from the A/Perth/16/2009 (H3N2) strain.
+\subsection*{Deep mutational scanning of HA from a recent strain of human H3N2 influenza virus}
+We performed a deep mutational scan of H3 HA from the A/Perth/16/2009 (H3N2) strain. 
 This strain was the H3N2 component of the influenza vaccine from 2010-2012.
-We first rescued virus carrying the wildtype Perth/2009 HA and NA and the WSN internal genes by reverse genetics, and serially passaged the virus in cell culture a total of six times.
+The HA gene we used in our experiments carries two mutations, G78D and T212I, relative to the Perth/2009 HA sequence deposited in GenBank.
+The G78D mutation has appeared sporadically at low frequency in natural H3N2 sequences. 
+Although the T212I mutation is not appreciable in nature \comment{reword}, the T212A mutation rose to fixation in the human population in $\sim$2011.
+We found that the G78D-T212I Perth/2009 HA mutant supported viral growth to $\sim$15- to 20-fold higher titers than the wildtype HA.
+%We first generated virus carrying the wildtype Perth/2009 HA and NA and the WSN internal genes by reverse genetics, and serially passaged the virus in cell culture a total of six times.
 %This passaging was completed to allow the virus to adapt to growth in cell culture and to find cell-culture adaptation mutations that emerged during the course of passaging.
-Two mutations, G78D and T212I, arose together from passaging, and the G78D-T212I Perth/2009 HA mutant supported viral growth to $\sim$15- to 20-fold higher titers than the wildtype HA \comment{(supplementary data?)}.
-We decided to perform all deep mutational scanning experiments in the background of the Perth/2009 HA carrying these two mutations to support higher viral growth and to limit these cell culture adaptation mutations from swamping out the signal from other mutations.
+%Two mutations, G78D and T212I, arose together from passaging.
+We therefore decided to perform all deep mutational scanning experiments in the background of the Perth/2009 HA carrying these two mutations to support higher viral growth.
 
-We mutagenized the Perth/2009 HA gene to create mutant plasmid HA libraries encompassing all 567 codons and harboring an average of $\sim$1.4 codon mutations per clone.
-We then used a helper-virus system previously established in~\cite{doud2016accurate} to bypass the bottlenecks associated with reverse genetics to generate complex mutant virus libraries from the mutant plasmids (Figure~\ref{fig:dms_overview}A).
-In order to maximize viral titers and further avoid bottlenecking the diversity of the initially generated mutant viruses, we used an HA-deficient helper virus carrying WSN/1933 internal and NA genes to rescue the mutant viruses, thereby pairing the mutant HA's with the WSN NA.
-Additionally, we used MDCK-SIAT1 cells constitutively expressing the TMPRSS2 protease, which is found endogenously in the human airway and facilitates HA cleavage and activation \comment{cite B\"{o}ttcher 2006 JVI, B\"{o}ttcher-Friebertsh\"{a}user 2010 JVI}.
+We mutagenized the HA gene to create mutant plasmid libraries encompassing all 567 codons and harboring an average of $\sim$1.4 codon mutations per clone.
+We then generated mutant virus libraries from the mutant plasmids using a helper virus system that enables the efficient generation of complex virus libraries~\citep{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
+In order to reduce biosafety concerns and to maximize viral titers, we used an HA-deficient helper virus carrying WSN/1933 internal and NA genes to rescue the mutant viruses.
+Additionally, we used MDCK-SIAT1 cells constitutively expressing the TMPRSS2 protease, which facilitates HA cleavage and activation in the human airway \comment{cite B\"{o}ttcher 2006 JVI, B\"{o}ttcher-Friebertsh\"{a}user 2010 JVI}.
 All of the experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B). 
@@
 \caption{\label{fig:dms_overview}
-{\bf Overview of deep mutational scanning experiments of H3 hemagglutinin.}
+{\bf Deep mutational scanning experiments of H3 HA.}
 (A) We used a helper virus approach previously described in Doud (2016) to generate the mutant virus libraries. 
@@
 
-\subsection*{H3 site-specific amino-acid preferences}
+\subsection*{The HA preferences improve phylogenetic fit of natural H3 sequences}
 How well do the Perth/2009 HA preferences inferred from experimental measurements describe the evolution of H3N2 influenza virus in nature?
@@
 
+\begin{figure}
+\centerline{\includegraphics[width=\textwidth]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}}
+\caption{\label{fig:logoplot}
+{\bf The site-specific amino-acid preferences of the Perth/2009 HA.}
+This logoplot shows the site-specific amino-acid preferences for the averaged replicates rescaled by the stringency parameter (Table 1) estimated by phydms.
+The height of each letter is proportional to its preference at that site, and the preferences for all sites are normalized to sum to 1.
+The sites are in H3 numbering.
+The top overlay bar shows the relative solvent accessibility.
+The bottom overlay bar is colored by the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
+The letters directly above each logo indicate the wildtype amino acid at that site.
+}
+\end{figure}
+
+\subsection*{The H3 stalk is relatively mutationally tolerant}
 We next sought to investigate the inherent mutational tolerance of the Perth/2009 HA. 
@@
 These findings are consistent with the high functional constraint of the receptor binding pocket, while the mutability of sites surrounding the pocket that are under strong immune pressure may contribute to antigenic evolution~\citep{wilson1981structure, wiley1981structural}.
-
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}}
-\caption{\label{fig:logoplot}
-{\bf The site-specific amino-acid preferences of H3 hemagglutinin.}
-This logoplot shows the site-specific amino-acid preferences for the averaged replicates rescaled by the stringency parameter (Table 1) estimated by phydms.
-The height of each letter is proportional to its preference at that site, and the preferences for all sites are normalized to sum to 1.
-The sites are in H3 numbering.
-The top overlay bar shows the relative solvent accessibility.
-The bottom overlay bar is colored by the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
-The letters directly above each logo indicate the wildtype amino acid at that site.
-}
-\end{figure}
 
@@
 
-\subsection*{Estimating mutational effects from an H3N2 phylogeny}
+\subsection*{The H3 preferences reveal differences in seasonal influenza virus lineages}
+The human H3N2 phylogeny has a characteristic ladder-like structure, with a persistent trunk lineage and side branches that rapidly die out.
+Do the preferences reveal inherent differences between lineages of human H3N2 influenza virus that have succeeded and those that have died out?
+To examine this, we calculated the effect of trunk and side branch mutations from an H3N2 phylogeny, shown in Figure~\ref{fig:trunkvssidebranch}A, from 1968-2012 using our preference dataset. 
+Figure~\ref{fig:trunkvssidebranch}B shows the median log$_{2}$ mutational effects and the interquartile ranges for trunk and side branch mutations in five-year windows for every year from 1968-2012. 
+On average, trunk mutations tend to be towards more preferred amino acids compared to side branch mutations, and this was true for all windows.
+Next we calculated the effect of all trunk and side branch mutations from the entirety of the phylogenetic tree.
+We found that on average, trunk mutations are significantly more favorable than side branch mutations (Figure~\ref{fig:trunkvssidebranch}C).
+Because a number of tip nodes are likely from egg- or cell-passaged isolates and our experiments were performed in cell culture, we were curious to see if our measurements would lead to higher estimations of mutational effects for terminal nodes.
+We found that terminal node mutations are on average towards less preferred amino acids than internal node mutations, and that both internal and terminal node mutations are significantly less favorable than those on the trunk.
+Therefore, strains that have accumulated mutations that we measured in our experiments to be unpreferred generally die out in nature.
+These findings demonstrate the importance of the functional impacts of HA mutations in determining lineage success.
 
@@
 \caption{\label{fig:trunkvssidebranch}
-{\bf The trunk of a human H3N2 phylogeny has higher mutational effects than those of side branches.}
+{\bf Mutations in evolutionary successful strains tend to be more favorable than in strains that die out.}
 (A) Phylogenetic tree of human H3N2 influenza virus from 1968-present. 
-\comment{We downloaded X sequences from the Influenza Virus Resource ?.... etc. inferred the tree, ancestral state reconstruction, visualized the tree. Mark Perth/2009 on the tree}
-To parse out trunk mutations from side branch mutations, we first defined a set of recent nodes sampled on or after Jan. 1, 2017, and traced these nodes back to their most recent common ancestor. 
-All branches ancestral to the MRCA of the recent nodes were defined as the trunk (shown in red), and all other branches were defined as side branches (shown in blue).
+The trunk is shown in red, and side branches are shown in blue.
+The gray branches represent the part of the tree for which we cannot yet distinguish the trunk from side branches.
 (B) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
@@
 (C) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel) are shown.
-To estimate significance, we performed 10,000 randomizations of the preferences to calculate the median difference in trunk vs side branch mutational effects, and counted how many of the randomizations exceeded the true difference in median trunk vs side branch mutational effects.
+The preferences were randomized for a number of iterations to estimate significance.
 The effects of trunk mutations are higher than side branch, internal and terminal node, mutations.
@@
 \end{figure}
+
+Rather than estimating an average effect of mutations for trunk and side branch lineages, we can directly compare nodes in the tree by scoring the HA sequence of every node using the preferences to quantify a metric which we will refer to as \textit{sequence preference}.
+Figure~\ref{fig:sequence_preference} shows that the sequence preference of trunk nodes tends to be higher than that of side branch nodes.
+Interestingly, the sequence preference of all nodes increases over time until it reaches the Perth/2009 node and its closely related nodes, which exhibit higher sequence preference than the trunk. 
+The observation that the strain in which we performed our deep mutational scan has one of the highest sequence preferences illustrates epistatic interactions among mutations such that an unpreferred mutation in one background may be preferred in another.
+% Due to the sheer number of possible pairwise combinations of mutations, it is impossible for us to experimentally characterize these epistatic interactions among mutations. 
+Yet, we are able to distinguish trunk mutations as more favorable than side branch mutations, indicating that the preferences are still of utility over short evolutionary timescales.
 
@@
 
+Can we then distinguish lineage-specific mutational effects using the preferences measured in a related HA homolog?
+Figure~\ref{fig:WSN_trunkvssidebranch} shows the effect of H3 trunk and side branch mutations as quantified by the WSN/1933 H1 preferences from~\cite{doud2016accurate}.
+It is evident that we do not see trunk mutations significantly more favored than side branch mutations using these preferences, suggesting that our ability to discriminate successful and unsuccessful lineages degrades over sufficiently long evolutionary distances.
+
+\comment{may add text here on using H1 preferences for H1N1 tree}
+
 \begin{figure}
@@
 (A) We calculated the log$_{2}$ mutational effects of the same set of trunk and side branch mutations from the inferred H3N2 phylogeny in Figure~\ref{fig:trunkvssidebranch} using the WSN/1933 H1 preferences.
-We again randomized the preferences for a total of 10,000 iterations and counted the number of randomizations that exceeded that true difference in median trunk vs side branch mutational effects to calculate a p-value.
-Using the WSN/1933 H1 preferences, there is not a significant difference in trunk vs side branch mutational effects.
+There is not a significant difference in trunk vs side branch mutational effects.
 (B) We also performed the same sliding window analysis shown in Figure~\ref{fig:trunkvssidebranch}B, but using the WSN/1933 H1 preferences.
@@
 The WSN/1933 H1 and the Perth/2009 H3 share $\sim$42\% amino-acid identity.
-(B) The correlation for the amino-acid preferences for replicates both within and between the two HA homologs.
+(B) The correlations of the amino-acid preferences for replicates both within and between the two HA homologs. 
+The within-Perth/2009 and the within-WSN/1933 correlations are shown in blue and red, respectively.
+The between homolog correlations are in gray.
+The correlations for replicates within a homolog are higher than for replicates between homologs.
 (C) The distribution of shifts in preference for various homolog pairs are shown.
@@
 \subsection*{Inference of phylogenetic trees}
+\comment{We downloaded X sequences from the Influenza Virus Resource ?.... etc. inferred the tree, ancestral state reconstruction, visualized the tree. Mark Perth/2009 on the tree}
+To parse out trunk mutations from side branch mutations, we first defined a set of recent nodes sampled on or after Jan. 1, 2017, and traced these nodes back to their most recent common ancestor.
 

### 2017-12-20 paper/manuscript.tex
Added a bit to results section 4

@@
 \item Mutations that contribute to antigenic evolution largely determine strain success.
-\item However, mutations in HA that impact viral growth may also influence strain success.
+\item RNA viruses can accumulate deleterious mutations, and because the influenza does not appreciably recombine, deleterious mutations are linked to beneficial ones.
+\item Mutations in HA that impact viral growth may influence strain success.
 \item We need to understand the functional impact of mutations in HA.
@@
 \hline
-\bf{Model} & \bf{$\Delta$AIC} & \bf{Log Likelihood} & \bf{Parameters} \\ \hline
-ExpCM & 0.0 & -8439.33 & $\beta = 2.44, \kappa = 5.78, \omega = 0.91$ \\
+\bf{Model} & \bf{$\Delta$AIC} & \bf{Log Likelihood} & \bf{Stringency} \\ \hline
+ExpCM & 0.0 & -8439.33 & $\beta = 2.44, \omega = 0.91$ \\
 Goldman-Yang M5 & 2166.06 & -9516.36 & $\omega_{\alpha} = 0.30, \omega_{\beta} = 0.84, \kappa = 5.10$ \\
@@
 \caption{\label{fig:mut_tolerance}
-{\bf Different regions of HA exhibit varying degrees of mutational tolerance.}
+{\bf Mutational tolerance mapped onto each site of HA.}
 Mutational tolerance as calculated by the Shannon entropy of a given site's amino-acid preferences are mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}) and the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}), with both trimers in approximately the same orientation.
@@
 \subsection*{The H3 preferences reveal differences in seasonal influenza virus lineages}
-The human H3N2 phylogeny has a characteristic ladder-like structure, with a persistent trunk lineage and side branches that rapidly die out.
+% The human H3N2 phylogeny has a characteristic ladder-like structure, with a persistent trunk lineage and side branches that rapidly die out.
 Do the preferences reveal inherent differences between lineages of human H3N2 influenza virus that have succeeded and those that have died out?
 To examine this, we calculated the effect of trunk and side branch mutations from an H3N2 phylogeny, shown in Figure~\ref{fig:trunkvssidebranch}A, from 1968-2012 using our preference dataset. 
+Strains with mutations that we measured to be more beneficial to viral growth tend to succeed in nature.
 Figure~\ref{fig:trunkvssidebranch}B shows the median log$_{2}$ mutational effects and the interquartile ranges for trunk and side branch mutations in five-year windows for every year from 1968-2012. 
@@
 We found that terminal node mutations are on average towards less preferred amino acids than internal node mutations, and that both internal and terminal node mutations are significantly less favorable than those on the trunk.
-Therefore, strains that have accumulated mutations that we measured in our experiments to be unpreferred generally die out in nature.
+Therefore, strains that have accumulated mutations that we measured in our experiments to be unpreferred generally die out in nature, while more favored mutations  provide a selective advantage to the surviving trunk.
+\comment{The average effect of trunk mutations near zero may also suggest that there is a balance between preferred and unpreferred mutations on the trunk.}
 These findings demonstrate the importance of the functional impacts of HA mutations in determining lineage success.

### 2017-12-22 paper/manuscript.tex
Edit manuscript text up to trunk / tip section

@@
 \author{
-Juhye M. Lee$^{1,4,5,\dagger}$ \and 
-John Huddleston$^{2,6,\dagger}$ \and 
-Michael B. Doud$^{1,4,5}$ \and 
-Kathryn A. Hooper$^{1,6}$ \and
+Juhye M. Lee$^{1,4,5,\dagger}$, \and 
+John Huddleston$^{2,6,\dagger}$, \and 
+Michael B. Doud$^{1,4,5}$, \and 
+Kathryn A. Hooper$^{1,6}$, \and
 Trevor Bedford$^{2,3*}$ and \and 
@@
 \subsection*{Deep mutational scanning of HA from a recent strain of human H3N2 influenza virus}
-We performed a deep mutational scan of H3 HA from the A/Perth/16/2009 (H3N2) strain. 
+We performed a deep mutational scan to measure the effects of all amino-acid mutations to the A/Perth/16/2009 (H3N2) HA on viral replication in cell culture. 
 This strain was the H3N2 component of the influenza vaccine from 2010-2012.
-The HA gene we used in our experiments carries two mutations, G78D and T212I, relative to the Perth/2009 HA sequence deposited in GenBank.
-The G78D mutation has appeared sporadically at low frequency in natural H3N2 sequences. 
-Although the T212I mutation is not appreciable in nature \comment{reword}, the T212A mutation rose to fixation in the human population in $\sim$2011.
-We found that the G78D-T212I Perth/2009 HA mutant supported viral growth to $\sim$15- to 20-fold higher titers than the wildtype HA.
+We used a variant of this HA that supported better viral replication in cell culture than did the sequence in GenBank \comment{Figure S1}.
+This HA variant contained two mutations: G78D and T212I.
+The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.
 %We first generated virus carrying the wildtype Perth/2009 HA and NA and the WSN internal genes by reverse genetics, and serially passaged the virus in cell culture a total of six times.
 %This passaging was completed to allow the virus to adapt to growth in cell culture and to find cell-culture adaptation mutations that emerged during the course of passaging.
-%Two mutations, G78D and T212I, arose together from passaging.
-We therefore decided to perform all deep mutational scanning experiments in the background of the Perth/2009 HA carrying these two mutations to support higher viral growth.
 
-We mutagenized the HA gene to create mutant plasmid libraries encompassing all 567 codons and harboring an average of $\sim$1.4 codon mutations per clone.
+We mutagenized the entire gene at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone.
 We then generated mutant virus libraries from the mutant plasmids using a helper virus system that enables the efficient generation of complex virus libraries~\citep{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
 In order to reduce biosafety concerns and to maximize viral titers, we used an HA-deficient helper virus carrying WSN/1933 internal and NA genes to rescue the mutant viruses.
-Additionally, we used MDCK-SIAT1 cells constitutively expressing the TMPRSS2 protease, which facilitates HA cleavage and activation in the human airway \comment{cite B\"{o}ttcher 2006 JVI, B\"{o}ttcher-Friebertsh\"{a}user 2010 JVI}.
+Additionally, we used MDCK-SIAT1 cells constitutively expressing the TMPRSS2 protease, which facilitates HA cleavage and activation in the human airway~\citep{bottcher2006proteolytic, bottcher2010cleavage}.
 All of the experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B). 
 We also passaged and deep sequenced library 3 in technical replicate (denoted as library 3-1 and 3-2) to gauge to the amount of experimental noise occurring \textit{within} a single biological replicate.
+As a control, we used the unmutated gene to generate and passage viruses carrying wildtype HA.
 
-Figure~\ref{fig:dms_overview}C shows the mutation frequencies of the mutant plasmids, mutant viruses, wildtype plasmids and wildtype viruses determined from deep sequencing. 
-There is selection against non-functional HA variants as revealed by the reduced mutation frequencies of the mutant viruses compared to their starting frequencies in the mutant plasmids.
+Deep sequencing revealed selection against non-functional HA variants as evidenced by the reduced mutation frequencies of the mutant viruses compared to their starting frequencies in the mutant plasmids (Figure~\ref{fig:dms_overview}C).
 Specifically, stop codons were purged to 20-45\% of their original starting frequencies, after correcting for error rates from the wildtype controls.
-Although the majority of stop codons were purged, incomplete purging of stop codons is likely due to complementation of defective viral particles by infectious virions. 
-We also observed purging of most nonsynonymous mutations to 30-40\% of their initial starting frequencies after error correction, suggesting strong selection against deleterious HA variants.
+Although the majority of stop codons were purged, incomplete purging of stop codons is likely due to complementation of virions. 
+We also observed purging of nonsynonymous mutations to 30-40\% of their initial starting frequencies after error correction, suggesting strong selection against deleterious HA variants.
 
@@
 {\bf Deep mutational scanning experiments of H3 HA.}
-(A) We used a helper virus approach previously described in Doud (2016) to generate the mutant virus libraries. 
-We transfected MDCK-SIAT1-TMPRSS2 cells with the mutant plasmid library carrying all possible amino-acid mutations to the A/Perth/16/2009 (H3N2) HA, in addition to protein expression plasmids encoding the viral ribonucleoprotein complex.
-After transfection, we infected the cells with an HA-deficient helper virus carrying all of the WSN/1933 (H1N1) influenza genes. 
+(A) We generated Perth/2009 HA mutant virus libraries in MDCK-SIAT1-TMPRSS2 cells using a helper virus approach described in~\cite{doud2016accurate}.
 We then passaged the initially generated pool of mutant viruses at low MOI to establish a genotype-phenotype linkage and select for functional HA variants. 
-(B) All of the experiments were completed in biological triplicate, starting from independent preps of the wildtype HA genes to create the mutant plasmids. 
-In addition, we passaged and deep sequenced library 3 in technical replicate, denoted 3-1 and 3-2, to estimate the amount of experimental noise within a single biological replicate.
-(C) Mutation frequencies of nonsynonymous, stop, and synonymous mutations for the mutant DNA, mutant virus, wildtype DNA, and wildtype virus samples. 
+Deep sequencing of the variants before and after selection allowed us to infer the H3 site-specific amino-acid preferences.
+(B) The DMS experiments were completed in biological triplicate, starting from independent preps of the wildtype HA genes to create the mutant plasmids. 
+In addition, we passaged and deep sequenced library 3 in technical replicate, denoted 3-1 and 3-2, to estimate the experimental noise within a single biological replicate.
+(C) Mutation frequencies of nonsynonymous, stop, and synonymous mutations for the mutant DNA, mutant virus, and unmutated DNA and virus controls. 
 There is selection against nonsynonymous and stop codons in the mutant viruses. 
 The percentages signify the frequency of stop codons remaining in the passaged mutant viruses relative to their starting frequency in the mutant plasmid libraries after correcting for the stop codon frequencies in the wildtype DNA and viruses.
-(D) The correlations and the Pearson correlation coefficient for the amino-acid preferences between each pair of replicates are shown. 
-The biological replicates are fairly well-correlated, as are the technical replicates, indicating some degree of bottlenecking of variants during the viral passage which can contribute to experimental noise.
+(D) The Pearson correlations for the amino-acid preferences between each pair of replicates are shown. 
+The correlation between the technical replicates is comparable to that between biological replicates, indicating that the low MOI passage contributes to much of the experimental noise. 
 }
@@
 \centering
-\begin{tabular}{cccc}
+\begin{tabular}{cccccccc}
 \hline
-\bf{Model} & \bf{$\Delta$AIC} & \bf{Log Likelihood} & \bf{Stringency} \\ \hline
-ExpCM & 0.0 & -8439.33 & $\beta = 2.44, \omega = 0.91$ \\
-Goldman-Yang M5 & 2166.06 & -9516.36 & $\omega_{\alpha} = 0.30, \omega_{\beta} = 0.84, \kappa = 5.10$ \\
-ExpCM, averaged across sites & 2504.18 & -9691.42 & $\beta = 0.68, \kappa = 5.58, \omega = 0.32$ \\
-Goldman-Yang M0 & 2607.92 & -9738.29 & $\kappa = 5.05, \omega = 0.31$ \\
+\bf{Model} & \bf{$\Delta$AIC} & \bf{Log Likelihood} & \bf{Stringency} & \bf{$\omega$} & \bf{$\overline{\omega}$} & \bf{$\omega_{\alpha}$} & \bf{$\omega_{\beta}$} \\ \hline
+ExpCM & 0.0 & -8439.33 & $2.44$ & $0.91$ & -- & -- & -- \\
+Goldman-Yang M5 & 2166.06 & -9516.36 & -- & -- & $0.36$ & $0.30$ & $0.84$ \\
+ExpCM, averaged across sites & 2504.18 & -9691.42 & $0.68$ & $0.32$ & -- & -- & -- \\
+Goldman-Yang M0 & 2607.92 & -9738.29 & -- & $0.31$ & -- & -- & -- \\
 \hline
@@
 An experimentally-informed codon substitution model (ExpCM) built from the preferences averaged across all replicates performs better than conventional substitution models, specifically the M0 and M5 models in~\cite{yang2000codon}.
-A non-site-specific ExpCM informed by preferences averaged across all sites does not perform as well as the M5 model, but still has a higher log likelihood than M0.
+A non-site-specific ExpCM informed by preferences averaged across all sites performs comparably to the GY94 class of models, indicating the importance of site-specificity in the ExpCM.
 The optimized parameters for each model are also shown.
@@
 Figure~\ref{fig:mut_tolerance} shows the mutational tolerance as calculated from the rescaled Perth/2009 H3 preferences and the rescaled WSN/1933 H1 preferences mapped onto the HA crystal structures.
-Interestingly, the Perth/2009 H3 stalk is relatively mutationally tolerant compared to the tolerance of the globular head domain. 
-In particular, the shorter $\alpha$-helix (helix A) of the stalk domain, antigenic site C, and the most distal portion of the globular head near antigenic site B are all highly mutationally tolerant.
-% may add more details here
-On the other hand, the head domain of the WSN/1933 H1 is more mutationally tolerant relative to its stalk domain.
-The receptor binding pockets of both HA's are relatively mutationally intolerant, although the residues surrounding the receptor binding pocket are fairly tolerant of mutations.
-These findings are consistent with the high functional constraint of the receptor binding pocket, while the mutability of sites surrounding the pocket that are under strong immune pressure may contribute to antigenic evolution~\citep{wilson1981structure, wiley1981structural}.
+We found antigenic site C and the most distal portion of the globular head near antigenic site B in the Perth/2009 H3 to be tolerant of mutations.
+Interestingly, the H3 stalk including the shorter $\alpha$-helix (helix A) is relatively mutationally tolerant compared to the tolerance of the globular head domain. 
+This observation suggests that the stalk may be prone to escape from antibodies, and agrees with previous work demonstrating that it is possible to select for antigenic mutants in H3 by broadly-neutralizing stalk-targeting antibodies~\citep{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human}.
+% In contrast, the head domain of the WSN/1933 H1 is more mutationally tolerant relative to its stalk domain.
+
+The sites inside the receptor binding pockets are highly functionally constrained and were found to be relatively mutationally intolerant in both H3 and H1~\citep{wilson1981structure}.
+In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to antigenic evolution as these sites are under strong immune pressure.~\citep{wiley1981structural}.
 
@@
 
-\subsection*{The H3 preferences reveal differences in seasonal influenza virus lineages}
+\subsection*{The experimental measurements can help discriminate successful influenza virus lineages}
 % The human H3N2 phylogeny has a characteristic ladder-like structure, with a persistent trunk lineage and side branches that rapidly die out.
@@
 \section*{DISCUSSION}
+We have measured the effect of all possible single amino-acid mutations to Perth/2009 H3 on viral growth in cell culture.
 

### 2017-12-28 paper/manuscript.tex
Switched order of distributions in distance_distribution fig
Add G78D-T212I supplemental figure

@@
 This strain was the H3N2 component of the influenza vaccine from 2010-2012.
-We used a variant of this HA that supported better viral replication in cell culture than did the sequence in GenBank \comment{Figure S1}.
+We used a variant of this HA that supported better viral replication in cell culture than did the sequence in GenBank (Figure~\ref{suppfig:Perth2009_mut}).
 This HA variant contained two mutations: G78D and T212I.
@@
 \begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/trunkvssidebranch/trunkvssidebranch.pdf}}
+\centerline{\includegraphics[width=0.75\textwidth]{figs/trunkvssidebranch/trunkvssidebranch.pdf}}
 \caption{\label{fig:trunkvssidebranch}
@@
 \begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/sequence_preference/sequence_preference.pdf}}
+\centerline{\includegraphics[width=0.75\textwidth]{figs/sequence_preference/sequence_preference.pdf}}
 \caption{\label{fig:sequence_preference}
@@
 
+\section*{Supplementary Material}
+\FloatBarrier
+\pagenumbering{arabic}% resets `page` counter to 1
+\renewcommand*{\thepage}{S\arabic{page}}
+
+\begin{suppfigure}
+\centerline{\includegraphics[width=\textwidth]{figs/S01_G78D-T212I/G78D-T212I.pdf}}
+\caption{\label{suppfig:Perth2009_mut}
+{\bf Characterization of the G78D-T212I Perth/2009 HA variant.} 
+(A) 
+The G78D-T212I Perth/2009 HA variant grows to higher titers than do viruses carrying the wildtype Perth/2009 HA.
+Each virus was generated in duplicate by reverse genetics and passaged once at MOI = 0.01 in MDCK-SIAT1-TMPRSS2 cells.
+The rescue and passage viral supernatants were titered, with the points marking each duplicate and the bar marking the mean.
+(B)
+The D78 variant remained at a low frequency in natural human H3N2 sequences over the past $~\sim$10 years.
+The A212 variant rose to fixation in $~\sim$2011, replacing the T212 variant.
+}
+\end{suppfigure}
+
 \end{document}

### 2017-12-30 paper/manuscript.tex
Edit text of first four results sections

@@
 
-\title{Deep mutational scanning of an H3 hemagglutinin can inform evolutionary forecasting of human H3N2 influenza virus}
+\title{Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fate of human H3N2 influenza virus}
 
@@
 This codon is part of the signal peptide and is cleaved from the mature HA protein.
-One possible explanation for why we do not see a strong preference for Met at this site is due to alternative translation initiation occurring at a downstream or upstream start site, as has been described in \comment{need to list citations here, inc.
-Girard 2011 Infection Genetics and Evolution, Chen 2001 Nat Med}.
+One possible explanation for why we do not see a strong preference for Met at this site is due to alternative translation initiation occurring at a downstream or upstream start site, as has been described for HA~\citep{girard2011upstream}.
 % No upstream or downstream "ATG"'s are present.
@@
 Figure~\ref{fig:mut_tolerance} shows the mutational tolerance as calculated from the rescaled Perth/2009 H3 preferences and the rescaled WSN/1933 H1 preferences mapped onto the HA crystal structures.
-We found antigenic site C and the most distal portion of the globular head near antigenic site B in the Perth/2009 H3 to be tolerant of mutations.
+We found antigenic site C and the most distal portion of the globular head near the 190-helix in the Perth/2009 H3 to be tolerant of mutations.
 Interestingly, the H3 stalk including the shorter $\alpha$-helix (helix A) is relatively mutationally tolerant compared to the tolerance of the globular head domain. 
@@
 The sialic acid receptor is shown as black sticks.
-The Perth/2009 H3 shows relatively high mutational tolerance in the stalk region compared to the head region. 
+The Perth/2009 H3 shows relatively high mutational tolerance in the stalk region, particularly in helix A, compared to the head region. 
+High mutational tolerance in H3 was also observed near antigenic site C and near the 190 helix.
 The head region of the WSN/1933 H1 is mutationally tolerant compared to the relatively intolerant stalk region. 
@@
 \subsection*{The experimental measurements can help discriminate successful influenza virus lineages}
-% The human H3N2 phylogeny has a characteristic ladder-like structure, with a persistent trunk lineage and side branches that rapidly die out.
-Do the preferences reveal inherent differences between lineages of human H3N2 influenza virus that have succeeded and those that have died out?
-To examine this, we calculated the effect of trunk and side branch mutations from an H3N2 phylogeny, shown in Figure~\ref{fig:trunkvssidebranch}A, from 1968-2012 using our preference dataset. 
-Strains with mutations that we measured to be more beneficial to viral growth tend to succeed in nature.
-Figure~\ref{fig:trunkvssidebranch}B shows the median log$_{2}$ mutational effects and the interquartile ranges for trunk and side branch mutations in five-year windows for every year from 1968-2012. 
-On average, trunk mutations tend to be towards more preferred amino acids compared to side branch mutations, and this was true for all windows.
-Next we calculated the effect of all trunk and side branch mutations from the entirety of the phylogenetic tree.
-We found that on average, trunk mutations are significantly more favorable than side branch mutations (Figure~\ref{fig:trunkvssidebranch}C).
-Because a number of tip nodes are likely from egg- or cell-passaged isolates and our experiments were performed in cell culture, we were curious to see if our measurements would lead to higher estimations of mutational effects for terminal nodes.
-We found that terminal node mutations are on average towards less preferred amino acids than internal node mutations, and that both internal and terminal node mutations are significantly less favorable than those on the trunk.
-Therefore, strains that have accumulated mutations that we measured in our experiments to be unpreferred generally die out in nature, while more favored mutations  provide a selective advantage to the surviving trunk.
-\comment{The average effect of trunk mutations near zero may also suggest that there is a balance between preferred and unpreferred mutations on the trunk.}
-These findings demonstrate the importance of the functional impacts of HA mutations in determining lineage success.
+A challenge in vaccine strain selection is predicting which strain will dominate the upcoming influenza season.
+An important question to thereby address is if our experimental measurements are useful in differentiating between strains of human H3N2 influenza virus that have succeeded and those that have died out.
+To investigate this, we used our preference dataset and an H3N2 phylogeny from 1968-2012 (Figure~\ref{fig:trunkvssidebranch}A) to calculate the effects of mutations for the evolutionarily successful trunk lineage and for side branches which have died out.
+We found that strains with mutations measured to be more beneficial to viral growth in our experiments tend to succeed in nature.
+Figure~\ref{fig:trunkvssidebranch}B shows the effects of trunk and side branch mutations in five-year intervals for every year from 1968-2012. 
+On average, trunk mutations are towards more preferred amino acids compared to side branch mutations, and this was true for all intervals.
+Importantly, trunk mutations are significantly more favorable than side branch mutations when calculating the effects from the entirety of the phylogenetic tree (Figure~\ref{fig:trunkvssidebranch}C).
+
+Because tip nodes can contain egg- or cell-passaged isolates~\citep{wu2017structural,mcwhite2016sequence,skowronski2016mutations} and our experiments were performed in cell culture, we examined the effects of mutations on terminal side branch nodes to see if these would rank more highly than internal node mutations.
+Instead, terminal node mutations are on average towards less preferred amino acids than internal node mutations (Figure~\ref{fig:trunkvssidebranch}C), and both internal and terminal node mutations are significantly less favorable than those on the trunk.
+Therefore, strains that have accumulated mutations that we experimentally measured to be unpreferred generally die out in nature, while more favored mutations provide a selective advantage to the trunk.
+These findings demonstrate the importance of the functional impacts of HA mutations in determining strain success.
 
@@
 The shaded region demarcates the interquartile range of trunk and side branch mutational effects.
+Negative numbers signify mutations towards less preferred amino acids, while positive numbers signify more preferred mutations.
 The median trunk mutational effects are consistently higher than the median side branch mutational effects for all windows.
 (C) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel) are shown.
-The preferences were randomized for a number of iterations to estimate significance.
-The effects of trunk mutations are higher than side branch, internal and terminal node, mutations.
+The preferences were randomized 10,000 times to estimate significance.
+The effects of trunk mutations are higher than side branch internal and terminal node mutations.
 }
@@
 
-Rather than estimating an average effect of mutations for trunk and side branch lineages, we can directly compare nodes in the tree by scoring the HA sequence of every node using the preferences to quantify a metric which we will refer to as \textit{sequence preference}.
-Figure~\ref{fig:sequence_preference} shows that the sequence preference of trunk nodes tends to be higher than that of side branch nodes.
-Interestingly, the sequence preference of all nodes increases over time until it reaches the Perth/2009 node and its closely related nodes, which exhibit higher sequence preference than the trunk. 
+How distantly can the preferences be extended to describe differences in successful and unsuccessful strains?
+To explore this question, we scored the complete HA sequence of every node in the tree using the preferences by quantifying a \textit{sequence preference} metric.
+Consistent with the finding that trunk mutations are generally more favorable than side branch mutations, Figure~\ref{fig:sequence_preference} shows that the sequences of trunk nodes also tend to be more highly preferred than those of side branch nodes.
+Interestingly, the sequence preferences increase over time as the nodes approach the Perth/2009 strain and its closely related nodes, which all exhibit sequence preferences that are higher than that of the trunk.
 The observation that the strain in which we performed our deep mutational scan has one of the highest sequence preferences illustrates epistatic interactions among mutations such that an unpreferred mutation in one background may be preferred in another.
 % Due to the sheer number of possible pairwise combinations of mutations, it is impossible for us to experimentally characterize these epistatic interactions among mutations.
-Yet, we are able to distinguish trunk mutations as more favorable than side branch mutations, indicating that the preferences are still of utility over short evolutionary timescales.
+
+\comment{this section will be fleshed out more once I figure out how to plot the substitutions from the root at epitope / non-epitope sites, as per Trevor's suggestion.}
+We reasoned that much of the increase in sequence preference toward Perth/2009 could be attributed to epitope sites, as most substitutions have occurred at sites under strong immune pressure.
+We therefore assessed each node's preference at epitope and non-epitope sites as defined by~\cite{wolf2006long}.
+Indeed, the preferences at epitope sites resembles those for all HA sites (Figure~\ref{fig:sequence_preference}), with preferences increasing over time.
+At non-epitope sites, however, the trunk preferences remain relatively constant while side branch preferences tend to drop below the trunk.
+These observations highlight the extensive epistasis among epitope sites, as has even been noted for sites within the 220-loop surrounding the receptor binding pocket~\citep{wu2017diversity}.
+Many successful epitope mutations are likely contingent upon the background in which they arise.
+Yet, we are able to distinguish trunk sequences as more favorable than side branch sequences, indicating that the preferences are still of utility over short evolutionary timescales.
 
@@
 \caption{\label{fig:sequence_preference}
-{\bf The trunk exhibits higher sequence preference than side branches.}
-figure text
+{\bf The HA sequences of trunk nodes tend to be more preferred than those of side branch nodes.}
+We used the preferences to calculate the HA sequence preference of every node in a human H3N2 phylogenetic tree.
+To calculate the sequence preference, we used the entire HA sequence (top panel), only epitope sites (center panel) or non-epitope sites (bottom panel) as defined by~\cite{wolf2006long}.
+Higher preferences are closer to zero while lower preferences are more negative.
+The preferences at all sites and at epitope sites increase as the tree approaches the Perth/2009 strain, indicating epistasis among epitope sites.
+The trunk nodes generally exhibit higher sequence preferences than do side branch nodes.
 }
@@
 
-Can we then distinguish lineage-specific mutational effects using the preferences measured in a related HA homolog?
-Figure~\ref{fig:WSN_trunkvssidebranch} shows the effect of H3 trunk and side branch mutations as quantified by the WSN/1933 H1 preferences from~\cite{doud2016accurate}.
-It is evident that we do not see trunk mutations significantly more favored than side branch mutations using these preferences, suggesting that our ability to discriminate successful and unsuccessful lineages degrades over sufficiently long evolutionary distances.
-
-\comment{may add text here on using H1 preferences for H1N1 tree}
+Can we then distinguish lineage-specific mutational effects using the preferences measured in a distantly related HA homolog?
+We used the preferences measured previously in the WSN/1933 H1~\citep{doud2016accurate} to quantify the effects of H3 trunk and side branch mutations, shown in Figure~\ref{fig:WSN_trunkvssidebranch}.
+It is evident that we do not see trunk mutations significantly more favored than side branch mutations, suggesting that our ability to discriminate successful and unsuccessful strains degrades over sufficiently long evolutionary distances.
 
@@
 {\bf The WSN/1933 H1 preferences do not reveal differences in trunk vs side branch mutational effects}
-(A) We calculated the log$_{2}$ mutational effects of the same set of trunk and side branch mutations from the inferred H3N2 phylogeny in Figure~\ref{fig:trunkvssidebranch} using the WSN/1933 H1 preferences.
+(A) We used the WSN/1933 H1 preferences from~\cite{doud2016accurate} to calculate the log$_{2}$ mutational effects of trunk and side branch mutations from the inferred H3N2 phylogeny in Figure~\ref{fig:trunkvssidebranch}.
 There is not a significant difference in trunk vs side branch mutational effects.
-(B) We also performed the same sliding window analysis shown in Figure~\ref{fig:trunkvssidebranch}B, but using the WSN/1933 H1 preferences.
+(B) We also performed a sliding window analysis with the WSN/1933 H1 preferences.
 There is not a distinct difference in trunk and side branch mutational effects.
@@
 
-\subsection*{Comparing H1 and H3 preferences}
+\subsection*{The H1 and H3 preferences at many sites have shifted}
+How shifted are the preferences between the distantly related Perth/2009 H3 and WSN/1933 H1 homologs?
+Although we have previously compared the preferences between related protein homologs of NP~\citep{doud2015site} and of Env \comment{cite haddox2017}, the HA homologs are considerably more diverged than either of these pairs.
 
@@
 (C) The distribution of shifts in preference for various homolog pairs are shown.
+The top two distributions show the distances between each of the HA homologs with the non-homologous HIV Env \comment{cite haddox2017}. 
+The center distribution shows the corrected distance between the two HIV Env homologs, which share 86\% amino-acid identity.
+The fourth distribution from the top is a null generated by randomizing the HA replicates and computing the distances.
+The bottom distribution is the corrected distance between the Perth/2009 H3 and WSN/1933 H1 homologs at all sites that align.
 }

### 2018-01-01 paper/manuscript.tex
Reword fig 1 caption

@@
 
-\title{Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fate of human H3N2 influenza virus}
+\title{Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fate of human H3N2 influenza virus lineages}
 
@@
 {\bf Deep mutational scanning experiments of H3 HA.}
-(A) We generated Perth/2009 HA mutant virus libraries in MDCK-SIAT1-TMPRSS2 cells using a helper virus approach described in~\cite{doud2016accurate}.
-We then passaged the initially generated pool of mutant viruses at low MOI to establish a genotype-phenotype linkage and select for functional HA variants. 
+(A) We generated Perth/2009 HA mutant virus libraries using a helper virus approach described in~\cite{doud2016accurate}.
+We then passaged this pool of mutant viruses at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants. 
 Deep sequencing of the variants before and after selection allowed us to infer the H3 site-specific amino-acid preferences.
-(B) The DMS experiments were completed in biological triplicate, starting from independent preps of the wildtype HA genes to create the mutant plasmids. 
-In addition, we passaged and deep sequenced library 3 in technical replicate, denoted 3-1 and 3-2, to estimate the experimental noise within a single biological replicate.
+(B) The DMS experiments were performed in biological triplicate, starting from independent preps of the wildtype HA genes to create the mutant plasmids. 
+We also passaged and deep sequenced library 3 in technical replicate, denoted 3-1 and 3-2, to estimate the experimental noise within a biological replicate.
 (C) Mutation frequencies of nonsynonymous, stop, and synonymous mutations for the mutant DNA, mutant virus, and unmutated DNA and virus controls. 
 There is selection against nonsynonymous and stop codons in the mutant viruses. 
-The percentages signify the frequency of stop codons remaining in the passaged mutant viruses relative to their starting frequency in the mutant plasmid libraries after correcting for the stop codon frequencies in the wildtype DNA and viruses.
 (D) The Pearson correlations for the amino-acid preferences between each pair of replicates are shown. 
-The correlation between the technical replicates is comparable to that between biological replicates, indicating that the low MOI passage contributes to much of the experimental noise. 
+The correlation between the technical replicates is comparable to that between biological replicates, indicating that much of the experimental noise comes from passaging at low MOI. 
 }

### 2018-01-01 paper/manuscript.tex
Tweak text in 2nd and 3rd sections of results

@@
 For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\citep{waterfield1981disulphide} possess high preference for cysteine.
-At residues involved in receptor binding, there are strong preferences for the amino acids at sites Y98, D190, W153, and S228.
+At residues involved in receptor binding, there are strong preferences for the amino acids at sites Y98, D190, W153, and S228~\citep{weis1988structure,martin1998studies}.
 A positively charged amino acid at site 329 is important for cleavage activation of the HA0 precursor, and indeed this site exhibits a high preference for arginine~\citep{kido1992isolation, stech2005new}.
@@
 
-\subsection*{The H3 stalk is relatively mutationally tolerant}
+\subsection*{The mutational tolerance of the head and the stalk shows less of a contrast in H3 than in H1}
 We next sought to investigate the inherent mutational tolerance of the Perth/2009 HA. 

### 2018-01-01 paper/manuscript.tex
Add some details to supp fig 1

@@
 Each virus was generated in duplicate by reverse genetics and passaged once at MOI = 0.01 in MDCK-SIAT1-TMPRSS2 cells.
-The rescue and passage viral supernatants were titered, with the points marking each duplicate and the bar marking the mean.
+The rescue and passage viral supernatants were collected at 72 hours post-transfection and 44 hours post-infection, respectively, and titered in MDCK-SIAT1-TMPRSS2 cells. 
+The points mark each duplicate and the bar marks the mean.
 (B)

### 2018-01-01 paper/manuscript.tex
Tweak some text in Results

@@
 The results in Table~\ref{tab:phydms} show that the ExpCM outperforms conventional substitution models in describing the evolution of human H3N2 HA. 
-The ExpCM also optimizes a stringency parameter ($\beta$) for the preferences to more closely reflect the strength of selection in nature.
+The ExpCM also optimizes a stringency parameter for the preferences to more closely reflect the strength of selection in nature.
 The stringency parameter in the ExpCM is equal to 2.44, which indicates that although the same amino acids are preferred, the strength of selection is more stringent in nature than in our experiments.
@@
 To explore this question, we scored the complete HA sequence of every node in the tree using the preferences by quantifying a \textit{sequence preference} metric.
+The sequence preference is defined as 
+$$\displaystyle\sum_{r} \ln \pi_{r, a}$$
+where $\pi_{r, a}$ is the preference for amino acid $a$ at site $r$.
 Consistent with the finding that trunk mutations are generally more favorable than side branch mutations, Figure~\ref{fig:sequence_preference} shows that the sequences of trunk nodes also tend to be more highly preferred than those of side branch nodes.
@@
 \subsection*{The H1 and H3 preferences at many sites have shifted}
-How shifted are the preferences between the distantly related Perth/2009 H3 and WSN/1933 H1 homologs?
-Although we have previously compared the preferences between related protein homologs of NP~\citep{doud2015site} and of Env \comment{cite haddox2017}, the HA homologs are considerably more diverged than either of these pairs.
+How shifted are the preferences between evolutionarily distant homologs such as H1 and H3?
+Although we have previously compared the preferences between related protein homologs of NP~\citep{doud2015site} and of Env \comment{cite haddox2017}, the HA homologs we have experimentally studied are considerably more diverged than either of these pairs.
+The WSN/1933 H1 and the Perth/2009 H3 fall into two phylogenetically distinct groups and share 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
+\begin{itemize}
+\item Correlations are higher within homolog replicate than between homologs
+\item Next aligned and compared preferences using a correction method to account for experimental noise
+\end{itemize}
 

### 2018-01-01 paper/manuscript.tex
Start working on last results section

@@
 {\bf The HA sequences of trunk nodes tend to be more preferred than those of side branch nodes.}
-We used the preferences to calculate the HA sequence preference of every node in a human H3N2 phylogenetic tree.
+We used the preferences to calculate the HA sequence preference normalized by the number of sites of every node in a human H3N2 phylogenetic tree.
 To calculate the sequence preference, we used the entire HA sequence (top panel), only epitope sites (center panel) or non-epitope sites (bottom panel) as defined by~\cite{wolf2006long}.
@@
 Although we have previously compared the preferences between related protein homologs of NP~\citep{doud2015site} and of Env \comment{cite haddox2017}, the HA homologs we have experimentally studied are considerably more diverged than either of these pairs.
-The WSN/1933 H1 and the Perth/2009 H3 fall into two phylogenetically distinct groups and share 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
-\begin{itemize}
-\item Correlations are higher within homolog replicate than between homologs
-\item Next aligned and compared preferences using a correction method to account for experimental noise
-\end{itemize}
+The WSN/1933 H1 and the Perth/2009 H3 fall into two phylogenetically distinct groups and share 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A), providing an opportunity to analyze what has shifted and what has remained conserved.
+Simply correlating the preferences between H1 and H3 reveals that the replicate measurements within a homolog are more correlated than between homologs (Figure~\ref{fig:distance_distribution}B).
+
+However, to quantify the extent of mutational shifts at a site-by-site level, we used an approach described in \comment{haddox2017}.
+We aligned Perth/2009 H3 and WSN/1933 H1 and at each alignable site calculated the difference in preferences between homologs while correcting for experimental noise within homologs.
+The distribution of shifts is shown in Figure~\ref{fig:distance_distribution}C.
 

### 2018-01-02 paper/manuscript.tex
Some more progress on last results section

@@
 The distribution of shifts is shown in Figure~\ref{fig:distance_distribution}C.
+Although many sites have small shifts near zero, a considerable number of sites have large shifts in preference, reaching a difference of 0.86 out of a maximum possible of 1.0.
+When we compare HA to the preferences measured in the non-homologous protein HIV Env \comment{haddox2017}, nearly all sites are shifted.
+On the other hand, when we randomize the HA replicates to generate a null distribution, there is very little shift in preferences. 
+For comparison, the preferences of the HIV Env homologs, which are 86\% identical at the amino-acid level, are mostly similar. 
+Although there are a small number of sites with larger shifts reaching a maximum of 0.52, we do not see a dramatic tail in the distribution as we see for the HA's.
+These observations suggest that as homologs diverge, their preferences increasingly shift.
 
@@
 {\bf Shifts in preferences mapped onto the structure of HA}
-(A) The preference shifts as calculated by $RMSD_{corrected}$ between the two HA homologs is mapped onto the structure of HA (PDB 4O5N, citation). 
+(A) The preference shifts as calculated by $RMSD_{corrected}$ between the two HA homologs is mapped onto the structure of HA (PDB 4O5N;~\cite{lee2014receptor}). 
 The left structure shows the HA trimer, and the right structure colors one of the monomers. 
 The sialic acid receptor is shown in black sticks.
-Gray indicates little shifts in preference, while red indicates large shifts in preference.
+Blue indicates small shifts in preference near zero, while red indicates large shifts in preference.
 The top ten most shifted sites are shown in spheres on the monomer.
-(B)
+(B) The stalk domain was found to be significantly less shifted than the head domain (left plot).
+Sites absolutely conserved all 18 HA subtypes were also found to be significantly less shifted than the remaining non-conserved sites (right plot).
+(C) 
 }

### 2018-01-02 paper/manuscript.tex
adjust table font size
Some more progress on last results section

@@
 \centering
+{\small
 \begin{tabular}{cccccccc}
@@
 \end{tabular}
+}
 \caption{\label{tab:phydms}
@@
 
+Upon mapping the shifts in preference onto the HA structure (Figure~\ref{fig:RMSD_heatmap}A), we did not see the shifted preferences obviously localize to specific regions in HA.
+However, we found the overall stalk domain to be significantly less shifted than the globular head domain (Figure~\ref{fig:RMSD_heatmap}B), which was expected given that the stalk domain is more conserved within and across HA subtypes.
+Sites absolutely conserved across all 18 HA subtypes were also found to be significantly less shifted than sites not conserved, which is suggestive of the high functional importance of the residues that have remained unchanged throughout the divergence of HA.
+
 \begin{figure}

### 2018-01-03 paper/manuscript.tex
Add a bit to last results section
phrasing edits to first subsection of results

@@
 \subsection*{Deep mutational scanning of HA from a recent strain of human H3N2 influenza virus}
-We performed a deep mutational scan to measure the effects of all amino-acid mutations to the A/Perth/16/2009 (H3N2) HA on viral replication in cell culture. 
-This strain was the H3N2 component of the influenza vaccine from 2010-2012.
-We used a variant of this HA that supported better viral replication in cell culture than did the sequence in GenBank (Figure~\ref{suppfig:Perth2009_mut}).
-This HA variant contained two mutations: G78D and T212I.
+We performed a deep mutational scan to measure the effects of all amino-acid mutations to HA from the A/Perth/16/2009 (H3N2) strain on viral replication in cell culture. 
+This strain was the H3N2 component of the influenza vaccine from 2010-2012~\comment{add citation}.
+Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral replication in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and \comment{supplementary file or dataset with our full reverse genetics plasmid sequence}).
 The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.
-%We first generated virus carrying the wildtype Perth/2009 HA and NA and the WSN internal genes by reverse genetics, and serially passaged the virus in cell culture a total of six times.
-%This passaging was completed to allow the virus to adapt to growth in cell culture and to find cell-culture adaptation mutations that emerged during the course of passaging.
 
-We mutagenized the entire gene at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone.
-We then generated mutant virus libraries from the mutant plasmids using a helper virus system that enables the efficient generation of complex virus libraries~\citep{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
-In order to reduce biosafety concerns and to maximize viral titers, we used an HA-deficient helper virus carrying WSN/1933 internal and NA genes to rescue the mutant viruses.
-Additionally, we used MDCK-SIAT1 cells constitutively expressing the TMPRSS2 protease, which facilitates HA cleavage and activation in the human airway~\citep{bottcher2006proteolytic, bottcher2010cleavage}.
+We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (\comment{supplementary figure with Sanger sequencing}).
+We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables the efficient generation of complex influenza virus libraries~\citep{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
+These mutant viruses derived all of their non-HA genes from the lab-adapted WSN/1933 strain.
+Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
+To further increase viral titers, we used MDCK-SIAT1 cells that we had engineered to constitutively express the TMPRSS2 protease, which facilitates HA cleavage and activation~\citep{bottcher2006proteolytic, bottcher2010cleavage}.
+
+After generating the mutant virus libraries, we passaged them at low MOI in cell culture to create a genotype-phenotype link and select for functional HA variants (Figure~\ref{fig:dms_overview}A).
 All of the experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B). 
-We also passaged and deep sequenced library 3 in technical replicate (denoted as library 3-1 and 3-2) to gauge to the amount of experimental noise occurring \textit{within} a single biological replicate.
-As a control, we used the unmutated gene to generate and passage viruses carrying wildtype HA.
+We also passaged and deep sequenced library 3 in duplicate (denoted as library 3-1 and 3-2) to gauge to the amount of experimental noise occurring \textit{within} a single biological replicate.
+As a control to measure sequencing and mutational errors, we used the unmutated HA gene to generate and passage viruses carrying wildtype HA.
 
-Deep sequencing revealed selection against non-functional HA variants as evidenced by the reduced mutation frequencies of the mutant viruses compared to their starting frequencies in the mutant plasmids (Figure~\ref{fig:dms_overview}C).
-Specifically, stop codons were purged to 20-45\% of their original starting frequencies, after correcting for error rates from the wildtype controls.
-Although the majority of stop codons were purged, incomplete purging of stop codons is likely due to complementation of virions. 
-We also observed purging of nonsynonymous mutations to 30-40\% of their initial starting frequencies after error correction, suggesting strong selection against deleterious HA variants.
+Deep sequencing of the initial plasmid mutant libraries and the passaged mutant viruses allowed us to observe selection for functional HA mutants.
+As expected, selection against non-functional HA variants as led to reduced mutation frequencies in the mutant viruses compared to the initial mutant plasmids (Figure~\ref{fig:dms_overview}C).
+Specifically, stop codons were purged to 20-45\% of their initial frequencies after correcting for error rates estimated by sequencing the wildtype controls.
+The incomplete purging of stop codons is likely because genetic complementation due to co-infection \comment{let's cite some paper here, maybe Chris Brooke one?} enabled the persistence of some virions with nonfunctional HAs.
+We also observed selection against many nonsynonymous mutations, with their frequencies falling to 30-40\% of their initial values after error correction.
 
 We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
-We inferred the amino-acid preferences at each site in the Perth/2009 HA using the method described in~\cite{bloom2015software} and implemented in the \texttt{dms\_tools2} software [\url{https://jbloomlab.github.io/dms_tools2/}].
-These preferences represent an estimate of the 567 sites $\times$ 20 amino acids $=$ 11340 experimental measurements and are normalized to sum to one at each site.
-The correlations of the amino-acid preferences between each pair of replicates is shown in Figure~\ref{fig:dms_overview}D.
-The biological replicates are fairly well-correlated, with a Pearson's $R$ ranging from 0.69 to 0.78. 
-Replicate 1 exhibits the least amount of correlation with the other biological replicates, consistent with the observation that this replicate showed the weakest selection against stop and nonsynonymous mutations and might therefore be subject to more experimental noise.
-Of note, the two technical replicates 3-1 and 3-2 were only slightly more reproducible than that between biological replicates.
-This suggests that bottlenecking of the virus library during the low MOI passage contributes to much of the noise observed in our experiments, as we are only able to passage a finite number of viral particles.
+We first used the deep sequencing data for each replicate to independently estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
+Because there are 567 residues in HA, there are $567 \times 20 = 11,340$ estimated amino-acid preferences.
+The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
+The biological replicates are fairly well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
+Replicate 1 exhibited the lowest correlation with the other replicates, consistent with the observation that this replicate also showed the weakest selection against stop and nonsynonymous mutations perhaps indicating more experimental noise.
+The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity after the reverse-genetics step contributes most of the experimental noise.
 
@@
 
+Despite their large sequence divergence, how is it that H1 and H3 adopt nearly identical protein folds \comment{cite Ha, Russell}?
+
 \begin{figure}

### 2018-01-03 paper/manuscript.tex
Start moving manuscript over to _PNAS_ LaTex template

--- 2018-01-02 paper/manuscript.tex
@@
-% Use only LaTeX2e, calling the article.cls class and 12-point type.
+\documentclass[9pt,twocolumn,twoside]{pnas-new}
+% Use the lineno option to display guide line numbers if required.
+% Note that the use of elements such as single-column equations
+% may affect the guide line number alignment.
 
-\documentclass[11pt]{article}
-\usepackage[round,semicolon]{natbib}
-\usepackage[margin=1.25in]{geometry}
-\usepackage{kpfonts}
-
-\usepackage{seqsplit}
-\usepackage{placeins}
-
-\usepackage{newfloat}
-\usepackage[labelfont=bf]{caption}
-\usepackage{nameref}
-\usepackage{rotating}
-\usepackage{color}
-\usepackage{float}
-
-\setcounter{topnumber}{8}
-\setcounter{bottomnumber}{8}
-\setcounter{totalnumber}{8}
-\renewcommand{\topfraction}{1}
-\renewcommand{\bottomfraction}{1}
-\renewcommand{\textfraction}{0}
-\renewcommand{\floatpagefraction}{1}
-
-\usepackage[font=small,labelfont=bf]{caption}
-
-\usepackage{newfloat}
-\DeclareFloatingEnvironment[name={Figure}]{suppfigure}
-\renewcommand{\thesuppfigure}{S\arabic{suppfigure}}
-\DeclareFloatingEnvironment[name={Table}]{supptable}
-\renewcommand{\thesupptable}{S\arabic{supptable}}
-\DeclareFloatingEnvironment[name={File}]{suppfile}
-\renewcommand{\thesuppfile}{S\arabic{suppfile}}
-
-\definecolor{darkblue}{rgb}{0, 0.0, 0.6}
-
-\usepackage{hyperref}
-\hypersetup{colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue}
-
-\usepackage{seqsplit}
-
-\usepackage{array}
-\newcolumntype{R}[1]{>{\raggedright\arraybackslash}p{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\newcommand{\comment}[1]{{\color{red}[\textsl{#1}]}}
-
-\usepackage{setspace}
-
-\renewcommand{\topfraction}{1}
-\renewcommand{\bottomfraction}{1}
-\renewcommand{\textfraction}{0}
-\renewcommand{\floatpagefraction}{1}
+\templatetype{pnasresearcharticle} % Choose template 
+% {pnasresearcharticle} = Template for a two-column research article
+% {pnasmathematics} = Template for a one-column mathematics article
+% {pnasinvited} = Template for a PNAS invited submission
 
@@
 
-\author{
-Juhye M. Lee$^{1,4,5,\dagger}$, \and 
-John Huddleston$^{2,6,\dagger}$, \and 
-Michael B. Doud$^{1,4,5}$, \and 
-Kathryn A. Hooper$^{1,6}$, \and
-Trevor Bedford$^{2,3*}$ and \and 
-Jesse D. Bloom$^{1,3,4*}$
-\\
-\scriptsize{$^1$Basic Sciences Division, $^2$Vaccine and Infectious Diseases Division, and $^3$Computational Biology Program,} \\
-\scriptsize{Fred Hutchinson Cancer Research Center, Seattle, WA, USA} \\
-\scriptsize{$^4$Department of Genome Sciences, $^5$Medical Scientist Training Program, and $^6$Molecular and Cellular Biology Program,} \\
-\scriptsize{University of Washington, Seattle, WA, USA} \\
-\scriptsize{$^{\dagger}$These authors contributed equally} \\
-\scriptsize{$^*$Correspondence: \href{trevor@bedford.io}{trevor@bedford.io}, \href{jbloom@fredhutch.org}{jbloom@fredhutch.org}}
-}
+% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
+\author[a,d,e,1]{Juhye M. Lee}
+\author[b,f,1]{John Huddleston} 
+\author[a,d,e]{Michael B. Doud}
+\author[a,f]{Kathryn A. Hooper}
+\author[b,c,2]{Trevor Bedford}
+\author[a,c,d,2]{Jesse D. Bloom}
 
-\date{}
+\affil[a]{Basic Sciences Division}
+\affil[b]{Vaccine and Infectious Diseases Division}
+\affil[c]{and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
+\affil[d]{Department of Genome Sciences}
+\affil[e]{Medical Scientist Training Program}
+\affil[f]{and Molecular and Cellular Biology Program, University of Washington, Seattle, WA 98195, USA}
+
+% Please give the surname of the lead author for the running footer
+\leadauthor{Lead author last name}
+
+% Please add here a significance statement to explain the relevance of your work
+\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality.
+The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership.
+The Significance Statement appears in the paper itself and is required for all research papers.}
+
+% Please include corresponding author, author contribution and author declaration information
+\authorcontributions{Please provide details of author contributions here.}
+\authordeclaration{Please declare any conflict of interest here.}
+\equalauthors{\textsuperscript{1}J.M.L. and J.H. contributed equally to this work.}
+\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: trevor@bedford.io, jbloom@fredhutch.org}
+
+% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
+\keywords{Keyword 1 $|$ Keyword 2 $|$ Keyword 3 $|$ ...}
+
+\begin{abstract}
+Please provide an abstract of no more than 250 words in a single paragraph.
+Abstracts should explain to the general reader the major contributions of the article.
+References in the abstract must be cited in full within the abstract itself and cited in the text.
+\end{abstract}
+
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 
@@
 
+% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
+% You should only change this length when you've finalised the article contents.
+\verticaladjustment{-2pt}
+
 \maketitle
-\onehalfspacing
+\thispagestyle{firststyle}
+\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
 
-\begin{abstract}
-Abstract text.
-\end{abstract}
+% If your first paragraph (i.e.
+with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation.
+If this is the case, add \parshape=0 to the end of the list environment.
+\dropcap{T}his PNAS journal template is provided to help you write your work in the correct journal format.  Instructions for use are provided below.
 
-\section*{INTRODUCTION}
-Very rough outline:
-\begin{itemize}
-\item Mutations are rampant in the evolution of human influenza virus. Seasonal H3N2 influenza virus in particular rapidly accumulates mutations.
-\item The evolution of H3N2 is also characterized by clade competition and population turnover.
-\item There have been efforts to predict evolutionary success.
-\item Mutations that contribute to antigenic evolution largely determine strain success.
-\item RNA viruses can accumulate deleterious mutations, and because the influenza does not appreciably recombine, deleterious mutations are linked to beneficial ones.
-\item Mutations in HA that impact viral growth may influence strain success.
-\item We need to understand the functional impact of mutations in HA.
-\item Previously, we measured the effect of all possible single amino-acid mutations to an H1 hemagglutinin from the A/WSN/1933 (H1N1) strain~\citep{thyagarajan2014inherent,doud2016accurate}. 
-\item However, this is a highly lab-adapted strain, and the measurements in this strain may not be particularly relevant for studying mutational processes of more contemporaneous strains circulating in the human population.
-\item We chose to study the Perth/2009 H3 HA.
-\item This also enabled a comparison of how the preferences have shifted for two diverged HA's.
-\end{itemize}
+Note: please start your introduction without including the word ``Introduction'' as a section heading (except for math articles in the Physical Sciences section); this heading is implied in the first paragraphs.
 
-\section*{RESULTS}
+\section*{Results}
 \label{sec:results}
@@
 \subsection*{Deep mutational scanning of HA from a recent strain of human H3N2 influenza virus}
-We performed a deep mutational scan to measure the effects of all amino-acid mutations to the A/Perth/16/2009 (H3N2) HA on viral replication in cell culture. 
-This strain was the H3N2 component of the influenza vaccine from 2010-2012.
-We used a variant of this HA that supported better viral replication in cell culture than did the sequence in GenBank (Figure~\ref{suppfig:Perth2009_mut}).
-This HA variant contained two mutations: G78D and T212I.
-The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.
-%We first generated virus carrying the wildtype Perth/2009 HA and NA and the WSN internal genes by reverse genetics, and serially passaged the virus in cell culture a total of six times.
-%This passaging was completed to allow the virus to adapt to growth in cell culture and to find cell-culture adaptation mutations that emerged during the course of passaging.
 
-We mutagenized the entire gene at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone.
-We then generated mutant virus libraries from the mutant plasmids using a helper virus system that enables the efficient generation of complex virus libraries~\citep{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
-In order to reduce biosafety concerns and to maximize viral titers, we used an HA-deficient helper virus carrying WSN/1933 internal and NA genes to rescue the mutant viruses.
-Additionally, we used MDCK-SIAT1 cells constitutively expressing the TMPRSS2 protease, which facilitates HA cleavage and activation in the human airway~\citep{bottcher2006proteolytic, bottcher2010cleavage}.
-All of the experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B). 
-We also passaged and deep sequenced library 3 in technical replicate (denoted as library 3-1 and 3-2) to gauge to the amount of experimental noise occurring \textit{within} a single biological replicate.
-As a control, we used the unmutated gene to generate and passage viruses carrying wildtype HA.
+\subsection*{Submitting Manuscripts}
 
-Deep sequencing revealed selection against non-functional HA variants as evidenced by the reduced mutation frequencies of the mutant viruses compared to their starting frequencies in the mutant plasmids (Figure~\ref{fig:dms_overview}C).
-Specifically, stop codons were purged to 20-45\% of their original starting frequencies, after correcting for error rates from the wildtype controls.
-Although the majority of stop codons were purged, incomplete purging of stop codons is likely due to complementation of virions. 
-We also observed purging of nonsynonymous mutations to 30-40\% of their initial starting frequencies after error correction, suggesting strong selection against deleterious HA variants.
+All authors must submit their articles at \href{http://www.pnascentral.org/cgi-bin/main.plex}{PNAScentral}.
+If you are using Overleaf to write your article, you can use the ``Submit to PNAS'' option in the top bar of the editor window.
 
-We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
-We inferred the amino-acid preferences at each site in the Perth/2009 HA using the method described in~\cite{bloom2015software} and implemented in the \texttt{dms\_tools2} software [\url{https://jbloomlab.github.io/dms_tools2/}].
-These preferences represent an estimate of the 567 sites $\times$ 20 amino acids $=$ 11340 experimental measurements and are normalized to sum to one at each site.
-The correlations of the amino-acid preferences between each pair of replicates is shown in Figure~\ref{fig:dms_overview}D.
-The biological replicates are fairly well-correlated, with a Pearson's $R$ ranging from 0.69 to 0.78. 
-Replicate 1 exhibits the least amount of correlation with the other biological replicates, consistent with the observation that this replicate showed the weakest selection against stop and nonsynonymous mutations and might therefore be subject to more experimental noise.
-Of note, the two technical replicates 3-1 and 3-2 were only slightly more reproducible than that between biological replicates.
-This suggests that bottlenecking of the virus library during the low MOI passage contributes to much of the noise observed in our experiments, as we are only able to passage a finite number of viral particles.
+\subsection*{Format}
 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/dms_overview/dms_overview.pdf}}
-\caption{\label{fig:dms_overview}
-{\bf Deep mutational scanning experiments of H3 HA.}
-(A) We generated Perth/2009 HA mutant virus libraries using a helper virus approach described in~\cite{doud2016accurate}.
-We then passaged this pool of mutant viruses at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants. 
-Deep sequencing of the variants before and after selection allowed us to infer the H3 site-specific amino-acid preferences.
-(B) The DMS experiments were performed in biological triplicate, starting from independent preps of the wildtype HA genes to create the mutant plasmids. 
-We also passaged and deep sequenced library 3 in technical replicate, denoted 3-1 and 3-2, to estimate the experimental noise within a biological replicate.
-(C) Mutation frequencies of nonsynonymous, stop, and synonymous mutations for the mutant DNA, mutant virus, and unmutated DNA and virus controls. 
-There is selection against nonsynonymous and stop codons in the mutant viruses. 
-(D) The Pearson correlations for the amino-acid preferences between each pair of replicates are shown. 
-The correlation between the technical replicates is comparable to that between biological replicates, indicating that much of the experimental noise comes from passaging at low MOI. 
-}
-\end{figure}
+Many authors find it useful to organize their manuscripts with the following order of sections;  Title, Author Affiliation, Keywords, Abstract, Significance Statement, Results, Discussion, Materials and methods, Acknowledgments, and References.
+Other orders and headings are permitted.
 
-\subsection*{The HA preferences improve phylogenetic fit of natural H3 sequences}
-How well do the Perth/2009 HA preferences inferred from experimental measurements describe the evolution of H3N2 influenza virus in nature?
-This question can be addressed by evaluating how well experimentally informed codon substitution models (ExpCM's) constructed from our laboratory measurements improve phylogenetic fit of H3N2 evolution~\citep{hilton2017phydms}.
-The results in Table~\ref{tab:phydms} show that the ExpCM outperforms conventional substitution models in describing the evolution of human H3N2 HA. 
-The ExpCM also optimizes a stringency parameter for the preferences to more closely reflect the strength of selection in nature.
-The stringency parameter in the ExpCM is equal to 2.44, which indicates that although the same amino acids are preferred, the strength of selection is more stringent in nature than in our experiments.
-Figure~\ref{fig:logoplot} shows a logo plot of the Perth/2009 HA amino-acid preferences rescaled by this stringency parameter.
+\subsection*{Manuscript Length}
 
-\begin{table}
+PNAS generally uses a two-column format averaging 67 characters, including spaces, per line.
+The maximum length of a Direct Submission research article is six pages and a PNAS PLUS research article is ten pages including all text, spaces, and the number of characters displaced by figures, tables, and equations.
+When submitting tables, figures, and/or equations in addition to text, keep the text for your manuscript under 39,000 characters (including spaces) for Direct Submissions and 72,000 characters (including spaces) for PNAS PLUS.
+
+\subsection*{References}
+
+References should be cited in numerical order as they appear in text; this will be done automatically via bibtex, e.g.
+\cite{belkin2002using} and \cite{berard1994embedding,coifman2005geometric}.
+All references should be included in the main manuscript file.
+
+\subsection*{Data Archival}
+
+PNAS must be able to archive the data essential to a published article.
+Where such archiving is not possible, deposition of data in public databases, such as GenBank, ArrayExpress, Protein Data Bank, Unidata, and others outlined in the Information for Authors, is acceptable.
+
+\subsection*{Language-Editing Services}
+Prior to submission, authors who believe their manuscripts would benefit from professional editing are encouraged to use a language-editing service (see list at www.pnas.org/site/authors/language-editing.xhtml).
+PNAS does not take responsibility for or endorse these services, and their use has no bearing on acceptance of a manuscript for publication.
+
+%\begin{figure}%[tbhp]
+%\centering
+%\includegraphics[width=.8\linewidth]{frog}
+%\caption{Placeholder image of a frog with a long example caption to show justification setting.}
+%\label{fig:frog}
+%\end{figure}
+
+%\begin{SCfigure*}[\sidecaptionrelwidth][t]
+%\centering
+%\includegraphics[width=11.4cm,height=11.4cm]{frog}
+%\caption{This caption would be placed at the side of the figure, rather than below it.}\label{fig:side}
+%\end{SCfigure*}
+
+\subsection*{Digital Figures}
+\label{sec:figures}
+
+Only TIFF, EPS, and high-resolution PDF for Mac or PC are allowed for figures that will appear in the main text, and images must be final size.
+Authors may submit U3D or PRC files for 3D images; these must be accompanied by 2D representations in TIFF, EPS, or high-resolution PDF format.
+Color images must be in RGB (red, green, blue) mode.
+Include the font files for any text.
+
+Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.
+
+Figure \ref{fig:frog} shows an example of how to insert a column-wide figure.
+To insert a figure wider than one column, please use the \verb|\begin{figure*}...\end{figure*}| environment.
+Figures wider than one column should be sized to 11.4 cm or 17.8 cm wide.
+Use \verb|\begin{SCfigure*}...\end{SCfigure*}| for a wide figure with side captions.
+
+\subsection*{Single column equations}
+
+Authors may use 1- or 2-column equations in their article, according to their preference.
+
+To allow an equation to span both columns, options are to use the \verb|\begin{figure*}...\end{figure*}| environment mentioned above for figures, or to use the \verb|\begin{widetext}...\end{widetext}| environment as shown in equation \ref{eqn:example} below.
+
+Please note that this option may run into problems with floats and footnotes, as mentioned in the \href{http://texdoc.net/pkg/cuted}{cuted package documentation}.
+In the case of problems with footnotes, it may be possible to correct the situation using commands \verb|\footnotemark| and \verb|\footnotetext|.
+
+%% Do not use widetext if paper is in single column.
+\begin{widetext}
+\begin{align*}
+(x+y)^3&=(x+y)(x+y)^2\\
+       &=(x+y)(x^2+2xy+y^2) \numberthis \label{eqn:example} \\
+       &=x^3+3x^2y+3xy^3+x^3. 
+\end{align*}
+\end{widetext}
+
+\begin{table}%[tbhp]
 \centering
-\begin{tabular}{cccccccc}
-\hline
-\bf{Model} & \bf{$\Delta$AIC} & \bf{Log Likelihood} & \bf{Stringency} & \bf{$\omega$} & \bf{$\overline{\omega}$} & \bf{$\omega_{\alpha}$} & \bf{$\omega_{\beta}$} \\ \hline
-ExpCM & 0.0 & -8439.33 & $2.44$ & $0.91$ & -- & -- & -- \\
-Goldman-Yang M5 & 2166.06 & -9516.36 & -- & -- & $0.36$ & $0.30$ & $0.84$ \\
-ExpCM, averaged across sites & 2504.18 & -9691.42 & $0.68$ & $0.32$ & -- & -- & -- \\
-Goldman-Yang M0 & 2607.92 & -9738.29 & -- & $0.31$ & -- & -- & -- \\
-\hline
+\caption{Comparison of the fitted potential energy surfaces and ab initio benchmark electronic energy calculations}
+\begin{tabular}{lrrr}
+Species & CBS & CV & G3 \\
+\midrule
+1. Acetaldehyde & 0.0 & 0.0 & 0.0 \\
+2. Vinyl alcohol & 9.1 & 9.6 & 13.5 \\
+3. Hydroxyethylidene & 50.8 & 51.2 & 54.0\\
+\bottomrule
 \end{tabular}
-\caption{\label{tab:phydms}
-{\bf The site-specific amino-acid preferences are informative for describing human H3N2 evolution in nature.}
-We implemented several codon substitution models for phylogenetic fitting of an alignment of human H3N2 HA sequences. 
-The maximum likelihood values for each model were compared using the Akaike information criteria ($\Delta$AIC)~\citep{posada2004model}.
-An experimentally-informed codon substitution model (ExpCM) built from the preferences averaged across all replicates performs better than conventional substitution models, specifically the M0 and M5 models in~\cite{yang2000codon}.
-A non-site-specific ExpCM informed by preferences averaged across all sites performs comparably to the GY94 class of models, indicating the importance of site-specificity in the ExpCM.
-The optimized parameters for each model are also shown.
-}
+
+\addtabletext{nomenclature for the TSs refers to the numbered species in the table.}
 \end{table}
 
-A closer examination of the logo plot reveals that the preferences generally agree with existing knowledge about HA's biochemistry.
-For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\citep{waterfield1981disulphide} possess high preference for cysteine.
-At residues involved in receptor binding, there are strong preferences for the amino acids at sites Y98, D190, W153, and S228~\citep{weis1988structure,martin1998studies}.
-A positively charged amino acid at site 329 is important for cleavage activation of the HA0 precursor, and indeed this site exhibits a high preference for arginine~\citep{kido1992isolation, stech2005new}.
-However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine. 
-This codon is part of the signal peptide and is cleaved from the mature HA protein.
-One possible explanation for why we do not see a strong preference for Met at this site is due to alternative translation initiation occurring at a downstream or upstream start site, as has been described for HA~\citep{girard2011upstream}.
-% No upstream or downstream "ATG"'s are present.
-I did find an in-frame "ATC" one codon upstream of the canonical start site (in the non-coding region), and interestingly, all Perth/2009 HA sequences deposited in GenBank have an "ATT" in this non-coding position, including Seema's Perth HA.
-% I6M(HA2) mutant slightly increases the fusion pH (Daniels 1985 Cell)
-% We see high preference for T,S,Q over the wildtype Ile at site 226.
-Although Q226 in an H3 background is characterized as having preference for avian-type receptors, most of these studies were done in older H3 strains such as Aichi.
-Wu 2017 Cell Host Microbe has shown extensive epistasis in the RBS, specifically in the 220-loop, so perhaps the amino acids tolerated in this loop have shifted over time.
-Indeed the wildtype amino acids at 225 and 226 are different in Perth compared to HK68 and Aichi.
+\subsection*{Supporting Information (SI)}
 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}}
-\caption{\label{fig:logoplot}
-{\bf The site-specific amino-acid preferences of the Perth/2009 HA.}
-This logoplot shows the site-specific amino-acid preferences for the averaged replicates rescaled by the stringency parameter (Table 1) estimated by phydms.
-The height of each letter is proportional to its preference at that site, and the preferences for all sites are normalized to sum to 1.
-The sites are in H3 numbering.
-The top overlay bar shows the relative solvent accessibility.
-The bottom overlay bar is colored by the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
-The letters directly above each logo indicate the wildtype amino acid at that site.
+The main text of the paper must stand on its own without the SI.
+Refer to SI in the manuscript at an appropriate point in the text.
+Number supporting figures and tables starting with S1, S2, etc.
+Authors are limited to no more than 10 SI files, not including movie files.
+Authors who place detailed materials and methods in SI must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the online methods.
+If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.
+Because PNAS edits SI and composes it into a single PDF, authors must provide the following file formats only.
+
+\subsubsection*{SI Text}
+
+Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference).
+
+\subsubsection*{SI Figures}
+
+Provide a brief legend for each supporting figure after the supporting text.
+Provide figure images in TIFF, EPS, high-resolution PDF, JPEG, or GIF format; figures may not be embedded in manuscript text.
+When saving TIFF files, use only LZW compression; do not use JPEG compression.
+Do not save figure numbers, legends, or author names as part of the image.
+Composite figures must be pre-assembled.
+
+\subsubsection*{3D Figures}
+
+Supply a composable U3D or PRC file so that it may be edited and composed. Authors may submit a PDF file but please note it will be published in raw format and will not be edited or composed.
+
+\subsubsection*{SI Tables}
+
+Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference); include only one table per file.
+Do not use tabs or spaces to separate columns in Word tables.
+
+\subsubsection*{SI Datasets}
+
+Supply Excel (.xls), RTF, or PDF files. This file type will be published in raw format and will not be edited or composed.
+
+\subsubsection*{SI Movies}
+
+Supply Audio Video Interleave (avi), Quicktime (mov), Windows Media (wmv), animated GIF (gif), or MPEG files and submit a brief legend for each movie in a Word or RTF file.
+All movies should be submitted at the desired reproduction size and length.
+Movies should be no more than 10 MB in size.
+
+\subsubsection*{Still images}
+
+Authors must provide a still image from each video file. Supply TIFF, EPS, high-resolution PDF, JPEG, or GIF files.
+
+\subsubsection*{Appendices}
+
+PNAS prefers that authors submit individual source files to ensure readability.
+If this is not possible, supply a single PDF file that contains all of the SI associated with the paper.
+This file type will be published in raw format and will not be edited or composed.
+
+\matmethods{Please describe your materials and methods here.
+This can be more than one paragraph, and may contain subsections and equations as required.
+Authors should include a statement in the methods section describing how readers will be able to access the data in the paper.
+
+\subsection*{Subsection for Method}
+Example text for subsection.
 }
-\end{figure}
 
-\subsection*{The mutational tolerance of the head and the stalk shows less of a contrast in H3 than in H1}
-We next sought to investigate the inherent mutational tolerance of the Perth/2009 HA. 
-Figure~\ref{fig:mut_tolerance} shows the mutational tolerance as calculated from the rescaled Perth/2009 H3 preferences and the rescaled WSN/1933 H1 preferences mapped onto the HA crystal structures.
-We found antigenic site C and the most distal portion of the globular head near the 190-helix in the Perth/2009 H3 to be tolerant of mutations.
-Interestingly, the H3 stalk including the shorter $\alpha$-helix (helix A) is relatively mutationally tolerant compared to the tolerance of the globular head domain. 
-This observation suggests that the stalk may be prone to escape from antibodies, and agrees with previous work demonstrating that it is possible to select for antigenic mutants in H3 by broadly-neutralizing stalk-targeting antibodies~\citep{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human}.
-% In contrast, the head domain of the WSN/1933 H1 is more mutationally tolerant relative to its stalk domain.
+\showmatmethods{} % Display the Materials and Methods section
 
-The sites inside the receptor binding pockets are highly functionally constrained and were found to be relatively mutationally intolerant in both H3 and H1~\citep{wilson1981structure}.
-In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to antigenic evolution as these sites are under strong immune pressure.~\citep{wiley1981structural}.
+\acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}
 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/mut_tolerance/entropy_heatmap.pdf}}
-\caption{\label{fig:mut_tolerance}
-{\bf Mutational tolerance mapped onto each site of HA.}
-Mutational tolerance as calculated by the Shannon entropy of a given site's amino-acid preferences are mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}) and the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}), with both trimers in approximately the same orientation.
-The site entropies were calculated from the preferences measured in the Perth/2009 H3 (left panel) from this study, or the preferences measured in the WSN/1933 H1 (right panel) from~\cite{doud2016accurate}.
-Lighter shades of blue or red signify low mutational tolerance, while darker shades of blue or red signify high mutational tolerance. 
-For each HA, the structure on the left side colors the full HA trimer, while the structure on the right side colors only one of the monomers.
-The sialic acid receptor is shown as black sticks.
-The Perth/2009 H3 shows relatively high mutational tolerance in the stalk region, particularly in helix A, compared to the head region. 
-High mutational tolerance in H3 was also observed near antigenic site C and near the 190 helix.
-The head region of the WSN/1933 H1 is mutationally tolerant compared to the relatively intolerant stalk region. 
-}
-\end{figure}
+\showacknow{} % Display the acknowledgments section
 
-\subsection*{The experimental measurements can help discriminate successful influenza virus lineages}
-A challenge in vaccine strain selection is predicting which strain will dominate the upcoming influenza season.
-An important question to thereby address is if our experimental measurements are useful in differentiating between strains of human H3N2 influenza virus that have succeeded and those that have died out.
-To investigate this, we used our preference dataset and an H3N2 phylogeny from 1968-2012 (Figure~\ref{fig:trunkvssidebranch}A) to calculate the effects of mutations for the evolutionarily successful trunk lineage and for side branches which have died out.
-We found that strains with mutations measured to be more beneficial to viral growth in our experiments tend to succeed in nature.
-Figure~\ref{fig:trunkvssidebranch}B shows the effects of trunk and side branch mutations in five-year intervals for every year from 1968-2012. 
-On average, trunk mutations are towards more preferred amino acids compared to side branch mutations, and this was true for all intervals.
-Importantly, trunk mutations are significantly more favorable than side branch mutations when calculating the effects from the entirety of the phylogenetic tree (Figure~\ref{fig:trunkvssidebranch}C).
+% \pnasbreak splits and balances the columns before the references.
+% Uncomment \pnasbreak to view the references in the PNAS-style
+% If you see unexpected formatting errors, try commenting out \pnasbreak
+% as it can run into problems with floats and footnotes on the final page.
+%\pnasbreak
 
-Because tip nodes can contain egg- or cell-passaged isolates~\citep{wu2017structural,mcwhite2016sequence,skowronski2016mutations} and our experiments were performed in cell culture, we examined the effects of mutations on terminal side branch nodes to see if these would rank more highly than internal node mutations.
-Instead, terminal node mutations are on average towards less preferred amino acids than internal node mutations (Figure~\ref{fig:trunkvssidebranch}C), and both internal and terminal node mutations are significantly less favorable than those on the trunk.
-Therefore, strains that have accumulated mutations that we experimentally measured to be unpreferred generally die out in nature, while more favored mutations provide a selective advantage to the trunk.
-These findings demonstrate the importance of the functional impacts of HA mutations in determining strain success.
-
-\begin{figure}
-\centerline{\includegraphics[width=0.75\textwidth]{figs/trunkvssidebranch/trunkvssidebranch.pdf}}
-\caption{\label{fig:trunkvssidebranch}
-{\bf Mutations in evolutionary successful strains tend to be more favorable than in strains that die out.}
-(A) Phylogenetic tree of human H3N2 influenza virus from 1968-present. 
-The trunk is shown in red, and side branches are shown in blue.
-The gray branches represent the part of the tree for which we cannot yet distinguish the trunk from side branches.
-(B) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
-The median log$_{2}$ mutational effect in a given window is shown as circles for trunk mutations and triangles for side branch mutations. 
-The shaded region demarcates the interquartile range of trunk and side branch mutational effects.
-Negative numbers signify mutations towards less preferred amino acids, while positive numbers signify more preferred mutations.
-The median trunk mutational effects are consistently higher than the median side branch mutational effects for all windows.
-(C) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel) are shown.
-The preferences were randomized 10,000 times to estimate significance.
-The effects of trunk mutations are higher than side branch internal and terminal node mutations.
-}
-\end{figure}
-
-How distantly can the preferences be extended to describe differences in successful and unsuccessful strains?
-To explore this question, we scored the complete HA sequence of every node in the tree using the preferences by quantifying a \textit{sequence preference} metric.
-The sequence preference is defined as 
-$$\displaystyle\sum_{r} \ln \pi_{r, a}$$
-where $\pi_{r, a}$ is the preference for amino acid $a$ at site $r$.
-Consistent with the finding that trunk mutations are generally more favorable than side branch mutations, Figure~\ref{fig:sequence_preference} shows that the sequences of trunk nodes also tend to be more highly preferred than those of side branch nodes.
-Interestingly, the sequence preferences increase over time as the nodes approach the Perth/2009 strain and its closely related nodes, which all exhibit sequence preferences that are higher than that of the trunk.
-The observation that the strain in which we performed our deep mutational scan has one of the highest sequence preferences illustrates epistatic interactions among mutations such that an unpreferred mutation in one background may be preferred in another.
-% Due to the sheer number of possible pairwise combinations of mutations, it is impossible for us to experimentally characterize these epistatic interactions among mutations.
-
-\comment{this section will be fleshed out more once I figure out how to plot the substitutions from the root at epitope / non-epitope sites, as per Trevor's suggestion.}
-We reasoned that much of the increase in sequence preference toward Perth/2009 could be attributed to epitope sites, as most substitutions have occurred at sites under strong immune pressure.
-We therefore assessed each node's preference at epitope and non-epitope sites as defined by~\cite{wolf2006long}.
-Indeed, the preferences at epitope sites resembles those for all HA sites (Figure~\ref{fig:sequence_preference}), with preferences increasing over time.
-At non-epitope sites, however, the trunk preferences remain relatively constant while side branch preferences tend to drop below the trunk.
-These observations highlight the extensive epistasis among epitope sites, as has even been noted for sites within the 220-loop surrounding the receptor binding pocket~\citep{wu2017diversity}.
-Many successful epitope mutations are likely contingent upon the background in which they arise.
-Yet, we are able to distinguish trunk sequences as more favorable than side branch sequences, indicating that the preferences are still of utility over short evolutionary timescales.
-
-\begin{figure}
-\centerline{\includegraphics[width=0.75\textwidth]{figs/sequence_preference/sequence_preference.pdf}}
-\caption{\label{fig:sequence_preference}
-{\bf The HA sequences of trunk nodes tend to be more preferred than those of side branch nodes.}
-We used the preferences to calculate the HA sequence preference normalized by the number of sites of every node in a human H3N2 phylogenetic tree.
-To calculate the sequence preference, we used the entire HA sequence (top panel), only epitope sites (center panel) or non-epitope sites (bottom panel) as defined by~\cite{wolf2006long}.
-Higher preferences are closer to zero while lower preferences are more negative.
-The preferences at all sites and at epitope sites increase as the tree approaches the Perth/2009 strain, indicating epistasis among epitope sites.
-The trunk nodes generally exhibit higher sequence preferences than do side branch nodes.
-}
-\end{figure}
-
-Can we then distinguish lineage-specific mutational effects using the preferences measured in a distantly related HA homolog?
-We used the preferences measured previously in the WSN/1933 H1~\citep{doud2016accurate} to quantify the effects of H3 trunk and side branch mutations, shown in Figure~\ref{fig:WSN_trunkvssidebranch}.
-It is evident that we do not see trunk mutations significantly more favored than side branch mutations, suggesting that our ability to discriminate successful and unsuccessful strains degrades over sufficiently long evolutionary distances.
-
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}}
-\caption{\label{fig:WSN_trunkvssidebranch}
-{\bf The WSN/1933 H1 preferences do not reveal differences in trunk vs side branch mutational effects}
-(A) We used the WSN/1933 H1 preferences from~\cite{doud2016accurate} to calculate the log$_{2}$ mutational effects of trunk and side branch mutations from the inferred H3N2 phylogeny in Figure~\ref{fig:trunkvssidebranch}.
-There is not a significant difference in trunk vs side branch mutational effects.
-(B) We also performed a sliding window analysis with the WSN/1933 H1 preferences.
-There is not a distinct difference in trunk and side branch mutational effects.
-}
-\end{figure}
-
-\subsection*{The H1 and H3 preferences at many sites have shifted}
-How shifted are the preferences between evolutionarily distant homologs such as H1 and H3?
-Although we have previously compared the preferences between related protein homologs of NP~\citep{doud2015site} and of Env \comment{cite haddox2017}, the HA homologs we have experimentally studied are considerably more diverged than either of these pairs.
-The WSN/1933 H1 and the Perth/2009 H3 fall into two phylogenetically distinct groups and share 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A), providing an opportunity to analyze what has shifted and what has remained conserved.
-Simply correlating the preferences between H1 and H3 reveals that the replicate measurements within a homolog are more correlated than between homologs (Figure~\ref{fig:distance_distribution}B).
-
-However, to quantify the extent of mutational shifts at a site-by-site level, we used an approach described in \comment{haddox2017}.
-We aligned Perth/2009 H3 and WSN/1933 H1 and at each alignable site calculated the difference in preferences between homologs while correcting for experimental noise within homologs.
-The distribution of shifts is shown in Figure~\ref{fig:distance_distribution}C.
-Although many sites have small shifts near zero, a considerable number of sites have large shifts in preference, reaching a difference of 0.86 out of a maximum possible of 1.0.
-When we compare HA to the preferences measured in the non-homologous protein HIV Env \comment{haddox2017}, nearly all sites are shifted.
-On the other hand, when we randomize the HA replicates to generate a null distribution, there is very little shift in preferences. 
-For comparison, the preferences of the HIV Env homologs, which are 86\% identical at the amino-acid level, are mostly similar. 
-Although there are a small number of sites with larger shifts reaching a maximum of 0.52, we do not see a dramatic tail in the distribution as we see for the HA's.
-These observations suggest that as homologs diverge, their preferences increasingly shift.
-
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/distance_distribution/distance_distribution.pdf}}
-\caption{\label{fig:distance_distribution}
-{\bf The HA homologs exhibit many large shifts in preference compared to shifts for other viral protein homologs}
-(A) A phylogenetic tree of the HA subtypes, with the two HA's, WSN/1933 H1 and Perth/2009 H3, for which we have measured amino-acid preferences denoted on the tree. 
-The WSN/1933 H1 and the Perth/2009 H3 share $\sim$42\% amino-acid identity.
-(B) The correlations of the amino-acid preferences for replicates both within and between the two HA homologs. 
-The within-Perth/2009 and the within-WSN/1933 correlations are shown in blue and red, respectively.
-The between homolog correlations are in gray.
-The correlations for replicates within a homolog are higher than for replicates between homologs.
-(C) The distribution of shifts in preference for various homolog pairs are shown.
-The top two distributions show the distances between each of the HA homologs with the non-homologous HIV Env \comment{cite haddox2017}. 
-The center distribution shows the corrected distance between the two HIV Env homologs, which share 86\% amino-acid identity.
-The fourth distribution from the top is a null generated by randomizing the HA replicates and computing the distances.
-The bottom distribution is the corrected distance between the Perth/2009 H3 and WSN/1933 H1 homologs at all sites that align.
-}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/RMSD_heatmap/RMSD_heatmap.pdf}}
-\caption{\label{fig:RMSD_heatmap}
-{\bf Shifts in preferences mapped onto the structure of HA}
-(A) The preference shifts as calculated by $RMSD_{corrected}$ between the two HA homologs is mapped onto the structure of HA (PDB 4O5N;~\cite{lee2014receptor}). 
-The left structure shows the HA trimer, and the right structure colors one of the monomers. 
-The sialic acid receptor is shown in black sticks.
-Blue indicates small shifts in preference near zero, while red indicates large shifts in preference.
-The top ten most shifted sites are shown in spheres on the monomer.
-(B) The stalk domain was found to be significantly less shifted than the head domain (left plot).
-Sites absolutely conserved all 18 HA subtypes were also found to be significantly less shifted than the remaining non-conserved sites (right plot).
-(C) 
-}
-\end{figure}
-
-\section*{DISCUSSION}
-We have measured the effect of all possible single amino-acid mutations to Perth/2009 H3 on viral growth in cell culture.
-
-\clearpage
-\small
-
-\section*{METHODS}
-\label{sec:methods}
-\subsection*{HA numbering}
-Unless otherwise indicated, all sites are in H3 numbering, with the signal peptide in negative numbers, the HA1 subunit in plain numbers, and the HA2 subunit denoted with "(HA2)".
-The conversion between sequential numbering of the A/Perth/16/2009 HA and H3 numbering was performed using an HA numbering Python script (available at \url{https://github.com/jbloomlab/HA_numbering}).
-
-\subsection*{Creation of MDCK-SIAT1-TMPRSS2 cell line}
-The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter and attached to mCherry through an IRES...etc etc
-\comment{Need to look at Katie's notebooks for this...}
-
-\subsection*{Generation of HA codon mutant plasmid libraries}
-Recombinant A/Perth/16/2009 (HA, NA) $\times$ A/Puerto Rico/8/1934 influenza virus, NIB-64, NR-41803 was ordered from BEI Resources, NIAID, NIH. 
-Bulk RNA from the viral sample was extracted using the QIAamp Viral RNA Mini Kit (QIAGEN) according to manufacturer's instructions.
-The Perth/2009 HA and NA genes were then reverse transcribed, PCR amplified, and cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2 \comment{cite?} plasmid backbones.
-
-The codon-mutant libraries were generated using a PCR-based approach described in~\cite{dingens2017comprehensive}.
-
-\subsection*{Generation and passaging of mutant viruses}
-The mutant virus libraries were generated and passaged using the approach described in~\cite{doud2016accurate} with several modifications.
-
-\subsection*{Barcoded subamplicon sequencing}
-
-\subsection*{Analysis of deep sequencing data}
-
-\subsection*{Inference of phylogenetic trees}
-\comment{We downloaded X sequences from the Influenza Virus Resource ?.... etc. inferred the tree, ancestral state reconstruction, visualized the tree. Mark Perth/2009 on the tree}
-To parse out trunk mutations from side branch mutations, we first defined a set of recent nodes sampled on or after Jan. 1, 2017, and traced these nodes back to their most recent common ancestor.
-
-\subsection*{Quantification of mutational effects and sequence preferences from an H3N2 phylogeny}
-
-\subsection*{Data availability and source code}
-Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
-Computer code used to analyze the data and produce the results in the paper are in...
-
-\subsection*{ACKNOWLEDGMENTS}
-We thank Sarah Hilton, Hugh Haddox, Sidney Bell...the Fred Hutch Genomics Core...
-Funding...
-
-\bibliographystyle{mbe}
-\bibliography{references.bib}
-
-\clearpage
-\normalsize
-
-\section*{Supplementary Material}
-\FloatBarrier
-\pagenumbering{arabic}% resets `page` counter to 1
-\renewcommand*{\thepage}{S\arabic{page}}
-
-\begin{suppfigure}
-\centerline{\includegraphics[width=\textwidth]{figs/S01_G78D-T212I/G78D-T212I.pdf}}
-\caption{\label{suppfig:Perth2009_mut}
-{\bf Characterization of the G78D-T212I Perth/2009 HA variant.} 
-(A) 
-The G78D-T212I Perth/2009 HA variant grows to higher titers than do viruses carrying the wildtype Perth/2009 HA.
-Each virus was generated in duplicate by reverse genetics and passaged once at MOI = 0.01 in MDCK-SIAT1-TMPRSS2 cells.
-The rescue and passage viral supernatants were collected at 72 hours post-transfection and 44 hours post-infection, respectively, and titered in MDCK-SIAT1-TMPRSS2 cells. 
-The points mark each duplicate and the bar marks the mean.
-(B)
-The D78 variant remained at a low frequency in natural human H3N2 sequences over the past $~\sim$10 years.
-The A212 variant rose to fixation in $~\sim$2011, replacing the T212 variant.
-}
-\end{suppfigure}
+% Bibliography
+\bibliography{pnas-sample}
 

### 2018-01-03 paper/manuscript.tex
Move more of the manuscript over to the new LaTex doc

--- 2018-01-03 paper/manuscript.tex
@@
-% Use only LaTeX2e, calling the article.cls class and 12-point type.
+\documentclass[9pt,twocolumn,twoside]{pnas-new}
+% Use the lineno option to display guide line numbers if required.
+% Note that the use of elements such as single-column equations
+% may affect the guide line number alignment.
 
-\documentclass[11pt]{article}
-\usepackage[round,semicolon]{natbib}
-\usepackage[margin=1.25in]{geometry}
-\usepackage{kpfonts}
-
-\usepackage{seqsplit}
-\usepackage{placeins}
-
-\usepackage{newfloat}
-\usepackage[labelfont=bf]{caption}
-\usepackage{nameref}
-\usepackage{rotating}
-\usepackage{color}
-\usepackage{float}
-
-\setcounter{topnumber}{8}
-\setcounter{bottomnumber}{8}
-\setcounter{totalnumber}{8}
-\renewcommand{\topfraction}{1}
-\renewcommand{\bottomfraction}{1}
-\renewcommand{\textfraction}{0}
-\renewcommand{\floatpagefraction}{1}
-
-\usepackage[font=small,labelfont=bf]{caption}
-
-\usepackage{newfloat}
-\DeclareFloatingEnvironment[name={Figure}]{suppfigure}
-\renewcommand{\thesuppfigure}{S\arabic{suppfigure}}
-\DeclareFloatingEnvironment[name={Table}]{supptable}
-\renewcommand{\thesupptable}{S\arabic{supptable}}
-\DeclareFloatingEnvironment[name={File}]{suppfile}
-\renewcommand{\thesuppfile}{S\arabic{suppfile}}
-
-\definecolor{darkblue}{rgb}{0, 0.0, 0.6}
-
-\usepackage{hyperref}
-\hypersetup{colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue}
-
-\usepackage{seqsplit}
-
-\usepackage{array}
-\newcolumntype{R}[1]{>{\raggedright\arraybackslash}p{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\templatetype{pnasresearcharticle} % Choose template 
+% {pnasresearcharticle} = Template for a two-column research article
+% {pnasmathematics} = Template for a one-column mathematics article
+% {pnasinvited} = Template for a PNAS invited submission
 
@@
 
-\usepackage{setspace}
-
-\renewcommand{\topfraction}{1}
-\renewcommand{\bottomfraction}{1}
-\renewcommand{\textfraction}{0}
-\renewcommand{\floatpagefraction}{1}
-
 \title{Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fate of human H3N2 influenza virus lineages}
 
-\author{
-Juhye M. Lee$^{1,4,5,\dagger}$, \and 
-John Huddleston$^{2,6,\dagger}$, \and 
-Michael B. Doud$^{1,4,5}$, \and 
-Kathryn A. Hooper$^{1,6}$, \and
-Trevor Bedford$^{2,3*}$ and \and 
-Jesse D. Bloom$^{1,3,4*}$
-\\
-\scriptsize{$^1$Basic Sciences Division, $^2$Vaccine and Infectious Diseases Division, and $^3$Computational Biology Program,} \\
-\scriptsize{Fred Hutchinson Cancer Research Center, Seattle, WA, USA} \\
-\scriptsize{$^4$Department of Genome Sciences, $^5$Medical Scientist Training Program, and $^6$Molecular and Cellular Biology Program,} \\
-\scriptsize{University of Washington, Seattle, WA, USA} \\
-\scriptsize{$^{\dagger}$These authors contributed equally} \\
-\scriptsize{$^*$Correspondence: \href{trevor@bedford.io}{trevor@bedford.io}, \href{jbloom@fredhutch.org}{jbloom@fredhutch.org}}
-}
+% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
+\author[a,d,e,1]{Juhye M. Lee}
+\author[b,f,1]{John Huddleston} 
+\author[a,d,e]{Michael B. Doud}
+\author[a,f]{Kathryn A. Hooper}
+\author[b,c,2]{Trevor Bedford}
+\author[a,c,d,2]{Jesse D. Bloom}
 
-\date{}
+\affil[a]{Basic Sciences Division}
+\affil[b]{Vaccine and Infectious Diseases Division}
+\affil[c]{and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
+\affil[d]{Department of Genome Sciences}
+\affil[e]{Medical Scientist Training Program}
+\affil[f]{and Molecular and Cellular Biology Program, University of Washington, Seattle, WA 98195, USA}
+
+% Please give the surname of the lead author for the running footer
+\leadauthor{Lead author last name}
+
+% Please add here a significance statement to explain the relevance of your work
+\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality.
+The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership.
+The Significance Statement appears in the paper itself and is required for all research papers.}
+
+% Please include corresponding author, author contribution and author declaration information
+\authorcontributions{Please provide details of author contributions here.}
+\authordeclaration{Please declare any conflict of interest here.}
+\equalauthors{\textsuperscript{1}J.M.L. and J.H. contributed equally to this work.}
+\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: trevor@bedford.io, jbloom@fredhutch.org}
+
+% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
+\keywords{Keyword 1 $|$ Keyword 2 $|$ Keyword 3 $|$ ...}
+
+\begin{abstract}
+Please provide an abstract of no more than 250 words in a single paragraph.
+Abstracts should explain to the general reader the major contributions of the article.
+References in the abstract must be cited in full within the abstract itself and cited in the text.
+\end{abstract}
+
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 
@@
 We performed a deep mutational scan to measure the effects of all amino-acid mutations to HA from the A/Perth/16/2009 (H3N2) strain on viral replication in cell culture. 
-This strain was the H3N2 component of the influenza vaccine from 2010-2012~\comment{add citation}.
+This strain was the H3N2 component of the influenza vaccine from 2010-2012~\cite{who2010d,who2011}.
 Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral replication in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and \comment{supplementary file or dataset with our full reverse genetics plasmid sequence}).
@@
 We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (\comment{supplementary figure with Sanger sequencing}).
-We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables the efficient generation of complex influenza virus libraries~\citep{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
+We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables the efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
 These mutant viruses derived all of their non-HA genes from the lab-adapted WSN/1933 strain.
 Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
-To further increase viral titers, we used MDCK-SIAT1 cells that we had engineered to constitutively express the TMPRSS2 protease, which facilitates HA cleavage and activation~\citep{bottcher2006proteolytic, bottcher2010cleavage}.
+To further increase viral titers, we used MDCK-SIAT1 cells that we had engineered to constitutively express the TMPRSS2 protease, which facilitates HA cleavage and activation~\cite{bottcher2006proteolytic, bottcher2010cleavage}.
 
@@
 
-We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
-We first used the deep sequencing data for each replicate to independently estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
-Because there are 567 residues in HA, there are $567 \times 20 = 11,340$ estimated amino-acid preferences.
-The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
-The biological replicates are fairly well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
-Replicate 1 exhibited the lowest correlation with the other replicates, consistent with the observation that this replicate also showed the weakest selection against stop and nonsynonymous mutations perhaps indicating more experimental noise.
-The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity after the reverse-genetics step contributes most of the experimental noise.
-
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/dms_overview/dms_overview.pdf}}
+\begin{figure*}
+\centering
+\includegraphics[width=13cm]{figs/dms_overview/dms_overview.pdf}
 \caption{\label{fig:dms_overview}
@@
 }
-\end{figure}
+\end{figure*}
+
+We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
+We first used the deep sequencing data for each replicate to independently estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
+Because there are 567 residues in HA, there are $567 \times 20 = 11,340$ estimated amino-acid preferences.
+The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
+The biological replicates are fairly well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
+Replicate 1 exhibited the lowest correlation with the other replicates, consistent with the observation that this replicate also showed the weakest selection against stop and nonsynonymous mutations perhaps indicating more experimental noise.
+The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity after the reverse-genetics step contributes most of the experimental noise.
 
@@
 How well do the Perth/2009 HA preferences inferred from experimental measurements describe the evolution of H3N2 influenza virus in nature?
-This question can be addressed by evaluating how well experimentally informed codon substitution models (ExpCM's) constructed from our laboratory measurements improve phylogenetic fit of H3N2 evolution~\citep{hilton2017phydms}.
+This question can be addressed by evaluating how well experimentally informed codon substitution models (ExpCM's) constructed from our laboratory measurements improve phylogenetic fit of H3N2 evolution~\cite{hilton2017phydms}.
 The results in Table~\ref{tab:phydms} show that the ExpCM outperforms conventional substitution models in describing the evolution of human H3N2 HA. 
@@
 
-\begin{table}
+\begin{figure*}
 \centering
-{\small
+\caption{\label{tab:phydms}
+{\bf The site-specific amino-acid preferences are informative for describing human H3N2 evolution in nature.}}
 \begin{tabular}{cccccccc}
@@
 \end{tabular}
-}
-\caption{\label{tab:phydms}
-{\bf The site-specific amino-acid preferences are informative for describing human H3N2 evolution in nature.}
-We implemented several codon substitution models for phylogenetic fitting of an alignment of human H3N2 HA sequences. 
-The maximum likelihood values for each model were compared using the Akaike information criteria ($\Delta$AIC)~\citep{posada2004model}.
+
+\addtabletext{We implemented several codon substitution models for phylogenetic fitting of an alignment of human H3N2 HA sequences. 
+The maximum likelihood values for each model were compared using the Akaike information criteria ($\Delta$AIC)~\cite{posada2004model}.
 An experimentally-informed codon substitution model (ExpCM) built from the preferences averaged across all replicates performs better than conventional substitution models, specifically the M0 and M5 models in~\cite{yang2000codon}.
 A non-site-specific ExpCM informed by preferences averaged across all sites performs comparably to the GY94 class of models, indicating the importance of site-specificity in the ExpCM.
-The optimized parameters for each model are also shown.
-}
-\end{table}
+The optimized parameters for each model are also shown.}
+\end{figure*}
 
 A closer examination of the logo plot reveals that the preferences generally agree with existing knowledge about HA's biochemistry.
-For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\citep{waterfield1981disulphide} possess high preference for cysteine.
-At residues involved in receptor binding, there are strong preferences for the amino acids at sites Y98, D190, W153, and S228~\citep{weis1988structure,martin1998studies}.
-A positively charged amino acid at site 329 is important for cleavage activation of the HA0 precursor, and indeed this site exhibits a high preference for arginine~\citep{kido1992isolation, stech2005new}.
+For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} possess high preference for cysteine.
+At residues involved in receptor binding, there are strong preferences for the amino acids at sites Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies}.
+A positively charged amino acid at site 329 is important for cleavage activation of the HA0 precursor, and indeed this site exhibits a high preference for arginine~\cite{kido1992isolation, stech2005new}.
 However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine. 
 This codon is part of the signal peptide and is cleaved from the mature HA protein.
-One possible explanation for why we do not see a strong preference for Met at this site is due to alternative translation initiation occurring at a downstream or upstream start site, as has been described for HA~\citep{girard2011upstream}.
+One possible explanation for why we do not see a strong preference for Met at this site is due to alternative translation initiation occurring at a downstream or upstream start site, as has been described for HA~\cite{girard2011upstream}.
 % No upstream or downstream "ATG"'s are present.
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}}
+\begin{figure*}
+\centering
+\includegraphics[width=17cm]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}
 \caption{\label{fig:logoplot}
@@
 }
-\end{figure}
+\end{figure*}
 
@@
 Interestingly, the H3 stalk including the shorter $\alpha$-helix (helix A) is relatively mutationally tolerant compared to the tolerance of the globular head domain. 
-This observation suggests that the stalk may be prone to escape from antibodies, and agrees with previous work demonstrating that it is possible to select for antigenic mutants in H3 by broadly-neutralizing stalk-targeting antibodies~\citep{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human}.
+This observation suggests that the stalk may be prone to escape from antibodies, and agrees with previous work demonstrating that it is possible to select for antigenic mutants in H3 by broadly-neutralizing stalk-targeting antibodies~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human}.
 % In contrast, the head domain of the WSN/1933 H1 is more mutationally tolerant relative to its stalk domain.
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/mut_tolerance/entropy_heatmap.pdf}}
+\begin{figure*}
+\centering
+\includegraphics[width=16cm]{figs/mut_tolerance/entropy_heatmap.pdf}
 \caption{\label{fig:mut_tolerance}
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=0.75\textwidth]{figs/trunkvssidebranch/trunkvssidebranch.pdf}}
+\begin{figure*}
+\centering
+\includegraphics[width=12cm]{figs/trunkvssidebranch/trunkvssidebranch.pdf}
 \caption{\label{fig:trunkvssidebranch}
@@
 The sequence preference is defined as 
-$$\displaystyle\sum_{r} \ln \pi_{r, a}$$
+$\displaystyle\sum_{r} \ln \pi_{r, a}$
 where $\pi_{r, a}$ is the preference for amino acid $a$ at site $r$.
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=0.75\textwidth]{figs/sequence_preference/sequence_preference.pdf}}
+\begin{figure*}
+\centering
+\includegraphics[width=12cm]{figs/sequence_preference/sequence_preference.pdf}
 \caption{\label{fig:sequence_preference}
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}}
+\begin{figure*}
+\centering
+\includegraphics[width=15cm]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}
 \caption{\label{fig:WSN_trunkvssidebranch}
@@
 }
-\end{figure}
+\end{figure*}
 
-\subsection*{The H1 and H3 preferences at many sites have shifted}
+\subsection*{The H1 and H3 preferences have shifted at many sites}
 How shifted are the preferences between evolutionarily distant homologs such as H1 and H3?
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/distance_distribution/distance_distribution.pdf}}
+\begin{figure*}
+\centering
+\includegraphics[width=16cm]{figs/distance_distribution/distance_distribution.pdf}
 \caption{\label{fig:distance_distribution}
@@
 }
-\end{figure}
+\end{figure*}
 
-\begin{figure}
-\centerline{\includegraphics[width=\textwidth]{figs/RMSD_heatmap/RMSD_heatmap.pdf}}
+\begin{figure*}
+\centering
+\includegraphics[width=16cm]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
 \caption{\label{fig:RMSD_heatmap}
@@
 }
-\end{figure}
+\end{figure*}
 
-\section*{DISCUSSION}
+\section*{Discussion}
+\label{sec:discussion}
 We have measured the effect of all possible single amino-acid mutations to Perth/2009 H3 on viral growth in cell culture.
 
-\clearpage
-\small
+\subsection*{References}
 
-\section*{METHODS}
-\label{sec:methods}
-\subsection*{HA numbering}
-Unless otherwise indicated, all sites are in H3 numbering, with the signal peptide in negative numbers, the HA1 subunit in plain numbers, and the HA2 subunit denoted with "(HA2)".
-The conversion between sequential numbering of the A/Perth/16/2009 HA and H3 numbering was performed using an HA numbering Python script (available at \url{https://github.com/jbloomlab/HA_numbering}).
+References should be cited in numerical order as they appear in text; this will be done automatically via bibtex, e.g.
+\cite{belkin2002using} and \cite{berard1994embedding,coifman2005geometric}.
+All references should be included in the main manuscript file.
 
-\subsection*{Creation of MDCK-SIAT1-TMPRSS2 cell line}
-The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter and attached to mCherry through an IRES...etc etc
-\comment{Need to look at Katie's notebooks for this...}
+\subsection*{Supporting Information (SI)}
 
-\subsection*{Generation of HA codon mutant plasmid libraries}
-Recombinant A/Perth/16/2009 (HA, NA) $\times$ A/Puerto Rico/8/1934 influenza virus, NIB-64, NR-41803 was ordered from BEI Resources, NIAID, NIH. 
-Bulk RNA from the viral sample was extracted using the QIAamp Viral RNA Mini Kit (QIAGEN) according to manufacturer's instructions.
-The Perth/2009 HA and NA genes were then reverse transcribed, PCR amplified, and cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2 \comment{cite?} plasmid backbones.
+The main text of the paper must stand on its own without the SI.
+Refer to SI in the manuscript at an appropriate point in the text.
+Number supporting figures and tables starting with S1, S2, etc.
+Authors are limited to no more than 10 SI files, not including movie files.
+Authors who place detailed materials and methods in SI must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the online methods.
+If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.
+Because PNAS edits SI and composes it into a single PDF, authors must provide the following file formats only.
 
-The codon-mutant libraries were generated using a PCR-based approach described in~\cite{dingens2017comprehensive}.
+\subsubsection*{SI Text}
 
-\subsection*{Generation and passaging of mutant viruses}
-The mutant virus libraries were generated and passaged using the approach described in~\cite{doud2016accurate} with several modifications.
+Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference).
 
-\subsection*{Barcoded subamplicon sequencing}
+\subsubsection*{SI Figures}
 
-\subsection*{Analysis of deep sequencing data}
+Provide a brief legend for each supporting figure after the supporting text.
+Provide figure images in TIFF, EPS, high-resolution PDF, JPEG, or GIF format; figures may not be embedded in manuscript text.
+When saving TIFF files, use only LZW compression; do not use JPEG compression.
+Do not save figure numbers, legends, or author names as part of the image.
+Composite figures must be pre-assembled.
 
-\subsection*{Inference of phylogenetic trees}
-\comment{We downloaded X sequences from the Influenza Virus Resource ?.... etc. inferred the tree, ancestral state reconstruction, visualized the tree. Mark Perth/2009 on the tree}
-To parse out trunk mutations from side branch mutations, we first defined a set of recent nodes sampled on or after Jan. 1, 2017, and traced these nodes back to their most recent common ancestor.
+\subsubsection*{SI Tables}
 
-\subsection*{Quantification of mutational effects and sequence preferences from an H3N2 phylogeny}
+Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference); include only one table per file.
+Do not use tabs or spaces to separate columns in Word tables.
 
-\subsection*{Data availability and source code}
-Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
-Computer code used to analyze the data and produce the results in the paper are in...
+\subsubsection*{SI Datasets}
 
-\subsection*{ACKNOWLEDGMENTS}
-We thank Sarah Hilton, Hugh Haddox, Sidney Bell...the Fred Hutch Genomics Core...
-Funding...
+Supply Excel (.xls), RTF, or PDF files. This file type will be published in raw format and will not be edited or composed.
 
-\bibliographystyle{mbe}
-\bibliography{references.bib}
+\subsubsection*{Appendices}
 
-\clearpage
-\normalsize
+PNAS prefers that authors submit individual source files to ensure readability.
+If this is not possible, supply a single PDF file that contains all of the SI associated with the paper.
+This file type will be published in raw format and will not be edited or composed.
 
-\section*{Supplementary Material}
-\FloatBarrier
-\pagenumbering{arabic}% resets `page` counter to 1
-\renewcommand*{\thepage}{S\arabic{page}}
+\matmethods{Please describe your materials and methods here.
+This can be more than one paragraph, and may contain subsections and equations as required.
+Authors should include a statement in the methods section describing how readers will be able to access the data in the paper.
 
-\begin{suppfigure}
-\centerline{\includegraphics[width=\textwidth]{figs/S01_G78D-T212I/G78D-T212I.pdf}}
-\caption{\label{suppfig:Perth2009_mut}
-{\bf Characterization of the G78D-T212I Perth/2009 HA variant.} 
-(A) 
-The G78D-T212I Perth/2009 HA variant grows to higher titers than do viruses carrying the wildtype Perth/2009 HA.
-Each virus was generated in duplicate by reverse genetics and passaged once at MOI = 0.01 in MDCK-SIAT1-TMPRSS2 cells.
-The rescue and passage viral supernatants were collected at 72 hours post-transfection and 44 hours post-infection, respectively, and titered in MDCK-SIAT1-TMPRSS2 cells. 
-The points mark each duplicate and the bar marks the mean.
-(B)
-The D78 variant remained at a low frequency in natural human H3N2 sequences over the past $~\sim$10 years.
-The A212 variant rose to fixation in $~\sim$2011, replacing the T212 variant.
+\subsection*{Subsection for Method}
+Example text for subsection.
 }
-\end{suppfigure}
+
+\showmatmethods{} % Display the Materials and Methods section
+
+\acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}
+
+\showacknow{} % Display the acknowledgments section
+
+% \pnasbreak splits and balances the columns before the references.
+% Uncomment \pnasbreak to view the references in the PNAS-style
+% If you see unexpected formatting errors, try commenting out \pnasbreak
+% as it can run into problems with floats and footnotes on the final page.
+%\pnasbreak
+
+% Bibliography
+\bibliography{pnas-sample}
 

### 2018-01-03 paper/manuscript.tex
Add pct identity to distance_distribution fig
Cleaned up LaTex for PNAS format

@@
 % {pnasinvited} = Template for a PNAS invited submission
+
+% added to enable supplementary figures
+\usepackage{placeins}
+\usepackage{newfloat}
+\DeclareFloatingEnvironment[name={Figure}]{suppfigure}
+\renewcommand{\thesuppfigure}{S\arabic{suppfigure}}
 
@@
 \centering
-\includegraphics[width=16cm]{figs/distance_distribution/distance_distribution.pdf}
+\includegraphics[width=17cm]{figs/distance_distribution/distance_distribution.pdf}
 \caption{\label{fig:distance_distribution}
@@
 
-\subsection*{References}
-
-References should be cited in numerical order as they appear in text; this will be done automatically via bibtex, e.g.
-\cite{belkin2002using} and \cite{berard1994embedding,coifman2005geometric}.
-All references should be included in the main manuscript file.
-
-\subsection*{Supporting Information (SI)}
-
-The main text of the paper must stand on its own without the SI.
-Refer to SI in the manuscript at an appropriate point in the text.
-Number supporting figures and tables starting with S1, S2, etc.
-Authors are limited to no more than 10 SI files, not including movie files.
-Authors who place detailed materials and methods in SI must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the online methods.
-If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.
-Because PNAS edits SI and composes it into a single PDF, authors must provide the following file formats only.
-
-\subsubsection*{SI Text}
-
-Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference).
-
-\subsubsection*{SI Figures}
-
-Provide a brief legend for each supporting figure after the supporting text.
-Provide figure images in TIFF, EPS, high-resolution PDF, JPEG, or GIF format; figures may not be embedded in manuscript text.
-When saving TIFF files, use only LZW compression; do not use JPEG compression.
-Do not save figure numbers, legends, or author names as part of the image.
-Composite figures must be pre-assembled.
-
-\subsubsection*{SI Tables}
-
-Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference); include only one table per file.
-Do not use tabs or spaces to separate columns in Word tables.
-
-\subsubsection*{SI Datasets}
-
-Supply Excel (.xls), RTF, or PDF files. This file type will be published in raw format and will not be edited or composed.
-
-\subsubsection*{Appendices}
-
-PNAS prefers that authors submit individual source files to ensure readability.
-If this is not possible, supply a single PDF file that contains all of the SI associated with the paper.
-This file type will be published in raw format and will not be edited or composed.
-
 \matmethods{Please describe your materials and methods here.
@@
 % Bibliography
-\bibliography{pnas-sample}
+\subsection*{References}
+\bibliography{references}
+
+% Supplementary material temporarily moved until AFTER everything else for initial submission
+
+\onecolumn
+
+\subsection*{Supporting Information (SI)}
+\FloatBarrier
+
+The main text of the paper must stand on its own without the SI.
+Refer to SI in the manuscript at an appropriate point in the text.
+Number supporting figures and tables starting with S1, S2, etc.
+Authors are limited to no more than 10 SI files, not including movie files.
+Authors who place detailed materials and methods in SI must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the online methods.
+If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.
+Because PNAS edits SI and composes it into a single PDF, authors must provide the following file formats only.
+
+\subsubsection*{SI Text}
+
+Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference).
+
+\subsubsection*{SI Figures}
+
+\begin{suppfigure}
+\centerline{\includegraphics[width=0.5\textwidth]{figs/S01_G78D-T212I/G78D-T212I.pdf}}
+\caption{\label{suppfig:Perth2009_mut}
+{\bf Characterization of the G78D-T212I Perth/2009 HA variant.} 
+(A) 
+The G78D-T212I Perth/2009 HA variant grows to higher titers than do viruses carrying the wildtype Perth/2009 HA.
+Each virus was generated in duplicate by reverse genetics and passaged once at MOI = 0.01 in MDCK-SIAT1-TMPRSS2 cells.
+The rescue and passage viral supernatants were collected at 72 hours post-transfection and 44 hours post-infection, respectively, and titered in MDCK-SIAT1-TMPRSS2 cells. 
+The points mark each duplicate and the bar marks the mean.
+(B)
+The D78 variant remained at a low frequency in natural human H3N2 sequences over the past $~\sim$10 years.
+The A212 variant rose to fixation in $~\sim$2011, replacing the T212 variant.
+}
+\end{suppfigure}
+
+\subsubsection*{SI Tables}
+
+Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference); include only one table per file.
+Do not use tabs or spaces to separate columns in Word tables.
+
+\subsubsection*{SI Datasets}
+
+Supply Excel (.xls), RTF, or PDF files. This file type will be published in raw format and will not be edited or composed.
 

### 2018-01-03 paper/manuscript.tex
some minor tweaks

@@
 \renewcommand{\thesuppfigure}{S\arabic{suppfigure}}
+
+\DeclareFloatingEnvironment[name={Dataset}]{suppdata}
+\renewcommand{\thesuppdata}{S\arabic{suppdata}}
 
@@
 % Please give the surname of the lead author for the running footer
-\leadauthor{Lead author last name}
+\leadauthor{Lee}
 
@@
 \equalauthors{\textsuperscript{1}J.M.L. and J.H. contributed equally to this work.}
-\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: trevor@bedford.io, jbloom@fredhutch.org}
+\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed: trevor@bedford.io, jbloom@fredhutch.org}
 
 % Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
-\keywords{Keyword 1 $|$ Keyword 2 $|$ Keyword 3 $|$ ...}
+\keywords{influenza virus $|$ hemagglutinin $|$ deep mutational scanning $|$ antigenic drift $|$ epistasis}
 
@@
 If this is the case, add \parshape=0 to the end of the list environment.
-\dropcap{T}his PNAS journal template is provided to help you write your work in the correct journal format.  Instructions for use are provided below.
-
-Note: please start your introduction without including the word ``Introduction'' as a section heading (except for math articles in the Physical Sciences section); this heading is implied in the first paragraphs.
+\dropcap{V}ery rough outline:
+\begin{itemize}
+\item Mutations are rampant in the evolution of human influenza virus. Seasonal H3N2 influenza virus in particular rapidly accumulates mutations.
+\item The evolution of H3N2 is also characterized by clade competition and population turnover.
+\item There have been efforts to predict evolutionary success.
+\item Mutations that contribute to antigenic evolution largely determine strain success.
+\item RNA viruses can accumulate deleterious mutations, and because the influenza does not appreciably recombine, deleterious mutations are linked to beneficial ones.
+\item Mutations in HA that impact viral growth may influence strain success.
+\item We need to understand the functional impact of mutations in HA.
+\item Previously, we measured the effect of all possible single amino-acid mutations to an H1 hemagglutinin from the A/WSN/1933 (H1N1) strain~\citep{thyagarajan2014inherent,doud2016accurate}. 
+\item However, this is a highly lab-adapted strain, and the measurements in this strain may not be particularly relevant for studying mutational processes of more contemporaneous strains circulating in the human population.
+\item We chose to study the Perth/2009 H3 HA.
+\item This also enabled a comparison of how the preferences have shifted for two diverged HA's.
+\end{itemize}
 
@@
 This strain was the H3N2 component of the influenza vaccine from 2010-2012~\cite{who2010d,who2011}.
-Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral replication in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and \comment{supplementary file or dataset with our full reverse genetics plasmid sequence}).
+Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral replication in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and Dataset~\ref{suppdata:PerthHA}).
 The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.
@@
 \caption{\label{fig:dms_overview}
-{\bf Deep mutational scanning experiments of H3 HA.}
-(A) We generated Perth/2009 HA mutant virus libraries using a helper virus approach described in~\cite{doud2016accurate}.
-We then passaged this pool of mutant viruses at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants. 
-Deep sequencing of the variants before and after selection allowed us to infer the H3 site-specific amino-acid preferences.
-(B) The DMS experiments were performed in biological triplicate, starting from independent preps of the wildtype HA genes to create the mutant plasmids. 
-We also passaged and deep sequenced library 3 in technical replicate, denoted 3-1 and 3-2, to estimate the experimental noise within a biological replicate.
-(C) Mutation frequencies of nonsynonymous, stop, and synonymous mutations for the mutant DNA, mutant virus, and unmutated DNA and virus controls. 
-There is selection against nonsynonymous and stop codons in the mutant viruses. 
-(D) The Pearson correlations for the amino-acid preferences between each pair of replicates are shown. 
-The correlation between the technical replicates is comparable to that between biological replicates, indicating that much of the experimental noise comes from passaging at low MOI. 
+{\bf Deep mutational scanning of the Perth/2009 H3 HA.}
+(A) We generated mutant virus libraries using a helper-virus approach~\cite{doud2016accurate}, and passaged them at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants.
+Deep sequencing of the variants before and after selection allowed us to estimate each site's amino-acid preferences.
+(B) The experiments were performed in full biological triplicate. 
+We also passaged and deep sequenced library 3 in duplicate.
+(C) Frequencies of nonsynonymous, stop, and synonymous mutations in the mutant plasmid DNA, the passaged mutant viruses, and wildtype DNA and virus controls. 
+(D) The Pearson correlations among the amino-acid preferences estimated in each replicate. 
 }
@@
 
-\subsection*{The HA preferences improve phylogenetic fit of natural H3 sequences}
-How well do the Perth/2009 HA preferences inferred from experimental measurements describe the evolution of H3N2 influenza virus in nature?
-This question can be addressed by evaluating how well experimentally informed codon substitution models (ExpCM's) constructed from our laboratory measurements improve phylogenetic fit of H3N2 evolution~\cite{hilton2017phydms}.
-The results in Table~\ref{tab:phydms} show that the ExpCM outperforms conventional substitution models in describing the evolution of human H3N2 HA. 
+\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
+How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
+This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describe H3N2 evolution compared to standard phylogenetic substitution models~\cite{hilton2017phydms}.
+Table~\ref{tab:phydms} shows that the ExpCM greatly outperforms conventional substitution models, showing that our experiments authentically capture some of the constraints on HA evolution. 
+\comment{Jesse and Juhye are here in sentence-level editing} 
 The ExpCM also optimizes a stringency parameter for the preferences to more closely reflect the strength of selection in nature.
@@
 
-Supply Excel (.xls), RTF, or PDF files. This file type will be published in raw format and will not be edited or composed.
+\begin{suppdata}
+\caption{\label{suppdata:PerthHA}
+Genbank file giving the full sequence of the bidirectional reverse-genetics plasmid pHW-Perth2009-HA-G78D-T212I, which encodes the wildtype HA sequence used in this study.
+}
+\end{suppdata}
 

### 2018-01-04 paper/manuscript.tex
Change coloring of mut tolerance heatmaps
Added small details to FigS1 caption

@@
 The site entropies were calculated from the preferences measured in the Perth/2009 H3 (left panel) from this study, or the preferences measured in the WSN/1933 H1 (right panel) from~\cite{doud2016accurate}.
-Lighter shades of blue or red signify low mutational tolerance, while darker shades of blue or red signify high mutational tolerance. 
+Lighter shades of purple or brown signify low mutational tolerance, while darker shades signify high mutational tolerance. 
 For each HA, the structure on the left side colors the full HA trimer, while the structure on the right side colors only one of the monomers.
@@
 (A) 
-The G78D-T212I Perth/2009 HA variant grows to higher titers than do viruses carrying the wildtype Perth/2009 HA.
-Each virus was generated in duplicate by reverse genetics and passaged once at MOI = 0.01 in MDCK-SIAT1-TMPRSS2 cells.
+The G78D-T212I Perth/2009 HA variant supports better viral growth than the wildtype Perth/2009 HA.
+Viruses were generated in duplicate by reverse genetics with the Perth/2009 NA and WSN internal genes, and passaged once at MOI = 0.01 in MDCK-SIAT1-TMPRSS2 cells.
 The rescue and passage viral supernatants were collected at 72 hours post-transfection and 44 hours post-infection, respectively, and titered in MDCK-SIAT1-TMPRSS2 cells. 

### 2018-01-04 paper/manuscript.tex
Add H3N2 phylogeny figure to intro
Add bare bones of materials and methods to manuscript

@@
 \end{itemize}
+
+\begin{SCfigure*}[\sidecaptionrelwidth][t]
+\centering
+\includegraphics[width=11.4cm]{figs/H3N2_phylogeny/H3N2_phylogeny.pdf}
+\caption{\label{fig:H3N2_phylogeny}
+{\bf Human H3N2 HA phylogeny.}
+}
+\end{SCfigure*}
 
@@
 
-\subsection*{Subsection for Method}
-Example text for subsection.
+\subsection*{HA numbering}
+Unless otherwise indicated, all sites are in H3 numbering, with the signal peptide in negative numbers, the HA1 subunit in plain numbers, and the HA2 subunit denoted with "(HA2)".
+The conversion between sequential numbering of the A/Perth/16/2009 HA and H3 numbering was performed using an HA numbering Python script (available at \url{https://github.com/jbloomlab/HA_numbering}).
+
+\subsection*{Creation of MDCK-SIAT1-TMPRSS2 cell line}
+The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter and attached to mCherry through an IRES...etc etc
+\comment{Need to look at Katie's notebooks for this...}
+
+\subsection*{Generation of HA codon mutant plasmid libraries}
+Recombinant A/Perth/16/2009 (HA, NA) $\times$ A/Puerto Rico/8/1934 influenza virus, NIB-64, NR-41803 was ordered from BEI Resources, NIAID, NIH. 
+Bulk RNA from the viral sample was extracted using the QIAamp Viral RNA Mini Kit (QIAGEN) according to manufacturer's instructions.
+The Perth/2009 HA and NA genes were then reverse transcribed, PCR amplified, and cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2 \comment{cite?} plasmid backbones.
+
+The codon-mutant libraries were generated using a PCR-based approach described in~\cite{bloom2014experimentally,dingens2017comprehensive}.
+
+\subsection*{Generation and passaging of mutant viruses}
+The mutant virus libraries were generated and passaged using the approach described in~\cite{doud2016accurate} with several modifications.
+
+\subsection*{Barcoded subamplicon sequencing}
+
+\subsection*{Analysis of deep sequencing data}
+
+\subsection*{Inference of phylogenetic trees}
+\comment{We downloaded X sequences from the Influenza Virus Resource ?.... etc. inferred the tree, ancestral state reconstruction, visualized the tree. Mark Perth/2009 on the tree}
+To parse out trunk mutations from side branch mutations, we first defined a set of recent nodes sampled on or after Jan. 1, 2017, and traced these nodes back to their most recent common ancestor.
+
+\subsection*{Quantification of mutational effects and sequence preferences from an H3N2 phylogeny}
+
+\subsection*{Data availability and source code}
+Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
+Computer code used to analyze the data and produce the results in the paper are in...
 }
@@
 
-\acknow{Please include your acknowledgments here, set in a single paragraph. Please do not include any acknowledgments in the Supporting Information, or anywhere else in the manuscript.}
+\acknow{We thank Sarah Hilton, Hugh Haddox, Sidney Bell...the Fred Hutch Genomics Core...
+Funding...}
 

### 2018-01-05 paper/manuscript.tex
Reorganize trunkvssidebranch figure
Added caption for H3N2 phylogeny figure

@@
 \caption{\label{fig:H3N2_phylogeny}
-{\bf Human H3N2 HA phylogeny.}
+{\bf Human H3N2 HA phylogeny from 1968-2017.}
+The trunk is shown in red, and side branches are shown in blue.
+The gray branches represent the part of the tree for which we cannot yet distinguish the trunk from side branches.
+The Perth/2009 strain is indicated with a star.
 }
@@
 
-\begin{figure*}
+\begin{SCfigure*}[\sidecaptionrelwidth][t]
 \centering
-\includegraphics[width=13cm]{figs/dms_overview/dms_overview.pdf}
+\includegraphics[width=12cm]{figs/dms_overview/dms_overview.pdf}
 \caption{\label{fig:dms_overview}
@@
 }
-\end{figure*}
+\end{SCfigure*}
 
@@
 
-\begin{figure*}
+\begin{SCfigure*}[\sidecaptionrelwidth][t]
 \centering
-\includegraphics[width=16cm]{figs/mut_tolerance/entropy_heatmap.pdf}
+\includegraphics[width=11.4cm]{figs/mut_tolerance/entropy_heatmap.pdf}
 \caption{\label{fig:mut_tolerance}
@@
 
-\begin{figure*}
+\begin{SCfigure*}[\sidecaptionrelwidth][t]
 \centering
-\includegraphics[width=12cm]{figs/trunkvssidebranch/trunkvssidebranch.pdf}
+\includegraphics[width=11.4cm]{figs/trunkvssidebranch/trunkvssidebranch.pdf}
 \caption{\label{fig:trunkvssidebranch}
 {\bf Mutations in evolutionary successful strains tend to be more favorable than in strains that die out.}
-(A) Phylogenetic tree of human H3N2 influenza virus from 1968-present. 
-The trunk is shown in red, and side branches are shown in blue.
-The gray branches represent the part of the tree for which we cannot yet distinguish the trunk from side branches.
-(B) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
+(A) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
 The median log$_{2}$ mutational effect in a given window is shown as circles for trunk mutations and triangles for side branch mutations. 
@@
 The median trunk mutational effects are consistently higher than the median side branch mutational effects for all windows.
+(B) The sequence preference normalized by the number of sites of every node in the human H3N2 phylogenetic tree shown in Fig.~\ref{fig:H3N2_phylogeny}.
+Trunk nodes are in red, side branch nodes in blue, and the Perth/2009 strain is marked with a star.
+Higher preferences are closer to zero while lower preferences are more negative.
 (C) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel) are shown.
@@
 To calculate the sequence preference, we used the entire HA sequence (top panel), only epitope sites (center panel) or non-epitope sites (bottom panel) as defined by~\cite{wolf2006long}.
-Higher preferences are closer to zero while lower preferences are more negative.
 The preferences at all sites and at epitope sites increase as the tree approaches the Perth/2009 strain, indicating epistasis among epitope sites.

### 2018-01-05 paper/manuscript.tex
Added Sanger seq supp fig to manuscript

@@
 
-We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (\comment{supplementary figure with Sanger sequencing}).
+We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (Figure~\ref{suppfig:SangerSeq}).
 We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables the efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
@@
 
+\begin{suppfigure}
+\centerline{\includegraphics[width=\textwidth]{figs/S02_SangerSeq/SangerSeq.pdf}}
+\caption{\label{suppfig:SangerSeq}
+{\bf Sanger sequencing of 30 clones from the mutant plasmid libraries.} 
+(A) 
+(B)
+(C)
+(D)
+(E)
+}
+\end{suppfigure}
+
 \subsubsection*{SI Tables}

### 2018-01-05 paper/manuscript.tex
Rework sequence preference figure

@@
 {\bf Mutations in evolutionary successful strains tend to be more favorable than in strains that die out.}
-(A) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of 5 years for every year from 1968-2013. 
+(A) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of five years for every year from 1968-2013. 
+The central year in a given five-year window is denoted on the x-axis.
 The median log$_{2}$ mutational effect in a given window is shown as circles for trunk mutations and triangles for side branch mutations. 
@@
 
-\begin{figure*}
+\begin{SCfigure*}[\sidecaptionrelwidth][t]
 \centering
-\includegraphics[width=12cm]{figs/sequence_preference/sequence_preference.pdf}
+\includegraphics[width=11.4cm]{figs/sequence_preference/sequence_preference.pdf}
 \caption{\label{fig:sequence_preference}
-{\bf The HA sequences of trunk nodes tend to be more preferred than those of side branch nodes.}
-We used the preferences to calculate the HA sequence preference normalized by the number of sites of every node in a human H3N2 phylogenetic tree.
-To calculate the sequence preference, we used the entire HA sequence (top panel), only epitope sites (center panel) or non-epitope sites (bottom panel) as defined by~\cite{wolf2006long}.
-The preferences at all sites and at epitope sites increase as the tree approaches the Perth/2009 strain, indicating epistasis among epitope sites.
-The trunk nodes generally exhibit higher sequence preferences than do side branch nodes.
+{\bf Epistasis has been driven by substitutions at epitope sites.}
+(A) We calculated the HA sequence preference normalized by the number of sites for epitope and non-epitope sites, as defined by~\cite{wolf2006long}.
+The preferences at all sites and at epitope sites increase as the tree approaches the Perth/2009 strain.
+The sequence preference of non-epitope sites is higher than that of epitope sites.
+(B) The log$_{2}$ mutational effect for side branch and trunk mutations at non-epitope (left) and epitope (right) sites are shown.
 }
-\end{figure*}
+\end{SCfigure*}
 

### 2018-01-05 paper/manuscript.tex
Add clade specific boxplots to RMSD_heatmap figure

@@
 \centering
-\includegraphics[width=16cm]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
+\includegraphics[width=17cm]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
 \caption{\label{fig:RMSD_heatmap}
-{\bf Shifts in preferences mapped onto the structure of HA}
-(A) The preference shifts as calculated by $RMSD_{corrected}$ between the two HA homologs is mapped onto the structure of HA (PDB 4O5N;~\cite{lee2014receptor}). 
+{\bf Characterization of the preference shifts}
+(A) The preference shifts as calculated by $\textit{RMSD}_\textit{corrected}$ between the two HA homologs is mapped onto the structure of HA (PDB 4O5N;~\cite{lee2014receptor}). 
 The left structure shows the HA trimer, and the right structure colors one of the monomers. 
@@
 Blue indicates small shifts in preference near zero, while red indicates large shifts in preference.
-The top ten most shifted sites are shown in spheres on the monomer.
 (B) The stalk domain was found to be significantly less shifted than the head domain (left plot).
-Sites absolutely conserved all 18 HA subtypes were also found to be significantly less shifted than the remaining non-conserved sites (right plot).
-(C) 
+Sites absolutely conserved all 18 HA subtypes were also found to be significantly less shifted than the remaining non-conserved sites (middle plot).
+Sites specific to the clade containing H1, H2, H5, and H6 and specific to the H3, H4, and H14 clade are significantly more shifted than non-clade-specific sites (right plot).
+(C) HA group-specific interactions between sites 107 and 75(HA2) play a role in determining the upward translation of the globular head domain relative to the stalk domain.
+The amino acids at sites 107 and 75(HA2) are shown in spheres on the structure of H1 and H3. 
+One monomer is shown in dark gray, while the HA1 domain of the neighboring monomer is in lighter gray.
+The sphere colors correspond to the shift in preference at that site.
+The preferences for each site in WSN/1933 and Perth/2009 are also shown.
 }
@@
 \begin{suppfigure}
-\centerline{\includegraphics[width=\textwidth]{figs/S02_SangerSeq/SangerSeq.pdf}}
+\centerline{\includegraphics[width=0.6\textwidth]{figs/S02_SangerSeq/SangerSeq.pdf}}
 \caption{\label{suppfig:SangerSeq}

### 2018-01-06 paper/manuscript.tex
Some tweaks to manuscript

@@
 \caption{\label{suppfig:SangerSeq}
-{\bf Sanger sequencing of 30 clones from the mutant plasmid libraries.} 
-(A) 
-(B)
-(C)
-(D)
-(E)
+{\bf Sanger sequencing of 31 clones from the mutant plasmid libraries.} 
+(A) There were an average of $\sim$1.4 codon mutations per clone across the three plasmid mutant libraries.
+(B) A mixture of one-, two-, and three-nucleotide mutations were present, with slightly fewer triple-nucleotide changes than expected.
+(C) Nucleotide frequencies were uniform in the codon mutations.
+(D) The mutations were distributed relatively evenly across the length of the HA coding sequence.
+(E) We calculated the pairwise distances between mutations for clones carrying more than one mutation.
+The cumulative distribution of the distances is shown in the red line.  
 }

### 2018-01-06 paper/manuscript.tex
edits to first 3 sections of Results

@@
 
-\begin{SCfigure*}[\sidecaptionrelwidth][t]
+%\begin{SCfigure*}[\sidecaptionrelwidth][t]
+\begin{figure}
 \centering
-\includegraphics[width=11.4cm]{figs/H3N2_phylogeny/H3N2_phylogeny.pdf}
+\includegraphics[width=\columnwidth]{figs/H3N2_phylogeny/H3N2_phylogeny.pdf}
 \caption{\label{fig:H3N2_phylogeny}
@@
 }
-\end{SCfigure*}
+\end{figure}
+%\end{SCfigure*}
 
@@
 How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
-This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describe H3N2 evolution compared to standard phylogenetic substitution models~\cite{hilton2017phydms}.
-Table~\ref{tab:phydms} shows that the ExpCM greatly outperforms conventional substitution models, showing that our experiments authentically capture some of the constraints on HA evolution. 
-\comment{Jesse and Juhye are here in sentence-level editing} 
-The ExpCM also optimizes a stringency parameter for the preferences to more closely reflect the strength of selection in nature.
-The stringency parameter in the ExpCM is equal to 2.44, which indicates that although the same amino acids are preferred, the strength of selection is more stringent in nature than in our experiments.
-Figure~\ref{fig:logoplot} shows a logo plot of the Perth/2009 HA amino-acid preferences rescaled by this stringency parameter.
+This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describe H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
+Table~\ref{tab:phydms} shows that the ExpCM using the replicate-average measurements greatly outperforms conventional substitution models, showing that our experiments authentically capture some of the constraints on HA evolution.
+Note that while the relative ratio of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models, this ratio is close to one for the ExpCM that accounts for the constraints measured in the experiments.
+ExpCM contain a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
+We fit a stringency parameter of 2.44 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments, but with greater stringency.
+Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
+These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.
 
-\begin{figure*}
-\centering
+\begin{table*}
 \caption{\label{tab:phydms}
-{\bf The site-specific amino-acid preferences are informative for describing human H3N2 evolution in nature.}}
+{\bf Substitution models informed by the experiments describe H3N2's natural evolution better than traditional substitution models.}}
+\begin{center} 
 \begin{tabular}{cccccccc}
 \hline
-\bf{Model} & \bf{$\Delta$AIC} & \bf{Log Likelihood} & \bf{Stringency} & \bf{$\omega$} & \bf{$\overline{\omega}$} & \bf{$\omega_{\alpha}$} & \bf{$\omega_{\beta}$} \\ \hline
-ExpCM & 0.0 & -8439.33 & $2.44$ & $0.91$ & -- & -- & -- \\
-Goldman-Yang M5 & 2166.06 & -9516.36 & -- & -- & $0.36$ & $0.30$ & $0.84$ \\
-ExpCM, averaged across sites & 2504.18 & -9691.42 & $0.68$ & $0.32$ & -- & -- & -- \\
-Goldman-Yang M0 & 2607.92 & -9738.29 & -- & $0.31$ & -- & -- & -- \\
+\bf{Model} & \bf{$\Delta$AIC} & \bf{Log Likelihood} & \bf{Stringency} & \bf{$\omega$}  \\ \hline
+ExpCM & 0.0 & -8439.33 & 2.44 & 0.91 \\
+Goldman-Yang M5 & 2166.06 & -9516.36 & -- & 0.30, 0.84, 0.36 \\
+ExpCM, averaged across sites & 2504.18 & -9691.42 & 0.68 & 0.32 \\
+Goldman-Yang M0 & 2607.92 & -9738.29 & -- & 0.31 \\
 \hline
 \end{tabular}
+ \end{center}
+\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 influenza HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites, and the M0 and M5 versions of the Goldman-Yang model~\cite{yang2000codon}.
+Models are compared by AIC~\cite{posada2004model}.
+The $\omega$ parameter is dN/dS for the Goldman-Yang models, and the relative dN/dS after accounting for the measurements for the ExpCM.
+For the M5 model, we give the shape parameter, rate parameter, and mean of the gamma distribution over $\omega$.
+}
+\end{table*}
 
-\addtabletext{We implemented several codon substitution models for phylogenetic fitting of an alignment of human H3N2 HA sequences. 
-The maximum likelihood values for each model were compared using the Akaike information criteria ($\Delta$AIC)~\cite{posada2004model}.
-An experimentally-informed codon substitution model (ExpCM) built from the preferences averaged across all replicates performs better than conventional substitution models, specifically the M0 and M5 models in~\cite{yang2000codon}.
-A non-site-specific ExpCM informed by preferences averaged across all sites performs comparably to the GY94 class of models, indicating the importance of site-specificity in the ExpCM.
-The optimized parameters for each model are also shown.}
-\end{figure*}
-
-A closer examination of the logo plot reveals that the preferences generally agree with existing knowledge about HA's biochemistry.
+A closer examination of Figure~\ref{fig:logoplot} reveals that the experimentally measured amino-acid preferences generally agree with existing knowledge about HA's structure and function.
 For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} possess high preference for cysteine.
-At residues involved in receptor binding, there are strong preferences for the amino acids at sites Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies}.
+At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies}\comment{These old citations are for X-31 HA, you might also see if Paulson and/or Ian Wilson have newer structural papers on receptor-binding in recent H3 -- if so, add them}.
 A positively charged amino acid at site 329 is important for cleavage activation of the HA0 precursor, and indeed this site exhibits a high preference for arginine~\cite{kido1992isolation, stech2005new}.
@@
 This codon is part of the signal peptide and is cleaved from the mature HA protein.
-One possible explanation for why we do not see a strong preference for Met at this site is due to alternative translation initiation occurring at a downstream or upstream start site, as has been described for HA~\cite{girard2011upstream}.
+One possible reason we do not see a strong preference for methionine at this site is due to alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
 % No upstream or downstream "ATG"'s are present.
@@
 \caption{\label{fig:logoplot}
-{\bf The site-specific amino-acid preferences of the Perth/2009 HA.}
-This logoplot shows the site-specific amino-acid preferences for the averaged replicates rescaled by the stringency parameter (Table 1) estimated by phydms.
-The height of each letter is proportional to its preference at that site, and the preferences for all sites are normalized to sum to 1.
+{\bf The site-specific amino-acid preferences of the Perth/2009 HA measure in our experiments.}
+The height of each letter is the preference for that amino acid, after taking the average of our experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter in Table~\ref{tab:phydms}.
 The sites are in H3 numbering.
-The top overlay bar shows the relative solvent accessibility.
-The bottom overlay bar is colored by the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
-The letters directly above each logo indicate the wildtype amino acid at that site.
+The top overlay bar indicates whether or not a site is part of the set of epitope residues delineated in \cite{wolf2006long}
+The bottom overlay bar indicates the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
+The letters directly above each logo stack indicate the wildtype amino acid at that site.
 }
@@
 
-\subsection*{The mutational tolerance of the head and the stalk shows less of a contrast in H3 than in H1}
-We next sought to investigate the inherent mutational tolerance of the Perth/2009 HA. 
-Figure~\ref{fig:mut_tolerance} shows the mutational tolerance as calculated from the rescaled Perth/2009 H3 preferences and the rescaled WSN/1933 H1 preferences mapped onto the HA crystal structures.
-We found antigenic site C and the most distal portion of the globular head near the 190-helix in the Perth/2009 H3 to be tolerant of mutations.
-Interestingly, the H3 stalk including the shorter $\alpha$-helix (helix A) is relatively mutationally tolerant compared to the tolerance of the globular head domain. 
-This observation suggests that the stalk may be prone to escape from antibodies, and agrees with previous work demonstrating that it is possible to select for antigenic mutants in H3 by broadly-neutralizing stalk-targeting antibodies~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human}.
-% In contrast, the head domain of the WSN/1933 H1 is more mutationally tolerant relative to its stalk domain.
+\subsection*{There is less difference in mutational tolerance between the HA head and stalk domains for H3 than for H1}
+Our experiments measure which amino acids are tolerated at each HA site under selection for protein function.
+We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we define as the Shannon entropy of the re-scaled preferences.
+In our prior deep mutational scanning of the WSN/1933, we found that the stalk domain of this H1 HA was substantially less mutationally tolerant than its globular head~\cite{thyagarajan2014inherent,doud2016accurate}.
+\comment{Do the Nicholas Wu papers with Ren Sun compare stalk and head at all?
+If so, maybe we should generalize this sentence to include them.}
 
-The sites inside the receptor binding pockets are highly functionally constrained and were found to be relatively mutationally intolerant in both H3 and H1~\citep{wilson1981structure}.
-In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to antigenic evolution as these sites are under strong immune pressure.~\citep{wiley1981structural}.
+We performed a similar analysis using the current data for the Perth/2009 H3 HA.
+Surprisingly, there was much less contrast in mutational tolerance between the stalk and head domains for the H3 HA than the H1 (Figure~\ref{fig:mut_tolerance}).
+For instance, the in H3 HA, the short helix A in the stalk is very mutationally tolerant.
+Interestingly, more studies have reported selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{doud2017quantifying}\comment{C179 paper} HAs.
+
+We also see high mutational tolerance in many of the known antigenic sites of H3 HA~\comment{cite some paper on these sites}.
+For instance, in recent H3N2 strains, antigenic site B is immunodominant and the location of most major antigenic drift mutations~\comment{is there some Scott Hensley that we can cite?}.
+We find that the most distal portion of the globular head near the 190-helix, which is part of antigenic site B, is highly tolerant of mutation (Figure~\ref{fig:mut_tolerance}).
+Antigenic site C is also notably mutationally tolerant.
+
+Many sites inside HA's receptor binding pocket are known to be highly functionally constrained~\citep{wilson1981structure,martin1998studies}, and our data indicates that these sites are relatively mutationally intolerant in both H3 and H1 HAs.
+In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, as these sites are under strong immune pressure.~\citep{wiley1981structural}.\comment{also the Koel et al paper}.
+
+\comment{One other issue we should think about is use of ``site'', ``residue'', and ``region''.
+I think we should probably be consistent with ``site'' for amino-site, because we've locked ourselves into that nomenclature.
+So I wonder if we should say ``antigenic region A'', etc to be clear?
+Maybe consider going back in here and changing that.}
 
@@
 \caption{\label{fig:mut_tolerance}
-{\bf Mutational tolerance mapped onto each site of HA.}
-Mutational tolerance as calculated by the Shannon entropy of a given site's amino-acid preferences are mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}) and the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}), with both trimers in approximately the same orientation.
-The site entropies were calculated from the preferences measured in the Perth/2009 H3 (left panel) from this study, or the preferences measured in the WSN/1933 H1 (right panel) from~\cite{doud2016accurate}.
-Lighter shades of purple or brown signify low mutational tolerance, while darker shades signify high mutational tolerance. 
-For each HA, the structure on the left side colors the full HA trimer, while the structure on the right side colors only one of the monomers.
-The sialic acid receptor is shown as black sticks.
-The Perth/2009 H3 shows relatively high mutational tolerance in the stalk region, particularly in helix A, compared to the head region. 
-High mutational tolerance in H3 was also observed near antigenic site C and near the 190 helix.
-The head region of the WSN/1933 H1 is mutationally tolerant compared to the relatively intolerant stalk region. 
+{\bf Mutational tolerance of each site in H3 and H1 HAs.}
+Mutational tolerance as measured in the current study is mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}).
+Mutational tolerance of the WSN/1933 H1 HA as measured in \cite{doud2016accurate} is mapped on the structure of the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}).
+Different color scales are used because measurements are comparable among sites within the same HA, but not necessarily across HAs.
+Both trimers are shown in approximately the same orientation. 
+For each HA, the structure at left shows a surface representation of the full trimer, while the structure at right side shows a ribbon representation of just one monomer.
+The sialic acid receptor is shown in black sticks.
 }

### 2018-01-08 paper/manuscript.tex
Re-edit first four results sections, draft of last two results sections, intro, abstract

@@
 \begin{abstract}
-Please provide an abstract of no more than 250 words in a single paragraph.
-Abstracts should explain to the general reader the major contributions of the article.
-References in the abstract must be cited in full within the abstract itself and cited in the text.
+% no more than 250 words. (200 now)
+The evolution of seasonal H3N2 influenza virus is mediated by the rapid accumulation of mutations. 
+A better understanding of how mutations affect viral growth is critical for studying the pressures and constraints underlying the evolution of the influenza virus in nature. 
+Here we experimentally measure the effect of every possible single amino-acid mutation to an H3 influenza hemagglutinin on viral growth in cell culture. 
+We find the stalk domain to be fairly mutationally tolerant, and there is not a large disparity in tolerance between the head and stalk domains.
+Furthermore, our measurements reveal mutations in successful seasonal influenza viral lineages to be significantly more preferred than mutations in lineages that have died out, and this was true at both epitope and non-epitope sites.
+Epistasis in hemagglutinin is also seemingly driven by antigenic drift.
+The preferences measured in a distantly related protein homolog do not reveal such differences in viral lineages.
+A comparison of the preferences between the two hemagglutinin homologs suggests that mutational effects increasingly shift as proteins diverge.
+Overall, our work highlights how experimental measurements of mutational effects can be leveraged to gain insight into the evolutionary fates of viral strains and potentially be used to predict viral evolution.
 \end{abstract}
@@
 If this is the case, add \parshape=0 to the end of the list environment.
-\dropcap{V}ery rough outline:
-\begin{itemize}
-\item Mutations are rampant in the evolution of human influenza virus. Seasonal H3N2 influenza virus in particular rapidly accumulates mutations.
-\item The evolution of H3N2 is also characterized by clade competition and population turnover.
-\item There have been efforts to predict evolutionary success.
-\item Mutations that contribute to antigenic evolution largely determine strain success.
-\item RNA viruses can accumulate deleterious mutations, and because the influenza does not appreciably recombine, deleterious mutations are linked to beneficial ones.
-\item Mutations in HA that impact viral growth may influence strain success.
-\item We need to understand the functional impact of mutations in HA.
-\item Previously, we measured the effect of all possible single amino-acid mutations to an H1 hemagglutinin from the A/WSN/1933 (H1N1) strain~\citep{thyagarajan2014inherent,doud2016accurate}. 
-\item However, this is a highly lab-adapted strain, and the measurements in this strain may not be particularly relevant for studying mutational processes of more contemporaneous strains circulating in the human population.
-\item We chose to study the Perth/2009 H3 HA.
-\item This also enabled a comparison of how the preferences have shifted for two diverged HA's.
-\end{itemize}
+\dropcap{S}easonal H3N2 influenza virus evolves rapidly, readily accumulating mutations in the hemagglutinin (HA) surface protein that contribute to antigenic drift.
+The evolution of human H3N2 influenza virus is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2015global}, producing a distinct spindly shape in its phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
+Given the rather distinctive pattern of H3N2 evolution and that vaccination remains an effective way of reducing disease burden, a key challenge in selecting a vaccine strain lies in being able to predict the dominant strain in the upcoming influenza season.
+
+Vaccine strain selection decisions mainly rely on experimentally characterizing the antigenicity of circulating strains.
+To aid in the timeliness and throughput of making such decisions, previous work have endeavored to accurately predict the evolutionary outcomes of viral strains based on the antigenicity and/or tree structure of viral clades~\cite{steinbruck2014computational,neher2014predicting}.
+However, the accuracy of these predictions are subject to uneven geographical and temporal sampling, and for~\cite{neher2014predicting}, the lack of information about a mutation's effect on antigenicity.
+Others have attempted to model the antigenic properties of viral strains using sequence information or phylogenetic relationships~\cite{sun2013using,harvey2016identification,neher2016prediction}.
+
+While mutations that contribute to antigenic evolution of the virus certainly determine strain success, the non-antigenic impacts of mutations play an important role in determining the persistence of viral strains~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
+Due to the high mutation rate of RNA viruses \comment{citation here} and because the influenza virus does not appreciably recombine within a given segment, deleterious mutations become linked to beneficial ones.
+A predictive fitness model takes into account mutations at both antigenic and non-antigenic sites in HA to infer clade fitness~\cite{luksza2014predictive}.
+In this model, mutations at epitope sites are beneficial to viral fitness, while mutations at non-epitope sites incur a fitness cost.
+Although this assumption may generally be true, the effect of mutations on viral fitness, and particularly on protein function, are much more nuanced.
+We therefore need a greater understanding of the phenotypic effects of mutations on viral growth to comprehend the forces and constraints shaping influenza virus evolution in nature.
 
@@
 \centering
-\includegraphics[width=\columnwidth]{figs/H3N2_phylogeny/H3N2_phylogeny.pdf}
+\includegraphics[width=\linewidth]{figs/H3N2_phylogeny/H3N2_phylogeny.pdf}
 \caption{\label{fig:H3N2_phylogeny}
@@
 %\end{SCfigure*}
+
+We can measure such mutational effects in the lab using deep mutational scanning approaches~\cite{fowler2010high}.
+Previously, we experimentally measured the effect of all possible single amino-acid mutations to an H1 HA from the A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,doud2016accurate}.
+However, the WSN/1933 is a highly lab-adapted strain, and these measurements may not be particularly relevant for studying the mutational processes of more contemporaneous strains circulating in the human population.
+Here we measure the effects of mutations to the HA from the A/Perth/16/2009 (H3N2) strain on viral growth in cell culture.
+% Our measurements generally agree with what is already known about HA structure and function.
+We apply these experimental measurements to examine differences in the trunk and side branch lineages of human H3N2 influenza virus.
+Trunk mutations were found to be more favorable than side branch mutations, both at epitope and non-epitope sites.
+We also find evidence that epistasis in HA has primarily been driven by antigenic drift.
+The measurements also enabled a comparison of how the amino-acid preferences have shifted between two diverged HAs.
+The extent of shifts in amino-acid preferences increase as homologs diverge.
+Our work highlights the potential for using high-throughput experimental measurements of mutational effects to inform evolutionary forecasting of human seasonal influenza virus.
 
@@
 Specifically, stop codons were purged to 20-45\% of their initial frequencies after correcting for error rates estimated by sequencing the wildtype controls.
-The incomplete purging of stop codons is likely because genetic complementation due to co-infection \comment{let's cite some paper here, maybe Chris Brooke one?} enabled the persistence of some virions with nonfunctional HAs.
+The incomplete purging of stop codons is likely because genetic complementation due to co-infection~\cite{marshall2013influenza} enabled the persistence of some virions with nonfunctional HAs. 
 We also observed selection against many nonsynonymous mutations, with their frequencies falling to 30-40\% of their initial values after error correction.
@@
 For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} possess high preference for cysteine.
-At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies}\comment{These old citations are for X-31 HA, you might also see if Paulson and/or Ian Wilson have newer structural papers on receptor-binding in recent H3 -- if so, add them}.
+At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change}\comment{These old citations are for X-31 HA, you might also see if Paulson and/or Ian Wilson have newer structural papers on receptor-binding in recent H3 -- if so, add them}.
 A positively charged amino acid at site 329 is important for cleavage activation of the HA0 precursor, and indeed this site exhibits a high preference for arginine~\cite{kido1992isolation, stech2005new}.
@@
 We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we define as the Shannon entropy of the re-scaled preferences.
-In our prior deep mutational scanning of the WSN/1933, we found that the stalk domain of this H1 HA was substantially less mutationally tolerant than its globular head~\cite{thyagarajan2014inherent,doud2016accurate}.
-\comment{Do the Nicholas Wu papers with Ren Sun compare stalk and head at all?
-If so, maybe we should generalize this sentence to include them.}
+In prior mutational studies of the WSN/1933, the stalk domain of this H1 HA was found to be substantially less mutationally tolerant than its globular head~\cite{thyagarajan2014inherent,doud2016accurate}.
 
 We performed a similar analysis using the current data for the Perth/2009 H3 HA.
-Surprisingly, there was much less contrast in mutational tolerance between the stalk and head domains for the H3 HA than the H1 (Figure~\ref{fig:mut_tolerance}).
+Surprisingly, there was much less contrast in mutational tolerance between the stalk and head domains for the H3 HA than for the H1 (Figure~\ref{fig:mut_tolerance}).
 For instance, the in H3 HA, the short helix A in the stalk is very mutationally tolerant.
-Interestingly, more studies have reported selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{doud2017quantifying}\comment{C179 paper} HAs.
+Interestingly, more studies have reported selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2017quantifying} HAs.
 
-We also see high mutational tolerance in many of the known antigenic sites of H3 HA~\comment{cite some paper on these sites}.
-For instance, in recent H3N2 strains, antigenic site B is immunodominant and the location of most major antigenic drift mutations~\comment{is there some Scott Hensley that we can cite?}.
-We find that the most distal portion of the globular head near the 190-helix, which is part of antigenic site B, is highly tolerant of mutation (Figure~\ref{fig:mut_tolerance}).
-Antigenic site C is also notably mutationally tolerant.
+We also see high mutational tolerance in many of the known antigenic regions of H3 HA~\cite{wiley1981structural}.
+For instance, in recent H3N2 strains, antigenic region B is immunodominant and the location of most major antigenic drift mutations~\cite{chambers2015identification,koel2013substitutions,popova2012immunodominance}.
+We find that the most distal portion of the globular head near the 190-helix, which is part of antigenic region B, is highly tolerant of mutations (Figure~\ref{fig:mut_tolerance}).
+Antigenic region C is also notably mutationally tolerant.
 
-Many sites inside HA's receptor binding pocket are known to be highly functionally constrained~\citep{wilson1981structure,martin1998studies}, and our data indicates that these sites are relatively mutationally intolerant in both H3 and H1 HAs.
-In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, as these sites are under strong immune pressure.~\citep{wiley1981structural}.\comment{also the Koel et al paper}.
-
-\comment{One other issue we should think about is use of ``site'', ``residue'', and ``region''.
-I think we should probably be consistent with ``site'' for amino-site, because we've locked ourselves into that nomenclature.
-So I wonder if we should say ``antigenic region A'', etc to be clear?
-Maybe consider going back in here and changing that.}
+Many residues inside HA's receptor binding pocket are known to be highly functionally constrained~\cite{wilson1981structure,martin1998studies}, and our data indicates that these sites are relatively mutationally intolerant in both H3 and H1 HAs.
+In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, as these sites are under strong immune pressure.~\cite{wiley1981structural,koel2013substitutions}.
 
@@
 An important question to thereby address is if our experimental measurements are useful in differentiating between strains of human H3N2 influenza virus that have succeeded and those that have died out.
-To investigate this, we used our preference dataset and an H3N2 phylogeny from 1968-2012 (Figure~\ref{fig:trunkvssidebranch}A) to calculate the effects of mutations for the evolutionarily successful trunk lineage and for side branches which have died out.
+To investigate this, we used our preference dataset and an H3N2 phylogeny from 1968-2012 (Figure~\ref{fig:H3N2_phylogeny}) to calculate the effects of mutations for the evolutionarily successful trunk lineage and for side branches which have died out.
 We found that strains with mutations measured to be more beneficial to viral growth in our experiments tend to succeed in nature.
-Figure~\ref{fig:trunkvssidebranch}B shows the effects of trunk and side branch mutations in five-year intervals for every year from 1968-2012. 
+Figure~\ref{fig:trunkvssidebranch}A shows the effects of trunk and side branch mutations in five-year intervals for every year from 1968-2012. 
 On average, trunk mutations are towards more preferred amino acids compared to side branch mutations, and this was true for all intervals.
+
+How distantly can the preferences be extended to describe differences in successful and unsuccessful strains?
+To explore this question, we scored the complete HA sequence of every node in the tree using the preferences by quantifying a \textit{sequence preference} metric.
+The sequence preference is defined as 
+$\displaystyle\sum_{r} \ln \pi_{r, a}$
+where $\pi_{r, a}$ is the preference for amino acid $a$ at site $r$.
+The sequence preference is then divided by the number of sites.
+Consistent with the finding that trunk mutations are generally more favorable than side branch mutations, Figure~\ref{fig:trunkvssidebranch}B shows that the sequences of trunk nodes also tend to be more highly preferred than those of side branch nodes.
+This indicates that our experimental measurements reveal the selective advantage and thereby persistence of the trunk.
+Interestingly, the sequence preferences increase over time as the nodes approach the Perth/2009 strain and its closely related nodes, which all exhibit sequence preferences that are higher than that of the trunk.
+The observation that the strain in which we performed our deep mutational scan has one of the highest sequence preferences illustrates epistatic interactions among mutations such that an unpreferred mutation in one background may be preferred in another \comment{cite Jesse's NP paper?}.
+% Due to the sheer number of possible pairwise combinations of mutations, it is impossible for us to experimentally characterize these epistatic interactions among mutations.
+
 Importantly, trunk mutations are significantly more favorable than side branch mutations when calculating the effects from the entirety of the phylogenetic tree (Figure~\ref{fig:trunkvssidebranch}C).
-
 Because tip nodes can contain egg- or cell-passaged isolates~\citep{wu2017structural,mcwhite2016sequence,skowronski2016mutations} and our experiments were performed in cell culture, we examined the effects of mutations on terminal side branch nodes to see if these would rank more highly than internal node mutations.
@@
 
-How distantly can the preferences be extended to describe differences in successful and unsuccessful strains?
-To explore this question, we scored the complete HA sequence of every node in the tree using the preferences by quantifying a \textit{sequence preference} metric.
-The sequence preference is defined as 
-$\displaystyle\sum_{r} \ln \pi_{r, a}$
-where $\pi_{r, a}$ is the preference for amino acid $a$ at site $r$.
-Consistent with the finding that trunk mutations are generally more favorable than side branch mutations, Figure~\ref{fig:sequence_preference} shows that the sequences of trunk nodes also tend to be more highly preferred than those of side branch nodes.
-Interestingly, the sequence preferences increase over time as the nodes approach the Perth/2009 strain and its closely related nodes, which all exhibit sequence preferences that are higher than that of the trunk.
-The observation that the strain in which we performed our deep mutational scan has one of the highest sequence preferences illustrates epistatic interactions among mutations such that an unpreferred mutation in one background may be preferred in another.
-% Due to the sheer number of possible pairwise combinations of mutations, it is impossible for us to experimentally characterize these epistatic interactions among mutations.
-
-\comment{this section will be fleshed out more once I figure out how to plot the substitutions from the root at epitope / non-epitope sites, as per Trevor's suggestion.}
 We reasoned that much of the increase in sequence preference toward Perth/2009 could be attributed to epitope sites, as most substitutions have occurred at sites under strong immune pressure.
 We therefore assessed each node's preference at epitope and non-epitope sites as defined by~\cite{wolf2006long}.
-Indeed, the preferences at epitope sites resembles those for all HA sites (Figure~\ref{fig:sequence_preference}), with preferences increasing over time.
-At non-epitope sites, however, the trunk preferences remain relatively constant while side branch preferences tend to drop below the trunk.
-These observations highlight the extensive epistasis among epitope sites, as has even been noted for sites within the 220-loop surrounding the receptor binding pocket~\citep{wu2017diversity}.
-Many successful epitope mutations are likely contingent upon the background in which they arise.
-Yet, we are able to distinguish trunk sequences as more favorable than side branch sequences, indicating that the preferences are still of utility over short evolutionary timescales.
+The preferences at epitope sites resembles those for all HA sites, with preferences increasing over time (Figure~\ref{fig:sequence_preference}A).
+At non-epitope sites, however, the sequence preferences are higher than at epitope sites, and remain relatively constant over time.
+In the timespan from 1968 to 2012, the trunk fixed X epitope mutations per site and Y non-epitope mutations per site.
+Epistasis in HA is seemingly predominantly driven by antigenic drift \comment{cite kryazhimskiy2011prevalence?}. 
+Extensive epistasis has even been noted for epitope sites within the 220-loop surrounding the receptor binding pocket~\cite{wu2017diversity}.
+The evolutionary success of many epitope mutations are also likely contingent upon the background in which they arise.
+Yet, we are able to distinguish trunk mutations as significantly more favorable than side branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B). 
+This suggests that the preferences are still of utility in distinguishing evolutionarily successful strains from non-successful ones over short evolutionary timescales.
 
@@
 Can we then distinguish lineage-specific mutational effects using the preferences measured in a distantly related HA homolog?
-We used the preferences measured previously in the WSN/1933 H1~\citep{doud2016accurate} to quantify the effects of H3 trunk and side branch mutations, shown in Figure~\ref{fig:WSN_trunkvssidebranch}.
+We used the WSN/1933 H1 preferences measured in~\cite{doud2016accurate} to quantify the effects of H3 trunk and side branch mutations, shown in Figure~\ref{fig:WSN_trunkvssidebranch}.
 It is evident that we do not see trunk mutations significantly more favored than side branch mutations, suggesting that our ability to discriminate successful and unsuccessful strains degrades over sufficiently long evolutionary distances.
@@
 {\bf The WSN/1933 H1 preferences do not reveal differences in trunk vs side branch mutational effects}
-(A) We used the WSN/1933 H1 preferences from~\cite{doud2016accurate} to calculate the log$_{2}$ mutational effects of trunk and side branch mutations from the inferred H3N2 phylogeny in Figure~\ref{fig:trunkvssidebranch}.
-There is not a significant difference in trunk vs side branch mutational effects.
+(A) We used the WSN/1933 H1 preferences experimentally measured in~\cite{doud2016accurate} to calculate the log$_{2}$ mutational effects of trunk and side branch mutations from the inferred H3N2 phylogeny in Figure~\ref{fig:trunkvssidebranch}.
 (B) We also performed a sliding window analysis with the WSN/1933 H1 preferences.
-There is not a distinct difference in trunk and side branch mutational effects.
 }
@@
 How shifted are the preferences between evolutionarily distant homologs such as H1 and H3?
-Although we have previously compared the preferences between related protein homologs of NP~\citep{doud2015site} and of Env \comment{cite haddox2017}, the HA homologs we have experimentally studied are considerably more diverged than either of these pairs.
-The WSN/1933 H1 and the Perth/2009 H3 fall into two phylogenetically distinct groups and share 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A), providing an opportunity to analyze what has shifted and what has remained conserved.
+Although we have previously compared the experimentally measured preferences between related protein homologs of NP~\cite{doud2015site} and of Env~\cite{haddox2017mapping}, the HA homologs we have studied are considerably more diverged than either of these pairs.
+The WSN/1933 H1 and the Perth/2009 H3 fall into two phylogenetically distinct groups and share 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A), allowing us to examine what has shifted and what has remained conserved across such diverged homologs.
 Simply correlating the preferences between H1 and H3 reveals that the replicate measurements within a homolog are more correlated than between homologs (Figure~\ref{fig:distance_distribution}B).
 
-However, to quantify the extent of mutational shifts at a site-by-site level, we used an approach described in \comment{haddox2017}.
+However, to quantify the extent of mutational shifts at a site-by-site level, we used an approach described in~\cite{doud2015site,haddox2017mapping}.
 We aligned Perth/2009 H3 and WSN/1933 H1 and at each alignable site calculated the difference in preferences between homologs while correcting for experimental noise within homologs.
@@
 Although many sites have small shifts near zero, a considerable number of sites have large shifts in preference, reaching a difference of 0.86 out of a maximum possible of 1.0.
-When we compare HA to the preferences measured in the non-homologous protein HIV Env \comment{haddox2017}, nearly all sites are shifted.
+When we compare HA to the preferences measured in the non-homologous protein HIV Env~\cite{haddox2017mapping}, nearly all sites are shifted.
 On the other hand, when we randomize the HA replicates to generate a null distribution, there is very little shift in preferences. 
@@
 Although there are a small number of sites with larger shifts reaching a maximum of 0.52, we do not see a dramatic tail in the distribution as we see for the HA's.
-These observations suggest that as homologs diverge, their preferences increasingly shift.
-
-Upon mapping the shifts in preference onto the HA structure (Figure~\ref{fig:RMSD_heatmap}A), we did not see the shifted preferences obviously localize to specific regions in HA.
-However, we found the overall stalk domain to be significantly less shifted than the globular head domain (Figure~\ref{fig:RMSD_heatmap}B), which was expected given that the stalk domain is more conserved within and across HA subtypes.
-Sites absolutely conserved across all 18 HA subtypes were also found to be significantly less shifted than sites not conserved, which is suggestive of the high functional importance of the residues that have remained unchanged throughout the divergence of HA.
-
-Despite their large sequence divergence, how is it that H1 and H3 adopt nearly identical protein folds \comment{cite Ha, Russell}?
+These observations suggest that as protein homologs diverge, their preferences increasingly shift.
 
@@
 (B) The correlations of the amino-acid preferences for replicates both within and between the two HA homologs. 
-The within-Perth/2009 and the within-WSN/1933 correlations are shown in blue and red, respectively.
+The within-Perth/2009 and the within-WSN/1933 correlations are shown in purple and brown, respectively.
 The between homolog correlations are in gray.
-The correlations for replicates within a homolog are higher than for replicates between homologs.
 (C) The distribution of shifts in preference for various homolog pairs are shown.
-The top two distributions show the distances between each of the HA homologs with the non-homologous HIV Env \comment{cite haddox2017}. 
+The top two distributions show the distances between each of the HA homologs with the non-homologous HIV Env~\cite{haddox2017mapping}. %BF520
 The center distribution shows the corrected distance between the two HIV Env homologs, which share 86\% amino-acid identity.
@@
 \end{figure*}
+
+We next asked if shifted sites tend to cluster together in the HA structure.
+However, the sites of large shifts were interspersed across HA and did not seem to obviously localize together (Figure~\ref{fig:RMSD_heatmap}A).
+Yet, at the domain level, we found the stalk domain to be significantly less shifted than the globular head domain (Figure~\ref{fig:RMSD_heatmap}B). 
+This is unsurprising given that the stalk domain is more conserved both within and across HA subtypes~\cite{nobusawa1991comparison,hai2012influenza,mallajosyula2014influenza}.
+Sites absolutely conserved across all 18 HA subtypes were also found to be significantly less shifted than sites not conserved (Figure~\ref{fig:RMSD_heatmap}B), which is suggestive of the high functional importance of the sites that have remained unchanged throughout the divergence of HA.
+
+Despite their large sequence divergence, H1 and H3 adopt nearly identical protein folds~\cite{ha2002h5,russell2004h1}.
+Previous work has found that although the subdomain structures are highly similar, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among the four HA clades represented by H1 \& H5, H9, H7, and H3~\cite{ha2002h5,russell2004h1}.
+One explanation for such clade-specific subdomain position differences could be local and/or long-range conformational interactions between sites that are variable between clades \comment{needs better wording}.
+We would expect such sites to exhibit large shifts in preference.
+Indeed, clade-specific sites that are conserved within the H1 clade (including H1, H2, H5, and H6) or within the H3 clade (including H3, H4, and H14) are significantly shifted compared to non-clade-specific sites.
+
+In particular, the clade-specific upward shift of the globular head relative to the stalk has largely been attributed to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
+The H1 HA has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
+In the WSN/1933 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
+In contrast, the interhelical loop in H3 HA makes a sharper and shorter turn which is facilitated by a Gly at 75(HA2).
+This site prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations suggesting that the amino-acid identity at this site plays a smaller role in determining the conformation of the interhelical loop in H3.
 
@@
 \subsection*{Inference of phylogenetic trees}
-\comment{We downloaded X sequences from the Influenza Virus Resource ?.... etc. inferred the tree, ancestral state reconstruction, visualized the tree. Mark Perth/2009 on the tree}
+\comment{We downloaded X sequences from the Influenza Virus Resource ?.... etc. inferred the tree, ancestral state reconstruction, visualized the tree...}
 To parse out trunk mutations from side branch mutations, we first defined a set of recent nodes sampled on or after Jan. 1, 2017, and traced these nodes back to their most recent common ancestor.
@@
 
+\begin{suppfigure}
+\centerline{\includegraphics[width=\textwidth]{figs/S03_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
+\caption{\label{suppfig:WSNprefs_logoplot}
+{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.} 
+}
+\end{suppfigure}
+
 \subsubsection*{SI Tables}

### 2018-01-08 paper/manuscript.tex
distinguish lineages section

@@
 
-%\begin{SCfigure*}[\sidecaptionrelwidth][t]
 \begin{figure}
@@
 \end{figure}
-%\end{SCfigure*}
 
@@
 
-\subsection*{The experimental measurements can help discriminate successful influenza virus lineages}
-A challenge in vaccine strain selection is predicting which strain will dominate the upcoming influenza season.
-An important question to thereby address is if our experimental measurements are useful in differentiating between strains of human H3N2 influenza virus that have succeeded and those that have died out.
-To investigate this, we used our preference dataset and an H3N2 phylogeny from 1968-2012 (Figure~\ref{fig:H3N2_phylogeny}) to calculate the effects of mutations for the evolutionarily successful trunk lineage and for side branches which have died out.
-We found that strains with mutations measured to be more beneficial to viral growth in our experiments tend to succeed in nature.
-Figure~\ref{fig:trunkvssidebranch}A shows the effects of trunk and side branch mutations in five-year intervals for every year from 1968-2012. 
-On average, trunk mutations are towards more preferred amino acids compared to side branch mutations, and this was true for all intervals.
+\subsection*{The experimental measurements can help discriminate between successful and unsuccessful influenza virus lineages}
+A major goal in the study of rapidly evolving viruses such as influenza is to forecast evolutionary trajectories\comment{Michael Lassig has a Nature Ecology \& Evolution review}.
+For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection\comment{add reference}. 
+Evolutionary forecasts must ultimately distinguish between successful and unsuccessful viral lineages, which in the case of human influenza virus means distinguishing between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).
 
-How distantly can the preferences be extended to describe differences in successful and unsuccessful strains?
-To explore this question, we scored the complete HA sequence of every node in the tree using the preferences by quantifying a \textit{sequence preference} metric.
-The sequence preference is defined as 
-$\displaystyle\sum_{r} \ln \pi_{r, a}$
-where $\pi_{r, a}$ is the preference for amino acid $a$ at site $r$.
-The sequence preference is then divided by the number of sites.
-Consistent with the finding that trunk mutations are generally more favorable than side branch mutations, Figure~\ref{fig:trunkvssidebranch}B shows that the sequences of trunk nodes also tend to be more highly preferred than those of side branch nodes.
-This indicates that our experimental measurements reveal the selective advantage and thereby persistence of the trunk.
-Interestingly, the sequence preferences increase over time as the nodes approach the Perth/2009 strain and its closely related nodes, which all exhibit sequence preferences that are higher than that of the trunk.
-The observation that the strain in which we performed our deep mutational scan has one of the highest sequence preferences illustrates epistatic interactions among mutations such that an unpreferred mutation in one background may be preferred in another \comment{cite Jesse's NP paper?}.
-% Due to the sheer number of possible pairwise combinations of mutations, it is impossible for us to experimentally characterize these epistatic interactions among mutations.
+To investigate whether our experiments can aid in distinguishing trunk and side branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the phylogeny in Figure~\ref{fig:H3N2_phylogeny}.
+The mutations that occurred on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
+This difference in the effects of mutations between the trunk and side branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
+Some influenza sequences are determined using egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{wu2017structural,mcwhite2016sequence,skowronski2016mutations}.
+These mutations will largely occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
+We therefore re-calculated the statistics separating side branches to tip and internal nodes, and again found that the difference between the trunk and both sets of side branches was statistically significant (Figure~\ref{fig:trunkvssidebranch}B).
+Therefore, strains with mutations that we experimentally measure to be favorable for viral replication tend to do better in nature than strains with mutations that we measure to be less favorable.
 
-Importantly, trunk mutations are significantly more favorable than side branch mutations when calculating the effects from the entirety of the phylogenetic tree (Figure~\ref{fig:trunkvssidebranch}C).
-Because tip nodes can contain egg- or cell-passaged isolates~\citep{wu2017structural,mcwhite2016sequence,skowronski2016mutations} and our experiments were performed in cell culture, we examined the effects of mutations on terminal side branch nodes to see if these would rank more highly than internal node mutations.
-Instead, terminal node mutations are on average towards less preferred amino acids than internal node mutations (Figure~\ref{fig:trunkvssidebranch}C), and both internal and terminal node mutations are significantly less favorable than those on the trunk.
-Therefore, strains that have accumulated mutations that we experimentally measured to be unpreferred generally die out in nature, while more favored mutations provide a selective advantage to the trunk.
-These findings demonstrate the importance of the functional impacts of HA mutations in determining strain success.
+Our experiments were performed on the Perth/2009 HA, but we are scoring mutations on a phylogenetic tree that extends from 1968 to 2017.
+Because the effects of mutations can vary with genetic background\comment{cite some papers about epistasis, amino-acid shifts, etc}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
+To explore this question, we scored the complete HA sequence of every node in the phylogenetic tree by quantifying its average per-site sequence preference for HA sequence $\mathbf{s}$, which is defined as
+\begin{equation}
+F\left(\mathbf{s}\right) = \frac{1}{L}\displaystyle\sum_{r=1}^L \ln \pi_{r, s_r},
+\end{equation}
+where $\pi_{r, a}$ is the preference for amino-acid $a$ at site $r$ as measured in our experiments (e.g., Figure~\ref{fig:logoplot}), $s_r$ is the amino-acid at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
+Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to be more highly preferred than those of side branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side branch mutations.
+However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
+The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
+These results suggest that there is some HA background-dependence in the effects of mutations, but that despite this fact our experimental measurements consistently reveal the selective advantage of the trunk over a more than half-century of viral evolution.
 
-\begin{SCfigure*}[\sidecaptionrelwidth][t]
+\begin{figure}
 \centering
-\includegraphics[width=11.4cm]{figs/trunkvssidebranch/trunkvssidebranch.pdf}
+\includegraphics[width=\columnwidth]{figs/trunkvssidebranch/trunkvssidebranch.pdf}
 \caption{\label{fig:trunkvssidebranch}
-{\bf Mutations in evolutionary successful strains tend to be more favorable than in strains that die out.}
-(A) Using the Perth/2009 H3 preferences, we calculated the log$_{2}$ mutational effect for trunk and side branch mutations in windows of five years for every year from 1968-2013. 
-The central year in a given five-year window is denoted on the x-axis.
-The median log$_{2}$ mutational effect in a given window is shown as circles for trunk mutations and triangles for side branch mutations. 
-The shaded region demarcates the interquartile range of trunk and side branch mutational effects.
-Negative numbers signify mutations towards less preferred amino acids, while positive numbers signify more preferred mutations.
-The median trunk mutational effects are consistently higher than the median side branch mutational effects for all windows.
-(B) The sequence preference normalized by the number of sites of every node in the human H3N2 phylogenetic tree shown in Fig.~\ref{fig:H3N2_phylogeny}.
-Trunk nodes are in red, side branch nodes in blue, and the Perth/2009 strain is marked with a star.
-Higher preferences are closer to zero while lower preferences are more negative.
-(C) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), in addition to all mutations in internal nodes and terminal nodes on the side branches (right panel) are shown.
-The preferences were randomized 10,000 times to estimate significance.
-The effects of trunk mutations are higher than side branch internal and terminal node mutations.
+{\bf Mutations on the trunk are more favorable than those on the side branches according to our experimental measurements.}
+\comment{Legend will be written under assumption that panels (B) and (C) are swapped. Also add sequence preference \emph{per site} to y-axis of the new (C).}
+(A) We used our experiments to calculate the log$_{2}$ mutational effect for all trunk and side branch mutations in five-year windows from 1968 to 2013. 
+The central year in each window is denoted on the x-axis.
+Median mutational effects in each window are shown as circles for the trunk and triangles for side branches. 
+The shaded regions demarcate the interquartile range.
+Negative numbers signify mutations towards less preferred amino acids, while positive numbers signify mutations to more preferred mutations.
+(B) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), and the same data but separating side branches to terminal and internal nodes.
+$P$-values were computed by randomizing the experimental measurements among sites 10,000 times, and determining how often the difference in median mutational effect between the trunk and side branches for the randomized data exceeded the actual value.
+(C) The average per-site sequence preference of every node in the phylogenetic tree in Fig.~\ref{fig:H3N2_phylogeny}.
+Trunk nodes are in red, side branch nodes in blue, and Perth/2009 is marked with a yellow star.
+More preferred sequences have larger values.
 }
-\end{SCfigure*}
+\end{figure}
+
+\subsection*{Another subsection}
 

### 2018-01-08 paper/manuscript.tex
epitope vs non-epitope section

@@
 
-\subsection*{Another subsection}
+\subsection*{Our experiments suggest different patterns of evolution at epitope and non-epitope sites}
+HA is under selection both to maintain its essential function in viral growth and to escape pre-existing immunity.
+Most of the immune selection is focused on a subset of so-called ``epitope sites'' that are targets of the immunodominant antibody response.
+Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites.
+In the timeframe from 1968 to 2012, the trunk of the tree in Figure~\ref{fig:H3N2_phylogeny} fixed \comment{X} epitope mutations per site, but only Y non-epitope mutations per site.
+Since our experiments only measure how mutations affect viral growth, they might be expected to describe the evolution of epitope and non-epitope sites differently.
 
-We reasoned that much of the increase in sequence preference toward Perth/2009 could be attributed to epitope sites, as most substitutions have occurred at sites under strong immune pressure.
-We therefore assessed each node's preference at epitope and non-epitope sites as defined by~\cite{wolf2006long}.
-The preferences at epitope sites resembles those for all HA sites, with preferences increasing over time (Figure~\ref{fig:sequence_preference}A).
-At non-epitope sites, however, the sequence preferences are higher than at epitope sites, and remain relatively constant over time.
-In the timespan from 1968 to 2012, the trunk fixed X epitope mutations per site and Y non-epitope mutations per site.
-Epistasis in HA is seemingly predominantly driven by antigenic drift \comment{cite kryazhimskiy2011prevalence?}. 
-Extensive epistasis has even been noted for epitope sites within the 220-loop surrounding the receptor binding pocket~\cite{wu2017diversity}.
-The evolutionary success of many epitope mutations are also likely contingent upon the background in which they arise.
-Yet, we are able to distinguish trunk mutations as significantly more favorable than side branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B). 
-This suggests that the preferences are still of utility in distinguishing evolutionarily successful strains from non-successful ones over short evolutionary timescales.
+Mutations that occur on the trunk are scored as significantly more favorable in our experiments than side-branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}A).
+Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and ones that are also under immune pressure.
+The fact that our experiments can discriminate between the trunk and side branch at epitope sites despite only assaying for viral growth underscores the fact that most epitope sites are still important for HA function\comment{cite?}.
 
-\begin{SCfigure*}[\sidecaptionrelwidth][t]
+But interestingly, there are clear differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
+The per-site sequence preference at epitope sites increases over time (Figure~\ref{fig:sequence_preference}A \comment{and new associated supplement that zooms in}), whereas at non-epitope sites it remains relatively constant.
+In addition, the per-site sequence preference is consistently higher at non-epitope than epitope sites, perhaps consistent with the fact that strong immune selection can lead to the fixation of less functionally favorable mutations at epitope sites.
+The fact that the epitope sites exhibit more of an increase in sequence preference over time may be due to the fact that they fix more mutations.
+It is known that as mutations accumulate at protein sites, they can lead to epistatic shifts in the effects of mutations\comment{general epistasis appears}.
+Such epistasis has been experimentally demonstrated for HA\comment{Yewdell paper, maybe Scott Hensley has something else}, including at some specific epitope sites in H3 HA~\cite{wu2017diversity}.
+We hypothesize that the rapid evolution of epitope sites in HA leads to epistatic interactions that shift the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA.
+
+\begin{figure}
 \centering
-\includegraphics[width=11.4cm]{figs/sequence_preference/sequence_preference.pdf}
+\includegraphics[width=\columnwidth]{figs/sequence_preference/sequence_preference.pdf}
 \caption{\label{fig:sequence_preference}
-{\bf Epistasis has been driven by substitutions at epitope sites.}
-(A) We calculated the HA sequence preference normalized by the number of sites for epitope and non-epitope sites, as defined by~\cite{wolf2006long}.
-The preferences at all sites and at epitope sites increase as the tree approaches the Perth/2009 strain.
-The sequence preference of non-epitope sites is higher than that of epitope sites.
-(B) The log$_{2}$ mutational effect for side branch and trunk mutations at non-epitope (left) and epitope (right) sites are shown.
+\comment{Switch panels A and B, the legend is already written to reflect that.}
+{\bf Effects of mutations at epitope and non-epitope sites during HA evolution.}
+We partitioned HA into epitope and non-epitope sites using the definitions of Wolf \textit{et al}~\cite{wolf2006long}.
+(A) The log$_{2}$ mutational effects for side branch and trunk mutations at non-epitope (left) and epitope (right) sites.
+$P$-values were computed as in Figure~\ref{fig:trunkvssidebranch} but only performing the randomizations over the appropriate set of sites (non-epitope or epitope).
+(B) 
+The average per-site sequence preference for all nodes in the phylogenetic tree, calculated separately for each set of sites.
+\comment{Write the P-value as $<10^{-4}$ if \emph{all} $10^4$ randomizations are less than the actual value. Otherwise write something like 0.0007 or (maybe if a lot of zeros $7\times10^{-4}$.
+Add supplementary figure for the zoom-in of non-epitope sites and then refer to it in the Figure legend here.
+You could even just add the old figure that split them and shows both epitope and non-epitope on their own y-axis as the supplement.
+I think the randomizations should be done just within the same set of sites.}
 }
-\end{SCfigure*}
+\end{figure}
+
+\subsection*{Somehow merged into next subsection}
 

### 2018-01-08 paper/manuscript.tex
edits to legends for H3 vs H1

@@
 
-\begin{figure*}
+\begin{figure}
 \centering
-\includegraphics[width=15cm]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}
+\includegraphics[width=\columnwidth]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}
 \caption{\label{fig:WSN_trunkvssidebranch}
-{\bf The WSN/1933 H1 preferences do not reveal differences in trunk vs side branch mutational effects}
-(A) We used the WSN/1933 H1 preferences experimentally measured in~\cite{doud2016accurate} to calculate the log$_{2}$ mutational effects of trunk and side branch mutations from the inferred H3N2 phylogeny in Figure~\ref{fig:trunkvssidebranch}.
-(B) We also performed a sliding window analysis with the WSN/1933 H1 preferences.
+{\bf Experimental measurements on an H1 HA do \emph{not} provide evolutionarily relevant information for H3 HAs.}
+(A), (B) This figure is analogous to that in Figure~\ref{fig:trunkvssidebranch} except that it scores the H3 sequences using experimental measurements made on the lab-adapted WSN/1933 HA~\cite{doud2016accurate}.
+\comment{Make into two vertical panels}
 }
-\end{figure*}
+\end{figure}
 
@@
 \caption{\label{fig:distance_distribution}
-{\bf The HA homologs exhibit many large shifts in preference compared to shifts for other viral protein homologs}
-(A) A phylogenetic tree of the HA subtypes, with the two HA's, WSN/1933 H1 and Perth/2009 H3, for which we have measured amino-acid preferences denoted on the tree. 
-The WSN/1933 H1 and the Perth/2009 H3 share $\sim$42\% amino-acid identity.
-(B) The correlations of the amino-acid preferences for replicates both within and between the two HA homologs. 
-The within-Perth/2009 and the within-WSN/1933 correlations are shown in purple and brown, respectively.
-The between homolog correlations are in gray.
-(C) The distribution of shifts in preference for various homolog pairs are shown.
-The top two distributions show the distances between each of the HA homologs with the non-homologous HIV Env~\cite{haddox2017mapping}. %BF520
-The center distribution shows the corrected distance between the two HIV Env homologs, which share 86\% amino-acid identity.
-The fourth distribution from the top is a null generated by randomizing the HA replicates and computing the distances.
-The bottom distribution is the corrected distance between the Perth/2009 H3 and WSN/1933 H1 homologs at all sites that align.
+{\bf There are substantial differences in the effects of mutations between H1 and H3 HAs.}
+(A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled. 
+These HAs have $\sim$42\% amino-acid identity\comment{get the exact number rounded the percent and remove the $\sim$ symbol}.
+(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study  and the three replicates in prior deep mutational scanning of an H1 HA~\cite{doud2016accurate}.
+Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray. 
+$R$ indicates the Pearson correlation coefficient.
+(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2017mapping}, and plotted the distribution of shifts for all sites.
+The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences to experimental noise.
+The shifts are also much larger than those previously observed between two variants of HIV envelope protein (Env) that share 86\% amino-acid identity (pink).
+However, the shifts between H3 and H1 are still less than the differences between either HA and HIV Env (green).
+\comment{Let's remove RMSDcorrected from legend.}
 }

### 2018-01-09 paper/manuscript.tex
updates to shifts between HAs section

@@
 
-\subsection*{Somehow merged into next subsection}
-
-Can we then distinguish lineage-specific mutational effects using the preferences measured in a distantly related HA homolog?
-We used the WSN/1933 H1 preferences measured in~\cite{doud2016accurate} to quantify the effects of H3 trunk and side branch mutations, shown in Figure~\ref{fig:WSN_trunkvssidebranch}.
-It is evident that we do not see trunk mutations significantly more favored than side branch mutations, suggesting that our ability to discriminate successful and unsuccessful strains degrades over sufficiently long evolutionary distances.
+\subsection*{There are large differences between between H3 and H1 HAs in the amino-acid preferences of many sites}
+The results above show that measurements of mutational effects in the Perth/2009 HA are informative about the evolutionary fate of other human H3N2 HAs.
+How broadly can experimental measurements be generalized across HAs?
+To address this question, we repeated the trunk versus side-branch analyses using amino-acid preferences from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
+Figure~\ref{fig:WSN_trunkvssidebranch} shows that the H1 measurements are not informative for distinguishing the trunk and side branches of an H3 phylogeny.
+This fact indicates that the utility of an experiment for discriminating successful and unsuccessful strains degrades over sufficiently long evolutionary distances.
 
@@
 
-\subsection*{The H1 and H3 preferences have shifted at many sites}
-How shifted are the preferences between evolutionarily distant homologs such as H1 and H3?
-Although we have previously compared the experimentally measured preferences between related protein homologs of NP~\cite{doud2015site} and of Env~\cite{haddox2017mapping}, the HA homologs we have studied are considerably more diverged than either of these pairs.
-The WSN/1933 H1 and the Perth/2009 H3 fall into two phylogenetically distinct groups and share 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A), allowing us to examine what has shifted and what has remained conserved across such diverged homologs.
-Simply correlating the preferences between H1 and H3 reveals that the replicate measurements within a homolog are more correlated than between homologs (Figure~\ref{fig:distance_distribution}B).
+An obvious hypothesis for why the H1 deep mutational scanning is not informative about the evolutionary fate of H3 strains is that the effect of the same mutation will often by very different between these two HA subtypes.
+To determine if this is the case, we can examine how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
+Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity\comment{check this number}~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2017mapping}.
+However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
+One simple way to investigate the extent of shifts in amino-acid preferences is simply to correlate measurements from independent deep mutational scanning replicates on H1 and H3 HAs.
+Figure~\ref{fig:distance_distribution}B clearly reveals that replicate measurements on the same HA variant are more correlated than those on different HA variants.
 
-However, to quantify the extent of mutational shifts at a site-by-site level, we used an approach described in~\cite{doud2015site,haddox2017mapping}.
-We aligned Perth/2009 H3 and WSN/1933 H1 and at each alignable site calculated the difference in preferences between homologs while correcting for experimental noise within homologs.
-The distribution of shifts is shown in Figure~\ref{fig:distance_distribution}C.
-Although many sites have small shifts near zero, a considerable number of sites have large shifts in preference, reaching a difference of 0.86 out of a maximum possible of 1.0.
-When we compare HA to the preferences measured in the non-homologous protein HIV Env~\cite{haddox2017mapping}, nearly all sites are shifted.
-On the other hand, when we randomize the HA replicates to generate a null distribution, there is very little shift in preferences. 
+To more rigorously quantify shifts in amino-acid preferences after correcting for experimental noise, we used the statistical approach described in~\cite{doud2015site,haddox2017mapping}.
+Figure~\ref{fig:distance_distribution}C shows the distribution of shifts in amino-acid preferences between H3 and H1 HAs after correcting for experimental noise.
+Although some sites have small shifts near zero, many sites have large shifts in preference.
+These shifts between H3 and H1 are much larger than expected from the null distribution that would be observed purely from experimental noise.
+They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2017mapping}.
+However, the shifts are still generally smaller than those observed when comparing HA to the non-homologous protein HIV Env protein. 
 For comparison, the preferences of the HIV Env homologs, which are 86\% identical at the amino-acid level, are mostly similar. 
-Although there are a small number of sites with larger shifts reaching a maximum of 0.52, we do not see a dramatic tail in the distribution as we see for the HA's.
-These observations suggest that as protein homologs diverge, their preferences increasingly shift.
+These observations show that there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations still remain more similar than for non-homologous proteins.
 
-\begin{figure*}
+\begin{SCfigure*}[\sidecaptionrelwidth][t]
 \centering
-\includegraphics[width=17cm]{figs/distance_distribution/distance_distribution.pdf}
+\includegraphics[width=12cm]{figs/distance_distribution/distance_distribution.pdf}
 \caption{\label{fig:distance_distribution}
@@
 However, the shifts between H3 and H1 are still less than the differences between either HA and HIV Env (green).
-\comment{Let's remove RMSDcorrected from legend.}
+\comment{Let's remove RMSDcorrected from legend.
+Maybe the correlation plot could be slightly smaller, and the other two panels a bit bigger, especially their fonts.}
 }
-\end{figure*}
+\end{SCfigure*}
+
+\subsection*{Properties associated with the shifts in amino-acid preferences between H3 and H1 HAs}
 

### 2018-01-09 paper/manuscript.tex
more updates to end of Results

@@
 \subsection*{Properties associated with the shifts in amino-acid preferences between H3 and H1 HAs}
-
-We next asked if shifted sites tend to cluster together in the HA structure.
-However, the sites of large shifts were interspersed across HA and did not seem to obviously localize together (Figure~\ref{fig:RMSD_heatmap}A).
-Yet, at the domain level, we found the stalk domain to be significantly less shifted than the globular head domain (Figure~\ref{fig:RMSD_heatmap}B). 
-This is unsurprising given that the stalk domain is more conserved both within and across HA subtypes~\cite{nobusawa1991comparison,hai2012influenza,mallajosyula2014influenza}.
-Sites absolutely conserved across all 18 HA subtypes were also found to be significantly less shifted than sites not conserved (Figure~\ref{fig:RMSD_heatmap}B), which is suggestive of the high functional importance of the sites that have remained unchanged throughout the divergence of HA.
+What features distinguish the sites that have shifted versus conserved amino-acid preferences between H3 and H1 HAs?
+The sites of large shifts are spread across HA, and do not obviously localize to one specific region of the protein's structure (Figure~\ref{fig:RMSD_heatmap}A).
+However, at the domain level, sites in the HA stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B). 
+The HA stalk domain is also more conserved at the amino-acid level across HA subtypes~\cite{nobusawa1991comparison,hai2012influenza,mallajosyula2014influenza}, suggesting that conservation of amino-acid sequence tends to be correlated with conservation of amino-acid preferences.
+Consistent with this idea, sites that are absolutely conserved across all 18 HA subtypes are significantly less shifted than sites that are variable across HA subtypes (Figure~\ref{fig:RMSD_heatmap}B).
+Presumably these sites are under consistent functional constraint across all HAs.
 
 Despite their large sequence divergence, H1 and H3 adopt nearly identical protein folds~\cite{ha2002h5,russell2004h1}.
-Previous work has found that although the subdomain structures are highly similar, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among the four HA clades represented by H1 \& H5, H9, H7, and H3~\cite{ha2002h5,russell2004h1}.
-One explanation for such clade-specific subdomain position differences could be local and/or long-range conformational interactions between sites that are variable between clades \comment{needs better wording}.
-We would expect such sites to exhibit large shifts in preference.
-Indeed, clade-specific sites that are conserved within the H1 clade (including H1, H2, H5, and H6) or within the H3 clade (including H3, H4, and H14) are significantly shifted compared to non-clade-specific sites.
-
-In particular, the clade-specific upward shift of the globular head relative to the stalk has largely been attributed to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
-The H1 HA has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
-In the WSN/1933 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
+Previous work has found that although the subdomain structures are highly similar, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among different HA subtypes~\cite{ha2002h5,russell2004h1}.
+This previous work has defined clades of closely related HA subtypes. 
+One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}).
+The clade containing H1 has an upward shift of the globular head relative to the clade containing H3.
+This shift in the head subdomain has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
+Specifically, clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
+In deep mutational scanning of the WSN/1933 H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
 In contrast, the interhelical loop in H3 HA makes a sharper and shorter turn which is facilitated by a Gly at 75(HA2).
-This site prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations suggesting that the amino-acid identity at this site plays a smaller role in determining the conformation of the interhelical loop in H3.
+In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations suggesting that the amino-acid identity at this site plays a smaller role in determining the conformation of the interhelical loop.
+Therefore, some of the shifts in HA amino-acid preferences can be directly rationalized in terms of changes in HA structure.
+\comment{We still need to site the last part of Figure~\ref{fig:RMSD_heatmap}B and also update this figure legend}
 

### 2018-01-09 paper/manuscript.tex
finished revisions to all of Results

@@
 
-After generating the mutant virus libraries, we passaged them at low MOI in cell culture to create a genotype-phenotype link and select for functional HA variants (Figure~\ref{fig:dms_overview}A).
-All of the experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B). 
-We also passaged and deep sequenced library 3 in duplicate (denoted as library 3-1 and 3-2) to gauge to the amount of experimental noise occurring \textit{within} a single biological replicate.
-As a control to measure sequencing and mutational errors, we used the unmutated HA gene to generate and passage viruses carrying wildtype HA.
-
-Deep sequencing of the initial plasmid mutant libraries and the passaged mutant viruses allowed us to observe selection for functional HA mutants.
-As expected, selection against non-functional HA variants as led to reduced mutation frequencies in the mutant viruses compared to the initial mutant plasmids (Figure~\ref{fig:dms_overview}C).
-Specifically, stop codons were purged to 20-45\% of their initial frequencies after correcting for error rates estimated by sequencing the wildtype controls.
-The incomplete purging of stop codons is likely because genetic complementation due to co-infection~\cite{marshall2013influenza} enabled the persistence of some virions with nonfunctional HAs. 
-We also observed selection against many nonsynonymous mutations, with their frequencies falling to 30-40\% of their initial values after error correction.
-
 \begin{SCfigure*}[\sidecaptionrelwidth][t]
@@
 \end{SCfigure*}
+
+After generating the mutant virus libraries, we passaged them at low MOI in cell culture to create a genotype-phenotype link and select for functional HA variants (Figure~\ref{fig:dms_overview}A).
+All of the experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B). 
+We also passaged and deep sequenced library 3 in duplicate (denoted as library 3-1 and 3-2) to gauge to the amount of experimental noise occurring \textit{within} a single biological replicate.
+As a control to measure sequencing and mutational errors, we used the unmutated HA gene to generate and passage viruses carrying wildtype HA.
+
+Deep sequencing of the initial plasmid mutant libraries and the passaged mutant viruses allowed us to observe selection for functional HA mutants.
+As expected, selection against non-functional HA variants as led to reduced mutation frequencies in the mutant viruses compared to the initial mutant plasmids (Figure~\ref{fig:dms_overview}C).
+Specifically, stop codons were purged to 20-45\% of their initial frequencies after correcting for error rates estimated by sequencing the wildtype controls.
+The incomplete purging of stop codons is likely because genetic complementation due to co-infection~\cite{marshall2013influenza} enabled the persistence of some virions with nonfunctional HAs. 
+We also observed selection against many nonsynonymous mutations, with their frequencies falling to 30-40\% of their initial values after error correction.
 
@@
 
-\begin{table*}
+\begin{table}
 \caption{\label{tab:phydms}
@@
 \hline
-\bf{Model} & \bf{$\Delta$AIC} & \bf{Log Likelihood} & \bf{Stringency} & \bf{$\omega$}  \\ \hline
-ExpCM & 0.0 & -8439.33 & 2.44 & 0.91 \\
-Goldman-Yang M5 & 2166.06 & -9516.36 & -- & 0.30, 0.84, 0.36 \\
-ExpCM, averaged across sites & 2504.18 & -9691.42 & 0.68 & 0.32 \\
-Goldman-Yang M0 & 2607.92 & -9738.29 & -- & 0.31 \\
+\bf{Model} & \bf{$\Delta$AIC} & \bf{LnL} & \bf{Stringency} & \bf{$\omega$}  \\ \hline
+ExpCM & 0.0 & -8439 & 2.44 & 0.91 \\
+GY94 M5 & 2166 & -9516 & -- & 0.36 (0.30, 0.84) \\
+ExpCM, averaged & 2504 & -9691 & 0.68 & 0.32 \\
+GY94 M0 & 2608 & -9738 & -- & 0.31 \\
 \hline
@@
  \end{center}
-\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 influenza HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites, and the M0 and M5 versions of the Goldman-Yang model~\cite{yang2000codon}.
-Models are compared by AIC~\cite{posada2004model}.
+\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 influenza HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites, and the M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
+Models are compared by AIC~\cite{posada2004model} computed from the log likelihood (LnL) and number of model parameters.
 The $\omega$ parameter is dN/dS for the Goldman-Yang models, and the relative dN/dS after accounting for the measurements for the ExpCM.
-For the M5 model, we give the shape parameter, rate parameter, and mean of the gamma distribution over $\omega$.
+For the M5 model, we give the mean followed by the shape and rate parameters of the gamma distribution over $\omega$.
 }
-\end{table*}
+\end{table}
 
@@
 
-\begin{figure*}
+\begin{figure*}[ht]
 \centering
@@
 
-\begin{SCfigure*}[\sidecaptionrelwidth][t]
+%\begin{SCfigure*}[\sidecaptionrelwidth][t]
+\begin{figure}
 \centering
-\includegraphics[width=11.4cm]{figs/mut_tolerance/entropy_heatmap.pdf}
+%\includegraphics[width=11.4cm]{figs/mut_tolerance/entropy_heatmap.pdf}
+\includegraphics[width=\columnwidth]{figs/mut_tolerance/entropy_heatmap.pdf}
 \caption{\label{fig:mut_tolerance}
@@
 }
-\end{SCfigure*}
+%\end{SCfigure*}
+\end{figure}
 
@@
 One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}).
+Sites that are conserved at different amino-acid identities in these two clades tend to have exceptionally large shifts in amino acid preferences (Figure~\ref{fig:RMSD_heatmap}B).
 The clade containing H1 has an upward shift of the globular head relative to the clade containing H3.
@@
 Therefore, some of the shifts in HA amino-acid preferences can be directly rationalized in terms of changes in HA structure.
-\comment{We still need to site the last part of Figure~\ref{fig:RMSD_heatmap}B and also update this figure legend}
 
-\begin{figure*}
+\begin{SCfigure*}[\sidecaptionrelwidth][t]
 \centering
-\includegraphics[width=17cm]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
+\includegraphics[width=11.4cm]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
 \caption{\label{fig:RMSD_heatmap}
-{\bf Characterization of the preference shifts}
-(A) The preference shifts as calculated by $\textit{RMSD}_\textit{corrected}$ between the two HA homologs is mapped onto the structure of HA (PDB 4O5N;~\cite{lee2014receptor}). 
-The left structure shows the HA trimer, and the right structure colors one of the monomers. 
-The sialic acid receptor is shown in black sticks.
-Blue indicates small shifts in preference near zero, while red indicates large shifts in preference.
-(B) The stalk domain was found to be significantly less shifted than the head domain (left plot).
-Sites absolutely conserved all 18 HA subtypes were also found to be significantly less shifted than the remaining non-conserved sites (middle plot).
-Sites specific to the clade containing H1, H2, H5, and H6 and specific to the H3, H4, and H14 clade are significantly more shifted than non-clade-specific sites (right plot).
-(C) HA group-specific interactions between sites 107 and 75(HA2) play a role in determining the upward translation of the globular head domain relative to the stalk domain.
-The amino acids at sites 107 and 75(HA2) are shown in spheres on the structure of H1 and H3. 
-One monomer is shown in dark gray, while the HA1 domain of the neighboring monomer is in lighter gray.
-The sphere colors correspond to the shift in preference at that site.
-The preferences for each site in WSN/1933 and Perth/2009 are also shown.
+{\bf Sites with strongly shifted amino-acid preferences between H3 and H1 HAs.}
+(A) The shift in amino-acid preferences at each site as calculated in Figure~\ref{fig:distance_distribution}C between the H3 and H1 HA is mapped onto the structure of the H3 HA. 
+(B) Amino-acid preferences of sites in the stalk domain are significantly less shifted than those in the head domain.
+Sites absolutely conserved in all 18 HA subtypes are significantly less shifted than other sites.
+Sites with one amino-acid identity in the clade containing H1, H2, H5, and H6 and another identity in the clade containing H3, H4, and H14 are significantly more shifted than other sites.
+(C) Sites 107 and 75(HA2) help determine the different orientation of the globular head domain in H1 versus H3 HAs.
+These sites are shown in spheres on the structure of H1 and H3. 
+One monomer is in dark gray, while the HA1 domain of the neighboring monomer is in lighter gray.
+The spheres are colored as in panel (A), and the experimentally measured amino-acid preferences in the WSN/1933 H1 and Perth/2009 H3 HAs are shown.
+\comment{some of the text needs to be larger in panel (B)}
 }
-\end{figure*}
+\end{SCfigure*}
 

### 2018-01-13 paper/manuscript.tex
Add zoomed sequence preference supplementary figure
First pass of edits to Results

@@
 If this is the case, add \parshape=0 to the end of the list environment.
-\dropcap{S}easonal H3N2 influenza virus evolves rapidly, readily accumulating mutations in the hemagglutinin (HA) surface protein that contribute to antigenic drift.
+\dropcap{S}easonal H3N2 influenza virus evolves rapidly, readily accumulating mutations in the hemagglutinin (HA) surface protein that contribute to antigenic drift~\cite{smith2004mapping,bhatt2011genomic}.
 The evolution of human H3N2 influenza virus is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2015global}, producing a distinct spindly shape in its phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
@@
 For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} possess high preference for cysteine.
-At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change}\comment{These old citations are for X-31 HA, you might also see if Paulson and/or Ian Wilson have newer structural papers on receptor-binding in recent H3 -- if so, add them}.
+At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change,yang2015structure}.
 A positively charged amino acid at site 329 is important for cleavage activation of the HA0 precursor, and indeed this site exhibits a high preference for arginine~\cite{kido1992isolation, stech2005new}.
@@
 \subsection*{The experimental measurements can help discriminate between successful and unsuccessful influenza virus lineages}
-A major goal in the study of rapidly evolving viruses such as influenza is to forecast evolutionary trajectories\comment{Michael Lassig has a Nature Ecology \& Evolution review}.
-For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection\comment{add reference}. 
+A major goal in the study of rapidly evolving viruses such as influenza is to forecast evolutionary trajectories~\cite{lassig2017predicting,morris2017predictive}.
+For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection~\cite{neher2015nextflu,lassig2017predicting,morris2017predictive}.
 Evolutionary forecasts must ultimately distinguish between successful and unsuccessful viral lineages, which in the case of human influenza virus means distinguishing between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).
@@
 Our experiments were performed on the Perth/2009 HA, but we are scoring mutations on a phylogenetic tree that extends from 1968 to 2017.
-Because the effects of mutations can vary with genetic background\comment{cite some papers about epistasis, amino-acid shifts, etc}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
+Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
 To explore this question, we scored the complete HA sequence of every node in the phylogenetic tree by quantifying its average per-site sequence preference for HA sequence $\mathbf{s}$, which is defined as
@@
 {\bf Mutations on the trunk are more favorable than those on the side branches according to our experimental measurements.}
-\comment{Legend will be written under assumption that panels (B) and (C) are swapped. Also add sequence preference \emph{per site} to y-axis of the new (C).}
 (A) We used our experiments to calculate the log$_{2}$ mutational effect for all trunk and side branch mutations in five-year windows from 1968 to 2013. 
@@
 Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites.
-In the timeframe from 1968 to 2012, the trunk of the tree in Figure~\ref{fig:H3N2_phylogeny} fixed \comment{X} epitope mutations per site, but only Y non-epitope mutations per site.
+In the timeframe from 1968 to 2012, the trunk of the tree in Figure~\ref{fig:H3N2_phylogeny} fixed 0.84 epitope mutations per site, but only 0.07 non-epitope mutations per site.
 Since our experiments only measure how mutations affect viral growth, they might be expected to describe the evolution of epitope and non-epitope sites differently.
@@
 Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and ones that are also under immune pressure.
-The fact that our experiments can discriminate between the trunk and side branch at epitope sites despite only assaying for viral growth underscores the fact that most epitope sites are still important for HA function\comment{cite?}.
+The fact that our experiments can discriminate between the trunk and side branch at epitope sites despite only assaying for viral growth underscores the fact that most epitope sites are still important for HA function~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.
 
 But interestingly, there are clear differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
-The per-site sequence preference at epitope sites increases over time (Figure~\ref{fig:sequence_preference}A \comment{and new associated supplement that zooms in}), whereas at non-epitope sites it remains relatively constant.
+The per-site sequence preference at epitope sites increases over time (Figure~\ref{fig:sequence_preference}A and Figure~\ref{suppfig:seqpref_zoomed}), whereas at non-epitope sites it remains relatively constant.
 In addition, the per-site sequence preference is consistently higher at non-epitope than epitope sites, perhaps consistent with the fact that strong immune selection can lead to the fixation of less functionally favorable mutations at epitope sites.
 The fact that the epitope sites exhibit more of an increase in sequence preference over time may be due to the fact that they fix more mutations.
-It is known that as mutations accumulate at protein sites, they can lead to epistatic shifts in the effects of mutations\comment{general epistasis appears}.
-Such epistasis has been experimentally demonstrated for HA\comment{Yewdell paper, maybe Scott Hensley has something else}, including at some specific epitope sites in H3 HA~\cite{wu2017diversity}.
+It is known that as mutations accumulate at protein sites, they can lead to epistatic shifts in the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
+Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some specific epitope sites in H3 HA~\cite{wu2017diversity}.
 We hypothesize that the rapid evolution of epitope sites in HA leads to epistatic interactions that shift the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA.
@@
 \caption{\label{fig:sequence_preference}
-\comment{Switch panels A and B, the legend is already written to reflect that.}
 {\bf Effects of mutations at epitope and non-epitope sites during HA evolution.}
@@
 The average per-site sequence preference for all nodes in the phylogenetic tree, calculated separately for each set of sites.
-\comment{Write the P-value as $<10^{-4}$ if \emph{all} $10^4$ randomizations are less than the actual value. Otherwise write something like 0.0007 or (maybe if a lot of zeros $7\times10^{-4}$.
-Add supplementary figure for the zoom-in of non-epitope sites and then refer to it in the Figure legend here.
-You could even just add the old figure that split them and shows both epitope and non-epitope on their own y-axis as the supplement.
-I think the randomizations should be done just within the same set of sites.}
+The sequence preferences for epitope and non-epitope sites on separate y-axes is shown in Figure~\ref{suppfig:seqpref_zoomed}.
 }
@@
 (A), (B) This figure is analogous to that in Figure~\ref{fig:trunkvssidebranch} except that it scores the H3 sequences using experimental measurements made on the lab-adapted WSN/1933 HA~\cite{doud2016accurate}.
-\comment{Make into two vertical panels}
 }
@@
 To determine if this is the case, we can examine how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
-Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity\comment{check this number}~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2017mapping}.
+Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2017mapping}.
 However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
@@
 (A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled. 
-These HAs have $\sim$42\% amino-acid identity\comment{get the exact number rounded the percent and remove the $\sim$ symbol}.
+These HAs have 42\% amino-acid identity.
 (B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study  and the three replicates in prior deep mutational scanning of an H1 HA~\cite{doud2016accurate}.
@@
 However, the shifts between H3 and H1 are still less than the differences between either HA and HIV Env (green).
-\comment{Let's remove RMSDcorrected from legend.
-Maybe the correlation plot could be slightly smaller, and the other two panels a bit bigger, especially their fonts.}
 }
@@
 The spheres are colored as in panel (A), and the experimentally measured amino-acid preferences in the WSN/1933 H1 and Perth/2009 H3 HAs are shown.
-\comment{some of the text needs to be larger in panel (B)}
 }
@@
 
+\begin{suppfigure}
+\centerline{\includegraphics[width=0.6\textwidth]{figs/S04_seqpref_zoomed/seqpref_zoomed.pdf}}
+\caption{\label{suppfig:seqpref_zoomed}
+{\bf The per-site sequence preference at epitope and non-epitope sites shown in Figure~\ref{fig:sequence_preference}B, but on their own axes.} 
+}
+\end{suppfigure}
+
 \subsubsection*{SI Tables}

### 2018-01-13 paper/manuscript.tex
Start writing Significance Statement

@@
 % Please add here a significance statement to explain the relevance of your work
-\significancestatement{Authors must submit a 120-word maximum statement about the significance of their research paper written at a level understandable to an undergraduate educated scientist outside their field of speciality.
-The primary goal of the Significance Statement is to explain the relevance of the work in broad context to a broad readership.
-The Significance Statement appears in the paper itself and is required for all research papers.}
+% 120-word max
+\significancestatement{Mutations drive the rapid evolution of seasonal influenza virus such that new strains emerge and co-circulate until one strain overtakes the population. 
+A key challenge in the study of seasonal influenza virus is to be able to differentiate between strains that are likely to persist and those that are likely to die out. 
+We therefore need a greater understanding of how mutations drive and constrain viral evolution in order to help identify persistent and nonpersistent strains. 
+We experimentally measured the effect of every possible amino-acid mutation to the surface protein hemagglutinin on viral growth.
+Using these measurements, we revealed differences between successful and unsuccessful viral lineages and examined how generalizable the measurements were for studying the evolutionary fate of viral lineages.
+\comment{Our work has potential for use in forecasting}}
 

### 2018-01-13 paper/manuscript.tex
Complete edits of results section

@@
 % Please add here a significance statement to explain the relevance of your work
-% 120-word max
-\significancestatement{Mutations drive the rapid evolution of seasonal influenza virus such that new strains emerge and co-circulate until one strain overtakes the population. 
+% 120 words max
+\significancestatement{
+Mutations drive the rapid evolution of seasonal influenza virus such that new strains emerge and co-circulate until one strain overtakes the population. 
 A key challenge in the study of seasonal influenza virus is to be able to differentiate between strains that are likely to persist and those that are likely to die out. 
@@
 Using these measurements, we revealed differences between successful and unsuccessful viral lineages and examined how generalizable the measurements were for studying the evolutionary fate of viral lineages.
-\comment{Our work has potential for use in forecasting}}
+\comment{Our work has potential for use in forecasting}
+}
 
@@
 {\bf Deep mutational scanning of the Perth/2009 H3 HA.}
-(A) We generated mutant virus libraries using a helper-virus approach~\cite{doud2016accurate}, and passaged them at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants.
+(A) We generated mutant virus libraries using a helper-virus approach~\cite{doud2016accurate}, and passaged the libraries at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants.
 Deep sequencing of the variants before and after selection allowed us to estimate each site's amino-acid preferences.
@@
 Deep sequencing of the initial plasmid mutant libraries and the passaged mutant viruses allowed us to observe selection for functional HA mutants.
-As expected, selection against non-functional HA variants as led to reduced mutation frequencies in the mutant viruses compared to the initial mutant plasmids (Figure~\ref{fig:dms_overview}C).
+As expected, selection against non-functional HA variants led to reduced mutation frequencies in the mutant viruses compared to the initial mutant plasmids (Figure~\ref{fig:dms_overview}C).
 Specifically, stop codons were purged to 20-45\% of their initial frequencies after correcting for error rates estimated by sequencing the wildtype controls.
@@
 The biological replicates are fairly well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
-Replicate 1 exhibited the lowest correlation with the other replicates, consistent with the observation that this replicate also showed the weakest selection against stop and nonsynonymous mutations perhaps indicating more experimental noise.
-The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity after the reverse-genetics step contributes most of the experimental noise.
+Replicate 1 exhibited the lowest correlation with the other replicates, consistent with the observation that this replicate also showed the weakest selection against stop and nonsynonymous mutations, perhaps indicating more experimental noise.
+The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity after the reverse-genetics step contributes to most of the experimental noise.
 
@@
 GY94 M5 & 2166 & -9516 & -- & 0.36 (0.30, 0.84) \\
-ExpCM, averaged & 2504 & -9691 & 0.68 & 0.32 \\
+ExpCM, site avg. & 2504 & -9691 & 0.68 & 0.32 \\
 GY94 M0 & 2608 & -9738 & -- & 0.31 \\
@@
  \end{center}
-\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 influenza HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites, and the M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
+\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 influenza HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites (site avg.), and the M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
 Models are compared by AIC~\cite{posada2004model} computed from the log likelihood (LnL) and number of model parameters.
@@
 The sites are in H3 numbering.
-The top overlay bar indicates whether or not a site is part of the set of epitope residues delineated in \cite{wolf2006long}
+The top overlay bar indicates whether or not a site is part of the set of epitope residues delineated in \cite{wolf2006long}.
 The bottom overlay bar indicates the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
@@
 We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we define as the Shannon entropy of the re-scaled preferences.
-In prior mutational studies of the WSN/1933, the stalk domain of this H1 HA was found to be substantially less mutationally tolerant than its globular head~\cite{thyagarajan2014inherent,doud2016accurate}.
+In prior mutational studies of the WSN/1933, the stalk domain of this H1 HA was found to be substantially less mutationally tolerant than its globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
 
@@
 Surprisingly, there was much less contrast in mutational tolerance between the stalk and head domains for the H3 HA than for the H1 (Figure~\ref{fig:mut_tolerance}).
-For instance, the in H3 HA, the short helix A in the stalk is very mutationally tolerant.
+For instance, in the H3 HA, the short helix A in the stalk is very mutationally tolerant.
 Interestingly, more studies have reported selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2017quantifying} HAs.
@@
 Many residues inside HA's receptor binding pocket are known to be highly functionally constrained~\cite{wilson1981structure,martin1998studies}, and our data indicates that these sites are relatively mutationally intolerant in both H3 and H1 HAs.
-In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, as these sites are under strong immune pressure.~\cite{wiley1981structural,koel2013substitutions}.
+In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, as these sites are under strong immune pressure~\cite{wiley1981structural,koel2013substitutions}.
 
@@
 Mutational tolerance as measured in the current study is mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}).
-Mutational tolerance of the WSN/1933 H1 HA as measured in \cite{doud2016accurate} is mapped on the structure of the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}).
+Mutational tolerance of the WSN/1933 H1 HA as measured in \cite{doud2016accurate} is mapped onto the structure of the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}).
 Different color scales are used because measurements are comparable among sites within the same HA, but not necessarily across HAs.
@@
 \end{equation}
-where $\pi_{r, a}$ is the preference for amino-acid $a$ at site $r$ as measured in our experiments (e.g., Figure~\ref{fig:logoplot}), $s_r$ is the amino-acid at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
+where $\pi_{r, s_r}$ is the preference as measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
 Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to be more highly preferred than those of side branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side branch mutations.
@@
 The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
-These results suggest that there is some HA background-dependence in the effects of mutations, but that despite this fact our experimental measurements consistently reveal the selective advantage of the trunk over a more than half-century of viral evolution.
+These results suggest that there is some HA background-dependence in the effects of mutations, but that despite this fact our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.
 
@@
 Mutations that occur on the trunk are scored as significantly more favorable in our experiments than side-branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}A).
-Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and ones that are also under immune pressure.
-The fact that our experiments can discriminate between the trunk and side branch at epitope sites despite only assaying for viral growth underscores the fact that most epitope sites are still important for HA function~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.
+Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and at sites that are also under immune pressure.
+The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that most epitope sites are still important for HA function~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.
 
 But interestingly, there are clear differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
-The per-site sequence preference at epitope sites increases over time (Figure~\ref{fig:sequence_preference}A and Figure~\ref{suppfig:seqpref_zoomed}), whereas at non-epitope sites it remains relatively constant.
+The per-site sequence preference at epitope sites increases over time (Figure~\ref{fig:sequence_preference}B and Figure~\ref{suppfig:seqpref_zoomed}), whereas at non-epitope sites it remains relatively constant.
 In addition, the per-site sequence preference is consistently higher at non-epitope than epitope sites, perhaps consistent with the fact that strong immune selection can lead to the fixation of less functionally favorable mutations at epitope sites.
@@
 
-\subsection*{There are large differences between between H3 and H1 HAs in the amino-acid preferences of many sites}
+\subsection*{There are large differences between H3 and H1 HAs in the amino-acid preferences of many sites}
 The results above show that measurements of mutational effects in the Perth/2009 HA are informative about the evolutionary fate of other human H3N2 HAs.
@@
 
-An obvious hypothesis for why the H1 deep mutational scanning is not informative about the evolutionary fate of H3 strains is that the effect of the same mutation will often by very different between these two HA subtypes.
+An obvious hypothesis for why the H1 deep mutational scanning is not informative about the evolutionary fate of H3 strains is that the effect of the same mutation will often be very different between these two HA subtypes.
 To determine if this is the case, we can examine how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
@@
 They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2017mapping}.
-However, the shifts are still generally smaller than those observed when comparing HA to the non-homologous protein HIV Env protein. 
-For comparison, the preferences of the HIV Env homologs, which are 86\% identical at the amino-acid level, are mostly similar. 
+However, the shifts are still generally smaller than those observed when comparing HA to the non-homologous HIV Env protein. 
 These observations show that there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations still remain more similar than for non-homologous proteins.
@@
 (C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2017mapping}, and plotted the distribution of shifts for all sites.
-The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences to experimental noise.
+The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
 The shifts are also much larger than those previously observed between two variants of HIV envelope protein (Env) that share 86\% amino-acid identity (pink).
@@
 This shift in the head subdomain has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
-Specifically, clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
+Specifically, the clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
 In deep mutational scanning of the WSN/1933 H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
@@
 We have measured the effect of all possible single amino-acid mutations to Perth/2009 H3 on viral growth in cell culture.
+
+\begin{itemize}
+\item{Antigenic characteristics play a large role in determining strain success. We therefore need to know how mutations affect antigenicity.}
+\item{non-HA segments are important too}
+\end{itemize}
 

### 2018-01-15 paper/manuscript.tex
Some restructuring of intro

@@
 \begin{abstract}
-% no more than 250 words. (200 now)
+% no more than 250 words. (199 now)
 The evolution of seasonal H3N2 influenza virus is mediated by the rapid accumulation of mutations. 
@@
 The preferences measured in a distantly related protein homolog do not reveal such differences in viral lineages.
-A comparison of the preferences between the two hemagglutinin homologs suggests that mutational effects increasingly shift as proteins diverge.
+A comparison of the preferences between the two hemagglutinin variants show substantial shifts in the effects of mutations.
 Overall, our work highlights how experimental measurements of mutational effects can be leveraged to gain insight into the evolutionary fates of viral strains and potentially be used to predict viral evolution.
@@
 The evolution of human H3N2 influenza virus is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2015global}, producing a distinct spindly shape in its phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
-Given the rather distinctive pattern of H3N2 evolution and that vaccination remains an effective way of reducing disease burden, a key challenge in selecting a vaccine strain lies in being able to predict the dominant strain in the upcoming influenza season.
+A key challenge in the study of H3N2 evolution lies in being able to predict the dominant strain in the upcoming influenza season, as this information can guide selection of a vaccine strain.
+Fundamentally, in order to make these predictions we must be able to distinguish between strains that will persist as the trunk and strains that will die out.
 
-Vaccine strain selection decisions mainly rely on experimentally characterizing the antigenicity of circulating strains.
-To aid in the timeliness and throughput of making such decisions, previous work have endeavored to accurately predict the evolutionary outcomes of viral strains based on the antigenicity and/or tree structure of viral clades~\cite{steinbruck2014computational,neher2014predicting}.
-However, the accuracy of these predictions are subject to uneven geographical and temporal sampling, and for~\cite{neher2014predicting}, the lack of information about a mutation's effect on antigenicity.
-Others have attempted to model the antigenic properties of viral strains using sequence information or phylogenetic relationships~\cite{sun2013using,harvey2016identification,neher2016prediction}.
-
-While mutations that contribute to antigenic evolution of the virus certainly determine strain success, the non-antigenic impacts of mutations play an important role in determining the persistence of viral strains~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
-Due to the high mutation rate of RNA viruses \comment{citation here} and because the influenza virus does not appreciably recombine within a given segment, deleterious mutations become linked to beneficial ones.
-A predictive fitness model takes into account mutations at both antigenic and non-antigenic sites in HA to infer clade fitness~\cite{luksza2014predictive}.
-In this model, mutations at epitope sites are beneficial to viral fitness, while mutations at non-epitope sites incur a fitness cost.
-Although this assumption may generally be true, the effect of mutations on viral fitness, and particularly on protein function, are much more nuanced.
-We therefore need a greater understanding of the phenotypic effects of mutations on viral growth to comprehend the forces and constraints shaping influenza virus evolution in nature.
+%To aid in the timeliness and throughput of making such decisions, previous work have endeavored to accurately predict the evolutionary outcomes of viral strains based on the antigenicity and/or tree structure of viral clades~\cite{steinbruck2014computational,neher2014predicting}.
+%However, the accuracy of these predictions are subject to uneven geographical and temporal sampling, and for~\cite{neher2014predicting}, the lack of information about a mutation's effect on antigenicity.
+One approach is to model the antigenic properties of circulating viral strains using sequence information or phylogenetic relationships and experimental measurements~\cite{sun2013using,harvey2016identification,neher2016prediction}.
+While mutations that contribute to antigenic evolution of the virus certainly determine strain success, the non-antigenic effects of mutations play an important role in determining the persistence of viral strains~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
+Due to the high mutation rate of RNA viruses~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and because the influenza virus does not appreciably recombine within a given segment, deleterious mutations become linked to beneficial ones.
+Strain fitness is thereby dependent on both the antigenic and non-antigenic effects of mutations, as modeled in~\cite{luksza2014predictive}.
+However, we lack a broad understanding of the phenotypic effects of mutations to HA, particularly on HA function and viral growth.
+We therefore need a greater understanding of the effects of mutations to HA to comprehend the forces and constraints shaping influenza virus evolution in nature.
 
@@
 Previously, we experimentally measured the effect of all possible single amino-acid mutations to an H1 HA from the A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,doud2016accurate}.
-However, the WSN/1933 is a highly lab-adapted strain, and these measurements may not be particularly relevant for studying the mutational processes of more contemporaneous strains circulating in the human population.
+However, WSN/1933 is a highly lab-adapted strain, and these measurements may not be particularly relevant for studying the mutational processes of more contemporaneous strains circulating in the human population.
 Here we measure the effects of mutations to the HA from the A/Perth/16/2009 (H3N2) strain on viral growth in cell culture.
@@
 We also find evidence that epistasis in HA has primarily been driven by antigenic drift.
-The measurements also enabled a comparison of how the amino-acid preferences have shifted between two diverged HAs.
+The measurements also enabled a comparison of the shifts in amino-acid preferences between two diverged HAs.
 The extent of shifts in amino-acid preferences increase as homologs diverge.

### 2018-01-15 paper/manuscript.tex
Start adding to Methods

@@
 \subsection*{Creation of MDCK-SIAT1-TMPRSS2 cell line}
-The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter and attached to mCherry through an IRES...etc etc
+The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter and attached to mCherry through an IRES. 
+We generated lentiviruses..etc etc transduced MDCK-SIAT1 cells, sorted by fluorescence activated cell sorting
 \comment{Need to look at Katie's notebooks for this...}
@@
 Bulk RNA from the viral sample was extracted using the QIAamp Viral RNA Mini Kit (QIAGEN) according to manufacturer's instructions.
-The Perth/2009 HA and NA genes were then reverse transcribed, PCR amplified, and cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2 \comment{cite?} plasmid backbones.
+The Perth/2009 HA and NA genes were then reverse transcribed using the SuperScript III First-Strand Synthesis SuperMix Kit (Invitrogen), PCR amplified, and cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2 \comment{cite?} plasmid backbones by In-Fusion.
 
-The codon-mutant libraries were generated using a PCR-based approach described in~\cite{bloom2014experimentally,dingens2017comprehensive}.
+To generate wildtype Perth/2009 H3N2 virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 cells with reverse genetics plasmids encoding the Perth/2009 HA and NA, WSN/1933 internal genes, and the TMPRSS2 lentiviral vector.
+\comment{add more details about reverse genetics?}
+The viral supernatant was harvested at 72 hours post-transfection, clarified at 1200 rpm for 5 min, and 0.01 to 1.0 ul of the supernatant was passaged onto $4 \times 10^{5}$ MDCK-SIAT1-TMPRSS2 cells in six-well plates for a total of six serial passages.
+After the final passage, RNA was extracted from 140 ul of the viral supernatant using the QIAamp Viral RNA Mini Kit.
+The HA and NA genes were reverse transcribed, PCR amplified, and sequenced by Sanger sequencing.
+Two mutations, G78D and T212I, arose after serial passaging. 
+The Perth/2009 HA carrying these two mutations were cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2 \comment{cite?} plasmid backbones.
+
+The codon-mutant libraries were generated in the Perth/2009 HA-G78D-T212I background using a PCR-based approach described in~\cite{bloom2014experimentally,dingens2017comprehensive} using two rounds of mutagenesis.
+The codon mutations were introduced into three independent replicates of the Perth/2009 HA variant.
+The mutant variants were then cloned into the pICR2 vector by digestion with BsmBI restriction sites, ligation by T4 DNA ligase, and electroporation and transformation into ElectroMAX DH10B competent cells (Invitrogen; 18290015).
+We pooled over 6 million transformants for each replicate, cultured in liquid LB + ampicillin at 37$^{\circ}$C for 3 h with shaking, and maxiprepped.
 
@@
 The mutant virus libraries were generated and passaged using the approach described in~\cite{doud2016accurate} with several modifications.
+We used the HA-deficient WSN/1933 helper virus generated in~\cite{doud2016accurate}.
+Briefly, these helper viruses were generated by transfecting seven WSN/1933 non-HA reverse-genetics plasmids plus a WSN/1933 HA protein expression plasmid into a coculture of 293T and MDCK-SIAT1-EF1$\alpha$-WSN-HA cells engineered to constitutively express the WSN/1933 HA.
+The transfection supernatant was then expanded by passaging onto a monolayer of MDCK-SIAT1-EF1$\alpha$-WSN-HA cells, and the viral supernatant was then harvested and aliquots were frozen at -80$^{\circ}$C.
+
+To generate the mutant virus libraries, we transfected $5 \times 10^5$ MDCK-SIAT1-TMPRSS2 cells in suspension with 937.5 ng each of four protein expression plasmids encoding the ribonucleoprotein complex (HDM-Nan95-PA, HDM-Nan95-PB1, HDM-Nan95-PB2, and HDM-Aichi68-NP)~\cite{gong2013stability}, and 1250 ng of one of the three pICR2-mutant-HA libraries (or the wildtype pICR2-Perth2009-HA-G78D-T212I control) using Lipofectamine3000 (ThermoFisher;L3000008).
+We allowed the transfected cells to adhere in 6-well plates, then four hours later changed the media to D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL).
+Eighteen hours after transfection, we infected the cells with WSN/1933 HA-deficient helper virus by preparing an inoculum of 500 TCID$_{50}$ per $\mu$L in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL), and aspirating the D10 media from the cells, and adding 2 mL of the helper-virus inoculum to each well.
+After three hours, we changed the media to fresh influenza growth media.
+Twenty-four hours after helper-virus infection, we harvested the viral supernatants for each replicate, froze aliquots at -80$^{\circ}$C, and titered in MDCK-SIAT1-TMPRSS2 cells.
+
+We passaged over $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
+To passage, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which the cells reached a density of $\sim 1.7 \times 10^7$ cells per dish.
+We then replaced the media of each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L.
+We collected viral supernatant for sequencing 48 hours post-infection.
 
 \subsection*{Barcoded subamplicon sequencing}
+To extract viral RNA from the three replicate HA virus libraries and the wildtype HA virus, we first ultracentrifuged 24 mL of the supernatant at 22,000 rpm for 1.5 h at 4$^{\circ}$C in a Beckman Coulter SW28 rotor.
+We then extracted the RNA using the Qiagen RNeasy Mini Kit by resuspending the viral pellet in 400 $\mu$L of buffer RLT supplemented with $\beta$-mercaptoethanol, pipetting 30 times, transferring the liquid to a microcentrifuge tube, adding 600 $\mu$L 70\% ethanol, and proceeding with the rest of the extraction according to manufacturer's instructions.
+The HA gene was then reverse-transcribed with AccuScript Reverse Transcriptase (Agilent 200820) using the primers P09-HA-For (5'-AGCAAAAGCAGGGGATAATTCTATTAATC-3') and P09-HA-Rev (5'-AGTAGAAACAAGGGTGTTTTTAATTACTAATACAC-3').
+
+We generated the HA PCR amplicons for each of the three plasmid libraries, the three virus libraries, one wildtype plasmid, and one wildtype virus samples using KOD Hot Start Master Mix (EMD Millipore;71842) according to the PCR reaction mixture and cycling conditions described in~\cite{bloom2014experimentally} and the P09-HA-For and P09-HA-Rev primers.
+\comment{used the same conditions as Mike - cite his Viruses paper to be succinct?}.
+We performed sequencing on a lane of a flow cell of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.
 
 \subsection*{Analysis of deep sequencing data}
+We used dms\_tools2...
 
 \subsection*{Inference of phylogenetic trees}
-\comment{We downloaded X sequences from the Influenza Virus Resource ?.... etc. inferred the tree, ancestral state reconstruction, visualized the tree...}
+\comment{John to add methods here. We downloaded X sequences from the Influenza Virus Resource ?.... etc. inferred the tree, ancestral state reconstruction, visualized the tree...}
 To parse out trunk mutations from side branch mutations, we first defined a set of recent nodes sampled on or after Jan. 1, 2017, and traced these nodes back to their most recent common ancestor.

### 2018-01-15 paper/manuscript.tex
Add a few bullet points to Discussion
Add a bit more to Methods

@@
 \begin{itemize}
+\item{These experimental measurements have utility in revealing differences between successful and unsuccessful lineages over 50 y of human influenza virus evolution.}
+\item{We have only measured the effect of mutations to HA on viral growth, but there are certainly other determinants of strain success}
 \item{Antigenic characteristics play a large role in determining strain success. We therefore need to know how mutations affect antigenicity.}
@@
 \subsection*{Inference of phylogenetic trees}
-\comment{John to add methods here. We downloaded X sequences from the Influenza Virus Resource ?.... etc. inferred the tree, ancestral state reconstruction, visualized the tree...}
+\comment{John to add methods here. We downloaded X sequences from the Influenza Virus Resource ?.... etc. subsampled how? inferred the tree, ancestral state reconstruction, visualized the tree...}
 To parse out trunk mutations from side branch mutations, we first defined a set of recent nodes sampled on or after Jan. 1, 2017, and traced these nodes back to their most recent common ancestor. 
+We defined all branches ancestral to this most recent common ancestor of recently sampled nodes as the trunk.
+All nodes descended from the trunk were defined as side branch nodes.
+Nodes descended from the most recent common ancestor were defined as unresolved.
 
 \subsection*{Quantification of mutational effects and sequence preferences from an H3N2 phylogeny}
+
+\subsection*{Analysis of mutational shifts}
 
@@
 
-\acknow{We thank Sarah Hilton, Hugh Haddox, Sidney Bell...the Fred Hutch Genomics Core...
+\acknow{We thank Sarah Hilton, Hugh Haddox, Sidney Bell for helpful discussions about data analysis...the Fred Hutch Genomics Core...
 Funding...}

### 2018-01-15 paper/manuscript.tex
Add some more to methods

@@
 \caption{\label{fig:logoplot}
-{\bf The site-specific amino-acid preferences of the Perth/2009 HA measure in our experiments.}
+{\bf The site-specific amino-acid preferences of the Perth/2009 HA measured in our experiments.}
 The height of each letter is the preference for that amino acid, after taking the average of our experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter in Table~\ref{tab:phydms}.
@@
 
-\subsection*{Inference of phylogenetic trees}
+\subsection*{Inference of human H3N2 phylogenetic tree}
 \comment{John to add methods here. We downloaded X sequences from the Influenza Virus Resource ?.... etc. subsampled how? inferred the tree, ancestral state reconstruction, visualized the tree...}
@@
 \subsection*{Quantification of mutational effects and sequence preferences from an H3N2 phylogeny}
+For a given mutation a1-r-a2, we calculated the mutational effect as:
+\begin{equation}
+\log_2 \frac{\pi_{r,a2}}{\pi_{r,a1}}
+\end{equation}
+where $\pi_{r,a1}$ and $\pi_{r,a2}$ are the preferences for amino-acid $a1$ or $a2$ at site $r$, and the preferences are re-scaled by the stringency parameter provided in Table~\ref{tab:phydms} and averaged across replicates.
+The WSN/1933 H1 HA amino-acid preferences were also re-scaled \comment{by stringency parameter in...} and averaged across replicates.
+The mutational effects for each five-year window were calculated for nodes with date $\geq$ YEAR1 and $<$ YEAR1 $+ 5$, and windows were incremented by one year.
+
+To calculate significance, we randomized the amino-acid preferences across sites and quantified the median difference in trunk and side branch mutational effects.
+We repeated this randomization for a total of 10,000 times and calculated how frequently the difference in median mutational effect between trunk and side branches exceeded the true difference to obtain a $P$-value.
+
+Sequence preference per site for every node on the phylogenetic tree shown in Figure~\ref{fig:H3N2_phylogeny} was calculated using Equation 1 for all sites, or the set of epitope and non-epitope sites defined in~\cite{wolf2006long}.
+
+\subsection*{Alignments and phylogenetic analysis of HA sequences}
+A description of the sequences and methods used to build the phylogenetic tree of HA subtypes shown in Figure~\ref{fig:distance_distribution}A is provided in~\cite{doud2017quantifying}.
+To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using mafft \comment{cite}.
+To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2017mapping} and based off the method presented in~\cite{doud2015site}.
+Briefly, for each site we calculated half of the sum of the absolute value of the difference between preferences for each amino acid.
+The RMSD$_{\text{corrected}}$ value sig
 
@@
 {\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.} 
+The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter.
+The sites are in H3 numbering.
+\comment{The top overlay bar indicates the relative solvent accessibility (RSA).}.
+The bottom overlay bar indicates the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
+The letters directly above each logo stack indicate the wildtype amino acid at that site.
 }
@@
 \caption{\label{suppfig:seqpref_zoomed}
-{\bf The per-site sequence preference at epitope and non-epitope sites shown in Figure~\ref{fig:sequence_preference}B, but on their own axes.} 
+{\bf The per-site sequence preference at epitope and non-epitope sites}
+The per-site sequence preferences shown in Figure~\ref{fig:sequence_preference}B, but on their own y-axes.
 }

### 2018-01-15 paper/manuscript.tex
some minor edits to Results

@@
 We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables the efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
-These mutant viruses derived all of their non-HA genes from the lab-adapted WSN/1933 strain.
+These mutant viruses derived all their non-HA genes from the lab-adapted WSN/1933 strain.
 Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
@@
 After generating the mutant virus libraries, we passaged them at low MOI in cell culture to create a genotype-phenotype link and select for functional HA variants (Figure~\ref{fig:dms_overview}A).
-All of the experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B). 
-We also passaged and deep sequenced library 3 in duplicate (denoted as library 3-1 and 3-2) to gauge to the amount of experimental noise occurring \textit{within} a single biological replicate.
+All the experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B). 
+We also passaged and deep sequenced library 3 in duplicate (denoted as library 3-1 and 3-2) to gauge the experimental noise occurring \textit{within} a single biological replicate.
 As a control to measure sequencing and mutational errors, we used the unmutated HA gene to generate and passage viruses carrying wildtype HA.
 
-Deep sequencing of the initial plasmid mutant libraries and the passaged mutant viruses allowed us to observe selection for functional HA mutants.
-As expected, selection against non-functional HA variants led to reduced mutation frequencies in the mutant viruses compared to the initial mutant plasmids (Figure~\ref{fig:dms_overview}C).
-Specifically, stop codons were purged to 20-45\% of their initial frequencies after correcting for error rates estimated by sequencing the wildtype controls.
+Deep sequencing of the initial plasmid mutant libraries and the passaged mutant viruses revealed selection for functional HA mutants.
+Specifically, stop codons were purged to 20-45\% of their initial frequencies after correcting for error rates estimated by sequencing the wildtype controls (Figure~\ref{fig:dms_overview}C).
 The incomplete purging of stop codons is likely because genetic complementation due to co-infection~\cite{marshall2013influenza} enabled the persistence of some virions with nonfunctional HAs. 
-We also observed selection against many nonsynonymous mutations, with their frequencies falling to 30-40\% of their initial values after error correction.
+We also observed selection against many nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), with their frequencies falling to 30-40\% of their initial values after error correction.
 
 We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
-We first used the deep sequencing data for each replicate to independently estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
+We first used the deep sequencing data for each replicate estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
 Because there are 567 residues in HA, there are $567 \times 20 = 11,340$ estimated amino-acid preferences.
@@
 The biological replicates are fairly well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
-Replicate 1 exhibited the lowest correlation with the other replicates, consistent with the observation that this replicate also showed the weakest selection against stop and nonsynonymous mutations, perhaps indicating more experimental noise.
-The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity after the reverse-genetics step contributes to most of the experimental noise.
+Replicate 1 exhibited the lowest correlation with the other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
+The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity after the reverse-genetics step contributes most of the experimental noise.
 
@@
 This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describe H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
-Table~\ref{tab:phydms} shows that the ExpCM using the replicate-average measurements greatly outperforms conventional substitution models, showing that our experiments authentically capture some of the constraints on HA evolution.
-Note that while the relative ratio of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models, this ratio is close to one for the ExpCM that accounts for the constraints measured in the experiments.
-ExpCM contain a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
-We fit a stringency parameter of 2.44 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments, but with greater stringency.
+Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
+This result indicates that our experiments authentically capture some of the constraints on HA evolution. 
+The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models.
+However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid constraints measured in our experiments (the $\omega$ for the ExpCM) is close to one, indicating that the deep mutational scanning captures much of the purifying selection on HA.
+ExpCM fit a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
+The stringency parameter for our HA ExpCM is 2.44 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments, but with greater stringency.
 Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
@@
 At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change,yang2015structure}.
-A positively charged amino acid at site 329 is important for cleavage activation of the HA0 precursor, and indeed this site exhibits a high preference for arginine~\cite{kido1992isolation, stech2005new}.
+A positively charged amino acid at site 329 is important for cleavage of the HA0 precursor into the mature form~\cite{kido1992isolation, stech2005new}, and this site strongly prefers arginine.
 However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine. 
 This codon is part of the signal peptide and is cleaved from the mature HA protein.
-One possible reason we do not see a strong preference for methionine at this site is due to alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
+One possible reason that our experiments do not show a strong preference for methionine at this site could be alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
 % No upstream or downstream "ATG"'s are present.
@@
 {\bf The site-specific amino-acid preferences of the Perth/2009 HA measured in our experiments.}
-The height of each letter is the preference for that amino acid, after taking the average of our experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter in Table~\ref{tab:phydms}.
+The height of each letter is the preference for that amino acid, after taking the average over experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter in Table~\ref{tab:phydms}.
 The sites are in H3 numbering.
-The top overlay bar indicates whether or not a site is part of the set of epitope residues delineated in \cite{wolf2006long}.
+The top overlay bar indicates whether or not a site is in the set of epitope residues delineated in \cite{wolf2006long}.
 The bottom overlay bar indicates the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
@@
 Our experiments measure which amino acids are tolerated at each HA site under selection for protein function.
-We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we define as the Shannon entropy of the re-scaled preferences.
-In prior mutational studies of the WSN/1933, the stalk domain of this H1 HA was found to be substantially less mutationally tolerant than its globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
+We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we quantify as the Shannon entropy of the re-scaled preferences.
+In prior mutational studies of the WSN/1933 H1 HA, the stalk domain was found to be substantially less mutationally tolerant than the globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
 
@@
 We also see high mutational tolerance in many of the known antigenic regions of H3 HA~\cite{wiley1981structural}.
-For instance, in recent H3N2 strains, antigenic region B is immunodominant and the location of most major antigenic drift mutations~\cite{chambers2015identification,koel2013substitutions,popova2012immunodominance}.
+For instance, in recent H3N2 strains, antigenic region B is immunodominant, and most recent major antigenic drift mutations have occurred in this region~\cite{chambers2015identification,koel2013substitutions,popova2012immunodominance}.
 We find that the most distal portion of the globular head near the 190-helix, which is part of antigenic region B, is highly tolerant of mutations (Figure~\ref{fig:mut_tolerance}).
@@
 Many residues inside HA's receptor binding pocket are known to be highly functionally constrained~\cite{wilson1981structure,martin1998studies}, and our data indicates that these sites are relatively mutationally intolerant in both H3 and H1 HAs.
-In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, as these sites are under strong immune pressure~\cite{wiley1981structural,koel2013substitutions}.
+In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, since mutations at these sites can have large effects on antigenicity~\cite{wiley1981structural,koel2013substitutions}.
 
@@
 Some influenza sequences are determined using egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{wu2017structural,mcwhite2016sequence,skowronski2016mutations}.
-These mutations will largely occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
+These mutations mostly occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
 We therefore re-calculated the statistics separating side branches to tip and internal nodes, and again found that the difference between the trunk and both sets of side branches was statistically significant (Figure~\ref{fig:trunkvssidebranch}B).
@@
 Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
-To explore this question, we scored the complete HA sequence of every node in the phylogenetic tree by quantifying its average per-site sequence preference for HA sequence $\mathbf{s}$, which is defined as
+To explore this question, we scored the complete HA sequence $\mathbf{s}$ of every node in the phylogenetic tree by quantifying its average per-site sequence preference as 
 \begin{equation}
@@
 \end{equation}
-where $\pi_{r, s_r}$ is the preference as measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
-Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to be more highly preferred than those of side branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side branch mutations.
+where $\pi_{r, s_r}$ is the preference measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid $s_r$ at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
+Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side branch mutations.
 However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
@@
 HA is under selection both to maintain its essential function in viral growth and to escape pre-existing immunity.
-Most of the immune selection is focused on a subset of so-called ``epitope sites'' that are targets of the immunodominant antibody response.
-Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites.
+Most immune selection is focused on a subset of so-called ``epitope sites'' that are targets of the immunodominant antibody response.
+Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites\comment{Maybe add some references to this point?}.
+There are various classifications of HA epitope sites; here we will use the classification of Wolf \textit{et al}~\cite{wolf2006long}, which defines \comment{X} epitope sites among the 567 residues in HA.
 In the timeframe from 1968 to 2012, the trunk of the tree in Figure~\ref{fig:H3N2_phylogeny} fixed 0.84 epitope mutations per site, but only 0.07 non-epitope mutations per site.
@@
 Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some specific epitope sites in H3 HA~\cite{wu2017diversity}.
-We hypothesize that the rapid evolution of epitope sites in HA leads to epistatic interactions that shift the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA.
+It is also known that positive selection for immune escape can increase the amount of epistasis among substitutions by favoring the fixation of combinations of individually deleterious immune-escape mutations and counterbalancing secondary mutations~\cite{gong2014epistatically}.
+We hypothesize that immune selection increases the accumulation of epistatic interactions involving substitutions at epitope sites.
+Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA.
 
@@
 {\bf Effects of mutations at epitope and non-epitope sites during HA evolution.}
-We partitioned HA into epitope and non-epitope sites using the definitions of Wolf \textit{et al}~\cite{wolf2006long}.
+We partitioned HA into epitope and non-epitope sites~\cite{wolf2006long}.
 (A) The log$_{2}$ mutational effects for side branch and trunk mutations at non-epitope (left) and epitope (right) sites.
@@
 The sites of large shifts are spread across HA, and do not obviously localize to one specific region of the protein's structure (Figure~\ref{fig:RMSD_heatmap}A).
-However, at the domain level, sites in the HA stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B). 
+However, at the domain level, sites in HA's stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B). 
 The HA stalk domain is also more conserved at the amino-acid level across HA subtypes~\cite{nobusawa1991comparison,hai2012influenza,mallajosyula2014influenza}, suggesting that conservation of amino-acid sequence tends to be correlated with conservation of amino-acid preferences.
@@
 
-Despite their large sequence divergence, H1 and H3 adopt nearly identical protein folds~\cite{ha2002h5,russell2004h1}.
-Previous work has found that although the subdomain structures are highly similar, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among different HA subtypes~\cite{ha2002h5,russell2004h1}.
-This previous work has defined clades of closely related HA subtypes. 
+Despite their high sequence divergence, H1 and H3 adopt very similar protein folds~\cite{ha2002h5,russell2004h1}.
+However, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among different HA subtypes~\cite{ha2002h5,russell2004h1}.
+Previous work has defined clades of structurally related HA subtypes~\cite{ha2002h5,russell2004h1}. 
 One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}).
@@
 The clade containing H1 has an upward shift of the globular head relative to the clade containing H3.
-This shift in the head subdomain has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
+This structural shift has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
 Specifically, the clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
@@
 {\bf Sites with strongly shifted amino-acid preferences between H3 and H1 HAs.}
-(A) The shift in amino-acid preferences at each site as calculated in Figure~\ref{fig:distance_distribution}C between the H3 and H1 HA is mapped onto the structure of the H3 HA. 
-(B) Amino-acid preferences of sites in the stalk domain are significantly less shifted than those in the head domain.
-Sites absolutely conserved in all 18 HA subtypes are significantly less shifted than other sites.
-Sites with one amino-acid identity in the clade containing H1, H2, H5, and H6 and another identity in the clade containing H3, H4, and H14 are significantly more shifted than other sites.
+(A) The shift in amino-acid preferences between the H3 and H1 HA at each site as calculated in Figure~\ref{fig:distance_distribution}C is mapped onto the structure of the H3 HA. 
+(B) Amino-acid preferences of sites in the stalk domain are less shifted than those in the head domain.
+Sites absolutely conserved in all 18 HA subtypes are less shifted than other sites.
+Sites with one amino-acid identity in the clade containing H1, H2, H5, and H6 and another identity in the clade containing H3, H4, and H14 are more shifted than other sites.
 (C) Sites 107 and 75(HA2) help determine the different orientation of the globular head domain in H1 versus H3 HAs.
@@
 The spheres are colored as in panel (A), and the experimentally measured amino-acid preferences in the WSN/1933 H1 and Perth/2009 H3 HAs are shown.
+\comment{Could B be a thinner panel at the bottom, and then C be to the right of A, possibly with H1 and H3 vertically oriented?
+It may be OK to shrink the A panel a bit if that save space in this new layout.}
 }
@@
 \caption{\label{suppfig:SangerSeq}
-{\bf Sanger sequencing of 31 clones from the mutant plasmid libraries.} 
+{\bf Sanger sequencing of 31 randomly chosen clones from the mutant plasmid libraries.} 
 (A) There were an average of $\sim$1.4 codon mutations per clone across the three plasmid mutant libraries.
@@
 (E) We calculated the pairwise distances between mutations for clones carrying more than one mutation.
-The cumulative distribution of the distances is shown in the red line.  
+The cumulative distribution of these distances is shown in the red line. 
+The blue line indicates the expected distribution if mutations in multiply mutated genes are randomly disbursed along the sequence. 
 }
@@
 {\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.} 
-The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter.
+The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter \comment{somehow indicate where this stringency parameter}.
 The sites are in H3 numbering.
-\comment{The top overlay bar indicates the relative solvent accessibility (RSA).}.
+\comment{I would suggest just getting rid of RSA. You might be able to replace the next two sentences by just saying: the overlays show the same information as in Figure X}.
 The bottom overlay bar indicates the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).

### 2018-01-15 paper/manuscript.tex
some progress on Intro

@@
 
-% If your first paragraph (i.e.
-with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation.
-If this is the case, add \parshape=0 to the end of the list environment.
-\dropcap{S}easonal H3N2 influenza virus evolves rapidly, readily accumulating mutations in the hemagglutinin (HA) surface protein that contribute to antigenic drift~\cite{smith2004mapping,bhatt2011genomic}.
-The evolution of human H3N2 influenza virus is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2015global}, producing a distinct spindly shape in its phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
-A key challenge in the study of H3N2 evolution lies in being able to predict the dominant strain in the upcoming influenza season, as this information can guide selection of a vaccine strain.
-Fundamentally, in order to make these predictions we must be able to distinguish between strains that will persist as the trunk and strains that will die out.
-
-%To aid in the timeliness and throughput of making such decisions, previous work have endeavored to accurately predict the evolutionary outcomes of viral strains based on the antigenicity and/or tree structure of viral clades~\cite{steinbruck2014computational,neher2014predicting}.
-%However, the accuracy of these predictions are subject to uneven geographical and temporal sampling, and for~\cite{neher2014predicting}, the lack of information about a mutation's effect on antigenicity.
-One approach is to model the antigenic properties of circulating viral strains using sequence information or phylogenetic relationships and experimental measurements~\cite{sun2013using,harvey2016identification,neher2016prediction}.
-While mutations that contribute to antigenic evolution of the virus certainly determine strain success, the non-antigenic effects of mutations play an important role in determining the persistence of viral strains~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
-Due to the high mutation rate of RNA viruses~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and because the influenza virus does not appreciably recombine within a given segment, deleterious mutations become linked to beneficial ones.
-Strain fitness is thereby dependent on both the antigenic and non-antigenic effects of mutations, as modeled in~\cite{luksza2014predictive}.
-However, we lack a broad understanding of the phenotypic effects of mutations to HA, particularly on HA function and viral growth.
-We therefore need a greater understanding of the effects of mutations to HA to comprehend the forces and constraints shaping influenza virus evolution in nature.
+\dropcap{S}easonal H3N2 influenza virus evolves rapidly, fixing 3 to 4 amino-acid mutations per year in its hemagglutinin (HA) surface protein.
+ Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping,bhatt2011genomic}.
+This evolution is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects,bedford2015global}, producing a spindly phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
+Several lines of evidence indicate that the trunk lineage outcompetes other strains at least in part because it has higher fitness~\cite{strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects}.
+A key challenge in the study of H3N2 evolution is to forecast which strain will dominate the upcoming influenza season, as this information can guide selection of a vaccine strain~\cite{morris2017predictive}.
+In order to make these forecasts, it is important to identify the features that enable the trunk strain to persist as other strains die out.
 
@@
 \end{figure}
+
+%To aid in the timeliness and throughput of making such decisions, previous work have endeavored to accurately predict the evolutionary outcomes of viral strains based on the antigenicity and/or tree structure of viral clades~\cite{steinbruck2014computational,neher2014predicting}.
+%However, the accuracy of these predictions are subject to uneven geographical and temporal sampling, and for~\cite{neher2014predicting}, the lack of information about a mutation's effect on antigenicity.
+
+Two main features distinguish the evolutionarily successful trunk from its competitors: greater antigenic change, and fewer mutations that disrupt replication and transmission.
+In principle, experiments could be informative for identifying how mutations affect these features.
+Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenic properties of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction}.
+While the antigenic properties of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
+Specifically, due to influenza virus's high mutation rate~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and lack of intra-segment recombination~\cite{boni2008homologous}, deleterious mutations become linked to beneficial ones.
+Strain fitness is thereby dependent on both the antigenic and non-antigenic effects of mutations~\cite{luksza2014predictive}.
+However, while extensive work has gone into characterizing which mutations have the largest antigenic effects on HA~\comment{epitope site papers, Koel paper, Hensley papers, Kawaoka paper}, we lack any large-scale quantitative characterization of how mutations to H3N2 HA affect the non-antigenic properties crucial to viral fitness.
+
+\comment{This paragraph would then be obsolete: One approach is to model the antigenic properties of circulating viral strains using sequence information or phylogenetic relationships and experimental measurements~\cite{sun2013using,harvey2016identification,neher2016prediction}.
+While mutations that contribute to antigenic evolution of the virus certainly determine strain success, the non-antigenic effects of mutations play an important role in determining the persistence of viral strains~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
+Due to the high mutation rate of RNA viruses~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and because the influenza virus does not appreciably recombine within a given segment, deleterious mutations become linked to beneficial ones.
+Strain fitness is thereby dependent on both the antigenic and non-antigenic effects of mutations, as modeled in~\cite{luksza2014predictive}.
+However, we lack a broad understanding of the phenotypic effects of mutations to HA, particularly on HA function and viral growth.
+We therefore need a greater understanding of the effects of mutations to HA to comprehend the forces and constraints shaping influenza virus evolution in nature.}
 
@@
 Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites\comment{Maybe add some references to this point?}.
-There are various classifications of HA epitope sites; here we will use the classification of Wolf \textit{et al}~\cite{wolf2006long}, which defines \comment{X} epitope sites among the 567 residues in HA.
+There are various classifications of HA epitope sites; here we will use the classification of Wolf \textit{et al}~\cite{wolf2006long}, which defines 129 epitope sites among the 567 residues in HA.
 In the timeframe from 1968 to 2012, the trunk of the tree in Figure~\ref{fig:H3N2_phylogeny} fixed 0.84 epitope mutations per site, but only 0.07 non-epitope mutations per site.

### 2018-01-15 paper/manuscript.tex
revised Intro

@@
 
-Two main features distinguish the evolutionarily successful trunk from its competitors: greater antigenic change, and fewer mutations that disrupt replication and transmission.
+Two main features distinguish the evolutionarily successful trunk from its competitors: greater antigenic change, and efficient viral growth and transmission.
 In principle, experiments could be informative for identifying how mutations affect these features.
-Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenic properties of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction}.
-While the antigenic properties of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
+Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction}.
+While the antigenic features of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
 Specifically, due to influenza virus's high mutation rate~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and lack of intra-segment recombination~\cite{boni2008homologous}, deleterious mutations become linked to beneficial ones.
-Strain fitness is thereby dependent on both the antigenic and non-antigenic effects of mutations~\cite{luksza2014predictive}.
-However, while extensive work has gone into characterizing which mutations have the largest antigenic effects on HA~\comment{epitope site papers, Koel paper, Hensley papers, Kawaoka paper}, we lack any large-scale quantitative characterization of how mutations to H3N2 HA affect the non-antigenic properties crucial to viral fitness.
+The resulting accumulation of deleterious mutations can affect non-antigenic properties that are central to viral fitness~\cite{luksza2014predictive}.
+However, while extensive work has gone into characterizing how mutations affect HA antigenicity~\comment{epitope site papers, Koel paper, Hensley papers, Kawaoka paper}, we lack any large-scale quantitative characterization of how mutations to H3N2 HA affect non-antigenic properties crucial to viral fitness.
 
-\comment{This paragraph would then be obsolete: One approach is to model the antigenic properties of circulating viral strains using sequence information or phylogenetic relationships and experimental measurements~\cite{sun2013using,harvey2016identification,neher2016prediction}.
-While mutations that contribute to antigenic evolution of the virus certainly determine strain success, the non-antigenic effects of mutations play an important role in determining the persistence of viral strains~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
-Due to the high mutation rate of RNA viruses~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and because the influenza virus does not appreciably recombine within a given segment, deleterious mutations become linked to beneficial ones.
-Strain fitness is thereby dependent on both the antigenic and non-antigenic effects of mutations, as modeled in~\cite{luksza2014predictive}.
-However, we lack a broad understanding of the phenotypic effects of mutations to HA, particularly on HA function and viral growth.
-We therefore need a greater understanding of the effects of mutations to HA to comprehend the forces and constraints shaping influenza virus evolution in nature.}
-
-We can measure such mutational effects in the lab using deep mutational scanning approaches~\cite{fowler2010high}.
-Previously, we experimentally measured the effect of all possible single amino-acid mutations to an H1 HA from the A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,doud2016accurate}.
-However, WSN/1933 is a highly lab-adapted strain, and these measurements may not be particularly relevant for studying the mutational processes of more contemporaneous strains circulating in the human population.
-Here we measure the effects of mutations to the HA from the A/Perth/16/2009 (H3N2) strain on viral growth in cell culture.
-% Our measurements generally agree with what is already known about HA structure and function.
-We apply these experimental measurements to examine differences in the trunk and side branch lineages of human H3N2 influenza virus.
-Trunk mutations were found to be more favorable than side branch mutations, both at epitope and non-epitope sites.
-We also find evidence that epistasis in HA has primarily been driven by antigenic drift.
-The measurements also enabled a comparison of the shifts in amino-acid preferences between two diverged HAs.
-The extent of shifts in amino-acid preferences increase as homologs diverge.
+It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of large numbers of mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high}.
+However, the only HA for which so such measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
+Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain, A/Perth/16/2009.
+We then examine whether these measurements are useful for distinguishing between H3N2 trunk and side branch lineages, and find that mutations on the trunk are measured to be more favorable than those on the side branches, both at epitope and non-epitope sites.
+However, the experiments are useful for assessing the effects of mutations only in closely related viral strains, since we show that mutations often have disparate effects on H3 and H1 HAs.
 Our work highlights the potential for using high-throughput experimental measurements of mutational effects to inform evolutionary forecasting of human seasonal influenza virus.
@@
 We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables the efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
-These mutant viruses derived all their non-HA genes from the lab-adapted WSN/1933 strain.
+These mutant viruses derived all their non-HA genes from the lab-adapted A/WSN/1933 strain.
 Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.

### 2018-01-15 paper/manuscript.tex
rough outline of Discussion

@@
 \end{figure}
-
-%To aid in the timeliness and throughput of making such decisions, previous work have endeavored to accurately predict the evolutionary outcomes of viral strains based on the antigenicity and/or tree structure of viral clades~\cite{steinbruck2014computational,neher2014predicting}.
-%However, the accuracy of these predictions are subject to uneven geographical and temporal sampling, and for~\cite{neher2014predicting}, the lack of information about a mutation's effect on antigenicity.
 
@@
 \label{sec:discussion}
-We have measured the effect of all possible single amino-acid mutations to Perth/2009 H3 on viral growth in cell culture.
+We have measured the effect of all possible single amino-acid mutations to Perth/2009 H3 on viral growth in cell culture, and shown that these measurements are informative about the evolutionary fate of human H3N2 viral strains in nature.
+Specifically, strains the trunk of the phylogenetic tree tend to have mutations that that our experiments score as more favorable than mutations on the side branches.
+Distinguishing between the trunk and side branches is the equivalent of identifying which strains succeed in nature, and so the ability of our experiments to do this show that they are useful.
 
-\begin{itemize}
-\item{These experimental measurements have utility in revealing differences between successful and unsuccessful lineages over 50 y of human influenza virus evolution.}
-\item{We have only measured the effect of mutations to HA on viral growth, but there are certainly other determinants of strain success}
-\item{Antigenic characteristics play a large role in determining strain success. We therefore need to know how mutations affect antigenicity.}
-\item{non-HA segments are important too}
-\end{itemize}
+It may seem surprising that simple measurements of viral growth in cell culture are informative about fitness in nature.
+There are many examples of mutations that have different effects in the lab and in nature, and our experiments definitely don't measure all the relevant factors.
+In addition, we only look at mutations to HA.
+Nonetheless, prior to our work there was no comprehensive estimates the functional effects of mutations on viral fitness outside of antigenicity.
+Therefore, all prior quantitative modeling either neglected non-antigenic effects or simply assumed that all mutations outside epitope sites had an equivalent deleterious effect on viral fitness.
+While our measurements are certainly imperfect, they are better than this.
+... also suggests that measurements in more realistic experimental models should be even more useful.
+
+Another caveat is that we are making measurements of the effects of mutations in a single HA variant, Perth/2009.
+We are then extrapolating those measurements across the full spectrum of seasonal human H3N2.
+These extrapolations will only be valid to the extent that the effects of mutations remain conserved during HA's evolution.
+It is known from extensive work on protein evolution that mutational effects can shift due to epistasis, and the extent to which this occurs is an important question in its own right.
+The success of our measurements across all H3N2 show that these shifts aren't too huge for relatively similar strains.
+But we also compared our measurements for H3 to prior measurements on H1 and found that there were large shifts --- much greater than those observed in prior protein-wide comparisons of more closely related homologs.
+The effective use of experiments for studying evolution in nature will require further analysis of how mutational effects shift in order to figure out how widely the results of any given experiment can be applied.
+
+Other things are important too.
+We did not look at antigenicity (whereas prior studies have only looked at that).
+In the future, our experimental approach could be used to look at antigenicity too.
 

### 2018-01-15 paper/manuscript.tex
updated Significance Statement

@@
 \significancestatement{ 
-Mutations drive the rapid evolution of seasonal influenza virus such that new strains emerge and co-circulate until one strain overtakes the population. 
-A key challenge in the study of seasonal influenza virus is to be able to differentiate between strains that are likely to persist and those that are likely to die out. 
-We therefore need a greater understanding of how mutations drive and constrain viral evolution in order to help identify persistent and nonpersistent strains. 
-We experimentally measured the effect of every possible amino-acid mutation to the surface protein hemagglutinin on viral growth.
-Using these measurements, we revealed differences between successful and unsuccessful viral lineages and examined how generalizable the measurements were for studying the evolutionary fate of viral lineages.
-\comment{Our work has potential for use in forecasting}
+A key goal in the study of influenza virus evolution is to forecast which viral strains will persist and which ones will die out. 
+Here we experimentally measure the effects of all amino-acid mutations to the hemagglutinin protein from a human H3N2 influenza strain on viral growth in cell culture.
+We show that these measurements have utility for distinguishing among viral strains that do and do not succeed in nature.
+Overall, our work suggest that new high-throughput experimental approaches may be useful for evolutionary forecasting.
 }

### 2018-01-15 paper/manuscript.tex
revisions to Abstract

@@
 % no more than 250 words. (199 now)
-The evolution of seasonal H3N2 influenza virus is mediated by the rapid accumulation of mutations. 
-A better understanding of how mutations affect viral growth is critical for studying the pressures and constraints underlying the evolution of the influenza virus in nature. 
-Here we experimentally measure the effect of every possible single amino-acid mutation to an H3 influenza hemagglutinin on viral growth in cell culture. 
-We find the stalk domain to be fairly mutationally tolerant, and there is not a large disparity in tolerance between the head and stalk domains.
-Furthermore, our measurements reveal mutations in successful seasonal influenza viral lineages to be significantly more preferred than mutations in lineages that have died out, and this was true at both epitope and non-epitope sites.
-Epistasis in hemagglutinin is also seemingly driven by antigenic drift.
-The preferences measured in a distantly related protein homolog do not reveal such differences in viral lineages.
-A comparison of the preferences between the two hemagglutinin variants show substantial shifts in the effects of mutations.
-Overall, our work highlights how experimental measurements of mutational effects can be leveraged to gain insight into the evolutionary fates of viral strains and potentially be used to predict viral evolution.
+Human influenza virus rapidly accumulates mutations in its major surface protein hemagglutinin (HA).
+Forecasting the evolutionary success of influenza virus strains requires understanding how these mutations affect the protein's functionality and antigenicity.
+Here we experimentally measure the effects on viral growth in cell culture of all single amino-acid mutations to the HA from a recent human H3N2 influenza virus strain. 
+We show that these measurements are informative for understanding viral evolution in nature, since mutations in evolutionarily successful H3N2 viral lineages are measured to generally be more favorable than mutations in lineages that die out.
+We also compare our measurements to prior data for an H1 HA, and find that the amino-acid preferences of many sites have substantially shifted between H1 and H3 HAs.
+For instance, there is much less of a disparity in mutational tolerance between the head and stalk domains for the H3 HA.
+Overall, our work suggests that experimental measurements of mutational effects might be leveraged to help forecast the evolutionary fates of viral lineages in nature.
 \end{abstract}

### 2018-01-15 paper/manuscript.tex
edits to Methods about TMPRSS2

@@
 \subsection*{Creation of MDCK-SIAT1-TMPRSS2 cell line}
-The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter and attached to mCherry through an IRES. 
-We generated lentiviruses..etc etc transduced MDCK-SIAT1 cells, sorted by fluorescence activated cell sorting
-\comment{Need to look at Katie's notebooks for this...}
+\comment{A sentence or two about the rationale: Normally trypsin must be added. In order to obviate the need for trypsin.}
+The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry.
+We used the lentiviral vector to transduce MDCK-SIAT1 cells, and sorted an intermediate mCherry population by flow cytometry.
+We refer to the resulting bulk population of sorted cells as MDCK-SIAT1-TMPRSS2 cells.
+There is no selectable marker for the TMPRSS2; however, we generally maintain the cells at low passage number, and have seen no indication that lose their ability to support the replication of viruses with H3 HAs in the absence of exogenous trypsin.
 

### 2018-01-16 paper/manuscript.tex
Add new WSN logoplot to supplementary figures
Full draft of Discussion

@@
 \begin{abstract}
-% no more than 250 words. (199 now)
+% no more than 250 words. (170 now)
 Human influenza virus rapidly accumulates mutations in its major surface protein hemagglutinin (HA).
@@
 \label{sec:discussion}
-We have measured the effect of all possible single amino-acid mutations to Perth/2009 H3 on viral growth in cell culture, and shown that these measurements are informative about the evolutionary fate of human H3N2 viral strains in nature.
-Specifically, strains the trunk of the phylogenetic tree tend to have mutations that that our experiments score as more favorable than mutations on the side branches.
-Distinguishing between the trunk and side branches is the equivalent of identifying which strains succeed in nature, and so the ability of our experiments to do this show that they are useful.
+We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 on viral growth in cell culture, and demonstrated that these measurements can be informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
+Specifically, strains on the trunk of the phylogenetic tree tend to have mutations that are beneficial for viral growth in our experiments when compared to side-branch mutations.
+Differentiating between trunk and side branch lineages can enable forecasting of influenza virus evolution by identifying which strains are likely to succeed in nature.
+The fact that our experiments are able to reveal differences between trunk and side branch lineages highlight the potential for incorporating experimental measurements of the non-antigenic effects of mutations for evolutionary forecasting.
 
-It may seem surprising that simple measurements of viral growth in cell culture are informative about fitness in nature.
-There are many examples of mutations that have different effects in the lab and in nature, and our experiments definitely don't measure all the relevant factors.
-In addition, we only look at mutations to HA.
-Nonetheless, prior to our work there was no comprehensive estimates the functional effects of mutations on viral fitness outside of antigenicity.
-Therefore, all prior quantitative modeling either neglected non-antigenic effects or simply assumed that all mutations outside epitope sites had an equivalent deleterious effect on viral fitness.
-While our measurements are certainly imperfect, they are better than this.
-... also suggests that measurements in more realistic experimental models should be even more useful.
+Surprisingly, our simple experimental measurements of viral growth in cell culture helped to elucidate differences in fitness of viral strains in nature.
+Mutations may have effects in the laboratory that are distinct from their effects in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}, and our experiments certainly don't capture all of the relevant selective pressures on HA that exist in nature.
+Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
+Yet, prior to our work presented here, there were no comprehensive studies of the functional effects of mutations to HA on viral fitness outside of antigenicity.
+Previous quantitative modeling work either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or made the simplifying assumption that all mutations outside epitope sites had equivalent deleterious effects on viral fitness~\cite{luksza2014predictive}.
+While our measurements may be imperfect, we have demonstrated that these measurements are still of utility in distinguishing the selective advantage of successful influenza viral lineages over $\sim$50 years of evolution.
+This also suggests that measurements in experimental models that more faithfully recapitulate the selective forces in nature should be of even greater utility for differentiating evolutionarily successful viral lineages.
 
-Another caveat is that we are making measurements of the effects of mutations in a single HA variant, Perth/2009.
-We are then extrapolating those measurements across the full spectrum of seasonal human H3N2.
-These extrapolations will only be valid to the extent that the effects of mutations remain conserved during HA's evolution.
-It is known from extensive work on protein evolution that mutational effects can shift due to epistasis, and the extent to which this occurs is an important question in its own right.
-The success of our measurements across all H3N2 show that these shifts aren't too huge for relatively similar strains.
-But we also compared our measurements for H3 to prior measurements on H1 and found that there were large shifts --- much greater than those observed in prior protein-wide comparisons of more closely related homologs.
-The effective use of experiments for studying evolution in nature will require further analysis of how mutational effects shift in order to figure out how widely the results of any given experiment can be applied.
+In addition, we are measuring the effects of mutations in a single HA variant from the Perth/2009 strain and extrapolating these measurements to other seasonal human H3N2 strains across a 50-year timespan.
+However, these extrapolations will only be valid to the extent that the effects of mutations remain conserved during HA's evolution.
+Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
+% and the extent to which this occurs is an important question in its own right.
+The observation that our measurements score trunk mutations more favorably than side branch mutations across most of H3N2 evolution suggests that for relatively closely related strains, the mutational effects are not shifted to a large extent.
+On the other hand, when we compared our measurements for H3 to prior measurements on H1, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs.
+Further analysis of how mutational effects shift as protein variants diverge can provide insight into how broadly the results of a given experiment can be generalized across variants.
 
-Other things are important too.
-We did not look at antigenicity (whereas prior studies have only looked at that).
-In the future, our experimental approach could be used to look at antigenicity too.
+Importantly, the antigenic properties of HA are integral in determining strain success, but we did not study the effect of mutations on antigenicity.
+However, our high-throughput experimental approach can be harnessed to profile mutations that confer escape from antibodies, as has been performed for the WSN/1933 H1 HA~\cite{doud2017complete,doud2017quantifying} and HIV Env~\cite{dingens2017comprehensive}.
+In future, these antigenic studies can be combined with functional measurements to inform forecasting of the evolution of seasonal influenza virus.
 
@@
 The sites are in H3 numbering.
-\comment{I would suggest just getting rid of RSA. You might be able to replace the next two sentences by just saying: the overlays show the same information as in Figure X}.
-The bottom overlay bar indicates the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
-The letters directly above each logo stack indicate the wildtype amino acid at that site.
+The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
 }

### 2018-01-16 paper/manuscript.tex
Re-edit Introduction

@@
 
-Two main features distinguish the evolutionarily successful trunk from its competitors: greater antigenic change, and efficient viral growth and transmission.
+Two main characteristics distinguish the evolutionarily successful trunk from its competitors: greater antigenic change, and efficient viral growth and transmission.
 In principle, experiments could be informative for identifying how mutations affect these features.
@@
 The resulting accumulation of deleterious mutations can affect non-antigenic properties that are central to viral fitness~\cite{luksza2014predictive}.
-However, while extensive work has gone into characterizing how mutations affect HA antigenicity~\comment{epitope site papers, Koel paper, Hensley papers, Kawaoka paper}, we lack any large-scale quantitative characterization of how mutations to H3N2 HA affect non-antigenic properties crucial to viral fitness.
+However, while extensive work has gone into characterizing how mutations affect HA antigenicity~\cite{laver1979antigenic, webster1980determination,wiley1981structural,koel2013substitutions,chambers2015identification,li2016selection}, we lack any large-scale quantitative characterization of how mutations to H3N2 HA affect non-antigenic properties crucial to viral fitness.
 
@@
 For instance, in the H3 HA, the short helix A in the stalk is very mutationally tolerant.
-Interestingly, more studies have reported selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2017quantifying} HAs.
+Interestingly, more studies have reported selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2017quantifying,anderson2017natural} HAs.
 

### 2018-01-17 paper/manuscript.tex
Rearrange and resize RMSD_heatmap figure

@@
 
-\begin{SCfigure*}[\sidecaptionrelwidth][t]
+%\begin{SCfigure*}[\sidecaptionrelwidth][t]
+\begin{figure}
 \centering
-\includegraphics[width=11.4cm]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
+%\includegraphics[width=11.4cm]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
+\includegraphics[width=\columnwidth]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
 \caption{\label{fig:RMSD_heatmap}
@@
 The spheres are colored as in panel (A), and the experimentally measured amino-acid preferences in the WSN/1933 H1 and Perth/2009 H3 HAs are shown.
-\comment{Could B be a thinner panel at the bottom, and then C be to the right of A, possibly with H1 and H3 vertically oriented?
-It may be OK to shrink the A panel a bit if that save space in this new layout.}
 }
-\end{SCfigure*}
+%\end{SCfigure*}
+\end{figure}
 

### 2018-01-17 paper/manuscript.tex
Finalize edits to Abstract and Intro

@@
 It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of large numbers of mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high}.
-However, the only HA for which so such measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
+However, the only HA for which such measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
 Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain, A/Perth/16/2009.

### 2018-01-17 paper/manuscript.tex
Add more to Methods

@@
 Most immune selection is focused on a subset of so-called ``epitope sites'' that are targets of the immunodominant antibody response.
-Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites\comment{Maybe add some references to this point?}.
+Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites~\cite{fitch1991positive,wolf2006long,bhatt2011genomic}.
 There are various classifications of HA epitope sites; here we will use the classification of Wolf \textit{et al}~\cite{wolf2006long}, which defines 129 epitope sites among the 567 residues in HA.
@@
 
-\matmethods{Please describe your materials and methods here.
-This can be more than one paragraph, and may contain subsections and equations as required.
-Authors should include a statement in the methods section describing how readers will be able to access the data in the paper.
+\matmethods{Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610. 
+Computer code used to analyze the data and produce the results in the paper are in...
 
@@
 \subsection*{Creation of MDCK-SIAT1-TMPRSS2 cell line}
-\comment{A sentence or two about the rationale: Normally trypsin must be added. In order to obviate the need for trypsin.}
+When growing influenza virus, TPCK-trypsin must normally be added in order to cleave HA into its mature form.
+In order to obviate the need for trypsin, we engineered an MDCK-SIAT1 cell line to constitutively express the TMPRSS2 protease.
+This protease cleaves and activates HA and is found endogenously in the human airway~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
 The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry.
@@
 Bulk RNA from the viral sample was extracted using the QIAamp Viral RNA Mini Kit (QIAGEN) according to manufacturer's instructions.
-The Perth/2009 HA and NA genes were then reverse transcribed using the SuperScript III First-Strand Synthesis SuperMix Kit (Invitrogen), PCR amplified, and cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2 \comment{cite?} plasmid backbones by In-Fusion.
+The Perth/2009 HA and NA genes were then reverse transcribed using the SuperScript III First-Strand Synthesis SuperMix Kit (Invitrogen), PCR amplified, and cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} plasmid backbones by In-Fusion.
 
@@
 Two mutations, G78D and T212I, arose after serial passaging. 
-The Perth/2009 HA carrying these two mutations were cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2 \comment{cite?} plasmid backbones.
+The Perth/2009 HA carrying these two mutations were cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} plasmid backbones.
 
@@
 We generated the HA PCR amplicons for each of the three plasmid libraries, the three virus libraries, one wildtype plasmid, and one wildtype virus samples using KOD Hot Start Master Mix (EMD Millipore;71842) according to the PCR reaction mixture and cycling conditions described in~\cite{bloom2014experimentally} and the P09-HA-For and P09-HA-Rev primers.
-\comment{used the same conditions as Mike - cite his Viruses paper to be succinct?}.
+We prepared the sequencing libraries using a barcoded-subamplicon strategy~\cite{wu2014high} to increase the accuracy from deep sequencing. 
+The approach we used is described in~\cite{doud2016accurate} (also see \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
 We performed sequencing on a lane of a flow cell of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.
@@
 \subsection*{Analysis of deep sequencing data}
-We used dms\_tools2...
+We used the dms\_tools2 software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}) to analyze the deep sequencing data.
+The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
 
@@
 where $\pi_{r,a1}$ and $\pi_{r,a2}$ are the preferences for amino-acid $a1$ or $a2$ at site $r$, and the preferences are re-scaled by the stringency parameter provided in Table~\ref{tab:phydms} and averaged across replicates.
-The WSN/1933 H1 HA amino-acid preferences were also re-scaled \comment{by stringency parameter in...} and averaged across replicates.
-The mutational effects for each five-year window were calculated for nodes with date $\geq$ YEAR1 and $<$ YEAR1 $+ 5$, and windows were incremented by one year.
+The WSN/1933 H1 HA amino-acid preferences were also re-scaled by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}) and averaged across replicates.
+The mutational effects were also calculated for five-year windows from 1968 to 2012, and windows were incremented by one year.
 
-To calculate significance, we randomized the amino-acid preferences across sites and quantified the median difference in trunk and side branch mutational effects.
+To calculate significance, we randomized the amino-acid preferences across all sites, epitope sites, or non-epitope sites, and quantified the median difference in trunk and side branch mutational effects.
 We repeated this randomization for a total of 10,000 times and calculated how frequently the difference in median mutational effect between trunk and side branches exceeded the true difference to obtain a $P$-value.
@@
 \subsection*{Alignments and phylogenetic analysis of HA sequences}
+To build the ExpCM using our measurements, we first downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
+The subsampled sequences were then aligned using MAFFT~\cite{katoh2013mafft} and used to build a phylogenetic tree using RAxML~\cite{stamatakis2006raxml}.
+We then used the phydms program~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms})
+
 A description of the sequences and methods used to build the phylogenetic tree of HA subtypes shown in Figure~\ref{fig:distance_distribution}A is provided in~\cite{doud2017quantifying}.
-To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using mafft \comment{cite}.
+
+\subsection*{Analysis of mutational shifts}
+To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using MAFFT~\cite{katoh2013mafft}.
 To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2017mapping} and based off the method presented in~\cite{doud2015site}.
 Briefly, for each site we calculated half of the sum of the absolute value of the difference between preferences for each amino acid.
-The RMSD$_{\text{corrected}}$ value sig
-
-\subsection*{Analysis of mutational shifts}
-
-\subsection*{Data availability and source code}
-Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
-Computer code used to analyze the data and produce the results in the paper are in...
+The RMSD$_{\text{corrected}}$ value signifies the shift in preference.
+\comment{add more here about absolutely conserved, clade-specific sites, etc}
 }
@@
 
-\acknow{We thank Sarah Hilton, Hugh Haddox, Sidney Bell for helpful discussions about data analysis...the Fred Hutch Genomics Core...
-Funding...}
+\acknow{We thank Sarah Hilton, Hugh Haddox, and Sidney Bell for helpful discussions about data analysis.
+We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
+This work was supported by...}
 

### 2018-01-17 paper/manuscript.tex
Complete first draft of methods

@@
 The HA and NA genes were reverse transcribed, PCR amplified, and sequenced by Sanger sequencing.
-Two mutations, G78D and T212I, arose after serial passaging. 
+Two mutations, G78D and T212I, arose to fixation after serial passaging. 
 The Perth/2009 HA carrying these two mutations were cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} plasmid backbones.
+
+To validate that the variant Perth/2009 HA could support viral growth, we generated in duplicate by reverse genetics viruses with the original Perth/2009 HA or the Perth/2009 HA-G78D-T212I variant on a background of Perth/2009 NA $+$ WSN internal genes.
+The transfection supernatant was collected 72 hours post-transfection and passaged onto a monolayer of MDCK-SIAT1-TMPRSS2 cells at an MOI = 0.01.
+The viral supernatant was collected at 44 hours post-infection.
+Both the transfection and passage supernatants were titered by TCID$_{50}$ in MDCK-SIAT1-TMPRSS2 cells.
 
@@
 We pooled over 6 million transformants for each replicate, cultured in liquid LB + ampicillin at 37$^{\circ}$C for 3 h with shaking, and maxiprepped.
+We randomly chose 31 clones to Sanger sequence, and summary statistics from sequencing are shown in Figure~\ref{suppfig:SangerSeq}.
 
@@
 The approach we used is described in~\cite{doud2016accurate} (also see \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
+The primers we used to generate the subamplicons are in Dataset~\ref{suppdata:bcsubamp_primers}.
 We performed sequencing on a lane of a flow cell of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.
@@
 The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
+The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Datasets \comment{suppfile}.
+
+\subsection*{Alignments and phylogenetic analysis of HA sequences}
+To build the ExpCM using our measurements, we first downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
+The subsampled sequences were then aligned using MAFFT~\cite{katoh2013mafft} and used to build a phylogenetic tree using RAxML~\cite{stamatakis2006raxml}.
+We then used the phydms program~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms})
+
+A description of the sequences and methods used to build the phylogenetic tree of HA subtypes shown in Figure~\ref{fig:distance_distribution}A is provided in~\cite{doud2017quantifying}.
 
@@
 
-\subsection*{Alignments and phylogenetic analysis of HA sequences}
-To build the ExpCM using our measurements, we first downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
-The subsampled sequences were then aligned using MAFFT~\cite{katoh2013mafft} and used to build a phylogenetic tree using RAxML~\cite{stamatakis2006raxml}.
-We then used the phydms program~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms})
-
-A description of the sequences and methods used to build the phylogenetic tree of HA subtypes shown in Figure~\ref{fig:distance_distribution}A is provided in~\cite{doud2017quantifying}.
-
 \subsection*{Analysis of mutational shifts}
@@
 The RMSD$_{\text{corrected}}$ value signifies the shift in preference.
-\comment{add more here about absolutely conserved, clade-specific sites, etc}
+
+For the plots shown in Figure~\ref{fig:RMSD_heatmap}B, any residues falling between Cys-52 and Cys-277 were defined as the head domain, and all other residues were defined as the stalk domain.
+We used a multiple sequence alignment of the HA subtype sequences provided in~\cite{doud2017quantifying} to list sites that are absolutely conserved across all subtypes.
+To find clade-specific sites, we listed sites with a single amino-acid identity in all of H1, H2, H5, and H6, and a different identity in H3, H4, and H14. 
 }
@@
 {\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.} 
-The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter \comment{somehow indicate where this stringency parameter}.
+The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
 The sites are in H3 numbering.
@@
 
+\begin{suppdata}
+\caption{\label{suppdata:bcsubamp_primers}
+Excel file providing the primers used to generate the barcoded subamplicons for Perth/2009 HA deep sequencing.
+}
+\end{suppdata}
+
+\begin{suppdata}
+\caption{\label{suppdata:rep1_prefs}
+CSV file providing the amino-acid preferences for replicate 1 in sequential HA numbering.
+}
+\end{suppdata}
+
+\begin{suppdata}
+\caption{\label{suppdata:rep2_prefs}
+CSV file providing the amino-acid preferences for replicate 2 in sequential HA numbering.
+}
+\end{suppdata}
+
+\begin{suppdata}
+\caption{\label{suppdata:rep3-1_prefs}
+CSV file providing the amino-acid preferences for replicate 3-1 in sequential HA numbering.
+}
+\end{suppdata}
+
+\begin{suppdata}
+\caption{\label{suppdata:rep3-2_prefs}
+CSV file providing the amino-acid preferences for replicate 3-2 in sequential HA numbering.
+}
+\end{suppdata}
+
+\begin{suppdata}
+\caption{\label{suppdata:avg_prefs}
+CSV file providing the across-replicate averaged and re-scaled amino-acid preferences in H3 numbering.
+}
+\end{suppdata}
+
 \end{document}

### 2018-01-17 paper/manuscript.tex
Some additions of detail to methods

@@
 \matmethods{Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610. 
-Computer code used to analyze the data and produce the results in the paper are in...
+Computer code used to analyze the data and produce the results in the paper are in \comment{link to repository}.
 
@@
 
-To generate wildtype Perth/2009 H3N2 virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 cells with reverse genetics plasmids encoding the Perth/2009 HA and NA, WSN/1933 internal genes, and the TMPRSS2 lentiviral vector.
+To generate wildtype Perth/2009 H3N2 virus, we transfected in triplicate co-cultures of 293T and MDCK-SIAT1-TMPRSS2 cells with reverse genetics plasmids encoding the Perth/2009 HA and NA, WSN/1933 internal genes, and the TMPRSS2 lentiviral vector.
 \comment{add more details about reverse genetics?}
 The viral supernatant was harvested at 72 hours post-transfection, clarified at 1200 rpm for 5 min, and 0.01 to 1.0 ul of the supernatant was passaged onto $4 \times 10^{5}$ MDCK-SIAT1-TMPRSS2 cells in six-well plates for a total of six serial passages.
-After the final passage, RNA was extracted from 140 ul of the viral supernatant using the QIAamp Viral RNA Mini Kit.
+After the final passage, RNA was extracted from 140 ul of the viral supernatant for each replicate using the QIAamp Viral RNA Mini Kit.
 The HA and NA genes were reverse transcribed, PCR amplified, and sequenced by Sanger sequencing.
-Two mutations, G78D and T212I, arose to fixation after serial passaging. 
+Two mutations, G78D and T212I, arose to fixation in one of the replicates after serial passaging. 
 The Perth/2009 HA carrying these two mutations were cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} plasmid backbones.
@@
 The codon-mutant libraries were generated in the Perth/2009 HA-G78D-T212I background using a PCR-based approach described in~\cite{bloom2014experimentally,dingens2017comprehensive} using two rounds of mutagenesis.
+The script used to design the mutagenesis primers is on \url{https://github.com/jbloomlab/CodonTilingPrimers}.
 The codon mutations were introduced into three independent replicates of the Perth/2009 HA variant.
@@
 We allowed the transfected cells to adhere in 6-well plates, then four hours later changed the media to D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL).
-Eighteen hours after transfection, we infected the cells with WSN/1933 HA-deficient helper virus by preparing an inoculum of 500 TCID$_{50}$ per $\mu$L in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL), and aspirating the D10 media from the cells, and adding 2 mL of the helper-virus inoculum to each well.
+Eighteen hours after transfection, we infected the cells with WSN/1933 HA-deficient helper virus by preparing an inoculum of 500 TCID$_{50}$ per $\mu$L in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL), aspirating the D10 media from the cells, and adding 2 mL of the helper-virus inoculum to each well.
 After three hours, we changed the media to fresh influenza growth media.
@@
 The subsampled sequences were then aligned using MAFFT~\cite{katoh2013mafft} and used to build a phylogenetic tree using RAxML~\cite{stamatakis2006raxml}.
-We then used the phydms program~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms})
+We then used the phydms program~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}) to fit phylogenetic substitution models using the aligned subsampled sequences.
 
@@
 \caption{\label{suppfig:seqpref_zoomed}
-{\bf The per-site sequence preference at epitope and non-epitope sites}
-The per-site sequence preferences shown in Figure~\ref{fig:sequence_preference}B, but on their own y-axes.
+{\bf The per-site sequence preference at epitope and non-epitope sites.}
+The per-site sequence preferences shown in Figure~\ref{fig:sequence_preference}B, but on separate y-axes.
 }

### 2018-01-18 paper/manuscript.tex
moved up `\begin{document}` command to enable `latexdiff` tracking
Jesse edits to Abstract

@@
 
-\title{Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fate of human H3N2 influenza virus lineages}
+\begin{document}
+
+\title{Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fates of human H3N2 influenza virus lineages}
 
@@
 Human influenza virus rapidly accumulates mutations in its major surface protein hemagglutinin (HA).
-Forecasting the evolutionary success of influenza virus strains requires understanding how these mutations affect the protein's functionality and antigenicity.
+The evolutionary success of influenza virus lineages depends on how these mutations affect HA's functionality and antigenicity.
 Here we experimentally measure the effects on viral growth in cell culture of all single amino-acid mutations to the HA from a recent human H3N2 influenza virus strain. 
-We show that these measurements are informative for understanding viral evolution in nature, since mutations in evolutionarily successful H3N2 viral lineages are measured to generally be more favorable than mutations in lineages that die out.
-We also compare our measurements to prior data for an H1 HA, and find that the amino-acid preferences of many sites have substantially shifted between H1 and H3 HAs.
-For instance, there is much less of a disparity in mutational tolerance between the head and stalk domains for the H3 HA.
-Overall, our work suggests that experimental measurements of mutational effects might be leveraged to help forecast the evolutionary fates of viral lineages in nature.
+We then show that mutations in evolutionarily successful H3N2 viral lineages are measured to generally be more favorable than mutations in lineages that die out.
+Therefore, despite the well-known caveats about cell-culture measurements of viral fitness, such measurements can still be informative for understanding virus evolution in nature.
+We also compare our measurements for an H3 HA to similar data previously generated for a distantly related H1 HA, and find substantial differences in which amino acids are preferred at many sites.
+For instance, the H3 HA has less disparity in mutational tolerance between the head and stalk domains than the H1 HA.
+Overall, our work suggests that experimental measurements of mutational effects can be leveraged to help forecast the evolutionary fates of viral lineages in nature --- but only when the measurements are made on a viral strain similar to the ones being studied in nature.
 \end{abstract}
@@
 \doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
-
-\begin{document}
 

### 2018-01-18 paper/manuscript.tex
Jesse edits to Intro and Results

@@
 \subsection*{Deep mutational scanning of HA from a recent strain of human H3N2 influenza virus}
-We performed a deep mutational scan to measure the effects of all amino-acid mutations to HA from the A/Perth/16/2009 (H3N2) strain on viral replication in cell culture. 
+We performed a deep mutational scan to measure the effects of all amino-acid mutations to HA from the A/Perth/16/2009 (H3N2) strain on viral growth in cell culture. 
 This strain was the H3N2 component of the influenza vaccine from 2010-2012~\cite{who2010d,who2011}.
-Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral replication in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and Dataset~\ref{suppdata:PerthHA}).
+Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral growth in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and Dataset~\ref{suppdata:PerthHA}).
 The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.
@@
 We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (Figure~\ref{suppfig:SangerSeq}).
-We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables the efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
+We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
 These mutant viruses derived all their non-HA genes from the lab-adapted A/WSN/1933 strain.
 Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
-To further increase viral titers, we used MDCK-SIAT1 cells that we had engineered to constitutively express the TMPRSS2 protease, which facilitates HA cleavage and activation~\cite{bottcher2006proteolytic, bottcher2010cleavage}.
+To further increase viral titers, we used MDCK-SIAT1 cells that we had engineered to constitutively express the TMPRSS2 protease, which cleaves the HA precursor to activate it for membrane fusion~\cite{bottcher2006proteolytic, bottcher2010cleavage}.
 
@@
 After generating the mutant virus libraries, we passaged them at low MOI in cell culture to create a genotype-phenotype link and select for functional HA variants (Figure~\ref{fig:dms_overview}A).
-All the experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B). 
+All experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B). 
 We also passaged and deep sequenced library 3 in duplicate (denoted as library 3-1 and 3-2) to gauge the experimental noise occurring \textit{within} a single biological replicate.
@@
 We first used the deep sequencing data for each replicate estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
-Because there are 567 residues in HA, there are $567 \times 20 = 11,340$ estimated amino-acid preferences.
+Because there are 567 residues in HA, there are $567 \times 20 = 11,340$ estimated amino-acid preferences, corresponding to $567 \times 19 = 10,773$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
 The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
 The biological replicates are fairly well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
-Replicate 1 exhibited the lowest correlation with the other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
+Replicate 1 exhibited the weakest correlation with the other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
 The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity after the reverse-genetics step contributes most of the experimental noise.
@@
 This result indicates that our experiments authentically capture some of the constraints on HA evolution. 
-The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models.
-However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid constraints measured in our experiments (the $\omega$ for the ExpCM) is close to one, indicating that the deep mutational scanning captures much of the purifying selection on HA.
+The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
+However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid constraints measured in our experiments (the $\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that the deep mutational scanning captures much of the purifying selection on HA.
 ExpCM fit a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
@@
 For instance, in the H3 HA, the short helix A in the stalk is very mutationally tolerant.
-Interestingly, more studies have reported selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2017quantifying,anderson2017natural} HAs.
+Interestingly, there are more reports of readily selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2017quantifying,anderson2017natural} HAs.
 
@@
 This difference in the effects of mutations between the trunk and side branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
-Some influenza sequences are determined using egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{wu2017structural,mcwhite2016sequence,skowronski2016mutations}.
+Some influenza sequences are derived from egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{wu2017structural,mcwhite2016sequence,skowronski2016mutations}.
 These mutations mostly occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
 We therefore re-calculated the statistics separating side branches to tip and internal nodes, and again found that the difference between the trunk and both sets of side branches was statistically significant (Figure~\ref{fig:trunkvssidebranch}B).
-Therefore, strains with mutations that we experimentally measure to be favorable for viral replication tend to do better in nature than strains with mutations that we measure to be less favorable.
+Therefore, strains with mutations that we experimentally measure to be favorable for viral growth tend to do better in nature than strains with mutations that we measure to be less favorable.
 
@@
 Median mutational effects in each window are shown as circles for the trunk and triangles for side branches. 
-The shaded regions demarcate the interquartile range.
-Negative numbers signify mutations towards less preferred amino acids, while positive numbers signify mutations to more preferred mutations.
-(B) The log$_{2}$ mutational effect for all side branch and all trunk mutations (left panel), and the same data but separating side branches to terminal and internal nodes.
-$P$-values were computed by randomizing the experimental measurements among sites 10,000 times, and determining how often the difference in median mutational effect between the trunk and side branches for the randomized data exceeded the actual value.
+Shaded regions demarcate the interquartile range.
+Negative numbers signify mutations to less preferred amino acids, while positive numbers signify mutations to more preferred mutations.
+(B) The log$_{2}$ mutational effect for all side branch and trunk mutations (left panel), and the same data but separating side branches to terminal and internal nodes.
+$P$-values were computed by randomizing the experimental measurements among sites 10,000 times, and determining how often the difference in median mutational effect between the trunk and side branches exceeded the actual value.
 (C) The average per-site sequence preference of every node in the phylogenetic tree in Fig.~\ref{fig:H3N2_phylogeny}.
@@
 
-But interestingly, there are clear differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
-The per-site sequence preference at epitope sites increases over time (Figure~\ref{fig:sequence_preference}B and Figure~\ref{suppfig:seqpref_zoomed}), whereas at non-epitope sites it remains relatively constant.
-In addition, the per-site sequence preference is consistently higher at non-epitope than epitope sites, perhaps consistent with the fact that strong immune selection can lead to the fixation of less functionally favorable mutations at epitope sites.
-The fact that the epitope sites exhibit more of an increase in sequence preference over time may be due to the fact that they fix more mutations.
-It is known that as mutations accumulate at protein sites, they can lead to epistatic shifts in the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
-Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some specific epitope sites in H3 HA~\cite{wu2017diversity}.
-It is also known that positive selection for immune escape can increase the amount of epistasis among substitutions by favoring the fixation of combinations of individually deleterious immune-escape mutations and counterbalancing secondary mutations~\cite{gong2014epistatically}.
-We hypothesize that immune selection increases the accumulation of epistatic interactions involving substitutions at epitope sites.
-Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA.
+However, there are differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
+The per-site sequence preference at epitope sites increases over time, whereas at non-epitope sites it remains relatively constant  (Figure~\ref{fig:sequence_preference}B and Figure~\ref{suppfig:seqpref_zoomed}).
+In addition, the per-site sequence preference is consistently higher at non-epitope than epitope sites, perhaps because immune pressure sometimes drives the fixation of functionally less favorable mutations at epitope sites.
+The fact that only epitope sites exhibit an increase in sequence preference over time can also be rationalized in terms of prior knowledge about protein evolution.
+As proteins evolve, there can be epistatic shifts in the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
+Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some epitope sites in H3 HA~\cite{wu2017diversity}.
+It is also known that immune pressure can increase the amount of epistasis among substitutions by selecting for combinations of functionally deleterious immune-escape mutations and counterbalancing secondary mutations~\cite{gong2014epistatically}.
+We hypothesize that immune pressure increases the accumulation of epistatic interactions involving substitutions at epitope sites.
+Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA (Figure~\ref{fig:sequence_preference}B and Figure~\ref{suppfig:seqpref_zoomed}).
+Such a trend is not apparent for non-epitope sites, probably because they are under less selection to fix adaptive mutations.
 
@@
 The results above show that measurements of mutational effects in the Perth/2009 HA are informative about the evolutionary fate of other human H3N2 HAs.
-How broadly can experimental measurements be generalized across HAs?
+How broadly can such measurements be generalized across HAs?
 To address this question, we repeated the trunk versus side-branch analyses using amino-acid preferences from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
@@
 However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
-One simple way to investigate the extent of shifts in amino-acid preferences is simply to correlate measurements from independent deep mutational scanning replicates on H1 and H3 HAs.
-Figure~\ref{fig:distance_distribution}B clearly reveals that replicate measurements on the same HA variant are more correlated than those on different HA variants.
+One simple way to investigate the extent of shifts in amino-acid preferences is to correlate measurements from independent deep mutational scanning replicates on H1 and H3 HAs.
+Figure~\ref{fig:distance_distribution}B shows that replicate measurements on the same HA variant are more correlated than those on different HA variants.
 
-To more rigorously quantify shifts in amino-acid preferences after correcting for experimental noise, we used the statistical approach described in~\cite{doud2015site,haddox2017mapping}.
+To more rigorously quantify shifts in amino-acid preferences after correcting for experimental noise, we used the statistical approach in~\cite{doud2015site,haddox2017mapping}.
 Figure~\ref{fig:distance_distribution}C shows the distribution of shifts in amino-acid preferences between H3 and H1 HAs after correcting for experimental noise.
-Although some sites have small shifts near zero, many sites have large shifts in preference.
+Although some sites have small shifts near zero, many sites have large shifts.
 These shifts between H3 and H1 are much larger than expected from the null distribution that would be observed purely from experimental noise.
 They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2017mapping}.
-However, the shifts are still generally smaller than those observed when comparing HA to the non-homologous HIV Env protein. 
-These observations show that there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations still remain more similar than for non-homologous proteins.
+However, the shifts are still smaller than those observed when comparing HA to the non-homologous HIV Env protein. 
+Therefore, there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations remain more similar than for non-homologous proteins.
 
@@
 \subsection*{Properties associated with the shifts in amino-acid preferences between H3 and H1 HAs}
-What features distinguish the sites that have shifted versus conserved amino-acid preferences between H3 and H1 HAs?
-The sites of large shifts are spread across HA, and do not obviously localize to one specific region of the protein's structure (Figure~\ref{fig:RMSD_heatmap}A).
+What features distinguish the sites with shifted amino-acid preferences between H3 and H1 HAs?
+The sites of large shifts do not obviously to one specific region of HA's structure (Figure~\ref{fig:RMSD_heatmap}A).
 However, at the domain level, sites in HA's stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B). 
-The HA stalk domain is also more conserved at the amino-acid level across HA subtypes~\cite{nobusawa1991comparison,hai2012influenza,mallajosyula2014influenza}, suggesting that conservation of amino-acid sequence tends to be correlated with conservation of amino-acid preferences.
+The HA stalk domain is also more conserved in sequence~\cite{nobusawa1991comparison,hai2012influenza,mallajosyula2014influenza}, suggesting that conservation of amino-acid sequence tends to be correlated with conservation of amino-acid preferences.
 Consistent with this idea, sites that are absolutely conserved across all 18 HA subtypes are significantly less shifted than sites that are variable across HA subtypes (Figure~\ref{fig:RMSD_heatmap}B).
@@
 Specifically, the clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
-In deep mutational scanning of the WSN/1933 H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
+In deep mutational scanning of the H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
 In contrast, the interhelical loop in H3 HA makes a sharper and shorter turn which is facilitated by a Gly at 75(HA2).
-In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations suggesting that the amino-acid identity at this site plays a smaller role in determining the conformation of the interhelical loop.
+In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations.
 Therefore, some of the shifts in HA amino-acid preferences can be directly rationalized in terms of changes in HA structure.
@@
 (C) Sites 107 and 75(HA2) help determine the different orientation of the globular head domain in H1 versus H3 HAs.
-These sites are shown in spheres on the structure of H1 and H3. 
+These sites are shown in spheres on the structure of H1 and H3 and colored as in panel (A), and the experimentally measured amino-acid preferences in the H1 and H3 HAs are shown.
 One monomer is in dark gray, while the HA1 domain of the neighboring monomer is in lighter gray.
-The spheres are colored as in panel (A), and the experimentally measured amino-acid preferences in the WSN/1933 H1 and Perth/2009 H3 HAs are shown.
 }
@@
 We refer to the resulting bulk population of sorted cells as MDCK-SIAT1-TMPRSS2 cells.
-There is no selectable marker for the TMPRSS2; however, we generally maintain the cells at low passage number, and have seen no indication that lose their ability to support the replication of viruses with H3 HAs in the absence of exogenous trypsin.
+There is no selectable marker for the TMPRSS2; however, we generally maintain the cells at low passage number, and have seen no indication that lose their ability to support the growth of viruses with H3 HAs in the absence of exogenous trypsin.
 

### 2018-01-19 paper/manuscript.tex
added skeleton for post-Perth/2009 tree

@@
 \renewcommand{\thesuppdata}{S\arabic{suppdata}}
+
+\setcounter{topnumber}{10}
+\setcounter{bottomnumber}{10}
+\setcounter{totalnumber}{10}
 
@@
 
-We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (Figure~\ref{suppfig:SangerSeq}).
-We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
-These mutant viruses derived all their non-HA genes from the lab-adapted A/WSN/1933 strain.
-Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
-To further increase viral titers, we used MDCK-SIAT1 cells that we had engineered to constitutively express the TMPRSS2 protease, which cleaves the HA precursor to activate it for membrane fusion~\cite{bottcher2006proteolytic, bottcher2010cleavage}.
-
 \begin{SCfigure*}[\sidecaptionrelwidth][t]
@@
 \end{SCfigure*}
+
+We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (Figure~\ref{suppfig:SangerSeq}).
+We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
+These mutant viruses derived all their non-HA genes from the lab-adapted A/WSN/1933 strain.
+Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
+To further increase viral titers, we used MDCK-SIAT1 cells that we had engineered to constitutively express the TMPRSS2 protease, which cleaves the HA precursor to activate it for membrane fusion~\cite{bottcher2006proteolytic, bottcher2010cleavage}.
 
@@
 The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
-These results suggest that there is some HA background-dependence in the effects of mutations, but that despite this fact our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.
+We do not think that this result means that Perth/2009 is actually higher fitness than the other strains.
+Rather, if there are modest shifts in the effects of mutations as HA evolves, then sequences more distant from Perth/2009 could be scored as having lower preference simply because these shifts start to degrade the accuracy of our measurements.
+But importantly, our experiments consistently show the trunk to be more favorable than the side branches regardless of whether we look at sequences that precede Perth/2009 (most of Figure~\ref{fig:trunkvssidebranch}C) or sequences that occur after Perth/2009 has diverged from the trunk (Figure~\ref{suppfig:future}).
+Therefore, our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.
 
@@
 \begin{suppfigure}
+\comment{add figure}
+\caption{\label{suppfig:future}
+Our experiments show a selective advantage for the trunk even if we limit the analysis to sequences that occur after the split of Perth/2009 off the trunk.
+(A) Phylogenetic tree showing all nodes that branch from the trunk before or contemporaneously with Perth/2009 (orange), all nodes that occur after the branch to Perth/2009 for which we can resolve the trunk (red) or side branches (blue), and all nodes for which it is not yet clear which sequences will be on the trunk or side branches.
+In this figure, we restrict the analysis to the nodes in red or blue.
+(B)  
+Average per-site sequence preference (similar to Figure~\ref{fig:trunkvssidebranch}C) for all trunk and side-branch nodes that occur after the branch to Perth/2009.
+(C)
+The sequence preferences of nodes on the trunk generally exceed those of nodes on the side branches along the post-Perth/2009 portion of the tree.
+Note that because there are relatively few post-Perth/2009 sequences for which we can currently distinguish trunk from side branch, the number of points in this plot is small, precluding meaningful statistical analyses of the type performed in Figure~\ref{fig:trunkvssidebranch}B.
+}
+\end{suppfigure}
+
+\begin{suppfigure}
 \centerline{\includegraphics[width=0.6\textwidth]{figs/S04_seqpref_zoomed/seqpref_zoomed.pdf}}
@@
 
-\subsubsection*{SI Tables}
-
-Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference); include only one table per file.
-Do not use tabs or spaces to separate columns in Word tables.
+\clearpage
 
 \subsubsection*{SI Datasets} 
+\FloatBarrier
 

### 2018-01-19 paper/manuscript.tex
revisions to Discussion

@@
 \label{sec:discussion}
-We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 on viral growth in cell culture, and demonstrated that these measurements can be informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
-Specifically, strains on the trunk of the phylogenetic tree tend to have mutations that are beneficial for viral growth in our experiments when compared to side-branch mutations.
-Differentiating between trunk and side branch lineages can enable forecasting of influenza virus evolution by identifying which strains are likely to succeed in nature.
-The fact that our experiments are able to reveal differences between trunk and side branch lineages highlight the potential for incorporating experimental measurements of the non-antigenic effects of mutations for evolutionary forecasting.
+We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture, and demonstrated that these measurements are informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
+Specifically, HA sequences on the trunk of the phylogenetic tree tend to have mutations that our experiments measure to be more beneficial for viral growth than mutations that occur in side-branch HA sequences.
+The fact that our experiments can help distinguish between trunk and side-branch lineages suggests that they might inform evolutionary forecasting.
+In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, Luksza and Lassig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
+Our work suggests that deep mutational scanning can provide such data.
 
-Surprisingly, our simple experimental measurements of viral growth in cell culture helped to elucidate differences in fitness of viral strains in nature.
-Mutations may have effects in the laboratory that are distinct from their effects in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}, and our experiments certainly don't capture all of the relevant selective pressures on HA that exist in nature.
-Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
-Yet, prior to our work presented here, there were no comprehensive studies of the functional effects of mutations to HA on viral fitness outside of antigenicity.
-Previous quantitative modeling work either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or made the simplifying assumption that all mutations outside epitope sites had equivalent deleterious effects on viral fitness~\cite{luksza2014predictive}.
-While our measurements may be imperfect, we have demonstrated that these measurements are still of utility in distinguishing the selective advantage of successful influenza viral lineages over $\sim$50 years of evolution.
-This also suggests that measurements in experimental models that more faithfully recapitulate the selective forces in nature should be of even greater utility for differentiating evolutionarily successful viral lineages.
+It is important to emphasize that measurements of viral growth in cell culture do \emph{not} represent true fitness in nature.
+Indeed, a vast amount of work in virology has chronicled the many ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}.
+Mutations in viral genes other than HA are also certainly important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
+Given these caveats, it might seem surprising that simply measuring viral growth in cell culture can be informative about the success of viral strains in nature.
+Yet, prior to our work, there were no comprehensive studies of the functional effects of mutations to H3 HA on any property that even resembled viral fitness in nature, and modeling work has either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or simply assumed that all non-epitope mutations had equivalent deleterious effects~\cite{luksza2014predictive}.
+The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative --- and represent a vast improvement over no information at all.
+That this is true is demonstrated by the fact that these measurements have utility for distinguishing successful and unsuccessful viral lineages over $\sim$50 years of H3N2 evolution.
+We suspect that performing the experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures~\cite{?}) might further improve their utility.
 
-In addition, we are measuring the effects of mutations in a single HA variant from the Perth/2009 strain and extrapolating these measurements to other seasonal human H3N2 strains across a 50-year timespan.
-However, these extrapolations will only be valid to the extent that the effects of mutations remain conserved during HA's evolution.
-Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
-% and the extent to which this occurs is an important question in its own right.
-The observation that our measurements score trunk mutations more favorably than side branch mutations across most of H3N2 evolution suggests that for relatively closely related strains, the mutational effects are not shifted to a large extent.
-On the other hand, when we compared our measurements for H3 to prior measurements on H1, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs.
-Further analysis of how mutational effects shift as protein variants diverge can provide insight into how broadly the results of a given experiment can be generalized across variants.
+We measured the effects of all single amino-acid mutations to a specific HA, and then generalized these measurements to other H3N2 HAs from a 50-year timespan.
+These generalizations will only be valid to the extent that the effects of mutations are conserved during HA's evolution.
+Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}, and the extent to which this occurs is an important question in its own right.
+The fact that our measurements score trunk mutations more favorably than side branch ones across a 50-year timespan suggests that the mutational effects have not shifted to dramatically among these relatively closely related sequences.
+On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2017mapping}.
+Further investigation of how mutational effects shift as proteins diverge will be important for determining how broadly any given experiment can be generalized when attempting to make evolutionary forecasts.
 
-Importantly, the antigenic properties of HA are integral in determining strain success, but we did not study the effect of mutations on antigenicity.
-However, our high-throughput experimental approach can be harnessed to profile mutations that confer escape from antibodies, as has been performed for the WSN/1933 H1 HA~\cite{doud2017complete,doud2017quantifying} and HIV Env~\cite{dingens2017comprehensive}.
-In future, these antigenic studies can be combined with functional measurements to inform forecasting of the evolution of seasonal influenza virus.
+Our work did not characterize the antigenic effects of mutations, which also play an important role in determining strain success in nature~\cite{?}.
+Indeed, some of our results (e.g., Figure~\ref{fig:sequence_preference}) suggest as much by showing that the experiments better capture selection at non-epitope than epitope sites.
+However, our basic selection and deep-sequencing approach can be harnessed to completely map how mutations affect antibody recognition~\cite{doud2017complete,doud2017quantifying}.
+But so far, experiments using this approach have either failed to examine antibodies or sera that are relevant to driving the evolution of H3N2 influenza~\cite{doud2017complete,doud2017quantifying}, or have used relevant sera but examined a non-comprehensive set of mutations~\cite{li2016selection}.
+Future experiments that completely map how HA mutations affect recognition by human sera seem likely to be especially fruitful for informing viral forecasting.
 
@@
 \caption{\label{suppdata:rep1_prefs}
-CSV file providing the amino-acid preferences for replicate 1 in sequential HA numbering.
-}
-\end{suppdata}
-
-\begin{suppdata}
-\caption{\label{suppdata:rep2_prefs}
-CSV file providing the amino-acid preferences for replicate 2 in sequential HA numbering.
-}
-\end{suppdata}
-
-\begin{suppdata}
-\caption{\label{suppdata:rep3-1_prefs}
-CSV file providing the amino-acid preferences for replicate 3-1 in sequential HA numbering.
-}
-\end{suppdata}
-
-\begin{suppdata}
-\caption{\label{suppdata:rep3-2_prefs}
-CSV file providing the amino-acid preferences for replicate 3-2 in sequential HA numbering.
-}
-\end{suppdata}
-
-\begin{suppdata}
-\caption{\label{suppdata:avg_prefs}
-CSV file providing the across-replicate averaged and re-scaled amino-acid preferences in H3 numbering.
+Excel file giving the amino-acid preferences in sequential 1, 2, ... numbering of the Perth/2009 HA.
+The unscaled preferences for replicates 1, 2, 3-1, and 3-2 are each in a separate tab of the file.
+Additional tabs give the across-replicate averaged and re-scaled amino-acid preferences in sequential numbering and in H3 numbering as shown in Figure~\ref{fig:logoplot}.
+There are also tabs that give the conversion from sequential to H3 numbering, and the epitope sites in H3 numbering as defined by Wolf et al and used in this study.
+Each tab can simply be exported to CSV for computational analyses.
 }

### 2018-01-19 paper/manuscript_origmethods.tex
Jesse a bit of work on Methods

--- 2018-01-19 paper/manuscript.tex
@@
 What features distinguish the sites with shifted amino-acid preferences between H3 and H1 HAs?
-The sites of large shifts do not obviously to one specific region of HA's structure (Figure~\ref{fig:RMSD_heatmap}A).
+The sites of large shifts do not obviously localize to one specific region of HA's structure (Figure~\ref{fig:RMSD_heatmap}A).
 However, at the domain level, sites in HA's stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B). 
@@
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing. 
-This work was supported by...}
+This work was supported by grant R01AI127897 from the NIAID of the NIH. 
+JML was supported in part by the Center for Inference and Dynamics of Infectious Diseases (CIDID), which is funded by grant U54GM111274 from the NIGMS of the NIH.
+The research of JDB is supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation, and a Burroughs Wellcome Young Investigator in the Pathogenesis of Infectious Diseases grant.
+\comment{Add any additional relevant grants for Trevor and John}
+}
 

### 2018-01-19 paper/manuscript.tex
Tiny change to sentence in Results
edits to Methods

@@
 
-\matmethods{Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610. 
+\matmethods{
+\subsection*{Data and computer code}
+Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610. 
 Computer code used to analyze the data and produce the results in the paper are in \comment{link to repository}.
@@
 \subsection*{HA numbering}
-Unless otherwise indicated, all sites are in H3 numbering, with the signal peptide in negative numbers, the HA1 subunit in plain numbers, and the HA2 subunit denoted with "(HA2)".
-The conversion between sequential numbering of the A/Perth/16/2009 HA and H3 numbering was performed using an HA numbering Python script (available at \url{https://github.com/jbloomlab/HA_numbering}).
+Sites are in H3 numbering, with the signal peptide in negative numbers, HA1 in plain numbers, and HA2 denoted with "(HA2)".
+Sequential 1, 2, ...
+numbering the Perth/2009 HA can be converted to H3 numbering by subtracting 16.
 
-\subsection*{Creation of MDCK-SIAT1-TMPRSS2 cell line}
-When growing influenza virus, TPCK-trypsin must normally be added in order to cleave HA into its mature form.
-In order to obviate the need for trypsin, we engineered an MDCK-SIAT1 cell line to constitutively express the TMPRSS2 protease.
-This protease cleaves and activates HA and is found endogenously in the human airway~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
-The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656), PCR amplified, and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry.
+\subsection*{Creation of MDCK-SIAT1-TMPRSS2 cells}
+When growing influenza virus in cell culture, trypsin is normally added in order to cleave HA into its mature form.
+To obviate the need for trypsin, we engineered an MDCK-SIAT1 cell line to constitutively express the TMPRSS2 protease, which cleaves and activates HA in the human airways~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
+The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656) and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry to create plasmid pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
 We used the lentiviral vector to transduce MDCK-SIAT1 cells, and sorted an intermediate mCherry population by flow cytometry.
-We refer to the resulting bulk population of sorted cells as MDCK-SIAT1-TMPRSS2 cells.
-There is no selectable marker for the TMPRSS2; however, we generally maintain the cells at low passage number, and have seen no indication that lose their ability to support the growth of viruses with H3 HAs in the absence of exogenous trypsin.
+We refer to the sorted bulk population as MDCK-SIAT1-TMPRSS2 cells.
+There is no selectable marker for the TMPRSS2; however, we maintain the cells at low passage number, and have seen no indication that lose their ability to support the growth of viruses with H3 HAs in the absence of exogenous trypsin.
 
 \subsection*{Generation of HA codon mutant plasmid libraries}
-Recombinant A/Perth/16/2009 (HA, NA) $\times$ A/Puerto Rico/8/1934 influenza virus, NIB-64, NR-41803 was ordered from BEI Resources, NIAID, NIH. 
-Bulk RNA from the viral sample was extracted using the QIAamp Viral RNA Mini Kit (QIAGEN) according to manufacturer's instructions.
-The Perth/2009 HA and NA genes were then reverse transcribed using the SuperScript III First-Strand Synthesis SuperMix Kit (Invitrogen), PCR amplified, and cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} plasmid backbones by In-Fusion.
+The HA and NA genes for the Perth/2009 viral strain were reverse-transcribed and amplified from recombinant virus obtained from BEI Resources (NR-41803), and cloned into the pHW2000~\citep{hoffmann2000dna} influenza reverse-genetics plasmids to create pHW-Perth09-HA and pHW-Perth09-NA.
 
-To generate wildtype Perth/2009 H3N2 virus, we transfected in triplicate co-cultures of 293T and MDCK-SIAT1-TMPRSS2 cells with reverse genetics plasmids encoding the Perth/2009 HA and NA, WSN/1933 internal genes, and the TMPRSS2 lentiviral vector.
-\comment{add more details about reverse genetics?}
-The viral supernatant was harvested at 72 hours post-transfection, clarified at 1200 rpm for 5 min, and 0.01 to 1.0 ul of the supernatant was passaged onto $4 \times 10^{5}$ MDCK-SIAT1-TMPRSS2 cells in six-well plates for a total of six serial passages.
-After the final passage, RNA was extracted from 140 ul of the viral supernatant for each replicate using the QIAamp Viral RNA Mini Kit.
-The HA and NA genes were reverse transcribed, PCR amplified, and sequenced by Sanger sequencing.
-Two mutations, G78D and T212I, arose to fixation in one of the replicates after serial passaging. 
-The Perth/2009 HA carrying these two mutations were cloned into the pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} plasmid backbones.
+We initially created a virus with the HA and NA from Perth/2009 and internal genes from WSN/1933, and passaged it in cell culture to test its genetic stability.
+To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amount of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
+The next day we changed the media to influenza growth media (abbreviated IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL --- no trypsin was added since there was TMPRSS2), and then we collected the viral supernatant at 72 hours post-transfection.
+This viral supernatant was then blind passaged in MDCK-SIAT1-TMPRSS2 a total of six additional times.
+We then isolated viral RNA from these passaged viruses and sequenced the HA gene.
+The passaged HA had two mutations, G78D and T212I, which we found enhanced viral growth as shown in Figure~\ref{suppfig:Perth2009_mut}.
+The HA with these two mutations was cloned into pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} to create pHW-Perth09-HA-G78D-T212I and pICR2-Perth09-HA-G78D-T212I.
+For all subsequent experiments, we used viruses with the HA containing these two mutations in order to improve titers and viral genetic stability, and this is the HA that we refer to as Perth/2009 in the manuscript text.
+We also used all non-HA genes (including NA) from WSN/1933 to help increase titers and reduce biosafety concerns.
 
-To validate that the variant Perth/2009 HA could support viral growth, we generated in duplicate by reverse genetics viruses with the original Perth/2009 HA or the Perth/2009 HA-G78D-T212I variant on a background of Perth/2009 NA $+$ WSN internal genes.
-The transfection supernatant was collected 72 hours post-transfection and passaged onto a monolayer of MDCK-SIAT1-TMPRSS2 cells at an MOI = 0.01.
-The viral supernatant was collected at 44 hours post-infection.
-Both the transfection and passage supernatants were titered by TCID$_{50}$ in MDCK-SIAT1-TMPRSS2 cells.
-
-The codon-mutant libraries were generated in the Perth/2009 HA-G78D-T212I background using a PCR-based approach described in~\cite{bloom2014experimentally,dingens2017comprehensive} using two rounds of mutagenesis.
-The script used to design the mutagenesis primers is on \url{https://github.com/jbloomlab/CodonTilingPrimers}.
-The codon mutations were introduced into three independent replicates of the Perth/2009 HA variant.
-The mutant variants were then cloned into the pICR2 vector by digestion with BsmBI restriction sites, ligation by T4 DNA ligase, and electroporation and transformation into ElectroMAX DH10B competent cells (Invitrogen; 18290015).
-We pooled over 6 million transformants for each replicate, cultured in liquid LB + ampicillin at 37$^{\circ}$C for 3 h with shaking, and maxiprepped.
-We randomly chose 31 clones to Sanger sequence, and summary statistics from sequencing are shown in Figure~\ref{suppfig:SangerSeq}.
+The codon-mutant libraries were generated in the Perth/2009 HA-G78D-T212I background using the PCR-based approach described in \cite{bloom2014experimentally} with the primer melting-temperature modifications described in \cite{dingens2017comprehensive}, using two rounds of mutagenesis.
+The script to design the mutagenesis primers is at \url{https://github.com/jbloomlab/CodonTilingPrimers}.
+We created three independent libraries, one for each biological replicate.
+The mutant variants were then cloned at high efficiency into the pICR2 vector using digestion with BsmBI, ligation with T4 DNA ligase, and electroporation into ElectroMAX DH10B competent cells (Invitrogen 18290015).
+We obtained $>$6 million transformants for each replicate.
+We scraped the plates, expanded the cultures in liquid LB + ampicillin at 37$^{\circ}$C for 3 h with shaking, and then maxiprepped.
+We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (Figure~\ref{suppfig:SangerSeq}).
 
 \subsection*{Generation and passaging of mutant viruses}
-The mutant virus libraries were generated and passaged using the approach described in~\cite{doud2016accurate} with several modifications.
-We used the HA-deficient WSN/1933 helper virus generated in~\cite{doud2016accurate}.
-Briefly, these helper viruses were generated by transfecting seven WSN/1933 non-HA reverse-genetics plasmids plus a WSN/1933 HA protein expression plasmid into a coculture of 293T and MDCK-SIAT1-EF1$\alpha$-WSN-HA cells engineered to constitutively express the WSN/1933 HA.
-The transfection supernatant was then expanded by passaging onto a monolayer of MDCK-SIAT1-EF1$\alpha$-WSN-HA cells, and the viral supernatant was then harvested and aliquots were frozen at -80$^{\circ}$C.
+The mutant virus libraries were generated using the helper-virus approach described in \cite{doud2016accurate} with several modifications, most notably changing the cell line used.
+Briefly, we transfected $5 \times 10^5$ MDCK-SIAT1-TMPRSS2 cells in suspension with 937.5 ng each of four protein expression plasmids encoding the ribonucleoprotein complex (HDM-Nan95-PA, HDM-Nan95-PB1, HDM-Nan95-PB2, and HDM-Aichi68-NP)~\cite{gong2013stability}, and 1250 ng of one of the three pICR2-mutant-HA libraries (or the wildtype control) using Lipofectamine 3000 (ThermoFisher L3000008).
+We allowed the transfected cells to adhere in 6-well plates, four hours later changed the media to D10 media.
+Eighteen hours after transfection, we infected the cells with the WSN/1933 HA-deficient helper virus~\cite{doud2016accurate} by preparing an inoculum of 500 TCID$_{50}$ per $\mu$L of helper virus (as computed on HA-expressing cells) in IGM, aspirating the D10 media from the cells, and adding 2 mL of the helper-virus inoculum to each well.
+After three hours, we changed the media to fresh IGM.
+At 24 hours after helper-virus infection, we harvested the viral supernatants for each replicate, froze aliquots at -80$^{\circ}$C, and titered in MDCK-SIAT1-TMPRSS2 cells.
+The titers were 92, 536, 536, and 734 TCID$_{50}$ per $\mu$L for the three library replicates and the wildtype control, respectively.
 
-To generate the mutant virus libraries, we transfected $5 \times 10^5$ MDCK-SIAT1-TMPRSS2 cells in suspension with 937.5 ng each of four protein expression plasmids encoding the ribonucleoprotein complex (HDM-Nan95-PA, HDM-Nan95-PB1, HDM-Nan95-PB2, and HDM-Aichi68-NP)~\cite{gong2013stability}, and 1250 ng of one of the three pICR2-mutant-HA libraries (or the wildtype pICR2-Perth2009-HA-G78D-T212I control) using Lipofectamine3000 (ThermoFisher;L3000008).
-We allowed the transfected cells to adhere in 6-well plates, then four hours later changed the media to D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL).
-Eighteen hours after transfection, we infected the cells with WSN/1933 HA-deficient helper virus by preparing an inoculum of 500 TCID$_{50}$ per $\mu$L in influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL), aspirating the D10 media from the cells, and adding 2 mL of the helper-virus inoculum to each well.
-After three hours, we changed the media to fresh influenza growth media.
-Twenty-four hours after helper-virus infection, we harvested the viral supernatants for each replicate, froze aliquots at -80$^{\circ}$C, and titered in MDCK-SIAT1-TMPRSS2 cells.
-
-We passaged over $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
-To passage, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which the cells reached a density of $\sim 1.7 \times 10^7$ cells per dish.
-We then replaced the media of each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L.
+We passaged $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
+To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they had reached a density of $\sim 1.7 \times 10^7$ cells per dish.
+We then replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L.
 We collected viral supernatant for sequencing 48 hours post-infection.
@@
 \subsection*{Barcoded subamplicon sequencing}
-To extract viral RNA from the three replicate HA virus libraries and the wildtype HA virus, we first ultracentrifuged 24 mL of the supernatant at 22,000 rpm for 1.5 h at 4$^{\circ}$C in a Beckman Coulter SW28 rotor.
-We then extracted the RNA using the Qiagen RNeasy Mini Kit by resuspending the viral pellet in 400 $\mu$L of buffer RLT supplemented with $\beta$-mercaptoethanol, pipetting 30 times, transferring the liquid to a microcentrifuge tube, adding 600 $\mu$L 70\% ethanol, and proceeding with the rest of the extraction according to manufacturer's instructions.
+To extract viral RNA from the three replicate HA virus libraries and the wildtype HA virus, we ultracentrifuged 24 mL of the supernatant at 22,000 rpm for 1.5 h at 4$^{\circ}$C in a Beckman Coulter SW28 rotor, and extracted RNA using the Qiagen RNeasy Mini Kit by resuspending the viral pellet in 400 $\mu$L of buffer RLT supplemented with $\beta$-mercaptoethanol, pipetting 30 times, transferring the liquid to a microcentrifuge tube, adding 600 $\mu$L 70\% ethanol, and proceeding with the extraction according to the manufacturer's instructions.
 The HA gene was then reverse-transcribed with AccuScript Reverse Transcriptase (Agilent 200820) using the primers P09-HA-For (5'-AGCAAAAGCAGGGGATAATTCTATTAATC-3') and P09-HA-Rev (5'-AGTAGAAACAAGGGTGTTTTTAATTACTAATACAC-3').
 
-We generated the HA PCR amplicons for each of the three plasmid libraries, the three virus libraries, one wildtype plasmid, and one wildtype virus samples using KOD Hot Start Master Mix (EMD Millipore;71842) according to the PCR reaction mixture and cycling conditions described in~\cite{bloom2014experimentally} and the P09-HA-For and P09-HA-Rev primers.
+We generated the HA PCR amplicons for the three plasmid libraries, the three virus libraries, the wildtype plasmid control, and the wildtype virus control using KOD Hot Start Master Mix (EMD Millipore;71842) using the PCR reaction mixture and cycling conditions described in~\cite{bloom2014experimentally} and the P09-HA-For and P09-HA-Rev primers.
 We prepared the sequencing libraries using a barcoded-subamplicon strategy~\cite{wu2014high} to increase the accuracy from deep sequencing. 
-The approach we used is described in~\cite{doud2016accurate} (also see \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
-The primers we used to generate the subamplicons are in Dataset~\ref{suppdata:bcsubamp_primers}.
-We performed sequencing on a lane of a flow cell of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.
+The exact details of this approach are described in~\cite{doud2016accurate} (also see \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
+The primers used to generate the subamplicons are in Dataset~\ref{suppdata:bcsubamp_primers}.
+We performed deep sequencing on a lane of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.
 
 \subsection*{Analysis of deep sequencing data}
-We used the dms\_tools2 software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}) to analyze the deep sequencing data.
+We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, \comment{version ?}) to analyze the deep sequencing data.
 The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
 The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Datasets \comment{suppfile}.
+Computer code that performs the entirety of this analysis is available at \comment{add this}
 
-\subsection*{Alignments and phylogenetic analysis of HA sequences}
-To build the ExpCM using our measurements, we first downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
-The subsampled sequences were then aligned using MAFFT~\cite{katoh2013mafft} and used to build a phylogenetic tree using RAxML~\cite{stamatakis2006raxml}.
-We then used the phydms program~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}) to fit phylogenetic substitution models using the aligned subsampled sequences.
+\subsection*{Phylogenetic model comparison and fitting of a stringency parameter}
+For the analysis in Table~\ref{tab:phydms}, we downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
+These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
+We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, \comment{?}) to fit the substitution models listed in Table~\ref{tab:phydms}.
+The amino-acid preferences were re-scaled by the stringency parameter using the approach described in \cite{hilton2017phydms}.
 
-A description of the sequences and methods used to build the phylogenetic tree of HA subtypes shown in Figure~\ref{fig:distance_distribution}A is provided in~\cite{doud2017quantifying}.
+The phylogenetic tree of HA subtypes in Figure~\ref{fig:distance_distribution}A is was generated as described in~\cite{doud2017quantifying}.
 
@@
 \comment{John to add methods here. We downloaded X sequences from the Influenza Virus Resource ?.... etc. subsampled how? inferred the tree, ancestral state reconstruction, visualized the tree...}
-To parse out trunk mutations from side branch mutations, we first defined a set of recent nodes sampled on or after Jan. 1, 2017, and traced these nodes back to their most recent common ancestor. 
-We defined all branches ancestral to this most recent common ancestor of recently sampled nodes as the trunk.
-All nodes descended from the trunk were defined as side branch nodes.
-Nodes descended from the most recent common ancestor were defined as unresolved.
+To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after Jan-1-2017, and traced these nodes back to their most recent common ancestor. 
+We defined all branches ancestral to this most recent common ancestor as the trunk.
+All nodes that do not have descendants isolated after 2017 were defined as side branch nodes.
+Nodes descended from the most recent common ancestor were defined as unresolved, as it is still unclear whether they will ultimately be on the trunk or side branches.
+\comment{add something about how the JSON file with all the data is in the computer code to be posted on GitHub}
 
 \subsection*{Quantification of mutational effects and sequence preferences from an H3N2 phylogeny}
-For a given mutation a1-r-a2, we calculated the mutational effect as:
+We calculated the effects of mutations from the amino-acid preferences that were estimated from the deep mutational scanning data.
+The effect of mutating site $r$ from amino-acid $a_1$ to $a_2$ was calculated as
 \begin{equation}
-\log_2 \frac{\pi_{r,a2}}{\pi_{r,a1}}
+\log_2 \frac{\pi_{r,a_2}}{\pi_{r,a_1}}
 \end{equation}
-where $\pi_{r,a1}$ and $\pi_{r,a2}$ are the preferences for amino-acid $a1$ or $a2$ at site $r$, and the preferences are re-scaled by the stringency parameter provided in Table~\ref{tab:phydms} and averaged across replicates.
-The WSN/1933 H1 HA amino-acid preferences were also re-scaled by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}) and averaged across replicates.
-The mutational effects were also calculated for five-year windows from 1968 to 2012, and windows were incremented by one year.
+where $\pi_{r,a_1}$ and $\pi_{r,a_2}$ are the re-scaled preferences for amino acids $a_1$ or $a_2$ at site $r$ as shown in Figure~\ref{fig:logoplot}.
+The WSN/1933 H1 HA amino-acid preferences are the replicate-average values reported in \cite{doud2016accurate}, re-scaled by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
+The mutational effects were also calculated for five-year windows from 1968 to 2012.
 
-To calculate significance, we randomized the amino-acid preferences across all sites, epitope sites, or non-epitope sites, and quantified the median difference in trunk and side branch mutational effects.
-We repeated this randomization for a total of 10,000 times and calculated how frequently the difference in median mutational effect between trunk and side branches exceeded the true difference to obtain a $P$-value.
-
-Sequence preference per site for every node on the phylogenetic tree shown in Figure~\ref{fig:H3N2_phylogeny} was calculated using Equation 1 for all sites, or the set of epitope and non-epitope sites defined in~\cite{wolf2006long}.
+The average sequence preference per site for every node on the phylogenetic tree shown in Figure~\ref{fig:H3N2_phylogeny} was calculated using Equation 1 for all sites, or the set of epitope and non-epitope sites defined in~\cite{wolf2006long} and provided in \comment{suppdata that lists these}.
 
@@
 To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using MAFFT~\cite{katoh2013mafft}.
-To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2017mapping} and based off the method presented in~\cite{doud2015site}.
-Briefly, for each site we calculated half of the sum of the absolute value of the difference between preferences for each amino acid.
-The RMSD$_{\text{corrected}}$ value signifies the shift in preference.
+To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2017mapping} and use the RMSD$_{\text{corrected}}$ values as our quantification of the extent of each shift.
 
 For the plots shown in Figure~\ref{fig:RMSD_heatmap}B, any residues falling between Cys-52 and Cys-277 were defined as the head domain, and all other residues were defined as the stalk domain.
-We used a multiple sequence alignment of the HA subtype sequences provided in~\cite{doud2017quantifying} to list sites that are absolutely conserved across all subtypes.
-To find clade-specific sites, we listed sites with a single amino-acid identity in all of H1, H2, H5, and H6, and a different identity in H3, H4, and H14. 
+We used the multiple sequence alignment of the HA subtype sequences from \cite{doud2017quantifying} to identify sites that are absolutely conserved across all subtypes, or in the different clades described in Figure~\ref{fig:RMSD_heatmap}.
 }

### 2018-01-21 paper/manuscript.tex
Add post-Perth2009 seqpref supp figure
Complete edits of manuscript text

@@
 Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction}.
-While the antigenic features of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,kucharski2011influenza,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
+While the antigenic features of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
 Specifically, due to influenza virus's high mutation rate~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and lack of intra-segment recombination~\cite{boni2008homologous}, deleterious mutations become linked to beneficial ones.
@@
 It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of large numbers of mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high}.
-However, the only HA for which such measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
+However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
 Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain, A/Perth/16/2009.
@@
 We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
-We first used the deep sequencing data for each replicate estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
+We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
 Because there are 567 residues in HA, there are $567 \times 20 = 11,340$ estimated amino-acid preferences, corresponding to $567 \times 19 = 10,773$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
@@
 For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} possess high preference for cysteine.
-At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change,yang2015structure}.
-A positively charged amino acid at site 329 is important for cleavage of the HA0 precursor into the mature form~\cite{kido1992isolation, stech2005new}, and this site strongly prefers arginine.
+At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such as Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change,yang2015structure}.
+A positively charged amino acid at site 329 is important for cleavage of the HA0 precursor into the mature form~\cite{stech2005new}, and this site strongly prefers arginine.
 However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine. 
@@
 This difference in the effects of mutations between the trunk and side branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
-Some influenza sequences are derived from egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{wu2017structural,mcwhite2016sequence,skowronski2016mutations}.
+Some influenza sequences are derived from egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{mcwhite2016sequence}.
 These mutations mostly occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
@@
 Negative numbers signify mutations to less preferred amino acids, while positive numbers signify mutations to more preferred mutations.
-(B) The log$_{2}$ mutational effect for all side branch and trunk mutations (left panel), and the same data but separating side branches to terminal and internal nodes.
+(B) The log$_{2}$ mutational effect for all side branch and trunk mutations (left panel), and the same data but separating side branches to terminal and internal nodes (right panel).
 $P$-values were computed by randomizing the experimental measurements among sites 10,000 times, and determining how often the difference in median mutational effect between the trunk and side branches exceeded the actual value.
@@
 The fact that our experiments can help distinguish between trunk and side-branch lineages suggests that they might inform evolutionary forecasting.
-In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, Luksza and Lassig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
+In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, {\L}uksza and L{\"a}ssig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
 Our work suggests that deep mutational scanning can provide such data.
@@
 That this is true is demonstrated by the fact that these measurements have utility for distinguishing successful and unsuccessful viral lineages over $\sim$50 years of H3N2 evolution.
-We suspect that performing the experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures~\cite{?}) might further improve their utility.
+We suspect that performing the experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility.
 
@@
 Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}, and the extent to which this occurs is an important question in its own right.
-The fact that our measurements score trunk mutations more favorably than side branch ones across a 50-year timespan suggests that the mutational effects have not shifted to dramatically among these relatively closely related sequences.
+The fact that our measurements score trunk mutations more favorably than side branch ones across a 50-year timespan suggests that the mutational effects have not shifted too dramatically among these relatively closely related sequences.
 On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2017mapping}.
@@
 
-Our work did not characterize the antigenic effects of mutations, which also play an important role in determining strain success in nature~\cite{?}.
-Indeed, some of our results (e.g., Figure~\ref{fig:sequence_preference}) suggest as much by showing that the experiments better capture selection at non-epitope than epitope sites.
+Our work did not characterize the antigenic effects of mutations, which also play an important role in determining strain success in nature~\cite{koel2013substitutions,neher2016prediction}.
+Indeed, some of our results (\textit{e.g.}, Figure~\ref{fig:sequence_preference}) suggest as much by showing that the experiments better capture selection at non-epitope than epitope sites.
 However, our basic selection and deep-sequencing approach can be harnessed to completely map how mutations affect antibody recognition~\cite{doud2017complete,doud2017quantifying}.
@@
 Sequential 1, 2, ...
-numbering the Perth/2009 HA can be converted to H3 numbering by subtracting 16.
+numbering of the Perth/2009 HA can be converted to H3 numbering by subtracting 16 for the HA1 subunit, and subtracting 345 for the HA2 subunit.
 
@@
 We refer to the sorted bulk population as MDCK-SIAT1-TMPRSS2 cells.
-There is no selectable marker for the TMPRSS2; however, we maintain the cells at low passage number, and have seen no indication that lose their ability to support the growth of viruses with H3 HAs in the absence of exogenous trypsin.
+There is no selectable marker for the TMPRSS2; however, we maintain the cells at low passage number, and have seen no indication that they lose their ability to support the growth of viruses with H3 HAs in the absence of exogenous trypsin.
 
@@
 We initially created a virus with the HA and NA from Perth/2009 and internal genes from WSN/1933, and passaged it in cell culture to test its genetic stability.
-To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amount of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
+To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amounts of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
 The next day we changed the media to influenza growth media (abbreviated IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL --- no trypsin was added since there was TMPRSS2), and then we collected the viral supernatant at 72 hours post-transfection.
@@
 \subsection*{Generation and passaging of mutant viruses}
-The mutant virus libraries were generated using the helper-virus approach described in \cite{doud2016accurate} with several modifications, most notably changing the cell line used.
+The mutant virus libraries were generated using the helper-virus approach described in \cite{doud2016accurate} with several modifications, most notably the cell line used.
 Briefly, we transfected $5 \times 10^5$ MDCK-SIAT1-TMPRSS2 cells in suspension with 937.5 ng each of four protein expression plasmids encoding the ribonucleoprotein complex (HDM-Nan95-PA, HDM-Nan95-PB1, HDM-Nan95-PB2, and HDM-Aichi68-NP)~\cite{gong2013stability}, and 1250 ng of one of the three pICR2-mutant-HA libraries (or the wildtype control) using Lipofectamine 3000 (ThermoFisher L3000008).
-We allowed the transfected cells to adhere in 6-well plates, four hours later changed the media to D10 media.
+We allowed the transfected cells to adhere in 6-well plates and four hours later changed the media to D10 media.
 Eighteen hours after transfection, we infected the cells with the WSN/1933 HA-deficient helper virus~\cite{doud2016accurate} by preparing an inoculum of 500 TCID$_{50}$ per $\mu$L of helper virus (as computed on HA-expressing cells) in IGM, aspirating the D10 media from the cells, and adding 2 mL of the helper-virus inoculum to each well.
@@
 
-We generated the HA PCR amplicons for the three plasmid libraries, the three virus libraries, the wildtype plasmid control, and the wildtype virus control using KOD Hot Start Master Mix (EMD Millipore;71842) using the PCR reaction mixture and cycling conditions described in~\cite{bloom2014experimentally} and the P09-HA-For and P09-HA-Rev primers.
+We generated the HA PCR amplicons for the three plasmid libraries, the three virus libraries, the wildtype plasmid control, and the wildtype virus control using KOD Hot Start Master Mix (EMD Millipore 71842) using the PCR reaction mixture and cycling conditions described in~\cite{bloom2014experimentally} and the P09-HA-For and P09-HA-Rev primers.
 We prepared the sequencing libraries using a barcoded-subamplicon strategy~\cite{wu2014high} to increase the accuracy from deep sequencing. 
@@
 \subsection*{Analysis of deep sequencing data}
-We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, \comment{version ?}) to analyze the deep sequencing data.
+We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, version \texttt{2.2.0}) to analyze the deep sequencing data.
 The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
-The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Datasets \comment{suppfile}.
+The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~\ref{suppdata:prefsfiles}.
 Computer code that performs the entirety of this analysis is available at \comment{add this}
@@
 These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
-We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, \comment{?}) to fit the substitution models listed in Table~\ref{tab:phydms}.
+We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.0}) to fit the substitution models listed in Table~\ref{tab:phydms}.
 The amino-acid preferences were re-scaled by the stringency parameter using the approach described in \cite{hilton2017phydms}.
 
-The phylogenetic tree of HA subtypes in Figure~\ref{fig:distance_distribution}A is was generated as described in~\cite{doud2017quantifying}.
+The phylogenetic tree of HA subtypes in Figure~\ref{fig:distance_distribution}A was generated as described in~\cite{doud2017quantifying}.
 
@@
 All nodes that do not have descendants isolated after 2017 were defined as side branch nodes.
-Nodes descended from the most recent common ancestor were defined as unresolved, as it is still unclear whether they will ultimately be on the trunk or side branches.
-\comment{add something about how the JSON file with all the data is in the computer code to be posted on GitHub}
+Nodes descended from the most recent common ancestor were defined as unresolved, as it is still unclear whether these nodes will ultimately be on the trunk or side branches.
+The phylogenetic tree is available as a JSON file on \comment{link to repository}.
 
@@
 
-The average sequence preference per site for every node on the phylogenetic tree shown in Figure~\ref{fig:H3N2_phylogeny} was calculated using Equation 1 for all sites, or the set of epitope and non-epitope sites defined in~\cite{wolf2006long} and provided in \comment{suppdata that lists these}.
+The average sequence preference per site for every node on the phylogenetic tree shown in Figure~\ref{fig:H3N2_phylogeny} was calculated using Equation 1 for all sites, or the set of epitope and non-epitope sites defined in~\cite{wolf2006long} and provided in Dataset~\ref{suppdata:prefsfiles}.
 
@@
 To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using MAFFT~\cite{katoh2013mafft}.
-To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2017mapping} and use the RMSD$_{\text{corrected}}$ values as our quantification of the extent of each shift.
+To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2017mapping} and used the RMSD$_{\text{corrected}}$ values as our quantification of the extent of each shift.
 
@@
 The cumulative distribution of these distances is shown in the red line. 
-The blue line indicates the expected distribution if mutations in multiply mutated genes are randomly disbursed along the sequence. 
+The blue line indicates the expected distribution if mutations in multiply mutated genes are randomly dispersed along the sequence. 
 }
@@
 \begin{suppfigure}
-\comment{add figure}
+\centerline{\includegraphics[width=0.75\textwidth]{figs/S04_post_Perth2009/post_Perth2009_seqpref.pdf}}
 \caption{\label{suppfig:future}
+{\bf The sequence preference of nodes that occur after the split of Perth/2009 off the trunk.}
 Our experiments show a selective advantage for the trunk even if we limit the analysis to sequences that occur after the split of Perth/2009 off the trunk.
 (A) Phylogenetic tree showing all nodes that branch from the trunk before or contemporaneously with Perth/2009 (orange), all nodes that occur after the branch to Perth/2009 for which we can resolve the trunk (red) or side branches (blue), and all nodes for which it is not yet clear which sequences will be on the trunk or side branches.
-In this figure, we restrict the analysis to the nodes in red or blue.
+In this figure, we restrict the analysis to the nodes in red and blue.
 (B)  
@@
 \begin{suppfigure}
-\centerline{\includegraphics[width=0.6\textwidth]{figs/S04_seqpref_zoomed/seqpref_zoomed.pdf}}
+\centerline{\includegraphics[width=0.6\textwidth]{figs/S05_seqpref_zoomed/seqpref_zoomed.pdf}}
 \caption{\label{suppfig:seqpref_zoomed}
@@
 \caption{\label{suppdata:PerthHA}
-Genbank file giving the full sequence of the bidirectional reverse-genetics plasmid pHW-Perth2009-HA-G78D-T212I, which encodes the wildtype HA sequence used in this study.
+Genbank file giving the full sequence of the bidirectional reverse-genetics plasmid pHW-Perth09-HA-G78D-T212I, which encodes the wildtype HA sequence used in this study.
 }
@@
 \begin{suppdata}
-\caption{\label{suppdata:rep1_prefs}
+\caption{\label{suppdata:prefsfiles}
 Excel file giving the amino-acid preferences in sequential 1, 2, ... numbering of the Perth/2009 HA.

### 2018-01-23 paper/manuscript.tex
Change color of sialic acid receptor in mut_tolerance figure
Small tweak to dms_overview fig, update phydms table

@@
 ExpCM fit a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
-The stringency parameter for our HA ExpCM is 2.44 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments, but with greater stringency.
+The stringency parameter for our HA ExpCM is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments, but with greater stringency.
 Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
@@
 \bf{Model} & \bf{$\Delta$AIC} & \bf{LnL} & \bf{Stringency} & \bf{$\omega$}  \\ \hline
-ExpCM & 0.0 & -8439 & 2.44 & 0.91 \\
-GY94 M5 & 2166 & -9516 & -- & 0.36 (0.30, 0.84) \\
-ExpCM, site avg. & 2504 & -9691 & 0.68 & 0.32 \\
-GY94 M0 & 2608 & -9738 & -- & 0.31 \\
+ExpCM & 0.0 & -8441 & 2.47 & 0.91 \\
+GY94 M5 & 2162 & -9516 & -- & 0.36 (0.30, 0.84) \\
+ExpCM, site avg. & 2501 & -9692 & 0.67 & 0.32 \\
+GY94 M0 & 2604 & -9738 & -- & 0.31 \\
 \hline
@@
 For each HA, the structure at left shows a surface representation of the full trimer, while the structure at right side shows a ribbon representation of just one monomer.
-The sialic acid receptor is shown in black sticks.
+The sialic acid receptor is shown in red sticks.
 }

### 2018-01-23 paper/manuscript.tex
one tiny edit

@@
 Our experiments show a selective advantage for the trunk even if we limit the analysis to sequences that occur after the split of Perth/2009 off the trunk.
-(A) Phylogenetic tree showing all nodes that branch from the trunk before or contemporaneously with Perth/2009 (orange), all nodes that occur after the branch to Perth/2009 for which we can resolve the trunk (red) or side branches (blue), and all nodes for which it is not yet clear which sequences will be on the trunk or side branches.
+(A) Phylogenetic tree showing all nodes that branch from the trunk before or contemporaneously with Perth/2009 (orange), all nodes that occur after the branch to Perth/2009 for which we can resolve the trunk (red) or side branches (blue), and all nodes for which it is not yet clear which sequences will be on the trunk or side branches (gray).
 In this figure, we restrict the analysis to the nodes in red and blue.

### 2018-01-29 paper/manuscript.tex
Add author contributions and conflict of interest statement
Started adding methods for phylogenetic tree.

@@
 % Please include corresponding author, author contribution and author declaration information
-\authorcontributions{Please provide details of author contributions here.}
-\authordeclaration{Please declare any conflict of interest here.}
+\authorcontributions{J.M.L.
+performed the experiments; J.M.L.
+and J.H.
+analyzed data with input from T.B.
+and J.D.B.; J.M.L.
+and J.D.B.
+wrote the paper with input from J.H.
+and T.B.; all authors designed and planned the project}
+\authordeclaration{The authors declare no conflict of interest.}
 \equalauthors{\textsuperscript{1}J.M.L. and J.H. contributed equally to this work.}
@@
 \subsection*{Inference of human H3N2 phylogenetic tree}
-\comment{John to add methods here. We downloaded X sequences from the Influenza Virus Resource ?.... etc. subsampled how? inferred the tree, ancestral state reconstruction, visualized the tree...}
+We downloaded H3N2 HA sequences from 1968 to 2017 using a curated database of publicly available sequences from GISAID (https://github.com/nextstrain/fauna; commit 1281ed9).
+We constructed an annotated phylogenetic tree from these sequences using Nextstrain's augur tool \cite{Hadfield224048} (https://github.com/nextstrain/augur; commit 006896d) with sequences subsampled to six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
+This subsampling selected 2,145 strains of H3N2's HA.
+The tree was constructed with RAxML 8.2.10~\cite{stamatakis2006raxml}.
+Ancestral state reconstruction and branch length timing were performed with treetime~\cite{Sagulenko2018}.
+
 To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after Jan-1-2017, and traced these nodes back to their most recent common ancestor.

### 2018-01-29 paper/manuscript.tex
Clarified phylogenetic methods.

@@
 \subsection*{Inference of human H3N2 phylogenetic tree}
-We downloaded H3N2 HA sequences from 1968 to 2017 using a curated database of publicly available sequences from GISAID (https://github.com/nextstrain/fauna; commit 1281ed9).
-We constructed an annotated phylogenetic tree from these sequences using Nextstrain's augur tool \cite{Hadfield224048} (https://github.com/nextstrain/augur; commit 006896d) with sequences subsampled to six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
-This subsampling selected 2,145 strains of H3N2's HA.
-The tree was constructed with RAxML 8.2.10~\cite{stamatakis2006raxml}.
+We applied Nextstrain's augur pipeline~\cite{Hadfield224048} (https://github.com/nextstrain/augur; commit 006896d) to publicly available H3N2 HA sequences from GISAID, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
+We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
 Ancestral state reconstruction and branch length timing were performed with treetime~\cite{Sagulenko2018}.
+We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
 
-To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after Jan-1-2017, and traced these nodes back to their most recent common ancestor.
+To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after January 1, 2017, and traced these nodes back to their most recent common ancestor.
 We defined all branches ancestral to this most recent common ancestor as the trunk.

### 2018-01-29 paper/manuscript.tex
Added baltic mention.
Minor syntax and tense edits.

@@
 How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
-This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describe H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
+This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
 Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
@@
 
-Our experiments were performed on the Perth/2009 HA, but we are scoring mutations on a phylogenetic tree that extends from 1968 to 2017.
+Our experiments were performed on the Perth/2009 HA, but we scored mutations on a phylogenetic tree that extends from 1968 to 2017.
 Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
@@
 An obvious hypothesis for why the H1 deep mutational scanning is not informative about the evolutionary fate of H3 strains is that the effect of the same mutation will often be very different between these two HA subtypes.
-To determine if this is the case, we can examine how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
+To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
 Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2017mapping}.
@@
 Previous work has defined clades of structurally related HA subtypes~\cite{ha2002h5,russell2004h1}.
-One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}).
+One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}A).
 Sites that are conserved at different amino-acid identities in these two clades tend to have exceptionally large shifts in amino acid preferences (Figure~\ref{fig:RMSD_heatmap}B).
@@
 \label{sec:discussion}
-We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture, and demonstrated that these measurements are informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
+We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements are informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
 Specifically, HA sequences on the trunk of the phylogenetic tree tend to have mutations that our experiments measure to be more beneficial for viral growth than mutations that occur in side-branch HA sequences.
@@
 The phylogenetic tree is available as a JSON file on \comment{link to repository}.
+The tree was visualized using baltic (https://github.com/blab/baltic).
 

### 2018-01-29 paper/manuscript.tex
More minor syntax edits.

@@
 ExpCM fit a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
-The stringency parameter for our HA ExpCM is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments, but with greater stringency.
+The stringency parameter for our HA ExpCM is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
 Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.

### 2018-01-30 paper/manuscript.tex
Minor revisions

--- 2018-01-29 paper/manuscript.tex
@@
 \affil[a]{Basic Sciences Division}
-\affil[b]{Vaccine and Infectious Diseases Division}
+\affil[b]{Vaccine and Infectious Disease Division}
 \affil[c]{and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
@@
 For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection~\cite{neher2015nextflu,lassig2017predicting,morris2017predictive}.
-Evolutionary forecasts must ultimately distinguish between successful and unsuccessful viral lineages, which in the case of human influenza virus means distinguishing between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).
+Evolutionary forecasts may target fixation and loss of alleles in the viral population, which means distinguishing between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).
 
-To investigate whether our experiments can aid in distinguishing trunk and side branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the phylogeny in Figure~\ref{fig:H3N2_phylogeny}.
+To investigate whether our experiments can aid in distinguishing trunk and side branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the H3N2 phylogeny (Figure~\ref{fig:H3N2_phylogeny}).
 The mutations that occurred on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
@@
 \subsection*{Inference of human H3N2 phylogenetic tree}
-We applied Nextstrain's augur pipeline~\cite{Hadfield224048} (https://github.com/nextstrain/augur; commit 006896d) to publicly available H3N2 HA sequences from GISAID, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
-We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
-Ancestral state reconstruction and branch length timing were performed with treetime~\cite{Sagulenko2018}.
-We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
-
-To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after January 1, 2017, and traced these nodes back to their most recent common ancestor.
+\comment{John to add methods here. We downloaded X sequences from the Influenza Virus Resource ?.... etc. subsampled how? inferred the tree, ancestral state reconstruction, visualized the tree...}
+To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after Jan-1-2017, and traced these nodes back to their most recent common ancestor. 
 We defined all branches ancestral to this most recent common ancestor as the trunk.
@@
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
-This work was supported by grant R01AI127897 from the NIAID of the NIH. 
+This work was supported by grant R01 AI127893 from the NIAID of the NIH to JDB and TB and grant U19 AI117891 to TB.
 JML was supported in part by the Center for Inference and Dynamics of Infectious Diseases (CIDID), which is funded by grant U54GM111274 from the NIGMS of the NIH.
 The research of JDB is supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation, and a Burroughs Wellcome Young Investigator in the Pathogenesis of Infectious Diseases grant.
-\comment{Add any additional relevant grants for Trevor and John}
+TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774.
 }

### 2018-01-30 paper/manuscript.tex
Minor revision

--- 2018-01-29 paper/manuscript.tex
@@
 
-\acknow{We thank Sarah Hilton, Hugh Haddox, and Sidney Bell for helpful discussions about data analysis. 
+\acknow{We thank Sarah Hilton, Hugh Haddox, and Sidney Bell for helpful discussions about data analysis and we thank Richard Neher for sharing analysis code.
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
-This work was supported by grant R01AI127897 from the NIAID of the NIH. 
+This work was supported by grant R01 AI127893 from the NIAID of the NIH to JDB and TB and grant U19 AI117891 to TB.
 JML was supported in part by the Center for Inference and Dynamics of Infectious Diseases (CIDID), which is funded by grant U54GM111274 from the NIGMS of the NIH.
 The research of JDB is supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation, and a Burroughs Wellcome Young Investigator in the Pathogenesis of Infectious Diseases grant.
-\comment{Add any additional relevant grants for Trevor and John}
+TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774.
 }

### 2018-01-30 paper/manuscript.tex
Change 567 sites to 566 per John suggestion

@@
 We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
-Because there are 567 residues in HA, there are $567 \times 20 = 11,340$ estimated amino-acid preferences, corresponding to $567 \times 19 = 10,773$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
+Because there are 566 residues in HA, there are $566 \times 20 = 11,320$ estimated amino-acid preferences, corresponding to $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
 The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
@@
 Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites~\cite{fitch1991positive,wolf2006long,bhatt2011genomic}.
-There are various classifications of HA epitope sites; here we will use the classification of Wolf \textit{et al}~\cite{wolf2006long}, which defines 129 epitope sites among the 567 residues in HA.
+There are various classifications of HA epitope sites; here we will use the classification of Wolf \textit{et al}~\cite{wolf2006long}, which defines 129 epitope sites among the 566 residues in HA.
 In the timeframe from 1968 to 2012, the trunk of the tree in Figure~\ref{fig:H3N2_phylogeny} fixed 0.84 epitope mutations per site, but only 0.07 non-epitope mutations per site.

### 2018-01-30 paper/manuscript.tex
Fixed phydms table in manuscript

@@
 ExpCM & 0.0 & -8441 & 2.47 & 0.91 \\
-GY94 M5 & 2162 & -9516 & -- & 0.36 (0.30, 0.84) \\
+GY94 M5 & 2094 & -9482 & -- & 0.36 (0.30, 0.84) \\
 ExpCM, site avg. & 2501 & -9692 & 0.67 & 0.32 \\
-GY94 M0 & 2604 & -9738 & -- & 0.31 \\
+GY94 M0 & 2536 & -9704 & -- & 0.31 \\
 \hline

### 2018-01-30 paper/manuscript.tex
added some of John edits

--- 2018-01-30 paper/manuscript.tex
@@
 
-\title{Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fates of human H3N2 influenza virus lineages}
+\title{Deep mutational scanning of hemagglutinin helps identify evolutionarily successful human H3N2 influenza viruses}
 
@@
 The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
-However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid constraints measured in our experiments (the $\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that the deep mutational scanning captures much of the purifying selection on HA.
-ExpCM fit a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
-The stringency parameter for our HA ExpCM is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
+However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments (the $\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for by the constraints measured in the deep mutational scanning.
+The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
 Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
@@
 The results above show that measurements of mutational effects in the Perth/2009 HA are informative about the evolutionary fate of other human H3N2 HAs.
-How broadly can such measurements be generalized across HAs?
-To address this question, we repeated the trunk versus side-branch analyses using amino-acid preferences from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
+To determine how broadly experimental measurements be generalized across HAs, we repeated the trunk versus side-branch analyses using amino-acid preferences from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
 Figure~\ref{fig:WSN_trunkvssidebranch} shows that the H1 measurements are not informative for distinguishing the trunk and side branches of an H3 phylogeny.
@@
 Sites that are conserved at different amino-acid identities in these two clades tend to have exceptionally large shifts in amino acid preferences (Figure~\ref{fig:RMSD_heatmap}B).
-The clade containing H1 has an upward shift of the globular head relative to the clade containing H3.
+The clade containing H1 has an upward translation of the globular head relative to the clade containing H3.
 This structural shift has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.

### 2018-01-30 paper/manuscript.tex
addd some Trevor edits

@@
 This evolution is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects,bedford2015global}, producing a spindly phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
-Several lines of evidence indicate that the trunk lineage outcompetes other strains at least in part because it has higher fitness~\cite{strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects}.
+Several lines of evidence indicate that the trunk has higher fitness than other lineages~\cite{strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects}.
 A key challenge in the study of H3N2 evolution is to forecast which strain will dominate the upcoming influenza season, as this information can guide selection of a vaccine strain~\cite{morris2017predictive}.
@@
 For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection~\cite{neher2015nextflu,lassig2017predicting,morris2017predictive}.
-Evolutionary forecasts may target fixation and loss of alleles in the viral population, which means distinguishing between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).
+Evolutionary forecasts should be able to distinguish successful and unsuccessful viral lineages, which includes discriminating between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).
 
@@
 Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some epitope sites in H3 HA~\cite{wu2017diversity}.
-It is also known that immune pressure can increase the amount of epistasis among substitutions by selecting for combinations of functionally deleterious immune-escape mutations and counterbalancing secondary mutations~\cite{gong2014epistatically}.
+It is also known that immune pressure can increase epistasis among substitutions by selecting for combinations of functionally deleterious immune-escape mutations and counterbalancing secondary mutations~\cite{gong2014epistatically}.
 We hypothesize that immune pressure increases the accumulation of epistatic interactions involving substitutions at epitope sites.
 Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA (Figure~\ref{fig:sequence_preference}B and Figure~\ref{suppfig:seqpref_zoomed}).
-Such a trend is not apparent for non-epitope sites, probably because they are under less selection to fix adaptive mutations.
+Such a trend is not apparent for non-epitope sites, perhaps because they are under less selection to fix adaptive mutations.
 

### 2018-01-30 paper/manuscript.tex
added some Mike comments

@@
 We show that these measurements have utility for distinguishing among viral strains that do and do not succeed in nature.
-Overall, our work suggest that new high-throughput experimental approaches may be useful for evolutionary forecasting.
+Overall, our work suggests that new high-throughput experimental approaches may be useful for evolutionary forecasting.
 }
@@
 Replicate 1 exhibited the weakest correlation with the other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
-The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity after the reverse-genetics step contributes most of the experimental noise.
+The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity during viral passage contributes most of the experimental noise.
 
@@
 We do not think that this result means that Perth/2009 is actually higher fitness than the other strains.
-Rather, if there are modest shifts in the effects of mutations as HA evolves, then sequences more distant from Perth/2009 could be scored as having lower preference simply because these shifts start to degrade the accuracy of our measurements.
+Rather, if there are modest shifts in the effects of mutations as HA evolves, then sequences more distant from Perth/2009 could be scored as having lower preference simply because these shifts start to degrade the applicability of our measurements.
 But importantly, our experiments consistently show the trunk to be more favorable than the side branches regardless of whether we look at sequences that precede Perth/2009 (most of Figure~\ref{fig:trunkvssidebranch}C) or sequences that occur after Perth/2009 has diverged from the trunk (Figure~\ref{suppfig:future}).
@@
 Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and at sites that are also under immune pressure.
-The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that most epitope sites are still important for HA function~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.
+The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that mutations at epitope sites still have significant non-antigenic functional consequences~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.
 
@@
 They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2017mapping}.
-However, the shifts are still smaller than those observed when comparing HA to the non-homologous HIV Env protein. 
+However, the typical shift between H3 and H1 is still smaller than that observed when comparing HA to the non-homologous HIV Env protein. 
 Therefore, there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations remain more similar than for non-homologous proteins.
@@
 The shifts are also much larger than those previously observed between two variants of HIV envelope protein (Env) that share 86\% amino-acid identity (pink).
-However, the shifts between H3 and H1 are still less than the differences between either HA and HIV Env (green).
+However, the shifts between H3 and H1 are usually less than the differences between either HA and HIV Env (green).
 }
@@
 Our work did not characterize the antigenic effects of mutations, which also play an important role in determining strain success in nature~\cite{koel2013substitutions,neher2016prediction}.
-Indeed, some of our results (\textit{e.g.}, Figure~\ref{fig:sequence_preference}) suggest as much by showing that the experiments better capture selection at non-epitope than epitope sites.
+%Indeed, some of our results (\textit{e.g.}, Figure~\ref{fig:sequence_preference}) suggest as much by showing that the experiments better capture selection at non-epitope than epitope sites.
 However, our basic selection and deep-sequencing approach can be harnessed to completely map how mutations affect antibody recognition~\cite{doud2017complete,doud2017quantifying}.
-But so far, experiments using this approach have either failed to examine antibodies or sera that are relevant to driving the evolution of H3N2 influenza~\cite{doud2017complete,doud2017quantifying}, or have used relevant sera but examined a non-comprehensive set of mutations~\cite{li2016selection}.
+But so far, experiments using this approach have not examined antibodies or sera that are relevant to driving the evolution of H3N2 influenza~\cite{doud2017complete,doud2017quantifying}, or have used relevant sera but examined a non-comprehensive set of mutations~\cite{li2016selection}.
 Future experiments that completely map how HA mutations affect recognition by human sera seem likely to be especially fruitful for informing viral forecasting.

### 2018-02-01 paper/manuscript.tex
Add Katie edit
Add mut effect plot to supp post-Perth2009 figure

@@
 The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656) and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry to create plasmid pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
-We used the lentiviral vector to transduce MDCK-SIAT1 cells, and sorted an intermediate mCherry population by flow cytometry.
+We used the lentiviral vector to transduce MDCK-SIAT1 cells, and sorted an intermediate mCherry-positive population by flow cytometry.
 We refer to the sorted bulk population as MDCK-SIAT1-TMPRSS2 cells.
@@
 The sequence preferences of nodes on the trunk generally exceed those of nodes on the side branches along the post-Perth/2009 portion of the tree.
-Note that because there are relatively few post-Perth/2009 sequences for which we can currently distinguish trunk from side branch, the number of points in this plot is small, precluding meaningful statistical analyses of the type performed in Figure~\ref{fig:trunkvssidebranch}B.
-}
-\end{suppfigure}
-
-\begin{suppfigure}
-\centerline{\includegraphics[width=0.6\textwidth]{figs/S05_seqpref_zoomed/seqpref_zoomed.pdf}}
-\caption{\label{suppfig:seqpref_zoomed}
-{\bf The per-site sequence preference at epitope and non-epitope sites.}
-The per-site sequence preferences shown in Figure~\ref{fig:sequence_preference}B, but on separate y-axes.
+(D)
+Post-Perth/2009 trunk mutations on average tend to be more favorable than side branch mutations.
+Note that there are relatively few post-Perth/2009 mutations (13 trunk mutations and 324 side branch mutations).
 }

### 2018-02-01 paper/manuscript.tex
Clean up manuscript text

@@
 We also see high mutational tolerance in many of the known antigenic regions of H3 HA~\cite{wiley1981structural}.
-For instance, in recent H3N2 strains, antigenic region B is immunodominant, and most recent major antigenic drift mutations have occurred in this region~\cite{chambers2015identification,koel2013substitutions,popova2012immunodominance}.
+For instance, antigenic region B is an immunodominant area, and many recent major antigenic drift mutations have occurred in this region~\cite{chambers2015identification,koel2013substitutions,popova2012immunodominance}.
 We find that the most distal portion of the globular head near the 190-helix, which is part of antigenic region B, is highly tolerant of mutations (Figure~\ref{fig:mut_tolerance}).
@@
 However, there are differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
-The per-site sequence preference at epitope sites increases over time, whereas at non-epitope sites it remains relatively constant  (Figure~\ref{fig:sequence_preference}B and Figure~\ref{suppfig:seqpref_zoomed}).
-In addition, the per-site sequence preference is consistently higher at non-epitope than epitope sites, perhaps because immune pressure sometimes drives the fixation of functionally less favorable mutations at epitope sites.
+The per-site sequence preference at epitope sites increases over time, whereas at non-epitope sites it remains relatively constant  (Figure~\ref{fig:sequence_preference}B).
+%In addition, the per-site sequence preference is consistently higher at non-epitope than epitope sites, perhaps because immune pressure sometimes drives the fixation of functionally less favorable mutations at epitope sites.
 The fact that only epitope sites exhibit an increase in sequence preference over time can also be rationalized in terms of prior knowledge about protein evolution.
@@
 We hypothesize that immune pressure increases the accumulation of epistatic interactions involving substitutions at epitope sites.
-Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA (Figure~\ref{fig:sequence_preference}B and Figure~\ref{suppfig:seqpref_zoomed}).
+Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA (Figure~\ref{fig:sequence_preference}B).
 Such a trend is not apparent for non-epitope sites, perhaps because they are under less selection to fix adaptive mutations.
@@
 The average per-site sequence preference for all nodes in the phylogenetic tree, calculated separately for each set of sites.
-The sequence preferences for epitope and non-epitope sites on separate y-axes is shown in Figure~\ref{suppfig:seqpref_zoomed}.
 }
@@
 To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amounts of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
-The next day we changed the media to influenza growth media (abbreviated IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL --- no trypsin was added since there was TMPRSS2), and then we collected the viral supernatant at 72 hours post-transfection.
+The next day we changed the media to influenza growth media (IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL --- no trypsin was added since there was TMPRSS2), and then we collected the viral supernatant at 72 hours post-transfection.
 This viral supernatant was then blind passaged in MDCK-SIAT1-TMPRSS2 a total of six additional times.
-We then isolated viral RNA from these passaged viruses and sequenced the HA gene.
+We isolated viral RNA from these passaged viruses and sequenced the HA gene.
 The passaged HA had two mutations, G78D and T212I, which we found enhanced viral growth as shown in Figure~\ref{suppfig:Perth2009_mut}.
@@
 \subsection*{Inference of human H3N2 phylogenetic tree}
-We applied Nextstrain's augur pipeline~\cite{Hadfield224048} (https://github.com/nextstrain/augur; commit 006896d) to publicly available H3N2 HA sequences from GISAID, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
+We applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
 We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
@@
 The phylogenetic tree is available as a JSON file on \comment{link to repository}.
-The tree was visualized using baltic (https://github.com/blab/baltic).
+The tree was visualized using baltic (\url{https://github.com/blab/baltic}).
 

### 2018-02-01 paper/manuscript.tex
fixed hyphenation of *side branch*

@@
 Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain, A/Perth/16/2009.
-We then examine whether these measurements are useful for distinguishing between H3N2 trunk and side branch lineages, and find that mutations on the trunk are measured to be more favorable than those on the side branches, both at epitope and non-epitope sites.
+We then examine whether these measurements are useful for distinguishing between H3N2 trunk and side-branch lineages, and find that mutations on the trunk are measured to be more favorable than those on the side branches, both at epitope and non-epitope sites.
 However, the experiments are useful for assessing the effects of mutations only in closely related viral strains, since we show that mutations often have disparate effects on H3 and H1 HAs.
@@
 
-To investigate whether our experiments can aid in distinguishing trunk and side branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the H3N2 phylogeny (Figure~\ref{fig:H3N2_phylogeny}).
+To investigate whether our experiments can aid in distinguishing trunk and side-branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the H3N2 phylogeny (Figure~\ref{fig:H3N2_phylogeny}).
 The mutations that occurred on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
-This difference in the effects of mutations between the trunk and side branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
+This difference in the effects of mutations between the trunk and side-branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
 Some influenza sequences are derived from egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{mcwhite2016sequence}.
@@
 where $\pi_{r, s_r}$ is the preference measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid $s_r$ at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
-Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side branch mutations.
+Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side-branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side-branch mutations.
 However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
@@
 {\bf Mutations on the trunk are more favorable than those on the side branches according to our experimental measurements.}
-(A) We used our experiments to calculate the log$_{2}$ mutational effect for all trunk and side branch mutations in five-year windows from 1968 to 2013. 
+(A) We used our experiments to calculate the log$_{2}$ mutational effect for all trunk and side-branch mutations in five-year windows from 1968 to 2013. 
 The central year in each window is denoted on the x-axis.
@@
 (C) The average per-site sequence preference of every node in the phylogenetic tree in Fig.~\ref{fig:H3N2_phylogeny}.
-Trunk nodes are in red, side branch nodes in blue, and Perth/2009 is marked with a yellow star.
+Trunk nodes are in red, side-branch nodes in blue, and Perth/2009 is marked with a yellow star.
 More preferred sequences have larger values.
@@
 We partitioned HA into epitope and non-epitope sites~\cite{wolf2006long}.
-(A) The log$_{2}$ mutational effects for side branch and trunk mutations at non-epitope (left) and epitope (right) sites.
+(A) The log$_{2}$ mutational effects for side-branch and trunk mutations at non-epitope (left) and epitope (right) sites.
 $P$-values were computed as in Figure~\ref{fig:trunkvssidebranch} but only performing the randomizations over the appropriate set of sites (non-epitope or epitope).
@@
 
-To parse trunk mutations from side branch mutations, we defined a set of recent nodes sampled on or after January 1, 2017, and traced these nodes back to their most recent common ancestor.
+To parse trunk mutations from side-branch mutations, we defined a set of recent nodes sampled on or after January 1, 2017, and traced these nodes back to their most recent common ancestor.
 We defined all branches ancestral to this most recent common ancestor as the trunk.
-All nodes that do not have descendants isolated after 2017 were defined as side branch nodes.
+All nodes that do not have descendants isolated after 2017 were defined as side-branch nodes.
 Nodes descended from the most recent common ancestor were defined as unresolved, as it is still unclear whether these nodes will ultimately be on the trunk or side branches.
@@
 (D)
-Post-Perth/2009 trunk mutations on average tend to be more favorable than side branch mutations.
-Note that there are relatively few post-Perth/2009 mutations (13 trunk mutations and 324 side branch mutations).
+Post-Perth/2009 trunk mutations on average tend to be more favorable than side-branch mutations.
+Note that there are relatively few post-Perth/2009 mutations (13 trunk mutations and 324 side-branch mutations).
 }

### 2018-02-01 paper/manuscript.tex
updated author contributions
final revisions

@@
 % Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[a,d,e,1]{Juhye M. Lee}
-\author[b,f,1]{John Huddleston} 
+\author[a,d,e]{Juhye M. Lee}
+\author[b,f]{John Huddleston} 
 \author[a,d,e]{Michael B. Doud}
 \author[a,f]{Kathryn A. Hooper}
-\author[b,c,2]{Trevor Bedford}
-\author[a,c,d,2]{Jesse D. Bloom}
+\author[b,c,1]{Trevor Bedford}
+\author[a,c,d,1]{Jesse D. Bloom}
 
@@
 analyzed data with input from T.B.
-and J.D.B.; J.M.L.
+and J.D.B.; M.B.D.
+contributed essential reagents and expertise; K.A.H.
+contributed essential reagents; J.M.L.
 and J.D.B.
@@
 \authordeclaration{The authors declare no conflict of interest.}
-\equalauthors{\textsuperscript{1}J.M.L. and J.H. contributed equally to this work.}
-\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed: trevor@bedford.io, jbloom@fredhutch.org}
+\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed: trevor@bedford.io, jbloom@fredhutch.org}
 
@@
 In principle, experiments could be informative for identifying how mutations affect these features.
-Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction}.
+Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction,koel2013substitutions,chambers2015identification,li2016selection}.
 While the antigenic features of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
@@
 The resulting accumulation of deleterious mutations can affect non-antigenic properties that are central to viral fitness~\cite{luksza2014predictive}.
-However, while extensive work has gone into characterizing how mutations affect HA antigenicity~\cite{laver1979antigenic, webster1980determination,wiley1981structural,koel2013substitutions,chambers2015identification,li2016selection}, we lack any large-scale quantitative characterization of how mutations to H3N2 HA affect non-antigenic properties crucial to viral fitness.
+However, there are no large-scale quantitative characterizations of how mutations to H3N2 HA affect non-antigenic properties crucial to viral fitness.
 
@@
 However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
-Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain, A/Perth/16/2009.
+Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
 We then examine whether these measurements are useful for distinguishing between H3N2 trunk and side-branch lineages, and find that mutations on the trunk are measured to be more favorable than those on the side branches, both at epitope and non-epitope sites.
@@
 We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
-Because there are 566 residues in HA, there are $566 \times 20 = 11,320$ estimated amino-acid preferences, corresponding to $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
+Because there are 566 residues in HA, there are $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
 The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
-The biological replicates are fairly well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
-Replicate 1 exhibited the weakest correlation with the other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
+The biological replicates were well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
+Replicate 1 exhibited the weakest correlation with other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
 The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity during viral passage contributes most of the experimental noise.
@@
 The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
-However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments (the $\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for by the constraints measured in the deep mutational scanning.
+However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for by the constraints measured in the deep mutational scanning.
 The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
@@
 
-A closer examination of Figure~\ref{fig:logoplot} reveals that the experimentally measured amino-acid preferences generally agree with existing knowledge about HA's structure and function.
+Examination of Figure~\ref{fig:logoplot} reveals that the experimentally measured amino-acid preferences generally agree with existing knowledge about HA's structure and function.
 For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} possess high preference for cysteine.
@@
 We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we quantify as the Shannon entropy of the re-scaled preferences.
-In prior mutational studies of the WSN/1933 H1 HA, the stalk domain was found to be substantially less mutationally tolerant than the globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
+In prior mutational studies of H1 HAs, the stalk domain was found to be substantially less mutationally tolerant than the globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,heaton2013genome}.
 
@@
 For instance, in the H3 HA, the short helix A in the stalk is very mutationally tolerant.
-Interestingly, there are more reports of readily selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2017quantifying,anderson2017natural} HAs.
+Interestingly, there are more reports of readily selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.
 
@@
 To investigate whether our experiments can aid in distinguishing trunk and side-branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the H3N2 phylogeny (Figure~\ref{fig:H3N2_phylogeny}).
-The mutations that occurred on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
+The mutations on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
 This difference in the effects of mutations between the trunk and side-branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
@@
 
-Mutations that occur on the trunk are scored as significantly more favorable in our experiments than side-branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}A).
+Mutations on the trunk are scored as significantly more favorable in our experiments than side-branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}A).
 Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and at sites that are also under immune pressure.
-The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that mutations at epitope sites still have significant non-antigenic functional consequences~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.
+The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that mutations at these sites still have non-antigenic functional consequences~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.
 
@@
 The per-site sequence preference at epitope sites increases over time, whereas at non-epitope sites it remains relatively constant  (Figure~\ref{fig:sequence_preference}B).
-%In addition, the per-site sequence preference is consistently higher at non-epitope than epitope sites, perhaps because immune pressure sometimes drives the fixation of functionally less favorable mutations at epitope sites.
-The fact that only epitope sites exhibit an increase in sequence preference over time can also be rationalized in terms of prior knowledge about protein evolution.
+This difference is unlikely to be explained purely by differences in the number of mutations at these two sets of sites: there have been $\sim$4-fold more total trunk mutations at epitope than non-epitope sites (109 versus 28), but the large increase in sequence preference for the epitope sites is $\gg$4-fold larger than the nearly imperceptible increase for non-epitope sites.
+The fact that only epitope sites exhibit an increase in sequence preference over time can be rationalized in terms of prior knowledge about protein evolution.
 As proteins evolve, there can be epistatic shifts in the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
 Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some epitope sites in H3 HA~\cite{wu2017diversity}.
-It is also known that immune pressure can increase epistasis among substitutions by selecting for combinations of functionally deleterious immune-escape mutations and counterbalancing secondary mutations~\cite{gong2014epistatically}.
+It is also known that immune pressure can increase epistasis among substitutions~\cite{gong2014epistatically}.
 We hypothesize that immune pressure increases the accumulation of epistatic interactions involving substitutions at epitope sites.
@@
 
-An obvious hypothesis for why the H1 deep mutational scanning is not informative about the evolutionary fate of H3 strains is that the effect of the same mutation will often be very different between these two HA subtypes.
+An obvious hypothesis for why the H1 deep mutational scanning is not informative about the evolutionary fate of H3 strains is that the effect of the same mutation is often different between these two HA subtypes.
 To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
@@
 The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
-The shifts are also much larger than those previously observed between two variants of HIV envelope protein (Env) that share 86\% amino-acid identity (pink).
-However, the shifts between H3 and H1 are usually less than the differences between either HA and HIV Env (green).
+The shifts are also much larger than those previously observed between two variants of HIV Env that share 86\% amino-acid identity (pink).
+However, the shifts between H3 and H1 are usually less than the differences between HA and HIV Env (green).
 }
@@
 However, at the domain level, sites in HA's stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B). 
-The HA stalk domain is also more conserved in sequence~\cite{nobusawa1991comparison,hai2012influenza,mallajosyula2014influenza}, suggesting that conservation of amino-acid sequence tends to be correlated with conservation of amino-acid preferences.
+The HA stalk domain is also more conserved in sequence~\cite{nobusawa1991comparison}, suggesting that conservation of amino-acid sequence is correlated with conservation of amino-acid preferences.
 Consistent with this idea, sites that are absolutely conserved across all 18 HA subtypes are significantly less shifted than sites that are variable across HA subtypes (Figure~\ref{fig:RMSD_heatmap}B).
@@
 In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations.
-Therefore, some of the shifts in HA amino-acid preferences can be directly rationalized in terms of changes in HA structure.
+Therefore, some of the shifts in HA amino-acid preferences can be rationalized in terms of changes in HA structure.
 
@@
 \label{sec:discussion}
-We have measured the effect of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements are informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
+We have measured the effects of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements are informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
 Specifically, HA sequences on the trunk of the phylogenetic tree tend to have mutations that our experiments measure to be more beneficial for viral growth than mutations that occur in side-branch HA sequences.
@@
 Indeed, a vast amount of work in virology has chronicled the many ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}.
-Mutations in viral genes other than HA are also certainly important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
-Given these caveats, it might seem surprising that simply measuring viral growth in cell culture can be informative about the success of viral strains in nature.
-Yet, prior to our work, there were no comprehensive studies of the functional effects of mutations to H3 HA on any property that even resembled viral fitness in nature, and modeling work has either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or simply assumed that all non-epitope mutations had equivalent deleterious effects~\cite{luksza2014predictive}.
+Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
+Given these caveats, it might seem surprising that measuring viral growth in cell culture can be informative about the success of viral strains in nature.
+Yet, prior to our work, there were no comprehensive studies of the functional effects of mutations to H3 HA on any property that even resembled viral fitness in nature, and modeling work has either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or assumed that all non-epitope mutations had equivalent deleterious effects~\cite{luksza2014predictive}.
 The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative --- and represent a vast improvement over no information at all.
@@
 %Indeed, some of our results (\textit{e.g.}, Figure~\ref{fig:sequence_preference}) suggest as much by showing that the experiments better capture selection at non-epitope than epitope sites.
-However, our basic selection and deep-sequencing approach can be harnessed to completely map how mutations affect antibody recognition~\cite{doud2017complete,doud2017quantifying}.
-But so far, experiments using this approach have not examined antibodies or sera that are relevant to driving the evolution of H3N2 influenza~\cite{doud2017complete,doud2017quantifying}, or have used relevant sera but examined a non-comprehensive set of mutations~\cite{li2016selection}.
+However, our basic selection and deep-sequencing approach can be harnessed to completely map how mutations affect antibody recognition~\cite{doud2017complete,doud2018quantifying}.
+But so far, experiments using this approach have not examined antibodies or sera that are relevant to driving the evolution of H3N2 influenza~\cite{doud2017complete,doud2018quantifying}, or have used relevant sera but examined a non-comprehensive set of mutations~\cite{li2016selection}.
 Future experiments that completely map how HA mutations affect recognition by human sera seem likely to be especially fruitful for informing viral forecasting.
@@
 Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610. 
-Computer code used to analyze the data and produce the results in the paper are in \comment{link to repository}.
+Computer code used to analyze the data and produce the results in the paper are on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
 
@@
 We created three independent libraries, one for each biological replicate.
-The mutant variants were then cloned at high efficiency into the pICR2 vector using digestion with BsmBI, ligation with T4 DNA ligase, and electroporation into ElectroMAX DH10B competent cells (Invitrogen 18290015).
+The mutant variants were then cloned at high efficiency into the pICR2~\cite{ashenberg2017deep} vector using digestion with BsmBI, ligation with T4 DNA ligase, and electroporation into ElectroMAX DH10B competent cells (Invitrogen 18290015).
 We obtained $>$6 million transformants for each replicate.
@@
 The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~\ref{suppdata:prefsfiles}.
-Computer code that performs the entirety of this analysis is available at \comment{add this}
+Computer code that performs the entirety of this analysis is available on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
 
@@
 These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
-We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.0}) to fit the substitution models listed in Table~\ref{tab:phydms}.
+We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.1}) to fit the substitution models listed in Table~\ref{tab:phydms}.
 The amino-acid preferences were re-scaled by the stringency parameter using the approach described in \cite{hilton2017phydms}.
 
-The phylogenetic tree of HA subtypes in Figure~\ref{fig:distance_distribution}A was generated as described in~\cite{doud2017quantifying}.
+The phylogenetic tree of HA subtypes in Figure~\ref{fig:distance_distribution}A was generated as described in~\cite{doud2018quantifying}.
 
@@
 Nodes descended from the most recent common ancestor were defined as unresolved, as it is still unclear whether these nodes will ultimately be on the trunk or side branches.
-The phylogenetic tree is available as a JSON file on \comment{link to repository}.
+The phylogenetic tree is available as a JSON file on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
 The tree was visualized using baltic (\url{https://github.com/blab/baltic}).
@@
 For the plots shown in Figure~\ref{fig:RMSD_heatmap}B, any residues falling between Cys-52 and Cys-277 were defined as the head domain, and all other residues were defined as the stalk domain.
-We used the multiple sequence alignment of the HA subtype sequences from \cite{doud2017quantifying} to identify sites that are absolutely conserved across all subtypes, or in the different clades described in Figure~\ref{fig:RMSD_heatmap}.
+We used the multiple sequence alignment of the HA subtype sequences from \cite{doud2018quantifying} to identify sites that are absolutely conserved across all subtypes, or in the different clades described in Figure~\ref{fig:RMSD_heatmap}.
+
 }
@@
 
+\clearpage
 \subsection*{Supporting Information (SI)}
-\FloatBarrier
+\clearpage
 
-The main text of the paper must stand on its own without the SI.
-Refer to SI in the manuscript at an appropriate point in the text.
-Number supporting figures and tables starting with S1, S2, etc.
-Authors are limited to no more than 10 SI files, not including movie files.
-Authors who place detailed materials and methods in SI must provide sufficient detail in the main text methods to enable a reader to follow the logic of the procedures and results and also must reference the online methods.
-If a paper is fundamentally a study of a new method or technique, then the methods must be described completely in the main text.
-Because PNAS edits SI and composes it into a single PDF, authors must provide the following file formats only.
-
-\subsubsection*{SI Text}
-
-Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference).
-
-\subsubsection*{SI Figures}
-
-\begin{suppfigure}
+\begin{suppfigure}[H]
 \centerline{\includegraphics[width=0.5\textwidth]{figs/S01_G78D-T212I/G78D-T212I.pdf}}
@@
 
-\begin{suppfigure}
+\begin{suppfigure}[H]
 \centerline{\includegraphics[width=0.6\textwidth]{figs/S02_SangerSeq/SangerSeq.pdf}}
@@
 
-\begin{suppfigure}
+\begin{suppfigure}[H]
 \centerline{\includegraphics[width=\textwidth]{figs/S03_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
 \caption{\label{suppfig:WSNprefs_logoplot}
-{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.} 
+{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA as measured in \cite{doud2016accurate}.} 
 The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
@@
 
-\begin{suppfigure}
+\begin{suppfigure}[H]
 \centerline{\includegraphics[width=0.75\textwidth]{figs/S04_post_Perth2009/post_Perth2009_seqpref.pdf}}
@@
 Post-Perth/2009 trunk mutations on average tend to be more favorable than side-branch mutations.
-Note that there are relatively few post-Perth/2009 mutations (13 trunk mutations and 324 side-branch mutations).
+This effect is of similar magnitude to pre-Perth/2009 mutations (see Figure~\ref{fig:sequence_preference}) but is not statistically significant because there are relatively few post-Perth/2009 mutations (13 trunk mutations and 324 side-branch mutations).
 }
@@
 
-\clearpage
-
-\subsubsection*{SI Datasets} 
-\FloatBarrier
-
-\begin{suppdata}
+\begin{suppdata}[H]
 \caption{\label{suppdata:PerthHA}
@@
 
-\begin{suppdata}
+\begin{suppdata}[H]
 \caption{\label{suppdata:bcsubamp_primers}
@@
 
-\begin{suppdata}
+\begin{suppdata}[H]
 \caption{\label{suppdata:prefsfiles}

### 2018-03-18 paper/manuscript.tex
added Nicholas Wu as co-author for his key intellectual contributions
Incorporate new mut tolerance figures into manuscript

@@
 \author[a,f]{Kathryn A. Hooper}
+\author[g]{Nicholas C. Wu}
 \author[b,c,1]{Trevor Bedford}
@@
 \affil[f]{and Molecular and Cellular Biology Program, University of Washington, Seattle, WA 98195, USA}
+\affil[g]{Department of Integrative Structural and Computational Biology, The Scripps Research Institute, La Jolla, CA  92037}
 
@@
 contributed essential reagents and expertise; K.A.H.
-contributed essential reagents; J.M.L.
+contributed essential reagents; N.C.W.
+made key intellectual contributions; J.M.L.
 and J.D.B.
 wrote the paper with input from J.H.
-and T.B.; all authors designed and planned the project}
+and T.B.; all authors were involved in project design and revising the paper}
 \authordeclaration{The authors declare no conflict of interest.}
@@
 Surprisingly, there was much less contrast in mutational tolerance between the stalk and head domains for the H3 HA than for the H1 (Figure~\ref{fig:mut_tolerance}).
-For instance, in the H3 HA, the short helix A in the stalk is very mutationally tolerant.
+Specifically, for solvent-exposed residues, the head domain is more mutationally tolerant than the stalk domain for the WSN/1933 H1 HA, whereas such a difference in tolerances was not seen for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B), and this was true using different relative solvent accessibility cutoffs to define solvent-exposed residues (Figure~\ref{suppfig:head_stalk_RSA}).
+Furthermore, in the H3 HA, the short helix A in the stalk was found to be very mutationally tolerant.
 Interestingly, there are more reports of readily selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.
@@
 %\begin{SCfigure*}[\sidecaptionrelwidth][t]
-\begin{figure}
+\begin{figure*}[ht]
 \centering
-%\includegraphics[width=11.4cm]{figs/mut_tolerance/entropy_heatmap.pdf}
-\includegraphics[width=\columnwidth]{figs/mut_tolerance/entropy_heatmap.pdf}
+\includegraphics[width=\textwidth]{figs/mut_tolerance/mut_tolerance_new.pdf}
 \caption{\label{fig:mut_tolerance}
 {\bf Mutational tolerance of each site in H3 and H1 HAs.}
-Mutational tolerance as measured in the current study is mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}).
+(A) Mutational tolerance as measured in the current study is mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}).
 Mutational tolerance of the WSN/1933 H1 HA as measured in \cite{doud2016accurate} is mapped onto the structure of the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}).
@@
 The sialic acid receptor is shown in red sticks.
+(B) The mutational tolerance (Shannon entropy of re-scaled preferences) of solvent-exposed residues with a relative solvent accessibility $\geq$ 0.2 in the head and stalk domains of the Perth/2009 H3 HA (purple) and WSN/1933 H1 HA (gold).
+We quantified the absolute solvent accessibility of each site using DSSP~\cite{kabsch1983} and then normalized to a relative solvent accessibility using the absolute accessibilities in~\cite{tien2013}.
+Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
+Note that the mutational tolerance values are not comparable between the two HAs.
 }
 %\end{SCfigure*}
-\end{figure}
+\end{figure*}
 
@@
 
+\begin{suppfigure}[H]
+\centerline{\includegraphics[width=0.75\textwidth]{figs/supp_head_stalk_RSA/head_stalk_RSA.pdf}}
+\caption{\label{suppfig:head_stalk_RSA}
+{\bf Mutational tolerances of the head and stalk domains at various RSA cutoffs.}
+The mutational tolerances of the head and stalk domains show less disparity for the Perth/2009 H3 HA compared to those for the WSN/1933 H1 HA. 
+We used relative solvent accessibility (RSA) cutoffs of $0.1$, $0.2$, and $0.3$ to define solvent-exposed residues and plotted the mutational tolerances (Shannon entropy of re-scaled preferences) of these residues in the head and stalk domains for the Perth/2009 H3 HA (purple) and the WSN/1933 H1 HA (brown).
+Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
+The HA structures color the residues that are defined as solvent exposed at a given RSA cutoff. 
+One monomer is shown in surface representation and another monomer shown in ribbon representation. 
+Residues in lighter shades of purple or brown are in the head domain, while residues in darker shades are in the stalk domain.
+Note that the mutational tolerance values are not comparable between the two HAs.
+}
+\end{suppfigure}
+
 \begin{suppdata}[H]

### 2018-03-24 paper/manuscript.tex
Start re-write of Results max freq section

@@
 Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping,bhatt2011genomic}.
-This evolution is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects,bedford2015global}, producing a spindly phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
-Several lines of evidence indicate that the trunk has higher fitness than other lineages~\cite{strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects}.
-A key challenge in the study of H3N2 evolution is to forecast which strain will dominate the upcoming influenza season, as this information can guide selection of a vaccine strain~\cite{morris2017predictive}.
-In order to make these forecasts, it is important to identify the features that enable the trunk strain to persist as other strains die out.
+This evolution is further characterized by competition and turnover among groups of strains or clades.
+As a result, mutations that differentiate clades exhibit dynamic frequency trajectories as these clades appear, circulate, or die out.
+Several lines of evidence indicate that successful clades, which reach high maximum frequencies or even fix in the global population, have higher fitness than clades that remain at low frequency.
+A key goal in the study of H3N2 evolution is to identify the features that enable certain clades to succeed as others die out.
+% This evolution is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects,bedford2015global}, producing a spindly phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
+% Several lines of evidence indicate that the trunk has higher fitness than other lineages~\cite{strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects}.
+% A key challenge in the study of H3N2 evolution is to forecast which strain will dominate the upcoming influenza season, as this information can guide selection of a vaccine strain~\cite{morris2017predictive}.
+% In order to make these forecasts, it is important to identify the features that enable the trunk strain to persist as other strains die out.
 
@@
 
-Two main characteristics distinguish the evolutionarily successful trunk from its competitors: greater antigenic change, and efficient viral growth and transmission.
+Two main characteristics distinguish evolutionarily successful clades from their competitors: greater antigenic change, and efficient viral growth and transmission.
 In principle, experiments could be informative for identifying how mutations affect these features.
@@
 Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
-We then examine whether these measurements are useful for distinguishing between H3N2 trunk and side-branch lineages, and find that mutations on the trunk are measured to be more favorable than those on the side branches, both at epitope and non-epitope sites.
-However, the experiments are useful for assessing the effects of mutations only in closely related viral strains, since we show that mutations often have disparate effects on H3 and H1 HAs.
-Our work highlights the potential for using high-throughput experimental measurements of mutational effects to inform evolutionary forecasting of human seasonal influenza virus.
+We then apply these measurements to examine differences in maximum clade frequencies.
+%We then examine whether these measurements are useful for distinguishing between H3N2 trunk and side-branch lineages, and find that mutations on the trunk are measured to be more favorable than those on the side branches, both at epitope and non-epitope sites.
+%However, the experiments are useful for assessing the effects of mutations only in closely related viral strains, since we show that mutations often have disparate effects on H3 and H1 HAs.
+%Our work highlights the potential for using high-throughput experimental measurements of mutational effects to inform evolutionary forecasting of human seasonal influenza virus.
 
@@
 
-\subsection*{The experimental measurements can help discriminate between successful and unsuccessful influenza virus lineages}
-A major goal in the study of rapidly evolving viruses such as influenza is to forecast evolutionary trajectories~\cite{lassig2017predicting,morris2017predictive}.
-For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection~\cite{neher2015nextflu,lassig2017predicting,morris2017predictive}.
-Evolutionary forecasts should be able to distinguish successful and unsuccessful viral lineages, which includes discriminating between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).
+\subsection*{Our measurements reveal differences between clades of low and high maximum frequencies}
+The maximum frequencies achieved by particular mutations are a reflection of the fitnesses of strains carrying those mutations.
+Do the experimentally-determined preferences therefore reveal underlying differences in strains that reach various maximum frequencies?
+To explore this question, we examined the relationship between the experimentally measured effects of all mutations and the maximum frequency achieved by all nodes with at least one tip in a maximum-likelihood H3N2 phylogeny.
+We estimated the frequency trajectories for every node in the tree by... \comment {I'll let John / Trevor add more here}.
+Evolutionarily successful nodes that occurred \textit{before} the Perth/2009 strain diverged from the trunk of the phylogenetic tree tend to carry mutations that move towards the Perth/2009 sequence. 
+Because the effects of mutations may vary depending on genetic background, we analyzed nodes that preceded the Perth/2009 strain separately from nodes that occurred after the split of Perth/2009 from the trunk of the phylogenetic tree \comment{need better explanation of pre- and post-Perth/2009 rationale here since it seems like we are going to completely drop the trunk/tip analysis from the paper?
+Also this is where we would reference the tree supplemental figure}.
+We found that mutations measured to be very deleterious to viral growth in our experiments were not observed at high frequencies in nature.
+Rather, only mutations with beneficial or small deleterious effects in our experiments appeared to reach high frequencies or even fix in nature, as evidenced by the slight but significant\comment{?} trend towards higher frequencies with more favorable mutations (Figure ~\ref{fig:muteffect_maxfreq}).
+Furthermore, shifts toward more favorable mutations with increasing maximum frequency were also apparent from plotting the distribution of mutational effects for different frequency bins.
+Mutations were on average scored as being more detrimental in the lower frequency bins than at higher frequencies.
+The trend towards more beneficial mutations with increasing maximum frequency was stronger for the pre-Perth/2009 nodes, which may be somewhat driven by high-frequency mutations generally moving towards the Perth/2009 sequence and thereby scored more favorably.
+Despite this, the post-Perth/2009 nodes exhibited mutations that were on average more favorable for viral growth at higher frequency nodes.
+Notably, nodes that reached moderate global frequencies ($> 10 \%$) had relatively few mutations that were highly deleterious compared to nodes that remained below $2 \%$ frequency, suggesting that unfavorable mutations may help determine whether or not a strain is able to surpass low frequencies in the global population.
+% Many low-frequency nodes also have mutations that are considered to be beneficial or have little effect in our experiments, indicating that factors other than viral growth influence clade dynamics.
+Overall, these results demonstrate that measurements of mutational effects in the Perth/2009 HA reveal differences in fitness between clades that circulate at low frequencies and those that achieve fixation or near-fixation.
 
-To investigate whether our experiments can aid in distinguishing trunk and side-branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the H3N2 phylogeny (Figure~\ref{fig:H3N2_phylogeny}).
-The mutations on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
-This difference in the effects of mutations between the trunk and side-branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
-Some influenza sequences are derived from egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{mcwhite2016sequence}.
-These mutations mostly occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
-We therefore re-calculated the statistics separating side branches to tip and internal nodes, and again found that the difference between the trunk and both sets of side branches was statistically significant (Figure~\ref{fig:trunkvssidebranch}B).
-Therefore, strains with mutations that we experimentally measure to be favorable for viral growth tend to do better in nature than strains with mutations that we measure to be less favorable.
+To determine how broadly experimental measurements can be generalized across HAs, we repeated the maximum frequency analyses using amino-acid preferences from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
+Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the H1 measurements do \textit{not} reveal a trend towards more beneficial mutations for higher-frequency nodes of an H3 phylogeny.
+This observation indicates that the utility of an experiment for revealing differences between low and high frequency clades degrades over sufficiently long evolutionary distances.
 
-Our experiments were performed on the Perth/2009 HA, but we scored mutations on a phylogenetic tree that extends from 1968 to 2017.
-Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
-To explore this question, we scored the complete HA sequence $\mathbf{s}$ of every node in the phylogenetic tree by quantifying its average per-site sequence preference as 
-\begin{equation}
-F\left(\mathbf{s}\right) = \frac{1}{L}\displaystyle\sum_{r=1}^L \ln \pi_{r, s_r},
-\end{equation}
-where $\pi_{r, s_r}$ is the preference measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid $s_r$ at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
-Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side-branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side-branch mutations.
-However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
-The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
-We do not think that this result means that Perth/2009 is actually higher fitness than the other strains.
-Rather, if there are modest shifts in the effects of mutations as HA evolves, then sequences more distant from Perth/2009 could be scored as having lower preference simply because these shifts start to degrade the applicability of our measurements.
-But importantly, our experiments consistently show the trunk to be more favorable than the side branches regardless of whether we look at sequences that precede Perth/2009 (most of Figure~\ref{fig:trunkvssidebranch}C) or sequences that occur after Perth/2009 has diverged from the trunk (Figure~\ref{suppfig:future}).
-Therefore, our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.
+\begin{figure*}[ht]
+\centering
+\includegraphics[width=\textwidth]{figs/muteffect_by_maxfreq.pdf}
+\caption{\label{fig:muteffect_maxfreq}
+{\bf Highly deleterious mutations are not observed in high-frequency nodes.}
+Figure caption.
+}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}[ht]
 \centering
-\includegraphics[width=\columnwidth]{figs/trunkvssidebranch/trunkvssidebranch.pdf}
-\caption{\label{fig:trunkvssidebranch}
-{\bf Mutations on the trunk are more favorable than those on the side branches according to our experimental measurements.}
-(A) We used our experiments to calculate the log$_{2}$ mutational effect for all trunk and side-branch mutations in five-year windows from 1968 to 2013. 
-The central year in each window is denoted on the x-axis.
-Median mutational effects in each window are shown as circles for the trunk and triangles for side branches. 
-Shaded regions demarcate the interquartile range.
-Negative numbers signify mutations to less preferred amino acids, while positive numbers signify mutations to more preferred mutations.
-(B) The log$_{2}$ mutational effect for all side branch and trunk mutations (left panel), and the same data but separating side branches to terminal and internal nodes (right panel).
-$P$-values were computed by randomizing the experimental measurements among sites 10,000 times, and determining how often the difference in median mutational effect between the trunk and side branches exceeded the actual value.
-(C) The average per-site sequence preference of every node in the phylogenetic tree in Fig.~\ref{fig:H3N2_phylogeny}.
-Trunk nodes are in red, side-branch nodes in blue, and Perth/2009 is marked with a yellow star.
-More preferred sequences have larger values.
+\includegraphics[width=\textwidth]{figs/muteffect_by_maxfreq_WSN.pdf}
+\caption{\label{fig:muteffect_maxfreq_WSN}
+{\bf Experimental measurements on an H1 HA do not reveal differences between low- and high-frequency nodes.}
+Figure caption.
 }
-\end{figure}
+\end{figure*}
 
-\subsection*{Our experiments suggest different patterns of evolution at epitope and non-epitope sites}
-HA is under selection both to maintain its essential function in viral growth and to escape pre-existing immunity.
-Most immune selection is focused on a subset of so-called ``epitope sites'' that are targets of the immunodominant antibody response.
-Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites~\cite{fitch1991positive,wolf2006long,bhatt2011genomic}.
-There are various classifications of HA epitope sites; here we will use the classification of Wolf \textit{et al}~\cite{wolf2006long}, which defines 129 epitope sites among the 566 residues in HA.
-In the timeframe from 1968 to 2012, the trunk of the tree in Figure~\ref{fig:H3N2_phylogeny} fixed 0.84 epitope mutations per site, but only 0.07 non-epitope mutations per site.
-Since our experiments only measure how mutations affect viral growth, they might be expected to describe the evolution of epitope and non-epitope sites differently.
+%\subsection*{The experimental measurements can help discriminate between successful and unsuccessful influenza virus lineages}
+%A major goal in the study of rapidly evolving viruses such as influenza is to forecast evolutionary trajectories~\cite{lassig2017predicting,morris2017predictive}.
+%For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection~\cite{neher2015nextflu,lassig2017predicting,morris2017predictive}.
+%Evolutionary forecasts should be able to distinguish successful and unsuccessful viral lineages, which includes discriminating between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).
+%
+%To investigate whether our experiments can aid in distinguishing trunk and side-branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the H3N2 phylogeny (Figure~\ref{fig:H3N2_phylogeny}).
+%The mutations on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
+%This difference in the effects of mutations between the trunk and side-branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
+%Some influenza sequences are derived from egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{mcwhite2016sequence}.
+%These mutations mostly occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
+%We therefore re-calculated the statistics separating side branches to tip and internal nodes, and again found that the difference between the trunk and both sets of side branches was statistically significant (Figure~\ref{fig:trunkvssidebranch}B).
+%Therefore, strains with mutations that we experimentally measure to be favorable for viral growth tend to do better in nature than strains with mutations that we measure to be less favorable. 
+%
+%Our experiments were performed on the Perth/2009 HA, but we scored mutations on a phylogenetic tree that extends from 1968 to 2017.
+%Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
+%To explore this question, we scored the complete HA sequence $\mathbf{s}$ of every node in the phylogenetic tree by quantifying its average per-site sequence preference as 
+%\begin{equation}
+%F\left(\mathbf{s}\right) = \frac{1}{L}\displaystyle\sum_{r=1}^L \ln \pi_{r, s_r},
+%\end{equation}
+%where $\pi_{r, s_r}$ is the preference measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid $s_r$ at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
+%Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side-branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side-branch mutations.
+%However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
+%The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
+%We do not think that this result means that Perth/2009 is actually higher fitness than the other strains.
+%Rather, if there are modest shifts in the effects of mutations as HA evolves, then sequences more distant from Perth/2009 could be scored as having lower preference simply because these shifts start to degrade the applicability of our measurements.
+%But importantly, our experiments consistently show the trunk to be more favorable than the side branches regardless of whether we look at sequences that precede Perth/2009 (most of Figure~\ref{fig:trunkvssidebranch}C) or sequences that occur after Perth/2009 has diverged from the trunk (Figure~\ref{suppfig:future}).
+%Therefore, our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.
 
-Mutations on the trunk are scored as significantly more favorable in our experiments than side-branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}A).
-Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and at sites that are also under immune pressure.
-The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that mutations at these sites still have non-antigenic functional consequences~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.
+%\begin{figure}
+%\centering
+%\includegraphics[width=\columnwidth]{figs/trunkvssidebranch/trunkvssidebranch.pdf}
+%\caption{\label{fig:trunkvssidebranch}
+%{\bf Mutations on the trunk are more favorable than those on the side branches according to our experimental measurements.}
+%(A) We used our experiments to calculate the log$_{2}$ mutational effect for all trunk and side-branch mutations in five-year windows from 1968 to 2013. 
+%The central year in each window is denoted on the x-axis.
+%Median mutational effects in each window are shown as circles for the trunk and triangles for side branches. 
+%Shaded regions demarcate the interquartile range.
+%Negative numbers signify mutations to less preferred amino acids, while positive numbers signify mutations to more preferred mutations.
+%(B) The log$_{2}$ mutational effect for all side branch and trunk mutations (left panel), and the same data but separating side branches to terminal and internal nodes (right panel).
+%$P$-values were computed by randomizing the experimental measurements among sites 10,000 times, and determining how often the difference in median mutational effect between the trunk and side branches exceeded the actual value.
+%(C) The average per-site sequence preference of every node in the phylogenetic tree in Fig.~\ref{fig:H3N2_phylogeny}.
+%Trunk nodes are in red, side-branch nodes in blue, and Perth/2009 is marked with a yellow star.
+%More preferred sequences have larger values.
+%}
+%\end{figure}
 
-However, there are differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
-The per-site sequence preference at epitope sites increases over time, whereas at non-epitope sites it remains relatively constant  (Figure~\ref{fig:sequence_preference}B).
-This difference is unlikely to be explained purely by differences in the number of mutations at these two sets of sites: there have been $\sim$4-fold more total trunk mutations at epitope than non-epitope sites (109 versus 28), but the large increase in sequence preference for the epitope sites is $\gg$4-fold larger than the nearly imperceptible increase for non-epitope sites.
-The fact that only epitope sites exhibit an increase in sequence preference over time can be rationalized in terms of prior knowledge about protein evolution.
-As proteins evolve, there can be epistatic shifts in the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
-Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some epitope sites in H3 HA~\cite{wu2017diversity}.
-It is also known that immune pressure can increase epistasis among substitutions~\cite{gong2014epistatically}.
-We hypothesize that immune pressure increases the accumulation of epistatic interactions involving substitutions at epitope sites.
-Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA (Figure~\ref{fig:sequence_preference}B).
-Such a trend is not apparent for non-epitope sites, perhaps because they are under less selection to fix adaptive mutations.
-
-\begin{figure}
-\centering
-\includegraphics[width=\columnwidth]{figs/sequence_preference/sequence_preference.pdf}
-\caption{\label{fig:sequence_preference}
-{\bf Effects of mutations at epitope and non-epitope sites during HA evolution.}
-We partitioned HA into epitope and non-epitope sites~\cite{wolf2006long}.
-(A) The log$_{2}$ mutational effects for side-branch and trunk mutations at non-epitope (left) and epitope (right) sites.
-$P$-values were computed as in Figure~\ref{fig:trunkvssidebranch} but only performing the randomizations over the appropriate set of sites (non-epitope or epitope).
-(B) 
-The average per-site sequence preference for all nodes in the phylogenetic tree, calculated separately for each set of sites.
-}
-\end{figure}
+%\subsection*{Our experiments suggest different patterns of evolution at epitope and non-epitope sites}
+%HA is under selection both to maintain its essential function in viral growth and to escape pre-existing immunity.
+%Most immune selection is focused on a subset of so-called ``epitope sites'' that are targets of the immunodominant antibody response.
+%Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites~\cite{fitch1991positive,wolf2006long,bhatt2011genomic}.
+%There are various classifications of HA epitope sites; here we will use the classification of Wolf \textit{et al}~\cite{wolf2006long}, which defines 129 epitope sites among the 566 residues in HA.
+%In the timeframe from 1968 to 2012, the trunk of the tree in Figure~\ref{fig:H3N2_phylogeny} fixed 0.84 epitope mutations per site, but only 0.07 non-epitope mutations per site.
+%Since our experiments only measure how mutations affect viral growth, they might be expected to describe the evolution of epitope and non-epitope sites differently.
+%
+%Mutations on the trunk are scored as significantly more favorable in our experiments than side-branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}A).
+%Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and at sites that are also under immune pressure.
+%The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that mutations at these sites still have non-antigenic functional consequences~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.
+%
+%However, there are differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
+%The per-site sequence preference at epitope sites increases over time, whereas at non-epitope sites it remains relatively constant  (Figure~\ref{fig:sequence_preference}B).
+%This difference is unlikely to be explained purely by differences in the number of mutations at these two sets of sites: there have been $\sim$4-fold more total trunk mutations at epitope than non-epitope sites (109 versus 28), but the large increase in sequence preference for the epitope sites is $\gg$4-fold larger than the nearly imperceptible increase for non-epitope sites.
+%The fact that only epitope sites exhibit an increase in sequence preference over time can be rationalized in terms of prior knowledge about protein evolution.
+%As proteins evolve, there can be epistatic shifts in the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
+%Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some epitope sites in H3 HA~\cite{wu2017diversity}.
+%It is also known that immune pressure can increase epistasis among substitutions~\cite{gong2014epistatically}.
+%We hypothesize that immune pressure increases the accumulation of epistatic interactions involving substitutions at epitope sites.
+%Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA (Figure~\ref{fig:sequence_preference}B).
+%Such a trend is not apparent for non-epitope sites, perhaps because they are under less selection to fix adaptive mutations.
 
 \subsection*{There are large differences between H3 and H1 HAs in the amino-acid preferences of many sites}
-The results above show that measurements of mutational effects in the Perth/2009 HA are informative about the evolutionary fate of other human H3N2 HAs.
-To determine how broadly experimental measurements be generalized across HAs, we repeated the trunk versus side-branch analyses using amino-acid preferences from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
-Figure~\ref{fig:WSN_trunkvssidebranch} shows that the H1 measurements are not informative for distinguishing the trunk and side branches of an H3 phylogeny.
-This fact indicates that the utility of an experiment for discriminating successful and unsuccessful strains degrades over sufficiently long evolutionary distances.
 
-\begin{figure}
-\centering
-\includegraphics[width=\columnwidth]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}
-\caption{\label{fig:WSN_trunkvssidebranch}
-{\bf Experimental measurements on an H1 HA do \emph{not} provide evolutionarily relevant information for H3 HAs.}
-(A), (B) This figure is analogous to that in Figure~\ref{fig:trunkvssidebranch} except that it scores the H3 sequences using experimental measurements made on the lab-adapted WSN/1933 HA~\cite{doud2016accurate}.
-}
-\end{figure}
+%\begin{figure}
+%\centering
+%\includegraphics[width=\columnwidth]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}
+%\caption{\label{fig:WSN_trunkvssidebranch}
+%{\bf Experimental measurements on an H1 HA do \emph{not} provide evolutionarily relevant information for H3 HAs.}
+%(A), (B) This figure is analogous to that in Figure~\ref{fig:trunkvssidebranch} except that it scores the H3 sequences using experimental measurements made on the lab-adapted WSN/1933 HA~\cite{doud2016accurate}.
+%}
+%\end{figure}
 
-An obvious hypothesis for why the H1 deep mutational scanning is not informative about the evolutionary fate of H3 strains is that the effect of the same mutation is often different between these two HA subtypes.
+An obvious hypothesis for why the H1 deep mutational scanning do not reveal differences among H3 strains is that the effect of the same mutation is often different between these two HA subtypes.
 To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.

### 2018-03-27 paper/manuscript.tex
Update with Jesse comments

@@
 Here we experimentally measure the effects on viral growth in cell culture of all single amino-acid mutations to the HA from a recent human H3N2 influenza virus strain. 
-We then show that mutations in evolutionarily successful H3N2 viral lineages are measured to generally be more favorable than mutations in lineages that die out.
-Therefore, despite the well-known caveats about cell-culture measurements of viral fitness, such measurements can still be informative for understanding virus evolution in nature.
+We then show that mutations in evolutionarily successful H3N2 viral lineages are measured to generally be more favorable than mutations in lineages that quickly die out.
+Therefore, despite the well-known caveats about cell-culture measurements of viral fitness, such measurements can still be informative for understanding evolution in nature.
 We also compare our measurements for an H3 HA to similar data previously generated for a distantly related H1 HA, and find substantial differences in which amino acids are preferred at many sites.
 For instance, the H3 HA has less disparity in mutational tolerance between the head and stalk domains than the H1 HA.
-Overall, our work suggests that experimental measurements of mutational effects can be leveraged to help forecast the evolutionary fates of viral lineages in nature --- but only when the measurements are made on a viral strain similar to the ones being studied in nature.
+Overall, our work suggests that experimental measurements of mutational effects can be leveraged to help understand the evolutionary fates of viral lineages in nature --- but only when the measurements are made on a viral strain similar to the ones being studied in nature.
 \end{abstract}
@@
 Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping,bhatt2011genomic}.
-This evolution is further characterized by competition and turnover among groups of strains or clades.
+This evolution is further characterized by competition and turnover among groups of strains or clades~\cite{fitch1997long,strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects,bedford2015global}.
 As a result, mutations that differentiate clades exhibit dynamic frequency trajectories as these clades appear, circulate, or die out.
-Several lines of evidence indicate that successful clades, which reach high maximum frequencies or even fix in the global population, have higher fitness than clades that remain at low frequency.
+Several lines of evidence indicate that successful clades, which reach high frequencies or even fix in the global population, have higher fitness than clades that remain at low frequency~\cite{strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects}.
 A key goal in the study of H3N2 evolution is to identify the features that enable certain clades to succeed as others die out.
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{figs/H3N2_phylogeny/H3N2_phylogeny.pdf}
-\caption{\label{fig:H3N2_phylogeny}
-{\bf Human H3N2 HA phylogeny from 1968-2017.}
-The trunk is shown in red, and side branches are shown in blue.
-The gray branches represent the part of the tree for which we cannot yet distinguish the trunk from side branches.
-The Perth/2009 strain is indicated with a star.
-}
-\end{figure}
+%\begin{figure}
+%\centering
+%\includegraphics[width=\linewidth]{figs/H3N2_phylogeny/H3N2_phylogeny.pdf}
+%\caption{\label{fig:H3N2_phylogeny}
+%{\bf Human H3N2 HA phylogeny from 1968-2017.}
+%The trunk is shown in red, and side branches are shown in blue.
+%The gray branches represent the part of the tree for which we cannot yet distinguish the trunk from side branches.
+%The Perth/2009 strain is indicated with a star.
+%}
+%\end{figure}
 
@@
 Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
-We then apply these measurements to examine differences in maximum clade frequencies.
+We then examine these measurements in the context of the natural evolution of human influenza, and show that they can help discriminate evolutionarily successful viral clades from lineages that quickly die out.
+However, the utility of experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are not useful for understanding the evolutionary fate of H3 viral strains.
 %We then examine whether these measurements are useful for distinguishing between H3N2 trunk and side-branch lineages, and find that mutations on the trunk are measured to be more favorable than those on the side branches, both at epitope and non-epitope sites.
@@
 Examination of Figure~\ref{fig:logoplot} reveals that the experimentally measured amino-acid preferences generally agree with existing knowledge about HA's structure and function.
-For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} possess high preference for cysteine.
+For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} strongly prefer cysteine.
 At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such as Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change,yang2015structure}.
@@
 Surprisingly, there was much less contrast in mutational tolerance between the stalk and head domains for the H3 HA than for the H1 (Figure~\ref{fig:mut_tolerance}).
-Specifically, for solvent-exposed residues, the head domain is more mutationally tolerant than the stalk domain for the WSN/1933 H1 HA, whereas such a difference in tolerances was not seen for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B), and this was true using different relative solvent accessibility cutoffs to define solvent-exposed residues (Figure~\ref{suppfig:head_stalk_RSA}).
+Specifically, for solvent-exposed residues, the head domain was more mutationally tolerant than the stalk domain for the WSN/1933 H1 HA, whereas such a difference in tolerances was not seen for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
+This difference between H1 and H3 HAs in the disparity of mutational tolerance between surface residues in the head and stalk domains was robust to the solvent accessibility cutoff used to define surface residues (Figure~\ref{suppfig:head_stalk_RSA}).
 Furthermore, in the H3 HA, the short helix A in the stalk was found to be very mutationally tolerant.
@@
 
-\subsection*{Our measurements reveal differences between clades of low and high maximum frequencies}
-The maximum frequencies achieved by particular mutations are a reflection of the fitnesses of strains carrying those mutations.
-Do the experimentally-determined preferences therefore reveal underlying differences in strains that reach various maximum frequencies?
-To explore this question, we examined the relationship between the experimentally measured effects of all mutations and the maximum frequency achieved by all nodes with at least one tip in a maximum-likelihood H3N2 phylogeny.
+\subsection*{Our measurements can help distinguish between viral clades that reach low and high frequencies in nature}
+A major goal in the study of rapidly evolving viruses such as influenza is to better understand the features that distinguish successful and unsuccessful clades, as this can help us identify clades that are likely to fix from those that will quickly die out~\cite{lassig2017predicting,morris2017predictive}.
+In principle, the maximum frequencies achieved by particular strains are a reflection of the fitnesses of those strains~\cite{strelkowa2012clonal,neher2014predicting,koelle2015effects}.
+We therefore investigated whether our experiments would reveal underlying differences in strains that reach various frequencies in the global population.
+
+We first examined the relationship between the effects of all mutations and the maximum frequency achieved by all nodes with at least one tip in a maximum-likelihood H3N2 phylogeny.
 We estimated the frequency trajectories for every node in the tree by... \comment {I'll let John / Trevor add more here}.
@@
 Also this is where we would reference the tree supplemental figure}.
+
 We found that mutations measured to be very deleterious to viral growth in our experiments were not observed at high frequencies in nature.
@@
 Mutations were on average scored as being more detrimental in the lower frequency bins than at higher frequencies.
+
 The trend towards more beneficial mutations with increasing maximum frequency was stronger for the pre-Perth/2009 nodes, which may be somewhat driven by high-frequency mutations generally moving towards the Perth/2009 sequence and thereby scored more favorably.
 Despite this, the post-Perth/2009 nodes exhibited mutations that were on average more favorable for viral growth at higher frequency nodes.
+
 Notably, nodes that reached moderate global frequencies ($> 10 \%$) had relatively few mutations that were highly deleterious compared to nodes that remained below $2 \%$ frequency, suggesting that unfavorable mutations may help determine whether or not a strain is able to surpass low frequencies in the global population.
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex.png}
+\caption{\label{fig:frequency_trajectory}
+{\bf Frequency trajectories of seasonal influenza viral clades.}
+Figure caption
+}
+\end{figure}
+
+\subsection*{Measurements made for an H1 HA are not useful for understanding the evolution of H3 influenza}
 To determine how broadly experimental measurements can be generalized across HAs, we repeated the maximum frequency analyses using amino-acid preferences from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.

### 2018-03-28 paper/manuscript.tex
small tweaks to title / abstract / introduction

@@
 
-\title{Deep mutational scanning of hemagglutinin helps identify evolutionarily successful human H3N2 influenza viruses}
+\title{Deep mutational scanning of hemagglutinin helps predict the evolutionary fates of human H3N2 influenza virus variants}
 
@@
 We show that these measurements have utility for distinguishing among viral strains that do and do not succeed in nature.
-Overall, our work suggests that new high-throughput experimental approaches may be useful for evolutionary forecasting.
+Overall, our work suggests that new high-throughput experimental approaches may be useful for understanding virus evolution in nature.
 }
@@
 \begin{abstract}
-% no more than 250 words. (170 now)
+% no more than 250 words. 
 Human influenza virus rapidly accumulates mutations in its major surface protein hemagglutinin (HA).
@@
 A key goal in the study of H3N2 evolution is to identify the features that enable certain clades to succeed as others die out.
-% This evolution is further characterized by strain competition and frequent population turnover~\cite{fitch1997long,strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects,bedford2015global}, producing a spindly phylogenetic tree with a persistent trunk lineage and short-lived side branches~\cite{fitch1991positive} (Figure~\ref{fig:H3N2_phylogeny}).
-% Several lines of evidence indicate that the trunk has higher fitness than other lineages~\cite{strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects}.
-% A key challenge in the study of H3N2 evolution is to forecast which strain will dominate the upcoming influenza season, as this information can guide selection of a vaccine strain~\cite{morris2017predictive}.
-% In order to make these forecasts, it is important to identify the features that enable the trunk strain to persist as other strains die out.
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=\linewidth]{figs/H3N2_phylogeny/H3N2_phylogeny.pdf}
-%\caption{\label{fig:H3N2_phylogeny}
-%{\bf Human H3N2 HA phylogeny from 1968-2017.}
-%The trunk is shown in red, and side branches are shown in blue.
-%The gray branches represent the part of the tree for which we cannot yet distinguish the trunk from side branches.
-%The Perth/2009 strain is indicated with a star.
-%}
-%\end{figure}
 
@@
 
-It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of large numbers of mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high}.
+It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of all single amino-acid mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high}.
 However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
 Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
-We then examine these measurements in the context of the natural evolution of human influenza, and show that they can help discriminate evolutionarily successful viral clades from lineages that quickly die out.
-However, the utility of experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are not useful for understanding the evolutionary fate of H3 viral strains.
-%We then examine whether these measurements are useful for distinguishing between H3N2 trunk and side-branch lineages, and find that mutations on the trunk are measured to be more favorable than those on the side branches, both at epitope and non-epitope sites.
-%However, the experiments are useful for assessing the effects of mutations only in closely related viral strains, since we show that mutations often have disparate effects on H3 and H1 HAs.
-%Our work highlights the potential for using high-throughput experimental measurements of mutational effects to inform evolutionary forecasting of human seasonal influenza virus.
+We then show that these experimental measurements can help discriminate evolutionarily successful viral clades from lineages that quickly die out.
+However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are not useful for understanding the evolutionary fate of H3 viral strains.
 

### 2018-03-28 paper/manuscript.tex
edits to head vs stalk

@@
 
-We performed a similar analysis using the current data for the Perth/2009 H3 HA.
-Surprisingly, there was much less contrast in mutational tolerance between the stalk and head domains for the H3 HA than for the H1 (Figure~\ref{fig:mut_tolerance}).
-Specifically, for solvent-exposed residues, the head domain was more mutationally tolerant than the stalk domain for the WSN/1933 H1 HA, whereas such a difference in tolerances was not seen for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
-This difference between H1 and H3 HAs in the disparity of mutational tolerance between surface residues in the head and stalk domains was robust to the solvent accessibility cutoff used to define surface residues (Figure~\ref{suppfig:head_stalk_RSA}).
-Furthermore, in the H3 HA, the short helix A in the stalk was found to be very mutationally tolerant.
-Interestingly, there are more reports of readily selecting escape mutants from broadly neutralizing anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.
+We performed a similar analysis using our new data for the Perth/2009 H3 HA.
+Surprisingly, the head domain of the H3 HA is \emph{not} more mutationally tolerant than the stalk domain (Figure~\ref{fig:mut_tolerance}).
+Specifically, whereas solvent-exposed residues in the head domain are substantially more mutationally tolerant than those in the stalk domain for the WSN/1933 H1 HA, the trend is actually reversed for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
+This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (Figure~\ref{suppfig:head_stalk_RSA}).
+For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
+Helix A is a component of the epitope of many broadly neutralizing anti-stalk antibodies\comment{cite something}, and interestingly, there are more reports of readily selecting escape mutants from such anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.
 
@@
 For instance, antigenic region B is an immunodominant area, and many recent major antigenic drift mutations have occurred in this region~\cite{chambers2015identification,koel2013substitutions,popova2012immunodominance}.
-We find that the most distal portion of the globular head near the 190-helix, which is part of antigenic region B, is highly tolerant of mutations (Figure~\ref{fig:mut_tolerance}).
+We find that the most distal portion of the globular head near the 190-helix, which is part of antigenic region B, is highly tolerant of mutations (Figure~\ref{fig:mut_tolerance}A).
 Antigenic region C is also notably mutationally tolerant.
 
-Many residues inside HA's receptor binding pocket are known to be highly functionally constrained~\cite{wilson1981structure,martin1998studies}, and our data indicates that these sites are relatively mutationally intolerant in both H3 and H1 HAs.
+Many residues inside HA's receptor binding pocket are known to be highly functionally constrained~\cite{wilson1981structure,martin1998studies}, and our data indicates that these sites are relatively mutationally intolerant in both H3 and H1 HAs (Figure~\ref{fig:mut_tolerance}A).
 In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, since mutations at these sites can have large effects on antigenicity~\cite{wiley1981structural,koel2013substitutions}.
@@
 Different color scales are used because measurements are comparable among sites within the same HA, but not necessarily across HAs.
-Both trimers are shown in approximately the same orientation. 
+Both trimers are shown in the same orientation. 
 For each HA, the structure at left shows a surface representation of the full trimer, while the structure at right side shows a ribbon representation of just one monomer.
 The sialic acid receptor is shown in red sticks.
-(B) The mutational tolerance (Shannon entropy of re-scaled preferences) of solvent-exposed residues with a relative solvent accessibility $\geq$ 0.2 in the head and stalk domains of the Perth/2009 H3 HA (purple) and WSN/1933 H1 HA (gold).
-We quantified the absolute solvent accessibility of each site using DSSP~\cite{kabsch1983} and then normalized to a relative solvent accessibility using the absolute accessibilities in~\cite{tien2013}.
+(B) The mutational tolerance of solvent-exposed residues in the head and stalk domains of the Perth/2009 H3 HA (purple) and WSN/1933 H1 HA (gold). 
+\comment{Move this sentence to Method: We quantified the absolute solvent accessibility of each site using DSSP~\cite{kabsch1983} and then normalized to a relative solvent accessibility using the absolute accessibilities in~\cite{tien2013}.}
 Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
-Note that the mutational tolerance values are not comparable between the two HAs.
+A residue was classified as solvent exposed if its relative solvent accessibility was $\geq$ 0.2; Figure~\ref{suppfig:head_stalk_RSA} shows that the results are robust to the choice of solvent accessibility cutoff.
+Note that the mutational tolerance values are not comparable between the two HAs, but are comparable between domains of the same HA.
 }
@@
 \caption{\label{suppfig:head_stalk_RSA}
-{\bf Mutational tolerances of the head and stalk domains at various RSA cutoffs.}
+{\bf Mutational tolerances of the head and stalk domains at various relative solvent accessibility cutoffs.}
 The mutational tolerances of the head and stalk domains show less disparity for the Perth/2009 H3 HA compared to those for the WSN/1933 H1 HA. 

### 2018-03-28 paper/manuscript.tex
more edits to Results / Discussion

@@
 (B) The mutational tolerance of solvent-exposed residues in the head and stalk domains of the Perth/2009 H3 HA (purple) and WSN/1933 H1 HA (gold). 
-\comment{Move this sentence to Method: We quantified the absolute solvent accessibility of each site using DSSP~\cite{kabsch1983} and then normalized to a relative solvent accessibility using the absolute accessibilities in~\cite{tien2013}.}
+\comment{Juhye: Move this sentence to Method: We quantified the absolute solvent accessibility of each site using DSSP~\cite{kabsch1983} and then normalized to a relative solvent accessibility using the absolute accessibilities in~\cite{tien2013}.}
 Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
@@
 \subsection*{Our measurements can help distinguish between viral clades that reach low and high frequencies in nature}
-A major goal in the study of rapidly evolving viruses such as influenza is to better understand the features that distinguish successful and unsuccessful clades, as this can help us identify clades that are likely to fix from those that will quickly die out~\cite{lassig2017predicting,morris2017predictive}.
-In principle, the maximum frequencies achieved by particular strains are a reflection of the fitnesses of those strains~\cite{strelkowa2012clonal,neher2014predicting,koelle2015effects}.
-We therefore investigated whether our experiments would reveal underlying differences in strains that reach various frequencies in the global population.
+A major goal in the study of rapidly evolving viruses such as influenza is to understand the features that distinguish successful and unsuccessful clades, as this can help us identify clades that are likely to reach high frequencies from those that will quickly die out~\cite{lassig2017predicting,morris2017predictive}.
+The frequencies that viral strains reach nature depend in large part on their fitnesses~\cite{strelkowa2012clonal,neher2014predicting,koelle2015effects}.
+Our experiments measure viral growth in cell culture, which is a partial albeit imperfect proxy for viral fitness.
+We therefore investigated whether these measurements could help distinguish among strains that do and do not reach high frequencies during natural evolution.
 
-We first examined the relationship between the effects of all mutations and the maximum frequency achieved by all nodes with at least one tip in a maximum-likelihood H3N2 phylogeny.
-We estimated the frequency trajectories for every node in the tree by... \comment {I'll let John / Trevor add more here}.
-Evolutionarily successful nodes that occurred \textit{before} the Perth/2009 strain diverged from the trunk of the phylogenetic tree tend to carry mutations that move towards the Perth/2009 sequence. 
-Because the effects of mutations may vary depending on genetic background, we analyzed nodes that preceded the Perth/2009 strain separately from nodes that occurred after the split of Perth/2009 from the trunk of the phylogenetic tree \comment{need better explanation of pre- and post-Perth/2009 rationale here since it seems like we are going to completely drop the trunk/tip analysis from the paper?
-Also this is where we would reference the tree supplemental figure}.
-
-We found that mutations measured to be very deleterious to viral growth in our experiments were not observed at high frequencies in nature.
-Rather, only mutations with beneficial or small deleterious effects in our experiments appeared to reach high frequencies or even fix in nature, as evidenced by the slight but significant\comment{?} trend towards higher frequencies with more favorable mutations (Figure ~\ref{fig:muteffect_maxfreq}).
-Furthermore, shifts toward more favorable mutations with increasing maximum frequency were also apparent from plotting the distribution of mutational effects for different frequency bins.
-Mutations were on average scored as being more detrimental in the lower frequency bins than at higher frequencies.
-
-The trend towards more beneficial mutations with increasing maximum frequency was stronger for the pre-Perth/2009 nodes, which may be somewhat driven by high-frequency mutations generally moving towards the Perth/2009 sequence and thereby scored more favorably.
-Despite this, the post-Perth/2009 nodes exhibited mutations that were on average more favorable for viral growth at higher frequency nodes.
-
-Notably, nodes that reached moderate global frequencies ($> 10 \%$) had relatively few mutations that were highly deleterious compared to nodes that remained below $2 \%$ frequency, suggesting that unfavorable mutations may help determine whether or not a strain is able to surpass low frequencies in the global population.
-% Many low-frequency nodes also have mutations that are considered to be beneficial or have little effect in our experiments, indicating that factors other than viral growth influence clade dynamics.
-Overall, these results demonstrate that measurements of mutational effects in the Perth/2009 HA reveal differences in fitness between clades that circulate at low frequencies and those that achieve fixation or near-fixation.
+Specifically, we examined the relationship between our experimental measurements and the maximum frequency reached by the clade descended from each branch on the tree on which an amino-acid mutation has occurred.
+\comment{Trevor: is there some general paper we should cite for the concept of clade frequencies, or should we just cite as here?}
+This concept of relating the experimentally measured effects of mutations to the maximum clade frequency is illustrated in Figure~\ref{fig:frequency_trajectory} for sequences isolated between 2004 and 2014.
+If a mutation is deleterious, we would expect the clade descended from the branch containing that mutation to quickly die out, whereas branches with beneficial mutations are more likely to lead to clades that expand in frequency.
+Qualitative inspection of Figure~\ref{fig:frequency_trajectory} suggests that mutations that are experimentally measured to be neutral or beneficial are more likely than those measured to be deleterious to occur on branches leading to high frequency clades.
 
@@
 \caption{\label{fig:frequency_trajectory}
-{\bf Frequency trajectories of seasonal influenza viral clades.}
-Figure caption
+{\bf Frequency trajectories of viral clades and their relation to the effects of individual mutations.}
+The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} windowed from 2004 to 2014. 
+Circles indicate amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
+The bottom panel shows the frequency trajectory of the clade downstream of each mutation, with clades colored according to the mutational effects as in the top panel.
+It is obvious that most clades that reach high frequency have mutations that are measured to be relatively favorable in our experiments.
+\comment{John: Can you update figure?
+In particular, maybe better colors.
+Also, let's increase the size of the legend and make it $<-10$, and $> 0$ rather than having the \emph{inf} in the labels.
+Also, can you make y-axis at bottom ``clade frequency.''}
 }
@@
 
-\subsection*{Measurements made for an H1 HA are not useful for understanding the evolution of H3 influenza}
-To determine how broadly experimental measurements can be generalized across HAs, we repeated the maximum frequency analyses using amino-acid preferences from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
-Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the H1 measurements do \textit{not} reveal a trend towards more beneficial mutations for higher-frequency nodes of an H3 phylogeny.
-This observation indicates that the utility of an experiment for revealing differences between low and high frequency clades degrades over sufficiently long evolutionary distances.
+To analyze the relationship between mutational effect and clade frequency more quantitatively, we next partitioned the phylogeny of human H3N2 influenza HA into sequences that split from the line of descent (trunk of the phylogenetic tree) before or after the Perth/2009 strain used in our experiments (see Figure~\ref{suppfig:tree}).
+The rationale for partitioning into pre- and post-Perth/2009 sequences is that we wanted to test the utility of our measurements for prospective as well as post-hoc analyses.
+\comment{John: do you have a sentence to add about how we used max tips to avoid passaging artifacts?}
+Figure~\ref{fig:muteffect_maxfreq} shows that there is a statistically significant trend for clades that reach high frequency to have more favorable mutations than clades that quickly die out.
+Importantly, the effect size is similar for both the pre- and post-Perth/2009 sequence groups.
 
@@
 \caption{\label{fig:muteffect_maxfreq}
-{\bf Highly deleterious mutations are not observed in high-frequency nodes.}
-Figure caption.
+{\bf The experimental measurements are informative about the evolutionary fate of viral clades.}
+(A) The correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by viral clades.
+The plots show the Spearman correlation $\rho$ and a $P$-value obtained by repeatedly randomizing the experimental measurements among sites \comment{John: is this right about how the $P$-value was calculated?}.
+(B) The distribution of mutational effects among clades binned by the maximum frequency that they reach.
+Note that the analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{fig:tree}).
+\comment{Split the top and bottom into A and B, adjust font sizes (particularly, tick labels too big), adjust capitalization to be consistent with other figures (which usually start lower-case unless proper noun), increasing spacing between A and B panels.
+Also, no need for a title at top and figure can probably get smaller.}
 }
 \end{figure*}
+
+The trends in Figure~\ref{fig:muteffect_maxfreq} appear to be most strongly driven by the fact that mutations we measure to be substantially deleterious to viral growth almost never occur in clades that reach high frequency.
+Among clades that reach high frequency, the mutations are overwhelmingly measured to be either beneficial or close to neutral.
+Overall, these results demonstrate that measurements of how mutations affect viral growth in cell culture are informative for understanding the fates of the viral clades in nature: in particular, if a clade has a mutation that is measurably deleterious to viral growth, that clade is unlikely to prosper in nature.
+
+\subsection*{Measurements made for an H1 HA are not useful for understanding the evolution of H3 influenza}
+To determine how broadly experimental measurements can be generalized across HAs, we repeated the clade frequency analyses using mutational effects from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
+Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the H1 measurements do \textit{not} reveal a trend for more beneficial mutations to be enriched in higher-frequency clades of an H3 phylogeny.
+This result is not unsurprising given that H1 and H3 HAs only have 42\% sequence identity, but it does underscore that the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
 
@@
 \caption{\label{fig:muteffect_maxfreq_WSN}
-{\bf Experimental measurements on an H1 HA do not reveal differences between low- and high-frequency nodes.}
-Figure caption.
+{\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 viral clades.}
+This figure repeats the analysis of the H3N2 clade frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
+Figure \comment{?} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
+\comment{Juhye: Maybe you can make a template for this. Just trim this figure to only show the correlations, and then put the histogram data as a supplement.}
 }
@@
 
-%\subsection*{The experimental measurements can help discriminate between successful and unsuccessful influenza virus lineages}
-%A major goal in the study of rapidly evolving viruses such as influenza is to forecast evolutionary trajectories~\cite{lassig2017predicting,morris2017predictive}.
-%For instance, forecasting which viral lineages will dominate the upcoming influenza season is an integral part of vaccine-strain selection~\cite{neher2015nextflu,lassig2017predicting,morris2017predictive}.
-%Evolutionary forecasts should be able to distinguish successful and unsuccessful viral lineages, which includes discriminating between the trunk and side branches of the phylogenetic tree (Figure~\ref{fig:H3N2_phylogeny}).
-%
-%To investigate whether our experiments can aid in distinguishing trunk and side-branch lineages, we calculated the experimentally measured effects of all mutations in a maximum-likelihood reconstruction of the H3N2 phylogeny (Figure~\ref{fig:H3N2_phylogeny}).
-%The mutations on the trunk of the tree were consistently more beneficial for viral growth according to our experimental measurements (Figure~\ref{fig:trunkvssidebranch}A).
-%This difference in the effects of mutations between the trunk and side-branch lineages was statistically significant when taken across the entire phylogeny (Figure~\ref{fig:trunkvssidebranch}B).
-%Some influenza sequences are derived from egg- or cell-passaged isolates that contain lab-adaptation mutations~\citep{mcwhite2016sequence}.
-%These mutations mostly occur on the terminal side branches that lead to tip nodes on the phylogenetic tree.
-%We therefore re-calculated the statistics separating side branches to tip and internal nodes, and again found that the difference between the trunk and both sets of side branches was statistically significant (Figure~\ref{fig:trunkvssidebranch}B).
-%Therefore, strains with mutations that we experimentally measure to be favorable for viral growth tend to do better in nature than strains with mutations that we measure to be less favorable. 
-%
-%Our experiments were performed on the Perth/2009 HA, but we scored mutations on a phylogenetic tree that extends from 1968 to 2017.
-%Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
-%To explore this question, we scored the complete HA sequence $\mathbf{s}$ of every node in the phylogenetic tree by quantifying its average per-site sequence preference as 
-%\begin{equation}
-%F\left(\mathbf{s}\right) = \frac{1}{L}\displaystyle\sum_{r=1}^L \ln \pi_{r, s_r},
-%\end{equation}
-%where $\pi_{r, s_r}$ is the preference measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid $s_r$ at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
-%Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side-branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side-branch mutations.
-%However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
-%The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
-%We do not think that this result means that Perth/2009 is actually higher fitness than the other strains.
-%Rather, if there are modest shifts in the effects of mutations as HA evolves, then sequences more distant from Perth/2009 could be scored as having lower preference simply because these shifts start to degrade the applicability of our measurements.
-%But importantly, our experiments consistently show the trunk to be more favorable than the side branches regardless of whether we look at sequences that precede Perth/2009 (most of Figure~\ref{fig:trunkvssidebranch}C) or sequences that occur after Perth/2009 has diverged from the trunk (Figure~\ref{suppfig:future}).
-%Therefore, our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=\columnwidth]{figs/trunkvssidebranch/trunkvssidebranch.pdf}
-%\caption{\label{fig:trunkvssidebranch}
-%{\bf Mutations on the trunk are more favorable than those on the side branches according to our experimental measurements.}
-%(A) We used our experiments to calculate the log$_{2}$ mutational effect for all trunk and side-branch mutations in five-year windows from 1968 to 2013. 
-%The central year in each window is denoted on the x-axis.
-%Median mutational effects in each window are shown as circles for the trunk and triangles for side branches. 
-%Shaded regions demarcate the interquartile range.
-%Negative numbers signify mutations to less preferred amino acids, while positive numbers signify mutations to more preferred mutations.
-%(B) The log$_{2}$ mutational effect for all side branch and trunk mutations (left panel), and the same data but separating side branches to terminal and internal nodes (right panel).
-%$P$-values were computed by randomizing the experimental measurements among sites 10,000 times, and determining how often the difference in median mutational effect between the trunk and side branches exceeded the actual value.
-%(C) The average per-site sequence preference of every node in the phylogenetic tree in Fig.~\ref{fig:H3N2_phylogeny}.
-%Trunk nodes are in red, side-branch nodes in blue, and Perth/2009 is marked with a yellow star.
-%More preferred sequences have larger values.
-%}
-%\end{figure}
-
-%\subsection*{Our experiments suggest different patterns of evolution at epitope and non-epitope sites}
-%HA is under selection both to maintain its essential function in viral growth and to escape pre-existing immunity.
-%Most immune selection is focused on a subset of so-called ``epitope sites'' that are targets of the immunodominant antibody response.
-%Although both epitope and non-epitope sites evolve rapidly, immune selection drives a higher rate of evolution at epitope sites~\cite{fitch1991positive,wolf2006long,bhatt2011genomic}.
-%There are various classifications of HA epitope sites; here we will use the classification of Wolf \textit{et al}~\cite{wolf2006long}, which defines 129 epitope sites among the 566 residues in HA.
-%In the timeframe from 1968 to 2012, the trunk of the tree in Figure~\ref{fig:H3N2_phylogeny} fixed 0.84 epitope mutations per site, but only 0.07 non-epitope mutations per site.
-%Since our experiments only measure how mutations affect viral growth, they might be expected to describe the evolution of epitope and non-epitope sites differently.
-%
-%Mutations on the trunk are scored as significantly more favorable in our experiments than side-branch mutations at both epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}A).
-%Therefore, our experiments can distinguish evolutionarily favorable mutations both at sites that are predominantly under functional constraint and at sites that are also under immune pressure.
-%The fact that our experiments can discriminate between the trunk and side branches at epitope sites despite only assaying for viral growth underscores the fact that mutations at these sites still have non-antigenic functional consequences~\cite{nakajima2003restriction,das2013defining,koel2013substitutions}.
-%
-%However, there are differences in how the average per-site sequence preference changes over time at the epitope and non-epitope sites (Figure~\ref{fig:sequence_preference}B).
-%The per-site sequence preference at epitope sites increases over time, whereas at non-epitope sites it remains relatively constant  (Figure~\ref{fig:sequence_preference}B).
-%This difference is unlikely to be explained purely by differences in the number of mutations at these two sets of sites: there have been $\sim$4-fold more total trunk mutations at epitope than non-epitope sites (109 versus 28), but the large increase in sequence preference for the epitope sites is $\gg$4-fold larger than the nearly imperceptible increase for non-epitope sites.
-%The fact that only epitope sites exhibit an increase in sequence preference over time can be rationalized in terms of prior knowledge about protein evolution.
-%As proteins evolve, there can be epistatic shifts in the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}.
-%Such epistasis has been experimentally demonstrated for HA~\cite{das2013defining,myers2013compensatory}, including at some epitope sites in H3 HA~\cite{wu2017diversity}.
-%It is also known that immune pressure can increase epistasis among substitutions~\cite{gong2014epistatically}.
-%We hypothesize that immune pressure increases the accumulation of epistatic interactions involving substitutions at epitope sites.
-%Over time, these epistatic interactions could manifest as shifts in the effects of mutations, leading to an apparent increase in per-site sequence preference over time when mutational effects are measured in the Perth/2009 HA (Figure~\ref{fig:sequence_preference}B).
-%Such a trend is not apparent for non-epitope sites, perhaps because they are under less selection to fix adaptive mutations.
-
 \subsection*{There are large differences between H3 and H1 HAs in the amino-acid preferences of many sites}
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=\columnwidth]{figs/WSN_trunkvssidebranch/WSN_trunkvssidebranch.pdf}
-%\caption{\label{fig:WSN_trunkvssidebranch}
-%{\bf Experimental measurements on an H1 HA do \emph{not} provide evolutionarily relevant information for H3 HAs.}
-%(A), (B) This figure is analogous to that in Figure~\ref{fig:trunkvssidebranch} except that it scores the H3 sequences using experimental measurements made on the lab-adapted WSN/1933 HA~\cite{doud2016accurate}.
-%}
-%\end{figure}
-
-An obvious hypothesis for why the H1 deep mutational scanning do not reveal differences among H3 strains is that the effect of the same mutation is often different between these two HA subtypes.
+An obvious hypothesis for why the H1 deep mutational scanning is not useful for understanding the evolution of H3N2 influenza viruses is that the effect of the same mutation is often different between these two HA subtypes.
 To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
@@
 \label{sec:discussion}
-We have measured the effects of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements are informative for distinguishing the evolutionary fate of human H3N2 viral strains in nature.
-Specifically, HA sequences on the trunk of the phylogenetic tree tend to have mutations that our experiments measure to be more beneficial for viral growth than mutations that occur in side-branch HA sequences.
-The fact that our experiments can help distinguish between trunk and side-branch lineages suggests that they might inform evolutionary forecasting.
+We have measured the effects of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements have some value for understanding the evolutionary fate of human H3N2 viral strains in nature.
+Specifically, viral clades that reach high frequency tend to have mutations that our experiments measure to be more beneficial for viral growth than the mutations that occur in clades that quickly die out.
+The fact that our experiments can be used to help identify some of the features of successful viral clades suggests that they might inform evolutionary forecasting.
 In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, {\L}uksza and L{\"a}ssig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
-Our work suggests that deep mutational scanning can provide such data.
+Our work suggests that deep mutational scanning may be able to provide such data.
 
@@
 Yet, prior to our work, there were no comprehensive studies of the functional effects of mutations to H3 HA on any property that even resembled viral fitness in nature, and modeling work has either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or assumed that all non-epitope mutations had equivalent deleterious effects~\cite{luksza2014predictive}.
-The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative --- and represent a vast improvement over no information at all.
-That this is true is demonstrated by the fact that these measurements have utility for distinguishing successful and unsuccessful viral lineages over $\sim$50 years of H3N2 evolution.
-We suspect that performing the experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility.
+The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative --- and so represent an improvement over no information at all.
+We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility.
 
@@
 Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}, and the extent to which this occurs is an important question in its own right.
-The fact that our measurements score trunk mutations more favorably than side branch ones across a 50-year timespan suggests that the mutational effects have not shifted too dramatically among these relatively closely related sequences.
+Our work suggests that measurements on a HA from single human H3N2 viral strain can be usefully generalized to at least some extent across the entire evolutionary history of human H3N2 HA.
 On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2017mapping}.
@@
 Our work did not characterize the antigenic effects of mutations, which also play an important role in determining strain success in nature~\cite{koel2013substitutions,neher2016prediction}.
-%Indeed, some of our results (\textit{e.g.}, Figure~\ref{fig:sequence_preference}) suggest as much by showing that the experiments better capture selection at non-epitope than epitope sites.
 However, our basic selection and deep-sequencing approach can be harnessed to completely map how mutations affect antibody recognition~\cite{doud2017complete,doud2018quantifying}.
@@
 
-\subsection*{Inference of human H3N2 phylogenetic tree}
-We applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
+\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum clade frequencies}
+To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
 We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
@@
 We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
-
-To parse trunk mutations from side-branch mutations, we defined a set of recent nodes sampled on or after January 1, 2017, and traced these nodes back to their most recent common ancestor.
-We defined all branches ancestral to this most recent common ancestor as the trunk.
-All nodes that do not have descendants isolated after 2017 were defined as side-branch nodes.
-Nodes descended from the most recent common ancestor were defined as unresolved, as it is still unclear whether these nodes will ultimately be on the trunk or side branches.
 The phylogenetic tree is available as a JSON file on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
 The tree was visualized using baltic (\url{https://github.com/blab/baltic}).
+\comment{John: can you update this section to be accurate?}
 
-\subsection*{Quantification of mutational effects and sequence preferences from an H3N2 phylogeny}
+\comment{John: can you add a section on calculation of max clade frequencies?}
+
+\subsection*{Quantification of mutational effects}
 We calculated the effects of mutations from the amino-acid preferences that were estimated from the deep mutational scanning data.
@@
 The WSN/1933 H1 HA amino-acid preferences are the replicate-average values reported in \cite{doud2016accurate}, re-scaled by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
-The mutational effects were also calculated for five-year windows from 1968 to 2012.
-
-The average sequence preference per site for every node on the phylogenetic tree shown in Figure~\ref{fig:H3N2_phylogeny} was calculated using Equation 1 for all sites, or the set of epitope and non-epitope sites defined in~\cite{wolf2006long} and provided in Dataset~\ref{suppdata:prefsfiles}.
 
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.75\textwidth]{figs/S04_post_Perth2009/post_Perth2009_seqpref.pdf}}
-\caption{\label{suppfig:future}
-{\bf The sequence preference of nodes that occur after the split of Perth/2009 off the trunk.}
-Our experiments show a selective advantage for the trunk even if we limit the analysis to sequences that occur after the split of Perth/2009 off the trunk.
-(A) Phylogenetic tree showing all nodes that branch from the trunk before or contemporaneously with Perth/2009 (orange), all nodes that occur after the branch to Perth/2009 for which we can resolve the trunk (red) or side branches (blue), and all nodes for which it is not yet clear which sequences will be on the trunk or side branches (gray).
-In this figure, we restrict the analysis to the nodes in red and blue.
-(B)  
-Average per-site sequence preference (similar to Figure~\ref{fig:trunkvssidebranch}C) for all trunk and side-branch nodes that occur after the branch to Perth/2009.
-(C)
-The sequence preferences of nodes on the trunk generally exceed those of nodes on the side branches along the post-Perth/2009 portion of the tree.
-(D)
-Post-Perth/2009 trunk mutations on average tend to be more favorable than side-branch mutations.
-This effect is of similar magnitude to pre-Perth/2009 mutations (see Figure~\ref{fig:sequence_preference}) but is not statistically significant because there are relatively few post-Perth/2009 mutations (13 trunk mutations and 324 side-branch mutations).
+\centerline{\includegraphics[clip=true, trim=0 5.5in 0 0,width=0.7\textwidth]{figs/S04_post_Perth2009/post_Perth2009_seqpref.pdf}}
+\caption{\label{suppfig:tree}
+{\bf A phylogenetic tree of all HA sequences used in our analysis of clade frequencies.}
+The Perth/2009 strain used in our experiments is indicated.
+The rest of the tree is partitioned into nodes that preceded the split of the Perth/2009 strain from the trunk of the tree, and nodes that branches off the trunk after the clade containing Perth/2009.
+In Figure~\ref{fig:muteffect_maxfreq}, these two partitions of the tree are analyzed separately.
+\comment{John: can you make an updated tree that shows the current sequence set being used and partitions into pre- and post-Perth, as well as the Perth clade itself.
+Maybe you can make the colors match Figure~\ref{fig:muteffect_maxfreq}.}
 }
@@
 Additional tabs give the across-replicate averaged and re-scaled amino-acid preferences in sequential numbering and in H3 numbering as shown in Figure~\ref{fig:logoplot}.
-There are also tabs that give the conversion from sequential to H3 numbering, and the epitope sites in H3 numbering as defined by Wolf et al and used in this study.
+There are also tabs that give the conversion from sequential to H3 numbering. 
+\comment{Juhye, get rid of this part: and the epitope sites in H3 numbering as defined by Wolf et al and used in this study.}
 Each tab can simply be exported to CSV for computational analyses.

### 2018-03-29 paper/manuscript.tex
Addressed comments for Juhye, cleaned up supp data file

@@
 For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
-Helix A is a component of the epitope of many broadly neutralizing anti-stalk antibodies\comment{cite something}, and interestingly, there are more reports of readily selecting escape mutants from such anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.
+Helix A is a component of the epitope of many broadly neutralizing anti-stalk antibodies~\cite{mallajosyula2014influenza,laursen2013broadly,chai2016two}, and interestingly, there are more reports of readily selecting escape mutants from such anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.
 
@@
 (B) The mutational tolerance of solvent-exposed residues in the head and stalk domains of the Perth/2009 H3 HA (purple) and WSN/1933 H1 HA (gold). 
-\comment{Juhye: Move this sentence to Method: We quantified the absolute solvent accessibility of each site using DSSP~\cite{kabsch1983} and then normalized to a relative solvent accessibility using the absolute accessibilities in~\cite{tien2013}.}
 Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
@@
 \centering
-\includegraphics[width=\textwidth]{figs/muteffect_by_maxfreq_WSN.pdf}
+\includegraphics[width=\textwidth]{figs/muteffect_by_maxfreq_WSN_1.pdf}
 \caption{\label{fig:muteffect_maxfreq_WSN}
@@
 This figure repeats the analysis of the H3N2 clade frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
-Figure \comment{?} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
-\comment{Juhye: Maybe you can make a template for this. Just trim this figure to only show the correlations, and then put the histogram data as a supplement.}
+Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
 }
@@
 
+\subsection*{Quantification of relative solvent accessibility}
+We quantified the absolute solvent accessibility of each site of the H3 HA (PDB 4O5N;~\cite{lee2014receptor}) or the H1 HA (PDB 1RVX;~\cite{gamblin2004structure}) structure using DSSP~\cite{kabsch1983}.
+We then normalized to a relative solvent accessibility using the absolute accessibilities in~\cite{tien2013}.
+
 \subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum clade frequencies}
@@
 
+\begin{suppfigure}[H]
+\centerline{\includegraphics[width=0.75\textwidth]{figs/muteffect_by_maxfreq_WSN_2.pdf}}
+\caption{\label{suppfig:muteffect_maxfreq_WSN_supp}
+{\bf The distribution of mutational effects measured in H1 HA among H3N2 clades binned by the maximum frequency that they reach.}
+This figure repeats the analysis of the H3N2 clade frequencies in Figure~\ref{fig:muteffect_maxfreq}B, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
+}
+\end{suppfigure}
+
 \begin{suppdata}[H]
@@
 There are also tabs that give the conversion from sequential to H3 numbering. 
-\comment{Juhye, get rid of this part: and the epitope sites in H3 numbering as defined by Wolf et al and used in this study.}
 Each tab can simply be exported to CSV for computational analyses.

### 2018-03-29 paper/manuscript.tex
few small tweaks

@@
 A major goal in the study of rapidly evolving viruses such as influenza is to understand the features that distinguish successful and unsuccessful clades, as this can help us identify clades that are likely to reach high frequencies from those that will quickly die out~\cite{lassig2017predicting,morris2017predictive}.
-The frequencies that viral strains reach nature depend in large part on their fitnesses~\cite{strelkowa2012clonal,neher2014predicting,koelle2015effects}.
+The frequency that a viral strain reaches in nature depends in large part on its fitness~\cite{strelkowa2012clonal,neher2014predicting,koelle2015effects}.
 Our experiments measure viral growth in cell culture, which is a partial albeit imperfect proxy for viral fitness.
@@
 Specifically, we examined the relationship between our experimental measurements and the maximum frequency reached by the clade descended from each branch on the tree on which an amino-acid mutation has occurred.
-\comment{Trevor: is there some general paper we should cite for the concept of clade frequencies, or should we just cite as here?}
+\comment{Trevor: is there some general paper we should cite for the concept of clade frequencies, or should we just state what we did without a citation?}
 This concept of relating the experimentally measured effects of mutations to the maximum clade frequency is illustrated in Figure~\ref{fig:frequency_trajectory} for sequences isolated between 2004 and 2014.
@@
 (B) The distribution of mutational effects among clades binned by the maximum frequency that they reach.
-Note that the analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{fig:tree}).
+Note that the analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
 \comment{Split the top and bottom into A and B, adjust font sizes (particularly, tick labels too big), adjust capitalization to be consistent with other figures (which usually start lower-case unless proper noun), increasing spacing between A and B panels.
@@
 
-\subsection*{Measurements made for an H1 HA are not useful for understanding the evolution of H3 influenza}
+\subsection*{Measurements made on an H1 HA are not useful for understanding the evolution of H3 influenza}
 To determine how broadly experimental measurements can be generalized across HAs, we repeated the clade frequency analyses using mutational effects from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.

### 2018-03-29 paper/manuscript.tex
elaborated on comments to @jlhudd in manuscript text

@@
 \comment{John: Can you update figure?
-In particular, maybe better colors.
-Also, let's increase the size of the legend and make it $<-10$, and $> 0$ rather than having the \emph{inf} in the labels.
-Also, can you make y-axis at bottom ``clade frequency.''}
+We defer to you and Trevor on choice of colors, I know there was some Slack discussion of this but can't remember answer.
+But definitely let's increase the size of the legend and make it $<-10$, and $> 0$ rather than having the \emph{inf} in the labels.
+Also, can you make y-axis at bottom ``clade frequency'' rather than ``mutation frequency'' as we use that terminology in the text instead.}
 }
@@
 Note that the analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
-\comment{Split the top and bottom into A and B, adjust font sizes (particularly, tick labels too big), adjust capitalization to be consistent with other figures (which usually start lower-case unless proper noun), increasing spacing between A and B panels.
-Also, no need for a title at top and figure can probably get smaller.}
+\comment{John: label the top and bottom into A and B as indicated in the legend above, adjust font sizes (particularly, tick labels too big), adjust capitalization to be consistent with other figures (which usually start lower-case unless proper noun), increasing spacing between A and B panels.
+Also, no need for a title at top and figure can probably get smaller.
+Some of this might be best done in Inkscape working with Juhye.}
 }
@@
 Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
+\comment{John: Can you match any formatting changes that you make to Figure~\ref{fig:muteffect_maxfreq}A here so that they look consistent.}
 }
@@
 This figure repeats the analysis of the H3N2 clade frequencies in Figure~\ref{fig:muteffect_maxfreq}B, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
+\comment{John: Can you match any formatting changes that you make to Figure~\ref{fig:muteffect_maxfreq}B here so that they look consistent.}
 }

### 2018-03-30 paper/manuscript.tex
Edits to maximum frequency and mutational effects section and methods.

--- 2018-03-29 paper/manuscript.tex
@@
 \subsection*{Our measurements can help distinguish between viral clades that reach low and high frequencies in nature}
-A major goal in the study of rapidly evolving viruses such as influenza is to understand the features that distinguish successful and unsuccessful clades, as this can help us identify clades that are likely to reach high frequencies from those that will quickly die out~\cite{lassig2017predicting,morris2017predictive}.
-The frequency that a viral strain reaches in nature depends in large part on its fitness~\cite{strelkowa2012clonal,neher2014predicting,koelle2015effects}.
-Our experiments measure viral growth in cell culture, which is a partial albeit imperfect proxy for viral fitness.
-We therefore investigated whether these measurements could help distinguish among strains that do and do not reach high frequencies during natural evolution.
+A major goal in the study of rapidly evolving viruses such as influenza is to understand the features that distinguish successful and unsuccessful clades~\cite{lassig2017predicting,morris2017predictive}.
+The frequency that a clade reaches in nature depends in large part on the fitness of its component viral strains~\cite{strelkowa2012clonal,neher2014predicting,koelle2015effects}.
+Our experiments provide a potential proxy for viral fitness, by measuring viral growth in cell culture.
+We therefore investigated whether these measurements could help distinguish clades that do and do not reach high frequencies during natural evolution.
 
-Specifically, we examined the relationship between our experimental measurements and the maximum frequency reached by the clade descended from each branch on the tree on which an amino-acid mutation has occurred.
+Specifically, we examined the relationship between the global frequency of naturally-occuring clades and the sum of mutational effects on the branch leading to each clade.
+We constructed a phylogenetic tree of H3N2 sequences sampled at a rate of six viruses per month between 1968 and 2018 (Figure~\ref{suppfig:tree}).
+From this tree, we estimated the frequency of each node at each time point in the sampled interval.
 \comment{Trevor: is there some general paper we should cite for the concept of clade frequencies, or should we just state what we did without a citation?}
-This concept of relating the experimentally measured effects of mutations to the maximum clade frequency is illustrated in Figure~\ref{fig:frequency_trajectory} for sequences isolated between 2004 and 2014.
-If a mutation is deleterious, we would expect the clade descended from the branch containing that mutation to quickly die out, whereas branches with beneficial mutations are more likely to lead to clades that expand in frequency.
-Qualitative inspection of Figure~\ref{fig:frequency_trajectory} suggests that mutations that are experimentally measured to be neutral or beneficial are more likely than those measured to be deleterious to occur on branches leading to high frequency clades.
+We identified all mutations on the branch leading to each node and assigned a single mutational effect value to each node based on the sum of effects for each mutation on the branch.
+We anticipated that clades with deleterious mutations represented by low mutational effects would never reach high global frequencies.
+In contrast, we expected clades with beneficial mutations represented by high mutational effects to dominate globally and fix.
+Correspondingly, we observed few clades in our phylogeny with strongly deleterious mutational effects and rarely observed these clades reaching high frequency (Figure~\ref{fig:frequency_trajectory}).
+Clades that reached high global frequencies generally had neutral or beneficial mutational effects.
 
@@
 It is obvious that most clades that reach high frequency have mutations that are measured to be relatively favorable in our experiments.
-\comment{John: Can you update figure?
-In particular, maybe better colors.
-Also, let's increase the size of the legend and make it $<-10$, and $> 0$ rather than having the \emph{inf} in the labels.
-Also, can you make y-axis at bottom ``clade frequency.''}
 }
@@
 
-To analyze the relationship between mutational effect and clade frequency more quantitatively, we next partitioned the phylogeny of human H3N2 influenza HA into sequences that split from the line of descent (trunk of the phylogenetic tree) before or after the Perth/2009 strain used in our experiments (see Figure~\ref{suppfig:tree}).
-The rationale for partitioning into pre- and post-Perth/2009 sequences is that we wanted to test the utility of our measurements for prospective as well as post-hoc analyses.
-\comment{John: do you have a sentence to add about how we used max tips to avoid passaging artifacts?}
-Figure~\ref{fig:muteffect_maxfreq} shows that there is a statistically significant trend for clades that reach high frequency to have more favorable mutations than clades that quickly die out.
-Importantly, the effect size is similar for both the pre- and post-Perth/2009 sequence groups.
+We further quantified this relationship by calculating a Spearman rank correlation between mutational effects and maximum global frequencies of clades in the H3N2 phylogeny.
+To minimize effects related to the genetic background of the DMS strain, we excluded clades closely related to the DMS strain itself and partitioned the remaining clades into those pre- and post-dating the Perth/2009 strain.
+We additionally excluded clades from the post-Perth partition that were sampled in 2014 or after and that have not had enough evolutionary time to reach fixation.
+We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
+In both pre-Perth and post-Perth clades, we found a modest, statistically significant relationship between maximum clade frequency and mutational effect (Figure~\ref{fig:muteffect_maxfreq}A).
+The similar effect size for both partitions suggests that our experimental measurements can be used to predict clade frequencies despite the genetic background of the DMS strain.
 
@@
 (A) The correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by viral clades.
-The plots show the Spearman correlation $\rho$ and a $P$-value obtained by repeatedly randomizing the experimental measurements among sites \comment{John: is this right about how the $P$-value was calculated?}.
+The plots show the Spearman correlation $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which the Spearman correlation $\rho$ between maximum frequency and randomized mutational effects was greater than or equal to the observed $\rho$.
 (B) The distribution of mutational effects among clades binned by the maximum frequency that they reach.
 Note that the analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
-\comment{Split the top and bottom into A and B, adjust font sizes (particularly, tick labels too big), adjust capitalization to be consistent with other figures (which usually start lower-case unless proper noun), increasing spacing between A and B panels.
-Also, no need for a title at top and figure can probably get smaller.}
 }
@@
 {\bf A phylogenetic tree of all HA sequences used in our analysis of clade frequencies.}
+HA sequences were sampled at a rate of six viruses per month from January 1, 1968 through February 1, 2018.
 The Perth/2009 strain used in our experiments is indicated.
-The rest of the tree is partitioned into nodes that preceded the split of the Perth/2009 strain from the trunk of the tree, and nodes that branches off the trunk after the clade containing Perth/2009.
+The rest of the tree is partitioned into nodes that preceded the split of the Perth/2009 strain from the trunk of the tree (blue) and nodes that branched off the trunk after the clade containing Perth/2009 (orange).
 In Figure~\ref{fig:muteffect_maxfreq}, these two partitions of the tree are analyzed separately.
-\comment{John: can you make an updated tree that shows the current sequence set being used and partitions into pre- and post-Perth, as well as the Perth clade itself.
-Maybe you can make the colors match Figure~\ref{fig:muteffect_maxfreq}.}
+Nodes in the clade containing the Perth/2009 strain and nodes sampled in 2014 or after were excluded from our analyses.
 }

### 2018-03-30 paper/manuscript.tex
TB edits

--- 2018-03-29 paper/manuscript.tex
@@
 
-\title{Deep mutational scanning of hemagglutinin helps predict the evolutionary fates of human H3N2 influenza virus variants}
+\title{Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants}
 
@@
 \authordeclaration{The authors declare no conflict of interest.}
-\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed: trevor@bedford.io, jbloom@fredhutch.org}
+\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed: tbedford@fredhutch.org, jbloom@fredhutch.org}
 
@@
 Here we experimentally measure the effects on viral growth in cell culture of all single amino-acid mutations to the HA from a recent human H3N2 influenza virus strain.
-We then show that mutations in evolutionarily successful H3N2 viral lineages are measured to generally be more favorable than mutations in lineages that quickly die out.
+We show that mutations that are measured to be more favorable in cell culture tend to result in more evolutionarily successful H3N2 viral lineages than mutations that are less favorable in cell culture.
 Therefore, despite the well-known caveats about cell-culture measurements of viral fitness, such measurements can still be informative for understanding evolution in nature.
-We also compare our measurements for an H3 HA to similar data previously generated for a distantly related H1 HA, and find substantial differences in which amino acids are preferred at many sites.
+We also compare our measurements for H3 HA to similar data previously generated for a distantly related H1 HA, and find substantial differences in which amino acids are preferred at many sites.
 For instance, the H3 HA has less disparity in mutational tolerance between the head and stalk domains than the H1 HA.
@@
 
-\dropcap{S}easonal H3N2 influenza virus evolves rapidly, fixing 3 to 4 amino-acid mutations per year in its hemagglutinin (HA) surface protein.
-Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping,bhatt2011genomic}.
-This evolution is further characterized by competition and turnover among groups of strains or clades~\cite{fitch1997long,strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects,bedford2015global}.
-As a result, mutations that differentiate clades exhibit dynamic frequency trajectories as these clades appear, circulate, or die out.
-Several lines of evidence indicate that successful clades, which reach high frequencies or even fix in the global population, have higher fitness than clades that remain at low frequency~\cite{strelkowa2012clonal,bedford2011,neher2014predicting,koelle2015effects}.
+\dropcap{S}easonal H3N2 influenza virus evolves rapidly, fixing 3 to 4 amino-acid mutations per year in its hemagglutinin (HA) surface protein~\cite{fitch1997long, bhatt2011genomic}.
+Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping}.
+This evolution is further characterized by competition and turnover among groups of strains or clades bearing different complements of mutations~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,koelle2015effects,bedford2015global}.
+Clades vary widely in their evolutionary success with some dieing out soon after emergence and others going on to take over the virus population.
+Several lines of evidence indicate that successful clades, which reach high frequencies or even fix in the global population, have higher fitness than clades that remain at low frequency~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,luksza2014predictive}.
 A key goal in the study of H3N2 evolution is to identify the features that enable certain clades to succeed as others die out.
@@
 However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
-Here we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
+Here, we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
 We then show that these experimental measurements can help discriminate evolutionarily successful viral clades from lineages that quickly die out.
@@
 \subsection*{Our measurements can help distinguish between viral clades that reach low and high frequencies in nature}
-A major goal in the study of rapidly evolving viruses such as influenza is to understand the features that distinguish successful and unsuccessful clades, as this can help us identify clades that are likely to reach high frequencies from those that will quickly die out~\cite{lassig2017predicting,morris2017predictive}.
-The frequency that a viral strain reaches in nature depends in large part on its fitness~\cite{strelkowa2012clonal,neher2014predicting,koelle2015effects}.
-Our experiments measure viral growth in cell culture, which is a partial albeit imperfect proxy for viral fitness.
-We therefore investigated whether these measurements could help distinguish among strains that do and do not reach high frequencies during natural evolution.
-
-Specifically, we examined the relationship between our experimental measurements and the maximum frequency reached by the clade descended from each branch on the tree on which an amino-acid mutation has occurred.
-\comment{Trevor: is there some general paper we should cite for the concept of clade frequencies, or should we just state what we did without a citation?}
-This concept of relating the experimentally measured effects of mutations to the maximum clade frequency is illustrated in Figure~\ref{fig:frequency_trajectory} for sequences isolated between 2004 and 2014.
-If a mutation is deleterious, we would expect the clade descended from the branch containing that mutation to quickly die out, whereas branches with beneficial mutations are more likely to lead to clades that expand in frequency.
-Qualitative inspection of Figure~\ref{fig:frequency_trajectory} suggests that mutations that are experimentally measured to be neutral or beneficial are more likely than those measured to be deleterious to occur on branches leading to high frequency clades.
+Mutations occurring in the H3N2 population have widely varying evolutionary outcomes~\ref{fig:frequency_trajectory}.
+Some mutations appear, spread and fix in the population, while others briefly circulate before dieing out.
+We take the maximum frequency reached by a mutation as a coarse indicator of its fitness, under the assumption that fitter mutations are expected to reach higher frequency on average than less fit mutations \cite{ewens2012mathematical}.
+If a mutation is deleterious, we would expect it to quickly die out, whereas beneficial mutations are more likely to expand in frequency.
+Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, i.e.\ although R142G occurs multiple times on the phylogeny we track each of these mutations separately.
+As such, each mutation is shown as a separate circle occuring on a separate branch in Figure \ref{fig:frequency_trajectory}.
 
@@
 
+Here, we examine to what degree our experimental measurements correlate with realized evolutionary outcomes.
+Qualitative inspection of Figure~\ref{fig:frequency_trajectory} suggests that mutations that are experimentally measured to be neutral or beneficial are more likely than those measured to be deleterious to occur on branches leading to high frequency clades.
 To analyze the relationship between mutational effect and clade frequency more quantitatively, we next partitioned the phylogeny of human H3N2 influenza HA into sequences that split from the line of descent (trunk of the phylogenetic tree) before or after the Perth/2009 strain used in our experiments (see Figure~\ref{suppfig:tree}).

### 2018-03-31 paper/manuscript.tex

@@
 If a mutation is deleterious, we would expect it to quickly die out, whereas beneficial mutations are more likely to expand in frequency.
+We anticipated that clades with deleterious mutations represented by low mutational effects would never reach high global frequencies. 
 Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, i.e.\ although R142G occurs multiple times on the phylogeny we track each of these mutations separately.
 As such, each mutation is shown as a separate circle occuring on a separate branch in Figure \ref{fig:frequency_trajectory}.
+We identified all mutations on a phylogeny branch and assigned a single mutational effect value based on the sum of effects for each mutation on the branch.
+
+In contrast, we expected clades with beneficial mutations represented by high mutational effects to dominate globally and fix. 
+Correspondingly, we observed few clades in our phylogeny with strongly deleterious mutational effects and rarely observed these clades reaching high frequency (Figure~\ref{fig:frequency_trajectory}). 
+Clades that reached high global frequencies generally had neutral or beneficial mutational effects.
 
@@
 It is obvious that most clades that reach high frequency have mutations that are measured to be relatively favorable in our experiments.
-\comment{John: Can you update figure?
-We defer to you and Trevor on choice of colors, I know there was some Slack discussion of this but can't remember answer.
-But definitely let's increase the size of the legend and make it $<-10$, and $> 0$ rather than having the \emph{inf} in the labels.
-Also, can you make y-axis at bottom ``clade frequency'' rather than ``mutation frequency'' as we use that terminology in the text instead.}
 }
@@
 
-Here, we examine to what degree our experimental measurements correlate with realized evolutionary outcomes.
+We further quantified this relationship by calculating a Spearman rank correlation between mutational effects and maximum global frequencies of clades in the H3N2 phylogeny.
 Qualitative inspection of Figure~\ref{fig:frequency_trajectory} suggests that mutations that are experimentally measured to be neutral or beneficial are more likely than those measured to be deleterious to occur on branches leading to high frequency clades.
-To analyze the relationship between mutational effect and clade frequency more quantitatively, we next partitioned the phylogeny of human H3N2 influenza HA into sequences that split from the line of descent (trunk of the phylogenetic tree) before or after the Perth/2009 strain used in our experiments (see Figure~\ref{suppfig:tree}).
-The rationale for partitioning into pre- and post-Perth/2009 sequences is that we wanted to test the utility of our measurements for prospective as well as post-hoc analyses.
-\comment{John: do you have a sentence to add about how we used max tips to avoid passaging artifacts?}
-Figure~\ref{fig:muteffect_maxfreq} shows that there is a statistically significant trend for clades that reach high frequency to have more favorable mutations than clades that quickly die out.
-Importantly, the effect size is similar for both the pre- and post-Perth/2009 sequence groups.
+To minimize effects related to the genetic background of the DMS strain, we excluded clades closely related to the DMS strain itself and partitioned the remaining clades into those pre- and post-dating the Perth/2009 strain.
+We additionally excluded clades from the post-Perth partition that were sampled in 2014 or after and that have not had enough evolutionary time to reach fixation.
+We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
+In both pre-Perth and post-Perth clades, we found a modest, statistically significant relationship between maximum clade frequency and mutational effect (Figure~\ref{fig:muteffect_maxfreq}A).
+The similar effect size for both partitions suggests that our experimental measurements can be used to predict clade frequencies despite the genetic background of the DMS strain.
 
@@
 (A) The correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by viral clades.
-The plots show the Spearman correlation $\rho$ and a $P$-value obtained by repeatedly randomizing the experimental measurements among sites \comment{John: is this right about how the $P$-value was calculated?}.
+The plots show the Spearman correlation $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which the Spearman correlation $\rho$ between maximum frequency and randomized mutational effects was greater than or equal to the observed $\rho$.
 (B) The distribution of mutational effects among clades binned by the maximum frequency that they reach.
 Note that the analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
-\comment{John: label the top and bottom into A and B as indicated in the legend above, adjust font sizes (particularly, tick labels too big), adjust capitalization to be consistent with other figures (which usually start lower-case unless proper noun), increasing spacing between A and B panels.
-Also, no need for a title at top and figure can probably get smaller.
-Some of this might be best done in Inkscape working with Juhye.}
 }
@@
 This figure repeats the analysis of the H3N2 clade frequencies in Figure~\ref{fig:muteffect_maxfreq}B, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
-\comment{John: Can you match any formatting changes that you make to Figure~\ref{fig:muteffect_maxfreq}B here so that they look consistent.}
 }

### 2018-03-31 paper/manuscript.tex
Further revisions of Results

@@
 Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping}.
-This evolution is further characterized by competition and turnover among groups of strains or clades bearing different complements of mutations~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,koelle2015effects,bedford2015global}.
+This evolution is further characterized by competition and turnover among groups of strains called clades bearing different complements of mutations~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,koelle2015effects,bedford2015global}.
 Clades vary widely in their evolutionary success with some dieing out soon after emergence and others going on to take over the virus population.
@@
 Here, we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
-We then show that these experimental measurements can help discriminate evolutionarily successful viral clades from lineages that quickly die out.
+We then show that these experimental measurements can help discriminate evolutionarily successful mutations from those that quickly die out.
 However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are not useful for understanding the evolutionary fate of H3 viral strains.
@@
 
-\subsection*{Our measurements can help distinguish between viral clades that reach low and high frequencies in nature}
-Mutations occurring in the H3N2 population have widely varying evolutionary outcomes~\ref{fig:frequency_trajectory}.
+\subsection*{Our measurements can help distinguish between mutations that reach low and high frequencies in nature}
+Mutations occurring in the H3N2 virus population have widely varying evolutionary outcomes (Figure~\ref{fig:frequency_trajectory}).
 Some mutations appear, spread and fix in the population, while others briefly circulate before dieing out.
 We take the maximum frequency reached by a mutation as a coarse indicator of its fitness, under the assumption that fitter mutations are expected to reach higher frequency on average than less fit mutations \cite{ewens2012mathematical}.
-If a mutation is deleterious, we would expect it to quickly die out, whereas beneficial mutations are more likely to expand in frequency.
-We anticipated that clades with deleterious mutations represented by low mutational effects would never reach high global frequencies. 
-Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, i.e.\ although R142G occurs multiple times on the phylogeny we track each of these mutations separately.
+If a mutation is highly deleterious, we would expect it to quickly die out, whereas less deleterious or beneficial mutations are more likely to expand in frequency.
+Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, i.e.\ although R142G occurs multiple times on the phylogeny we track each of these mutations occurring on different backgrounds separately.
 As such, each mutation is shown as a separate circle occuring on a separate branch in Figure \ref{fig:frequency_trajectory}.
-We identified all mutations on a phylogeny branch and assigned a single mutational effect value based on the sum of effects for each mutation on the branch.
-
-In contrast, we expected clades with beneficial mutations represented by high mutational effects to dominate globally and fix. 
-Correspondingly, we observed few clades in our phylogeny with strongly deleterious mutational effects and rarely observed these clades reaching high frequency (Figure~\ref{fig:frequency_trajectory}). 
-Clades that reached high global frequencies generally had neutral or beneficial mutational effects.
+However, because multiple mutations on the same phylogeny branch cannot be disentangled, we identified all mutations on a branch and assigned a single mutational effect based on the sum of effects for each mutation on the branch.
+We observed few circulating mutations in our phylogeny with strongly deleterious mutational effects and rarely observed these mutations reaching high frequency (Figure~\ref{fig:frequency_trajectory}).
+Mutations that reached high global frequencies generally had neutral or beneficial mutational effects.
 
@@
 \caption{\label{fig:frequency_trajectory}
-{\bf Frequency trajectories of viral clades and their relation to the effects of individual mutations.}
+{\bf Frequency trajectories of individual mutations and their relation to the experimentally determined effects of these mutations.}
 The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} windowed from 2004 to 2014.
-Circles indicate amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
-The bottom panel shows the frequency trajectory of the clade downstream of each mutation, with clades colored according to the mutational effects as in the top panel.
-It is obvious that most clades that reach high frequency have mutations that are measured to be relatively favorable in our experiments.
+Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
+The bottom panel shows the frequency trajectory of each mutation, with mutation trajectories colored according to the mutational effects as in the top panel.
+It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
 }
@@
 
-We further quantified this relationship by calculating a Spearman rank correlation between mutational effects and maximum global frequencies of clades in the H3N2 phylogeny.
-Qualitative inspection of Figure~\ref{fig:frequency_trajectory} suggests that mutations that are experimentally measured to be neutral or beneficial are more likely than those measured to be deleterious to occur on branches leading to high frequency clades.
-To minimize effects related to the genetic background of the DMS strain, we excluded clades closely related to the DMS strain itself and partitioned the remaining clades into those pre- and post-dating the Perth/2009 strain.
-We additionally excluded clades from the post-Perth partition that were sampled in 2014 or after and that have not had enough evolutionary time to reach fixation.
+We next sought to quantify the correlation between mutation effects and maximum frequencies attained.
+To minimize effects related to the genetic background of the DMS strain, we excluded mutations closely related to the DMS strain itself and partitioned the remaining mutations into XXX \comment{John: please add number} mutations pre-dating and XXX \comment{John: please add number} mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
+We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
 We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
-In both pre-Perth and post-Perth clades, we found a modest, statistically significant relationship between maximum clade frequency and mutational effect (Figure~\ref{fig:muteffect_maxfreq}A).
-The similar effect size for both partitions suggests that our experimental measurements can be used to predict clade frequencies despite the genetic background of the DMS strain.
+We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
+In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).
+The similar effect size for both partitions suggests that our experimental measurements can be used to predict mutation frequencies despite the genetic background of the DMS strain.
 
@@
 \caption{\label{fig:muteffect_maxfreq}
-{\bf The experimental measurements are informative about the evolutionary fate of viral clades.}
-(A) The correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by viral clades.
-The plots show the Spearman correlation $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which the Spearman correlation $\rho$ between maximum frequency and randomized mutational effects was greater than or equal to the observed $\rho$.
-(B) The distribution of mutational effects among clades binned by the maximum frequency that they reach.
-Note that the analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
+{\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
+(A) Correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by these mutations.
+The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which permuted $\rho$ was greater than or equal to the observed $\rho$.
+(B) The distribution of mutational effects partitioned by maximum mutation frequency.
+The vertical black line shows the mean \comment{John: or is it median?} mutation effect for each category.
+The analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
 }
@@
 
-The trends in Figure~\ref{fig:muteffect_maxfreq} appear to be most strongly driven by the fact that mutations we measure to be substantially deleterious to viral growth almost never occur in clades that reach high frequency.
-Among clades that reach high frequency, the mutations are overwhelmingly measured to be either beneficial or close to neutral.
-Overall, these results demonstrate that measurements of how mutations affect viral growth in cell culture are informative for understanding the fates of the viral clades in nature: in particular, if a clade has a mutation that is measurably deleterious to viral growth, that clade is unlikely to prosper in nature.
+The trends in Figure~\ref{fig:muteffect_maxfreq}A appear to be most strongly driven by behavior of strongly deleterious mutations.
+We investigated this further by partitioning mutations in those circulating at low, medium and high levels and those that fix in the population (Figure~\ref{fig:muteffect_maxfreq}B).
+We observe both a shift in the mean \comment{John: or is it median?} mutation effect and in the distribution of mutation effects as we increase in circulation category.
+Mutations that we measure to be substantially deleterious almost never reach high frequency.
+Overall, these results demonstrate that measurements of how mutations affect viral growth in cell culture are informative for understanding the fates of these mutations in nature: in particular, if a mutation is measurably deleterious to viral growth, that mutation is unlikely to prosper in nature.
 
 \subsection*{Measurements made on an H1 HA are not useful for understanding the evolution of H3 influenza}
-To determine how broadly experimental measurements can be generalized across HAs, we repeated the clade frequency analyses using mutational effects from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
-Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the H1 measurements do \textit{not} reveal a trend for more beneficial mutations to be enriched in higher-frequency clades of an H3 phylogeny.
-This result is not unsurprising given that H1 and H3 HAs only have 42\% sequence identity, but it does underscore that the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
+To determine how broadly experimental measurements can be generalized across HAs, we repeated the analysis of H3N2 mutation frequencies, but using mutational effects from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
+Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the H1 measurements do \textit{not} reveal a trend for mutation effect on H1 to correlate with H3 maximum mutation frequency.
+This result is not surprising given that H1 and H3 HAs only have 42\% sequence identity, but it does underscore that the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
 
@@
 \caption{\label{fig:muteffect_maxfreq_WSN}
-{\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 viral clades.}
-This figure repeats the analysis of the H3N2 clade frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
+{\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 mutations.}
+This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}, i.e.\
 Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
@@
 \label{sec:discussion}
-We have measured the effects of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements have some value for understanding the evolutionary fate of human H3N2 viral strains in nature.
-Specifically, viral clades that reach high frequency tend to have mutations that our experiments measure to be more beneficial for viral growth than the mutations that occur in clades that quickly die out.
-The fact that our experiments can be used to help identify some of the features of successful viral clades suggests that they might inform evolutionary forecasting.
+We have measured the effects of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements have some value for understanding the evolutionary fate of these mutations in nature.
+Specifically, mutations measured to be more beneficial for viral growth tend to circulate at higher frequency in nature than mutations measured to be more deleterious for viral growth.
+The fact that our experiments can be used to help identify evolutionary successful mutations suggests that they might inform evolutionary forecasting.
 In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, {\L}uksza and L{\"a}ssig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
@@
 
-\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum clade frequencies}
+\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
 To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
@@
 The tree was visualized using baltic (\url{https://github.com/blab/baltic}).
+
 \comment{John: can you update this section to be accurate?}
 
-\comment{John: can you add a section on calculation of max clade frequencies?}
+\comment{John: can you add a section on calculation of max mutation frequencies?}
 
@@
 \caption{\label{suppfig:tree}
-{\bf A phylogenetic tree of all HA sequences used in our analysis of clade frequencies.}
+{\bf A phylogenetic tree of all HA sequences used in our analysis of mutation frequencies.}
 HA sequences were sampled at a rate of six viruses per month from January 1, 1968 through February 1, 2018.
@@
 \caption{\label{suppfig:muteffect_maxfreq_WSN_supp}
-{\bf The distribution of mutational effects measured in H1 HA among H3N2 clades binned by the maximum frequency that they reach.}
-This figure repeats the analysis of the H3N2 clade frequencies in Figure~\ref{fig:muteffect_maxfreq}B, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
+{\bf The distribution of mutational effects measured in H1 HA among H3N2 mutations binned by the maximum frequency that they reach.}
+This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}B, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
 }

### 2018-03-31 paper/manuscript.tex

@@
 \centering
-\includegraphics[width=\textwidth]{figs/muteffect_by_maxfreq.pdf}
+\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq.pdf}
 \caption{\label{fig:muteffect_maxfreq}
@@
 \centering
-\includegraphics[width=\textwidth]{figs/muteffect_by_maxfreq_WSN_1.pdf}
+\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq_WSN_1.pdf}
 \caption{\label{fig:muteffect_maxfreq_WSN}
@@
 
-\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
-To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
-We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
-Ancestral state reconstruction and branch length timing were performed with treetime~\cite{Sagulenko2018}.
-We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
-The phylogenetic tree is available as a JSON file on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
-The tree was visualized using baltic (\url{https://github.com/blab/baltic}).
-
-\comment{John: can you update this section to be accurate?}
-
-\comment{John: can you add a section on calculation of max mutation frequencies?}
-
 \subsection*{Quantification of mutational effects}
@@
 The WSN/1933 H1 HA amino-acid preferences are the replicate-average values reported in \cite{doud2016accurate}, re-scaled by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
+
+\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
+To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid}, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
+We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
+Ancestral state reconstruction and branch length timing were performed with TreeTime~\cite{Sagulenko2018}.
+We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
+The phylogenetic tree is available as a JSON file on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
+The tree was visualized using BALTIC (\url{https://github.com/blab/baltic}).
+
+% notation for frequencies model
+\newcommand{\dx}{\mathrm{d}x}						% change in x
+\newcommand{\dy}{\mathrm{d}y}						% change in y
+\newcommand{\dt}{\mathrm{d}t}						% change in t
+\newcommand{\inertia}{\epsilon}			    % inertia parameter
+\newcommand{\normal}{\mathcal{N}}				% normal distribution
+
+The frequency trajectory of each individual mutation on the phylogeny is estimated following Nextstrain's augur pipeline and as first implemented in Nextflu \cite{neher2015nextflu}.
+Herein, mutation frequency dynamics are modeled according to a Brownian motion diffusion process discretized to one-month intervals.
+Relative to a simple Brownian motion, the expectation includes an `inertia' term $\inertia$ that adds velocity to the diffusion and the variance includes a term $x(1-x)$ to scale variance according to frequency following a Wright-Fisher population genetic process.
+This results in the following diffusion process
+\begin{equation}
+x(t+\dt) = \normal\big( x(t) + \inertia \, \dx , \; \dt \, \sigma^2 \, x(t) \, (1-x(t)) \big),
+\end{equation}
+with `volatility' parameter $\sigma^2$.
+The term $\dx$ is the increment in the previous timestep, so that $\dx = x(t) - x(t-\dt)$.
+We used $\inertia = 0.7$ and $\sigma^2 = 0.05$ to maximize fit to empirical trajectory behavior.
+
+We also include an Bernoulli observation model for mutation presence / absence among sampled viruses at timestep $t$.
+This observation model follows
+\begin{equation}
+f(x,t) = \prod_{v\in V} x(t) \prod_{v\notin V} (1-x(t)),
+\end{equation}
+where $v\in V$ represents the set of viruses that have the mutation and $v\notin V$ represents the set of viruses that do not have the mutation.
+Each frequency trajectory is estimated by simultaneously maximizing the likelihood of the process model and the likelihood of the observation model via adjusting frequency trajectory $\mathbf{x}=(x_1, \ldots, x_n)$.
 

### 2018-03-31 paper/manuscript.tex
Remove unnecessary whitespace at line endings
Move figures to appear on page that they are mentioned in text
some minor tweaks

@@
 Here we experimentally measure the effects on viral growth in cell culture of all single amino-acid mutations to the HA from a recent human H3N2 influenza virus strain.
-We show that mutations that are measured to be more favorable in cell culture tend to result in more evolutionarily successful H3N2 viral lineages than mutations that are less favorable in cell culture.
+We show that mutations that are measured to be more favorable for viral growth are enriched in evolutionarily successful H3N2 viral lineages relative to mutations that are measured to be less favorable for viral growth.
 Therefore, despite the well-known caveats about cell-culture measurements of viral fitness, such measurements can still be informative for understanding evolution in nature.
@@
 This evolution is further characterized by competition and turnover among groups of strains called clades bearing different complements of mutations~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,koelle2015effects,bedford2015global}.
-Clades vary widely in their evolutionary success with some dieing out soon after emergence and others going on to take over the virus population.
+Clades vary widely in their evolutionary success with some dying out soon after emergence and others going on to take over the virus population.
 Several lines of evidence indicate that successful clades, which reach high frequencies or even fix in the global population, have higher fitness than clades that remain at low frequency~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,luksza2014predictive}.
 A key goal in the study of H3N2 evolution is to identify the features that enable certain clades to succeed as others die out.
-
-Two main characteristics distinguish evolutionarily successful clades from their competitors: greater antigenic change, and efficient viral growth and transmission.
-In principle, experiments could be informative for identifying how mutations affect these features.
-Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction,koel2013substitutions,chambers2015identification,li2016selection}.
-While the antigenic features of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
-Specifically, due to influenza virus's high mutation rate~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and lack of intra-segment recombination~\cite{boni2008homologous}, deleterious mutations become linked to beneficial ones.
-The resulting accumulation of deleterious mutations can affect non-antigenic properties that are central to viral fitness~\cite{luksza2014predictive}.
-However, there are no large-scale quantitative characterizations of how mutations to H3N2 HA affect non-antigenic properties crucial to viral fitness.
-
-It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of all single amino-acid mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high}.
-However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
-Here, we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
-We then show that these experimental measurements can help discriminate evolutionarily successful mutations from those that quickly die out.
-However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are not useful for understanding the evolutionary fate of H3 viral strains.
-
-\section*{Results}
-\label{sec:results}
-
-\subsection*{Deep mutational scanning of HA from a recent strain of human H3N2 influenza virus}
-We performed a deep mutational scan to measure the effects of all amino-acid mutations to HA from the A/Perth/16/2009 (H3N2) strain on viral growth in cell culture.
-This strain was the H3N2 component of the influenza vaccine from 2010-2012~\cite{who2010d,who2011}.
-Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral growth in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and Dataset~\ref{suppdata:PerthHA}).
-The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.
 
@@
 \end{SCfigure*}
+
+Two main characteristics distinguish evolutionarily successful clades from their competitors: greater antigenic change, and efficient viral growth and transmission.
+In principle, experiments could be informative for identifying how mutations affect these features.
+Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction,koel2013substitutions,chambers2015identification,li2016selection}.
+While the antigenic features of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
+Specifically, due to influenza virus's high mutation rate~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and lack of intra-segment recombination~\cite{boni2008homologous}, deleterious mutations become linked to beneficial ones.
+The resulting accumulation of deleterious mutations can affect non-antigenic properties that are central to viral fitness~\cite{luksza2014predictive}.
+However, there are no large-scale quantitative characterizations of how mutations to H3N2 HA affect non-antigenic properties crucial to viral fitness.
+
+It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of all single amino-acid mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high,haddox2018mapping}.
+However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
+Here, we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
+We then show that these experimental measurements can help discriminate evolutionarily successful mutations from those that are found in strains that quickly die out.
+However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are not useful for understanding the evolutionary fate of H3 viral strains.
+
+\section*{Results}
+\label{sec:results}
+
+\subsection*{Deep mutational scanning of HA from a recent strain of human H3N2 influenza virus}
+We performed a deep mutational scan to measure the effects of all amino-acid mutations to HA from the A/Perth/16/2009 (H3N2) strain on viral growth in cell culture.
+This strain was the H3N2 component of the influenza vaccine from 2010-2012~\cite{who2010d,who2011}.
+Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral growth in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and Dataset~\ref{suppdata:PerthHA}).
+The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.
 
@@
 
-We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates. 
-We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
-Because there are 566 residues in HA, there are $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
-The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
-The biological replicates were well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78. 
-Replicate 1 exhibited the weakest correlation with other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
-The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity during viral passage contributes most of the experimental noise.
-
-\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
-How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
-This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
-Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
-This result indicates that our experiments authentically capture some of the constraints on HA evolution. 
-The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
-However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for by the constraints measured in the deep mutational scanning.
-The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
-Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
-These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.
-
-\begin{table}
+\begin{table}[b!]
 \caption{\label{tab:phydms}
@@
 
-Examination of Figure~\ref{fig:logoplot} reveals that the experimentally measured amino-acid preferences generally agree with existing knowledge about HA's structure and function.
-For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} strongly prefer cysteine.
-At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such as Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change,yang2015structure}.
-A positively charged amino acid at site 329 is important for cleavage of the HA0 precursor into the mature form~\cite{stech2005new}, and this site strongly prefers arginine.
-However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine. 
-This codon is part of the signal peptide and is cleaved from the mature HA protein.
-One possible reason that our experiments do not show a strong preference for methionine at this site could be alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
-% No upstream or downstream "ATG"'s are present.
-I did find an in-frame "ATC" one codon upstream of the canonical start site (in the non-coding region), and interestingly, all Perth/2009 HA sequences deposited in GenBank have an "ATT" in this non-coding position, including Seema's Perth HA.
-% I6M(HA2) mutant slightly increases the fusion pH (Daniels 1985 Cell)
-% We see high preference for T,S,Q over the wildtype Ile at site 226.
-Although Q226 in an H3 background is characterized as having preference for avian-type receptors, most of these studies were done in older H3 strains such as Aichi.
-Wu 2017 Cell Host Microbe has shown extensive epistasis in the RBS, specifically in the 220-loop, so perhaps the amino acids tolerated in this loop have shifted over time.
-Indeed the wildtype amino acids at 225 and 226 are different in Perth compared to HK68 and Aichi.
-
 \begin{figure*}[ht]
@@
 
-\subsection*{There is less difference in mutational tolerance between the HA head and stalk domains for H3 than for H1}
-Our experiments measure which amino acids are tolerated at each HA site under selection for protein function.
-We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we quantify as the Shannon entropy of the re-scaled preferences.
-In prior mutational studies of H1 HAs, the stalk domain was found to be substantially less mutationally tolerant than the globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,heaton2013genome}.
+We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates.
+We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
+Because there are 566 residues in HA, there are $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
+The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
+The biological replicates were well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78.
+Replicate 1 exhibited the weakest correlation with other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
+The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity during viral passage contributes most of the experimental noise.
 
-We performed a similar analysis using our new data for the Perth/2009 H3 HA.
-Surprisingly, the head domain of the H3 HA is \emph{not} more mutationally tolerant than the stalk domain (Figure~\ref{fig:mut_tolerance}).
-Specifically, whereas solvent-exposed residues in the head domain are substantially more mutationally tolerant than those in the stalk domain for the WSN/1933 H1 HA, the trend is actually reversed for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
-This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (Figure~\ref{suppfig:head_stalk_RSA}).
-For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
-Helix A is a component of the epitope of many broadly neutralizing anti-stalk antibodies~\cite{mallajosyula2014influenza,laursen2013broadly,chai2016two}, and interestingly, there are more reports of readily selecting escape mutants from such anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.
-
-We also see high mutational tolerance in many of the known antigenic regions of H3 HA~\cite{wiley1981structural}.
-For instance, antigenic region B is an immunodominant area, and many recent major antigenic drift mutations have occurred in this region~\cite{chambers2015identification,koel2013substitutions,popova2012immunodominance}.
-We find that the most distal portion of the globular head near the 190-helix, which is part of antigenic region B, is highly tolerant of mutations (Figure~\ref{fig:mut_tolerance}A).
-Antigenic region C is also notably mutationally tolerant.
-
-Many residues inside HA's receptor binding pocket are known to be highly functionally constrained~\cite{wilson1981structure,martin1998studies}, and our data indicates that these sites are relatively mutationally intolerant in both H3 and H1 HAs (Figure~\ref{fig:mut_tolerance}A).
-In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, since mutations at these sites can have large effects on antigenicity~\cite{wiley1981structural,koel2013substitutions}.
+\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
+How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
+This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
+Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
+This result indicates that our experiments authentically capture some of the constraints on HA evolution.
+The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
+However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for by the constraints measured in the deep mutational scanning.
+The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
+Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
+These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.
 
@@
 
-\subsection*{Our measurements can help distinguish between mutations that reach low and high frequencies in nature}
-Mutations occurring in the H3N2 virus population have widely varying evolutionary outcomes (Figure~\ref{fig:frequency_trajectory}).
-Some mutations appear, spread and fix in the population, while others briefly circulate before dieing out.
-We take the maximum frequency reached by a mutation as a coarse indicator of its fitness, under the assumption that fitter mutations are expected to reach higher frequency on average than less fit mutations \cite{ewens2012mathematical}.
-If a mutation is highly deleterious, we would expect it to quickly die out, whereas less deleterious or beneficial mutations are more likely to expand in frequency.
-Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, i.e.\ although R142G occurs multiple times on the phylogeny we track each of these mutations occurring on different backgrounds separately.
-As such, each mutation is shown as a separate circle occuring on a separate branch in Figure \ref{fig:frequency_trajectory}.
-However, because multiple mutations on the same phylogeny branch cannot be disentangled, we identified all mutations on a branch and assigned a single mutational effect based on the sum of effects for each mutation on the branch.
-We observed few circulating mutations in our phylogeny with strongly deleterious mutational effects and rarely observed these mutations reaching high frequency (Figure~\ref{fig:frequency_trajectory}).
-Mutations that reached high global frequencies generally had neutral or beneficial mutational effects.
+Examination of Figure~\ref{fig:logoplot} reveals that the experimentally measured amino-acid preferences generally agree with existing knowledge about HA's structure and function.
+For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} strongly prefer cysteine.
+At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such as Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change,yang2015structure}.
+A positively charged amino acid at site 329 is important for cleavage of the HA0 precursor into the mature form~\cite{stech2005new}, and this site strongly prefers arginine.
+However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine.
+This codon is part of the signal peptide and is cleaved from the mature HA protein.
+One possible reason that our experiments do not show a strong preference for methionine at this site could be alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
+% No upstream or downstream "ATG"'s are present.
+I did find an in-frame "ATC" one codon upstream of the canonical start site (in the non-coding region), and interestingly, all Perth/2009 HA sequences deposited in GenBank have an "ATT" in this non-coding position, including Seema's Perth HA.
+% I6M(HA2) mutant slightly increases the fusion pH (Daniels 1985 Cell)
+% We see high preference for T,S,Q over the wildtype Ile at site 226.
+Although Q226 in an H3 background is characterized as having preference for avian-type receptors, most of these studies were done in older H3 strains such as Aichi.
+Wu 2017 Cell Host Microbe has shown extensive epistasis in the RBS, specifically in the 220-loop, so perhaps the amino acids tolerated in this loop have shifted over time.
+Indeed the wildtype amino acids at 225 and 226 are different in Perth compared to HK68 and Aichi.
 
-\begin{figure}
+\subsection*{There is less difference in mutational tolerance between the HA head and stalk domains for H3 than for H1}
+Our experiments measure which amino acids are tolerated at each HA site under selection for viral growth.
+We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we quantify as the Shannon entropy of the re-scaled preferences.
+In prior mutational studies of H1 HAs, the stalk domain was found to be substantially less mutationally tolerant than the globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,heaton2013genome}.
+
+We performed a similar analysis using our new data for the Perth/2009 H3 HA.
+Surprisingly, the head domain of the H3 HA is \emph{not} more mutationally tolerant than the stalk domain (Figure~\ref{fig:mut_tolerance}).
+Specifically, whereas solvent-exposed residues in the head domain are substantially more mutationally tolerant than those in the stalk domain for the WSN/1933 H1 HA, the trend is actually reversed for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
+This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (Figure~\ref{suppfig:head_stalk_RSA}).
+For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
+Helix A forms part of the epitope of many broadly neutralizing anti-stalk antibodies~\cite{mallajosyula2014influenza,laursen2013broadly,chai2016two}, and interestingly, there are more reports of readily selecting escape mutants from such anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.
+
+We also see high mutational tolerance in many of the known antigenic regions of H3 HA~\cite{wiley1981structural}.
+For instance, antigenic region B is an immunodominant area, and many recent major antigenic drift mutations have occurred in this region~\cite{chambers2015identification,koel2013substitutions,popova2012immunodominance}.
+We find that the most distal portion of the globular head near the 190-helix, which is part of antigenic region B, is highly tolerant of mutations (Figure~\ref{fig:mut_tolerance}A).
+Antigenic region C is also notably mutationally tolerant.
+
+Many residues inside HA's receptor binding pocket are known to be highly functionally constrained~\cite{wilson1981structure,martin1998studies}, and our data indicates that these sites are relatively mutationally intolerant in both H3 and H1 HAs (Figure~\ref{fig:mut_tolerance}A).
+In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, since mutations at these sites can have large effects on antigenicity~\cite{wiley1981structural,koel2013substitutions}.
+
+\begin{figure}[h!]
 \centering
@@
 \end{figure}
-
-We next sought to quantify the correlation between mutation effects and maximum frequencies attained.
-To minimize effects related to the genetic background of the DMS strain, we excluded mutations closely related to the DMS strain itself and partitioned the remaining mutations into XXX \comment{John: please add number} mutations pre-dating and XXX \comment{John: please add number} mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
-We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
-We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
-We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
-In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).
-The similar effect size for both partitions suggests that our experimental measurements can be used to predict mutation frequencies despite the genetic background of the DMS strain.
 
@@
 
-The trends in Figure~\ref{fig:muteffect_maxfreq}A appear to be most strongly driven by behavior of strongly deleterious mutations.
-We investigated this further by partitioning mutations in those circulating at low, medium and high levels and those that fix in the population (Figure~\ref{fig:muteffect_maxfreq}B).
-We observe both a shift in the mean \comment{John: or is it median?} mutation effect and in the distribution of mutation effects as we increase in circulation category.
-Mutations that we measure to be substantially deleterious almost never reach high frequency.
-Overall, these results demonstrate that measurements of how mutations affect viral growth in cell culture are informative for understanding the fates of these mutations in nature: in particular, if a mutation is measurably deleterious to viral growth, that mutation is unlikely to prosper in nature.
+\subsection*{Our measurements can help distinguish between mutations that reach low and high frequencies in nature}
+Mutations occurring in the H3N2 virus population have widely varying evolutionary outcomes (Figure~\ref{fig:frequency_trajectory}).
+Some mutations appear, spread and fix in the population, while others briefly circulate before dying out.
+We take the maximum frequency reached by a mutation as a coarse indicator of its fitness, under the assumption that fitter mutations are expected to reach higher frequency on average than less fit mutations \cite{ewens2012mathematical}.
+If a mutation is highly deleterious, we would expect it to quickly die out, whereas less deleterious or beneficial mutations are more likely to expand in frequency.
+Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, i.e.\ although R142G occurs multiple times on the phylogeny we track each of these mutations occurring on different backgrounds separately.
+As such, each mutation is shown as a separate circle occurring on a separate branch in Figure \ref{fig:frequency_trajectory}.
+However, because multiple mutations on the same phylogeny branch cannot be disentangled, when multiple mutations occurred on a single branch, we assigned a single mutational effect based on the sum of effects of each mutation.
 
-\subsection*{Measurements made on an H1 HA are not useful for understanding the evolution of H3 influenza}
-To determine how broadly experimental measurements can be generalized across HAs, we repeated the analysis of H3N2 mutation frequencies, but using mutational effects from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
-Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the H1 measurements do \textit{not} reveal a trend for mutation effect on H1 to correlate with H3 maximum mutation frequency.
-This result is not surprising given that H1 and H3 HAs only have 42\% sequence identity, but it does underscore that the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
+After annotating mutations and their frequencies on the phylogeny in this way (Figure~\ref{fig:frequency_trajectory}), it is visually obvious that there are relatively few circulating mutations with strongly deleterious mutational effects---and that such deleterious mutations rarely reach high frequency when they do occur.
+Mutations that reached high global frequencies generally had neutral or beneficial mutational effects.
+
+We next sought to quantify the correlation between mutation effects and maximum frequencies attained.
+To minimize effects related to the genetic background of the DMS strain, we excluded mutations closely related to the DMS strain itself and partitioned the remaining mutations into XXX \comment{John: please add number} mutations pre-dating and XXX \comment{John: please add number} mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
+We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
+We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
+We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
+In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).
+The similar effect size for both partitions suggests that our experimental measurements can be used to predict mutation frequencies despite the genetic background of the DMS strain.
 
@@
 
+The trends in Figure~\ref{fig:muteffect_maxfreq}A appear to be most strongly driven by behavior of strongly deleterious mutations.
+We investigated this further by partitioning mutations in those circulating at low, medium and high levels and those that fix in the population (Figure~\ref{fig:muteffect_maxfreq}B).
+We observe both a shift in the mean \comment{John: or is it median?} mutation effect and in the distribution of mutation effects as we increase in circulation category.
+Mutations that we measure to be substantially deleterious almost never reach high frequency.
+Overall, these results demonstrate that measurements of how mutations affect viral growth in cell culture are informative for understanding the fates of these mutations in nature: in particular, if a mutation is measurably deleterious to viral growth, that mutation is unlikely to prosper in nature.
+
+\subsection*{Measurements made on an H1 HA are not useful for understanding the evolution of H3 influenza}
+To determine how broadly experimental measurements can be generalized across HAs, we repeated the analysis of H3N2 mutation frequencies, but using mutational effects from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
+Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the H1 measurements do \textit{not} reveal a trend for mutation effect on H1 to correlate with H3 maximum mutation frequency.
+This result is not surprising given that H1 and H3 HAs only have 42\% sequence identity, but it does underscore that the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
+
 \subsection*{There are large differences between H3 and H1 HAs in the amino-acid preferences of many sites}
@@
 To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
-Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2017mapping}.
+Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2018mapping}.
 However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
@@
 
-To more rigorously quantify shifts in amino-acid preferences after correcting for experimental noise, we used the statistical approach in~\cite{doud2015site,haddox2017mapping}.
+To more rigorously quantify shifts in amino-acid preferences after correcting for experimental noise, we used the statistical approach in~\cite{doud2015site,haddox2018mapping}.
 Figure~\ref{fig:distance_distribution}C shows the distribution of shifts in amino-acid preferences between H3 and H1 HAs after correcting for experimental noise.
@@
 These shifts between H3 and H1 are much larger than expected from the null distribution that would be observed purely from experimental noise.
-They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2017mapping}.
+They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2018mapping}.
 However, the typical shift between H3 and H1 is still smaller than that observed when comparing HA to the non-homologous HIV Env protein.
@@
 $R$ indicates the Pearson correlation coefficient.
-(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2017mapping}, and plotted the distribution of shifts for all sites.
+(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2018mapping}, and plotted the distribution of shifts for all sites.
 The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
@@
 These generalizations will only be valid to the extent that the effects of mutations are conserved during HA's evolution.
-Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2017mapping}, and the extent to which this occurs is an important question in its own right.
+Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2018mapping}, and the extent to which this occurs is an important question in its own right.
 Our work suggests that measurements on a HA from single human H3N2 viral strain can be usefully generalized to at least some extent across the entire evolutionary history of human H3N2 HA.
-On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2017mapping}.
+On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2018mapping}.
 Further investigation of how mutational effects shift as proteins diverge will be important for determining how broadly any given experiment can be generalized when attempting to make evolutionary forecasts.
@@
 Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
-Computer code used to analyze the data and produce the results in the paper are on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
+Computer code used to analyze the data and produce the results in the paper are on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript}.
 
@@
 The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~\ref{suppdata:prefsfiles}.
-Computer code that performs the entirety of this analysis is available on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
+Computer code that performs the entirety of this analysis is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript} \comment{Juhye, add additional details and link to the exact notebook}.
 
@@
 We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
-The phylogenetic tree is available as a JSON file on GitHub at \comment{repository will be made public upon acceptance of paper; if you are a reviewer who wants access please contact the editor}.
+The phylogenetic tree is available as a JSON file on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript} \comment{Juhye: add link to exact file}.
 The tree was visualized using BALTIC (\url{https://github.com/blab/baltic}).
@@
 To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using MAFFT~\cite{katoh2013mafft}.
-To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2017mapping} and used the RMSD$_{\text{corrected}}$ values as our quantification of the extent of each shift.
+To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2018mapping} and used the RMSD$_{\text{corrected}}$ values as our quantification of the extent of each shift.
 

### 2018-04-02 paper/manuscript.tex
Addressed comments about mean/median and number of mutations.

@@
 (B) The distribution of mutational effects partitioned by maximum mutation frequency.
-The vertical black line shows the mean \comment{John: or is it median?} mutation effect for each category.
+The vertical black line shows the mean mutation effect for each category.
 The analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
@@
 We next sought to quantify the correlation between mutation effects and maximum frequencies attained.
-To minimize effects related to the genetic background of the DMS strain, we excluded mutations closely related to the DMS strain itself and partitioned the remaining mutations into XXX \comment{John: please add number} mutations pre-dating and XXX \comment{John: please add number} mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
+To minimize effects related to the genetic background of the DMS strain, we excluded mutations closely related to the DMS strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
 We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
@@
 Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
-\comment{John: Can you match any formatting changes that you make to Figure~\ref{fig:muteffect_maxfreq}A here so that they look consistent.}
+The empirical $P$-value represents the results of 1,000 permutations.
 }
@@
 We investigated this further by partitioning mutations in those circulating at low, medium and high levels and those that fix in the population (Figure~\ref{fig:muteffect_maxfreq}B).
-We observe both a shift in the mean \comment{John: or is it median?} mutation effect and in the distribution of mutation effects as we increase in circulation category.
+We observe both a shift in the mean mutation effect and in the distribution of mutation effects as we increase in circulation category.
 Mutations that we measure to be substantially deleterious almost never reach high frequency.

### 2018-04-02 paper/manuscript.tex
Added PDF of frequency trajectory examples resized to fit in one column.
Updated WSN scatterplots for max frequency by mut effect.
Updated WSN histograms for max frequency by mut effect.

@@
 \centering
-\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex.png}
+\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex.pdf}
 \caption{\label{fig:frequency_trajectory}
@@
 \centering
-\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq_WSN_1.pdf}
+\includegraphics[width=1.0\textwidth]{figs/muteffect_by_maxfreq_WSN_1.pdf}
 \caption{\label{fig:muteffect_maxfreq_WSN}
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.75\textwidth]{figs/muteffect_by_maxfreq_WSN_2.pdf}}
+\centerline{\includegraphics[width=1.0\textwidth]{figs/muteffect_by_maxfreq_WSN_2.pdf}}
 \caption{\label{suppfig:muteffect_maxfreq_WSN_supp}

### 2018-04-06 paper/manuscript.tex
Updated 1968-2018 phylogeny.
Update figures, incorporate into manuscript

@@
 \centering
-\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex.pdf}
+\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex/frequency_trajectory_ex.pdf}
 \caption{\label{fig:frequency_trajectory}
@@
 
-\begin{figure*}[ht]
+\begin{SCfigure*}[\sidecaptionrelwidth][t]
+%\begin{figure*}[ht]
 \centering
-\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq.pdf}
+\includegraphics[width=12cm]{figs/muteffect_by_maxfreq/muteffect_by_maxfreq.pdf}
+%\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq/muteffect_by_maxfreq.pdf}
 \caption{\label{fig:muteffect_maxfreq}
@@
 }
-\end{figure*}
+\end{SCfigure*}
+%\end{figure*}
 
@@
 
-\begin{figure*}[ht]
+\begin{SCfigure*}[\sidecaptionrelwidth][t]
+%\begin{figure*}[ht]
 \centering
-\includegraphics[width=1.0\textwidth]{figs/muteffect_by_maxfreq_WSN_1.pdf}
+\includegraphics[width=12cm]{figs/muteffect_by_maxfreq_WSN/muteffect_by_maxfreq_WSN.pdf}
+%\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq_WSN/muteffect_by_maxfreq_WSN.pdf}
 \caption{\label{fig:muteffect_maxfreq_WSN}
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.5\textwidth]{figs/S01_G78D-T212I/G78D-T212I.pdf}}
+\centerline{\includegraphics[width=0.5\textwidth]{figs/supp_G78D-T212I/G78D-T212I.pdf}}
 \caption{\label{suppfig:Perth2009_mut}
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.6\textwidth]{figs/S02_SangerSeq/SangerSeq.pdf}}
+\centerline{\includegraphics[width=0.6\textwidth]{figs/supp_SangerSeq/SangerSeq.pdf}}
 \caption{\label{suppfig:SangerSeq}
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=\textwidth]{figs/S03_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
+\centerline{\includegraphics[width=\textwidth]{figs/supp_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
 \caption{\label{suppfig:WSNprefs_logoplot}
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[clip=true, trim=0 5.5in 0 0,width=0.7\textwidth]{figs/S04_post_Perth2009/post_Perth2009_seqpref.pdf}}
+%%\centerline{\includegraphics[clip=true, trim=0 5.5in 0 0,width=0.7\textwidth]{figs/S04_post_Perth2009/post_Perth2009_seqpref.pdf}}
+\centerline{\includegraphics[width=1.0\textwidth]{figs/supp_H3N2_phylogeny/H3N2_phylogeny.pdf}}
 \caption{\label{suppfig:tree}
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=1.0\textwidth]{figs/muteffect_by_maxfreq_WSN_2.pdf}}
+\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_muteffect_maxfreq_WSN/muteffect_maxfreq_WSN_supp.pdf}}
 \caption{\label{suppfig:muteffect_maxfreq_WSN_supp}
@@
 
+\begin{suppdata}[H]
+\caption{\label{suppdata:gisaid}
+Excel file providing acknowledgments and accessions for sequences downloaded from GISAID.
+}
+\end{suppdata}
+
 \end{document}

### 2018-04-06 paper/manuscript.tex
Add appropriate repository links in Methods

@@
 The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~\ref{suppdata:prefsfiles}.
-Computer code that performs the entirety of this analysis is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript} \comment{Juhye, add additional details and link to the exact notebook}.
+Computer code that performs the entirety of this analysis is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/analysis_code/analysis_notebook.ipynb}.
 
@@
 We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
-The phylogenetic tree is available as a JSON file on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript} \comment{Juhye: add link to exact file}.
+The phylogenetic tree is available as a JSON file on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/analysis_code/flu_h3n2_ha_1968_2018_6v_tree.json.gz}.
 The tree was visualized using BALTIC (\url{https://github.com/blab/baltic}).

### 2018-04-07 paper/manuscript.tex
revision to text

@@
 This evolution is further characterized by competition and turnover among groups of strains called clades bearing different complements of mutations~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,koelle2015effects,bedford2015global}.
-Clades vary widely in their evolutionary success with some dying out soon after emergence and others going on to take over the virus population.
-Several lines of evidence indicate that successful clades, which reach high frequencies or even fix in the global population, have higher fitness than clades that remain at low frequency~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,luksza2014predictive}.
+Clades vary widely in their evolutionary success, with some dying out soon after emergence and others going on to take over the virus population.
+Several lines of evidence indicate that successful clades have higher fitness than clades that remain at low frequency~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,luksza2014predictive}.
 A key goal in the study of H3N2 evolution is to identify the features that enable certain clades to succeed as others die out.
@@
 Specifically, stop codons were purged to 20-45\% of their initial frequencies after correcting for error rates estimated by sequencing the wildtype controls (Figure~\ref{fig:dms_overview}C).
-The incomplete purging of stop codons is likely because genetic complementation due to co-infection~\cite{marshall2013influenza} enabled the persistence of some virions with nonfunctional HAs.
+The incomplete purging of stop codons is likely because genetic complementation due to co-infection~\cite{marshall2013influenza, brooke2013most} enabled the persistence of some virions with nonfunctional HAs.
 We also observed selection against many nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), with their frequencies falling to 30-40\% of their initial values after error correction.
@@
 \caption{\label{fig:frequency_trajectory}
-{\bf Frequency trajectories of individual mutations and their relation to the experimentally determined effects of these mutations.}
-The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} windowed from 2004 to 2014.
+{\bf Frequency trajectories of individual mutations and their relation to the experimentally measured effects of these mutations.}
+The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} from 2004 to 2014.
 Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
-The bottom panel shows the frequency trajectory of each mutation, with mutation trajectories colored according to the mutational effects as in the top panel.
+The bottom panel shows the frequency trajectory of each mutation, with trajectories colored according to the mutational effects as in the top panel.
 It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
@@
 {\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
-(A) Correlation between the effects of mutations as measured in our experiments and the maximum frequency reached by these mutations.
-The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations for which permuted $\rho$ was greater than or equal to the observed $\rho$.
+(A) Correlation between the effects of mutations as measured in our deep mutational scanning of the Perth/2009 HA and the maximum frequency reached by these mutations in nature.
+The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations of the experimental measurements for which the permuted $\rho$ was greater than or equal to the observed $\rho$.
 (B) The distribution of mutational effects partitioned by maximum mutation frequency.
@@
 \subsection*{Our measurements can help distinguish between mutations that reach low and high frequencies in nature}
-Mutations occurring in the H3N2 virus population have widely varying evolutionary outcomes (Figure~\ref{fig:frequency_trajectory}).
-Some mutations appear, spread and fix in the population, while others briefly circulate before dying out.
-We take the maximum frequency reached by a mutation as a coarse indicator of its fitness, under the assumption that fitter mutations are expected to reach higher frequency on average than less fit mutations \cite{ewens2012mathematical}.
-If a mutation is highly deleterious, we would expect it to quickly die out, whereas less deleterious or beneficial mutations are more likely to expand in frequency.
-Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, i.e.\ although R142G occurs multiple times on the phylogeny we track each of these mutations occurring on different backgrounds separately.
-As such, each mutation is shown as a separate circle occurring on a separate branch in Figure \ref{fig:frequency_trajectory}.
+Mutations occurring in the H3N2 virus population experience widely varying evolutionary fates (Figure~\ref{fig:frequency_trajectory}).
+Some mutations appear, spread and fix in the population, while others briefly circulate before disappearing.
+We take the maximum frequency reached by a mutation as a coarse indicator of its effect on fitness, since favorable mutations generally reach higher frequency than unfavorable ones~\cite{ewens2012mathematical}.
+%Highly deleterious mutations will usually die out, whereas less deleterious or beneficial ones are more likely to expand in frequency.
+Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, e.g.\ although R142G occurs multiple times on the phylogeny we track each of these mutations occurring on different backgrounds separately.
+As such, each mutation is shown as a separate circle on a separate branch in Figure \ref{fig:frequency_trajectory}.
 However, because multiple mutations on the same phylogeny branch cannot be disentangled, when multiple mutations occurred on a single branch, we assigned a single mutational effect based on the sum of effects of each mutation.
 
-After annotating mutations and their frequencies on the phylogeny in this way (Figure~\ref{fig:frequency_trajectory}), it is visually obvious that there are relatively few circulating mutations with strongly deleterious mutational effects---and that such deleterious mutations rarely reach high frequency when they do occur.
-Mutations that reached high global frequencies generally had neutral or beneficial mutational effects.
+After annotating mutations and their frequencies on the phylogeny in this way (Figure~\ref{fig:frequency_trajectory}), it is visually obvious that there are relatively few circulating mutations that we measure to be strongly deleterious---and that such deleterious mutations rarely reach high frequency when they do occur.
+Mutations that reach high frequencies generally have neutral or beneficial effects according to our experimental measurements.
 
-We next sought to quantify the correlation between mutation effects and maximum frequencies attained.
-To minimize effects related to the genetic background of the DMS strain, we excluded mutations closely related to the DMS strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
+We next sought to quantify the correlation between a mutation's experimentally measured effect and the maximum frequency it attained during natural evoluton.
+To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
 We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
@@
 In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).
-The similar effect size for both partitions suggests that our experimental measurements can be used to predict mutation frequencies despite the genetic background of the DMS strain.
+The similar effect sizes for both partitions shows that our experimental measurements can help explain the evolutionary fates of mutations in strains that post-date the experimentally studied strain, as well as to retrospectively analyze mutations that precede the experimental strain.
 
@@
 
-The trends in Figure~\ref{fig:muteffect_maxfreq}A appear to be most strongly driven by behavior of strongly deleterious mutations.
-We investigated this further by partitioning mutations in those circulating at low, medium and high levels and those that fix in the population (Figure~\ref{fig:muteffect_maxfreq}B).
-We observe both a shift in the mean mutation effect and in the distribution of mutation effects as we increase in circulation category.
-Mutations that we measure to be substantially deleterious almost never reach high frequency.
+The trends in Figure~\ref{fig:muteffect_maxfreq}A are most strongly driven by the behavior of substantially deleterious mutations.
+We investigated this further by partitioning mutations into those that reach low, medium and high frequencies, and those that fix in the population (Figure~\ref{fig:muteffect_maxfreq}B).
+The mutations that reach higher frequency have a more favorable mean effect.
+Mutations measured to be substantially deleterious almost never reach high frequency.
 Overall, these results demonstrate that measurements of how mutations affect viral growth in cell culture are informative for understanding the fates of these mutations in nature: in particular, if a mutation is measurably deleterious to viral growth, that mutation is unlikely to prosper in nature.
@@
 \subsection*{Measurements made on an H1 HA are not useful for understanding the evolution of H3 influenza}
-To determine how broadly experimental measurements can be generalized across HAs, we repeated the analysis of H3N2 mutation frequencies, but using mutational effects from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
-Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the H1 measurements do \textit{not} reveal a trend for mutation effect on H1 to correlate with H3 maximum mutation frequency.
-This result is not surprising given that H1 and H3 HAs only have 42\% sequence identity, but it does underscore that the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
+To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
+Figure~\ref{fig:muteffect_maxfreq_WSN} shows that there is no correlation between the H1 experimental measurements and maximum frequency that mutations reach during H3N2 viral evolution.
+Therefore, the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
 
@@
 \caption{\label{fig:distance_distribution}
-{\bf There are substantial differences in the effects of mutations between H1 and H3 HAs.}
+{\bf There are large shifts in the effects of mutations between H1 and H3 HAs.}
 (A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled.
-These HAs have 42\% amino-acid identity.
-(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study  and the three replicates in prior deep mutational scanning of an H1 HA~\cite{doud2016accurate}.
+(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study and the three replicates in prior deep mutational scanning of the H1 HA~\cite{doud2016accurate}.
 Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray.
@@
 The shifts are also much larger than those previously observed between two variants of HIV Env that share 86\% amino-acid identity (pink).
-However, the shifts between H3 and H1 are usually less than the differences between HA and HIV Env (green).
+However, the shifts between H3 and H1 are less than the differences between HA and HIV Env (green).
 }
@@
 We have measured the effects of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements have some value for understanding the evolutionary fate of these mutations in nature.
-Specifically, mutations measured to be more beneficial for viral growth tend to circulate at higher frequency in nature than mutations measured to be more deleterious for viral growth.
-The fact that our experiments can be used to help identify evolutionary successful mutations suggests that they might inform evolutionary forecasting.
+Specifically, mutations measured to be more beneficial for viral growth tend to reach higher frequencies in nature than mutations measured to be more deleterious for viral growth.
+The fact that our experiments can help identify evolutionary successful mutations suggests that they might inform evolutionary forecasting.
 In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, {\L}uksza and L{\"a}ssig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
@@
 It is important to emphasize that measurements of viral growth in cell culture do \emph{not} represent true fitness in nature.
-Indeed, a vast amount of work in virology has chronicled the many ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}.
+Indeed, a vast amount of work in virology has chronicled the ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}.
 Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
@@
 Yet, prior to our work, there were no comprehensive studies of the functional effects of mutations to H3 HA on any property that even resembled viral fitness in nature, and modeling work has either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or assumed that all non-epitope mutations had equivalent deleterious effects~\cite{luksza2014predictive}.
-The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative --- and so represent an improvement over no information at all.
+The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative---and so represent an improvement over no information at all.
 We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility.
@@
 These generalizations will only be valid to the extent that the effects of mutations are conserved during HA's evolution.
-Extensive work on protein evolution has shown that epistasis can shift the effects of mutations~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative,haddox2018mapping}, and the extent to which this occurs is an important question in its own right.
+Extensive prior work has shown that epistasis can shift the effects of mutations as proteins evolve~\cite{pollock2012amino,shah2015contingency,gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}.
 Our work suggests that measurements on a HA from single human H3N2 viral strain can be usefully generalized to at least some extent across the entire evolutionary history of human H3N2 HA.
-On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites --- much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2018mapping}.
+On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites---much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2018mapping}.
 Further investigation of how mutational effects shift as proteins diverge will be important for determining how broadly any given experiment can be generalized when attempting to make evolutionary forecasts.
@@
 \subsection*{Creation of MDCK-SIAT1-TMPRSS2 cells}
-When growing influenza virus in cell culture, trypsin is normally added in order to cleave HA into its mature form.
-To obviate the need for trypsin, we engineered an MDCK-SIAT1 cell line to constitutively express the TMPRSS2 protease, which cleaves and activates HA in the human airways~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
+When growing influenza virus in cell culture, trypsin is normally added to cleave HA into its mature form.
+To obviate the need for trypsin, we engineered MDCK-SIAT1 cells to constitutively express the TMPRSS2 protease, which cleaves and activates HA in the human airways~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
 The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656) and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry to create plasmid pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
@@
 \subsection*{Generation of HA codon mutant plasmid libraries}
-The HA and NA genes for the Perth/2009 viral strain were reverse-transcribed and amplified from recombinant virus obtained from BEI Resources (NR-41803), and cloned into the pHW2000~\citep{hoffmann2000dna} influenza reverse-genetics plasmids to create pHW-Perth09-HA and pHW-Perth09-NA.
+HA and NA genes for the Perth/2009 viral strain were cloned from recombinant virus obtained from BEI Resources (NR-41803) into the pHW2000~\citep{hoffmann2000dna} influenza reverse-genetics plasmids to create pHW-Perth09-HA and pHW-Perth09-NA.
 
@@
 To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amounts of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
-The next day we changed the media to influenza growth media (IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL --- no trypsin was added since there was TMPRSS2), and then we collected the viral supernatant at 72 hours post-transfection.
-This viral supernatant was then blind passaged in MDCK-SIAT1-TMPRSS2 a total of six additional times.
+The next day we changed the media to influenza growth media (IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL (no trypsin was added since there was TMPRSS2), and then collected the viral supernatant at 72 hours post-transfection.
+This viral supernatant was blind passaged in MDCK-SIAT1-TMPRSS2 a total of six additional times.
 We isolated viral RNA from these passaged viruses and sequenced the HA gene.
-The passaged HA had two mutations, G78D and T212I, which we found enhanced viral growth as shown in Figure~\ref{suppfig:Perth2009_mut}.
+The passaged HA had two mutations, G78D and T212I, which enhanced viral growth as shown in Figure~\ref{suppfig:Perth2009_mut}.
 The HA with these two mutations was cloned into pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} to create pHW-Perth09-HA-G78D-T212I and pICR2-Perth09-HA-G78D-T212I.
-For all subsequent experiments, we used viruses with the HA containing these two mutations in order to improve titers and viral genetic stability, and this is the HA that we refer to as Perth/2009 in the manuscript text.
-We also used all non-HA genes (including NA) from WSN/1933 to help increase titers and reduce biosafety concerns.
+For all subsequent experiments, we used viruses with the HA containing these two mutations to improve titers and viral genetic stability, and this is the HA that we refer to as Perth/2009 in the manuscript text.
+We used all non-HA genes (including NA) from WSN/1933 to help increase titers and reduce biosafety concerns.
 
@@
 We passaged $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
-To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they had reached a density of $\sim 1.7 \times 10^7$ cells per dish.
-We then replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L.
+To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they were at $\sim 1.7 \times 10^7$ cells per dish.
+We replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L in IGM.
 We collected viral supernatant for sequencing 48 hours post-infection.
@@
 \subsection*{Barcoded subamplicon sequencing}
-To extract viral RNA from the three replicate HA virus libraries and the wildtype HA virus, we ultracentrifuged 24 mL of the supernatant at 22,000 rpm for 1.5 h at 4$^{\circ}$C in a Beckman Coulter SW28 rotor, and extracted RNA using the Qiagen RNeasy Mini Kit by resuspending the viral pellet in 400 $\mu$L of buffer RLT supplemented with $\beta$-mercaptoethanol, pipetting 30 times, transferring the liquid to a microcentrifuge tube, adding 600 $\mu$L 70\% ethanol, and proceeding with the extraction according to the manufacturer's instructions.
-The HA gene was then reverse-transcribed with AccuScript Reverse Transcriptase (Agilent 200820) using the primers P09-HA-For (5'-AGCAAAAGCAGGGGATAATTCTATTAATC-3') and P09-HA-Rev (5'-AGTAGAAACAAGGGTGTTTTTAATTACTAATACAC-3').
+To extract viral RNA from the three replicate HA virus libraries and the wildtype HA virus, we ultracentrifuged 24 mL of supernatant at 22,000 rpm for 1.5 h at 4$^{\circ}$C in a Beckman Coulter SW28 rotor, and extracted RNA using the Qiagen RNeasy Mini Kit by resuspending the viral pellet in 400 $\mu$L of buffer RLT supplemented with $\beta$-mercaptoethanol, pipetting 30 times, transferring the liquid to a microcentrifuge tube, adding 600 $\mu$L 70\% ethanol, and proceeding according to the manufacturer's instructions.
+The HA gene was reverse-transcribed with AccuScript Reverse Transcriptase (Agilent 200820) using primers P09-HA-For (5'-AGCAAAAGCAGGGGATAATTCTATTAATC-3') and P09-HA-Rev (5'-AGTAGAAACAAGGGTGTTTTTAATTACTAATACAC-3').
 
@@
 The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~\ref{suppdata:prefsfiles}.
-Computer code that performs the entirety of this analysis is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/analysis_code/analysis_notebook.ipynb}.
+Computer code that performs the entirety of this analysis and shows many detailed plots about read depth and other quality control metrics is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/analysis_notebook.ipynb}.
 
@@
 \subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
-To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid}, sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
+To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid} (see Dataset~\ref{suppdata:gisaid}), sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
 We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
@@
 We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
-The phylogenetic tree is available as a JSON file on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/analysis_code/flu_h3n2_ha_1968_2018_6v_tree.json.gz}.
+The phylogenetic tree is available as a JSON file on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/data/flu_h3n2_ha_1968_2018_6v_tree.json.gz}.
 The tree was visualized using BALTIC (\url{https://github.com/blab/baltic}).
@@
 Herein, mutation frequency dynamics are modeled according to a Brownian motion diffusion process discretized to one-month intervals.
-Relative to a simple Brownian motion, the expectation includes an `inertia' term $\inertia$ that adds velocity to the diffusion and the variance includes a term $x(1-x)$ to scale variance according to frequency following a Wright-Fisher population genetic process.
+Relative to a simple Brownian motion, the expectation includes an ``inertia'' term $\inertia$ that adds velocity to the diffusion and the variance includes a term $x(1-x)$ to scale variance according to frequency following a Wright-Fisher population genetic process.
 This results in the following diffusion process
@@
 
-\acknow{We thank Sarah Hilton, Hugh Haddox, and Sidney Bell for helpful discussions about data analysis and we thank Richard Neher for sharing analysis code.
+\acknow{We thank Sarah Hilton, Hugh Haddox, and Sidney Bell for helpful discussions about data analysis and we thank Richard Neher for sharing analysis code and providing helpful comments on an initial version of the manuscript.
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
@@
 \begin{suppfigure}[H]
-%%\centerline{\includegraphics[clip=true, trim=0 5.5in 0 0,width=0.7\textwidth]{figs/S04_post_Perth2009/post_Perth2009_seqpref.pdf}}
 \centerline{\includegraphics[width=1.0\textwidth]{figs/supp_H3N2_phylogeny/H3N2_phylogeny.pdf}}
@@
 Nodes in the clade containing the Perth/2009 strain and nodes sampled in 2014 or after were excluded from our analyses.
+The Perth/2009 strain was excluded to avoid artifacts related to mutations that occurred on the branches leading to the HA sequence used in the experiment.
+The post-2014 nodes were excluded because the evolutionary fates of many sequences after this date are not yet full resolved.
 }

### 2018-04-07 paper/manuscript.tex
tweaks to fix figure placement on pages

@@
 In principle, experiments could be informative for identifying how mutations affect these features.
-Most work on influenza evolution to date has utilized experimental data primarily to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction,koel2013substitutions,chambers2015identification,li2016selection}.
-While the antigenic features of a virus certainly contribute to its success, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
+Most work on influenza evolution to date has utilized experimental data to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction,koel2013substitutions,chambers2015identification,li2016selection}.
+However, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
 Specifically, due to influenza virus's high mutation rate~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and lack of intra-segment recombination~\cite{boni2008homologous}, deleterious mutations become linked to beneficial ones.
-The resulting accumulation of deleterious mutations can affect non-antigenic properties that are central to viral fitness~\cite{luksza2014predictive}.
-However, there are no large-scale quantitative characterizations of how mutations to H3N2 HA affect non-antigenic properties crucial to viral fitness.
+The resulting accumulation of deleterious mutations can affect non-antigenic properties central to viral fitness~\cite{luksza2014predictive}.
+However, there are no large-scale quantitative characterizations of how mutations to H3N2 HA affect viral growth.
 
@@
 However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
-Here, we measure the effects on viral growth in cell culture of all mutations to the HA from a recent human H3N2 strain.
-We then show that these experimental measurements can help discriminate evolutionarily successful mutations from those that are found in strains that quickly die out.
+Here, we measure the effects on viral growth in cell culture of all mutations to the HA of a recent human H3N2 strain.
+We show that these experimental measurements can help discriminate evolutionarily successful mutations from those found in strains that quickly die out.
 However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are not useful for understanding the evolutionary fate of H3 viral strains.
@@
 All experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B).
-We also passaged and deep sequenced library 3 in duplicate (denoted as library 3-1 and 3-2) to gauge the experimental noise occurring \textit{within} a single biological replicate.
+We also passaged and deep sequenced library 3 in duplicate (library 3-1 and 3-2) to gauge experimental noise \textit{within} a single biological replicate.
 As a control to measure sequencing and mutational errors, we used the unmutated HA gene to generate and passage viruses carrying wildtype HA.
@@
 \caption{\label{tab:phydms}
-{\bf Substitution models informed by the experiments describe H3N2's natural evolution better than traditional substitution models.}}
+{\bf Substitution models informed by the experiments describe HA's evolution better than traditional models.}}
 \begin{center}
@@
  \end{center}
-\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 influenza HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites (site avg.), and the M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
+\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites (site avg.), and M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
 Models are compared by AIC~\cite{posada2004model} computed from the log likelihood (LnL) and number of model parameters.
@@
 
-We next quantified the reproducibility of our deep mutational scanning measurements across biological and technical replicates.
-We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids using the method described in~\cite{bloom2015software}.
-Because there are 566 residues in HA, there are $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site are constrained to sum to one~\cite{bloom2015software}).
+We next quantified the reproducibility of our deep mutational scanning across biological and technical replicates.
+We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids as described in~\cite{bloom2015software}.
+Because there are 566 residues in HA, there are $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site sum to one~\cite{bloom2015software}).
 The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
@@
 How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
-This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
+This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard substitution models~\cite{bloom2017identification,hilton2017phydms}.
 Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
@@
 The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
-However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for by the constraints measured in the deep mutational scanning.
+However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for in the deep mutational scanning.
 The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
-Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
+Throughout the rest of this paper, we use the amino-acid preferences re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
 These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.
@@
 \centering
-\includegraphics[width=\textwidth]{figs/mut_tolerance/mut_tolerance_new.pdf}
+\includegraphics[width=0.95\textwidth]{figs/mut_tolerance/mut_tolerance_new.pdf}
 \caption{\label{fig:mut_tolerance}
@@
 }
-%\end{SCfigure*}
 \end{figure*}
@@
 One possible reason that our experiments do not show a strong preference for methionine at this site could be alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
-% No upstream or downstream "ATG"'s are present.
-I did find an in-frame "ATC" one codon upstream of the canonical start site (in the non-coding region), and interestingly, all Perth/2009 HA sequences deposited in GenBank have an "ATT" in this non-coding position, including Seema's Perth HA.
-% I6M(HA2) mutant slightly increases the fusion pH (Daniels 1985 Cell)
-% We see high preference for T,S,Q over the wildtype Ile at site 226.
-Although Q226 in an H3 background is characterized as having preference for avian-type receptors, most of these studies were done in older H3 strains such as Aichi.
-Wu 2017 Cell Host Microbe has shown extensive epistasis in the RBS, specifically in the 220-loop, so perhaps the amino acids tolerated in this loop have shifted over time.
-Indeed the wildtype amino acids at 225 and 226 are different in Perth compared to HK68 and Aichi.
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[b!]
 \centering
-\includegraphics[width=\linewidth]{figs/frequency_trajectory_ex/frequency_trajectory_ex.pdf}
+\includegraphics[width=0.95\linewidth]{figs/frequency_trajectory_ex/frequency_trajectory_ex.pdf}
 \caption{\label{fig:frequency_trajectory}
@@
 
-\begin{SCfigure*}[\sidecaptionrelwidth][t]
-%\begin{figure*}[ht]
+\begin{SCfigure*}[\sidecaptionrelwidth][h!]
 \centering
 \includegraphics[width=12cm]{figs/muteffect_by_maxfreq/muteffect_by_maxfreq.pdf}
-%\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq/muteffect_by_maxfreq.pdf}
 \caption{\label{fig:muteffect_maxfreq}
@@
 \end{SCfigure*}
-%\end{figure*}
 
@@
 We take the maximum frequency reached by a mutation as a coarse indicator of its effect on fitness, since favorable mutations generally reach higher frequency than unfavorable ones~\cite{ewens2012mathematical}.
-%Highly deleterious mutations will usually die out, whereas less deleterious or beneficial ones are more likely to expand in frequency.
 Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, e.g.\ although R142G occurs multiple times on the phylogeny we track each of these mutations occurring on different backgrounds separately.
@@
 \begin{SCfigure*}[\sidecaptionrelwidth][t]
-%\begin{figure*}[ht]
 \centering
 \includegraphics[width=12cm]{figs/muteffect_by_maxfreq_WSN/muteffect_by_maxfreq_WSN.pdf}
-%\includegraphics[width=0.85\textwidth]{figs/muteffect_by_maxfreq_WSN/muteffect_by_maxfreq_WSN.pdf}
 \caption{\label{fig:muteffect_maxfreq_WSN}
@@
 Therefore, the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
+
+\begin{SCfigure*}[\sidecaptionrelwidth][b]
+\centering
+\includegraphics[width=12cm]{figs/distance_distribution/distance_distribution.pdf}
+\caption{\label{fig:distance_distribution}
+{\bf There are large shifts in the effects of mutations between H1 and H3 HAs.}
+(A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled.
+(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study and the three replicates in prior deep mutational scanning of the H1 HA~\cite{doud2016accurate}.
+Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray.
+$R$ indicates the Pearson correlation coefficient.
+(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2018mapping}, and plotted the distribution of shifts for all sites.
+The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
+The shifts are also much larger than those previously observed between two variants of HIV Env that share 86\% amino-acid identity (pink).
+However, the shifts between H3 and H1 are less than the differences between HA and HIV Env (green).
+}
+\end{SCfigure*}
 
@@
 Therefore, there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations remain more similar than for non-homologous proteins.
-
-\begin{SCfigure*}[\sidecaptionrelwidth][t]
-\centering
-\includegraphics[width=12cm]{figs/distance_distribution/distance_distribution.pdf}
-\caption{\label{fig:distance_distribution}
-{\bf There are large shifts in the effects of mutations between H1 and H3 HAs.}
-(A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled.
-(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study and the three replicates in prior deep mutational scanning of the H1 HA~\cite{doud2016accurate}.
-Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray.
-$R$ indicates the Pearson correlation coefficient.
-(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2018mapping}, and plotted the distribution of shifts for all sites.
-The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
-The shifts are also much larger than those previously observed between two variants of HIV Env that share 86\% amino-acid identity (pink).
-However, the shifts between H3 and H1 are less than the differences between HA and HIV Env (green).
-}
-\end{SCfigure*}
 

### 2018-04-07 paper/cover_letter.tex
added Cover Letter

\documentclass[a4paper,11pt]{letter}

\usepackage{graphicx}
\usepackage{hyperref}

\topmargin -2cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 23cm
\footskip 1.0cm

% sets Palatino as text font
\usepackage{mathpazo}

% Some of the article customisations are relevant for this class

\signature{
\vspace{-0.5in}\includegraphics[width=1.5in]{cover_letter_images/BloomSignature.png} \\
Jesse D. Bloom \\
} % Goes after the closing (ie at the end of the letter, with space for a signature)

\address{
\includegraphics[width=2.5in]{cover_letter_images/HutchLogo.png} \\ 
Jesse D. Bloom, Ph.D. \\
Fred Hutchinson Cancer Research Center \\
Seattle, WA  98109 \\
\href{mailto:jbloom@fredhutch.org}{jbloom@fredhutch.org} \\
\url{https://research.fhcrc.org/bloom/en.html}
}
% Alternatively, these may be set on an individual basis within each letter environment.

\begin{document}
\begin{letter}{}

\opening{Dear \textit{PNAS} Editorial Board,} % eg Hello.

I am writing to submit the enclosed manuscript, ``Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants,'' for consideration as a \textit{PNAS Plus} research report.

In this manuscript, we show that high-throughput experimental measurements of how mutations affect influenza virus growth in the lab can help explain the fates of viral lineages in nature.
Specifically, we use the new technique of deep mutational scanning to measure how \emph{all} amino-acid mutations to the hemagglutinin protein of a recent human H3N2 strain of influenza virus affect viral growth in the lab.
We then show that our measurements can help distinguish which viral strains succeed and which ones die out in nature.
Our work therefore takes an important step towards the goal of leveraging experiments to forecast virus evolution in nature.
For this reason, it will be of interest to evolutionary biologists, virologists, and computational biologists---making it suitable for a broad journal such as \textit{PNAS}.

Our work will also be of specific interest to virologists and structural biologists for several additional reasons.
Hemagglutinin is an important protein because it is a major target of anti-influenza immunity and a prototype viral membrane fusion protein.
Our detailed map of how all mutations to hemagglutinin affect viral growth is valuable for understanding its biochemistry.
Additionally, there has been recent work trying to elicit immunity that targets the stalk domain of hemagglutinin.
This work is motivated by the observation that the stalk domain of H1 influenza hemagglutinin is less mutationally tolerant than the immunodominant head domain.
Surprisingly, our work shows that the reduced mutational tolerance of the stalk does \emph{not} extend to H3 hemagglutinin, perhaps partially explaining why there is more literature on anti-stalk antibodies for H1 versus H3 hemagglutinins.

Finally, our work will be of specific interest to evolutionary biologists because it sheds light on the question of how the effects of mutations shift during evolution.
Prior work has experimentally compared the effects of mutations among closely related proteins, but ours is the first to compare them among highly diverged proteins.
We find that many mutations have different effects in H3 and H3 hemagglutinins, showing the stark effect of epistasis when comparing distant homologs.

For all of these reasons, we are confident that our work will be of substantial interest to the readership of \textit{PNAS}.
In addition, we would like to highlight how we have adhered to the ideals of open and transparent science by making all of our data and analysis, include the current version of the manuscript itself, publicly available at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript}. 
This openness will facilitate the efforts of others to understand, validate, and extend our work.

Because of the highly quantitative nature of our work, we suggest members of the Editorial Board and the NAS who have expertise in evolutionary biology, computation, and viruses. 
Specifically, we suggest the following \textit{PNAS} Editorial Board members:
\begin{itemize}
\item Richard Lenski: a leader in the study of evolution in the lab.
\item Simon Levin: an expert in quantitative biology and its application to viruses.
\item Peter Palese: a leading expert in influenza virus.
\end{itemize}

We suggest the following additional NAS members:
\begin{itemize}
\item Boris Shraiman: an expert in quantitative studies of influenza virus evolution.
\item James Bull: an expert in virus evolution.
\item Daniel Fisher: an expert in quantitative approaches in evolution.
\item Stephen Harrison: an expert on viral proteins, including hemagglutinin.
\item Robert Lamb: an expert on viruses including influenza virus.
\item Pamela Bjorkman: an expert on viral proteins, including hemagglutinin.
\end{itemize}

We suggest the following potential reviewers.
We have included reviewers with expertise that spans the range of fields covered by our study:
\begin{itemize}
\item Michael Lassig (Cologne, \href{mailto:lassig@thp.uni-koeln.de}{lassig@thp.uni-koeln.de}) is a leading expert in predictive models of influenza virus evolution.
\item Joshua Plotkin (Penn, \href{mailto:jplotkin@sas.upenn.edu}{jplotkin@sas.upenn.edu}) is an expert on quantitative studies of virus evolution.
\item Nicholas Heaton (Duke, \href{mailto:nicholas.heaton@dm.duke.edu}{nicholas.heaton@dm.duke.edu}) is an expert on influenza virus.
\item Katia Koelle (Emory, \href{mailto:katia.koelle@emory.edu}{katia.koelle@emory.edu}) is an expert on influenza virus evolution.
\item Ren Sun (UCLA, \href{mailto:rsun@mednet.ucla.edu}{rsun@mednet.ucla.edu}) is an expert on the deep mutational scanning of virus proteins.
\item Colin Russell (AMC Amsterdam, \href{mailto:c.a.russell@amc.uva.nl}{c.a.russell@amc.uva.nl}) is an expert on influenza virus evolution.
\end{itemize}

\closing{Sincerely} % eg Regards,

\end{letter}
\end{document}

### 2018-04-08 paper/cover_letter.tex
fixed cover letter typo noticed by Juhye

@@
 Prior work has experimentally compared the effects of mutations among closely related proteins, but ours is the first to compare them among highly diverged proteins.
-We find that many mutations have different effects in H3 and H3 hemagglutinins, showing the stark effect of epistasis when comparing distant homologs.
+We find that many mutations have different effects in H1 and H3 hemagglutinins, showing the stark effect of epistasis when comparing distant homologs.
 

### 2018-04-09 paper/manuscript.tex
Updated date interval and number of strains sampled for phylogeny.

@@
 \subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
-To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid} (see Dataset~\ref{suppdata:gisaid}), sampling six viruses per month over the time interval of January 1, 1968 to April 1, 2017.
-We aligned the resulting 2,139 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
+To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid} (see Dataset~\ref{suppdata:gisaid}), sampling six viruses per month over the time interval of January 1, 1968 to February 1, 2018.
+We aligned the resulting 2,189 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
 Ancestral state reconstruction and branch length timing were performed with TreeTime~\cite{Sagulenko2018}.
-We pruned seven strains with premature stop codons in their amino acid translations that represented likely sequencing errors, producing a final tree with 2,132 strains.
 The phylogenetic tree is available as a JSON file on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/data/flu_h3n2_ha_1968_2018_6v_tree.json.gz}.

### 2018-04-09 paper/cover_letter.tex
updates to cover letter

@@
 \documentclass[a4paper,11pt]{letter}
+
+\pagestyle{empty}
 
@@
 \item Michael Lassig (Cologne, \href{mailto:lassig@thp.uni-koeln.de}{lassig@thp.uni-koeln.de}) is a leading expert in predictive models of influenza virus evolution.
+\item Marta Luksza (Institute for Advanced Study, \href{mailto:mluksza@ias.edu}{mluksza@ias.edu}) is a leading expert in predictive models of influenza virus evolution.
 \item Joshua Plotkin (Penn, \href{mailto:jplotkin@sas.upenn.edu}{jplotkin@sas.upenn.edu}) is an expert on quantitative studies of virus evolution.
 \item Nicholas Heaton (Duke, \href{mailto:nicholas.heaton@dm.duke.edu}{nicholas.heaton@dm.duke.edu}) is an expert on influenza virus.
+\item Sergey Kryazhimskiy (UCSD, \href{mailto:skryazhi@ucsd.edu}{mailto:skryazhi@ucsd.edu}) is an expert on quantitative studies of viral evolution.
 \item Katia Koelle (Emory, \href{mailto:katia.koelle@emory.edu}{katia.koelle@emory.edu}) is an expert on influenza virus evolution.
-\item Ren Sun (UCLA, \href{mailto:rsun@mednet.ucla.edu}{rsun@mednet.ucla.edu}) is an expert on the deep mutational scanning of virus proteins.
-\item Colin Russell (AMC Amsterdam, \href{mailto:c.a.russell@amc.uva.nl}{c.a.russell@amc.uva.nl}) is an expert on influenza virus evolution.
+\item Ren Sun (UCLA, \href{mailto:rsun@mednet.ucla.edu}{rsun@mednet.ucla.edu}) is an expert on viral deep mutational scanning.
+\item Colin Russell (AMC Amsterdam, \href{mailto:c.a.russell@amc.uva.nl}{c.a.russell@amc.uva.nl}) is an expert on influenza virus.
 \end{itemize}

### 2018-04-09 paper/manuscript_bioRxiv.tex
added _bioRxiv_ version (no _PNAS_ labels)

--- 2018-04-09 paper/manuscript.tex
@@
-\documentclass[9pt,twocolumn,twoside]{pnas-new}
+\documentclass[9pt,twocolumn,twoside]{pnas-new-for-biorxiv}
 % Use the lineno option to display guide line numbers if required.
@@
 % Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[a,d,e]{Juhye M. Lee}
-\author[b,f]{John Huddleston}
+\author[a,d,e,1]{Juhye M. Lee}
+\author[b,f,1]{John Huddleston}
 \author[a,d,e]{Michael B. Doud}
@@
 wrote the paper with input from J.H.
-and T.B.; all authors were involved in project design and revising the paper}
-\authordeclaration{The authors declare no conflict of interest.}
+and T.B.; all authors were involved in project design and revising the paper.}
+\authordeclaration{The authors declare no conflicts of interest.}
+\equalauthors{\textsuperscript{1} J.M.L and J.H. contributed equally to this work.}
 \correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed: tbedford@fredhutch.org, jbloom@fredhutch.org}
@@
 
-\acknow{We thank Sarah Hilton, Hugh Haddox, and Sidney Bell for helpful discussions about data analysis and we thank Richard Neher for sharing analysis code and providing helpful comments on an initial version of the manuscript.
+\acknow{We thank Sarah Hilton, Hugh Haddox, and Sidney Bell for helpful discussions about data analysis, and Richard Neher for sharing analysis code and providing helpful comments on the manuscript.
 We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.

### 2018-05-08 paper/reviewer_response_PNAS.tex
PNAS reviews in LaTex document

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage{color}
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}

\newcommand{\comment}[1]{{\color{red}[\textsl{#1}]}}

\title{Response to reviews of ``Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants'' for \textit{PNAS}}
\author{Juhye M. Lee, John Huddleston, Michael B. Doud, Kathryn A. Hooper,\\Nicholas C. Wu, Trevor Bedford, Jesse D. Bloom}

\begin{document}
\maketitle

We thank the reviewers for the careful reading and suggestions about the manuscript.
In the revised version, we have performed additional experiments and analyses that should address all of the reviewer concerns.

\emph{Below, the reviewer comments {\color{blue} are in blue}, and our responses are in black.}

\color{blue}

\subsection*{Reviewer \#1 Comments}

\subsubsection*{Comments:} 
In their manuscript Lee et al. examine the tolerance of an H3 virus HA for single point mutations using deep mutational scanning. Using the data obtained from cell culture experiments the authors then try to understand if there is a correlation between the observed changes in vitro and actual virus evolution in vivo. Finally, they test if there is any correlation between H3 and H1 mutational tolerance and basically find that measurements made with H1 HA are not helpful to predict H3 evolution. The manuscript is written in a relatively confusing way and the experimental setup is problematic. Some of the obtained data also need re-evaluation/experimental validation.

\subsubsection*{Major points}

1) It is disturbing that there is no preference for having a methionine in the first position of the HA ORF. The authors discuss this and refer to one 2002 paper that shows additional start sites for non-human H2. This is a pretty weak explanation. If this is really the case the authors should actually isolate/rescue these viruses and test their fitness relative to wild type.

Of note, position 1 for the H1 HA in the supplementary figures is not even shown.

2) While other important features of the HA like several cysteines seem to be conserved, there is no pressure on other elements which are considered important for viral fitness including the NGT glycosylation site in the beginning of HA1 which is conserved in all group 2 HAs or the adjacent NAT glycosylation site which is conserved in H3, H7, H10 and H15. On the other hand, head glycosylation sites, which are highly variable in nature, seem to be conserved. Again, viruses without these glycosylation sites should be rescued and evaluated for fitness relative to wild type.

3) It is also surprising that the transmembrane domain seems to be highly plastic, a feature that is certainly not present in nature where it is highly conserved within H3 HAs.

4) Were sequences from passaged viruses (isolated on MDCKs and other cell lines) included in the comparison between in vitro mutagenesis data and natural evolution data? They need to be excluded since their presence might induce a bias towards what is tolerated/preferred in cell culture. Only sequences obtained directly from patient specimens should be used for the comparison.

5) It is unclear how the 'rescaling' is performed and what it exactly does. The authors need to described and justify this better and should also provide the non-rescaled data in a 'sequence logo' figure for comparison.

6) The authors state that it there are more reports of escape from stalk antibodies for H3 than H1. This is simply not true. Actually, the first reference used as evidence for easy selection of H3 anti-stalk escape mutants (Chai et al) reported three escape mutants which a) did not disrupt binding completely and b) were crippled in vitro and in vivo. The authors should remove their statement as there is not systematic data to support this.

7) The biggest issue with the manuscript is the experimental set up. The authors titrate their rescued virus using a TCID50 assay on MDCK cells. Based on the titer that they measure, they passage the virus from the initial rescue once onto fresh MDCK cells using a low MOI. After infection, the virus dilution stays on the cells and is then harvested without centrifugation followed by RNA extraction and sequencing. There are many issue with this process.

a) The TCID50 is likely underestimating the number of virus particles with HA genomic segments because not all of them (likely the minority) will be capable of inducting a full infectious cycle (but their presence might still be detected in the cell supernatants).

b) The infectious inoculum is never washed off the infected cells meaning that any particles that are transferred from the rescue transfection are going to be detected during sequencing. This might even apply to naked RNPs released from dead cells.

c) This would actually not be an issue, if the dilution from the rescue culture to the new culture would be large enough. But this is not the case. The authors passaged $9\times10^5$ TCID50 (and likely much more DIs/RNA) from the rescue cultures onto the new cells in a volume of 25 ml. This is, depending on experiment, between 10 and 1 ml of supernatant of the original rescue and leads to a 1:2.5 or 1:25 dilution respectively.

d) The supernatant is never centrifuged before RNA extraction. Cell debris is not removed which is likely contributing significantly to RNA sequences detected in the supernatant. However, RNA associated with cell debris is not representative of RNA from fit, replicating viruses.

The experiments should be repeated with at least one more passaging step, proper removal of inoculum and washing of the cells after infection. Centrifugation steps should also be included between the passages and before RNA extraction in the end.

\subsubsection*{Minor points}

1) Many abbreviations are not defined in the manuscript including MDCK-SIAT1, TMPRSS2, MOI, DMEM, FBS, LB.

2) The authors state on page 5 that mutations detected in cell culture as deleterious rarely reach high frequency in nature. Could you please give a few examples?

3) The authors picked 31 random clones to evaluate the mutation rate per clone. These sequences should be made available. Did this set include viruses without a proper start codon (see above)? What was their growth phenotype in vitro?

\subsection*{Reviewer \#2 Comments}

\subsubsection*{Comments:} 
The work presented in this manuscript is a thorough analysis of the tolerance of an H3 hemagglutinin to all possible single amino-acid mutations. H3 evolution is of particular relevance now in light of the most recent flu season where vaccine efficacy against H3N2 was estimated at only 25\%. The experimental setup presented in the manuscript is sound and there is strong correlation among the biological and technical replicates. The authors show that data from deep mutational scanning is better able to describe the evolution of H3 HA compared to conventional models. This is consistent with the results of their previous study using deep mutational scanning on the H1 HA of WSN. It is important to note that their approach is "systematic and quantitative" in capturing potential effects of mutations in the H3 HA on viral fitness in cell culture. The significance statement is accurate in describing the conclusions of the paper and is understandable to a general scientific audience.

\subsubsection*{Major points:}

The authors themselves acknowledge the most important limitation of this study: that the experimental measure of viral growth in cell culture is not reflective of true fitness in nature. Most importantly, the work presented here does not capture host-induced pressures on the antigenic regions of the viral protein. This is evidenced by the discrepancy between the high mutational tolerance they calculated for the stalk region and the actual change in stalk amino acid sequence. A simple protein BLAST comparison between the amino acid sequences of HA from H3N2 viruses Perth09 and HK68 shows a 77\% conservation in head amino acids but a 94\% conservation in stalk amino acids. Additionally, according to figure 2, there is no observable reversion of the G78D mutation they introduced despite the dominance of G78 in the population as seen in figure S1. The authors recognize this in their work and do not attempt to exaggerate the significance of these findings. Still, it would be useful to differentiate between head and stalk mutations when depicting individual amino acid mutations in figure 4 and their frequency trajectories to highlight that the predominance of mutations that occur in nature are in the head.

The data regarding greater mutational tolerance in the stalk compared to the head should be paired with data from deep mutational scanning utilizing stalk- and head-specific antibodies as a selection process as this group has previously done with WSN H1 (Doud et al., Nat Comm 2018, Doud et al., PLoS Path 2017) to see if tolerant sites are responsible for viral escape.

Additionally, the authors should directly show the discrepancies between this in vitro selection and the H3 HA's evolution in nature. This group has performed such an analysis in their previous deep mutational scanning work with the H1 HA of WSN (Doud and Bloom, Viruses 2016). Maximum likelihood phylogenetics can be used to estimate the difference between experimental amino acid preference and natural preference. One might expect that this would provide some insight into the nature of the bias introduced through cell culture.

The authors are upfront about the limited degree to which their calculated mutational effect and the maximum mutation frequency in nature correlate, describing it as "modest." One of the major points of the paper is that conclusions from this type of analysis cannot be generalized when using data from strains that are too divergent. The authors should address the possibility that data from deep mutational scanning can be generalized to divergent strains if the study is done in a system that is more reflective of nature.

The experimental approach of deep mutational scanning has been applied to viral proteins before, including HIV env, influenza H1 HA, and NP. The novelty in this paper is that it is the first time such an analysis has been performed for H3 specifically and that it demonstrates the (albeit limited) utility of deep mutational scanning in predicting the success of viral strains in nature. Ultimately, it is clear from the discussion that the authors see this study as an intermediate step, suggesting that their work is "an improvement over no information at all." While something is indeed better than nothing, in my opinion, this paper represents only an incremental step forward in the field of influenza evolutionary forecasting due to the aforementioned limitations, especially since this experiment and subsequent analysis could be performed using more stringent selection methods during viral passaging like treatment with human sera for data that is presumably more faithful to nature.

\subsubsection*{Minor Points:}

The correlation for prospective analysis (post-Perth) is smaller than the correlation for post-hoc analysis (pre-Perth). Since prospective analysis is more important when predicting evolutionary fates, it might be beneficial for the authors to comment on this difference.

A number of issues interfere with the clarity of this manuscript, particularly when appealing to a broader scientific audience. While this is understandable given the nature of this study, several adjustments should be made. A clear description of mutational effect is never provided in the results, though it forms a crucial component of the data. As such, it is not really intuitive as to why the range from -10-0 should be considered "neutral." Additionally, mutational tolerance is quantified as Shannon entropy of the re-scaled amino-acid preferences at each site, but it is unclear how this entropy is calculated. The entire first half of the second results section describing how the experimentally informed codon substitution model they generate is better than existing models and the associated Table 1 is very difficult to understand. AIC is never explained. The significance of dN/dS being much less than 1 for conventional models should be explained. The purpose of modeling of ExpCM, site avg. is unclear. 
In figure 4, it would be useful to show Perth09 on the tree and mention that closely related strains are excluded from analysis in the figure legend. As it stands, there is just an unlabeled cluster of viruses for which no individual amino acid mutations are characterized. It is also important for the authors to clearly state what frequency they are talking about. Specifically, in consideration of a particular mutation, what is the population in the denominator?

The graph labels in figure 8B could be cleaner. The question marks after "absolutely conserved" and "clade-specific" are extraneous. Indeed, the graph labels themselves ("HA domain," "absolutely conserved? ""clade-specific?") are unnecessary.

\end{document}

### 2018-05-08 paper/reviewer_response_PNAS.tex
outline responses to Reviewer 1

@@
 \newcommand{\comment}[1]{{\color{red}[\textsl{#1}]}}
+\newcommand{\response}[1]{{\color{black}#1}}
 
@@
 
-Of note, position 1 for the H1 HA in the supplementary figures is not even shown.
+\comment{To reply to this point, we need to generate one of these mutants and show that it grows in cell culture.
+{\bf Juhye:} we should discuss how to do this.}
+
+Of note, position 1 for the H1 HA in the supplementary figure is not even shown.
+
+\response{Figure~S3 shows data for the H1 HA as measured previously by Doud and Bloom (2016). 
+That study used mutant libraries in which all codons \emph{except} the one for the N-terminal methionine were mutagenized.
+Therefore, that study did not estimate the effects of mutations to the first codon, which is why it is not shown in Figure~S3.
+We have added a sentence to the legend for this supplementary figure explaining this fact.
+}
 
@@
 
+\comment{{\bf Juhye:} to reply to this point, we should generate several of these mutants and evaluate their growth in cell culture.}
+
 3) It is also surprising that the transmembrane domain seems to be highly plastic, a feature that is certainly not present in nature where it is highly conserved within H3 HAs.
+
+\comment{Here we need to explain that constraint in our experiments and conservation in nature are not expected to be entirely the same.
+Conservation in nature results from the convolution of constraint and diversifying selection from immunity, whereas our experiments only look at constraint.
+As evidence to explain this point, we can point to studies showing that the head of HA is much more conserved for swine influenza viruses where there is less pressure from immunity.
+{\bf Juhye:} To make this more convincing, we should also pick a site in the transmembrane domain that is highly conserved in nature but not under constraint in our experiments, and validate it.}
 
@@
 
-5) It is unclear how the 'rescaling' is performed and what it exactly does.
-The authors need to described and justify this better and should also provide the non-rescaled data in a 'sequence logo' figure for comparison.
+\comment{This is a good suggestion. 
+{\bf John:} I think we should add an additional analysis that does this only for sequences that were not passaged in cell-culture or eggs.
+I believe that GISAID annotates passage history.
+I would expect that we can only do this for the post-Perth sequences as I doubt that there are many unpasaged viruses prior to the early 2000's.
+We could just add this as another panel assuming it gives the same results.}
+
+5) It is unclear how the ``rescaling'' is performed and what it exactly does.
+The authors need to described and justify this better and should also provide the non-rescaled data in a ``sequence logo'' figure for comparison.
+
+\comment{This is easy to explain better; the only reason we didn't go into more detail before is because of space constraints.
+We can also show the non-rescaled logo plots, as Supplemental Information.
+I thought there was a limit on the number of Supplemental Information figures, but I can't find that right now so maybe we can just add them as more...
+If someone objects, we can then merge some of the SI figures into one large multi-panel figure.
+We can explain how the scaling factor doesn't effect any of the evolutionary analyses as it is just a constant factor that multiplies all mutational effects, and so will therefore simply correspond to a change in units on the x-axis of the Figure~5.}
 
@@
 The authors should remove their statement as there is not systematic data to support this.
+
+\comment{We looked at this pretty carefully, and there are definitely more reports of selecting anti-stalk antibodies for H1 and H3.
+The reviewer's counterpoint is to mention on of the H3 studies. 
+However, we didn't say that there are no H3 reports, just that there are a lot more H1 reports.
+Certainly this is not systematic proof, but our statement in the text goes to great pains to caveat that this is purely circumstantial and not systematic proof.
+But since this isn't a central claim of the paper of anyway, let's just remove it in a concession to the reviewer.
+The study we're helping Nicholas Wu with will eventually systematically test this point.
+}
 
@@
 
+\response{The reviewer has a number of questions and suggestions about the experimental design.
+Over the last few years, we have performed a number of studies optimizing procedures for deep mutational scanning of influenza virus libraries (see PMIDs 24859245, 25006036, 27271655) during which we have carefully considered all of the experimental parameters that the reviewer discusses.
+Below we briefly explain why each of our choices for these experimental parameters were appropriate.
+We do not deny that there are slightly different ways that the experiments could be done, but the in end, the ``proof is in the pudding.'' 
+In the manuscript, we show that the measurements made using the experimental parameters that we chose lead to useful data as judged by the following criterion:
+\begin{itemize}
+\item The results from fully independent experimental replicates are highly correlated (Fig.~1D), which would not be the case if the results were dominated by noise.
+\item The experimental measurements do a good job of predicting the effects of individually validated mutations \comment{Here we would refer to the validation experiments described above}.
+\item The experimental measurements can inform substitution models that describe influenza virus's evolution much better than standard models (Table~1).
+\item The experimental measurements provide information that can help predict the fates of influenza virus lineages in nature (Fig.~5).
+\end{itemize}
+We do not claim that other experimental parameter choices might not lead to equivalent or even slightly better results, but we have carefully considered all of our choices and clearly they lead to very good results.
+
+Below we explain our reasoning for each choice about which the reviewer makes suggestions:
+}
+
 a) The TCID50 is likely underestimating the number of virus particles with HA genomic segments because not all of them (likely the minority) will be capable of inducting a full infectious cycle (but their presence might still be detected in the cell supernatants).
+
+\response{The reviewer is certainly correct that TCID50 assays underestimate the number of HA genomic segments, as it is well known that the number of physical influenza virions exceeds the number of infectious units by 10-100 fold.
+This is why we used an extremely low MOI (0.0035) so that most cells are still infected by no more than one virion.
+It is impossible to ever grow a viral stock that is completely free of defective particles.
+However, in both our original rescue with helper virus and in the passage, we follow the steps suggested by Carolina Lopez and Scott Hensley (PMID 27047455) to minimize defective particles.
+These include low MOI (0.0035 TCID50 per cell is quite low), and short growth times to avoid accumulation of defective particles (the helper-virus rescue was allowed to propagate for just 24 hours, and the low-MOI passage for just 48 hours).
+\comment{Juhye, do you have HA-staining titers for these viruses too? Then we can explicitly compare HA-expressing units to TCID50.}
+}
 
@@
 This might even apply to naked RNPs released from dead cells.
+
+\response{It is true that the original inoculum remains and so will contribute to the variants detected during sequencing.
+However, the original inoculum makes a negligible contribution to the total number of virions that are present at the time of sequencing.
+Specifically, in the original inoculum placed on the cells, the titer is \comment{?} TCID50 / ml, whereas at the time of supernatant collection for sequencing the titer is \comment{?} TCID50 / ml.
+Therefore, the original inoculum constitutes just \comment{?}\% of the viral titer at the time of collection, and therefore makes a negligible contribution to the sequenced material.
+\comment{Juhye, can you calculate these numbers from the TCID50 of the collected viruses. 
+You can also mention those titers in the Methods.
+I expect that we will see at least a 10,000-fold dilution of the original inoculum.
+If we really want to nail this point (and the one after it), we could also do qPCR on the inoculum (we could re-constitute some from a bit of transfection supernatant) and the passage supernatant to show that the HA RNA amounts also changed comparably.
+I don't think this is really necessary since it's obvious, but we could do it if we want to really drive a stake through the heart of the reviewers' concern...}
+}
 
@@
 
+\comment{The reviewer's math is off here, we passaged $9\times 10^5$ TCID50 \emph{total} across all 15 cm dishes. 
+This could be clarified by adding a word to the Methods indicating how many 15-cm dishes we used (right now, it just says ``in 15-cm dishes.''
+Below that we say that we used an inoculum of 2.5 TCID50 / $\mu$l, which corresponds to dilutions that range from 1:37 to 1:294, not 1:2.5 or 1:25 as the reviewer says.
+In addition, we should point out that the supernatant comes from a helper-virus rescue that has occurred for just 24 hours, and so has low titers of both infectious virus and TCID50s.
+Again, we could really drive a stake through this question by doing the qPCR mentioned above.
+}
+
 d) The supernatant is never centrifuged before RNA extraction.
@@
 
+\response{The reviewer is correct that we never clarified the viral supernatant before RNA extraction.
+There are two reason for this.
+First, for a relatively short and very low MOI passage like the one we used (48 hours starting at a MOI of 0.0035), there was little cell debri since the cell monolayer was still largely intact at the end of the passage.
+Second, there is no reason to expect that the RNA in supernatant virions will better correspond to functional HA mutants that RNA in cell debris.
+If anything, the opposite might be expected.
+Functional HA is needed only for viral entry.
+The RNA in cell debris at least originated from a virus that was able to enter the cell, whereas the HA in released virions has not yet undergone selection beyond that which occurred at the infection stage.
+In practice, after 48 hours, most RNA is from viruses that have undergone multiple cycles of infection, and so there has been reasonably strong selection for HA function in both virion RNA and cell debris RNA.
+}
+
 The experiments should be repeated with at least one more passaging step, proper removal of inoculum and washing of the cells after infection.
 Centrifugation steps should also be included between the passages and before RNA extraction in the end.
+
+\response{For the reasons described above, it is not necessary to totally repeat the experiments.
+Certainly other experimental parameter choices could have been used, but this is always the case for complex experiments.
+Our choices will reasonable, and the four lines of evidence listed at the beginning of this comment show that they yield reproducible and meaningful results.}
 

### 2018-05-08 paper/manuscript.tex
added equal contributions for JML and JH
outline responses to Reviewer 1

@@
 The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
+Note that \cite{doud2016accurate} used libraries in which all codons were mutagenized \emph{except} for the one encoding N-terminal methionine.
+Therefore, no data is shown for the first codon in the gene.
 }

### 2018-05-08 paper/reviewer_response_PNAS.tex
full outline of proposed reviewer responses

@@
 This is consistent with the results of their previous study using deep mutational scanning on the H1 HA of WSN.
-It is important to note that their approach is "systematic and quantitative" in capturing potential effects of mutations in the H3 HA on viral fitness in cell culture.
+It is important to note that their approach is ``systematic and quantitative'' in capturing potential effects of mutations in the H3 HA on viral fitness in cell culture.
 The significance statement is accurate in describing the conclusions of the paper and is understandable to a general scientific audience.
+
+\comment{We thank the reviewer for the accurate and fair summary of our work.}
 
@@
 
+\response{We thank the reviewer for acknowledging our efforts to fairly present the considerable caveats associated with cell-culture measurements.
+
+The suggestion to make a version of Fig.~4 that distinguished between the stalk and the head is a good one.
+It is difficult to also overlay the structural location (stalk versus head) on the current Fig.~4, since colors are already used to indicate mutational effects.
+We have therefore created a new supplemental figure (Fig.~S\comment{?}) that shows these data---as the reviewer suspected, this shows that head mutations are much more common.
+\comment{{\bf John:} Can you make this figure? I'm actually not 100\% sure of the point, as I don't think the layout in Fig.~4 is the best way to show head versus stalk. 
+But since it's what the reviewer suggested, let's just do it and make it a supplement.
+This would be just like the current Fig.~4 except rather than coloring by mutational effect, we color by head versus stalk.
+Then I think we should also make a second panel that shows the number of head and stalk mutations that reach an indicated frequency over the whole timespan, as this is really a better way to show the point.
+So the x-axis would be mutation frequency, we would count mutations that reach each frequency, bin them and show this for head and stalk.
+This would definitely show that there are more head mutations overall, and would probably show that more head mutations reach high frequency.
+You could argue that it would be more informative to do this for epitope versus non-epitope, but since the reviewer asked for head versus stalk and we won't really discuss this much in the text anyway, let's just do head versus stalk...}
+
+The reviewer is correct that G78D appears to be a site where the correspondence between selection in our experiments and frequency in nature is low.
+This makes sense, as G78D is one of the two sites that we specifically identified as being a lab adaptation (see Figure~S1).
+We now explicitly mention this fact in the text \comment{still need to actually add this mention}.
+}
+
 The data regarding greater mutational tolerance in the stalk compared to the head should be paired with data from deep mutational scanning utilizing stalk- and head-specific antibodies as a selection process as this group has previously done with WSN H1 (Doud et al., Nat Comm 2018, Doud et al., PLoS Path 2017) to see if tolerant sites are responsible for viral escape.
+
+\comment{This would be interesting to do, but is a whole other study.
+This is the one point that I think we are just going to have to argue is beyond the scope of the manuscript.
+}
 
@@
 
-The authors are upfront about the limited degree to which their calculated mutational effect and the maximum mutation frequency in nature correlate, describing it as "modest." One of the major points of the paper is that conclusions from this type of analysis cannot be generalized when using data from strains that are too divergent.
+\comment{We can definitely do this pretty easily, perhaps adding Sarah Hilton as an author to help.
+Not sure where we would fit the data given that the paper is already too long, but I guess we could cram it in the supplement.
+}
+
+The authors are upfront about the limited degree to which their calculated mutational effect and the maximum mutation frequency in nature correlate, describing it as ``modest.'' One of the major points of the paper is that conclusions from this type of analysis cannot be generalized when using data from strains that are too divergent.
 The authors should address the possibility that data from deep mutational scanning can be generalized to divergent strains if the study is done in a system that is more reflective of nature.
+
+\comment{This is a possibility, although I highly doubt that it is true.
+However, we can easily add a sentence acknowledging the possibility.
+}
 
@@
 
+\comment{This is a fair point.
+However, I think we should argue that it's the first paper to show deep mutational scanning can actually provide at least some information that reflects outcomes in nature.
+This comment does make me worry that the reviewer just doesn't think the work is significant enough, but since the editor did ask for a revision you'd have to hope that if we answer all specific comments that would be enough.
+}
+
 \subsubsection*{Minor Points:}
@@
 Since prospective analysis is more important when predicting evolutionary fates, it might be beneficial for the authors to comment on this difference.
+
+\comment{Not sure if the difference between $\rho = 0.17$ and $\rho = 0.14$ is significant. 
+We could figure out how to calculate this, but first let's see what it looks like for the non-lab-passaged sequences requested by Reviewer 1.}
 
@@
 is unclear.
+
+\comment{We can clarify these easily.}
+
 In figure 4, it would be useful to show Perth09 on the tree and mention that closely related strains are excluded from analysis in the figure legend.
@@
 Specifically, in consideration of a particular mutation, what is the population in the denominator?
+
+\comment{We can clarify these easily.}
 
@@
 
+\comment{Easily fixed.}
+
 \end{document}

### 2018-05-08 paper/reviewer_response_PNAS.tex
added outline responses to minor points from Reviewer 1

@@
 
+\comment{Easy to address.}
+
 2) The authors state on page 5 that mutations detected in cell culture as deleterious rarely reach high frequency in nature. Could you please give a few examples?
+
+\comment{Not obvious how to give an example, as there are many mutations we identify as deleterious and many mutations that don't reach high frequency. 
+But I suppose if we validate some specific mutation (e.g., a cysteine mutation) as deleterious, we can point out that it doesn't reach high frequency...}
 
@@
 What was their growth phenotype in vitro?
+
+\comment{We can add these sequences to Figure~S1.}
 

### 2018-05-09 paper/reviewer_response_PNAS.tex
Added Juhye comments to reviewer responses and small updates to Methods

@@
 
+\comment{JML: OK, agreed.}
+
 Of note, position 1 for the H1 HA in the supplementary figure is not even shown.
@@
 
+\comment{JML: We can test some of these as well.
+But fitness in nature is not necessarily reflective of fitness in a cell culture system, and just because something is conserved in natural sequences does not necessarily mean it is indisposable for viral growth in the lab....
+This is related to point 3 below.}
+
 3) It is also surprising that the transmembrane domain seems to be highly plastic, a feature that is certainly not present in nature where it is highly conserved within H3 HAs.
@@
 {\bf Juhye:} To make this more convincing, we should also pick a site in the transmembrane domain that is highly conserved in nature but not under constraint in our experiments, and validate it.}
+
+\comment{JML: OK, can test a couple of these as well.
+I'd also like to point out (1) that the WSN HA also exhibits some degree of flexibility in amino acids that are tolerated in the TM domain, and (2) that various studies have found that chimeric HA's with transmembrane domains from {\bf completely different proteins} can still maintain HA cell surface expression and fusion activity (Melikyan et al 1999 Mol Biol Cell, Schroth-Diez et al 1998 JVI, Dong et al 1992 JVI, Roth et al 1986 JCB) and that the {\bf length} of the TM domain seems to play a large role in determining HA fusion activity (Armstrong et al 2000 JCB).}
 
@@
 
+\comment{JML: There is a limit of 10 supplementary files + figures combined, and we are currently at that limit.
+We could make a multipanel with the WSN HA logoplot, although I agree that this wouldn't be ideal.
+We could also just add it to the Jupyter notebook and link to that as well I suppose}
+
 6) The authors state that it there are more reports of escape from stalk antibodies for H3 than H1.
@@
 }
+
+\comment{JML: Agreed that it may be better just to remove that sentence.}
 
@@
 
+\comment{JML: I only titered by TCID50, but we used a super low MOI and see points below.}
+
 b) The infectious inoculum is never washed off the infected cells meaning that any particles that are transferred from the rescue transfection are going to be detected during sequencing.
@@
 However, the original inoculum makes a negligible contribution to the total number of virions that are present at the time of sequencing.
-Specifically, in the original inoculum placed on the cells, the titer is \comment{?} TCID50 / ml, whereas at the time of supernatant collection for sequencing the titer is \comment{?} TCID50 / ml.
+Specifically, in the original inoculum placed on the cells, the titer is \comment{100 to 700} TCID50 / ul, whereas at the time of supernatant collection for sequencing the titer is \comment{30 to 230} TCID50 / ul for unconcentrated virus \comment{but 400 to 7550 TCID50 / ul for the concentrated virus}.
 Therefore, the original inoculum constitutes just \comment{?}\% of the viral titer at the time of collection, and therefore makes a negligible contribution to the sequenced material.
@@
 }
+
+\comment{JML: Actually I {\bf did} change the media.
+I was really concerned about viral particles hanging around, so 3 hours post-infection, I changed the media to fresh IGM.
+I have updated the Methods to reflect this.
+So particles transferred from the rescue transfection, even if they make up a large proportion of the original inoculum are not going to be detected anyway.}
 
@@
 }
+
+\comment{JML: Yes, the $9\times 10^5$ TCID50 was across 15 x 15 cm plates. Also added this to the Methods.}
 
@@
 
+\comment{JML: Actually I {\bf did} centrifuge the viral supernatant at 2000xg for 5 min to clarify any cellular debris.
+I then took the clarified supernatant and ultracentrifuged 24 mL of this for RNA extraction.
+Updated the methods to reflect this.}
+
 The experiments should be repeated with at least one more passaging step, proper removal of inoculum and washing of the cells after infection.
@@
 Our choices will reasonable, and the four lines of evidence listed at the beginning of this comment show that they yield reproducible and meaningful results.}
+
+\comment{JML: see JML comments above}
 
@@
 \comment{We can add these sequences to Figure~S1.}
+
+\comment{JML: Two of the 31 clones had mutations in the start codon. I'll compile a list of the mutations in the 31 clones}.
 
@@
 
+\comment{JML: I've also been wanting to do this and can do this easily. But it depends on how much space we have (probably not much)}.
+
 The authors are upfront about the limited degree to which their calculated mutational effect and the maximum mutation frequency in nature correlate, describing it as ``modest.'' One of the major points of the paper is that conclusions from this type of analysis cannot be generalized when using data from strains that are too divergent.

### 2018-05-09 paper/manuscript.tex
Added Juhye comments to reviewer responses and small updates to Methods

@@
 We passaged $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
-To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they were at $\sim 1.7 \times 10^7$ cells per dish.
+To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in fifteen 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they were at $\sim 1.7 \times 10^7$ cells per dish.
 We replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L in IGM.
+Three hours post-infection, we replaced the inoculum with fresh IGM.
 We collected viral supernatant for sequencing 48 hours post-infection.
@@
 \subsection*{Barcoded subamplicon sequencing}
-To extract viral RNA from the three replicate HA virus libraries and the wildtype HA virus, we ultracentrifuged 24 mL of supernatant at 22,000 rpm for 1.5 h at 4$^{\circ}$C in a Beckman Coulter SW28 rotor, and extracted RNA using the Qiagen RNeasy Mini Kit by resuspending the viral pellet in 400 $\mu$L of buffer RLT supplemented with $\beta$-mercaptoethanol, pipetting 30 times, transferring the liquid to a microcentrifuge tube, adding 600 $\mu$L 70\% ethanol, and proceeding according to the manufacturer's instructions.
+To extract viral RNA from the three replicate HA virus libraries and the wildtype HA virus, we clarified the viral supernatant by centrifuging at 2000 x \textit{g} for five minutes, then ultracentrifuged 24 mL of the clarified supernatant at 22,000 rpm for 1.5 h at 4$^{\circ}$C in a Beckman Coulter SW28 rotor, and extracted RNA using the Qiagen RNeasy Mini Kit by resuspending the viral pellet in 400 $\mu$L of buffer RLT supplemented with $\beta$-mercaptoethanol, pipetting 30 times, transferring the liquid to a microcentrifuge tube, adding 600 $\mu$L 70\% ethanol, and proceeding according to the manufacturer's instructions.
 The HA gene was reverse-transcribed with AccuScript Reverse Transcriptase (Agilent 200820) using primers P09-HA-For (5'-AGCAAAAGCAGGGGATAATTCTATTAATC-3') and P09-HA-Rev (5'-AGTAGAAACAAGGGTGTTTTTAATTACTAATACAC-3').

### 2018-06-04 paper/reviewer_response_PNAS.tex
Start integrating new figures, update paper text, fig captions, start updating reviewer responses

@@
 
+\response{We thank the reviewer for this good suggestion and now include an analysis of the relationship between mutational effect and maximum frequency for mutations from unpassaged isolates.
+Due to the sparseness of unpassaged isolate sequences previous to the last decade, we analyzed unpassaged isolates from the post-Perth/2009 partition of the H3N2 HA phylogenetic tree.
+Figures 5, 6, and S6 now include panels showing the results for these unpassaged isolate mutations.
+The results in Figure 5 demonstrate that the correlation between mutational effect as determined by our experiments and maximum mutation frequency is stronger for unpassaged isolates than when we include all isolates ($\rho = 0.24$).
+We do not see such a correlation when we use the WSN/1933 H1 HA preferences to estimate the mutational effects for unpassaged isolate mutations.
+}
+
 \comment{This is a good suggestion. 
@@
 
+\response{Rescaling of the preferences is performed by fitting a stringency parameter that is a reflection of how strongly our experimentally determined preferences correspond to the gene's evolution in nature.
+This parameter is fit under an experimentally informed codon substitution model (ExpCM) using the phydms program \comment{reference here}.
+A stringency parameter greater than 1 indicates that natural selection favors the same amino acids as the experiments but with greater stringency, while a value less than 1 indicates that natural selection does not favor the amino acids with as much stringency as the experiments.
+Because selection in the lab often does not fully reflect the stringency of selection in nature and selection across experiments may not be entirely consistent, we re-scale the preferences by this stringency parameter.
+We now include a supplementary panel A to Figure S3 showing the un-scaled Perth/2009 H3 HA preferences.
+}
+
 \comment{This is easy to explain better; the only reason we didn't go into more detail before is because of space constraints.
@@
 We can explain how the scaling factor doesn't effect any of the evolutionary analyses as it is just a constant factor that multiplies all mutational effects, and so will therefore simply correspond to a change in units on the x-axis of the Figure~5.}
-
-\comment{JML: There is a limit of 10 supplementary files + figures combined, and we are currently at that limit.
-We could make a multipanel with the WSN HA logoplot, although I agree that this wouldn't be ideal.
-We could also just add it to the Jupyter notebook and link to that as well I suppose}
 
@@
 
+\response{We agree that this would be an interesting experiment to perform...}
+
 \comment{This would be interesting to do, but is a whole other study.
@@
 
+\response{We now have acknowledged this possibility in the Discussion in the sentence "We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains.}
+
 \comment{This is a possibility, although I highly doubt that it is true.
@@
 Since prospective analysis is more important when predicting evolutionary fates, it might be beneficial for the authors to comment on this difference.
+
+\response{We now include an analysis of mutational effects and maximum frequency for mutations of unpassaged isolates (see Figures 5, 6, and S6) from the post-Perth partition of the phylogenetic tree to limit any potential biases that our experimental measurements may have towards mutations from cell- or egg-passaged isolates.
+The results show that this correlation is stronger when considering mutations from unpassaged isolates in a post-hoc analysis, demonstrating the potential utility of our experimental measurements in a prospective analysis, specifically when passaged isolate sequences are excluded.
+}
 
@@
 
-\comment{Easily fixed.}
+\response{We have now fixed the graph labels in figure 8B accordingly.}
 

### 2018-06-04 paper/manuscript.tex
Start integrating new figures, update paper text, fig captions, start updating reviewer responses

@@
 The vertical black line shows the mean mutation effect for each category.
-The analysis is performed separately for the pre- and post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
+The analysis is performed separately for pre-Perth/2009, post-Perth/2009, and unpassaged isolates from the post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
 }
@@
 
-We next sought to quantify the correlation between a mutation's experimentally measured effect and the maximum frequency it attained during natural evoluton.
+We next sought to quantify the correlation between a mutation's experimentally measured effect and the maximum frequency it attained during natural evolution.
 To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
@@
 In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).
+To restrict potential biases of our experimental measurements towards mutations from cell- or egg-passaged isolates, we evaluated mutations from unpassaged isolates.
+Due to the limited availability of primary isolate sequences until the last decade, we analyzed mutations of unpassaged isolates from the post-Perth partition of the phylogenetic tree~\cite{mcwhite2016sequence}.
+The correlation between mutational effect and maximum frequency was stronger for mutations from unpassaged post-Perth isolates ($\rho = 0.24$).
 The similar effect sizes for both partitions shows that our experimental measurements can help explain the evolutionary fates of mutations in strains that post-date the experimentally studied strain, as well as to retrospectively analyze mutations that precede the experimental strain.
@@
 The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative---and so represent an improvement over no information at all.
-We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility.
+We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains.
 
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=\textwidth]{figs/supp_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
+\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_WSNprefs_logoplot/supp_logoplots.pdf}}
 \caption{\label{suppfig:WSNprefs_logoplot}
-{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA as measured in \cite{doud2016accurate}.}
-The amino-acid preferences from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
-The sites are in H3 numbering.
+{\bf The site-specific amino-acid preferences of the Perth/2009 H3 HA and the WSN/1933 H1 HA.}
+(A) The un-scaled amino-acid preferences of the Perth/2009 H3 HA.
+The overlays show the same information as in Figure~\ref{fig:logoplot} (domain, epitope site, and wildtype amino acid).
+(B) The amino-acid preferences of the WSN/1933 H1 HA from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
 The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
@@
 Therefore, no data is shown for the first codon in the gene.
+The sites are in H3 numbering for both logo plots.
 }
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=1.0\textwidth]{figs/supp_H3N2_phylogeny/H3N2_phylogeny.pdf}}
+\centerline{\includegraphics[width=0.9\textwidth]{figs/supp_H3N2_phylogeny/H3N2_phylogeny.pdf}}
 \caption{\label{suppfig:tree}
 {\bf A phylogenetic tree of all HA sequences used in our analysis of mutation frequencies.}
-HA sequences were sampled at a rate of six viruses per month from January 1, 1968 through February 1, 2018.
+(A) HA sequences were sampled at a rate of six viruses per month from January 1, 1968 through February 1, 2018.
 The Perth/2009 strain used in our experiments is indicated.
@@
 The post-2014 nodes were excluded because the evolutionary fates of many sequences after this date are not yet full resolved.
+(B) The post-Perth/2009 partition of the tree containing only sequences from unpassaged isolates.
 }

### 2018-06-08 paper/manuscript.tex
Added methods and reference for unpassaged virus analyses.

@@
 
+To determine the effect of virus passaging on the relationship between mutational effects and maximum frequencies in natural populations, we repeated the above analyses using only viruses that were sequenced directly without passaging.
+Routine direct sequencing did not begin until the early 2000s \cite{McWhite2016}.
+To construct a tree with a similar number of viruses as the original analysis, we sampled 30 viruses per month between January 1, 2000 and April 1, 2018, producing a tree with 2,374 unpassaged viruses with augur (commit: 6d9f708).
+We included the passaged DMS strain, A/Perth/16/2009, in the resulting tree to enable comparison between pre-Perth and post-Perth clades.
+
 \subsection*{Analysis of mutational shifts}

### 2018-06-20 paper/reviewer_response_PNAS.tex
Update supp figures & paper text, add to reviewer responses

--- 2018-05-08 paper/reviewer_response_PNAS.tex
@@
 
-\comment{To reply to this point, we need to generate one of these mutants and show that it grows in cell culture.
-{\bf Juhye:} we should discuss how to do this.}
+\response{We have now rescued by reverse genetics influenza GFP viruses carrying the M(-16)K mutation in the start codon, and titered by flow cytometry. 
+By 53 hours post-transfection when we harvested transfection supernatant, the M(-16)K mutant reached a titer of approximately one log less than that of wildtype virus.
+Although we may have expected the M(-16)K mutant to grow comparably to wildtype given that Met and Lys are both preferred at the start codon, we would like to emphasize that (1) we grew influenza GFP viruses which may not behave identically to live influenza viruses and (2) we did not capture the full viral growth kinetics of this mutant since we titered at a relatively early timepoint after transfection.
+In comparison, the C52A variant, which mutates the highly conserved Cys at site 52, grows very poorly, while the synonymous C52C mutant grows to titers comparable to those of wildtype.
+These results are now shown in Figure S9, and we have updated the Methods section describing how we validated these point mutants.
+}
 
@@
 
-\comment{{\bf Juhye:} to reply to this point, we should generate several of these mutants and evaluate their growth in cell culture.}
+\response{This is a valid point, and we have also individually evaluated the growth of three glycosylation site mutants.
+Two of the mutants, T24F and T40V, remove the aforementioned highly conserved but mutationally tolerant NGT and NAT glycosylation sites, respectively.
+The third mutant, S287A, removes a constrained glycosylation site.
+In Figure S9, we show that the T24F and T40V mutants grow to titers comparable to or higher than that of wildtype. 
+However, the titer of the S287A is approximately 2.8-fold lower than that of wildtype, suggesting that the glycan at sites 285-287 is important for viral growth in cell culture.
+
+Furthermore, constraint in our experiments and conservation in nature are not expected to be entirely the same.
+Conservation in nature results from the convolution of constraint and diversifying selection from immunity, whereas our experiments only look at constraint.
+\comment{As evidence to explain this point, we can point to studies showing that the head of HA is much more conserved for swine influenza viruses where there is less pressure from immunity.}
+}
+
+\comment{Alanine is not the least preferred amino acid at site 287, while it is the least preferred at site 52.}
 
@@
 
-\comment{Here we need to explain that constraint in our experiments and conservation in nature are not expected to be entirely the same.
-Conservation in nature results from the convolution of constraint and diversifying selection from immunity, whereas our experiments only look at constraint.
-As evidence to explain this point, we can point to studies showing that the head of HA is much more conserved for swine influenza viruses where there is less pressure from immunity.
-{\bf Juhye:} To make this more convincing, we should also pick a site in the transmembrane domain that is highly conserved in nature but not under constraint in our experiments, and validate it.}
+\response{We validated growth of a transmembrane domain mutant, C199(HA2)K, and show that this mutant attains titers approximately 1.6-fold higher than that of wildtype.
+In addition, various studies have found that chimeric HA's with transmembrane domains from completely different proteins can still maintain HA cell surface expression and fusion activity (Melikyan et al 1999 Mol Biol Cell, Schroth-Diez et al 1998 JVI, Dong et al 1992 JVI, Roth et al 1986 JCB), and that the length of the transmembrane domain seems to play a large role in determining HA fusion activity (Armstrong et al 2000 JCB).
+}
 
@@
 
-\comment{This is a good suggestion. 
-{\bf John:} I think we should add an additional analysis that does this only for sequences that were not passaged in cell-culture or eggs.
-I believe that GISAID annotates passage history.
-I would expect that we can only do this for the post-Perth sequences as I doubt that there are many unpasaged viruses prior to the early 2000's.
-We could just add this as another panel assuming it gives the same results.}
+\response{We thank the reviewer for this astute suggestion and now include an analysis of the relationship between mutational effect and maximum frequency for mutations from unpassaged isolates.
+Due to the sparseness of unpassaged isolate sequences previous to the last decade, we analyzed unpassaged isolates from the post-Perth/2009 partition of the H3N2 HA phylogenetic tree.
+Figures 5, 6, and S7 now include panels showing the results for these unpassaged isolate mutations.
+The results in Figure 5 demonstrate that the correlation between mutational effect as determined by our experiments and maximum mutation frequency is stronger for unpassaged isolates than when we include all isolates ($\rho = 0.24$).
+We do not see such a correlation when we use the WSN/1933 H1 HA preferences to estimate the mutational effects for unpassaged isolate mutations.
+}
 
@@
 
-\comment{This is easy to explain better; the only reason we didn't go into more detail before is because of space constraints.
-We can also show the non-rescaled logo plots, as Supplemental Information.
-I thought there was a limit on the number of Supplemental Information figures, but I can't find that right now so maybe we can just add them as more...
-If someone objects, we can then merge some of the SI figures into one large multi-panel figure.
-We can explain how the scaling factor doesn't effect any of the evolutionary analyses as it is just a constant factor that multiplies all mutational effects, and so will therefore simply correspond to a change in units on the x-axis of the Figure~5.}
+\response{Rescaling of the preferences is performed by fitting a stringency parameter that is a reflection of how strongly our experimentally determined preferences correspond to the gene's evolution in nature.
+This parameter is fit under an experimentally informed codon substitution model (ExpCM) using the phydms program (Hilton 2017 PeerJ).
+A stringency parameter greater than 1 indicates that natural selection favors the same amino acids as the experiments but with greater stringency, while a value less than 1 indicates that natural selection does not favor the amino acids with as much stringency as the experiments.
+Because selection in the lab often does not fully reflect the stringency of selection in nature and selection across experiments may not be entirely consistent, we re-scale the preferences by this stringency parameter.
+Importantly, re-scaling does not affect any of the evolutionary analyses performed in this paper since it is a constant that multiplies all mutational effects, simply corresponding a change in units on the x-axis of Figure~5.
+We now include a supplementary panel A to Figure S3 showing the un-scaled Perth/2009 H3 HA preferences.
+}
 
@@
 The authors should remove their statement as there is not systematic data to support this.
+
+\response{We have now removed this sentence from the paper.}
 
@@
 \item The results from fully independent experimental replicates are highly correlated (Fig.~1D), which would not be the case if the results were dominated by noise.
-\item The experimental measurements do a good job of predicting the effects of individually validated mutations \comment{Here we would refer to the validation experiments described above}.
+\item The experimental measurements do a good job of predicting the effects of individually validated mutations, as shown in Figure S8.
 \item The experimental measurements can inform substitution models that describe influenza virus's evolution much better than standard models (Table~1).
@@
 However, in both our original rescue with helper virus and in the passage, we follow the steps suggested by Carolina Lopez and Scott Hensley (PMID 27047455) to minimize defective particles.
-These include low MOI (0.0035 TCID50 per cell is quite low), and short growth times to avoid accumulation of defective particles (the helper-virus rescue was allowed to propagate for just 24 hours, and the low-MOI passage for just 48 hours).
-\comment{Juhye, do you have HA-staining titers for these viruses too? Then we can explicitly compare HA-expressing units to TCID50.}
+These include low MOI (0.0035 TCID50 per cell is quite low), short growth times to avoid accumulation of defective particles (the helper-virus rescue was allowed to propagate for just 24 hours, and the low-MOI passage for just 48 hours), and replacing the media three hours post-infection to remove residual defective particles.
 }
@@
 
-\response{It is true that the original inoculum remains and so will contribute to the variants detected during sequencing.
-However, the original inoculum makes a negligible contribution to the total number of virions that are present at the time of sequencing.
-Specifically, in the original inoculum placed on the cells, the titer is \comment{?} TCID50 / ml, whereas at the time of supernatant collection for sequencing the titer is \comment{?} TCID50 / ml.
-Therefore, the original inoculum constitutes just \comment{?}\% of the viral titer at the time of collection, and therefore makes a negligible contribution to the sequenced material.
-\comment{Juhye, can you calculate these numbers from the TCID50 of the collected viruses. 
-You can also mention those titers in the Methods.
-I expect that we will see at least a 10,000-fold dilution of the original inoculum.
-If we really want to nail this point (and the one after it), we could also do qPCR on the inoculum (we could re-constitute some from a bit of transfection supernatant) and the passage supernatant to show that the HA RNA amounts also changed comparably.
-I don't think this is really necessary since it's obvious, but we could do it if we want to really drive a stake through the heart of the reviewers' concern...}
+\response{For biological replicates 1, 2, and 3-2, we replaced the original infectious inoculum with 25 mL fresh IGM three hours post-infection.
+For replicate 3-1, we did not change the media. 
+However, for this replicate, the original inoculum makes a negligible contribution to the total number of virions that are present at the time of sequencing.
+Specifically, in the original inoculum placed on the cells, the titer is 2.5 TCID50 / ul, whereas at the time of supernatant collection for sequencing the titer is 32 TCID50 / ul for unconcentrated virus but 400 TCID50 / ul for the concentrated virus.
+Therefore, the original inoculum constitutes just 8\% of the viral titer at the time of collection, and therefore makes a negligible contribution to the sequenced material.
+We deeply apologize for not including this crucial step in the Methods, and have updated the Methods to now include completion of this step. 
+We have also included in the GitHub repository a PDF of the lab notebook describing the rescue and passage of the mutant virus libraries at %\url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/lab_notes/Perth09%20HA%20Library%20Rescue.pdf} (relevant details of the notebook can be found in pages 20-25).
 }
@@
 
-\comment{The reviewer's math is off here, we passaged $9\times 10^5$ TCID50 \emph{total} across all 15 cm dishes. 
-This could be clarified by adding a word to the Methods indicating how many 15-cm dishes we used (right now, it just says ``in 15-cm dishes.''
-Below that we say that we used an inoculum of 2.5 TCID50 / $\mu$l, which corresponds to dilutions that range from 1:37 to 1:294, not 1:2.5 or 1:25 as the reviewer says.
-In addition, we should point out that the supernatant comes from a helper-virus rescue that has occurred for just 24 hours, and so has low titers of both infectious virus and TCID50s.
-Again, we could really drive a stake through this question by doing the qPCR mentioned above.
+\response{We passaged for each mutant virus library $9\times10^5$ TCID50 total across 15 $\times$ 15-cm dishes.
+Because we used an inoculum of 2.5 TCID50 / $\mu$l across these fifteen dishes, this corresponds to dilutions that range from 1:37 to 1:294.
+The quantity of dishes used to passage each mutant virus library is now mentioned in the Methods.
 }
@@
 
-\response{The reviewer is correct that we never clarified the viral supernatant before RNA extraction.
-There are two reason for this.
-First, for a relatively short and very low MOI passage like the one we used (48 hours starting at a MOI of 0.0035), there was little cell debri since the cell monolayer was still largely intact at the end of the passage.
-Second, there is no reason to expect that the RNA in supernatant virions will better correspond to functional HA mutants that RNA in cell debris.
-If anything, the opposite might be expected.
-Functional HA is needed only for viral entry.
-The RNA in cell debris at least originated from a virus that was able to enter the cell, whereas the HA in released virions has not yet undergone selection beyond that which occurred at the infection stage.
-In practice, after 48 hours, most RNA is from viruses that have undergone multiple cycles of infection, and so there has been reasonably strong selection for HA function in both virion RNA and cell debris RNA.
+\response{We clarified the viral supernatant before RNA extraction by centrifuging the supernatant at 2000 $\times$ \textit{g} for five minutes to remove any cellular debris.
+We failed to mention this step in the Methods, and again deeply apologize for this critical oversight.
+We have updated the Methods to now include mention of this step.
 }
@@
 
-\comment{Easy to address.}
+\response{We now explain these abbreviations in the manuscript.}
 
@@
 
-\comment{We can add these sequences to Figure~S1.}
+\response{We have now added the list of mutations from the 31 clones to Dataset S3.
+Two of the clones had a mutation that removed the ATG start codon.
+}
+
+\comment{Not sure how to address the growth phenotype question}.
 
@@
 
-\comment{We thank the reviewer for the accurate and fair summary of our work.}
+\response{We thank the reviewer for an accurate and fair summary of our work.}
 
@@
 It is difficult to also overlay the structural location (stalk versus head) on the current Fig.~4, since colors are already used to indicate mutational effects.
-We have therefore created a new supplemental figure (Fig.~S\comment{?}) that shows these data---as the reviewer suspected, this shows that head mutations are much more common.
-\comment{{\bf John:} Can you make this figure? I'm actually not 100\% sure of the point, as I don't think the layout in Fig.~4 is the best way to show head versus stalk. 
-But since it's what the reviewer suggested, let's just do it and make it a supplement.
-This would be just like the current Fig.~4 except rather than coloring by mutational effect, we color by head versus stalk.
-Then I think we should also make a second panel that shows the number of head and stalk mutations that reach an indicated frequency over the whole timespan, as this is really a better way to show the point.
-So the x-axis would be mutation frequency, we would count mutations that reach each frequency, bin them and show this for head and stalk.
-This would definitely show that there are more head mutations overall, and would probably show that more head mutations reach high frequency.
-You could argue that it would be more informative to do this for epitope versus non-epitope, but since the reviewer asked for head versus stalk and we won't really discuss this much in the text anyway, let's just do head versus stalk...}
+We have therefore created a new supplemental figure (Fig.~S7) that shows these data---as the reviewer suspected, this shows that head mutations are much more common.
 
@@
 
-\comment{This would be interesting to do, but is a whole other study.
-This is the one point that I think we are just going to have to argue is beyond the scope of the manuscript.
+\response{We agree that this would be a very interesting experiment to perform.
+However, we believe such a study, which would require rigorous examination of escape from multiple head- and stalk-specific antibodies, is beyond the scope of this manuscript.
 }
@@
 
-\comment{We can definitely do this pretty easily, perhaps adding Sarah Hilton as an author to help.
-Not sure where we would fit the data given that the paper is already too long, but I guess we could cram it in the supplement.
+\response{This is a good suggestion and we have now added the results of this analysis in Figure S4.
 }
+
+\comment{Also need to add mention of this figure in the manuscript text somewhere...}
 
@@
 
-\comment{This is a possibility, although I highly doubt that it is true.
-However, we can easily add a sentence acknowledging the possibility.
-}
+\response{We now have acknowledged this possibility in the Discussion in the sentence "We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains.}
 
@@
 Ultimately, it is clear from the discussion that the authors see this study as an intermediate step, suggesting that their work is "an improvement over no information at all." While something is indeed better than nothing, in my opinion, this paper represents only an incremental step forward in the field of influenza evolutionary forecasting due to the aforementioned limitations, especially since this experiment and subsequent analysis could be performed using more stringent selection methods during viral passaging like treatment with human sera for data that is presumably more faithful to nature.
+
+\response{This is a fair assessment by the reviewer.
+However, this is the first paper to show that data from deep mutational scanning can actually provide some insight into evolutionary outcomes in nature, and it is possible that this can provide a useful framework for applying deep mutational scanning studies in the context of evolution in nature...
+}
 
@@
 
-\comment{Not sure if the difference between $\rho = 0.17$ and $\rho = 0.14$ is significant. 
-We could figure out how to calculate this, but first let's see what it looks like for the non-lab-passaged sequences requested by Reviewer 1.}
+\response{We now include an analysis of mutational effects and maximum frequency for mutations of unpassaged isolates (see Figures 5, 6, and S6) from the post-Perth partition of the phylogenetic tree to limit any potential biases that our experimental measurements may have towards mutations from cell- or egg-passaged isolates.
+The results show that this correlation is stronger when considering mutations from unpassaged isolates in a post-hoc analysis, demonstrating the potential utility of our experimental measurements in a prospective analysis, specifically when passaged isolate sequences are excluded.
+}
 
@@
 
-\comment{We can clarify these easily.}
+\comment{Fixed the figure. Still need to explain frequency}
 

### 2018-06-20 paper/manuscript.tex
Update supp figures & paper text, add to reviewer responses

@@
 Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
-To further increase viral titers, we used MDCK-SIAT1 cells that we had engineered to constitutively express the TMPRSS2 protease, which cleaves the HA precursor to activate it for membrane fusion~\cite{bottcher2006proteolytic, bottcher2010cleavage}.
+To further increase viral titers, we used MDCK-SIAT1 (Madin-Darby canine kidney cells with enhanced expression of sialic acids) cells that we had engineered to constitutively express the TMPRSS2 (Transmembrane Protease, Serine 2) protease, which cleaves the HA precursor to activate it for membrane fusion~\cite{bottcher2006proteolytic, bottcher2010cleavage}.
 
-After generating the mutant virus libraries, we passaged them at low MOI in cell culture to create a genotype-phenotype link and select for functional HA variants (Figure~\ref{fig:dms_overview}A).
+After generating the mutant virus libraries, we passaged them at low multiplicity of infection (MOI) in cell culture to create a genotype-phenotype link and select for functional HA variants (Figure~\ref{fig:dms_overview}A).
 All experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B).
@@
 For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
-Helix A forms part of the epitope of many broadly neutralizing anti-stalk antibodies~\cite{mallajosyula2014influenza,laursen2013broadly,chai2016two}, and interestingly, there are more reports of readily selecting escape mutants from such anti-stalk antibodies in H3~\cite{ekiert2011highly, friesen2014common, chai2016two, yamayoshi2017human} than in H1~\cite{okuno1993common,doud2018quantifying,anderson2017natural} HAs.
+Helix A forms part of the epitope of many broadly neutralizing anti-stalk antibodies~\cite{mallajosyula2014influenza,laursen2013broadly,chai2016two}.
 
@@
 Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
+The Perth/2009 strain is labeled with a star, and nodes in the clade containing the Perth/2009 strain were excluded from our analyses.
 The bottom panel shows the frequency trajectory of each mutation, with trajectories colored according to the mutational effects as in the top panel.
@@
 We initially created a virus with the HA and NA from Perth/2009 and internal genes from WSN/1933, and passaged it in cell culture to test its genetic stability.
-To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM supplemented with 10\% heat-inactivated FBS, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amounts of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
+To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM, or Dulbecco's Modified Eagle Medium, supplemented with 10\% heat-inactivated FBS, or Fetal Bovine Serum, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amounts of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
 The next day we changed the media to influenza growth media (IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL (no trypsin was added since there was TMPRSS2), and then collected the viral supernatant at 72 hours post-transfection.
@@
 We obtained $>$6 million transformants for each replicate.
-We scraped the plates, expanded the cultures in liquid LB + ampicillin at 37$^{\circ}$C for 3 h with shaking, and then maxiprepped.
+We scraped the plates, expanded the cultures in liquid LB (Luria-Bertani Broth) + ampicillin at 37$^{\circ}$C for 3 h with shaking, and then maxiprepped.
 We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (Figure~\ref{suppfig:SangerSeq}).
@@
 We replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L in IGM.
-Three hours post-infection, we replaced the inoculum with fresh IGM.
+Three hours post-infection, we replaced the inoculum with fresh IGM for replicates 1, 2, and 3-2.
 We collected viral supernatant for sequencing 48 hours post-infection.
@@
 To determine the effect of virus passaging on the relationship between mutational effects and maximum frequencies in natural populations, we repeated the above analyses using only viruses that were sequenced directly without passaging.
-Routine direct sequencing did not begin until the early 2000s \cite{McWhite2016}.
+Routine direct sequencing did not begin until the early 2000s \cite{mcwhite2016sequence}.
 To construct a tree with a similar number of viruses as the original analysis, we sampled 30 viruses per month between January 1, 2000 and April 1, 2018, producing a tree with 2,374 unpassaged viruses with augur (commit: 6d9f708).
@@
 We used the multiple sequence alignment of the HA subtype sequences from \cite{doud2018quantifying} to identify sites that are absolutely conserved across all subtypes, or in the different clades described in Figure~\ref{fig:RMSD_heatmap}.
+
+\subsection*{Validation of individual point mutants}
+To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, C199(HA2)K, T24F, T40V, and S287A, we used site-directed mutagenesis to introduce the amino-acid mutation into the Perth/2009 HA bidirectional reverse genetics plasmid, and verified the sequence of two clones for each mutant by Sanger sequencing.
+We generated these individual mutant viruses carrying GFP in the PB1 segment using a protocol described in \cite{hooper2013mutant}, and the PB2, PA, NP, NA, M, and NS segments of the A/WSN/1933 (H1N1) strain.
+To rescue each mutant GFP-carrying virus in duplicate, we transfected a co-culture of $4 \times 10^5$ 293T-CMV-PB1 and $0.5 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells with the eight reverse genetics plasmid and the pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W plasmid.
+Each well received a transfection mixture of 100 $\mu$L DMEM, 3 $\mu$L BioT transfection reagent, and 250 ng of each plasmid.
+We changed the media in each well with 2 mL IGM eight hours post-transfection.
+At 53 hours post-transfection, transfection supernatants were harvested, clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes, aliquoted, and frozen at -80$^\circ$C.
+
+To titer the GFP-carrying viruses, we plated $1 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells per well in 12-well plates in IGM.
+Four hours after plating, we infected cells with dilutions of viral supernatant.
+At 16 hours post-infection, we selected wells with 1 to 10\% of cells that were GFP-positive and analyzed the fraction of GFP-positive cells by flow cytometry to calculate the titer of infectious particles per $\mu$L.
 
@@
 \begin{suppfigure}[H]
+\centerline{\includegraphics[width=0.9\textwidth]{figs/supp_diffprefs/diffprefs.pdf}}
+\caption{\label{suppfig:diffprefs_logoplot}
+{\bf The differential preferences of the Perth/2009 H3 HA compared to amino-acid frequencies from a sequence alignment of natural human H3.}
+The differential preferences estimated by comparing the Perth/2009 H3 HA preferences to amino-acid frequencies from a sequence alignment of natural human H3N2 HA sequences.
+For each site, the height of each letter above and below the line indicate how differentially preferred that amino acid is in our experiments compared to in nature.
+The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
+The sites are in H3 numbering.
+}
+\end{suppfigure}
+
+\begin{suppfigure}[H]
 \centerline{\includegraphics[width=0.9\textwidth]{figs/supp_H3N2_phylogeny/H3N2_phylogeny.pdf}}
@@
 
+\begin{suppfigure}[H]
+\centerline{\includegraphics[width=0.5\textwidth]{figs/supp_head_stalk_mutfreq/head_stalk_mutfreq.pdf}}
+\caption{\label{suppfig:head_stalk_mutfreq_supp}
+{\bf Frequency trajectories of head and stalk domain mutations.}
+(A) This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:frequency_trajectory}, but colors amino-acid mutations by whether they occur in the head (purple) or the stalk (blue) domain.
+(B) Histogram of mutation maximum frequencies by the number of nodes for head and stalk domain mutations.
+It is clear that mutations in the head domain are more numerous than those in the stalk.
+}
+\end{suppfigure}
+
+\begin{suppfigure}[H]
+\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
+\caption{\label{suppfig:point_mut_validation_supp}
+{\bf Viral titers of individual Perth/2009 H3 HA point mutants.}
+Influenza GFP viruses carrying wildtype or single point mutants of the Perth/2009 HA were rescued by reverse genetics in duplicate. 
+Viral transfection supernatant was harvested at 53 hours post-transfection and titered by flow cytometry in MDCK-SIAT1-TMPRSS2 cells.
+Note that the wildtype, C52A and C52C mutant virus titers shown in each panel represent the same duplicate measurements.
+The points mark each duplicate and the bar marks the mean.
+(A) The start codon mutant M(-16)K reaches titers approximately one log less than those of wildtype. 
+The C52A mutant fails to reach high titers, and the C52C synonymous variant grows to titers comparable to those of wildtype virus.
+(B) The transmembrane domain mutant C199(HA2)K grows to higher titers than the wildtype.
+(C) The T24F and T40V mutants, which remove glycosylation motifs, attain titers similar to or higher than wildtype, respectively. 
+The S287A mutant, which also removes a glycosylation motif, does not grow as well as wildtype virus.
+}
+\end{suppfigure}
+
 \begin{suppdata}[H]
@@
 Additional tabs give the across-replicate averaged and re-scaled amino-acid preferences in sequential numbering and in H3 numbering as shown in Figure~\ref{fig:logoplot}.
-There are also tabs that give the conversion from sequential to H3 numbering.
+There are also tabs that give the conversion from sequential to H3 numbering, and a list of the mutations in the 31 clones we Sanger sequenced to evaluate the mutation rate of the mutant plasmid libraries.
 Each tab can simply be exported to CSV for computational analyses.

### 2018-06-21 paper/reviewer_response_PNAS.tex
Finish drafting Reviewer Responses, update manuscript revisions

@@
 
-\response{Rescaling of the preferences is performed by fitting a stringency parameter that is a reflection of how strongly our experimentally determined preferences correspond to the gene's evolution in nature.
+\response{Re-scaling of the preferences is performed by fitting a stringency parameter that is a reflection of how strongly our experimentally determined preferences correspond to the gene's evolution in nature.
 This parameter is fit under an experimentally informed codon substitution model (ExpCM) using the phydms program (Hilton 2017 PeerJ).
@@
 Certainly other experimental parameter choices could have been used, but this is always the case for complex experiments.
-Our choices will reasonable, and the four lines of evidence listed at the beginning of this comment show that they yield reproducible and meaningful results.}
+Our choices are reasonable, and the four lines of evidence listed at the beginning of this comment show that they yield reproducible and meaningful results.}
 
@@
 2) The authors state on page 5 that mutations detected in cell culture as deleterious rarely reach high frequency in nature. Could you please give a few examples?
+
+\response{We individually validated viral growth of a C52A mutant, which removes a Cys critical in forming a disulfide bond with Cys-277 and maintaining HA structure.
+We confirmed that this mutant grows very poorly when rescued, and mutations at site 52 never arise at a detectable level.
+
+As another example, mutation N170H fails to reach $>$1\% frequency in nature, and according to our dataset, is a mutation towards a highly unpreferred amino acid. 
+}
 
@@
 Two of the clones had a mutation that removed the ATG start codon.
+One of these resulted in an M(-16)V mutation, while the other resulted in an M(-16)Q mutation which was linked to a Q80I mutation.
+According to our preferences, Val and Gln are not very preferred at this site, suggesting that these mutants would not grow well compared to wildtype.
 }
-
-\comment{Not sure how to address the growth phenotype question}.
 
@@
 It is difficult to also overlay the structural location (stalk versus head) on the current Fig.~4, since colors are already used to indicate mutational effects.
-We have therefore created a new supplemental figure (Fig.~S7) that shows these data---as the reviewer suspected, this shows that head mutations are much more common.
+We have therefore created a new supplemental figure (Fig.~S8) that shows these data---as the reviewer suspected, this shows that head mutations are much more common.
 
@@
 This makes sense, as G78D is one of the two sites that we specifically identified as being a lab adaptation (see Figure~S1).
-We now explicitly mention this fact in the text \comment{still need to actually add this mention}.
+We now explicitly mention this fact in the Discussion as follows:
+
+\begin{quote}
+\textsl{Indeed, a vast amount of work in virology has chronicled the ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature.
+As an example of this, although we identified G78D as a lab adaptation mutation and the wildtype Gly is not highly preferred at this site, this mutation never reaches above 20\% frequency in nature. }
+\end{quote}
 }
@@
 \response{This is a good suggestion and we have now added the results of this analysis in Figure S4.
+As the logo plot shows, although there are some differences, at most sites there is little difference between the amino acids that are preferred in our experiments to their frequencies in nature.
 }
-
-\comment{Also need to add mention of this figure in the manuscript text somewhere...}
 
@@
 
-\response{We now have acknowledged this possibility in the Discussion in the sentence "We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains.}
+\response{We now have acknowledged this possibility in the Discussion in the sentence ``We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains."}
 
@@
 \response{This is a fair assessment by the reviewer.
-However, this is the first paper to show that data from deep mutational scanning can actually provide some insight into evolutionary outcomes in nature, and it is possible that this can provide a useful framework for applying deep mutational scanning studies in the context of evolution in nature...
-}
-
-\comment{This is a fair point.
-However, I think we should argue that it's the first paper to show deep mutational scanning can actually provide at least some information that reflects outcomes in nature.
-This comment does make me worry that the reviewer just doesn't think the work is significant enough, but since the editor did ask for a revision you'd have to hope that if we answer all specific comments that would be enough.
+However, this is the first paper to show that data from deep mutational scanning can actually reflect evolutionary outcomes in nature, and provides some insight into the extent by which functional constraints on HA shape its evolution.
+Furthermore, it is possible that this work can provide a useful framework for applying deep mutational scanning studies in the context of evolution in nature.
 }
@@
 
-\response{We now include an analysis of mutational effects and maximum frequency for mutations of unpassaged isolates (see Figures 5, 6, and S6) from the post-Perth partition of the phylogenetic tree to limit any potential biases that our experimental measurements may have towards mutations from cell- or egg-passaged isolates.
+\response{We now include an analysis of mutational effects and maximum frequency for mutations of unpassaged isolates (see Figures 5, 6, and S7) from the post-Perth partition of the phylogenetic tree to limit any potential biases that our experimental measurements may have towards mutations from cell- or egg-passaged isolates.
 The results show that this correlation is stronger when considering mutations from unpassaged isolates in a post-hoc analysis, demonstrating the potential utility of our experimental measurements in a prospective analysis, specifically when passaged isolate sequences are excluded.
@@
 
-\comment{We can clarify these easily.}
+\response{We appreciate these suggestions from the reviewer and have made the following changes:
+\begin{itemize}
+\item We have now included a description of how we calculated mutational effects in the Results section.
+\item We agree that a cutoff of -10 to 0 is arbitrary and removed the sentence ``Mutations that reach high frequencies generally have neutral or beneficial effects according to our experimental measurements."
+\item We included in the Methods a description of how we calculated Shannon entropy.
+\item We now define AIC in the legend of Table 1.
+\item We have now included an explanation of dN/dS $\ll$1 for conventional substitution models, and now describe the utility of building an ExpCM, site avg. substitution model
+\end{itemize}
+}
 
@@
 
-\comment{Fixed the figure. Still need to explain frequency}
+\response{We have now labeled the Perth/2009 strain on the tree in Figure 4 and now mention in the caption that we have excluded closely related strains from the analysis...
+}
+
+\comment{Fixed the figure. Maybe John and Trevor can do a better job than me at explaining the frequency denominator. Is it the population observed at a given pivot?}
 

### 2018-06-21 paper/manuscript.tex
Finish drafting Reviewer Responses, update manuscript revisions

@@
 \addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites (site avg.), and M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
-Models are compared by AIC~\cite{posada2004model} computed from the log likelihood (LnL) and number of model parameters.
+Models are compared by Akaike information criterion (AIC)~\cite{posada2004model} computed from the log likelihood (LnL) and number of model parameters.
 The $\omega$ parameter is dN/dS for the Goldman-Yang models, and the relative dN/dS after accounting for the measurements for the ExpCM.
@@
 This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard substitution models~\cite{bloom2017identification,hilton2017phydms}.
-Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
+Table~\ref{tab:phydms} shows that the phylogenetic fit of the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
 This result indicates that our experiments authentically capture some of the constraints on HA evolution.
-The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
+A substitution model in which the amino-acid preferences were averaged across all sites (ExpCM, site avg.) performs no better than conventional substitution models, demonstrating the importance of site-specificity for the preferences.
+The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}), illustrating that these models need to account for constraints on natural HA evolution.
 However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for in the deep mutational scanning.
@@
 These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.
+Furthermore, Figure~\ref{suppfig:diffprefs_logoplot} shows that when we compare these preferences to amino-acid frequencies from natural human H3N2 HA sequences, at most sites, there is little difference between the amino acids that are preferred in our experiments to their frequencies in nature.
 
@@
 After annotating mutations and their frequencies on the phylogeny in this way (Figure~\ref{fig:frequency_trajectory}), it is visually obvious that there are relatively few circulating mutations that we measure to be strongly deleterious---and that such deleterious mutations rarely reach high frequency when they do occur.
-Mutations that reach high frequencies generally have neutral or beneficial effects according to our experimental measurements.
 
 We next sought to quantify the correlation between a mutation's experimentally measured effect and the maximum frequency it attained during natural evolution.
+To calculate a given mutation's effect, we simply took the logarithm of the preference for the new amino-acid identity divided by the preference for the original amino-acid identity.
 To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
@@
 In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).
+
 To restrict potential biases of our experimental measurements towards mutations from cell- or egg-passaged isolates, we evaluated mutations from unpassaged isolates.
@@
 Indeed, a vast amount of work in virology has chronicled the ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}.
+As an example of this, although we identified G78D as a lab adaptation mutation and the wildtype Gly is not highly preferred at this site, this mutation never reaches above 20\% frequency in nature.
 Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
@@
 
-\subsection*{Quantification of relative solvent accessibility}
+\subsection*{Quantification of Shannon entropy and relative solvent accessibility}
+We calculated Shannon entropy, $h_r$ for site $r$ as $h_r = - \sum_{x} \pi_{r,x} \log(\pi_{r,x})$, where $\pi_{r,x}$ is the preference for amino acid $x$ at site $r$.
+
 We quantified the absolute solvent accessibility of each site of the H3 HA (PDB 4O5N;~\cite{lee2014receptor}) or the H1 HA (PDB 1RVX;~\cite{gamblin2004structure}) structure using DSSP~\cite{kabsch1983}.

### 2018-06-23 paper/reviewer_response_PNAS.tex
start codon revisions

@@
 								% TeX will automatically convert eps --> pdf in pdflatex
+
 \usepackage{amssymb}
@@
 We thank the reviewers for the careful reading and suggestions about the manuscript.
-In the revised version, we have performed additional experiments and analyses that should address all of the reviewer concerns.
+In the revised version, we have performed additional experiments and analyses that address all of the reviewer questions and concerns.
 
@@
 
+\response{We are sorry that the reviewer found some parts of the manuscript confusing.
+We have revised the parts of the text that the reviewer highlighted, performed additional validation experiments, and clarified the experimental setup.
+These modifications should fully address the reviewer's concerns.}
+
 \subsubsection*{Major points}
@@
 
-\response{We have now rescued by reverse genetics influenza GFP viruses carrying the M(-16)K mutation in the start codon, and titered by flow cytometry. 
-By 53 hours post-transfection when we harvested transfection supernatant, the M(-16)K mutant reached a titer of approximately one log less than that of wildtype virus.
-Although we may have expected the M(-16)K mutant to grow comparably to wildtype given that Met and Lys are both preferred at the start codon, we would like to emphasize that (1) we grew influenza GFP viruses which may not behave identically to live influenza viruses and (2) we did not capture the full viral growth kinetics of this mutant since we titered at a relatively early timepoint after transfection.
-In comparison, the C52A variant, which mutates the highly conserved Cys at site 52, grows very poorly, while the synonymous C52C mutant grows to titers comparable to those of wildtype.
-These results are now shown in Figure S9, and we have updated the Methods section describing how we validated these point mutants.
+\response{
+We have now used reverse genetics to generate viruses carrying a mutation of the initiating methionine to lysine.
+These start-codon mutant viruses grow to appreciable titers, suggesting that there is indeed some flexibility in the site of translation initiation in this HA.
+These data are in Figure \comment{S5} of the revised manuscript.
+As we note in the manuscript, this finding is concordant with prior work on suggesting additional translation initiation sites sites in H2 HAs.
+Because translation initiation of HA is not the focus of our manuscript, we have not performed further investigation of the exact mechanism or identity of the alternate start sites.
+However, the data we have added clearly show that the canonical ATG is not strictly necessary for viral replication.
 }
@@
 
-\response{Figure~S3 shows data for the H1 HA as measured previously by Doud and Bloom (2016). 
-That study used mutant libraries in which all codons \emph{except} the one for the N-terminal methionine were mutagenized.
+\response{
+The reviewer is correct. 
+Figure~S3 shows data for a H1 HA as measured previously by Doud and Bloom (2016), and these prior data are simply presented here in H3 numbering for comparative purposes (the original Doud and Bloom paper used a different HA numbering scheme that makes site-to-site comparisons with the current study difficult).
+The Doud and Bloom study used mutant libraries in which all codons in HA \emph{except} for the one encoding the N-terminal methionine were mutagenized.
 Therefore, that study did not estimate the effects of mutations to the first codon, which is why it is not shown in Figure~S3.

### 2018-06-23 paper/manuscript.tex
Clarified the denominator for mutation frequency calculations in methods

@@
 Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
-To further increase viral titers, we used MDCK-SIAT1 (Madin-Darby canine kidney cells with enhanced expression of sialic acids) cells that we had engineered to constitutively express the TMPRSS2 (Transmembrane Protease, Serine 2) protease, which cleaves the HA precursor to activate it for membrane fusion~\cite{bottcher2006proteolytic, bottcher2010cleavage}.
+To further increase viral titers, we used MDCK-SIAT1 cells (Madin-Darby canine kidney cells overexpressing 2,6-sialyltransferase)~\cite{matrosovich2003overexpression} that we engineered to constitutively express TMPRSS2 (Transmembrane Protease, Serine 2), which cleaves the HA precursor to activate it for membrane fusion~\cite{bottcher2006proteolytic, bottcher2010cleavage}.
 
@@
 This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard substitution models~\cite{bloom2017identification,hilton2017phydms}.
-Table~\ref{tab:phydms} shows that the phylogenetic fit of the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
+Table~\ref{tab:phydms} shows that an ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
 This result indicates that our experiments authentically capture some of the constraints on HA evolution.
-A substitution model in which the amino-acid preferences were averaged across all sites (ExpCM, site avg.) performs no better than conventional substitution models, demonstrating the importance of site-specificity for the preferences.
-The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}), illustrating that these models need to account for constraints on natural HA evolution.
+A substitution model in which the amino-acid preferences were averaged across all sites (ExpCM, site avg.) performs no better than conventional substitution models, demonstrating that the reason that our measurements are informative is because they capture site-specific constraints.
+The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}), since the only way that these models can account for the constraints on HA evolution is via the heuristic use of the $\omega$ parameter to depress the rates of all nonysynonymous substitutions.
 However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for in the deep mutational scanning.
@@
 Furthermore, Figure~\ref{suppfig:diffprefs_logoplot} shows that when we compare these preferences to amino-acid frequencies from natural human H3N2 HA sequences, at most sites, there is little difference between the amino acids that are preferred in our experiments to their frequencies in nature.
+\comment{This needs to be improved / changed}
 
@@
 A positively charged amino acid at site 329 is important for cleavage of the HA0 precursor into the mature form~\cite{stech2005new}, and this site strongly prefers arginine.
-However, a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine.
+
+However, there are also some notable differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences.
+Most surprisingly, the start codon at position does not show a particularly strong preference for methionine.
+We validated that viruses carrying a mutation at this site from methionine to lysine do in fact reach appreciable titers (Figure~\ref{suppfig:startcodon_validation}), perhaps because of alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
+\comment{I AM HERE}
+a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine (Figure~\ref{fig:logoplot}).
 This codon is part of the signal peptide and is cleaved from the mature HA protein.
@@
 Herein, mutation frequency dynamics are modeled according to a Brownian motion diffusion process discretized to one-month intervals.
+The number of viruses sampled in each interval determines the denominator of the mutation frequency calculations.
 Relative to a simple Brownian motion, the expectation includes an ``inertia'' term $\inertia$ that adds velocity to the diffusion and the variance includes a term $x(1-x)$ to scale variance according to frequency following a Wright-Fisher population genetic process.
@@
 \caption{\label{suppfig:WSNprefs_logoplot}
-{\bf The site-specific amino-acid preferences of the Perth/2009 H3 HA and the WSN/1933 H1 HA.}
-(A) The un-scaled amino-acid preferences of the Perth/2009 H3 HA.
-The overlays show the same information as in Figure~\ref{fig:logoplot} (domain, epitope site, and wildtype amino acid).
-(B) The amino-acid preferences of the WSN/1933 H1 HA from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
+{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.}
+\comment{Get rid of unscaled Perth/2009 in this figure. It doesn't make sense to show them here. I've removed from legend; remove from the actual figure}
+The amino-acid preferences of the WSN/1933 H1 HA from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
 The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
@@
 Therefore, no data is shown for the first codon in the gene.
-The sites are in H3 numbering for both logo plots.
+The sites are in H3 numbering.
 }
@@
 The sites are in H3 numbering.
+}
+\end{suppfigure}
+
+\begin{suppfigure}[H]
+\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
+\caption{
+\comment{Just show the first panel (A) here, the others can go in later figures}
+{\bf Validation that the Perth/2009 HA is somewhat tolerant to mutation of the canonical start codon.}
+We created variants of the Perth/2009 HA in which the canonical start codon was mutated (amino-acid mutation M(-16)K, codon mutation \comment{what was it}), a conserved cysteine was mutated to alanine (C52A, codon mutation \comment{what was it?}), and the same cysteine was synonymously mutated (C52C, codon mutation \comment{what was it}).
+We selected these mutations because our deep mutational scanning results in Figure~\ref{fig:logoplot} surprisingly suggests that the start codon (position -16) is fairly tolerant of mutations, but that site 52 is highly intolerant of mutation to anything other than cysteine.
+The synonymous C52C mutation is a negative control mutation that is not expected to have any effect.
+We then used reverse genetics to generate viruses carrying the wildtype HA or each point mutant, with GFP packaged in the PB1 segment to enable easy titering by flow cytometry~\cite{bloom2010permissive,hooper2013mutant}.
+Each variant was generated in duplicate.
+The plots show the viral titer in the supernatant at 53 hours post-transfection.
+As expected, the C52A virus is essentially non-viable, while the C52C mutation is roughly equivalent to wildtype.
+the M(-16)K mutation is only moderately attenuated, validating that the canonical start codon is not completely essential for viral growth.
+}
+\label{suppfig:startcodon_validation}
+\end{suppfigure}
+
+\begin{suppfigure}[H]
+\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
+\caption{\label{XXX}
+{\bf Viral titers of individual Perth/2009 H3 HA point mutants.}
+Influenza GFP viruses carrying wildtype or single point mutants of the Perth/2009 HA were rescued by reverse genetics in duplicate. 
+Viral transfection supernatant was harvested at 53 hours post-transfection and titered by flow cytometry in MDCK-SIAT1-TMPRSS2 cells.
+Note that the wildtype, C52A and C52C mutant virus titers shown in each panel represent the same duplicate measurements.
+The points mark each duplicate and the bar marks the mean.
+(A) The start codon mutant M(-16)K reaches titers approximately one log less than those of wildtype. 
+The C52A mutant fails to reach high titers, and the C52C synonymous variant grows to titers comparable to those of wildtype virus.
+(B) The transmembrane domain mutant C199(HA2)K grows to higher titers than the wildtype.
+(C) The T24F and T40V mutants, which remove glycosylation motifs, attain titers similar to or higher than wildtype, respectively. 
+The S287A mutant, which also removes a glycosylation motif, does not grow as well as wildtype virus.
 }
@@
 
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
-\caption{\label{suppfig:point_mut_validation_supp}
-{\bf Viral titers of individual Perth/2009 H3 HA point mutants.}
-Influenza GFP viruses carrying wildtype or single point mutants of the Perth/2009 HA were rescued by reverse genetics in duplicate. 
-Viral transfection supernatant was harvested at 53 hours post-transfection and titered by flow cytometry in MDCK-SIAT1-TMPRSS2 cells.
-Note that the wildtype, C52A and C52C mutant virus titers shown in each panel represent the same duplicate measurements.
-The points mark each duplicate and the bar marks the mean.
-(A) The start codon mutant M(-16)K reaches titers approximately one log less than those of wildtype. 
-The C52A mutant fails to reach high titers, and the C52C synonymous variant grows to titers comparable to those of wildtype virus.
-(B) The transmembrane domain mutant C199(HA2)K grows to higher titers than the wildtype.
-(C) The T24F and T40V mutants, which remove glycosylation motifs, attain titers similar to or higher than wildtype, respectively. 
-The S287A mutant, which also removes a glycosylation motif, does not grow as well as wildtype virus.
-}
-\end{suppfigure}
-
 \begin{suppdata}[H]

### 2018-06-23 paper/reviewer_response_PNAS.tex
responded to points 2-3 of reviewer 1

@@
 These start-codon mutant viruses grow to appreciable titers, suggesting that there is indeed some flexibility in the site of translation initiation in this HA.
-These data are in Figure \comment{S5} of the revised manuscript.
+These data are in Figure S5 of the revised manuscript.
 As we note in the manuscript, this finding is concordant with prior work on suggesting additional translation initiation sites sites in H2 HAs.
@@
 
-\response{This is a valid point, and we have also individually evaluated the growth of three glycosylation site mutants.
-Two of the mutants, T24F and T40V, remove the aforementioned highly conserved but mutationally tolerant NGT and NAT glycosylation sites, respectively.
-The third mutant, S287A, removes a constrained glycosylation site.
-In Figure S9, we show that the T24F and T40V mutants grow to titers comparable to or higher than that of wildtype. 
-However, the titer of the S287A is approximately 2.8-fold lower than that of wildtype, suggesting that the glycan at sites 285-287 is important for viral growth in cell culture.
+\response{This is a valid point, and we have now also individually evaluated the reverse-genetics titers of three glycosylation site mutants.
+Two of the mutants, T24F and T40V, remove the aforementioned highly conserved but mutationally tolerant NGT and NAT glycosylation sites near the start of HA1.
+The third mutation, S287A, removes a glycosylation site that is highly variable in nature but that our experiments suggest is under functional constraint.
+The new Figure S6 shows that the T24F and T40V mutants grow to titers comparable to wildtype, supporting the conclusion that they are expendable at least in cell culture. 
+However, the titer of the S287A is modestly lower than wildtype, supporting the conclusion that this glycan contributes to viral growth in cell culture.
 
-Furthermore, constraint in our experiments and conservation in nature are not expected to be entirely the same.
-Conservation in nature results from the convolution of constraint and diversifying selection from immunity, whereas our experiments only look at constraint.
-\comment{As evidence to explain this point, we can point to studies showing that the head of HA is much more conserved for swine influenza viruses where there is less pressure from immunity.}
+One possible explanation is simply that cell culture does not fully capture the constraints in nature.
+But a more interesting possibility is that conservation in nature is not expected to strictly parallel mutational tolerance.
+The tendency for a site to change in nature depends on how much selection there is for the mutations.
+For instance, Wei et al (\textit{Science Translational Medicine}, 2:24ra21, 2010) have pointed out that glycosylation motifs evolved rapidly in the globular head of HA from human H1 influenza in the period from 1918 to 2008 (seasonal H1N1), but changed very little in the same period in the swine H1 lineages descended from the 1918 virus.
+The likely reason is differential strengths of selection: human influenza is under strong pressure from accumulated immune memory, whereas swine influenza is under less immune pressure because pigs are short lived and so are often naive when infected with influenza.
+This fact highlights that even mutationally tolerant sites will only change rapidly if there is a benefit to doing so, while even somewhat constrained sites can change if there is sufficient pressure.
+A strength of our experiments is that we have controlled the selection pressure (selection only for viral growth), and so can measure mutational tolerance in the absence of the confounding effects of immunity.
+
+In addition to the data in Figure S6, we have added the following text:
+
+\begin{quote}
+\textsl{
+The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function on nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.}
+\end{quote}
 }
-
-\comment{Alanine is not the least preferred amino acid at site 287, while it is the least preferred at site 52.}
 
@@
 
-\response{We validated growth of a transmembrane domain mutant, C199(HA2)K, and show that this mutant attains titers approximately 1.6-fold higher than that of wildtype.
-In addition, various studies have found that chimeric HA's with transmembrane domains from completely different proteins can still maintain HA cell surface expression and fusion activity (Melikyan et al 1999 Mol Biol Cell, Schroth-Diez et al 1998 JVI, Dong et al 1992 JVI, Roth et al 1986 JCB), and that the length of the transmembrane domain seems to play a large role in determining HA fusion activity (Armstrong et al 2000 JCB).
+\response{We have validated that one of these mutations in the transmembrane domain, C199(HA2)K, reaches reverse-genetics titers comparable to wildtype.
+These data are in the new Figure S7.
+In addition, various other studies have found that chimeric HAs with transmembrane domains from completely different proteins can still maintain HA cell surface expression and fusion activity (Melikyan et al 1999 Mol Biol Cell, Schroth-Diez et al 1998 JVI, Dong et al 1992 JVI, Roth et al 1986 JCB).
+
+We posit that the disparity between the conservation of the transmembrane domain in nature and its apparent mutational tolerance in our experiments could arise at least in part from a lack of immune pressure for mutations in nature, as discussed in the response to the previous point.
 }

### 2018-06-23 paper/manuscript.tex
responded to points 2-3 of reviewer 1

@@
 However, there are also some notable differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences.
-Most surprisingly, the start codon at position does not show a particularly strong preference for methionine.
-We validated that viruses carrying a mutation at this site from methionine to lysine do in fact reach appreciable titers (Figure~\ref{suppfig:startcodon_validation}), perhaps because of alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
-\comment{I AM HERE}
-a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine (Figure~\ref{fig:logoplot}).
-This codon is part of the signal peptide and is cleaved from the mature HA protein.
-One possible reason that our experiments do not show a strong preference for methionine at this site could be alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
+Most surprisingly, the start codon does not show a particularly strong preference for methionine (Figure~\ref{fig:logoplot}).
+We validated that a virus carrying a mutation at this site from methionine to lysine does in fact reach appreciable titers (Figure~\ref{suppfig:startcodon_validation}), perhaps because of alternative translation-initiation at a downstream or upstream start site as has been described for other HAs~\cite{girard2011upstream}.
+Our measurements also suggest mutational tolerance at some other sites that are relatively conserved among natural HAs, such as the N-linked glycosylation motifs near the beginning of HA1 and the transmembrane domain (Figure~\ref{fig:logoplot}).
+We validated that viruses with mutations to the glycosylation motifs at sites 22 or 38, or a site in the transmembrane domain, do in fact grow to high titers (Figures~\ref{suppfig:glycan_validation} and \ref{suppfig:TM_validation}, respectively).
+The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function on nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.
 
@@
 \caption{
-\comment{Just show the first panel (A) here, the others can go in later figures}
+\comment{Just show the first panel (A) here}
 {\bf Validation that the Perth/2009 HA is somewhat tolerant to mutation of the canonical start codon.}
@@
 We then used reverse genetics to generate viruses carrying the wildtype HA or each point mutant, with GFP packaged in the PB1 segment to enable easy titering by flow cytometry~\cite{bloom2010permissive,hooper2013mutant}.
-Each variant was generated in duplicate.
-The plots show the viral titer in the supernatant at 53 hours post-transfection.
+Each variant was generated in duplicate, and the plots show the viral titer in the supernatant at 53 hours post-transfection.
 As expected, the C52A virus is essentially non-viable, while the C52C mutation is roughly equivalent to wildtype.
@@
 \centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
-\caption{\label{XXX}
-{\bf Viral titers of individual Perth/2009 H3 HA point mutants.}
-Influenza GFP viruses carrying wildtype or single point mutants of the Perth/2009 HA were rescued by reverse genetics in duplicate. 
-Viral transfection supernatant was harvested at 53 hours post-transfection and titered by flow cytometry in MDCK-SIAT1-TMPRSS2 cells.
-Note that the wildtype, C52A and C52C mutant virus titers shown in each panel represent the same duplicate measurements.
-The points mark each duplicate and the bar marks the mean.
-(A) The start codon mutant M(-16)K reaches titers approximately one log less than those of wildtype. 
-The C52A mutant fails to reach high titers, and the C52C synonymous variant grows to titers comparable to those of wildtype virus.
-(B) The transmembrane domain mutant C199(HA2)K grows to higher titers than the wildtype.
-(C) The T24F and T40V mutants, which remove glycosylation motifs, attain titers similar to or higher than wildtype, respectively. 
-The S287A mutant, which also removes a glycosylation motif, does not grow as well as wildtype virus.
+\caption{\label{suppfig:glycan_validation}
+\comment{Just show the third panel (C) here, removing the cysteine mutants and adjusting the y-axis to a more appropriate for just these mutants.}
+{\bf Validation of viral variants with mutations at N-linked glycosylation motifs.}
+We created variants of the Perth/2009 HA with mutations that eliminated N-linked glycosylation motifs (Asn-Xaa-Ser/Thr) at asparagine residues 22, 38, and 285 (these are mutations T24F, T40V, and S287A, respectively).
+We then used reverse genetics to generate viruses carrying the wildtype HA or each of these mutants.
+Each variant was generated in duplicate, and the plots show the viral titer in the supernatant at 53 hours post-transfection.
+The viruses with mutations at the motifs at residues 22 and 38 reached titers at least as high as wildtype, whereas the virus with a mutation to the motif at residue 285 was modestly attenuated.
+}
+\end{suppfigure}
+
+\begin{suppfigure}[H]
+\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
+\caption{\label{suppfig:TM_validation}
+\comment{Just show TM mutant here, similar to previous figure}
+{\bf Validation of the mutational tolerance of a site in the transmembrane domain.}
+We created a variant of the Perth/2009 HA with a transmembrane domain mutation, C199(HA2)K, at a site that our deep mutational scanning suggests should be highly mutationally tolerant (Figure~\ref{fig:logoplot}).
+We then used reverse genetics to generate viruses carrying the wildtype HA or this mutants
+Each variant was generated in duplicate, and the plots show the viral titer in the supernatant at 53 hours post-transfection.
+The virus the mutation reached titers comparable to wildtype.
 }

### 2018-06-23 paper/reviewer_response_PNAS.tex
reviewer 1 point 4, lab passaging

@@
 
-\response{We thank the reviewer for this astute suggestion and now include an analysis of the relationship between mutational effect and maximum frequency for mutations from unpassaged isolates.
-Due to the sparseness of unpassaged isolate sequences previous to the last decade, we analyzed unpassaged isolates from the post-Perth/2009 partition of the H3N2 HA phylogenetic tree.
-Figures 5, 6, and S7 now include panels showing the results for these unpassaged isolate mutations.
-The results in Figure 5 demonstrate that the correlation between mutational effect as determined by our experiments and maximum mutation frequency is stronger for unpassaged isolates than when we include all isolates ($\rho = 0.24$).
-We do not see such a correlation when we use the WSN/1933 H1 HA preferences to estimate the mutational effects for unpassaged isolate mutations.
+\response{
+This is an excellent question.
+We now a separate analysis of the relationship between our measured mutational effects and maximum frequency in nature for mutations from unpassaged isolates only.
+Because sequences from unpassaged isolates have only become common relatively recently, we could perform this analysis only for the post-Perth/2009 partition of the H3N2 HA phylogenetic tree.
+The new Figure 5 shows that the correlation between our measurements and natural evolution is actually \emph{stronger} when we look only at unpassaged isolates.
+This makes sense---lab passage mutations will not do well in nature, but will be scored favorably in our experiments.
+Therefore, it makes sense that our approach works \emph{even better} when the analysis is restricted to unpassaged isolates.
+We thank the reviewer for this suggestion, which has strengthened our work.
 }
@@
 
-\response{We have now removed this sentence from the paper.}
+\response{We have removed this sentence from the paper.}
 

### 2018-06-23 paper/manuscript.tex
reviewer 1 point 4, lab passaging

@@
 We next sought to quantify the correlation between a mutation's experimentally measured effect and the maximum frequency it attained during natural evolution.
-To calculate a given mutation's effect, we simply took the logarithm of the preference for the new amino-acid identity divided by the preference for the original amino-acid identity.
+To calculate a given mutation's effect, we simply took the logarithm of the ratio of the preferences for the mutant and wildtype amino acids at that site.
 To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
@@
 In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).
+The similar effect sizes for both the pre- and post-Perth partitions shows that our experimental measurements can help explain the evolutionary fates of mutations in strains that post-date the experimentally studied strain, as well as to retrospectively analyze mutations that precede the experimental strain.
 
-To restrict potential biases of our experimental measurements towards mutations from cell- or egg-passaged isolates, we evaluated mutations from unpassaged isolates.
-Due to the limited availability of primary isolate sequences until the last decade, we analyzed mutations of unpassaged isolates from the post-Perth partition of the phylogenetic tree~\cite{mcwhite2016sequence}.
-The correlation between mutational effect and maximum frequency was stronger for mutations from unpassaged post-Perth isolates ($\rho = 0.24$).
-The similar effect sizes for both partitions shows that our experimental measurements can help explain the evolutionary fates of mutations in strains that post-date the experimentally studied strain, as well as to retrospectively analyze mutations that precede the experimental strain.
+Many of the HAs in sequence databases were obtained from virus isolates that were passaged in cell-culture or eggs, which can lead to the acquisition of lab-adaptation mutations that can confound evolutionary analyses~\cite{mcwhite2016sequence}.
+To check that our results were not confounded by such lab-adaptation mutations, we repeated our analysis using only HA sequences that were annotated as being derived from viruses that had not been passaged in the lab.
+Because sequencing of unpassaged primary isolates has only become commonplace relatively recently, we were only able to perform this analysis for the post-Perth partition of the phylogenetic tree.
+Figure~\ref{fig:muteffect_maxfreq}A shows that the correlation between our measured mutational effects and the maximum frequency was even stronger for mutations from unpassaged viral isolates ($\rho = 0.24$).
 

### 2018-06-23 paper/reviewer_response_PNAS.tex
comment 5 reviewer 1

@@
 \usepackage{amssymb}
+\usepackage{hyperref}
 
@@
 
-\response{Re-scaling of the preferences is performed by fitting a stringency parameter that is a reflection of how strongly our experimentally determined preferences correspond to the gene's evolution in nature.
-This parameter is fit under an experimentally informed codon substitution model (ExpCM) using the phydms program (Hilton 2017 PeerJ).
-A stringency parameter greater than 1 indicates that natural selection favors the same amino acids as the experiments but with greater stringency, while a value less than 1 indicates that natural selection does not favor the amino acids with as much stringency as the experiments.
-Because selection in the lab often does not fully reflect the stringency of selection in nature and selection across experiments may not be entirely consistent, we re-scale the preferences by this stringency parameter.
-Importantly, re-scaling does not affect any of the evolutionary analyses performed in this paper since it is a constant that multiplies all mutational effects, simply corresponding a change in units on the x-axis of Figure~5.
-We now include a supplementary panel A to Figure S3 showing the un-scaled Perth/2009 H3 HA preferences.
+\response{
+The re-scaling of amino-acid preferences by a stringency parameter is a well-defined mathematical procedure that is explained in detail in Hilton et al (\textit{PeerJ}, 5:e3657, 2017), which is cited in the manuscript.
+This re-scaling procedure has now been used fairly widely in the literature. 
+Importantly, re-scaling does \emph{not} affect any of the evolutionary analyses performed in this paper since it just corresponds to a constant that multiplies all mutational effects (e.g., a change in units on the x-axis of Figure~5).
+Specifically, if the re-scaling parameter is $\beta$, then the mutational effects calculated using Equation~1 in the manuscript are just multiplied by $\beta$ after re-scaling.
+Therefore, for analyses of mutational effects, the re-scaling simply corresponds to a change in units (e.g., comparable to showing an antibody titration in ng/mL versus $\mu$M), and therefore has no effect on the conclusions.
+We have added the following text in the Methods explaining this fact:
+
+\begin{quote}
+\textsl{The amino-acid preferences were re-scaled by the stringency parameter using the approach described in Hilton et al (2017).
+Note that the re-scaling simply puts the amino-acid preferences on a useful scale for visualization in logo plots, and has no effect on any of the quantitative conclusions relating the deep mutational scanning to natural evolution.
+The reason is that all of these conclusions use the effects of mutations as calculated using Equation~1, and the re-scaling simply acts as a constant multiplier on all mutational effects (e.g., a scale factor) when re-scaled preferences are converted to mutational effects.}
+\end{quote}
+
+Due to space constraints and the size of the logo plots, we cannot fit another main figure with the preferences prior to re-scaling, but such an image is now included in the Jupyter notebook in the manuscript's GitHub repository at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript}.
+The unscaled amino-acid preferences for each replicate are also provided in numerical form in Dataset S3.
 }
@@
 \response{We have removed this sentence from the paper.}
-
-\comment{We looked at this pretty carefully, and there are definitely more reports of selecting anti-stalk antibodies for H1 and H3.
-The reviewer's counterpoint is to mention on of the H3 studies. 
-However, we didn't say that there are no H3 reports, just that there are a lot more H1 reports.
-Certainly this is not systematic proof, but our statement in the text goes to great pains to caveat that this is purely circumstantial and not systematic proof.
-But since this isn't a central claim of the paper of anyway, let's just remove it in a concession to the reviewer.
-The study we're helping Nicholas Wu with will eventually systematically test this point.
-}
 

### 2018-06-23 paper/manuscript.tex
comment 5 reviewer 1

@@
 We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.1}) to fit the substitution models listed in Table~\ref{tab:phydms}.
+
 The amino-acid preferences were re-scaled by the stringency parameter using the approach described in \cite{hilton2017phydms}.
+Note that the re-scaling simply puts the amino-acid preferences on a useful scale for visualization in logo plots, and has no effect on any of the quantitative conclusions relating the deep mutational scanning to natural evolution.
+The reason is that all of these conclusions use the effects of mutations as calculated using Equation~\ref{eq:muteffect}, and the re-scaling simply acts as a constant multiplier on all mutational effects (e.g., a scale factor) when re-scaled preferences are converted to mutational effects.
 
@@
 \begin{equation}
+\label{eq:muteffect}
 \log_2 \frac{\pi_{r,a_2}}{\pi_{r,a_1}}

### 2018-06-23 paper/reviewer_response_PNAS.tex
respond to reviewer 1 comment 7

@@
 \response{The reviewer has a number of questions and suggestions about the experimental design.
-Over the last few years, we have performed a number of studies optimizing procedures for deep mutational scanning of influenza virus libraries (see PMIDs 24859245, 25006036, 27271655) during which we have carefully considered all of the experimental parameters that the reviewer discusses.
-Below we briefly explain why each of our choices for these experimental parameters were appropriate.
-We do not deny that there are slightly different ways that the experiments could be done, but the in end, the ``proof is in the pudding.'' 
-In the manuscript, we show that the measurements made using the experimental parameters that we chose lead to useful data as judged by the following criterion:
-\begin{itemize}
-\item The results from fully independent experimental replicates are highly correlated (Fig.~1D), which would not be the case if the results were dominated by noise.
-\item The experimental measurements do a good job of predicting the effects of individually validated mutations, as shown in Figure S8.
-\item The experimental measurements can inform substitution models that describe influenza virus's evolution much better than standard models (Table~1).
-\item The experimental measurements provide information that can help predict the fates of influenza virus lineages in nature (Fig.~5).
-\end{itemize}
-We do not claim that other experimental parameter choices might not lead to equivalent or even slightly better results, but we have carefully considered all of our choices and clearly they lead to very good results.
+Some of these suggestions are things that were already implemented in our experiments and we simply failed to explain in adequate detail, while others are things that we have evidence are unimportant.
+Below we describe the issues one-by-one.
 
-Below we explain our reasoning for each choice about which the reviewer makes suggestions:
+In addition, we have now included our \emph{full original lab notes} for the relevant experiments (\url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/tree/master/lab_notes}) so that readers can examine everything that we did in full detail.
+This level of transparency goes well beyond that usually taken in scientific papers, and will enable readers to fully understand our reasoning and procedures for each detail of the experiments.
 }
@@
 This is why we used an extremely low MOI (0.0035) so that most cells are still infected by no more than one virion.
-It is impossible to ever grow a viral stock that is completely free of defective particles.
+It is impossible to grow a viral stock that is completely free of defective particles.
 However, in both our original rescue with helper virus and in the passage, we follow the steps suggested by Carolina Lopez and Scott Hensley (PMID 27047455) to minimize defective particles.
-These include low MOI (0.0035 TCID50 per cell is quite low), short growth times to avoid accumulation of defective particles (the helper-virus rescue was allowed to propagate for just 24 hours, and the low-MOI passage for just 48 hours), and replacing the media three hours post-infection to remove residual defective particles.
+These include low MOI (0.0035 TCID50 per cell is quite low), short growth times to avoid accumulation of defective particles (the helper-virus rescue was allowed to propagate for just 24 hours, and the low-MOI passage for just 48 hours).
+
+In short, the presence of non-infectious particles is an inherent feature of all virology work.
+The best that can be done is to use MOIs and growth times that ensure that they do not dominate the experiments, and we have done that.
 }
@@
 
-\response{For biological replicates 1, 2, and 3-2, we replaced the original infectious inoculum with 25 mL fresh IGM three hours post-infection.
-For replicate 3-1, we did not change the media. 
-However, for this replicate, the original inoculum makes a negligible contribution to the total number of virions that are present at the time of sequencing.
-Specifically, in the original inoculum placed on the cells, the titer is 2.5 TCID50 / ul, whereas at the time of supernatant collection for sequencing the titer is 32 TCID50 / ul for unconcentrated virus but 400 TCID50 / ul for the concentrated virus.
-Therefore, the original inoculum constitutes just 8\% of the viral titer at the time of collection, and therefore makes a negligible contribution to the sequenced material.
-We deeply apologize for not including this crucial step in the Methods, and have updated the Methods to now include completion of this step. 
-We have also included in the GitHub repository a PDF of the lab notebook describing the rescue and passage of the mutant virus libraries at %\url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/lab_notes/Perth09%20HA%20Library%20Rescue.pdf} (relevant details of the notebook can be found in pages 20-25).
+\response{
+We apologize for not expanding on this detail in the Methods of the original paper.
+
+In fact, like the reviewer, we initially thought that a media change to wash aways the inoculum might be important. 
+Therefore, in our experiments, we performed such a media change to remove the initial inoculum for biological replicates 1 and 2, and for one of the technical replicates (3-2) of biological replicate 3.
+We did not perform such a change for the other technical replicate (3-2) of biological replicate 3.
+It turns out that the media change makes no discernible difference, which is why we neglected to expand on this point in the original manuscript. 
+We now have added the following explanatory text to the Methods:
+
+\begin{quote}
+\textsl{Three hours post-infection, we replaced the inoculum with fresh media for replicates 1, 2, and 3-2.
+We did not perform a media change for replicate 3-1.
+As can be seen in Figure~1D, the media change does not appear to have a substantial effect, as replicate 3-1 looks comparable to the other replicates.
+}
+\end{quote}
+The full details of the media changes can be found in the lab notes we have included at We have also included in the GitHub repository a PDF of the lab notebook describing the rescue and passage of the mutant virus libraries at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/lab_notes/Perth09_HA_Library_Rescue.pdf} (relevant details of the notebook can be found in pages 21-25).
+
+We think that the reason that washing away the original inoculum has no discernible effect on the results it that the low concentration of virus in the inoculum makes a negligible contribution to the total number of virions that are present at the time of sequencing, since vastly more new virions are produced from the infected cells.
 }
@@
 
-\response{We passaged for each mutant virus library $9\times10^5$ TCID50 total across 15 $\times$ 15-cm dishes.
+\response{
+We apologize if the original wording of our methods were unclear. 
+For each mutant virus library, we passaged a total of $9\times10^5$ TCID50, but this was passaged across \emph{fifteen} 15-cm dishes \emph{each} of which received 25 ml.
 Because we used an inoculum of 2.5 TCID50 / $\mu$l across these fifteen dishes, this corresponds to dilutions that range from 1:37 to 1:294.
-The quantity of dishes used to passage each mutant virus library is now mentioned in the Methods.
+The number of dishes used to passage each mutant virus library is now mentioned in the Methods; before it had to be calculated from the stated total number of cells and the cells per dish.
+
+Also, we note that the supernatants from the helper virus rescues are collected at just 24 hours post-infection, which leads to lower viral titers (which is why we have to passage relatively large volumes), but also presumably much lower amounts of DIs / RNA.
+This is probably the reason why (as described above), we did removing the inoculum did not yield discernibly affect the results.
 }
@@
 
-\response{We clarified the viral supernatant before RNA extraction by centrifuging the supernatant at 2000 $\times$ \textit{g} for five minutes to remove any cellular debris.
-We failed to mention this step in the Methods, and again deeply apologize for this critical oversight.
-We have updated the Methods to now include mention of this step.
+\response{We apologize for failing to provide sufficiently detailed methods stating that we ultra-centrifuged \emph{clarified} the viral supernatant before RNA extraction.
+The supernatant was clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes to remove any cellular debris, similar to the reviewer's suggestions.
+We have updated the Methods to now include this detail, and (as mentioned above) also now include our full lab notes so that all other details can be reviewed.
+Specifically, see \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/lab_notes/Perth09_HA_Library_Rescue.pdf} (relevant details of the notebook can be found in pages 21-25)
 }
@@
 \response{For the reasons described above, it is not necessary to totally repeat the experiments.
-Certainly other experimental parameter choices could have been used, but this is always the case for complex experiments.
-Our choices are reasonable, and the four lines of evidence listed at the beginning of this comment show that they yield reproducible and meaningful results.}
+We do not deny that there are slightly different ways that the experiments could be done, but the in end, the ``proof is in the pudding.'' 
+In the manuscript, we show that the measurements made using the experimental parameters that we chose lead to useful data as judged by the following criterion:
+\begin{itemize}
+\item The results from fully independent experimental replicates are highly correlated (Fig.~1D), which would not be the case if the results were dominated by noise.
+\item The experimental measurements can inform substitution models that describe influenza virus's evolution much better than standard models (Table~1).
+\item The experimental measurements provide information that can help predict the fates of influenza virus lineages in nature (Fig.~5).
+\end{itemize}
+For any experimental study there are an almost infinite number of slight methodological variations that could be employed, and it is impossible to test them all.
+Our experiments used reasonable parameter choices, were performed in full biological triplicate, and give meaningful / useful results with respect to the larger goals of the study.}
 

### 2018-06-23 paper/manuscript.tex
respond to reviewer 1 comment 7

@@
 Three hours post-infection, we replaced the inoculum with fresh IGM for replicates 1, 2, and 3-2.
+We did not perform a media change for replicate 3-1.
+As can be seen in Figure~\ref{fig:dms_overview}D, the media change does not appear to have a substantial effect, as replicate 3-1 looks comparable to the other replicates.
 We collected viral supernatant for sequencing 48 hours post-infection.
+
+Full lab notes from these experiments are at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/tree/master/lab_notes}.
 
@@
 We performed deep sequencing on a lane of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.
+
+Full lab notes for the library preparation are at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/tree/master/lab_notes}.
 

### 2018-06-23 paper/reviewer_response_PNAS.tex
finished responses to Reviewer 1

@@
 \response{We individually validated viral growth of a C52A mutant, which removes a Cys critical in forming a disulfide bond with Cys-277 and maintaining HA structure.
-We confirmed that this mutant grows very poorly when rescued, and mutations at site 52 never arise at a detectable level.
+We confirmed that this mutant grows very poorly when rescued, and mutations at site 52 never appear in sequenced isolates.
 
@@
 }
-
-\comment{Not obvious how to give an example, as there are many mutations we identify as deleterious and many mutations that don't reach high frequency. 
-But I suppose if we validate some specific mutation (e.g., a cysteine mutation) as deleterious, we can point out that it doesn't reach high frequency...}
 
@@
 Two of the clones had a mutation that removed the ATG start codon.
-One of these resulted in an M(-16)V mutation, while the other resulted in an M(-16)Q mutation which was linked to a Q80I mutation.
-According to our preferences, Val and Gln are not very preferred at this site, suggesting that these mutants would not grow well compared to wildtype.
+We did not test the growth of these clones. 
+As explained in Figure~S2, these were plasmid variants chosen randomly to Sanger sequence to evaluate the mutation rate prior to creating virus libraries; they were not evaluated as viruses.
+The new Figures S4 to S6 now show the growth phenotype of other selected mutants, one of which includes a start codon mutant.
 }

### 2018-06-23 paper/reviewer_response_PNAS.tex
some responses to Reviewer 2

@@
 We thank the reviewers for the careful reading and suggestions about the manuscript.
+We were gratified that the reviewers recognized the quality of our experimental work (one of the reviewers said, ``the experimental setup is sound and there is strong correlation among technical and biological replicates.'').
+
+We also appreciate the reviewer's recognition of the important step our work takes in linking experiments with quantitative models of influenza virus evolution, and their recognition that we adequately discuss the caveats as well as the strengths of our approach.
+
 In the revised version, we have performed additional experiments and analyses that address all of the reviewer questions and concerns.
@@
 It is difficult to also overlay the structural location (stalk versus head) on the current Fig.~4, since colors are already used to indicate mutational effects.
-We have therefore created a new supplemental figure (Fig.~S8) that shows these data---as the reviewer suspected, this shows that head mutations are much more common.
+We have therefore created a new supplemental figure (Fig.~S11) that shows these data---as the reviewer suspected, this figure shows that head mutations reach high frequencies much more commonly than stalk mutations.
 
@@
 \textsl{Indeed, a vast amount of work in virology has chronicled the ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature.
-As an example of this, although we identified G78D as a lab adaptation mutation and the wildtype Gly is not highly preferred at this site, this mutation never reaches above 20\% frequency in nature. }
+As an example, although we identified G78D as favorable for viral growth in cell culture, this mutation never fixes in nature. }
 \end{quote}
@@
 \response{We agree that this would be a very interesting experiment to perform.
-However, we believe such a study, which would require rigorous examination of escape from multiple head- and stalk-specific antibodies, is beyond the scope of this manuscript.
+However, such experiments would be far beyond the scope of our study.
+As stated in the Abstract and Significance Statement, the main goal of our study is to perform deep mutational scanning of a recent H3 HA and use the data to better predict H3N2 viral evolution in nature.
+A large set of additional experiments comparing stalk- and head-specific antibodies in H3 and H1 HAs is clearly beyond the scope of this goal.
 }
@@
 
-\response{We now have acknowledged this possibility in the Discussion in the sentence ``We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains."}
+\response{This is a good point, and we have added the following sentence to the Discussion: ``We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains."}
 
@@
 
-\response{This is a fair assessment by the reviewer.
-However, this is the first paper to show that data from deep mutational scanning can actually reflect evolutionary outcomes in nature, and provides some insight into the extent by which functional constraints on HA shape its evolution.
-Furthermore, it is possible that this work can provide a useful framework for applying deep mutational scanning studies in the context of evolution in nature.
+\response{The reviewer is correct that additional experiments might lead to further improvements.
+However, ours is the first paper to show that data from deep mutational scanning can actually help predict real evolutionary outcomes in nature---which is of itself a substantial step forward that we believes merits publication.
+We are excited about the future directions that the reviewer suggests.
+However, the current study is a complete contribution even if (as is almost invariably the case in science) future work can probably take things further.
 }
@@
 
-\response{We now include an analysis of mutational effects and maximum frequency for mutations of unpassaged isolates (see Figures 5, 6, and S7) from the post-Perth partition of the phylogenetic tree to limit any potential biases that our experimental measurements may have towards mutations from cell- or egg-passaged isolates.
-The results show that this correlation is stronger when considering mutations from unpassaged isolates in a post-hoc analysis, demonstrating the potential utility of our experimental measurements in a prospective analysis, specifically when passaged isolate sequences are excluded.
+\response{
+The reviewer is correct that on the complete data sets in the original submission, the correlation was slightly higher for the post-hoc (post-Perth) and prospective (pre-Perth) analysis.
+But at the suggestion of Reviewer \#1, we have now also included a post-Perth analysis using \emph{only} HA sequences from unpassaged viral isolates (there are not enough unpassaged isolates from before 2009 to perform a similar unpassaged pre-Perth analysis).
+As shown in the new Figure 5, the correlation for the unpassaged post-Perth is actually higher than either other set, demonstrating that there is not a systematic trend for poorer performance in prospective analyses.
+In addition, none of these differences seem sufficiently large ($\rho$ ranges from 0.17 to 0.24) that we are confident in claiming strong differences in performance across datasets.
 }
@@
 \item We now define AIC in the legend of Table 1.
-\item We have now included an explanation of dN/dS $\ll$1 for conventional substitution models, and now describe the utility of building an ExpCM, site avg. substitution model
+\item We have now included an explanation of dN/dS $\ll$1 for conventional substitution models.
+\item We now describe how the ``ExpCM, site avg.'' substitution model is a control that shows that the improvements are due to the site-specific nature of the data, since this model averages data across sites.
 \end{itemize}
@@
 
-\response{We have now labeled the Perth/2009 strain on the tree in Figure 4 and now mention in the caption that we have excluded closely related strains from the analysis...
+\response{We have now labeled the Perth/2009 strain on the tree in Figure 4 and now mention in the caption that we have excluded closely related strains from the analysis.
+We also now explain in the Methods that the frequencies are calculated in one month windows, and the denominator is the total number of sequences in that window.
 }
@@
 
-\response{We have now fixed the graph labels in figure 8B accordingly.}
+\response{We have now fixed the graph labels in Figure 8B as suggested by the reviewer.}
 

### 2018-06-23 paper/manuscript.tex
some responses to Reviewer 2

@@
 It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
+Figure~\ref{suppfig:head_stalk_mutfreq_supp} shows a similar layout but colors mutations by whether they are in HA's head or stalk domain.
 }
@@
 Indeed, a vast amount of work in virology has chronicled the ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}.
-As an example of this, although we identified G78D as a lab adaptation mutation and the wildtype Gly is not highly preferred at this site, this mutation never reaches above 20\% frequency in nature.
+As an example, although we identified G78D as favorable for viral growth in cell culture, this mutation never fixes in nature.
 Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
@@
 (B) Histogram of mutation maximum frequencies by the number of nodes for head and stalk domain mutations.
-It is clear that mutations in the head domain are more numerous than those in the stalk.
+It is clear that mutations in the head domain are more numerous than those in the stalk, particularly among mutations that reach high frequencies.
 }

### 2018-06-23 paper/reviewer_response_PNAS.tex
more tweaks to reviewer response

@@
 This is an excellent question.
-We now a separate analysis of the relationship between our measured mutational effects and maximum frequency in nature for mutations from unpassaged isolates only.
+We now perform a separate analysis of the relationship between our measured mutational effects and maximum frequency in nature for mutations from unpassaged isolates only.
 Because sequences from unpassaged isolates have only become common relatively recently, we could perform this analysis only for the post-Perth/2009 partition of the H3N2 HA phylogenetic tree.
@@
 Also, we note that the supernatants from the helper virus rescues are collected at just 24 hours post-infection, which leads to lower viral titers (which is why we have to passage relatively large volumes), but also presumably much lower amounts of DIs / RNA.
-This is probably the reason why (as described above), we did removing the inoculum did not yield discernibly affect the results.
+This is probably the reason why (as described above) removing the inoculum did not discernibly affect the results.
 }
@@
 \response{We apologize for failing to provide sufficiently detailed methods stating that we ultra-centrifuged \emph{clarified} the viral supernatant before RNA extraction.
-The supernatant was clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes to remove any cellular debris, similar to the reviewer's suggestions.
+The supernatant was clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes to remove any cellular debris.
 We have updated the Methods to now include this detail, and (as mentioned above) also now include our full lab notes so that all other details can be reviewed.
@@
 
-\response{For the reasons described above, it is not necessary to totally repeat the experiments.
+\response{For the reasons described above, it is not necessary to repeat the experiments.
 We do not deny that there are slightly different ways that the experiments could be done, but the in end, the ``proof is in the pudding.'' 
@@
 \response{This is a good suggestion and we have now added the results of this analysis in Figure S4.
-As the logo plot shows, although there are some differences, at most sites there is little difference between the amino acids that are preferred in our experiments to their frequencies in nature.
 }
@@
 \response{The reviewer is correct that additional experiments might lead to further improvements.
-However, ours is the first paper to show that data from deep mutational scanning can actually help predict real evolutionary outcomes in nature---which is of itself a substantial step forward that we believes merits publication.
+However, ours is the first paper to show that data from deep mutational scanning can actually help predict real evolutionary outcomes in nature---which is an important step forward that merits publication.
+
 We are excited about the future directions that the reviewer suggests.
-However, the current study is a complete contribution even if (as is almost invariably the case in science) future work can probably take things further.
+However, the current study is a complete and important contribution even if (as is almost invariably the case in science) future work can probably take things further.
+Researchers have been performing functional experiments on influenza virus for decades, and have been trying to forecast the virus's evolution for decades---yet ours is the first study to successfully use such functional experiments to predict the success of viral strains in nature.
+We have gone to great pains to fairly discuss limitations and caveats and how things could be improved in the future, but that should not detract from the advances in the current paper.
 }
@@
 
-\comment{Fixed the figure. Maybe John and Trevor can do a better job than me at explaining the frequency denominator. Is it the population observed at a given pivot?}
-
 The graph labels in figure 8B could be cleaner.

### 2018-06-23 paper/reviewer_response_PNAS.tex
finished Jesse revisions

@@
 These start-codon mutant viruses grow to appreciable titers, suggesting that there is indeed some flexibility in the site of translation initiation in this HA.
-These data are in Figure S5 of the revised manuscript.
+These data are in Figure S4 of the revised manuscript.
 As we note in the manuscript, this finding is concordant with prior work on suggesting additional translation initiation sites sites in H2 HAs.
@@
 The third mutation, S287A, removes a glycosylation site that is highly variable in nature but that our experiments suggest is under functional constraint.
-The new Figure S6 shows that the T24F and T40V mutants grow to titers comparable to wildtype, supporting the conclusion that they are expendable at least in cell culture. 
+The new Figure S5 shows that the T24F and T40V mutants grow to titers comparable to wildtype, supporting the conclusion that they are expendable at least in cell culture. 
 However, the titer of the S287A is modestly lower than wildtype, supporting the conclusion that this glycan contributes to viral growth in cell culture.
@@
 
-In addition to the data in Figure S6, we have added the following text:
+In addition to the data in Figure S5, we have added the following text:
 
@@
 \response{We have validated that one of these mutations in the transmembrane domain, C199(HA2)K, reaches reverse-genetics titers comparable to wildtype.
-These data are in the new Figure S7.
+These data are in the new Figure S6.
 In addition, various other studies have found that chimeric HAs with transmembrane domains from completely different proteins can still maintain HA cell surface expression and fusion activity (Melikyan et al 1999 Mol Biol Cell, Schroth-Diez et al 1998 JVI, Dong et al 1992 JVI, Roth et al 1986 JCB).
@@
 As explained in Figure~S2, these were plasmid variants chosen randomly to Sanger sequence to evaluate the mutation rate prior to creating virus libraries; they were not evaluated as viruses.
-The new Figures S4 to S6 now show the growth phenotype of other selected mutants, one of which includes a start codon mutant.
+The new Figures S5 to S7 now show the growth phenotype of other selected mutants, one of which includes a start codon mutant.
 }
@@
 
-\response{This is a good suggestion and we have now added the results of this analysis in Figure S4.
+\response{This is a good suggestion and we have now added the results of this analysis in Figure S3.
 }
@@
 As shown in the new Figure 5, the correlation for the unpassaged post-Perth is actually higher than either other set, demonstrating that there is not a systematic trend for poorer performance in prospective analyses.
-In addition, none of these differences seem sufficiently large ($\rho$ ranges from 0.17 to 0.24) that we are confident in claiming strong differences in performance across datasets.
 }

### 2018-06-23 paper/manuscript.tex
more tweaks to reviewer response
finished Jesse revisions

@@
 These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.
-Furthermore, Figure~\ref{suppfig:diffprefs_logoplot} shows that when we compare these preferences to amino-acid frequencies from natural human H3N2 HA sequences, at most sites, there is little difference between the amino acids that are preferred in our experiments to their frequencies in nature.
-\comment{This needs to be improved / changed}
 
@@
 
-However, there are also some notable differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences.
+However, there are also some differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences (Figure~\ref{suppfig:diffprefs_logoplot}).
 Most surprisingly, the start codon does not show a particularly strong preference for methionine (Figure~\ref{fig:logoplot}).
@@
 We validated that viruses with mutations to the glycosylation motifs at sites 22 or 38, or a site in the transmembrane domain, do in fact grow to high titers (Figures~\ref{suppfig:glycan_validation} and \ref{suppfig:TM_validation}, respectively).
-The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function on nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.
+The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function on nature, or could be because these sites are not under strong immune pressure and so mutations to them are not positively selected in nature.
 
@@
 
-\begin{figure}[b!]
+\begin{figure}[t!]
 \centering
@@
 
-Many of the HAs in sequence databases were obtained from virus isolates that were passaged in cell-culture or eggs, which can lead to the acquisition of lab-adaptation mutations that can confound evolutionary analyses~\cite{mcwhite2016sequence}.
-To check that our results were not confounded by such lab-adaptation mutations, we repeated our analysis using only HA sequences that were annotated as being derived from viruses that had not been passaged in the lab.
-Because sequencing of unpassaged primary isolates has only become commonplace relatively recently, we were only able to perform this analysis for the post-Perth partition of the phylogenetic tree.
+Many of the HAs in sequence databases are from viral isolates that were passaged in cell-culture or eggs, which can cause lab-adaptation mutations that confound evolutionary analyses~\cite{mcwhite2016sequence}.
+To check that our results were robust to such lab-adaptation mutations, we repeated our analysis using only HA sequences derived from viruses that had not been passaged in the lab.
+Because sequencing of unpassaged primary isolates has only recently become commonplace, we could only perform this analysis for the post-Perth partition of the phylogenetic tree.
 Figure~\ref{fig:muteffect_maxfreq}A shows that the correlation between our measured mutational effects and the maximum frequency was even stronger for mutations from unpassaged viral isolates ($\rho = 0.24$).
@@
 \subsection*{Measurements made on an H1 HA are not useful for understanding the evolution of H3 influenza}
-To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
+To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (Figure~\ref{suppfig:WSNprefs_logoplot}), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
 Figure~\ref{fig:muteffect_maxfreq_WSN} shows that there is no correlation between the H1 experimental measurements and maximum frequency that mutations reach during H3N2 viral evolution.
@@
 
-Full lab notes from these experiments are at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/tree/master/lab_notes}.
-
 \subsection*{Barcoded subamplicon sequencing}
@@
 
-Full lab notes for the library preparation are at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/tree/master/lab_notes}.
-
 \subsection*{Analysis of deep sequencing data}
-We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, version \texttt{2.2.0}) to analyze the deep sequencing data.
+We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, version \texttt{2.2.5}) to analyze the deep sequencing data.
 The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
@@
 These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
-We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.1}) to fit the substitution models listed in Table~\ref{tab:phydms}.
+We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.2}) to fit the substitution models listed in Table~\ref{tab:phydms}.
 
@@
 
-To determine the effect of virus passaging on the relationship between mutational effects and maximum frequencies in natural populations, we repeated the above analyses using only viruses that were sequenced directly without passaging.
+We also repeated the above analyses using only viruses that were sequenced directly without passaging.
 Routine direct sequencing did not begin until the early 2000s \cite{mcwhite2016sequence}.
@@
 To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, C199(HA2)K, T24F, T40V, and S287A, we used site-directed mutagenesis to introduce the amino-acid mutation into the Perth/2009 HA bidirectional reverse genetics plasmid, and verified the sequence of two clones for each mutant by Sanger sequencing.
-We generated these individual mutant viruses carrying GFP in the PB1 segment using a protocol described in \cite{hooper2013mutant}, and the PB2, PA, NP, NA, M, and NS segments of the A/WSN/1933 (H1N1) strain.
+We generated these individual mutant viruses carrying GFP in the PB1 segment using a protocol described in \cite{bloom2010permissive,hooper2013mutant}, and the PB2, PA, NP, NA, M, and NS segments of the A/WSN/1933 (H1N1) strain.
 To rescue each mutant GFP-carrying virus in duplicate, we transfected a co-culture of $4 \times 10^5$ 293T-CMV-PB1 and $0.5 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells with the eight reverse genetics plasmid and the pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W plasmid.
@@
 \clearpage
-\subsection*{Supporting Information (SI)}
-\clearpage
+\section*{Supporting Information (SI)}
+\FloatBarrier
+\pagenumbering{arabic}% resets `page` counter to 1
+\renewcommand*{\thepage}{S\arabic{page}}
 
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_WSNprefs_logoplot/supp_logoplots.pdf}}
-\caption{\label{suppfig:WSNprefs_logoplot}
-{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.}
-\comment{Get rid of unscaled Perth/2009 in this figure. It doesn't make sense to show them here. I've removed from legend; remove from the actual figure}
-The amino-acid preferences of the WSN/1933 H1 HA from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
-The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
-Note that \cite{doud2016accurate} used libraries in which all codons were mutagenized \emph{except} for the one encoding N-terminal methionine.
-Therefore, no data is shown for the first codon in the gene.
-The sites are in H3 numbering.
-}
-\end{suppfigure}
-
-\begin{suppfigure}[H]
 \centerline{\includegraphics[width=0.9\textwidth]{figs/supp_diffprefs/diffprefs.pdf}}
 \caption{\label{suppfig:diffprefs_logoplot}
-{\bf The differential preferences of the Perth/2009 H3 HA compared to amino-acid frequencies from a sequence alignment of natural human H3.}
-The differential preferences estimated by comparing the Perth/2009 H3 HA preferences to amino-acid frequencies from a sequence alignment of natural human H3N2 HA sequences.
-For each site, the height of each letter above and below the line indicate how differentially preferred that amino acid is in our experiments compared to in nature.
+{\bf Sites where there are strong differences between our experimental measurements and the amino-acid frequencies among natural HA sequences.}
+We calculated the distance between our H3 measurements and the alignment frequencies using the same approach as in Figure~\ref{fig:distance_distribution}, but using the alignment frequencies in place of the H1 preferences.
+For each site, the height of each letter above or below the line indicate how much more or less preferred that amino acid is in our experiments compared to its frequency nature.
 The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
 The sites are in H3 numbering.
+The HA alignment used to calculate the natural frequencies is the same one used in Table~\ref{tab:phydms}.
+\comment{Juhye: the notebook that creates this has a bug, as it never creates the \texttt{H3\_alignment\_frequences\_copy.csv}. 
+This must be something you did by hand but then didn't include the code for in the notebook: fix this!
+Also, it might be nice if you added a bit more explanation to this section of the Jupyter notebook.
+}
 }
@@
 \begin{suppfigure}[H]
+\centerline{\includegraphics[width=0.75\textwidth]{figs/supp_head_stalk_RSA/head_stalk_RSA.pdf}}
+\caption{\label{suppfig:head_stalk_RSA}
+{\bf Mutational tolerances of the head and stalk domains at various relative solvent accessibility cutoffs.}
+The mutational tolerances of the head and stalk domains show less disparity for the Perth/2009 H3 HA compared to those for the WSN/1933 H1 HA.
+We used relative solvent accessibility (RSA) cutoffs of $0.1$, $0.2$, and $0.3$ to define solvent-exposed residues and plotted the mutational tolerances (Shannon entropy of re-scaled preferences) of these residues in the head and stalk domains for the Perth/2009 H3 HA (purple) and the WSN/1933 H1 HA (brown).
+Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
+The HA structures color the residues that are defined as solvent exposed at a given RSA cutoff.
+One monomer is shown in surface representation and another monomer shown in ribbon representation.
+Residues in lighter shades of purple or brown are in the head domain, while residues in darker shades are in the stalk domain.
+Note that the mutational tolerance values are not comparable between the two HAs.
+}
+\end{suppfigure}
+
+\begin{suppfigure}[H]
 \centerline{\includegraphics[width=0.9\textwidth]{figs/supp_H3N2_phylogeny/H3N2_phylogeny.pdf}}
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.75\textwidth]{figs/supp_head_stalk_RSA/head_stalk_RSA.pdf}}
-\caption{\label{suppfig:head_stalk_RSA}
-{\bf Mutational tolerances of the head and stalk domains at various relative solvent accessibility cutoffs.}
-The mutational tolerances of the head and stalk domains show less disparity for the Perth/2009 H3 HA compared to those for the WSN/1933 H1 HA.
-We used relative solvent accessibility (RSA) cutoffs of $0.1$, $0.2$, and $0.3$ to define solvent-exposed residues and plotted the mutational tolerances (Shannon entropy of re-scaled preferences) of these residues in the head and stalk domains for the Perth/2009 H3 HA (purple) and the WSN/1933 H1 HA (brown).
-Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
-The HA structures color the residues that are defined as solvent exposed at a given RSA cutoff.
-One monomer is shown in surface representation and another monomer shown in ribbon representation.
-Residues in lighter shades of purple or brown are in the head domain, while residues in darker shades are in the stalk domain.
-Note that the mutational tolerance values are not comparable between the two HAs.
+\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_WSNprefs_logoplot/supp_logoplots.pdf}}
+\caption{\label{suppfig:WSNprefs_logoplot}
+{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.}
+\comment{Get rid of unscaled Perth/2009 in this figure. It doesn't make sense to show them here. I've removed from legend; remove from the actual figure}
+The amino-acid preferences of the WSN/1933 H1 HA from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
+The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
+Note that \cite{doud2016accurate} used libraries in which all codons were mutagenized \emph{except} for the one encoding N-terminal methionine.
+Therefore, no data is shown for the first codon in the gene.
+The sites are in H3 numbering.
 }

### 2018-06-24 paper/manuscript.tex
Split mut validation fig into separate figs, update captions

@@
 The HA alignment used to calculate the natural frequencies is the same one used in Table~\ref{tab:phydms}.
-\comment{Juhye: the notebook that creates this has a bug, as it never creates the \texttt{H3\_alignment\_frequences\_copy.csv}. 
-This must be something you did by hand but then didn't include the code for in the notebook: fix this!
-Also, it might be nice if you added a bit more explanation to this section of the Jupyter notebook.
-}
 }
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
+\centerline{\includegraphics[width=0.4\textwidth]{figs/supp_point_mut_validation/startcodon_titers.pdf}}
 \caption{
-\comment{Just show the first panel (A) here}
 {\bf Validation that the Perth/2009 HA is somewhat tolerant to mutation of the canonical start codon.}
-We created variants of the Perth/2009 HA in which the canonical start codon was mutated (amino-acid mutation M(-16)K, codon mutation \comment{what was it}), a conserved cysteine was mutated to alanine (C52A, codon mutation \comment{what was it?}), and the same cysteine was synonymously mutated (C52C, codon mutation \comment{what was it}).
+We created variants of the Perth/2009 HA in which the canonical start codon was mutated (amino-acid mutation M(-16)K, codon mutation ATG(-16)AAA), a conserved cysteine was mutated to alanine (C52A, codon mutation TGC52GCA), and the same cysteine was synonymously mutated (C52C, codon mutation TGC52TGT).
 We selected these mutations because our deep mutational scanning results in Figure~\ref{fig:logoplot} surprisingly suggests that the start codon (position -16) is fairly tolerant of mutations, but that site 52 is highly intolerant of mutation to anything other than cysteine.
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
+\centerline{\includegraphics[width=0.4\textwidth]{figs/supp_point_mut_validation/glycan_titers.pdf}}
 \caption{\label{suppfig:glycan_validation}
-\comment{Just show the third panel (C) here, removing the cysteine mutants and adjusting the y-axis to a more appropriate for just these mutants.}
 {\bf Validation of viral variants with mutations at N-linked glycosylation motifs.}
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
+\centerline{\includegraphics[width=0.45\textwidth]{figs/supp_point_mut_validation/tm_titers.pdf}}
 \caption{\label{suppfig:TM_validation}
-\comment{Just show TM mutant here, similar to previous figure}
 {\bf Validation of the mutational tolerance of a site in the transmembrane domain.}

### 2018-06-24 paper/manuscript.tex
Remove un-scaled Perth logoplot supp figure

@@
 We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
-In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).
+In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.15$).
 The similar effect sizes for both the pre- and post-Perth partitions shows that our experimental measurements can help explain the evolutionary fates of mutations in strains that post-date the experimentally studied strain, as well as to retrospectively analyze mutations that precede the experimental strain.
@@
 \begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_WSNprefs_logoplot/supp_logoplots.pdf}}
+\centerline{\includegraphics[width=0.9\textwidth]{figs/supp_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
 \caption{\label{suppfig:WSNprefs_logoplot}
 {\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.}
-\comment{Get rid of unscaled Perth/2009 in this figure. It doesn't make sense to show them here. I've removed from legend; remove from the actual figure}
 The amino-acid preferences of the WSN/1933 H1 HA from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).

### 2018-06-24 paper/manuscript.tex
Finish manuscript edits

@@
 We validated that viruses with mutations to the glycosylation motifs at sites 22 or 38, or a site in the transmembrane domain, do in fact grow to high titers (Figures~\ref{suppfig:glycan_validation} and \ref{suppfig:TM_validation}, respectively).
-The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function on nature, or could be because these sites are not under strong immune pressure and so mutations to them are not positively selected in nature.
+The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function in nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.
 
@@
 Some mutations appear, spread and fix in the population, while others briefly circulate before disappearing.
-We take the maximum frequency reached by a mutation as a coarse indicator of its effect on fitness, since favorable mutations generally reach higher frequency than unfavorable ones~\cite{ewens2012mathematical}.
+We take the maximum frequency reached by a mutation as a coarse indicator of its effect on fitness, since favorable mutations generally reach higher frequencies than unfavorable ones~\cite{ewens2012mathematical}.
 Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, e.g.\ although R142G occurs multiple times on the phylogeny we track each of these mutations occurring on different backgrounds separately.
@@
 To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
-We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
+We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after, since these mutations have not had enough time for their evolutionary fates to be fully resolved.
 We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
@@
 {\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 mutations.}
-This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}, i.e.\
+This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
 Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
-The empirical $P$-value represents the results of 1,000 permutations.
+The empirical $P$-value represents the result of 1,000 permutations.
 }
@@
 We investigated this further by partitioning mutations into those that reach low, medium and high frequencies, and those that fix in the population (Figure~\ref{fig:muteffect_maxfreq}B).
-The mutations that reach higher frequency have a more favorable mean effect.
+The mutations that reach higher frequencies have a more favorable mean effect.
 Mutations measured to be substantially deleterious almost never reach high frequency.
@@
 To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (Figure~\ref{suppfig:WSNprefs_logoplot}), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
-Figure~\ref{fig:muteffect_maxfreq_WSN} shows that there is no correlation between the H1 experimental measurements and maximum frequency that mutations reach during H3N2 viral evolution.
+Figure~\ref{fig:muteffect_maxfreq_WSN} shows that there is no correlation between the H1 experimental measurements and maximum frequency that mutations reach during H3N2 viral evolution, and only a weak correlation with mutations from unpassaged viral isolates.
 Therefore, the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
@@
 When growing influenza virus in cell culture, trypsin is normally added to cleave HA into its mature form.
-To obviate the need for trypsin, we engineered MDCK-SIAT1 cells to constitutively express the TMPRSS2 protease, which cleaves and activates HA in the human airways~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
+To obviate the need for trypsin, we engineered MDCK-SIAT1 cells and MDCK-SIAT1-CMV-PB1 cells to constitutively express the TMPRSS2 protease, which cleaves and activates HA in the human airways~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
 The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656) and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry to create plasmid pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
-We used the lentiviral vector to transduce MDCK-SIAT1 cells, and sorted an intermediate mCherry-positive population by flow cytometry.
-We refer to the sorted bulk population as MDCK-SIAT1-TMPRSS2 cells.
+We used the lentiviral vector to transduce MDCK-SIAT1 or MDCK-SIAT1-CMV-PB1 cells, and sorted an intermediate mCherry-positive population by flow cytometry.
+We refer to the sorted bulk population as MDCK-SIAT1-TMPRSS2 cells or MDCK-SIAT1-CMV-PB1-TMPRSS2 cells.
 There is no selectable marker for the TMPRSS2; however, we maintain the cells at low passage number, and have seen no indication that they lose their ability to support the growth of viruses with H3 HAs in the absence of exogenous trypsin.
@@
 We calculated the effects of mutations from the amino-acid preferences that were estimated from the deep mutational scanning data.
-The effect of mutating site $r$ from amino-acid $a_1$ to $a_2$ was calculated as
+The effect of mutating site $r$ from amino acid $a_1$ to $a_2$ was calculated as
 \begin{equation}
@@
 Routine direct sequencing did not begin until the early 2000s \cite{mcwhite2016sequence}.
-To construct a tree with a similar number of viruses as the original analysis, we sampled 30 viruses per month between January 1, 2000 and April 1, 2018, producing a tree with 2,374 unpassaged viruses with augur (commit: 6d9f708).
+To construct a tree with a similar number of viruses as the original analysis, we sampled 30 viruses per month between January 1, 2000 and April 1, 2018, producing a tree with 2,374 unpassaged viruses with augur (commit: \texttt{6d9f708}).
 We included the passaged DMS strain, A/Perth/16/2009, in the resulting tree to enable comparison between pre-Perth and post-Perth clades.
@@
 \subsection*{Validation of individual point mutants}
-To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, C199(HA2)K, T24F, T40V, and S287A, we used site-directed mutagenesis to introduce the amino-acid mutation into the Perth/2009 HA bidirectional reverse genetics plasmid, and verified the sequence of two clones for each mutant by Sanger sequencing.
+To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, T24F, T40V, S287A, and C199(HA2)K, we used site-directed mutagenesis to introduce the amino-acid mutation into the Perth/2009 HA bidirectional reverse genetics plasmid, and verified the sequence of two clones for each mutant by Sanger sequencing.
 We generated these individual mutant viruses carrying GFP in the PB1 segment using a protocol described in \cite{bloom2010permissive,hooper2013mutant}, and the PB2, PA, NP, NA, M, and NS segments of the A/WSN/1933 (H1N1) strain.
-To rescue each mutant GFP-carrying virus in duplicate, we transfected a co-culture of $4 \times 10^5$ 293T-CMV-PB1 and $0.5 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells with the eight reverse genetics plasmid and the pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W plasmid.
+To rescue each mutant GFP-carrying virus in duplicate, we transfected a co-culture of $4 \times 10^5$ 293T-CMV-PB1 and $0.5 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells with the eight reverse genetics plasmids and the pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W plasmid.
 Each well received a transfection mixture of 100 $\mu$L DMEM, 3 $\mu$L BioT transfection reagent, and 250 ng of each plasmid.
@@
 We calculated the distance between our H3 measurements and the alignment frequencies using the same approach as in Figure~\ref{fig:distance_distribution}, but using the alignment frequencies in place of the H1 preferences.
-For each site, the height of each letter above or below the line indicate how much more or less preferred that amino acid is in our experiments compared to its frequency nature.
+For each site, the height of each letter above or below the line indicates how much more or less preferred that amino acid is in our experiments compared to its frequency nature.
 The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
@@
 We created variants of the Perth/2009 HA in which the canonical start codon was mutated (amino-acid mutation M(-16)K, codon mutation ATG(-16)AAA), a conserved cysteine was mutated to alanine (C52A, codon mutation TGC52GCA), and the same cysteine was synonymously mutated (C52C, codon mutation TGC52TGT).
-We selected these mutations because our deep mutational scanning results in Figure~\ref{fig:logoplot} surprisingly suggests that the start codon (position -16) is fairly tolerant of mutations, but that site 52 is highly intolerant of mutation to anything other than cysteine.
+We selected these mutations because our deep mutational scanning results in Figure~\ref{fig:logoplot} surprisingly suggest that the start codon (position -16) is fairly tolerant of mutations, but that site 52 is highly intolerant of mutation to anything other than cysteine.
 The synonymous C52C mutation is a negative control mutation that is not expected to have any effect.
@@
 As expected, the C52A virus is essentially non-viable, while the C52C mutation is roughly equivalent to wildtype.
-the M(-16)K mutation is only moderately attenuated, validating that the canonical start codon is not completely essential for viral growth.
+The M(-16)K mutation is only moderately attenuated, validating that the canonical start codon is not completely essential for viral growth.
 }
@@
 We created a variant of the Perth/2009 HA with a transmembrane domain mutation, C199(HA2)K, at a site that our deep mutational scanning suggests should be highly mutationally tolerant (Figure~\ref{fig:logoplot}).
-We then used reverse genetics to generate viruses carrying the wildtype HA or this mutants
+We then used reverse genetics to generate viruses carrying the wildtype HA or this mutant.
 Each variant was generated in duplicate, and the plots show the viral titer in the supernatant at 53 hours post-transfection.
-The virus the mutation reached titers comparable to wildtype.
+The virus with the mutation reached titers comparable to wildtype.
 }
@@
 The Perth/2009 strain was excluded to avoid artifacts related to mutations that occurred on the branches leading to the HA sequence used in the experiment.
-The post-2014 nodes were excluded because the evolutionary fates of many sequences after this date are not yet full resolved.
+The post-2014 nodes were excluded because the evolutionary fates of many sequences after this date are not yet fully resolved.
 (B) The post-Perth/2009 partition of the tree containing only sequences from unpassaged isolates.

### 2018-06-24 paper/reviewer_response_PNAS.tex
Finish Juhye edits to reviewer responses

@@
 Two of the mutants, T24F and T40V, remove the aforementioned highly conserved but mutationally tolerant NGT and NAT glycosylation sites near the start of HA1.
-The third mutation, S287A, removes a glycosylation site that is highly variable in nature but that our experiments suggest is under functional constraint.
+The third mutation, S287A, removes a glycosylation site that our experiments suggest is under functional constraint.
 The new Figure S5 shows that the T24F and T40V mutants grow to titers comparable to wildtype, supporting the conclusion that they are expendable at least in cell culture. 
-However, the titer of the S287A is modestly lower than wildtype, supporting the conclusion that this glycan contributes to viral growth in cell culture.
+However, the titer of the S287A mutant is modestly lower than wildtype, supporting the conclusion that this glycan contributes to viral growth in cell culture.
 
-One possible explanation is simply that cell culture does not fully capture the constraints in nature.
+One possible explanation is simply that cell culture does not fully capture the constraints on HA in nature.
 But a more interesting possibility is that conservation in nature is not expected to strictly parallel mutational tolerance.
@@
 
-In addition to the data in Figure S5, we have added the following text:
+In addition to the data in Figure S5, we have added the following text to the Results:
 
@@
 \textsl{
-The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function on nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.}
+The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function in nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.}
 \end{quote}
@@
 However, in both our original rescue with helper virus and in the passage, we follow the steps suggested by Carolina Lopez and Scott Hensley (PMID 27047455) to minimize defective particles.
-These include low MOI (0.0035 TCID50 per cell is quite low), short growth times to avoid accumulation of defective particles (the helper-virus rescue was allowed to propagate for just 24 hours, and the low-MOI passage for just 48 hours).
+These include low MOI (0.0035 TCID50 per cell is quite low) and short growth times to avoid accumulation of defective particles (the helper-virus rescue was allowed to propagate for just 24 hours, and the low-MOI passage for just 48 hours).
 
@@
 
-In fact, like the reviewer, we initially thought that a media change to wash aways the inoculum might be important. 
+In fact, like the reviewer, we initially thought that a media change to wash away the inoculum might be important. 
 Therefore, in our experiments, we performed such a media change to remove the initial inoculum for biological replicates 1 and 2, and for one of the technical replicates (3-2) of biological replicate 3.
-We did not perform such a change for the other technical replicate (3-2) of biological replicate 3.
+We did not perform such a change for the other technical replicate (3-1) of biological replicate 3.
 It turns out that the media change makes no discernible difference, which is why we neglected to expand on this point in the original manuscript. 
@@
 We have updated the Methods to now include this detail, and (as mentioned above) also now include our full lab notes so that all other details can be reviewed.
-Specifically, see \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/lab_notes/Perth09_HA_Library_Rescue.pdf} (relevant details of the notebook can be found in pages 21-25)
+Specifically, see \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/lab_notes/Perth09_HA_Library_Rescue.pdf} (relevant details of the notebook can be found in pages 21-25).
 }
@@
 \response{For the reasons described above, it is not necessary to repeat the experiments.
-We do not deny that there are slightly different ways that the experiments could be done, but the in end, the ``proof is in the pudding.'' 
+We do not deny that there are slightly different ways that the experiments could be done, but in the end, the ``proof is in the pudding.'' 
 In the manuscript, we show that the measurements made using the experimental parameters that we chose lead to useful data as judged by the following criterion:
@@
 We did not test the growth of these clones. 
-As explained in Figure~S2, these were plasmid variants chosen randomly to Sanger sequence to evaluate the mutation rate prior to creating virus libraries; they were not evaluated as viruses.
-The new Figures S5 to S7 now show the growth phenotype of other selected mutants, one of which includes a start codon mutant.
+As explained in Figure~S2, these were plasmid variants chosen randomly to be Sanger sequenced to evaluate the mutation rate prior to creating virus libraries; they were not evaluated as viruses.
+The new Figures S4 to S6 now show the growth phenotype of other selected mutants, one of which includes a start codon mutant.
 }
@@
 
-The suggestion to make a version of Fig.~4 that distinguished between the stalk and the head is a good one.
+The suggestion to make a version of Fig.~4 that distinguished between head and stalk mutations is a good one.
 It is difficult to also overlay the structural location (stalk versus head) on the current Fig.~4, since colors are already used to indicate mutational effects.
@@
 
-\response{We agree that this would be a very interesting experiment to perform.
+\response{We agree that these would be very interesting experiments to perform.
 However, such experiments would be far beyond the scope of our study.
-As stated in the Abstract and Significance Statement, the main goal of our study is to perform deep mutational scanning of a recent H3 HA and use the data to better predict H3N2 viral evolution in nature.
+As stated in the Abstract and Significance Statement, the main goal of our study is to perform deep mutational scanning of a recent H3 HA and to use the data to better predict H3N2 viral evolution in nature.
 A large set of additional experiments comparing stalk- and head-specific antibodies in H3 and H1 HAs is clearly beyond the scope of this goal.
@@
 \item We have now included an explanation of dN/dS $\ll$1 for conventional substitution models.
-\item We now describe how the ``ExpCM, site avg.'' substitution model is a control that shows that the improvements are due to the site-specific nature of the data, since this model averages data across sites.
+\item We now describe how the ``ExpCM, site avg.'' substitution model is a control that shows that the improvements of the ExpCM are due to the site-specific nature of the data, since the ``ExpCM, site avg.'' model averages data across sites.
 \end{itemize}

### 2018-06-24 paper/reviewer_response_PNAS.tex
slight tweaks to reviewer response

@@
 
-We thank the reviewers for the careful reading and suggestions about the manuscript.
-We were gratified that the reviewers recognized the quality of our experimental work (one of the reviewers said, ``the experimental setup is sound and there is strong correlation among technical and biological replicates.'').
+We thank the reviewers for the careful reading of our manuscript.
+We were gratified that they recognized the quality of our work: one of the reviewers said, ``the experimental setup is sound and there is strong correlation among technical and biological replicates.''
 
-We also appreciate the reviewer's recognition of the important step our work takes in linking experiments with quantitative models of influenza virus evolution, and their recognition that we adequately discuss the caveats as well as the strengths of our approach.
+The main contribution of our work is to use show that deep mutational scanning of an H3 HA provides information that is informative about the evolutionary fates of H3N2 influenza virus variants in nature.
+We appreciated the reviewers' recognition of the important step our work takes in linking experiments with quantitative models of influenza virus evolution, and their recognition that we carefully discuss the caveats as well as the strengths of our approach.
+Reviewer \#1 did make an important suggestion to look at HA sequences only from unpassaged virus isolates---in the revised version we have done that, and found that our data are even more predictive of evolutionary fate for these higher quality sequences.
 
-In the revised version, we have performed additional experiments and analyses that address all of the reviewer questions and concerns.
+Most of the reviewer questions had to do with more tangential aspects of our work, such as validation of specific mutations and phylogenetic analyses.
+In the revised version, we have addressed all of these questions.
+Therefore, we think that you will find that the revisions have further strengthened the paper.
 
@@
 As we note in the manuscript, this finding is concordant with prior work on suggesting additional translation initiation sites sites in H2 HAs.
-Because translation initiation of HA is not the focus of our manuscript, we have not performed further investigation of the exact mechanism or identity of the alternate start sites.
-However, the data we have added clearly show that the canonical ATG is not strictly necessary for viral replication.
+Because translation initiation of HA is not the focus of our manuscript, we have not investigated the exact mechanism or identity of the alternate start sites.
+However, the data we have added clearly show that the canonical ATG is not strictly necessary for viral growth.
 }
@@
 
-\response{This is a valid point, and we have now also individually evaluated the reverse-genetics titers of three glycosylation site mutants.
+\response{We have now also individually evaluated the reverse-genetics titers of three glycosylation site mutants.
 Two of the mutants, T24F and T40V, remove the aforementioned highly conserved but mutationally tolerant NGT and NAT glycosylation sites near the start of HA1.
@@
 One possible explanation is simply that cell culture does not fully capture the constraints on HA in nature.
-But a more interesting possibility is that conservation in nature is not expected to strictly parallel mutational tolerance.
+But a more interesting possibility is that conservation in nature does not strictly parallel mutational tolerance.
 The tendency for a site to change in nature depends on how much selection there is for the mutations.
@@
 But at the suggestion of Reviewer \#1, we have now also included a post-Perth analysis using \emph{only} HA sequences from unpassaged viral isolates (there are not enough unpassaged isolates from before 2009 to perform a similar unpassaged pre-Perth analysis).
-As shown in the new Figure 5, the correlation for the unpassaged post-Perth is actually higher than either other set, demonstrating that there is not a systematic trend for poorer performance in prospective analyses.
+As shown in the new Figure 5, the correlation for the unpassaged post-Perth is actually higher than either other set, suggesting that there may not a systematic trend for poorer performance in prospective analyses.
 }

### 2018-06-24 paper/manuscript.tex
filler spots for additional codon mutations

@@
 When growing influenza virus in cell culture, trypsin is normally added to cleave HA into its mature form.
-To obviate the need for trypsin, we engineered MDCK-SIAT1 cells and MDCK-SIAT1-CMV-PB1 cells to constitutively express the TMPRSS2 protease, which cleaves and activates HA in the human airways~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
+To obviate the need for trypsin, we engineered MDCK-SIAT1 cells and MDCK-SIAT1-CMV-PB1~\cite{bloom2010permissive} cells to constitutively express the TMPRSS2 protease, which cleaves and activates HA in the human airways~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
 The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656) and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry to create plasmid pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
@@
 {\bf Validation that the Perth/2009 HA is somewhat tolerant to mutation of the canonical start codon.}
-We created variants of the Perth/2009 HA in which the canonical start codon was mutated (amino-acid mutation M(-16)K, codon mutation ATG(-16)AAA), a conserved cysteine was mutated to alanine (C52A, codon mutation TGC52GCA), and the same cysteine was synonymously mutated (C52C, codon mutation TGC52TGT).
+We created variants of the Perth/2009 HA in which the canonical start codon was mutated (amino-acid mutation M(-16)K, codon mutation ATG$\rightarrow$AAA), a conserved cysteine was mutated to alanine (C52A, codon mutation TGC$\rightarrow$GCA), and the same cysteine was synonymously mutated (C52C, codon mutation TGC$\rightarrow$TGT).
 We selected these mutations because our deep mutational scanning results in Figure~\ref{fig:logoplot} surprisingly suggest that the start codon (position -16) is fairly tolerant of mutations, but that site 52 is highly intolerant of mutation to anything other than cysteine.
@@
 We created variants of the Perth/2009 HA with mutations that eliminated N-linked glycosylation motifs (Asn-Xaa-Ser/Thr) at asparagine residues 22, 38, and 285 (these are mutations T24F, T40V, and S287A, respectively).
+\comment{The codon mutations were NNN$\rightarrow$NNN, NNN$\rightarrow$NNN, and NNN$\rightarrow$NNN, respectively.}
 We then used reverse genetics to generate viruses carrying the wildtype HA or each of these mutants.
@@
 {\bf Validation of the mutational tolerance of a site in the transmembrane domain.}
-We created a variant of the Perth/2009 HA with a transmembrane domain mutation, C199(HA2)K, at a site that our deep mutational scanning suggests should be highly mutationally tolerant (Figure~\ref{fig:logoplot}).
+We created a variant of the Perth/2009 HA with a transmembrane domain mutation, C199(HA2)K (\comment{codon mutation NNN$\rightarrow$NNN}), at a site that our deep mutational scanning suggests should be highly mutationally tolerant (Figure~\ref{fig:logoplot}).
 We then used reverse genetics to generate viruses carrying the wildtype HA or this mutant.

### 2018-06-24 paper/manuscript.tex
Added codon details

@@
 We created variants of the Perth/2009 HA with mutations that eliminated N-linked glycosylation motifs (Asn-Xaa-Ser/Thr) at asparagine residues 22, 38, and 285 (these are mutations T24F, T40V, and S287A, respectively).
-\comment{The codon mutations were NNN$\rightarrow$NNN, NNN$\rightarrow$NNN, and NNN$\rightarrow$NNN, respectively.}
+The codon mutations were ACG$\rightarrow$TTC, ACT$\rightarrow$GTA, and AGC$\rightarrow$GCG, respectively.
 We then used reverse genetics to generate viruses carrying the wildtype HA or each of these mutants.
@@
 {\bf Validation of the mutational tolerance of a site in the transmembrane domain.}
-We created a variant of the Perth/2009 HA with a transmembrane domain mutation, C199(HA2)K (\comment{codon mutation NNN$\rightarrow$NNN}), at a site that our deep mutational scanning suggests should be highly mutationally tolerant (Figure~\ref{fig:logoplot}).
+We created a variant of the Perth/2009 HA with a transmembrane domain mutation, C199(HA2)K (codon mutation TGT$\rightarrow$AAG), at a site that our deep mutational scanning suggests should be highly mutationally tolerant (Figure~\ref{fig:logoplot}).
 We then used reverse genetics to generate viruses carrying the wildtype HA or this mutant.

### 2018-06-26 paper/manuscript.tex
make track changes version
Update manuscript text around updated results

@@
 We show that these experimental measurements can help discriminate evolutionarily successful mutations from those found in strains that quickly die out.
-However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are not useful for understanding the evolutionary fate of H3 viral strains.
+However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are less informative for understanding the evolutionary fate of H3 viral strains.
 
@@
 \caption{\label{fig:muteffect_maxfreq_WSN}
-{\bf Experimental measurements on an H1 HA are not informative about the evolutionary fate of H3N2 mutations.}
+{\bf Experimental measurements on an H1 HA are less informative about the evolutionary fate of H3N2 mutations.}
 This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
@@
 
-\subsection*{Measurements made on an H1 HA are not useful for understanding the evolution of H3 influenza}
+\subsection*{Measurements made on an H1 HA are less informative for understanding the evolution of H3 influenza}
 To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (Figure~\ref{suppfig:WSNprefs_logoplot}), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
-Figure~\ref{fig:muteffect_maxfreq_WSN} shows that there is no correlation between the H1 experimental measurements and maximum frequency that mutations reach during H3N2 viral evolution, and only a weak correlation with mutations from unpassaged viral isolates.
+Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the correlations between the H1 experimental measurements and the maximum frequency that mutations reach during H3N2 viral evolution are consistently weaker than those using H3 experimental measurements (Figure~\ref{fig:muteffect_maxfreq}).
 Therefore, the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
@@
 
-\matmethods{
+{\small
+\section*{Materials and Methods}
 \subsection*{Data and computer code}
@@
 At 16 hours post-infection, we selected wells with 1 to 10\% of cells that were GFP-positive and analyzed the fraction of GFP-positive cells by flow cytometry to calculate the titer of infectious particles per $\mu$L.
-
 }
-
-\showmatmethods{} % Display the Materials and Methods section
 

### 2018-06-26 paper/manuscript.tex
tweaks to merge H1 corr update

@@
 To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (Figure~\ref{suppfig:WSNprefs_logoplot}), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
-Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the correlations between the H1 experimental measurements and the maximum frequency that mutations reach during H3N2 viral evolution are consistently weaker than those using H3 experimental measurements (Figure~\ref{fig:muteffect_maxfreq}).
+Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the correlations between the H1 experimental measurements and the maximum frequency that mutations reach during H3N2 viral evolution are consistently weaker than those using H3 experimental measurements (compare Figure~\ref{fig:muteffect_maxfreq_WSN} to Figure~\ref{fig:muteffect_maxfreq}A).
 Therefore, the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
@@
 \subsection*{There are large differences between H3 and H1 HAs in the amino-acid preferences of many sites}
-An obvious hypothesis for why the H1 deep mutational scanning is not useful for understanding the evolution of H3N2 influenza viruses is that the effect of the same mutation is often different between these two HA subtypes.
+An obvious hypothesis for why the H1 deep mutational scanning is less useful for understanding the evolution of H3N2 influenza viruses is that the effect of the same mutation is often different between these two HA subtypes.
 To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.

### 2018-06-27 paper/reviewer_response_PNAS.tex
Minor tweaks to reviewer response

@@
 
-The main contribution of our work is to use show that deep mutational scanning of an H3 HA provides information that is informative about the evolutionary fates of H3N2 influenza virus variants in nature.
+The main contribution of our work is to show that deep mutational scanning of an H3 HA provides information that is informative about the evolutionary fates of H3N2 influenza virus variants in nature.
 We appreciated the reviewers' recognition of the important step our work takes in linking experiments with quantitative models of influenza virus evolution, and their recognition that we carefully discuss the caveats as well as the strengths of our approach.
@@
 But at the suggestion of Reviewer \#1, we have now also included a post-Perth analysis using \emph{only} HA sequences from unpassaged viral isolates (there are not enough unpassaged isolates from before 2009 to perform a similar unpassaged pre-Perth analysis).
-As shown in the new Figure 5, the correlation for the unpassaged post-Perth is actually higher than either other set, suggesting that there may not a systematic trend for poorer performance in prospective analyses.
+As shown in the new Figure 5, the correlation for the unpassaged post-Perth is actually higher than either other set, suggesting that there may not be a systematic trend for poorer performance in prospective analyses.
 }

### 2018-06-28 paper/manuscript.tex
edited dN/dS text per Trevor suggestions

@@
 A substitution model in which the amino-acid preferences were averaged across all sites (ExpCM, site avg.) performs no better than conventional substitution models, demonstrating that the reason that our measurements are informative is because they capture site-specific constraints.
-The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}), since the only way that these models can account for the constraints on HA evolution is via the heuristic use of the $\omega$ parameter to depress the rates of all nonysynonymous substitutions.
+The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
 However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for in the deep mutational scanning.

### 2018-06-29 paper/SI.tex
split manuscript and supporting information

--- 2018-06-28 paper/manuscript.tex
@@
 \usepackage{newfloat}
+\usepackage{xr-hyper}
+\externaldocument{manuscript_noSI}
 \DeclareFloatingEnvironment[name={Figure}]{suppfigure}
@@
 
-\title{Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants}
-
-% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[a,d,e,1]{Juhye M. Lee}
-\author[b,f,1]{John Huddleston}
-\author[a,d,e]{Michael B. Doud}
-\author[a,f]{Kathryn A. Hooper}
-\author[g]{Nicholas C. Wu}
-\author[b,c,1]{Trevor Bedford}
-\author[a,c,d,1]{Jesse D. Bloom}
-
-\affil[a]{Basic Sciences Division}
-\affil[b]{Vaccine and Infectious Disease Division}
-\affil[c]{and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
-\affil[d]{Department of Genome Sciences}
-\affil[e]{Medical Scientist Training Program}
-\affil[f]{and Molecular and Cellular Biology Program, University of Washington, Seattle, WA 98195, USA}
-\affil[g]{Department of Integrative Structural and Computational Biology, The Scripps Research Institute, La Jolla, CA  92037}
-
 % Please give the surname of the lead author for the running footer
 \leadauthor{Lee}
-
-% Please add here a significance statement to explain the relevance of your work
-% 120 words max
-\significancestatement{
-A key goal in the study of influenza virus evolution is to forecast which viral strains will persist and which ones will die out.
-Here we experimentally measure the effects of all amino-acid mutations to the hemagglutinin protein from a human H3N2 influenza strain on viral growth in cell culture.
-We show that these measurements have utility for distinguishing among viral strains that do and do not succeed in nature.
-Overall, our work suggests that new high-throughput experimental approaches may be useful for understanding virus evolution in nature.
-}
-
-% Please include corresponding author, author contribution and author declaration information
-\authorcontributions{J.M.L.
-performed the experiments; J.M.L.
-and J.H.
-analyzed data with input from T.B.
-and J.D.B.; M.B.D.
-contributed essential reagents and expertise; K.A.H.
-contributed essential reagents; N.C.W.
-made key intellectual contributions; J.M.L.
-and J.D.B.
-wrote the paper with input from J.H.
-and T.B.; all authors were involved in project design and revising the paper.}
-\authordeclaration{The authors declare no conflicts of interest.}
-\equalauthors{\textsuperscript{1} J.M.L and J.H. contributed equally to this work.}
-\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed: tbedford@fredhutch.org, jbloom@fredhutch.org}
-
-% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
-\keywords{influenza virus $|$ hemagglutinin $|$ deep mutational scanning $|$ antigenic drift $|$ epistasis}
-
-\begin{abstract}
-% no more than 250 words.
-Human influenza virus rapidly accumulates mutations in its major surface protein hemagglutinin (HA).
-The evolutionary success of influenza virus lineages depends on how these mutations affect HA's functionality and antigenicity.
-Here we experimentally measure the effects on viral growth in cell culture of all single amino-acid mutations to the HA from a recent human H3N2 influenza virus strain.
-We show that mutations that are measured to be more favorable for viral growth are enriched in evolutionarily successful H3N2 viral lineages relative to mutations that are measured to be less favorable for viral growth.
-Therefore, despite the well-known caveats about cell-culture measurements of viral fitness, such measurements can still be informative for understanding evolution in nature.
-We also compare our measurements for H3 HA to similar data previously generated for a distantly related H1 HA, and find substantial differences in which amino acids are preferred at many sites.
-For instance, the H3 HA has less disparity in mutational tolerance between the head and stalk domains than the H1 HA.
-Overall, our work suggests that experimental measurements of mutational effects can be leveraged to help understand the evolutionary fates of viral lineages in nature --- but only when the measurements are made on a viral strain similar to the ones being studied in nature.
-\end{abstract}
 
@@
 \verticaladjustment{-2pt}
-
-\maketitle
-\thispagestyle{firststyle}
-\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
-
-\dropcap{S}easonal H3N2 influenza virus evolves rapidly, fixing 3 to 4 amino-acid mutations per year in its hemagglutinin (HA) surface protein~\cite{fitch1997long, bhatt2011genomic}.
-Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping}.
-This evolution is further characterized by competition and turnover among groups of strains called clades bearing different complements of mutations~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,koelle2015effects,bedford2015global}.
-Clades vary widely in their evolutionary success, with some dying out soon after emergence and others going on to take over the virus population.
-Several lines of evidence indicate that successful clades have higher fitness than clades that remain at low frequency~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,luksza2014predictive}.
-A key goal in the study of H3N2 evolution is to identify the features that enable certain clades to succeed as others die out.
-
-\begin{SCfigure*}[\sidecaptionrelwidth][t]
-\centering
-\includegraphics[width=12cm]{figs/dms_overview/dms_overview.pdf}
-\caption{\label{fig:dms_overview}
-{\bf Deep mutational scanning of the Perth/2009 H3 HA.}
-(A) We generated mutant virus libraries using a helper-virus approach~\cite{doud2016accurate}, and passaged the libraries at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants.
-Deep sequencing of the variants before and after selection allowed us to estimate each site's amino-acid preferences.
-(B) The experiments were performed in full biological triplicate.
-We also passaged and deep sequenced library 3 in duplicate.
-(C) Frequencies of nonsynonymous, stop, and synonymous mutations in the mutant plasmid DNA, the passaged mutant viruses, and wildtype DNA and virus controls.
-(D) The Pearson correlations among the amino-acid preferences estimated in each replicate.
-}
-\end{SCfigure*}
-
-Two main characteristics distinguish evolutionarily successful clades from their competitors: greater antigenic change, and efficient viral growth and transmission.
-In principle, experiments could be informative for identifying how mutations affect these features.
-Most work on influenza evolution to date has utilized experimental data to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction,koel2013substitutions,chambers2015identification,li2016selection}.
-However, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
-Specifically, due to influenza virus's high mutation rate~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and lack of intra-segment recombination~\cite{boni2008homologous}, deleterious mutations become linked to beneficial ones.
-The resulting accumulation of deleterious mutations can affect non-antigenic properties central to viral fitness~\cite{luksza2014predictive}.
-However, there are no large-scale quantitative characterizations of how mutations to H3N2 HA affect viral growth.
-
-It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of all single amino-acid mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high,haddox2018mapping}.
-However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
-Here, we measure the effects on viral growth in cell culture of all mutations to the HA of a recent human H3N2 strain.
-We show that these experimental measurements can help discriminate evolutionarily successful mutations from those found in strains that quickly die out.
-However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are less informative for understanding the evolutionary fate of H3 viral strains.
-
-\section*{Results}
-\label{sec:results}
-
-\subsection*{Deep mutational scanning of HA from a recent strain of human H3N2 influenza virus}
-We performed a deep mutational scan to measure the effects of all amino-acid mutations to HA from the A/Perth/16/2009 (H3N2) strain on viral growth in cell culture.
-This strain was the H3N2 component of the influenza vaccine from 2010-2012~\cite{who2010d,who2011}.
-Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral growth in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and Dataset~\ref{suppdata:PerthHA}).
-The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.
-
-We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (Figure~\ref{suppfig:SangerSeq}).
-We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
-These mutant viruses derived all their non-HA genes from the lab-adapted A/WSN/1933 strain.
-Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
-To further increase viral titers, we used MDCK-SIAT1 cells (Madin-Darby canine kidney cells overexpressing 2,6-sialyltransferase)~\cite{matrosovich2003overexpression} that we engineered to constitutively express TMPRSS2 (Transmembrane Protease, Serine 2), which cleaves the HA precursor to activate it for membrane fusion~\cite{bottcher2006proteolytic, bottcher2010cleavage}.
-
-After generating the mutant virus libraries, we passaged them at low multiplicity of infection (MOI) in cell culture to create a genotype-phenotype link and select for functional HA variants (Figure~\ref{fig:dms_overview}A).
-All experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B).
-We also passaged and deep sequenced library 3 in duplicate (library 3-1 and 3-2) to gauge experimental noise \textit{within} a single biological replicate.
-As a control to measure sequencing and mutational errors, we used the unmutated HA gene to generate and passage viruses carrying wildtype HA.
-
-Deep sequencing of the initial plasmid mutant libraries and the passaged mutant viruses revealed selection for functional HA mutants.
-Specifically, stop codons were purged to 20-45\% of their initial frequencies after correcting for error rates estimated by sequencing the wildtype controls (Figure~\ref{fig:dms_overview}C).
-The incomplete purging of stop codons is likely because genetic complementation due to co-infection~\cite{marshall2013influenza, brooke2013most} enabled the persistence of some virions with nonfunctional HAs.
-We also observed selection against many nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), with their frequencies falling to 30-40\% of their initial values after error correction.
-
-\begin{table}[b!]
-\caption{\label{tab:phydms}
-{\bf Substitution models informed by the experiments describe HA's evolution better than traditional models.}}
-\begin{center}
-\begin{tabular}{cccccccc}
-\hline
-\bf{Model} & \bf{$\Delta$AIC} & \bf{LnL} & \bf{Stringency} & \bf{$\omega$}  \\ \hline
-ExpCM & 0.0 & -8441 & 2.47 & 0.91 \\
-GY94 M5 & 2094 & -9482 & -- & 0.36 (0.30, 0.84) \\
-ExpCM, site avg. & 2501 & -9692 & 0.67 & 0.32 \\
-GY94 M0 & 2536 & -9704 & -- & 0.31 \\
-\hline
-\end{tabular}
- \end{center}
-\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites (site avg.), and M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
-Models are compared by Akaike information criterion (AIC)~\cite{posada2004model} computed from the log likelihood (LnL) and number of model parameters.
-The $\omega$ parameter is dN/dS for the Goldman-Yang models, and the relative dN/dS after accounting for the measurements for the ExpCM.
-For the M5 model, we give the mean followed by the shape and rate parameters of the gamma distribution over $\omega$.
-}
-\end{table}
-
-\begin{figure*}[ht]
-\centering
-\includegraphics[width=17cm]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}
-\caption{\label{fig:logoplot}
-{\bf The site-specific amino-acid preferences of the Perth/2009 HA measured in our experiments.}
-The height of each letter is the preference for that amino acid, after taking the average over experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter in Table~\ref{tab:phydms}.
-The sites are in H3 numbering.
-The top overlay bar indicates whether or not a site is in the set of epitope residues delineated in \cite{wolf2006long}.
-The bottom overlay bar indicates the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
-The letters directly above each logo stack indicate the wildtype amino acid at that site.
-}
-\end{figure*}
-
-We next quantified the reproducibility of our deep mutational scanning across biological and technical replicates.
-We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids as described in~\cite{bloom2015software}.
-Because there are 566 residues in HA, there are $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site sum to one~\cite{bloom2015software}).
-The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
-The biological replicates were well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78.
-Replicate 1 exhibited the weakest correlation with other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
-The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity during viral passage contributes most of the experimental noise.
-
-\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
-How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
-This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard substitution models~\cite{bloom2017identification,hilton2017phydms}.
-Table~\ref{tab:phydms} shows that an ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
-This result indicates that our experiments authentically capture some of the constraints on HA evolution.
-A substitution model in which the amino-acid preferences were averaged across all sites (ExpCM, site avg.) performs no better than conventional substitution models, demonstrating that the reason that our measurements are informative is because they capture site-specific constraints.
-The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
-However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for in the deep mutational scanning.
-The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
-Throughout the rest of this paper, we use the amino-acid preferences re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
-These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.
-
-%\begin{SCfigure*}[\sidecaptionrelwidth][t]
-\begin{figure*}[ht]
-\centering
-\includegraphics[width=0.95\textwidth]{figs/mut_tolerance/mut_tolerance_new.pdf}
-\caption{\label{fig:mut_tolerance}
-{\bf Mutational tolerance of each site in H3 and H1 HAs.}
-(A) Mutational tolerance as measured in the current study is mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}).
-Mutational tolerance of the WSN/1933 H1 HA as measured in \cite{doud2016accurate} is mapped onto the structure of the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}).
-Different color scales are used because measurements are comparable among sites within the same HA, but not necessarily across HAs.
-Both trimers are shown in the same orientation.
-For each HA, the structure at left shows a surface representation of the full trimer, while the structure at right side shows a ribbon representation of just one monomer.
-The sialic acid receptor is shown in red sticks.
-(B) The mutational tolerance of solvent-exposed residues in the head and stalk domains of the Perth/2009 H3 HA (purple) and WSN/1933 H1 HA (gold).
-Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
-A residue was classified as solvent exposed if its relative solvent accessibility was $\geq$ 0.2; Figure~\ref{suppfig:head_stalk_RSA} shows that the results are robust to the choice of solvent accessibility cutoff.
-Note that the mutational tolerance values are not comparable between the two HAs, but are comparable between domains of the same HA.
-}
-\end{figure*}
-
-Examination of Figure~\ref{fig:logoplot} reveals that the experimentally measured amino-acid preferences generally agree with existing knowledge about HA's structure and function.
-For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} strongly prefer cysteine.
-At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such as Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change,yang2015structure}.
-A positively charged amino acid at site 329 is important for cleavage of the HA0 precursor into the mature form~\cite{stech2005new}, and this site strongly prefers arginine.
-
-However, there are also some differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences (Figure~\ref{suppfig:diffprefs_logoplot}).
-Most surprisingly, the start codon does not show a particularly strong preference for methionine (Figure~\ref{fig:logoplot}).
-We validated that a virus carrying a mutation at this site from methionine to lysine does in fact reach appreciable titers (Figure~\ref{suppfig:startcodon_validation}), perhaps because of alternative translation-initiation at a downstream or upstream start site as has been described for other HAs~\cite{girard2011upstream}.
-Our measurements also suggest mutational tolerance at some other sites that are relatively conserved among natural HAs, such as the N-linked glycosylation motifs near the beginning of HA1 and the transmembrane domain (Figure~\ref{fig:logoplot}).
-We validated that viruses with mutations to the glycosylation motifs at sites 22 or 38, or a site in the transmembrane domain, do in fact grow to high titers (Figures~\ref{suppfig:glycan_validation} and \ref{suppfig:TM_validation}, respectively).
-The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function in nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.
-
-\subsection*{There is less difference in mutational tolerance between the HA head and stalk domains for H3 than for H1}
-Our experiments measure which amino acids are tolerated at each HA site under selection for viral growth.
-We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we quantify as the Shannon entropy of the re-scaled preferences.
-In prior mutational studies of H1 HAs, the stalk domain was found to be substantially less mutationally tolerant than the globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,heaton2013genome}.
-
-We performed a similar analysis using our new data for the Perth/2009 H3 HA.
-Surprisingly, the head domain of the H3 HA is \emph{not} more mutationally tolerant than the stalk domain (Figure~\ref{fig:mut_tolerance}).
-Specifically, whereas solvent-exposed residues in the head domain are substantially more mutationally tolerant than those in the stalk domain for the WSN/1933 H1 HA, the trend is actually reversed for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
-This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (Figure~\ref{suppfig:head_stalk_RSA}).
-For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
-Helix A forms part of the epitope of many broadly neutralizing anti-stalk antibodies~\cite{mallajosyula2014influenza,laursen2013broadly,chai2016two}.
-
-We also see high mutational tolerance in many of the known antigenic regions of H3 HA~\cite{wiley1981structural}.
-For instance, antigenic region B is an immunodominant area, and many recent major antigenic drift mutations have occurred in this region~\cite{chambers2015identification,koel2013substitutions,popova2012immunodominance}.
-We find that the most distal portion of the globular head near the 190-helix, which is part of antigenic region B, is highly tolerant of mutations (Figure~\ref{fig:mut_tolerance}A).
-Antigenic region C is also notably mutationally tolerant.
-
-Many residues inside HA's receptor binding pocket are known to be highly functionally constrained~\cite{wilson1981structure,martin1998studies}, and our data indicates that these sites are relatively mutationally intolerant in both H3 and H1 HAs (Figure~\ref{fig:mut_tolerance}A).
-In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, since mutations at these sites can have large effects on antigenicity~\cite{wiley1981structural,koel2013substitutions}.
-
-\begin{figure}[t!]
-\centering
-\includegraphics[width=0.95\linewidth]{figs/frequency_trajectory_ex/frequency_trajectory_ex.pdf}
-\caption{\label{fig:frequency_trajectory}
-{\bf Frequency trajectories of individual mutations and their relation to the experimentally measured effects of these mutations.}
-The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} from 2004 to 2014.
-Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
-The Perth/2009 strain is labeled with a star, and nodes in the clade containing the Perth/2009 strain were excluded from our analyses.
-The bottom panel shows the frequency trajectory of each mutation, with trajectories colored according to the mutational effects as in the top panel.
-It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
-Figure~\ref{suppfig:head_stalk_mutfreq_supp} shows a similar layout but colors mutations by whether they are in HA's head or stalk domain.
-}
-\end{figure}
-
-\begin{SCfigure*}[\sidecaptionrelwidth][h!]
-\centering
-\includegraphics[width=12cm]{figs/muteffect_by_maxfreq/muteffect_by_maxfreq.pdf}
-\caption{\label{fig:muteffect_maxfreq}
-{\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
-(A) Correlation between the effects of mutations as measured in our deep mutational scanning of the Perth/2009 HA and the maximum frequency reached by these mutations in nature.
-The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations of the experimental measurements for which the permuted $\rho$ was greater than or equal to the observed $\rho$.
-(B) The distribution of mutational effects partitioned by maximum mutation frequency.
-The vertical black line shows the mean mutation effect for each category.
-The analysis is performed separately for pre-Perth/2009, post-Perth/2009, and unpassaged isolates from the post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
-}
-\end{SCfigure*}
-
-\subsection*{Our measurements can help distinguish between mutations that reach low and high frequencies in nature}
-Mutations occurring in the H3N2 virus population experience widely varying evolutionary fates (Figure~\ref{fig:frequency_trajectory}).
-Some mutations appear, spread and fix in the population, while others briefly circulate before disappearing.
-We take the maximum frequency reached by a mutation as a coarse indicator of its effect on fitness, since favorable mutations generally reach higher frequencies than unfavorable ones~\cite{ewens2012mathematical}.
-Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, e.g.\ although R142G occurs multiple times on the phylogeny we track each of these mutations occurring on different backgrounds separately.
-As such, each mutation is shown as a separate circle on a separate branch in Figure \ref{fig:frequency_trajectory}.
-However, because multiple mutations on the same phylogeny branch cannot be disentangled, when multiple mutations occurred on a single branch, we assigned a single mutational effect based on the sum of effects of each mutation.
-
-After annotating mutations and their frequencies on the phylogeny in this way (Figure~\ref{fig:frequency_trajectory}), it is visually obvious that there are relatively few circulating mutations that we measure to be strongly deleterious---and that such deleterious mutations rarely reach high frequency when they do occur.
-
-We next sought to quantify the correlation between a mutation's experimentally measured effect and the maximum frequency it attained during natural evolution.
-To calculate a given mutation's effect, we simply took the logarithm of the ratio of the preferences for the mutant and wildtype amino acids at that site.
-To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
-We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after, since these mutations have not had enough time for their evolutionary fates to be fully resolved.
-We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
-We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
-In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.15$).
-The similar effect sizes for both the pre- and post-Perth partitions shows that our experimental measurements can help explain the evolutionary fates of mutations in strains that post-date the experimentally studied strain, as well as to retrospectively analyze mutations that precede the experimental strain.
-
-Many of the HAs in sequence databases are from viral isolates that were passaged in cell-culture or eggs, which can cause lab-adaptation mutations that confound evolutionary analyses~\cite{mcwhite2016sequence}.
-To check that our results were robust to such lab-adaptation mutations, we repeated our analysis using only HA sequences derived from viruses that had not been passaged in the lab.
-Because sequencing of unpassaged primary isolates has only recently become commonplace, we could only perform this analysis for the post-Perth partition of the phylogenetic tree.
-Figure~\ref{fig:muteffect_maxfreq}A shows that the correlation between our measured mutational effects and the maximum frequency was even stronger for mutations from unpassaged viral isolates ($\rho = 0.24$).
-
-\begin{SCfigure*}[\sidecaptionrelwidth][t]
-\centering
-\includegraphics[width=12cm]{figs/muteffect_by_maxfreq_WSN/muteffect_by_maxfreq_WSN.pdf}
-\caption{\label{fig:muteffect_maxfreq_WSN}
-{\bf Experimental measurements on an H1 HA are less informative about the evolutionary fate of H3N2 mutations.}
-This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
-Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
-The empirical $P$-value represents the result of 1,000 permutations.
-}
-\end{SCfigure*}
-
-The trends in Figure~\ref{fig:muteffect_maxfreq}A are most strongly driven by the behavior of substantially deleterious mutations.
-We investigated this further by partitioning mutations into those that reach low, medium and high frequencies, and those that fix in the population (Figure~\ref{fig:muteffect_maxfreq}B).
-The mutations that reach higher frequencies have a more favorable mean effect.
-Mutations measured to be substantially deleterious almost never reach high frequency.
-Overall, these results demonstrate that measurements of how mutations affect viral growth in cell culture are informative for understanding the fates of these mutations in nature: in particular, if a mutation is measurably deleterious to viral growth, that mutation is unlikely to prosper in nature.
-
-\subsection*{Measurements made on an H1 HA are less informative for understanding the evolution of H3 influenza}
-To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (Figure~\ref{suppfig:WSNprefs_logoplot}), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
-Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the correlations between the H1 experimental measurements and the maximum frequency that mutations reach during H3N2 viral evolution are consistently weaker than those using H3 experimental measurements (compare Figure~\ref{fig:muteffect_maxfreq_WSN} to Figure~\ref{fig:muteffect_maxfreq}A).
-Therefore, the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
-
-\begin{SCfigure*}[\sidecaptionrelwidth][b]
-\centering
-\includegraphics[width=12cm]{figs/distance_distribution/distance_distribution.pdf}
-\caption{\label{fig:distance_distribution}
-{\bf There are large shifts in the effects of mutations between H1 and H3 HAs.}
-(A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled.
-(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study and the three replicates in prior deep mutational scanning of the H1 HA~\cite{doud2016accurate}.
-Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray.
-$R$ indicates the Pearson correlation coefficient.
-(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2018mapping}, and plotted the distribution of shifts for all sites.
-The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
-The shifts are also much larger than those previously observed between two variants of HIV Env that share 86\% amino-acid identity (pink).
-However, the shifts between H3 and H1 are less than the differences between HA and HIV Env (green).
-}
-\end{SCfigure*}
-
-\subsection*{There are large differences between H3 and H1 HAs in the amino-acid preferences of many sites}
-An obvious hypothesis for why the H1 deep mutational scanning is less useful for understanding the evolution of H3N2 influenza viruses is that the effect of the same mutation is often different between these two HA subtypes.
-To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
-Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2018mapping}.
-However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
-One simple way to investigate the extent of shifts in amino-acid preferences is to correlate measurements from independent deep mutational scanning replicates on H1 and H3 HAs.
-Figure~\ref{fig:distance_distribution}B shows that replicate measurements on the same HA variant are more correlated than those on different HA variants.
-
-To more rigorously quantify shifts in amino-acid preferences after correcting for experimental noise, we used the statistical approach in~\cite{doud2015site,haddox2018mapping}.
-Figure~\ref{fig:distance_distribution}C shows the distribution of shifts in amino-acid preferences between H3 and H1 HAs after correcting for experimental noise.
-Although some sites have small shifts near zero, many sites have large shifts.
-These shifts between H3 and H1 are much larger than expected from the null distribution that would be observed purely from experimental noise.
-They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2018mapping}.
-However, the typical shift between H3 and H1 is still smaller than that observed when comparing HA to the non-homologous HIV Env protein.
-Therefore, there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations remain more similar than for non-homologous proteins.
-
-\subsection*{Properties associated with the shifts in amino-acid preferences between H3 and H1 HAs}
-What features distinguish the sites with shifted amino-acid preferences between H3 and H1 HAs?
-The sites of large shifts do not obviously localize to one specific region of HA's structure (Figure~\ref{fig:RMSD_heatmap}A).
-However, at the domain level, sites in HA's stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B).
-The HA stalk domain is also more conserved in sequence~\cite{nobusawa1991comparison}, suggesting that conservation of amino-acid sequence is correlated with conservation of amino-acid preferences.
-Consistent with this idea, sites that are absolutely conserved across all 18 HA subtypes are significantly less shifted than sites that are variable across HA subtypes (Figure~\ref{fig:RMSD_heatmap}B).
-Presumably these sites are under consistent functional constraint across all HAs.
-
-Despite their high sequence divergence, H1 and H3 adopt very similar protein folds~\cite{ha2002h5,russell2004h1}.
-However, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among different HA subtypes~\cite{ha2002h5,russell2004h1}.
-Previous work has defined clades of structurally related HA subtypes~\cite{ha2002h5,russell2004h1}.
-One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}A).
-Sites that are conserved at different amino-acid identities in these two clades tend to have exceptionally large shifts in amino acid preferences (Figure~\ref{fig:RMSD_heatmap}B).
-The clade containing H1 has an upward translation of the globular head relative to the clade containing H3.
-This structural shift has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
-Specifically, the clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
-In deep mutational scanning of the H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
-In contrast, the interhelical loop in H3 HA makes a sharper and shorter turn which is facilitated by a Gly at 75(HA2).
-In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations.
-Therefore, some of the shifts in HA amino-acid preferences can be rationalized in terms of changes in HA structure.
-
-%\begin{SCfigure*}[\sidecaptionrelwidth][t]
-\begin{figure}
-\centering
-%\includegraphics[width=11.4cm]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
-\includegraphics[width=\columnwidth]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
-\caption{\label{fig:RMSD_heatmap}
-{\bf Sites with strongly shifted amino-acid preferences between H3 and H1 HAs.}
-(A) The shift in amino-acid preferences between the H3 and H1 HA at each site as calculated in Figure~\ref{fig:distance_distribution}C is mapped onto the structure of the H3 HA.
-(B) Amino-acid preferences of sites in the stalk domain are less shifted than those in the head domain.
-Sites absolutely conserved in all 18 HA subtypes are less shifted than other sites.
-Sites with one amino-acid identity in the clade containing H1, H2, H5, and H6 and another identity in the clade containing H3, H4, and H14 are more shifted than other sites.
-(C) Sites 107 and 75(HA2) help determine the different orientation of the globular head domain in H1 versus H3 HAs.
-These sites are shown in spheres on the structure of H1 and H3 and colored as in panel (A), and the experimentally measured amino-acid preferences in the H1 and H3 HAs are shown.
-One monomer is in dark gray, while the HA1 domain of the neighboring monomer is in lighter gray.
-}
-%\end{SCfigure*}
-\end{figure}
-
-\section*{Discussion}
-\label{sec:discussion}
-We have measured the effects of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements have some value for understanding the evolutionary fate of these mutations in nature.
-Specifically, mutations measured to be more beneficial for viral growth tend to reach higher frequencies in nature than mutations measured to be more deleterious for viral growth.
-The fact that our experiments can help identify evolutionary successful mutations suggests that they might inform evolutionary forecasting.
-In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, {\L}uksza and L{\"a}ssig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
-Our work suggests that deep mutational scanning may be able to provide such data.
-
-It is important to emphasize that measurements of viral growth in cell culture do \emph{not} represent true fitness in nature.
-Indeed, a vast amount of work in virology has chronicled the ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}.
-As an example, although we identified G78D as favorable for viral growth in cell culture, this mutation never fixes in nature.
-Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
-Given these caveats, it might seem surprising that measuring viral growth in cell culture can be informative about the success of viral strains in nature.
-Yet, prior to our work, there were no comprehensive studies of the functional effects of mutations to H3 HA on any property that even resembled viral fitness in nature, and modeling work has either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or assumed that all non-epitope mutations had equivalent deleterious effects~\cite{luksza2014predictive}.
-The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative---and so represent an improvement over no information at all.
-We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains.
-
-We measured the effects of all single amino-acid mutations to a specific HA, and then generalized these measurements to other H3N2 HAs from a 50-year timespan.
-These generalizations will only be valid to the extent that the effects of mutations are conserved during HA's evolution.
-Extensive prior work has shown that epistasis can shift the effects of mutations as proteins evolve~\cite{pollock2012amino,shah2015contingency,gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}.
-Our work suggests that measurements on a HA from single human H3N2 viral strain can be usefully generalized to at least some extent across the entire evolutionary history of human H3N2 HA.
-On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites---much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2018mapping}.
-Further investigation of how mutational effects shift as proteins diverge will be important for determining how broadly any given experiment can be generalized when attempting to make evolutionary forecasts.
-
-Our work did not characterize the antigenic effects of mutations, which also play an important role in determining strain success in nature~\cite{koel2013substitutions,neher2016prediction}.
-However, our basic selection and deep-sequencing approach can be harnessed to completely map how mutations affect antibody recognition~\cite{doud2017complete,doud2018quantifying}.
-But so far, experiments using this approach have not examined antibodies or sera that are relevant to driving the evolution of H3N2 influenza~\cite{doud2017complete,doud2018quantifying}, or have used relevant sera but examined a non-comprehensive set of mutations~\cite{li2016selection}.
-Future experiments that completely map how HA mutations affect recognition by human sera seem likely to be especially fruitful for informing viral forecasting.
-
-{\small
-\section*{Materials and Methods}
-\subsection*{Data and computer code}
-Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
-Computer code used to analyze the data and produce the results in the paper are on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript}.
-
-\subsection*{HA numbering}
-Sites are in H3 numbering, with the signal peptide in negative numbers, HA1 in plain numbers, and HA2 denoted with "(HA2)".
-Sequential 1, 2, ...
-numbering of the Perth/2009 HA can be converted to H3 numbering by subtracting 16 for the HA1 subunit, and subtracting 345 for the HA2 subunit.
-
-\subsection*{Creation of MDCK-SIAT1-TMPRSS2 cells}
-When growing influenza virus in cell culture, trypsin is normally added to cleave HA into its mature form.
-To obviate the need for trypsin, we engineered MDCK-SIAT1 cells and MDCK-SIAT1-CMV-PB1~\cite{bloom2010permissive} cells to constitutively express the TMPRSS2 protease, which cleaves and activates HA in the human airways~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
-The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656) and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry to create plasmid pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
-We used the lentiviral vector to transduce MDCK-SIAT1 or MDCK-SIAT1-CMV-PB1 cells, and sorted an intermediate mCherry-positive population by flow cytometry.
-We refer to the sorted bulk population as MDCK-SIAT1-TMPRSS2 cells or MDCK-SIAT1-CMV-PB1-TMPRSS2 cells.
-There is no selectable marker for the TMPRSS2; however, we maintain the cells at low passage number, and have seen no indication that they lose their ability to support the growth of viruses with H3 HAs in the absence of exogenous trypsin.
-
-\subsection*{Generation of HA codon mutant plasmid libraries}
-HA and NA genes for the Perth/2009 viral strain were cloned from recombinant virus obtained from BEI Resources (NR-41803) into the pHW2000~\citep{hoffmann2000dna} influenza reverse-genetics plasmids to create pHW-Perth09-HA and pHW-Perth09-NA.
-
-We initially created a virus with the HA and NA from Perth/2009 and internal genes from WSN/1933, and passaged it in cell culture to test its genetic stability.
-To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM, or Dulbecco's Modified Eagle Medium, supplemented with 10\% heat-inactivated FBS, or Fetal Bovine Serum, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amounts of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
-The next day we changed the media to influenza growth media (IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL (no trypsin was added since there was TMPRSS2), and then collected the viral supernatant at 72 hours post-transfection.
-This viral supernatant was blind passaged in MDCK-SIAT1-TMPRSS2 a total of six additional times.
-We isolated viral RNA from these passaged viruses and sequenced the HA gene.
-The passaged HA had two mutations, G78D and T212I, which enhanced viral growth as shown in Figure~\ref{suppfig:Perth2009_mut}.
-The HA with these two mutations was cloned into pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} to create pHW-Perth09-HA-G78D-T212I and pICR2-Perth09-HA-G78D-T212I.
-For all subsequent experiments, we used viruses with the HA containing these two mutations to improve titers and viral genetic stability, and this is the HA that we refer to as Perth/2009 in the manuscript text.
-We used all non-HA genes (including NA) from WSN/1933 to help increase titers and reduce biosafety concerns.
-
-The codon-mutant libraries were generated in the Perth/2009 HA-G78D-T212I background using the PCR-based approach described in \cite{bloom2014experimentally} with the primer melting-temperature modifications described in \cite{dingens2017comprehensive}, using two rounds of mutagenesis.
-The script to design the mutagenesis primers is at \url{https://github.com/jbloomlab/CodonTilingPrimers}.
-We created three independent libraries, one for each biological replicate.
-The mutant variants were then cloned at high efficiency into the pICR2~\cite{ashenberg2017deep} vector using digestion with BsmBI, ligation with T4 DNA ligase, and electroporation into ElectroMAX DH10B competent cells (Invitrogen 18290015).
-We obtained $>$6 million transformants for each replicate.
-We scraped the plates, expanded the cultures in liquid LB (Luria-Bertani Broth) + ampicillin at 37$^{\circ}$C for 3 h with shaking, and then maxiprepped.
-We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (Figure~\ref{suppfig:SangerSeq}).
-
-\subsection*{Generation and passaging of mutant viruses}
-The mutant virus libraries were generated using the helper-virus approach described in \cite{doud2016accurate} with several modifications, most notably the cell line used.
-Briefly, we transfected $5 \times 10^5$ MDCK-SIAT1-TMPRSS2 cells in suspension with 937.5 ng each of four protein expression plasmids encoding the ribonucleoprotein complex (HDM-Nan95-PA, HDM-Nan95-PB1, HDM-Nan95-PB2, and HDM-Aichi68-NP)~\cite{gong2013stability}, and 1250 ng of one of the three pICR2-mutant-HA libraries (or the wildtype control) using Lipofectamine 3000 (ThermoFisher L3000008).
-We allowed the transfected cells to adhere in 6-well plates and four hours later changed the media to D10 media.
-Eighteen hours after transfection, we infected the cells with the WSN/1933 HA-deficient helper virus~\cite{doud2016accurate} by preparing an inoculum of 500 TCID$_{50}$ per $\mu$L of helper virus (as computed on HA-expressing cells) in IGM, aspirating the D10 media from the cells, and adding 2 mL of the helper-virus inoculum to each well.
-After three hours, we changed the media to fresh IGM.
-At 24 hours after helper-virus infection, we harvested the viral supernatants for each replicate, froze aliquots at -80$^{\circ}$C, and titered in MDCK-SIAT1-TMPRSS2 cells.
-The titers were 92, 536, 536, and 734 TCID$_{50}$ per $\mu$L for the three library replicates and the wildtype control, respectively.
-
-We passaged $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
-To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in fifteen 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they were at $\sim 1.7 \times 10^7$ cells per dish.
-We replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L in IGM.
-Three hours post-infection, we replaced the inoculum with fresh IGM for replicates 1, 2, and 3-2.
-We did not perform a media change for replicate 3-1.
-As can be seen in Figure~\ref{fig:dms_overview}D, the media change does not appear to have a substantial effect, as replicate 3-1 looks comparable to the other replicates.
-We collected viral supernatant for sequencing 48 hours post-infection.
-
-\subsection*{Barcoded subamplicon sequencing}
-To extract viral RNA from the three replicate HA virus libraries and the wildtype HA virus, we clarified the viral supernatant by centrifuging at 2000 x \textit{g} for five minutes, then ultracentrifuged 24 mL of the clarified supernatant at 22,000 rpm for 1.5 h at 4$^{\circ}$C in a Beckman Coulter SW28 rotor, and extracted RNA using the Qiagen RNeasy Mini Kit by resuspending the viral pellet in 400 $\mu$L of buffer RLT supplemented with $\beta$-mercaptoethanol, pipetting 30 times, transferring the liquid to a microcentrifuge tube, adding 600 $\mu$L 70\% ethanol, and proceeding according to the manufacturer's instructions.
-The HA gene was reverse-transcribed with AccuScript Reverse Transcriptase (Agilent 200820) using primers P09-HA-For (5'-AGCAAAAGCAGGGGATAATTCTATTAATC-3') and P09-HA-Rev (5'-AGTAGAAACAAGGGTGTTTTTAATTACTAATACAC-3').
-
-We generated the HA PCR amplicons for the three plasmid libraries, the three virus libraries, the wildtype plasmid control, and the wildtype virus control using KOD Hot Start Master Mix (EMD Millipore 71842) using the PCR reaction mixture and cycling conditions described in~\cite{bloom2014experimentally} and the P09-HA-For and P09-HA-Rev primers.
-We prepared the sequencing libraries using a barcoded-subamplicon strategy~\cite{wu2014high} to increase the accuracy from deep sequencing.
-The exact details of this approach are described in~\cite{doud2016accurate} (also see \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
-The primers used to generate the subamplicons are in Dataset~\ref{suppdata:bcsubamp_primers}.
-We performed deep sequencing on a lane of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.
-
-\subsection*{Analysis of deep sequencing data}
-We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, version \texttt{2.2.5}) to analyze the deep sequencing data.
-The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
-The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~\ref{suppdata:prefsfiles}.
-Computer code that performs the entirety of this analysis and shows many detailed plots about read depth and other quality control metrics is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/analysis_notebook.ipynb}.
-
-\subsection*{Phylogenetic model comparison and fitting of a stringency parameter}
-For the analysis in Table~\ref{tab:phydms}, we downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
-These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
-We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.2}) to fit the substitution models listed in Table~\ref{tab:phydms}.
-
-The amino-acid preferences were re-scaled by the stringency parameter using the approach described in \cite{hilton2017phydms}.
-Note that the re-scaling simply puts the amino-acid preferences on a useful scale for visualization in logo plots, and has no effect on any of the quantitative conclusions relating the deep mutational scanning to natural evolution.
-The reason is that all of these conclusions use the effects of mutations as calculated using Equation~\ref{eq:muteffect}, and the re-scaling simply acts as a constant multiplier on all mutational effects (e.g., a scale factor) when re-scaled preferences are converted to mutational effects.
-
-The phylogenetic tree of HA subtypes in Figure~\ref{fig:distance_distribution}A was generated as described in~\cite{doud2018quantifying}.
-
-\subsection*{Quantification of Shannon entropy and relative solvent accessibility}
-We calculated Shannon entropy, $h_r$ for site $r$ as $h_r = - \sum_{x} \pi_{r,x} \log(\pi_{r,x})$, where $\pi_{r,x}$ is the preference for amino acid $x$ at site $r$.
-
-We quantified the absolute solvent accessibility of each site of the H3 HA (PDB 4O5N;~\cite{lee2014receptor}) or the H1 HA (PDB 1RVX;~\cite{gamblin2004structure}) structure using DSSP~\cite{kabsch1983}.
-We then normalized to a relative solvent accessibility using the absolute accessibilities in~\cite{tien2013}.
-
-\subsection*{Quantification of mutational effects}
-We calculated the effects of mutations from the amino-acid preferences that were estimated from the deep mutational scanning data.
-The effect of mutating site $r$ from amino acid $a_1$ to $a_2$ was calculated as
-\begin{equation}
-\label{eq:muteffect}
-\log_2 \frac{\pi_{r,a_2}}{\pi_{r,a_1}}
-\end{equation}
-where $\pi_{r,a_1}$ and $\pi_{r,a_2}$ are the re-scaled preferences for amino acids $a_1$ or $a_2$ at site $r$ as shown in Figure~\ref{fig:logoplot}.
-The WSN/1933 H1 HA amino-acid preferences are the replicate-average values reported in \cite{doud2016accurate}, re-scaled by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
-
-\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
-To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid} (see Dataset~\ref{suppdata:gisaid}), sampling six viruses per month over the time interval of January 1, 1968 to February 1, 2018.
-We aligned the resulting 2,189 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
-Ancestral state reconstruction and branch length timing were performed with TreeTime~\cite{Sagulenko2018}.
-The phylogenetic tree is available as a JSON file on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/data/flu_h3n2_ha_1968_2018_6v_tree.json.gz}.
-The tree was visualized using BALTIC (\url{https://github.com/blab/baltic}).
-
-% notation for frequencies model
-\newcommand{\dx}{\mathrm{d}x}						% change in x
-\newcommand{\dy}{\mathrm{d}y}						% change in y
-\newcommand{\dt}{\mathrm{d}t}						% change in t
-\newcommand{\inertia}{\epsilon}			    % inertia parameter
-\newcommand{\normal}{\mathcal{N}}				% normal distribution
-
-The frequency trajectory of each individual mutation on the phylogeny is estimated following Nextstrain's augur pipeline and as first implemented in Nextflu \cite{neher2015nextflu}.
-Herein, mutation frequency dynamics are modeled according to a Brownian motion diffusion process discretized to one-month intervals.
-The number of viruses sampled in each interval determines the denominator of the mutation frequency calculations.
-Relative to a simple Brownian motion, the expectation includes an ``inertia'' term $\inertia$ that adds velocity to the diffusion and the variance includes a term $x(1-x)$ to scale variance according to frequency following a Wright-Fisher population genetic process.
-This results in the following diffusion process
-\begin{equation}
-x(t+\dt) = \normal\big( x(t) + \inertia \, \dx , \; \dt \, \sigma^2 \, x(t) \, (1-x(t)) \big),
-\end{equation}
-with `volatility' parameter $\sigma^2$.
-The term $\dx$ is the increment in the previous timestep, so that $\dx = x(t) - x(t-\dt)$.
-We used $\inertia = 0.7$ and $\sigma^2 = 0.05$ to maximize fit to empirical trajectory behavior.
-
-We also include an Bernoulli observation model for mutation presence / absence among sampled viruses at timestep $t$.
-This observation model follows
-\begin{equation}
-f(x,t) = \prod_{v\in V} x(t) \prod_{v\notin V} (1-x(t)),
-\end{equation}
-where $v\in V$ represents the set of viruses that have the mutation and $v\notin V$ represents the set of viruses that do not have the mutation.
-Each frequency trajectory is estimated by simultaneously maximizing the likelihood of the process model and the likelihood of the observation model via adjusting frequency trajectory $\mathbf{x}=(x_1, \ldots, x_n)$.
-
-We also repeated the above analyses using only viruses that were sequenced directly without passaging.
-Routine direct sequencing did not begin until the early 2000s \cite{mcwhite2016sequence}.
-To construct a tree with a similar number of viruses as the original analysis, we sampled 30 viruses per month between January 1, 2000 and April 1, 2018, producing a tree with 2,374 unpassaged viruses with augur (commit: \texttt{6d9f708}).
-We included the passaged DMS strain, A/Perth/16/2009, in the resulting tree to enable comparison between pre-Perth and post-Perth clades.
-
-\subsection*{Analysis of mutational shifts}
-To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using MAFFT~\cite{katoh2013mafft}.
-To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2018mapping} and used the RMSD$_{\text{corrected}}$ values as our quantification of the extent of each shift.
-
-For the plots shown in Figure~\ref{fig:RMSD_heatmap}B, any residues falling between Cys-52 and Cys-277 were defined as the head domain, and all other residues were defined as the stalk domain.
-We used the multiple sequence alignment of the HA subtype sequences from \cite{doud2018quantifying} to identify sites that are absolutely conserved across all subtypes, or in the different clades described in Figure~\ref{fig:RMSD_heatmap}.
-
-\subsection*{Validation of individual point mutants}
-To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, T24F, T40V, S287A, and C199(HA2)K, we used site-directed mutagenesis to introduce the amino-acid mutation into the Perth/2009 HA bidirectional reverse genetics plasmid, and verified the sequence of two clones for each mutant by Sanger sequencing.
-We generated these individual mutant viruses carrying GFP in the PB1 segment using a protocol described in \cite{bloom2010permissive,hooper2013mutant}, and the PB2, PA, NP, NA, M, and NS segments of the A/WSN/1933 (H1N1) strain.
-To rescue each mutant GFP-carrying virus in duplicate, we transfected a co-culture of $4 \times 10^5$ 293T-CMV-PB1 and $0.5 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells with the eight reverse genetics plasmids and the pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W plasmid.
-Each well received a transfection mixture of 100 $\mu$L DMEM, 3 $\mu$L BioT transfection reagent, and 250 ng of each plasmid.
-We changed the media in each well with 2 mL IGM eight hours post-transfection.
-At 53 hours post-transfection, transfection supernatants were harvested, clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes, aliquoted, and frozen at -80$^\circ$C.
-
-To titer the GFP-carrying viruses, we plated $1 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells per well in 12-well plates in IGM.
-Four hours after plating, we infected cells with dilutions of viral supernatant.
-At 16 hours post-infection, we selected wells with 1 to 10\% of cells that were GFP-positive and analyzed the fraction of GFP-positive cells by flow cytometry to calculate the titer of infectious particles per $\mu$L.
-}
-
-\acknow{We thank Sarah Hilton, Hugh Haddox, and Sidney Bell for helpful discussions about data analysis, and Richard Neher for sharing analysis code and providing helpful comments on the manuscript.
-We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
-This work was supported by grant R01 AI127893 from the NIAID of the NIH to JDB and TB and grant U19 AI117891 to TB.
-JML was supported in part by the Center for Inference and Dynamics of Infectious Diseases (CIDID), which is funded by grant U54GM111274 from the NIGMS of the NIH.
-The research of JDB is supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation, and a Burroughs Wellcome Young Investigator in the Pathogenesis of Infectious Diseases grant.
-TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774.
-}
-
-\showacknow{} % Display the acknowledgments section
-
-% \pnasbreak splits and balances the columns before the references.
-% Uncomment \pnasbreak to view the references in the PNAS-style
-% If you see unexpected formatting errors, try commenting out \pnasbreak
-% as it can run into problems with floats and footnotes on the final page.
-%\pnasbreak
-
-% Bibliography
-\subsection*{References}
-\bibliography{references}
-
-% Supplementary material temporarily moved until AFTER everything else for initial submission
 
@@
 (A) This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:frequency_trajectory}, but colors amino-acid mutations by whether they occur in the head (purple) or the stalk (blue) domain.
-(B) Histogram of mutation maximum frequencies by the number of nodes for head and stalk domain mutations.
+(B) Histogram of mutation maximum frequencies by the number of mutations in the head and stalk domains.
 It is clear that mutations in the head domain are more numerous than those in the stalk, particularly among mutations that reach high frequencies.
@@
 
+\subsection*{References}
+\bibliography{references}
+
 \end{document}

### 2018-06-29 paper/manuscript_noSI.tex
split manuscript and supporting information

--- 2018-06-28 paper/manuscript.tex
@@
 \renewcommand{\thesuppfigure}{S\arabic{suppfigure}}
+
+\usepackage{xr-hyper}
+\externaldocument{SI}
 
@@
 \centering
-\includegraphics[width=12cm]{figs/dms_overview/dms_overview.pdf}
+\includegraphics[width=12cm]{dms_overview.pdf}
 \caption{\label{fig:dms_overview}
@@
 This strain was the H3N2 component of the influenza vaccine from 2010-2012~\cite{who2010d,who2011}.
-Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral growth in cell culture, G78D and T212I (Figure~\ref{suppfig:Perth2009_mut} and Dataset~\ref{suppdata:PerthHA}).
+Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral growth in cell culture, G78D and T212I (Figure~S1 and Dataset~S1).
 The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.
 
-We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (Figure~\ref{suppfig:SangerSeq}).
+We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (Figure~S2).
 We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
@@
 \centering
-\includegraphics[width=17cm]{figs/prefslogoplot/rescaled-avgprefs_prefs.pdf}
+\includegraphics[width=17cm]{rescaled-avgprefs_prefs.pdf}
 \caption{\label{fig:logoplot}
@@
 \centering
-\includegraphics[width=0.95\textwidth]{figs/mut_tolerance/mut_tolerance_new.pdf}
+\includegraphics[width=0.95\textwidth]{mut_tolerance_new.pdf}
 \caption{\label{fig:mut_tolerance}
@@
 Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
-A residue was classified as solvent exposed if its relative solvent accessibility was $\geq$ 0.2; Figure~\ref{suppfig:head_stalk_RSA} shows that the results are robust to the choice of solvent accessibility cutoff.
+A residue was classified as solvent exposed if its relative solvent accessibility was $\geq$ 0.2; Figure~S7 shows that the results are robust to the choice of solvent accessibility cutoff.
 Note that the mutational tolerance values are not comparable between the two HAs, but are comparable between domains of the same HA.
@@
 
-However, there are also some differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences (Figure~\ref{suppfig:diffprefs_logoplot}).
+However, there are also some differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences (Figure~S3).
 Most surprisingly, the start codon does not show a particularly strong preference for methionine (Figure~\ref{fig:logoplot}).
-We validated that a virus carrying a mutation at this site from methionine to lysine does in fact reach appreciable titers (Figure~\ref{suppfig:startcodon_validation}), perhaps because of alternative translation-initiation at a downstream or upstream start site as has been described for other HAs~\cite{girard2011upstream}.
+We validated that a virus carrying a mutation at this site from methionine to lysine does in fact reach appreciable titers (Figure~S4), perhaps because of alternative translation-initiation at a downstream or upstream start site as has been described for other HAs~\cite{girard2011upstream}.
 Our measurements also suggest mutational tolerance at some other sites that are relatively conserved among natural HAs, such as the N-linked glycosylation motifs near the beginning of HA1 and the transmembrane domain (Figure~\ref{fig:logoplot}).
-We validated that viruses with mutations to the glycosylation motifs at sites 22 or 38, or a site in the transmembrane domain, do in fact grow to high titers (Figures~\ref{suppfig:glycan_validation} and \ref{suppfig:TM_validation}, respectively).
+We validated that viruses with mutations to the glycosylation motifs at sites 22 or 38, or a site in the transmembrane domain, do in fact grow to high titers (Figures~S5 and S6, respectively).
 The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function in nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.
@@
 Specifically, whereas solvent-exposed residues in the head domain are substantially more mutationally tolerant than those in the stalk domain for the WSN/1933 H1 HA, the trend is actually reversed for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
-This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (Figure~\ref{suppfig:head_stalk_RSA}).
+This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (Figure~S7).
 For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
@@
 \centering
-\includegraphics[width=0.95\linewidth]{figs/frequency_trajectory_ex/frequency_trajectory_ex.pdf}
+\includegraphics[width=0.95\linewidth]{frequency_trajectory_ex.pdf}
 \caption{\label{fig:frequency_trajectory}
 {\bf Frequency trajectories of individual mutations and their relation to the experimentally measured effects of these mutations.}
-The top panel shows the subset of the full H3N2 HA tree in Figure~\ref{suppfig:tree} from 2004 to 2014.
+The top panel shows the subset of the full H3N2 HA tree in Figure~S8 from 2004 to 2014.
 Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
@@
 It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
-Figure~\ref{suppfig:head_stalk_mutfreq_supp} shows a similar layout but colors mutations by whether they are in HA's head or stalk domain.
+Figure~S11 shows a similar layout but colors mutations by whether they are in HA's head or stalk domain.
 }
@@
 \centering
-\includegraphics[width=12cm]{figs/muteffect_by_maxfreq/muteffect_by_maxfreq.pdf}
+\includegraphics[width=12cm]{muteffect_by_maxfreq.pdf}
 \caption{\label{fig:muteffect_maxfreq}
@@
 The vertical black line shows the mean mutation effect for each category.
-The analysis is performed separately for pre-Perth/2009, post-Perth/2009, and unpassaged isolates from the post-Perth/2009 partitions of the tree (see Figure~\ref{suppfig:tree}).
+The analysis is performed separately for pre-Perth/2009, post-Perth/2009, and unpassaged isolates from the post-Perth/2009 partitions of the tree (see Figure~S8).
 }
@@
 To calculate a given mutation's effect, we simply took the logarithm of the ratio of the preferences for the mutant and wildtype amino acids at that site.
-To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
+To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~S8).
 We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after, since these mutations have not had enough time for their evolutionary fates to be fully resolved.
@@
 \centering
-\includegraphics[width=12cm]{figs/muteffect_by_maxfreq_WSN/muteffect_by_maxfreq_WSN.pdf}
+\includegraphics[width=12cm]{muteffect_by_maxfreq_WSN.pdf}
 \caption{\label{fig:muteffect_maxfreq_WSN}
@@
 This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
-Figure~\ref{suppfig:muteffect_maxfreq_WSN_supp} shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
+Figure~S10 shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
 The empirical $P$-value represents the result of 1,000 permutations.
@@
 \subsection*{Measurements made on an H1 HA are less informative for understanding the evolution of H3 influenza}
-To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (Figure~\ref{suppfig:WSNprefs_logoplot}), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
+To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (Figure~S9), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
 Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the correlations between the H1 experimental measurements and the maximum frequency that mutations reach during H3N2 viral evolution are consistently weaker than those using H3 experimental measurements (compare Figure~\ref{fig:muteffect_maxfreq_WSN} to Figure~\ref{fig:muteffect_maxfreq}A).
@@
 \centering
-\includegraphics[width=12cm]{figs/distance_distribution/distance_distribution.pdf}
+\includegraphics[width=12cm]{distance_distribution.pdf}
 \caption{\label{fig:distance_distribution}
@@
 \centering
-%\includegraphics[width=11.4cm]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
-\includegraphics[width=\columnwidth]{figs/RMSD_heatmap/RMSD_heatmap.pdf}
+\includegraphics[width=\columnwidth]{RMSD_heatmap.pdf}
 \caption{\label{fig:RMSD_heatmap}
@@
 We isolated viral RNA from these passaged viruses and sequenced the HA gene.
-The passaged HA had two mutations, G78D and T212I, which enhanced viral growth as shown in Figure~\ref{suppfig:Perth2009_mut}.
+The passaged HA had two mutations, G78D and T212I, which enhanced viral growth as shown in Figure~S1.
 The HA with these two mutations was cloned into pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} to create pHW-Perth09-HA-G78D-T212I and pICR2-Perth09-HA-G78D-T212I.
@@
 We scraped the plates, expanded the cultures in liquid LB (Luria-Bertani Broth) + ampicillin at 37$^{\circ}$C for 3 h with shaking, and then maxiprepped.
-We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (Figure~\ref{suppfig:SangerSeq}).
+We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (Figure~S2).
 
@@
 The exact details of this approach are described in~\cite{doud2016accurate} (also see \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
-The primers used to generate the subamplicons are in Dataset~\ref{suppdata:bcsubamp_primers}.
+The primers used to generate the subamplicons are in Dataset~S2.
 We performed deep sequencing on a lane of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.
@@
 The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
-The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~\ref{suppdata:prefsfiles}.
+The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~S3.
 Computer code that performs the entirety of this analysis and shows many detailed plots about read depth and other quality control metrics is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/analysis_notebook.ipynb}.
@@
 \subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
-To generate the tree shown in Figure~\ref{suppfig:tree}, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid} (see Dataset~\ref{suppdata:gisaid}), sampling six viruses per month over the time interval of January 1, 1968 to February 1, 2018.
+To generate the tree shown in Figure~S8, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid} (see Dataset~S4), sampling six viruses per month over the time interval of January 1, 1968 to February 1, 2018.
 We aligned the resulting 2,189 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
@@
 
-% Supplementary material temporarily moved until AFTER everything else for initial submission
-
-\onecolumn
-
-\clearpage
-\section*{Supporting Information (SI)}
-\FloatBarrier
-\pagenumbering{arabic}% resets `page` counter to 1
-\renewcommand*{\thepage}{S\arabic{page}}
-
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.5\textwidth]{figs/supp_G78D-T212I/G78D-T212I.pdf}}
-\caption{\label{suppfig:Perth2009_mut}
-{\bf Characterization of the G78D-T212I Perth/2009 HA variant.}
-(A)
-The G78D-T212I Perth/2009 HA variant supports better viral growth than the wildtype Perth/2009 HA.
-Viruses were generated in duplicate by reverse genetics with the Perth/2009 NA and WSN internal genes, and passaged once at MOI = 0.01 in MDCK-SIAT1-TMPRSS2 cells.
-The rescue and passage viral supernatants were collected at 72 hours post-transfection and 44 hours post-infection, respectively, and titered in MDCK-SIAT1-TMPRSS2 cells.
-The points mark each duplicate and the bar marks the mean.
-(B)
-The D78 variant remained at a low frequency in natural human H3N2 sequences over the past $~\sim$10 years.
-The A212 variant rose to fixation in $~\sim$2011, replacing the T212 variant.
-}
-\end{suppfigure}
-
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.6\textwidth]{figs/supp_SangerSeq/SangerSeq.pdf}}
-\caption{\label{suppfig:SangerSeq}
-{\bf Sanger sequencing of 31 randomly chosen clones from the mutant plasmid libraries.}
-(A) There were an average of $\sim$1.4 codon mutations per clone across the three plasmid mutant libraries.
-(B) A mixture of one-, two-, and three-nucleotide mutations were present, with slightly fewer triple-nucleotide changes than expected.
-(C) Nucleotide frequencies were uniform in the codon mutations.
-(D) The mutations were distributed relatively evenly across the length of the HA coding sequence.
-(E) We calculated the pairwise distances between mutations for clones carrying more than one mutation.
-The cumulative distribution of these distances is shown in the red line.
-The blue line indicates the expected distribution if mutations in multiply mutated genes are randomly dispersed along the sequence.
-}
-\end{suppfigure}
-
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.9\textwidth]{figs/supp_diffprefs/diffprefs.pdf}}
-\caption{\label{suppfig:diffprefs_logoplot}
-{\bf Sites where there are strong differences between our experimental measurements and the amino-acid frequencies among natural HA sequences.}
-We calculated the distance between our H3 measurements and the alignment frequencies using the same approach as in Figure~\ref{fig:distance_distribution}, but using the alignment frequencies in place of the H1 preferences.
-For each site, the height of each letter above or below the line indicates how much more or less preferred that amino acid is in our experiments compared to its frequency nature.
-The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
-The sites are in H3 numbering.
-The HA alignment used to calculate the natural frequencies is the same one used in Table~\ref{tab:phydms}.
-}
-\end{suppfigure}
-
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.4\textwidth]{figs/supp_point_mut_validation/startcodon_titers.pdf}}
-\caption{
-{\bf Validation that the Perth/2009 HA is somewhat tolerant to mutation of the canonical start codon.}
-We created variants of the Perth/2009 HA in which the canonical start codon was mutated (amino-acid mutation M(-16)K, codon mutation ATG$\rightarrow$AAA), a conserved cysteine was mutated to alanine (C52A, codon mutation TGC$\rightarrow$GCA), and the same cysteine was synonymously mutated (C52C, codon mutation TGC$\rightarrow$TGT).
-We selected these mutations because our deep mutational scanning results in Figure~\ref{fig:logoplot} surprisingly suggest that the start codon (position -16) is fairly tolerant of mutations, but that site 52 is highly intolerant of mutation to anything other than cysteine.
-The synonymous C52C mutation is a negative control mutation that is not expected to have any effect.
-We then used reverse genetics to generate viruses carrying the wildtype HA or each point mutant, with GFP packaged in the PB1 segment to enable easy titering by flow cytometry~\cite{bloom2010permissive,hooper2013mutant}.
-Each variant was generated in duplicate, and the plots show the viral titer in the supernatant at 53 hours post-transfection.
-As expected, the C52A virus is essentially non-viable, while the C52C mutation is roughly equivalent to wildtype.
-The M(-16)K mutation is only moderately attenuated, validating that the canonical start codon is not completely essential for viral growth.
-}
-\label{suppfig:startcodon_validation}
-\end{suppfigure}
-
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.4\textwidth]{figs/supp_point_mut_validation/glycan_titers.pdf}}
-\caption{\label{suppfig:glycan_validation}
-{\bf Validation of viral variants with mutations at N-linked glycosylation motifs.}
-We created variants of the Perth/2009 HA with mutations that eliminated N-linked glycosylation motifs (Asn-Xaa-Ser/Thr) at asparagine residues 22, 38, and 285 (these are mutations T24F, T40V, and S287A, respectively).
-The codon mutations were ACG$\rightarrow$TTC, ACT$\rightarrow$GTA, and AGC$\rightarrow$GCG, respectively.
-We then used reverse genetics to generate viruses carrying the wildtype HA or each of these mutants.
-Each variant was generated in duplicate, and the plots show the viral titer in the supernatant at 53 hours post-transfection.
-The viruses with mutations at the motifs at residues 22 and 38 reached titers at least as high as wildtype, whereas the virus with a mutation to the motif at residue 285 was modestly attenuated.
-}
-\end{suppfigure}
-
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.45\textwidth]{figs/supp_point_mut_validation/tm_titers.pdf}}
-\caption{\label{suppfig:TM_validation}
-{\bf Validation of the mutational tolerance of a site in the transmembrane domain.}
-We created a variant of the Perth/2009 HA with a transmembrane domain mutation, C199(HA2)K (codon mutation TGT$\rightarrow$AAG), at a site that our deep mutational scanning suggests should be highly mutationally tolerant (Figure~\ref{fig:logoplot}).
-We then used reverse genetics to generate viruses carrying the wildtype HA or this mutant.
-Each variant was generated in duplicate, and the plots show the viral titer in the supernatant at 53 hours post-transfection.
-The virus with the mutation reached titers comparable to wildtype.
-}
-\end{suppfigure}
-
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.75\textwidth]{figs/supp_head_stalk_RSA/head_stalk_RSA.pdf}}
-\caption{\label{suppfig:head_stalk_RSA}
-{\bf Mutational tolerances of the head and stalk domains at various relative solvent accessibility cutoffs.}
-The mutational tolerances of the head and stalk domains show less disparity for the Perth/2009 H3 HA compared to those for the WSN/1933 H1 HA.
-We used relative solvent accessibility (RSA) cutoffs of $0.1$, $0.2$, and $0.3$ to define solvent-exposed residues and plotted the mutational tolerances (Shannon entropy of re-scaled preferences) of these residues in the head and stalk domains for the Perth/2009 H3 HA (purple) and the WSN/1933 H1 HA (brown).
-Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
-The HA structures color the residues that are defined as solvent exposed at a given RSA cutoff.
-One monomer is shown in surface representation and another monomer shown in ribbon representation.
-Residues in lighter shades of purple or brown are in the head domain, while residues in darker shades are in the stalk domain.
-Note that the mutational tolerance values are not comparable between the two HAs.
-}
-\end{suppfigure}
-
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.9\textwidth]{figs/supp_H3N2_phylogeny/H3N2_phylogeny.pdf}}
-\caption{\label{suppfig:tree}
-{\bf A phylogenetic tree of all HA sequences used in our analysis of mutation frequencies.}
-(A) HA sequences were sampled at a rate of six viruses per month from January 1, 1968 through February 1, 2018.
-The Perth/2009 strain used in our experiments is indicated.
-The rest of the tree is partitioned into nodes that preceded the split of the Perth/2009 strain from the trunk of the tree (blue) and nodes that branched off the trunk after the clade containing Perth/2009 (orange).
-In Figure~\ref{fig:muteffect_maxfreq}, these two partitions of the tree are analyzed separately.
-Nodes in the clade containing the Perth/2009 strain and nodes sampled in 2014 or after were excluded from our analyses.
-The Perth/2009 strain was excluded to avoid artifacts related to mutations that occurred on the branches leading to the HA sequence used in the experiment.
-The post-2014 nodes were excluded because the evolutionary fates of many sequences after this date are not yet fully resolved.
-(B) The post-Perth/2009 partition of the tree containing only sequences from unpassaged isolates.
-}
-\end{suppfigure}
-
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.9\textwidth]{figs/supp_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
-\caption{\label{suppfig:WSNprefs_logoplot}
-{\bf The site-specific amino-acid preferences of the WSN/1933 H1 HA.}
-The amino-acid preferences of the WSN/1933 H1 HA from~\cite{doud2016accurate} after taking the average of the experimental replicates and re-scaling~\cite{hilton2017phydms} by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
-The overlays show the same information as in Figure~\ref{fig:logoplot} (domain and wildtype amino acid).
-Note that \cite{doud2016accurate} used libraries in which all codons were mutagenized \emph{except} for the one encoding N-terminal methionine.
-Therefore, no data is shown for the first codon in the gene.
-The sites are in H3 numbering.
-}
-\end{suppfigure}
-
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_muteffect_maxfreq_WSN/muteffect_maxfreq_WSN_supp.pdf}}
-\caption{\label{suppfig:muteffect_maxfreq_WSN_supp}
-{\bf The distribution of mutational effects measured in H1 HA among H3N2 mutations binned by the maximum frequency that they reach.}
-This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}B, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
-}
-\end{suppfigure}
-
-\begin{suppfigure}[H]
-\centerline{\includegraphics[width=0.5\textwidth]{figs/supp_head_stalk_mutfreq/head_stalk_mutfreq.pdf}}
-\caption{\label{suppfig:head_stalk_mutfreq_supp}
-{\bf Frequency trajectories of head and stalk domain mutations.}
-(A) This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:frequency_trajectory}, but colors amino-acid mutations by whether they occur in the head (purple) or the stalk (blue) domain.
-(B) Histogram of mutation maximum frequencies by the number of nodes for head and stalk domain mutations.
-It is clear that mutations in the head domain are more numerous than those in the stalk, particularly among mutations that reach high frequencies.
-}
-\end{suppfigure}
-
-\begin{suppdata}[H]
-\caption{\label{suppdata:PerthHA}
-Genbank file giving the full sequence of the bidirectional reverse-genetics plasmid pHW-Perth09-HA-G78D-T212I, which encodes the wildtype HA sequence used in this study.
-}
-\end{suppdata}
-
-\begin{suppdata}[H]
-\caption{\label{suppdata:bcsubamp_primers}
-Excel file providing the primers used to generate the barcoded subamplicons for Perth/2009 HA deep sequencing.
-}
-\end{suppdata}
-
-\begin{suppdata}[H]
-\caption{\label{suppdata:prefsfiles}
-Excel file giving the amino-acid preferences in sequential 1, 2, ... numbering of the Perth/2009 HA.
-The unscaled preferences for replicates 1, 2, 3-1, and 3-2 are each in a separate tab of the file.
-Additional tabs give the across-replicate averaged and re-scaled amino-acid preferences in sequential numbering and in H3 numbering as shown in Figure~\ref{fig:logoplot}.
-There are also tabs that give the conversion from sequential to H3 numbering, and a list of the mutations in the 31 clones we Sanger sequenced to evaluate the mutation rate of the mutant plasmid libraries.
-Each tab can simply be exported to CSV for computational analyses.
-}
-\end{suppdata}
-
-\begin{suppdata}[H]
-\caption{\label{suppdata:gisaid}
-Excel file providing acknowledgments and accessions for sequences downloaded from GISAID.
-}
-\end{suppdata}
-
 \end{document}

### 2018-06-30 paper/SI.tex
moved SI to PNAS LaTex template

@@
-\documentclass[9pt,twocolumn,twoside]{pnas-new}
+\documentclass[9pt,twoside,lineno]{pnas-new}
 % Use the lineno option to display guide line numbers if required.
-% Note that the use of elements such as single-column equations
-% may affect the guide line number alignment.
 
-\templatetype{pnasresearcharticle} % Choose template
-% {pnasresearcharticle} = Template for a two-column research article
-% {pnasmathematics} = Template for a one-column mathematics article
-% {pnasinvited} = Template for a PNAS invited submission
-
-% added to enable supplementary figures
-\usepackage{placeins}
-\usepackage{newfloat}
 \usepackage{xr-hyper}
 \externaldocument{manuscript_noSI}
-\DeclareFloatingEnvironment[name={Figure}]{suppfigure}
-\renewcommand{\thesuppfigure}{S\arabic{suppfigure}}
 
-\DeclareFloatingEnvironment[name={Dataset}]{suppdata}
-\renewcommand{\thesuppdata}{S\arabic{suppdata}}
+\templatetype{pnassupportinginfo}
+% \readytosubmit %% Uncomment this line before submitting, so that the instruction page is removed.
 
-\setcounter{topnumber}{10}
-\setcounter{bottomnumber}{10}
-\setcounter{totalnumber}{10}
-
-\newcommand{\comment}[1]{{\color{red}[\textsl{#1}]}}
+\title{Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants}
+\author{JM Lee, J Huddleston, MB Doud, KA Hooper, NC Wu, T Bedford, JD Bloom}
+\correspondingauthor{Trevor Bedford \\ tbedford@fredhutch.org \\ Jesse D. Bloom \\ jbloom@fredhutch.org}
 
@@
 
-% Please give the surname of the lead author for the running footer
-\leadauthor{Lee}
+%% Comment/remove this line before generating final copy for submission
+\instructionspage
 
-\dates{This manuscript was compiled on \today}
-\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
+\maketitle
 
-% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
-% You should only change this length when you've finalised the article contents.
-\verticaladjustment{-2pt}
+%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
+\SItext
 
-\onecolumn
+\subsection*{Subhead}
+Type or paste text here. This should be additional explanatory text such as an extended technical description of results, full details of mathematical models, extended lists of acknowledgments, etc.
 
-\clearpage
-\section*{Supporting Information (SI)}
-\FloatBarrier
-\pagenumbering{arabic}% resets `page` counter to 1
-\renewcommand*{\thepage}{S\arabic{page}}
+\section*{Heading}
+\subsection*{Subhead}
+Type or paste text here. You may break this section up into subheads as needed (e.g., one section on ``Materials'' and one on ``Methods'').
 
-\begin{suppfigure}[H]
+\subsection*{Materials}
+Add a Materials subsection if you need to.
+
+\subsection*{Methods}
+Add a Methods subsection if you need to.
+
+%%% Each figure should be on its own page
+\begin{figure}
 \centerline{\includegraphics[width=0.5\textwidth]{figs/supp_G78D-T212I/G78D-T212I.pdf}}
@@
 }
-\end{suppfigure}
+\end{figure}
+\FloatBarrier
 
-\begin{suppfigure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.6\textwidth]{figs/supp_SangerSeq/SangerSeq.pdf}}
@@
 }
-\end{suppfigure}
+\end{figure}
+\FloatBarrier
 
-\begin{suppfigure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.9\textwidth]{figs/supp_diffprefs/diffprefs.pdf}}
@@
 }
-\end{suppfigure}
+\end{figure}
+\FloatBarrier
 
-\begin{suppfigure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.4\textwidth]{figs/supp_point_mut_validation/startcodon_titers.pdf}}
@@
 \label{suppfig:startcodon_validation}
-\end{suppfigure}
+\end{figure}
+\FloatBarrier
 
-\begin{suppfigure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.4\textwidth]{figs/supp_point_mut_validation/glycan_titers.pdf}}
@@
 }
-\end{suppfigure}
+\end{figure}
+\FloatBarrier
 
-\begin{suppfigure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.45\textwidth]{figs/supp_point_mut_validation/tm_titers.pdf}}
@@
 }
-\end{suppfigure}
+\end{figure}
+\FloatBarrier
 
-\begin{suppfigure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.75\textwidth]{figs/supp_head_stalk_RSA/head_stalk_RSA.pdf}}
@@
 }
-\end{suppfigure}
+\end{figure}
+\FloatBarrier
 
-\begin{suppfigure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.9\textwidth]{figs/supp_H3N2_phylogeny/H3N2_phylogeny.pdf}}
@@
 }
-\end{suppfigure}
+\end{figure}
+\FloatBarrier
 
-\begin{suppfigure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.9\textwidth]{figs/supp_WSNprefs_logoplot/WSN-rescaled_prefs.pdf}}
@@
 }
-\end{suppfigure}
+\end{figure}
+\FloatBarrier
 
-\begin{suppfigure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.8\textwidth]{figs/supp_muteffect_maxfreq_WSN/muteffect_maxfreq_WSN_supp.pdf}}
@@
 }
-\end{suppfigure}
+\end{figure}
+\FloatBarrier
 
-\begin{suppfigure}[H]
+\begin{figure}
 \centerline{\includegraphics[width=0.5\textwidth]{figs/supp_head_stalk_mutfreq/head_stalk_mutfreq.pdf}}
@@
 }
-\end{suppfigure}
+\end{figure}
+\FloatBarrier
 
-\begin{suppdata}[H]
-\caption{\label{suppdata:PerthHA}
-Genbank file giving the full sequence of the bidirectional reverse-genetics plasmid pHW-Perth09-HA-G78D-T212I, which encodes the wildtype HA sequence used in this study.
-}
-\end{suppdata}
+\dataset{pHW_Perth09-HA-G78D-T212I.txt}{Genbank file giving the full sequence of the bidirectional reverse-genetics plasmid pHW-Perth09-HA-G78D-T212I, which encodes the wildtype HA sequence used in this study.}
 
-\begin{suppdata}[H]
-\caption{\label{suppdata:bcsubamp_primers}
-Excel file providing the primers used to generate the barcoded subamplicons for Perth/2009 HA deep sequencing.
-}
-\end{suppdata}
+\dataset{Perth2009_subamplicon_primers.xlsx}{Excel file providing the primers used to generate the barcoded subamplicons for Perth/2009 HA deep sequencing.}
 
-\begin{suppdata}[H]
-\caption{\label{suppdata:prefsfiles}
-Excel file giving the amino-acid preferences in sequential 1, 2, ... numbering of the Perth/2009 HA.
+\dataset{DatasetS3.xlsx}{Excel file giving the amino-acid preferences in sequential 1, 2, ... numbering of the Perth/2009 HA.
 The unscaled preferences for replicates 1, 2, 3-1, and 3-2 are each in a separate tab of the file.
@@
 There are also tabs that give the conversion from sequential to H3 numbering, and a list of the mutations in the 31 clones we Sanger sequenced to evaluate the mutation rate of the mutant plasmid libraries.
-Each tab can simply be exported to CSV for computational analyses.
-}
-\end{suppdata}
+Each tab can simply be exported to CSV for computational analyses.}
 
-\begin{suppdata}[H]
-\caption{\label{suppdata:gisaid}
-Excel file providing acknowledgments and accessions for sequences downloaded from GISAID.
-}
-\end{suppdata}
+\dataset{gisaid_acknowledgment_table.xls}{Excel file providing acknowledgments and accessions for sequences downloaded from GISAID.}
 
-\subsection*{References}
 \bibliography{references}

### 2018-06-30 paper/manuscript_noSI.tex
added `SI Appendix` to all SI references in manuscript

@@
 \usepackage{newfloat}
-\DeclareFloatingEnvironment[name={Figure}]{suppfigure}
-\renewcommand{\thesuppfigure}{S\arabic{suppfigure}}
-
-\usepackage{xr-hyper}
-\externaldocument{SI}
-
-\DeclareFloatingEnvironment[name={Dataset}]{suppdata}
-\renewcommand{\thesuppdata}{S\arabic{suppdata}}
 
@@
 This strain was the H3N2 component of the influenza vaccine from 2010-2012~\cite{who2010d,who2011}.
-Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral growth in cell culture, G78D and T212I (Figure~S1 and Dataset~S1).
+Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral growth in cell culture, G78D and T212I (see SI Appendix, Figure~S1 and Dataset~S1).
 The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.
 
-We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (Figure~S2).
+We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (see SI Appendix, Figure~S2).
 We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
@@
 Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
-A residue was classified as solvent exposed if its relative solvent accessibility was $\geq$ 0.2; Figure~S7 shows that the results are robust to the choice of solvent accessibility cutoff.
+A residue was classified as solvent exposed if its relative solvent accessibility was $\geq$ 0.2.
+The results are robust to the choice of solvent accessibility cutoff (see SI Appendix, Figure~S7).
 Note that the mutational tolerance values are not comparable between the two HAs, but are comparable between domains of the same HA.
@@
 
-However, there are also some differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences (Figure~S3).
+However, there are also some differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences (see SI Appendix, Figure~S3).
 Most surprisingly, the start codon does not show a particularly strong preference for methionine (Figure~\ref{fig:logoplot}).
-We validated that a virus carrying a mutation at this site from methionine to lysine does in fact reach appreciable titers (Figure~S4), perhaps because of alternative translation-initiation at a downstream or upstream start site as has been described for other HAs~\cite{girard2011upstream}.
+We validated that a virus carrying a mutation at this site from methionine to lysine does in fact reach appreciable titers (see SI Appendix, Figure~S4), perhaps because of alternative translation-initiation at a downstream or upstream start site as has been described for other HAs~\cite{girard2011upstream}.
 Our measurements also suggest mutational tolerance at some other sites that are relatively conserved among natural HAs, such as the N-linked glycosylation motifs near the beginning of HA1 and the transmembrane domain (Figure~\ref{fig:logoplot}).
@@
 Specifically, whereas solvent-exposed residues in the head domain are substantially more mutationally tolerant than those in the stalk domain for the WSN/1933 H1 HA, the trend is actually reversed for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
-This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (Figure~S7).
+This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (see SI Appendix, Figure~S7).
 For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
@@
 {\bf Frequency trajectories of individual mutations and their relation to the experimentally measured effects of these mutations.}
-The top panel shows the subset of the full H3N2 HA tree in Figure~S8 from 2004 to 2014.
+The top panel shows the subset of the full H3N2 HA tree in the SI Appendix (Figure~S8) from 2004 to 2014.
 Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
@@
 It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
-Figure~S11 shows a similar layout but colors mutations by whether they are in HA's head or stalk domain.
+The SI Appendix (Figure~S11) shows a similar layout but colors mutations by whether they are in HA's head or stalk domain.
 }
@@
 The vertical black line shows the mean mutation effect for each category.
-The analysis is performed separately for pre-Perth/2009, post-Perth/2009, and unpassaged isolates from the post-Perth/2009 partitions of the tree (see Figure~S8).
+The analysis is performed separately for pre-Perth/2009, post-Perth/2009, and unpassaged isolates from the post-Perth/2009 partitions of the tree (see SI Appendix, Figure~S8).
 }
@@
 To calculate a given mutation's effect, we simply took the logarithm of the ratio of the preferences for the mutant and wildtype amino acids at that site.
-To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~S8).
+To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (see SI Appendix, Figure~S8).
 We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after, since these mutations have not had enough time for their evolutionary fates to be fully resolved.
@@
 This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
-Figure~S10 shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
+The SI Appendix (Figure~S10) shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
 The empirical $P$-value represents the result of 1,000 permutations.
@@
 \subsection*{Measurements made on an H1 HA are less informative for understanding the evolution of H3 influenza}
-To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (Figure~S9), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
+To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (see SI Appendix, Figure~S9), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
 Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the correlations between the H1 experimental measurements and the maximum frequency that mutations reach during H3N2 viral evolution are consistently weaker than those using H3 experimental measurements (compare Figure~\ref{fig:muteffect_maxfreq_WSN} to Figure~\ref{fig:muteffect_maxfreq}A).
@@
 We isolated viral RNA from these passaged viruses and sequenced the HA gene.
-The passaged HA had two mutations, G78D and T212I, which enhanced viral growth as shown in Figure~S1.
+The passaged HA had two mutations, G78D and T212I, which enhanced viral growth as shown in the SI Appendix (see Figure~S1).
 The HA with these two mutations was cloned into pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} to create pHW-Perth09-HA-G78D-T212I and pICR2-Perth09-HA-G78D-T212I.
@@
 We scraped the plates, expanded the cultures in liquid LB (Luria-Bertani Broth) + ampicillin at 37$^{\circ}$C for 3 h with shaking, and then maxiprepped.
-We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (Figure~S2).
+We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (see SI Appendix, Figure~S2).
 
@@
 The exact details of this approach are described in~\cite{doud2016accurate} (also see \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
-The primers used to generate the subamplicons are in Dataset~S2.
+The primers used to generate the subamplicons are in the SI Appendix (Dataset~S2).
 We performed deep sequencing on a lane of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.
@@
 The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
-The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in Dataset~S3.
+The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in the SI Appendix (Dataset~S3).
 Computer code that performs the entirety of this analysis and shows many detailed plots about read depth and other quality control metrics is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/analysis_notebook.ipynb}.
@@
 \subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
-To generate the tree shown in Figure~S8, we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid} (see Dataset~S4), sampling six viruses per month over the time interval of January 1, 1968 to February 1, 2018.
+To generate the tree shown in the SI Appendix (Figure~S8), we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid} (see the SI Appendix, Dataset~S4), sampling six viruses per month over the time interval of January 1, 1968 to February 1, 2018.
 We aligned the resulting 2,189 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.

### 2018-06-30 paper/SI.tex
moved some Methods to SI to get to 10 pages

@@
 %% Comment/remove this line before generating final copy for submission
-\instructionspage
+%\instructionspage
 
@@
 
-\subsection*{Subhead}
-Type or paste text here. This should be additional explanatory text such as an extended technical description of results, full details of mathematical models, extended lists of acknowledgments, etc.
+\subsection*{Codon mutagenesis}
+The codon-mutant libraries were generated in the Perth/2009 HA-G78D-T212I background using the PCR-based approach described in \cite{bloom2014experimentally} with the primer melting-temperature modifications described in \cite{dingens2017comprehensive}, using two rounds of mutagenesis.
+The script to design the mutagenesis primers is at \url{https://github.com/jbloomlab/CodonTilingPrimers}.
+We created three independent libraries, one for each biological replicate.
+The mutant variants were then cloned at high efficiency into the pICR2~\cite{ashenberg2017deep} vector using digestion with BsmBI, ligation with T4 DNA ligase, and electroporation into ElectroMAX DH10B competent cells (Invitrogen 18290015).
+We obtained $>$6 million transformants for each replicate.
+We scraped the plates, expanded the cultures in liquid LB (Luria-Bertani Broth) + ampicillin at 37$^{\circ}$C for 3 h with shaking, and then maxiprepped.
+We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (see SI Appendix, Figure~S2).
 
-\section*{Heading}
-\subsection*{Subhead}
-Type or paste text here. You may break this section up into subheads as needed (e.g., one section on ``Materials'' and one on ``Methods'').
+\subsection*{Barcoded-subamplicon sequencing}
+We generated the HA PCR amplicons for the three plasmid libraries, the three virus libraries, the wildtype plasmid control, and the wildtype virus control using KOD Hot Start Master Mix (EMD Millipore 71842) using the PCR reaction mixture and cycling conditions described in~\cite{bloom2014experimentally} and the P09-HA-For and P09-HA-Rev primers.
+We prepared the sequencing libraries using a barcoded-subamplicon strategy~\cite{wu2014high} to increase the accuracy from deep sequencing.
+The exact details of this approach are described in~\cite{doud2016accurate} (also see \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
+The primers used to generate the subamplicons are in the SI Appendix (Dataset~S2).
+We performed deep sequencing on a lane of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.
 
-\subsection*{Materials}
-Add a Materials subsection if you need to.
+\subsection*{Phylogenetic model comparison and fitting of a stringency parameter}
+For the analysis in Table~\ref{tab:phydms}, we downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
+These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
+We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.2}) to fit the substitution models listed in Table~\ref{tab:phydms}.
 
-\subsection*{Methods}
-Add a Methods subsection if you need to.
+The amino-acid preferences were re-scaled by the stringency parameter using the approach described in \cite{hilton2017phydms}.
+Note that the re-scaling simply puts the amino-acid preferences on a useful scale for visualization in logo plots, and has no effect on any of the quantitative conclusions relating the deep mutational scanning to natural evolution.
+The reason is that all of these conclusions use the effects of mutations as calculated using Equation~\ref{eq:muteffect}, and the re-scaling simply acts as a constant multiplier on all mutational effects (e.g., a scale factor) when re-scaled preferences are converted to mutational effects.
+
+The phylogenetic tree of HA subtypes in Figure~\ref{fig:distance_distribution}A was generated as described in~\cite{doud2018quantifying}.
+
+\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
+To generate the tree shown in the SI Appendix (Figure~S8), we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid} (see the SI Appendix, Dataset~S4), sampling six viruses per month over the time interval of January 1, 1968 to February 1, 2018.
+We aligned the resulting 2,189 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
+Ancestral state reconstruction and branch length timing were performed with TreeTime~\cite{Sagulenko2018}.
+The phylogenetic tree is available as a JSON file on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/data/flu_h3n2_ha_1968_2018_6v_tree.json.gz}.
+The tree was visualized using BALTIC (\url{https://github.com/blab/baltic}).
+
+% notation for frequencies model
+\newcommand{\dx}{\mathrm{d}x}						% change in x
+\newcommand{\dy}{\mathrm{d}y}						% change in y
+\newcommand{\dt}{\mathrm{d}t}						% change in t
+\newcommand{\inertia}{\epsilon}			    % inertia parameter
+\newcommand{\normal}{\mathcal{N}}				% normal distribution
+
+The frequency trajectory of each individual mutation on the phylogeny is estimated following Nextstrain's augur pipeline and as first implemented in Nextflu \cite{neher2015nextflu}.
+Herein, mutation frequency dynamics are modeled according to a Brownian motion diffusion process discretized to one-month intervals.
+The number of viruses sampled in each interval determines the denominator of the mutation frequency calculations.
+Relative to a simple Brownian motion, the expectation includes an ``inertia'' term $\inertia$ that adds velocity to the diffusion and the variance includes a term $x(1-x)$ to scale variance according to frequency following a Wright-Fisher population genetic process.
+This results in the following diffusion process
+\begin{equation}
+x(t+\dt) = \normal\big( x(t) + \inertia \, \dx , \; \dt \, \sigma^2 \, x(t) \, (1-x(t)) \big),
+\end{equation}
+with `volatility' parameter $\sigma^2$.
+The term $\dx$ is the increment in the previous timestep, so that $\dx = x(t) - x(t-\dt)$.
+We used $\inertia = 0.7$ and $\sigma^2 = 0.05$ to maximize fit to empirical trajectory behavior.
+
+We also include an Bernoulli observation model for mutation presence / absence among sampled viruses at timestep $t$.
+This observation model follows
+\begin{equation}
+f(x,t) = \prod_{v\in V} x(t) \prod_{v\notin V} (1-x(t)),
+\end{equation}
+where $v\in V$ represents the set of viruses that have the mutation and $v\notin V$ represents the set of viruses that do not have the mutation.
+Each frequency trajectory is estimated by simultaneously maximizing the likelihood of the process model and the likelihood of the observation model via adjusting frequency trajectory $\mathbf{x}=(x_1, \ldots, x_n)$.
+
+We also repeated the above analyses using only viruses that were sequenced directly without passaging.
+Routine direct sequencing did not begin until the early 2000s \cite{mcwhite2016sequence}.
+To construct a tree with a similar number of viruses as the original analysis, we sampled 30 viruses per month between January 1, 2000 and April 1, 2018, producing a tree with 2,374 unpassaged viruses with augur (commit: \texttt{6d9f708}).
+We included the passaged DMS strain, A/Perth/16/2009, in the resulting tree to enable comparison between pre-Perth and post-Perth clades.
+
+\subsection*{Analysis of mutational shifts}
+To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using MAFFT~\cite{katoh2013mafft}.
+To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2018mapping} and used the RMSD$_{\text{corrected}}$ values as our quantification of the extent of each shift.
+
+For the plots shown in Figure~\ref{fig:RMSD_heatmap}B, any residues falling between Cys-52 and Cys-277 were defined as the head domain, and all other residues were defined as the stalk domain.
+We used the multiple sequence alignment of the HA subtype sequences from \cite{doud2018quantifying} to identify sites that are absolutely conserved across all subtypes, or in the different clades described in Figure~\ref{fig:RMSD_heatmap}.
+
+\subsection*{Validation of individual point mutants}
+To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, T24F, T40V, S287A, and C199(HA2)K, we used site-directed mutagenesis to introduce the amino-acid mutation into the Perth/2009 HA bidirectional reverse genetics plasmid, and verified the sequence of two clones for each mutant by Sanger sequencing.
+We generated these individual mutant viruses carrying GFP in the PB1 segment using a protocol described in \cite{bloom2010permissive,hooper2013mutant}, and the PB2, PA, NP, NA, M, and NS segments of the A/WSN/1933 (H1N1) strain.
+To rescue each mutant GFP-carrying virus in duplicate, we transfected a co-culture of $4 \times 10^5$ 293T-CMV-PB1 and $0.5 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells with the eight reverse genetics plasmids and the pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W plasmid.
+Each well received a transfection mixture of 100 $\mu$L DMEM, 3 $\mu$L BioT transfection reagent, and 250 ng of each plasmid.
+We changed the media in each well with 2 mL IGM eight hours post-transfection.
+At 53 hours post-transfection, transfection supernatants were harvested, clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes, aliquoted, and frozen at -80$^\circ$C.
+
+To titer the GFP-carrying viruses, we plated $1 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells per well in 12-well plates in IGM.
+Four hours after plating, we infected cells with dilutions of viral supernatant.
+At 16 hours post-infection, we selected wells with 1 to 10\% of cells that were GFP-positive and analyzed the fraction of GFP-positive cells by flow cytometry to calculate the titer of infectious particles per $\mu$L.
 

### 2018-06-30 paper/manuscript_noSI.tex
moved some Methods to SI to get to 10 pages

@@
 Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
-Computer code used to analyze the data and produce the results in the paper are on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript}.
+Computer code used to analyze the data is at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript}.
 
@@
 
-The codon-mutant libraries were generated in the Perth/2009 HA-G78D-T212I background using the PCR-based approach described in \cite{bloom2014experimentally} with the primer melting-temperature modifications described in \cite{dingens2017comprehensive}, using two rounds of mutagenesis.
-The script to design the mutagenesis primers is at \url{https://github.com/jbloomlab/CodonTilingPrimers}.
-We created three independent libraries, one for each biological replicate.
-The mutant variants were then cloned at high efficiency into the pICR2~\cite{ashenberg2017deep} vector using digestion with BsmBI, ligation with T4 DNA ligase, and electroporation into ElectroMAX DH10B competent cells (Invitrogen 18290015).
-We obtained $>$6 million transformants for each replicate.
-We scraped the plates, expanded the cultures in liquid LB (Luria-Bertani Broth) + ampicillin at 37$^{\circ}$C for 3 h with shaking, and then maxiprepped.
-We randomly chose 31 clones to Sanger sequence to evaluate the mutation rate (see SI Appendix, Figure~S2).
+The codon-mutant libraries were generated using the approach in \cite{bloom2014experimentally} with the modifications in \cite{dingens2017comprehensive}.
+See the SI Appendix (Supplementary text) for full details.
 
@@
 The HA gene was reverse-transcribed with AccuScript Reverse Transcriptase (Agilent 200820) using primers P09-HA-For (5'-AGCAAAAGCAGGGGATAATTCTATTAATC-3') and P09-HA-Rev (5'-AGTAGAAACAAGGGTGTTTTTAATTACTAATACAC-3').
-
-We generated the HA PCR amplicons for the three plasmid libraries, the three virus libraries, the wildtype plasmid control, and the wildtype virus control using KOD Hot Start Master Mix (EMD Millipore 71842) using the PCR reaction mixture and cycling conditions described in~\cite{bloom2014experimentally} and the P09-HA-For and P09-HA-Rev primers.
-We prepared the sequencing libraries using a barcoded-subamplicon strategy~\cite{wu2014high} to increase the accuracy from deep sequencing.
-The exact details of this approach are described in~\cite{doud2016accurate} (also see \url{https://jbloomlab.github.io/dms_tools2/bcsubamp.html}).
-The primers used to generate the subamplicons are in the SI Appendix (Dataset~S2).
-We performed deep sequencing on a lane of an Illumina HiSeq 2500 using 2 $\times$ 250 bp paired-end reads in rapid-run mode.
+The barcoded-subamplicon library prep and deep sequencing were then performed as in~\cite{doud2016accurate}.
+See the SI Appendix (Supplementary text) for full details.
 
@@
 We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, version \texttt{2.2.5}) to analyze the deep sequencing data.
-The algorithm used to estimate the site-specific amino-acid preferences from the deep sequencing counts is described in~\cite{bloom2015software}.
-The amino-acid preferences for each replicate and for the re-scaled, across-replicate average are provided in the SI Appendix (Dataset~S3).
-Computer code that performs the entirety of this analysis and shows many detailed plots about read depth and other quality control metrics is available on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/analysis_notebook.ipynb}.
+The amino-acid preferences are provided in the SI Appendix (Dataset~S3).
+Computer code and detailed plots about read depth and other quality control metrics are at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/analysis_notebook.ipynb}.
 
-\subsection*{Phylogenetic model comparison and fitting of a stringency parameter}
-For the analysis in Table~\ref{tab:phydms}, we downloaded all full-length H3 HA sequences from the Influenza Virus Resource~\cite{bao2008}, and randomly subsampled two sequences per year.
-These sequences were aligned using MAFFT~\cite{katoh2013mafft} and used to infer a phylogenetic tree using RAxML~\cite{stamatakis2006raxml} with a GTRCAT model of nucleotide substitution.
-We then used \texttt{phydms}~\cite{hilton2017phydms} (\url{https://github.com/jbloomlab/phydms}, version \texttt{2.2.2}) to fit the substitution models listed in Table~\ref{tab:phydms}.
+\subsection*{Phylogenetic model comparison and stringency parameter}
+Phylogenetic model comparisons and fitting of a stringency parameter were performed using \texttt{phydms} as described in~\cite{hilton2017phydms}.
+See the SI Appendix (Supplementary text) for full details.
 
-The amino-acid preferences were re-scaled by the stringency parameter using the approach described in \cite{hilton2017phydms}.
-Note that the re-scaling simply puts the amino-acid preferences on a useful scale for visualization in logo plots, and has no effect on any of the quantitative conclusions relating the deep mutational scanning to natural evolution.
-The reason is that all of these conclusions use the effects of mutations as calculated using Equation~\ref{eq:muteffect}, and the re-scaling simply acts as a constant multiplier on all mutational effects (e.g., a scale factor) when re-scaled preferences are converted to mutational effects.
-
-The phylogenetic tree of HA subtypes in Figure~\ref{fig:distance_distribution}A was generated as described in~\cite{doud2018quantifying}.
-
-\subsection*{Quantification of Shannon entropy and relative solvent accessibility}
+\subsection*{Shannon entropy and relative solvent accessibility}
 We calculated Shannon entropy, $h_r$ for site $r$ as $h_r = - \sum_{x} \pi_{r,x} \log(\pi_{r,x})$, where $\pi_{r,x}$ is the preference for amino acid $x$ at site $r$.
@@
 \subsection*{Quantification of mutational effects}
-We calculated the effects of mutations from the amino-acid preferences that were estimated from the deep mutational scanning data.
-The effect of mutating site $r$ from amino acid $a_1$ to $a_2$ was calculated as
+The effect of mutating site $r$ from amino acid $a_1$ to $a_2$ was quantified as
 \begin{equation}
@@
 
-\subsection*{Inference of human H3N2 phylogenetic tree and calculation of maximum mutation frequencies}
-To generate the tree shown in the SI Appendix (Figure~S8), we applied Nextstrain's augur pipeline~\cite{Hadfield224048} (\url{https://github.com/nextstrain/augur}; commit \texttt{006896d}) to publicly available H3N2 HA sequences from GISAID \cite{shu2017gisaid} (see the SI Appendix, Dataset~S4), sampling six viruses per month over the time interval of January 1, 1968 to February 1, 2018.
-We aligned the resulting 2,189 HA sequences with MAFFT v7.310~\cite{katoh2013mafft} and constructed a maximum likelihood phylogeny from this alignment with RAxML 8.2.10~\cite{stamatakis2006raxml}.
-Ancestral state reconstruction and branch length timing were performed with TreeTime~\cite{Sagulenko2018}.
-The phylogenetic tree is available as a JSON file on GitHub at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/data/flu_h3n2_ha_1968_2018_6v_tree.json.gz}.
-The tree was visualized using BALTIC (\url{https://github.com/blab/baltic}).
-
-% notation for frequencies model
-\newcommand{\dx}{\mathrm{d}x}						% change in x
-\newcommand{\dy}{\mathrm{d}y}						% change in y
-\newcommand{\dt}{\mathrm{d}t}						% change in t
-\newcommand{\inertia}{\epsilon}			    % inertia parameter
-\newcommand{\normal}{\mathcal{N}}				% normal distribution
-
-The frequency trajectory of each individual mutation on the phylogeny is estimated following Nextstrain's augur pipeline and as first implemented in Nextflu \cite{neher2015nextflu}.
-Herein, mutation frequency dynamics are modeled according to a Brownian motion diffusion process discretized to one-month intervals.
-The number of viruses sampled in each interval determines the denominator of the mutation frequency calculations.
-Relative to a simple Brownian motion, the expectation includes an ``inertia'' term $\inertia$ that adds velocity to the diffusion and the variance includes a term $x(1-x)$ to scale variance according to frequency following a Wright-Fisher population genetic process.
-This results in the following diffusion process
-\begin{equation}
-x(t+\dt) = \normal\big( x(t) + \inertia \, \dx , \; \dt \, \sigma^2 \, x(t) \, (1-x(t)) \big),
-\end{equation}
-with `volatility' parameter $\sigma^2$.
-The term $\dx$ is the increment in the previous timestep, so that $\dx = x(t) - x(t-\dt)$.
-We used $\inertia = 0.7$ and $\sigma^2 = 0.05$ to maximize fit to empirical trajectory behavior.
-
-We also include an Bernoulli observation model for mutation presence / absence among sampled viruses at timestep $t$.
-This observation model follows
-\begin{equation}
-f(x,t) = \prod_{v\in V} x(t) \prod_{v\notin V} (1-x(t)),
-\end{equation}
-where $v\in V$ represents the set of viruses that have the mutation and $v\notin V$ represents the set of viruses that do not have the mutation.
-Each frequency trajectory is estimated by simultaneously maximizing the likelihood of the process model and the likelihood of the observation model via adjusting frequency trajectory $\mathbf{x}=(x_1, \ldots, x_n)$.
-
-We also repeated the above analyses using only viruses that were sequenced directly without passaging.
-Routine direct sequencing did not begin until the early 2000s \cite{mcwhite2016sequence}.
-To construct a tree with a similar number of viruses as the original analysis, we sampled 30 viruses per month between January 1, 2000 and April 1, 2018, producing a tree with 2,374 unpassaged viruses with augur (commit: \texttt{6d9f708}).
-We included the passaged DMS strain, A/Perth/16/2009, in the resulting tree to enable comparison between pre-Perth and post-Perth clades.
+\subsection*{H3N2 phylogenetic tree and maximum mutation frequencies}
+The pylogenetic tree was generated using Nextstrain's augur pipeline~\cite{Hadfield224048}, and ancestral state reconstruction and branch length timing were performed with TreeTime~\cite{Sagulenko2018}.
+Frequency trajectories of mutations were estimated following Nextstrain's augur pipeline as first implemented in Nextflu~\cite{neher2015nextflu}.
+See the SI Appendix (Supplementary text) for full details.
 
 \subsection*{Analysis of mutational shifts}
-To compare the Perth/2009 H3 and WSN/1933 H1 HA preferences, we first aligned the wildtype HA sequences using MAFFT~\cite{katoh2013mafft}.
-To quantify the shifts in preference for every alignable site while accounting for experimental noise, we used the approach described in~\cite{haddox2018mapping} and used the RMSD$_{\text{corrected}}$ values as our quantification of the extent of each shift.
-
-For the plots shown in Figure~\ref{fig:RMSD_heatmap}B, any residues falling between Cys-52 and Cys-277 were defined as the head domain, and all other residues were defined as the stalk domain.
-We used the multiple sequence alignment of the HA subtype sequences from \cite{doud2018quantifying} to identify sites that are absolutely conserved across all subtypes, or in the different clades described in Figure~\ref{fig:RMSD_heatmap}.
+We compared the preferences for the Perth/2009 H3 and WSN/1933 H1 HAs using the approach in~\cite{haddox2018mapping}.
+See the SI Appendix (Supplementary text) for full details.
 
 \subsection*{Validation of individual point mutants}
-To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, T24F, T40V, S287A, and C199(HA2)K, we used site-directed mutagenesis to introduce the amino-acid mutation into the Perth/2009 HA bidirectional reverse genetics plasmid, and verified the sequence of two clones for each mutant by Sanger sequencing.
-We generated these individual mutant viruses carrying GFP in the PB1 segment using a protocol described in \cite{bloom2010permissive,hooper2013mutant}, and the PB2, PA, NP, NA, M, and NS segments of the A/WSN/1933 (H1N1) strain.
-To rescue each mutant GFP-carrying virus in duplicate, we transfected a co-culture of $4 \times 10^5$ 293T-CMV-PB1 and $0.5 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells with the eight reverse genetics plasmids and the pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W plasmid.
-Each well received a transfection mixture of 100 $\mu$L DMEM, 3 $\mu$L BioT transfection reagent, and 250 ng of each plasmid.
-We changed the media in each well with 2 mL IGM eight hours post-transfection.
-At 53 hours post-transfection, transfection supernatants were harvested, clarified by centrifugation at 2000 $\times$ \textit{g} for five minutes, aliquoted, and frozen at -80$^\circ$C.
-
-To titer the GFP-carrying viruses, we plated $1 \times 10^5$ MDCK-SIAT1-CMV-PB1-TMPRSS2 cells per well in 12-well plates in IGM.
-Four hours after plating, we infected cells with dilutions of viral supernatant.
-At 16 hours post-infection, we selected wells with 1 to 10\% of cells that were GFP-positive and analyzed the fraction of GFP-positive cells by flow cytometry to calculate the titer of infectious particles per $\mu$L.
+To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, T24F, T40V, S287A, and C199(HA2)K, we examined the supernatant titers of each of these variants after reverse-genetics generation in the context of PB1flank-GFP viruses~\cite{bloom2010permissive,hooper2013mutant}.
+See the SI Appendix (Supplementary text) for full details.
 }

### 2018-06-30 paper/Table1.tex
added Table 1 separate LaTex as well

--- 2018-06-30 paper/manuscript_noSI.tex
@@
-\documentclass[9pt,twocolumn,twoside]{pnas-new}
-% Use the lineno option to display guide line numbers if required.
-% Note that the use of elements such as single-column equations
-% may affect the guide line number alignment.
+\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
+\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
+\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
+%\geometry{landscape}                		% Activate for rotated page geometry
+%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
+\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
+								% TeX will automatically convert eps --> pdf in pdflatex		
+\usepackage{amssymb}
 
-\templatetype{pnasresearcharticle} % Choose template
-% {pnasresearcharticle} = Template for a two-column research article
-% {pnasmathematics} = Template for a one-column mathematics article
-% {pnasinvited} = Template for a PNAS invited submission
+%SetFonts
 
-% added to enable supplementary figures
-\usepackage{placeins}
-\usepackage{newfloat}
+%SetFonts
 
-\setcounter{topnumber}{10}
-\setcounter{bottomnumber}{10}
-\setcounter{totalnumber}{10}
-
-\newcommand{\comment}[1]{{\color{red}[\textsl{#1}]}}
+\pagestyle{empty}
 
 \begin{document}
-
-\title{Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants}
-
-% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-\author[a,d,e,1]{Juhye M. Lee}
-\author[b,f,1]{John Huddleston}
-\author[a,d,e]{Michael B. Doud}
-\author[a,f]{Kathryn A. Hooper}
-\author[g]{Nicholas C. Wu}
-\author[b,c,1]{Trevor Bedford}
-\author[a,c,d,1]{Jesse D. Bloom}
-
-\affil[a]{Basic Sciences Division}
-\affil[b]{Vaccine and Infectious Disease Division}
-\affil[c]{and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA}
-\affil[d]{Department of Genome Sciences}
-\affil[e]{Medical Scientist Training Program}
-\affil[f]{and Molecular and Cellular Biology Program, University of Washington, Seattle, WA 98195, USA}
-\affil[g]{Department of Integrative Structural and Computational Biology, The Scripps Research Institute, La Jolla, CA  92037}
-
-% Please give the surname of the lead author for the running footer
-\leadauthor{Lee}
-
-% Please add here a significance statement to explain the relevance of your work
-% 120 words max
-\significancestatement{
-A key goal in the study of influenza virus evolution is to forecast which viral strains will persist and which ones will die out.
-Here we experimentally measure the effects of all amino-acid mutations to the hemagglutinin protein from a human H3N2 influenza strain on viral growth in cell culture.
-We show that these measurements have utility for distinguishing among viral strains that do and do not succeed in nature.
-Overall, our work suggests that new high-throughput experimental approaches may be useful for understanding virus evolution in nature.
-}
-
-% Please include corresponding author, author contribution and author declaration information
-\authorcontributions{J.M.L.
-performed the experiments; J.M.L.
-and J.H.
-analyzed data with input from T.B.
-and J.D.B.; M.B.D.
-contributed essential reagents and expertise; K.A.H.
-contributed essential reagents; N.C.W.
-made key intellectual contributions; J.M.L.
-and J.D.B.
-wrote the paper with input from J.H.
-and T.B.; all authors were involved in project design and revising the paper.}
-\authordeclaration{The authors declare no conflicts of interest.}
-\equalauthors{\textsuperscript{1} J.M.L and J.H. contributed equally to this work.}
-\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed: tbedford@fredhutch.org, jbloom@fredhutch.org}
-
-% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
-\keywords{influenza virus $|$ hemagglutinin $|$ deep mutational scanning $|$ antigenic drift $|$ epistasis}
-
-\begin{abstract}
-% no more than 250 words.
-Human influenza virus rapidly accumulates mutations in its major surface protein hemagglutinin (HA).
-The evolutionary success of influenza virus lineages depends on how these mutations affect HA's functionality and antigenicity.
-Here we experimentally measure the effects on viral growth in cell culture of all single amino-acid mutations to the HA from a recent human H3N2 influenza virus strain.
-We show that mutations that are measured to be more favorable for viral growth are enriched in evolutionarily successful H3N2 viral lineages relative to mutations that are measured to be less favorable for viral growth.
-Therefore, despite the well-known caveats about cell-culture measurements of viral fitness, such measurements can still be informative for understanding evolution in nature.
-We also compare our measurements for H3 HA to similar data previously generated for a distantly related H1 HA, and find substantial differences in which amino acids are preferred at many sites.
-For instance, the H3 HA has less disparity in mutational tolerance between the head and stalk domains than the H1 HA.
-Overall, our work suggests that experimental measurements of mutational effects can be leveraged to help understand the evolutionary fates of viral lineages in nature --- but only when the measurements are made on a viral strain similar to the ones being studied in nature.
-\end{abstract}
-
-\dates{This manuscript was compiled on \today}
-\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
-
-% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
-% You should only change this length when you've finalised the article contents.
-\verticaladjustment{-2pt}
-
-\maketitle
-\thispagestyle{firststyle}
-\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
-
-\dropcap{S}easonal H3N2 influenza virus evolves rapidly, fixing 3 to 4 amino-acid mutations per year in its hemagglutinin (HA) surface protein~\cite{fitch1997long, bhatt2011genomic}.
-Many of these mutations contribute to the rapid antigenic drift that necessitates frequent updates to the annual influenza vaccine~\cite{smith2004mapping}.
-This evolution is further characterized by competition and turnover among groups of strains called clades bearing different complements of mutations~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,koelle2015effects,bedford2015global}.
-Clades vary widely in their evolutionary success, with some dying out soon after emergence and others going on to take over the virus population.
-Several lines of evidence indicate that successful clades have higher fitness than clades that remain at low frequency~\cite{bedford2011,strelkowa2012clonal,neher2014predicting,luksza2014predictive}.
-A key goal in the study of H3N2 evolution is to identify the features that enable certain clades to succeed as others die out.
-
-\begin{SCfigure*}[\sidecaptionrelwidth][t]
-\centering
-\includegraphics[width=12cm]{dms_overview.pdf}
-\caption{\label{fig:dms_overview}
-{\bf Deep mutational scanning of the Perth/2009 H3 HA.}
-(A) We generated mutant virus libraries using a helper-virus approach~\cite{doud2016accurate}, and passaged the libraries at low MOI to establish a genotype-phenotype linkage and to select for functional HA variants.
-Deep sequencing of the variants before and after selection allowed us to estimate each site's amino-acid preferences.
-(B) The experiments were performed in full biological triplicate.
-We also passaged and deep sequenced library 3 in duplicate.
-(C) Frequencies of nonsynonymous, stop, and synonymous mutations in the mutant plasmid DNA, the passaged mutant viruses, and wildtype DNA and virus controls.
-(D) The Pearson correlations among the amino-acid preferences estimated in each replicate.
-}
-\end{SCfigure*}
-
-Two main characteristics distinguish evolutionarily successful clades from their competitors: greater antigenic change, and efficient viral growth and transmission.
-In principle, experiments could be informative for identifying how mutations affect these features.
-Most work on influenza evolution to date has utilized experimental data to assess the antigenicity of circulating strains~\cite{sun2013using,harvey2016identification,neher2016prediction,koel2013substitutions,chambers2015identification,li2016selection}.
-However, the non-antigenic effects of mutations also play an important role~\cite{pybus2007phylogenetic,strelkowa2012clonal,luksza2014predictive,koelle2015effects}.
-Specifically, due to influenza virus's high mutation rate~\cite{holland1982rapid,steinhauer1987rapid,lauring2010quasispecies} and lack of intra-segment recombination~\cite{boni2008homologous}, deleterious mutations become linked to beneficial ones.
-The resulting accumulation of deleterious mutations can affect non-antigenic properties central to viral fitness~\cite{luksza2014predictive}.
-However, there are no large-scale quantitative characterizations of how mutations to H3N2 HA affect viral growth.
-
-It is now possible to use deep mutational scanning~\cite{fowler2014deep} to measure the functional effects of all single amino-acid mutations to viral proteins~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,haddox2016experimental,qi2015high,haddox2018mapping}.
-However, the only HA for which such large-scale measurements have previously been made is from the highly lab-adapted A/WSN/1933 (H1N1) strain~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate}.
-Here, we measure the effects on viral growth in cell culture of all mutations to the HA of a recent human H3N2 strain.
-We show that these experimental measurements can help discriminate evolutionarily successful mutations from those found in strains that quickly die out.
-However, the utility of the experiments for understanding natural evolution depends on the similarity between the experimental and natural strains: measurements made on an H1 HA are less informative for understanding the evolutionary fate of H3 viral strains.
-
-\section*{Results}
-\label{sec:results}
-
-\subsection*{Deep mutational scanning of HA from a recent strain of human H3N2 influenza virus}
-We performed a deep mutational scan to measure the effects of all amino-acid mutations to HA from the A/Perth/16/2009 (H3N2) strain on viral growth in cell culture.
-This strain was the H3N2 component of the influenza vaccine from 2010-2012~\cite{who2010d,who2011}.
-Relative to the consensus sequence for this HA in Genbank, we used a variant with two mutations that enhanced viral growth in cell culture, G78D and T212I (see SI Appendix, Figure~S1 and Dataset~S1).
-The G78D mutation occurs at low frequency in natural H3N2 sequences, and T212 is a site where a mutation to Ala rose to fixation in human influenza in $\sim$2011.
-
-We mutagenized the entire HA coding sequence at the codon level to create mutant plasmid libraries harboring an average of $\sim$1.4 codon mutations per clone (see SI Appendix, Figure~S2).
-We then generated mutant virus libraries from the mutant plasmids using a helper-virus system that enables efficient generation of complex influenza virus libraries~\cite{doud2016accurate} (Figure~\ref{fig:dms_overview}A).
-These mutant viruses derived all their non-HA genes from the lab-adapted A/WSN/1933 strain.
-Using WSN/1933 for the non-HA genes reduces biosafety concerns, and also helped increase viral titers.
-To further increase viral titers, we used MDCK-SIAT1 cells (Madin-Darby canine kidney cells overexpressing 2,6-sialyltransferase)~\cite{matrosovich2003overexpression} that we engineered to constitutively express TMPRSS2 (Transmembrane Protease, Serine 2), which cleaves the HA precursor to activate it for membrane fusion~\cite{bottcher2006proteolytic, bottcher2010cleavage}.
-
-After generating the mutant virus libraries, we passaged them at low multiplicity of infection (MOI) in cell culture to create a genotype-phenotype link and select for functional HA variants (Figure~\ref{fig:dms_overview}A).
-All experiments were completed in full biological triplicate (Figure~\ref{fig:dms_overview}B).
-We also passaged and deep sequenced library 3 in duplicate (library 3-1 and 3-2) to gauge experimental noise \textit{within} a single biological replicate.
-As a control to measure sequencing and mutational errors, we used the unmutated HA gene to generate and passage viruses carrying wildtype HA.
-
-Deep sequencing of the initial plasmid mutant libraries and the passaged mutant viruses revealed selection for functional HA mutants.
-Specifically, stop codons were purged to 20-45\% of their initial frequencies after correcting for error rates estimated by sequencing the wildtype controls (Figure~\ref{fig:dms_overview}C).
-The incomplete purging of stop codons is likely because genetic complementation due to co-infection~\cite{marshall2013influenza, brooke2013most} enabled the persistence of some virions with nonfunctional HAs.
-We also observed selection against many nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), with their frequencies falling to 30-40\% of their initial values after error correction.
 
@@
  \end{center}
-\addtabletext{Maximum likelihood phylogenetic fit to an alignment of human H3N2 HAs using ExpCM~\cite{hilton2017phydms}, ExpCM in which the experimental measurements are averaged across sites (site avg.), and M0 and M5 versions of the Goldman-Yang (GY94) model~\cite{yang2000codon}.
-Models are compared by Akaike information criterion (AIC)~\cite{posada2004model} computed from the log likelihood (LnL) and number of model parameters.
+{Maximum likelihood phylogenetic fit to an alignment of human H3N2 HAs using ExpCM, ExpCM in which the experimental measurements are averaged across sites (site avg.), and M0 and M5 versions of the Goldman-Yang (GY94) model.
+Models are compared by Akaike information criterion (AIC) computed from the log likelihood (LnL) and number of model parameters.
 The $\omega$ parameter is dN/dS for the Goldman-Yang models, and the relative dN/dS after accounting for the measurements for the ExpCM.
@@
 
-\begin{figure*}[ht]
-\centering
-\includegraphics[width=17cm]{rescaled-avgprefs_prefs.pdf}
-\caption{\label{fig:logoplot}
-{\bf The site-specific amino-acid preferences of the Perth/2009 HA measured in our experiments.}
-The height of each letter is the preference for that amino acid, after taking the average over experimental replicates and re-scaling~\cite{hilton2017phydms} by the stringency parameter in Table~\ref{tab:phydms}.
-The sites are in H3 numbering.
-The top overlay bar indicates whether or not a site is in the set of epitope residues delineated in \cite{wolf2006long}.
-The bottom overlay bar indicates the HA domain (sig. pep. = signal peptide, HA1 ecto. = HA1 ectodomain, HA2 ecto. = HA2 ectodomain, TM = transmembrane domain, cyto. tail. = cytoplasmic tail).
-The letters directly above each logo stack indicate the wildtype amino acid at that site.
-}
-\end{figure*}
-
-We next quantified the reproducibility of our deep mutational scanning across biological and technical replicates.
-We first used the deep sequencing data for each replicate to estimate the preference of each site in HA for all 20 amino acids as described in~\cite{bloom2015software}.
-Because there are 566 residues in HA, there are $566 \times 19 = 10,754$ distinct measurements (the 20 preferences at each site sum to one~\cite{bloom2015software}).
-The correlations of the amino-acid preferences between pairs of replicates are shown in Figure~\ref{fig:dms_overview}D.
-The biological replicates were well-correlated, with Pearson's $R$ ranging from 0.69 to 0.78.
-Replicate 1 exhibited the weakest correlation with other replicates; this replicate also showed the weakest selection against stop and nonsynonymous mutations (Figure~\ref{fig:dms_overview}C), perhaps indicating more experimental noise.
-The two technical replicates 3-1 and 3-2 were only slightly more correlated than pairs of biological replicates, suggesting that bottlenecking of library diversity during viral passage contributes most of the experimental noise.
-
-\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
-How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
-This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard substitution models~\cite{bloom2017identification,hilton2017phydms}.
-Table~\ref{tab:phydms} shows that an ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
-This result indicates that our experiments authentically capture some of the constraints on HA evolution.
-A substitution model in which the amino-acid preferences were averaged across all sites (ExpCM, site avg.) performs no better than conventional substitution models, demonstrating that the reason that our measurements are informative is because they capture site-specific constraints.
-The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
-However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments ($\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for in the deep mutational scanning.
-The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
-Throughout the rest of this paper, we use the amino-acid preferences re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
-These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.
-
-%\begin{SCfigure*}[\sidecaptionrelwidth][t]
-\begin{figure*}[ht]
-\centering
-\includegraphics[width=0.95\textwidth]{mut_tolerance_new.pdf}
-\caption{\label{fig:mut_tolerance}
-{\bf Mutational tolerance of each site in H3 and H1 HAs.}
-(A) Mutational tolerance as measured in the current study is mapped onto the structure of the H3 trimer (PDB 4O5N;~\cite{lee2014receptor}).
-Mutational tolerance of the WSN/1933 H1 HA as measured in \cite{doud2016accurate} is mapped onto the structure of the H1 trimer (PDB 1RVX;~\cite{gamblin2004structure}).
-Different color scales are used because measurements are comparable among sites within the same HA, but not necessarily across HAs.
-Both trimers are shown in the same orientation.
-For each HA, the structure at left shows a surface representation of the full trimer, while the structure at right side shows a ribbon representation of just one monomer.
-The sialic acid receptor is shown in red sticks.
-(B) The mutational tolerance of solvent-exposed residues in the head and stalk domains of the Perth/2009 H3 HA (purple) and WSN/1933 H1 HA (gold).
-Residues falling in between the two cysteines at sites 52 and 277 were defined as belonging to the head domain, while all other residues were defined as the stalk domain.
-A residue was classified as solvent exposed if its relative solvent accessibility was $\geq$ 0.2.
-The results are robust to the choice of solvent accessibility cutoff (see SI Appendix, Figure~S7).
-Note that the mutational tolerance values are not comparable between the two HAs, but are comparable between domains of the same HA.
-}
-\end{figure*}
-
-Examination of Figure~\ref{fig:logoplot} reveals that the experimentally measured amino-acid preferences generally agree with existing knowledge about HA's structure and function.
-For instance, sites that form structurally important disulfide bridges (sites 52 \& 277, 64 \& 76, 97 \& 139, 281 \& 305, 14 \& 137-HA2, 144-HA2 \& 148-HA2)~\cite{waterfield1981disulphide} strongly prefer cysteine.
-At residues involved in receptor binding, there are strong preferences for the amino acids that are known to be involved in binding sialic acid, such as Y98, D190, W153, and S228~\cite{weis1988structure,martin1998studies,nobusawa2000change,yang2015structure}.
-A positively charged amino acid at site 329 is important for cleavage of the HA0 precursor into the mature form~\cite{stech2005new}, and this site strongly prefers arginine.
-
-However, there are also some differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences (see SI Appendix, Figure~S3).
-Most surprisingly, the start codon does not show a particularly strong preference for methionine (Figure~\ref{fig:logoplot}).
-We validated that a virus carrying a mutation at this site from methionine to lysine does in fact reach appreciable titers (see SI Appendix, Figure~S4), perhaps because of alternative translation-initiation at a downstream or upstream start site as has been described for other HAs~\cite{girard2011upstream}.
-Our measurements also suggest mutational tolerance at some other sites that are relatively conserved among natural HAs, such as the N-linked glycosylation motifs near the beginning of HA1 and the transmembrane domain (Figure~\ref{fig:logoplot}).
-We validated that viruses with mutations to the glycosylation motifs at sites 22 or 38, or a site in the transmembrane domain, do in fact grow to high titers (Figures~S5 and S6, respectively).
-The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function in nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.
-
-\subsection*{There is less difference in mutational tolerance between the HA head and stalk domains for H3 than for H1}
-Our experiments measure which amino acids are tolerated at each HA site under selection for viral growth.
-We can therefore use our experimentally measured amino-acid preferences to calculate the inherent mutational tolerance of each site, which we quantify as the Shannon entropy of the re-scaled preferences.
-In prior mutational studies of H1 HAs, the stalk domain was found to be substantially less mutationally tolerant than the globular head~\cite{thyagarajan2014inherent,wu2014high,doud2016accurate,heaton2013genome}.
-
-We performed a similar analysis using our new data for the Perth/2009 H3 HA.
-Surprisingly, the head domain of the H3 HA is \emph{not} more mutationally tolerant than the stalk domain (Figure~\ref{fig:mut_tolerance}).
-Specifically, whereas solvent-exposed residues in the head domain are substantially more mutationally tolerant than those in the stalk domain for the WSN/1933 H1 HA, the trend is actually reversed for the Perth/2009 H3 HA (Figure~\ref{fig:mut_tolerance}B).
-This difference between the relative mutational tolerances of the H1 and H3 HAs is robust to the cutoff used to define surface residues (see SI Appendix, Figure~S7).
-For instance, for the H3 HA, the short helix A in the stalk domain is as mutationally tolerant as many surface-exposed residues in the head domain---something that is not the case for the H1 HA.
-Helix A forms part of the epitope of many broadly neutralizing anti-stalk antibodies~\cite{mallajosyula2014influenza,laursen2013broadly,chai2016two}.
-
-We also see high mutational tolerance in many of the known antigenic regions of H3 HA~\cite{wiley1981structural}.
-For instance, antigenic region B is an immunodominant area, and many recent major antigenic drift mutations have occurred in this region~\cite{chambers2015identification,koel2013substitutions,popova2012immunodominance}.
-We find that the most distal portion of the globular head near the 190-helix, which is part of antigenic region B, is highly tolerant of mutations (Figure~\ref{fig:mut_tolerance}A).
-Antigenic region C is also notably mutationally tolerant.
-
-Many residues inside HA's receptor binding pocket are known to be highly functionally constrained~\cite{wilson1981structure,martin1998studies}, and our data indicates that these sites are relatively mutationally intolerant in both H3 and H1 HAs (Figure~\ref{fig:mut_tolerance}A).
-In contrast, the residues surrounding the receptor binding pocket are fairly mutationally tolerant, which may contribute to the rapidity of influenza's antigenic evolution, since mutations at these sites can have large effects on antigenicity~\cite{wiley1981structural,koel2013substitutions}.
-
-\begin{figure}[t!]
-\centering
-\includegraphics[width=0.95\linewidth]{frequency_trajectory_ex.pdf}
-\caption{\label{fig:frequency_trajectory}
-{\bf Frequency trajectories of individual mutations and their relation to the experimentally measured effects of these mutations.}
-The top panel shows the subset of the full H3N2 HA tree in the SI Appendix (Figure~S8) from 2004 to 2014.
-Circles indicate individual amino-acid mutations, and are colored according to the mutational effect measured in our deep mutational scanning (negative values indicate mutations measured to be deleterious to viral growth).
-The Perth/2009 strain is labeled with a star, and nodes in the clade containing the Perth/2009 strain were excluded from our analyses.
-The bottom panel shows the frequency trajectory of each mutation, with trajectories colored according to the mutational effects as in the top panel.
-It is clear that most mutations that reach high frequency are measured to be relatively favorable in our experiments.
-The SI Appendix (Figure~S11) shows a similar layout but colors mutations by whether they are in HA's head or stalk domain.
-}
-\end{figure}
-
-\begin{SCfigure*}[\sidecaptionrelwidth][h!]
-\centering
-\includegraphics[width=12cm]{muteffect_by_maxfreq.pdf}
-\caption{\label{fig:muteffect_maxfreq}
-{\bf Experimental measurements are informative about the evolutionary fate of viral mutations.}
-(A) Correlation between the effects of mutations as measured in our deep mutational scanning of the Perth/2009 HA and the maximum frequency reached by these mutations in nature.
-The plots show Spearman $\rho$ and an empirical $P$-value representing the proportion of 10,000 permutations of the experimental measurements for which the permuted $\rho$ was greater than or equal to the observed $\rho$.
-(B) The distribution of mutational effects partitioned by maximum mutation frequency.
-The vertical black line shows the mean mutation effect for each category.
-The analysis is performed separately for pre-Perth/2009, post-Perth/2009, and unpassaged isolates from the post-Perth/2009 partitions of the tree (see SI Appendix, Figure~S8).
-}
-\end{SCfigure*}
-
-\subsection*{Our measurements can help distinguish between mutations that reach low and high frequencies in nature}
-Mutations occurring in the H3N2 virus population experience widely varying evolutionary fates (Figure~\ref{fig:frequency_trajectory}).
-Some mutations appear, spread and fix in the population, while others briefly circulate before disappearing.
-We take the maximum frequency reached by a mutation as a coarse indicator of its effect on fitness, since favorable mutations generally reach higher frequencies than unfavorable ones~\cite{ewens2012mathematical}.
-Here, we follow the population genetic definition of \textit{mutation} and track the outcome of each individual mutation event, e.g.\ although R142G occurs multiple times on the phylogeny we track each of these mutations occurring on different backgrounds separately.
-As such, each mutation is shown as a separate circle on a separate branch in Figure \ref{fig:frequency_trajectory}.
-However, because multiple mutations on the same phylogeny branch cannot be disentangled, when multiple mutations occurred on a single branch, we assigned a single mutational effect based on the sum of effects of each mutation.
-
-After annotating mutations and their frequencies on the phylogeny in this way (Figure~\ref{fig:frequency_trajectory}), it is visually obvious that there are relatively few circulating mutations that we measure to be strongly deleterious---and that such deleterious mutations rarely reach high frequency when they do occur.
-
-We next sought to quantify the correlation between a mutation's experimentally measured effect and the maximum frequency it attained during natural evolution.
-To calculate a given mutation's effect, we simply took the logarithm of the ratio of the preferences for the mutant and wildtype amino acids at that site.
-To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (see SI Appendix, Figure~S8).
-We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after, since these mutations have not had enough time for their evolutionary fates to be fully resolved.
-We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
-We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
-In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.15$).
-The similar effect sizes for both the pre- and post-Perth partitions shows that our experimental measurements can help explain the evolutionary fates of mutations in strains that post-date the experimentally studied strain, as well as to retrospectively analyze mutations that precede the experimental strain.
-
-Many of the HAs in sequence databases are from viral isolates that were passaged in cell-culture or eggs, which can cause lab-adaptation mutations that confound evolutionary analyses~\cite{mcwhite2016sequence}.
-To check that our results were robust to such lab-adaptation mutations, we repeated our analysis using only HA sequences derived from viruses that had not been passaged in the lab.
-Because sequencing of unpassaged primary isolates has only recently become commonplace, we could only perform this analysis for the post-Perth partition of the phylogenetic tree.
-Figure~\ref{fig:muteffect_maxfreq}A shows that the correlation between our measured mutational effects and the maximum frequency was even stronger for mutations from unpassaged viral isolates ($\rho = 0.24$).
-
-\begin{SCfigure*}[\sidecaptionrelwidth][t]
-\centering
-\includegraphics[width=12cm]{muteffect_by_maxfreq_WSN.pdf}
-\caption{\label{fig:muteffect_maxfreq_WSN}
-{\bf Experimental measurements on an H1 HA are less informative about the evolutionary fate of H3N2 mutations.}
-This figure repeats the analysis of the H3N2 mutation frequencies in Figure~\ref{fig:muteffect_maxfreq}A, but uses the deep mutational scanning data for an H1 HA as measured in \cite{doud2016accurate}.
-The SI Appendix (Figure~S10) shows the histograms comparable to those in Figure~\ref{fig:muteffect_maxfreq}B.
-The empirical $P$-value represents the result of 1,000 permutations.
-}
-\end{SCfigure*}
-
-The trends in Figure~\ref{fig:muteffect_maxfreq}A are most strongly driven by the behavior of substantially deleterious mutations.
-We investigated this further by partitioning mutations into those that reach low, medium and high frequencies, and those that fix in the population (Figure~\ref{fig:muteffect_maxfreq}B).
-The mutations that reach higher frequencies have a more favorable mean effect.
-Mutations measured to be substantially deleterious almost never reach high frequency.
-Overall, these results demonstrate that measurements of how mutations affect viral growth in cell culture are informative for understanding the fates of these mutations in nature: in particular, if a mutation is measurably deleterious to viral growth, that mutation is unlikely to prosper in nature.
-
-\subsection*{Measurements made on an H1 HA are less informative for understanding the evolution of H3 influenza}
-To determine how broadly experimental measurements can be generalized across HAs, we repeated the foregoing analysis of H3N2 mutation frequencies, but using mutational effects measured in our prior deep mutational scanning of the WSN/1933 H1 HA~\cite{doud2016accurate} (see SI Appendix, Figure~S9), which is highly diverged from the Perth/2009 H3 HA (the two HAs only have 42\% protein sequence identity).
-Figure~\ref{fig:muteffect_maxfreq_WSN} shows that the correlations between the H1 experimental measurements and the maximum frequency that mutations reach during H3N2 viral evolution are consistently weaker than those using H3 experimental measurements (compare Figure~\ref{fig:muteffect_maxfreq_WSN} to Figure~\ref{fig:muteffect_maxfreq}A).
-Therefore, the utility of an experiment for understanding natural evolution degrades as the experimental sequence becomes more diverged from the natural sequences that are being studied.
-
-\begin{SCfigure*}[\sidecaptionrelwidth][b]
-\centering
-\includegraphics[width=12cm]{distance_distribution.pdf}
-\caption{\label{fig:distance_distribution}
-{\bf There are large shifts in the effects of mutations between H1 and H3 HAs.}
-(A) Phylogenetic tree of HA subtypes, with the WSN/1933 H1 and Perth/2009 H3 HAs labeled.
-(B) All pairwise correlations of the amino-acid preferences measured in the three individual deep mutational scanning replicates in the current study and the three replicates in prior deep mutational scanning of the H1 HA~\cite{doud2016accurate}.
-Comparisons between H3 replicates are in purple, those between H1 replicates are in brown, and those across H1 and H3 replicates are in gray.
-$R$ indicates the Pearson correlation coefficient.
-(C) We calculated the shift in amino-acid preferences at each site between H3 and H1 HAs using the method in \cite{haddox2018mapping}, and plotted the distribution of shifts for all sites.
-The shifts between H3 and H1 (yellow) are much larger than the null distribution (blue) expected if all differences are due to experimental noise.
-The shifts are also much larger than those previously observed between two variants of HIV Env that share 86\% amino-acid identity (pink).
-However, the shifts between H3 and H1 are less than the differences between HA and HIV Env (green).
-}
-\end{SCfigure*}
-
-\subsection*{There are large differences between H3 and H1 HAs in the amino-acid preferences of many sites}
-An obvious hypothesis for why the H1 deep mutational scanning is less useful for understanding the evolution of H3N2 influenza viruses is that the effect of the same mutation is often different between these two HA subtypes.
-To determine if this is the case, we examined how much the amino-acid preferences of homologous sites have shifted between H3 and H1 HAs.
-Prior experiments have found only modest shifts in amino-acid preferences between two variants of influenza nucleoprotein with 94\% amino-acid identity~\cite{doud2015site} and variants of HIV envelope (Env) with 86\% amino-acid identity~\cite{haddox2018mapping}.
-However, the H1 and H3 HAs are far more diverged, with only 42\% amino-acid identity (Figure~\ref{fig:distance_distribution}A).
-One simple way to investigate the extent of shifts in amino-acid preferences is to correlate measurements from independent deep mutational scanning replicates on H1 and H3 HAs.
-Figure~\ref{fig:distance_distribution}B shows that replicate measurements on the same HA variant are more correlated than those on different HA variants.
-
-To more rigorously quantify shifts in amino-acid preferences after correcting for experimental noise, we used the statistical approach in~\cite{doud2015site,haddox2018mapping}.
-Figure~\ref{fig:distance_distribution}C shows the distribution of shifts in amino-acid preferences between H3 and H1 HAs after correcting for experimental noise.
-Although some sites have small shifts near zero, many sites have large shifts.
-These shifts between H3 and H1 are much larger than expected from the null distribution that would be observed purely from experimental noise.
-They are also much larger than the shifts previously observed between two HIV Envs with 86\% amino-acid identity~\cite{haddox2018mapping}.
-However, the typical shift between H3 and H1 is still smaller than that observed when comparing HA to the non-homologous HIV Env protein.
-Therefore, there are very substantial shifts in mutational effects between highly diverged HA homologs, although the effects of mutations remain more similar than for non-homologous proteins.
-
-\subsection*{Properties associated with the shifts in amino-acid preferences between H3 and H1 HAs}
-What features distinguish the sites with shifted amino-acid preferences between H3 and H1 HAs?
-The sites of large shifts do not obviously localize to one specific region of HA's structure (Figure~\ref{fig:RMSD_heatmap}A).
-However, at the domain level, sites in HA's stalk tend to have smaller shifts than sites in HA's globular head (Figure~\ref{fig:RMSD_heatmap}B).
-The HA stalk domain is also more conserved in sequence~\cite{nobusawa1991comparison}, suggesting that conservation of amino-acid sequence is correlated with conservation of amino-acid preferences.
-Consistent with this idea, sites that are absolutely conserved across all 18 HA subtypes are significantly less shifted than sites that are variable across HA subtypes (Figure~\ref{fig:RMSD_heatmap}B).
-Presumably these sites are under consistent functional constraint across all HAs.
-
-Despite their high sequence divergence, H1 and H3 adopt very similar protein folds~\cite{ha2002h5,russell2004h1}.
-However, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among different HA subtypes~\cite{ha2002h5,russell2004h1}.
-Previous work has defined clades of structurally related HA subtypes~\cite{ha2002h5,russell2004h1}.
-One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}A).
-Sites that are conserved at different amino-acid identities in these two clades tend to have exceptionally large shifts in amino acid preferences (Figure~\ref{fig:RMSD_heatmap}B).
-The clade containing H1 has an upward translation of the globular head relative to the clade containing H3.
-This structural shift has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
-Specifically, the clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
-In deep mutational scanning of the H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
-In contrast, the interhelical loop in H3 HA makes a sharper and shorter turn which is facilitated by a Gly at 75(HA2).
-In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations.
-Therefore, some of the shifts in HA amino-acid preferences can be rationalized in terms of changes in HA structure.
-
-%\begin{SCfigure*}[\sidecaptionrelwidth][t]
-\begin{figure}
-\centering
-\includegraphics[width=\columnwidth]{RMSD_heatmap.pdf}
-\caption{\label{fig:RMSD_heatmap}
-{\bf Sites with strongly shifted amino-acid preferences between H3 and H1 HAs.}
-(A) The shift in amino-acid preferences between the H3 and H1 HA at each site as calculated in Figure~\ref{fig:distance_distribution}C is mapped onto the structure of the H3 HA.
-(B) Amino-acid preferences of sites in the stalk domain are less shifted than those in the head domain.
-Sites absolutely conserved in all 18 HA subtypes are less shifted than other sites.
-Sites with one amino-acid identity in the clade containing H1, H2, H5, and H6 and another identity in the clade containing H3, H4, and H14 are more shifted than other sites.
-(C) Sites 107 and 75(HA2) help determine the different orientation of the globular head domain in H1 versus H3 HAs.
-These sites are shown in spheres on the structure of H1 and H3 and colored as in panel (A), and the experimentally measured amino-acid preferences in the H1 and H3 HAs are shown.
-One monomer is in dark gray, while the HA1 domain of the neighboring monomer is in lighter gray.
-}
-%\end{SCfigure*}
-\end{figure}
-
-\section*{Discussion}
-\label{sec:discussion}
-We have measured the effects of all possible single amino-acid mutations to the Perth/2009 H3 HA on viral growth in cell culture and demonstrated that these measurements have some value for understanding the evolutionary fate of these mutations in nature.
-Specifically, mutations measured to be more beneficial for viral growth tend to reach higher frequencies in nature than mutations measured to be more deleterious for viral growth.
-The fact that our experiments can help identify evolutionary successful mutations suggests that they might inform evolutionary forecasting.
-In their landmark paper introducing predictive viral fitness models that accounted for both antigenic and non-antigenic mutations, {\L}uksza and L{\"a}ssig~\cite{luksza2014predictive} noted that the models could in principle be improved by integrating ``diverse genotypic and phenotypic data'' that more realistically represented the effects of specific mutations.
-Our work suggests that deep mutational scanning may be able to provide such data.
-
-It is important to emphasize that measurements of viral growth in cell culture do \emph{not} represent true fitness in nature.
-Indeed, a vast amount of work in virology has chronicled the ways in which experiments can select for lab artifacts or fail to capture important pressures that are relevant in nature~\cite{daniels1985fusion,sun2010modifications,lee2013comparison,wu2017structural}.
-As an example, although we identified G78D as favorable for viral growth in cell culture, this mutation never fixes in nature.
-Mutations in viral genes other than HA are also important in determining strain success~\cite{memoli2009recent,raghwani2017selection}.
-Given these caveats, it might seem surprising that measuring viral growth in cell culture can be informative about the success of viral strains in nature.
-Yet, prior to our work, there were no comprehensive studies of the functional effects of mutations to H3 HA on any property that even resembled viral fitness in nature, and modeling work has either omitted the non-antigenic effects of mutations~\cite{sun2013using,harvey2016identification,neher2016prediction} or assumed that all non-epitope mutations had equivalent deleterious effects~\cite{luksza2014predictive}.
-The strength of our measurements are not that they perfectly capture fitness in nature, but that they are systematic and quantitative---and so represent an improvement over no information at all.
-We suspect that performing similar experiments using more realistic and complex selections (e.g., ferrets or primary human airway cultures) might further improve their utility and possibly their generalizability to more divergent strains.
-
-We measured the effects of all single amino-acid mutations to a specific HA, and then generalized these measurements to other H3N2 HAs from a 50-year timespan.
-These generalizations will only be valid to the extent that the effects of mutations are conserved during HA's evolution.
-Extensive prior work has shown that epistasis can shift the effects of mutations as proteins evolve~\cite{pollock2012amino,shah2015contingency,gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}.
-Our work suggests that measurements on a HA from single human H3N2 viral strain can be usefully generalized to at least some extent across the entire evolutionary history of human H3N2 HA.
-On the other hand, when we compared our measurements for an H3 HA to prior measurements on H1 HA, we found substantial shifts at many sites---much greater than those observed in prior protein-wide comparisons of more closely related homologs~\cite{doud2015site,haddox2018mapping}.
-Further investigation of how mutational effects shift as proteins diverge will be important for determining how broadly any given experiment can be generalized when attempting to make evolutionary forecasts.
-
-Our work did not characterize the antigenic effects of mutations, which also play an important role in determining strain success in nature~\cite{koel2013substitutions,neher2016prediction}.
-However, our basic selection and deep-sequencing approach can be harnessed to completely map how mutations affect antibody recognition~\cite{doud2017complete,doud2018quantifying}.
-But so far, experiments using this approach have not examined antibodies or sera that are relevant to driving the evolution of H3N2 influenza~\cite{doud2017complete,doud2018quantifying}, or have used relevant sera but examined a non-comprehensive set of mutations~\cite{li2016selection}.
-Future experiments that completely map how HA mutations affect recognition by human sera seem likely to be especially fruitful for informing viral forecasting.
-
-{\small
-\section*{Materials and Methods}
-\subsection*{Data and computer code}
-Deep sequencing data are available from the Sequence Read Archive under BioSample accessions SAMN08102609 and SAMN08102610.
-Computer code used to analyze the data is at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript}.
-
-\subsection*{HA numbering}
-Sites are in H3 numbering, with the signal peptide in negative numbers, HA1 in plain numbers, and HA2 denoted with "(HA2)".
-Sequential 1, 2, ...
-numbering of the Perth/2009 HA can be converted to H3 numbering by subtracting 16 for the HA1 subunit, and subtracting 345 for the HA2 subunit.
-
-\subsection*{Creation of MDCK-SIAT1-TMPRSS2 cells}
-When growing influenza virus in cell culture, trypsin is normally added to cleave HA into its mature form.
-To obviate the need for trypsin, we engineered MDCK-SIAT1 cells and MDCK-SIAT1-CMV-PB1~\cite{bloom2010permissive} cells to constitutively express the TMPRSS2 protease, which cleaves and activates HA in the human airways~\cite{bottcher2006proteolytic,bottcher2010cleavage}.
-The human TMPRSS2 cDNA ORF was ordered from OriGene (NM\_005656) and cloned into a pHAGE2 lentiviral vector under an EF1$\alpha$-Int promoter followed by an IRES driving expression of mCherry to create plasmid pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
-We used the lentiviral vector to transduce MDCK-SIAT1 or MDCK-SIAT1-CMV-PB1 cells, and sorted an intermediate mCherry-positive population by flow cytometry.
-We refer to the sorted bulk population as MDCK-SIAT1-TMPRSS2 cells or MDCK-SIAT1-CMV-PB1-TMPRSS2 cells.
-There is no selectable marker for the TMPRSS2; however, we maintain the cells at low passage number, and have seen no indication that they lose their ability to support the growth of viruses with H3 HAs in the absence of exogenous trypsin.
-
-\subsection*{Generation of HA codon mutant plasmid libraries}
-HA and NA genes for the Perth/2009 viral strain were cloned from recombinant virus obtained from BEI Resources (NR-41803) into the pHW2000~\citep{hoffmann2000dna} influenza reverse-genetics plasmids to create pHW-Perth09-HA and pHW-Perth09-NA.
-
-We initially created a virus with the HA and NA from Perth/2009 and internal genes from WSN/1933, and passaged it in cell culture to test its genetic stability.
-To generate this virus, we transfected a co-culture of 293T and MDCK-SIAT1-TMPRSS2 in D10 media (DMEM, or Dulbecco's Modified Eagle Medium, supplemented with 10\% heat-inactivated FBS, or Fetal Bovine Serum, 2 mM L-glutamine, 100 U of penicillin/mL, and 100 $\mu$g of streptomycin/mL) with equal amounts of pHW-Perth09-HA, pHW-Perth09-NA, the pHW18* series of plasmids~\cite{hoffmann2000dna} for all non-HA/NA viral genes, and pHAGE2-EF1aInt-TMPRSS2-IRES-mCherry-W.
-The next day we changed the media to influenza growth media (IGM, consisting of Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/mL, 100 $\mu$g of streptomycin/mL, and 100 $\mu$g of calcium chloride/mL (no trypsin was added since there was TMPRSS2), and then collected the viral supernatant at 72 hours post-transfection.
-This viral supernatant was blind passaged in MDCK-SIAT1-TMPRSS2 a total of six additional times.
-We isolated viral RNA from these passaged viruses and sequenced the HA gene.
-The passaged HA had two mutations, G78D and T212I, which enhanced viral growth as shown in the SI Appendix (see Figure~S1).
-The HA with these two mutations was cloned into pHW2000~\citep{hoffmann2000dna} and pICR2~\cite{ashenberg2017deep} to create pHW-Perth09-HA-G78D-T212I and pICR2-Perth09-HA-G78D-T212I.
-For all subsequent experiments, we used viruses with the HA containing these two mutations to improve titers and viral genetic stability, and this is the HA that we refer to as Perth/2009 in the manuscript text.
-We used all non-HA genes (including NA) from WSN/1933 to help increase titers and reduce biosafety concerns.
-
-The codon-mutant libraries were generated using the approach in \cite{bloom2014experimentally} with the modifications in \cite{dingens2017comprehensive}.
-See the SI Appendix (Supplementary text) for full details.
-
-\subsection*{Generation and passaging of mutant viruses}
-The mutant virus libraries were generated using the helper-virus approach described in \cite{doud2016accurate} with several modifications, most notably the cell line used.
-Briefly, we transfected $5 \times 10^5$ MDCK-SIAT1-TMPRSS2 cells in suspension with 937.5 ng each of four protein expression plasmids encoding the ribonucleoprotein complex (HDM-Nan95-PA, HDM-Nan95-PB1, HDM-Nan95-PB2, and HDM-Aichi68-NP)~\cite{gong2013stability}, and 1250 ng of one of the three pICR2-mutant-HA libraries (or the wildtype control) using Lipofectamine 3000 (ThermoFisher L3000008).
-We allowed the transfected cells to adhere in 6-well plates and four hours later changed the media to D10 media.
-Eighteen hours after transfection, we infected the cells with the WSN/1933 HA-deficient helper virus~\cite{doud2016accurate} by preparing an inoculum of 500 TCID$_{50}$ per $\mu$L of helper virus (as computed on HA-expressing cells) in IGM, aspirating the D10 media from the cells, and adding 2 mL of the helper-virus inoculum to each well.
-After three hours, we changed the media to fresh IGM.
-At 24 hours after helper-virus infection, we harvested the viral supernatants for each replicate, froze aliquots at -80$^{\circ}$C, and titered in MDCK-SIAT1-TMPRSS2 cells.
-The titers were 92, 536, 536, and 734 TCID$_{50}$ per $\mu$L for the three library replicates and the wildtype control, respectively.
-
-We passaged $9 \times 10^5$ TCID$_{50}$ of the transfection supernatants at an MOI of 0.0035 TCID$_{50}$ per cell.
-To do this, we plated $4.6 \times 10^6$ MDCK-SIAT1-TMPRSS2 cells per dish in fifteen 15-cm dishes in D10 media, and allowed the cells to grow for 24 hours, at which time they were at $\sim 1.7 \times 10^7$ cells per dish.
-We replaced the media in each dish with 25 mL of an inoculum of 2.5 TCID$_{50}$ of virus per $\mu$L in IGM.
-Three hours post-infection, we replaced the inoculum with fresh IGM for replicates 1, 2, and 3-2.
-We did not perform a media change for replicate 3-1.
-As can be seen in Figure~\ref{fig:dms_overview}D, the media change does not appear to have a substantial effect, as replicate 3-1 looks comparable to the other replicates.
-We collected viral supernatant for sequencing 48 hours post-infection.
-
-\subsection*{Barcoded subamplicon sequencing}
-To extract viral RNA from the three replicate HA virus libraries and the wildtype HA virus, we clarified the viral supernatant by centrifuging at 2000 x \textit{g} for five minutes, then ultracentrifuged 24 mL of the clarified supernatant at 22,000 rpm for 1.5 h at 4$^{\circ}$C in a Beckman Coulter SW28 rotor, and extracted RNA using the Qiagen RNeasy Mini Kit by resuspending the viral pellet in 400 $\mu$L of buffer RLT supplemented with $\beta$-mercaptoethanol, pipetting 30 times, transferring the liquid to a microcentrifuge tube, adding 600 $\mu$L 70\% ethanol, and proceeding according to the manufacturer's instructions.
-The HA gene was reverse-transcribed with AccuScript Reverse Transcriptase (Agilent 200820) using primers P09-HA-For (5'-AGCAAAAGCAGGGGATAATTCTATTAATC-3') and P09-HA-Rev (5'-AGTAGAAACAAGGGTGTTTTTAATTACTAATACAC-3').
-The barcoded-subamplicon library prep and deep sequencing were then performed as in~\cite{doud2016accurate}.
-See the SI Appendix (Supplementary text) for full details.
-
-\subsection*{Analysis of deep sequencing data}
-We used the \texttt{dms\_tools2} software package~\cite{bloom2015software} (\url{https://github.com/jbloomlab/dms_tools2}, version \texttt{2.2.5}) to analyze the deep sequencing data.
-The amino-acid preferences are provided in the SI Appendix (Dataset~S3).
-Computer code and detailed plots about read depth and other quality control metrics are at \url{https://github.com/jbloomlab/Perth2009-DMS-Manuscript/blob/master/analysis_code/analysis_notebook.ipynb}.
-
-\subsection*{Phylogenetic model comparison and stringency parameter}
-Phylogenetic model comparisons and fitting of a stringency parameter were performed using \texttt{phydms} as described in~\cite{hilton2017phydms}.
-See the SI Appendix (Supplementary text) for full details.
-
-\subsection*{Shannon entropy and relative solvent accessibility}
-We calculated Shannon entropy, $h_r$ for site $r$ as $h_r = - \sum_{x} \pi_{r,x} \log(\pi_{r,x})$, where $\pi_{r,x}$ is the preference for amino acid $x$ at site $r$.
-
-We quantified the absolute solvent accessibility of each site of the H3 HA (PDB 4O5N;~\cite{lee2014receptor}) or the H1 HA (PDB 1RVX;~\cite{gamblin2004structure}) structure using DSSP~\cite{kabsch1983}.
-We then normalized to a relative solvent accessibility using the absolute accessibilities in~\cite{tien2013}.
-
-\subsection*{Quantification of mutational effects}
-The effect of mutating site $r$ from amino acid $a_1$ to $a_2$ was quantified as
-\begin{equation}
-\label{eq:muteffect}
-\log_2 \frac{\pi_{r,a_2}}{\pi_{r,a_1}}
-\end{equation}
-where $\pi_{r,a_1}$ and $\pi_{r,a_2}$ are the re-scaled preferences for amino acids $a_1$ or $a_2$ at site $r$ as shown in Figure~\ref{fig:logoplot}.
-The WSN/1933 H1 HA amino-acid preferences are the replicate-average values reported in \cite{doud2016accurate}, re-scaled by a stringency parameter of 2.05 (see \url{https://github.com/jbloomlab/dms_tools2/blob/master/examples/Doud2016/analysis_notebook.ipynb}).
-
-\subsection*{H3N2 phylogenetic tree and maximum mutation frequencies}
-The pylogenetic tree was generated using Nextstrain's augur pipeline~\cite{Hadfield224048}, and ancestral state reconstruction and branch length timing were performed with TreeTime~\cite{Sagulenko2018}.
-Frequency trajectories of mutations were estimated following Nextstrain's augur pipeline as first implemented in Nextflu~\cite{neher2015nextflu}.
-See the SI Appendix (Supplementary text) for full details.
-
-\subsection*{Analysis of mutational shifts}
-We compared the preferences for the Perth/2009 H3 and WSN/1933 H1 HAs using the approach in~\cite{haddox2018mapping}.
-See the SI Appendix (Supplementary text) for full details.
-
-\subsection*{Validation of individual point mutants}
-To validate the viral growth of Perth/2009 HA point mutants M(-16)K, C52A, C52C, T24F, T40V, S287A, and C199(HA2)K, we examined the supernatant titers of each of these variants after reverse-genetics generation in the context of PB1flank-GFP viruses~\cite{bloom2010permissive,hooper2013mutant}.
-See the SI Appendix (Supplementary text) for full details.
-}
-
-\acknow{We thank Sarah Hilton, Hugh Haddox, and Sidney Bell for helpful discussions about data analysis, and Richard Neher for sharing analysis code and providing helpful comments on the manuscript.
-We thank the Fred Hutch Genomics Core for performing the Illumina deep sequencing.
-This work was supported by grant R01 AI127893 from the NIAID of the NIH to JDB and TB and grant U19 AI117891 to TB.
-JML was supported in part by the Center for Inference and Dynamics of Infectious Diseases (CIDID), which is funded by grant U54GM111274 from the NIGMS of the NIH.
-The research of JDB is supported in part by a Faculty Scholar grant from the Howard Hughes Medical Institute and the Simons Foundation, and a Burroughs Wellcome Young Investigator in the Pathogenesis of Infectious Diseases grant.
-TB is a Pew Biomedical Scholar and is supported by NIH R35 GM119774.
-}
-
-\showacknow{} % Display the acknowledgments section
-
-% \pnasbreak splits and balances the columns before the references.
-% Uncomment \pnasbreak to view the references in the PNAS-style
-% If you see unexpected formatting errors, try commenting out \pnasbreak
-% as it can run into problems with floats and footnotes on the final page.
-%\pnasbreak
-
-% Bibliography
-\subsection*{References}
-\bibliography{references}
-
 \end{document}
